# From Description to Inference {#description-to-inference}

```{r setup-01, include = FALSE}
source("_common.R")
library(ggplot2)
```

Almost everything in this book is an answer to one question:

> **You have looked at some of the data. What can you say about the part you
> did not look at?**

That sentence is the whole subject. Everything else — distributions, the two
great theorems, standard errors, $t$-statistics, regression — is machinery
built to answer it honestly.

## Two questions that look alike {#d2i-two-questions}

Start with something you can settle completely.

```{r d2i-class}
students <- read.csv("data/stats-class.csv")

mean(students$height, na.rm = TRUE)
```

The average height in this class is about
`r round(mean(read.csv("data/stats-class.csv")$height, na.rm = TRUE), 1)` cm.

Now: how confident are you in that number?

::: {.intuition}
Completely confident. There is nothing to be confident *about*. You measured
everyone in the class, so the average height of the class is simply a fact you
computed. There is no uncertainty, no margin of error, no $p$-value. Asking
"but is it statistically significant?" would be meaningless — significantly
different from what?

This is **descriptive statistics**: you have the whole group you care about,
and you are summarising it.
:::

Now a question that looks almost identical.

The Annual Survey of Industries records value added by registered factories in
each Indian state. Here is 2010:

```{r d2i-asi}
asi <- read.csv("data/asi-state-gva.csv")
asi_2010 <- subset(asi, yr == 2010 & !is.na(gva))

nrow(asi_2010)
mean(asi_2010$gva)
```

The ASI reports value added in ₹ lakh, so that average is about
₹`r format(round(mean(subset(read.csv("data/asi-state-gva.csv"), yr == 2010 & !is.na(gva))$gva) / 100), big.mark = ",")`
crore per state.

```{r d2i-crore}
# 1 lakh = 100,000;  1 crore = 100 lakh = 10,000,000
asi_2010$gva_crore <- asi_2010$gva / 100

summary(asi_2010$gva_crore)
```

Same arithmetic. Completely different epistemic situation — for two separate
reasons, and it is worth being precise about which is which.

::: {.watchout}
**Reason one: the ASI is a survey, not a census.** It does not visit every
factory. It samples them. So the state figures are estimates, each carrying
sampling error, before we compute anything at all.

**Reason two, and the one that matters more here: what do you want the number
*for*?** If you only want to describe the 32 states in 2010, the mean above is
a fact, like the class average. But nobody wants that. They want to say
something about *Indian industry* — including 2011, including unregistered
firms, including what would happen under a different policy. The moment you
want that, you have left description behind.
:::

**Inferential statistics** is the discipline of making that leap and being
honest about how far you have jumped.

::: {.keyidea}
Descriptive statistics summarises the data you have.

Inferential statistics uses the data you have to make claims about data you do
not have — and, crucially, attaches a statement of how uncertain those claims
are.

A number without an honest account of its uncertainty is not a finding. It is
an anecdote with decimal places.
:::

## The population you cannot see {#d2i-population}

The leap needs vocabulary.

The **population** is the entire collection you want to know about. Every
registered factory in India. Every agricultural labourer. Every household that
might be offered a job guarantee scheme.

The **sample** is what you actually observe.

A **parameter** is a number describing the population — the true mean $\mu$,
the true standard deviation $\sigma$. You never see these. Write them as Greek
letters to remember they are permanently out of reach.

A **statistic** is a number computed from your sample — $\bar{x}$, $s$. You can
always see these. They are your only evidence.

::: {.keyidea}
The entire problem of inference in one line:

$$\underbrace{\bar{x}}_{\text{you see this}} \quad\longrightarrow\quad \underbrace{\mu}_{\text{you want this}}$$

The arrow is not equality. Making that arrow trustworthy — and quantifying how
much to trust it — is what the rest of this book is about.
:::

Why can we say anything at all? Because a sample drawn properly from a
population carries information about it. Not perfect information. But
information whose reliability we can *calculate*, which turns out to be enough.

Two theorems make the calculation possible. The **Law of Large Numbers**
says that as samples grow, the sample
mean closes in on the population mean. The **Central Limit Theorem** says
something more surprising: it tells us the *shape* of the error, and that shape
turns out to be the same no matter what the population looks like. Together
they convert "we are probably close" into a number.

Before we get there, we need distributions — the objects that describe
populations in the first place.

## The average is not the distribution {#d2i-average}

Here is a habit worth forming now, before any technique gets in the way.

Almost all of inferential statistics is about averages. Nearly every method in
this book estimates a mean, compares two means, or fits a line through
conditional means. That focus is not an accident: means are well behaved,
they have tractable sampling distributions, and the two theorems above are
statements about them.

But the mean is only *one* summary of a distribution, and often not the one a
policy question actually cares about.

Look again at industrial value added across states:

```{r d2i-skew}
c(mean   = mean(asi_2010$gva_crore),
  median = median(asi_2010$gva_crore))

mean(asi_2010$gva_crore) / median(asi_2010$gva_crore)
```

The mean is more than **twice** the median. When those two diverge that
sharply, the average has stopped describing a typical state.

```{r d2i-skew-plot, echo = FALSE, fig.height = 4.2, fig.cap = "Industrial value added across Indian states, 2010. The mean sits far to the right of most states, dragged there by a handful of large ones."}
ggplot(asi_2010, aes(gva_crore)) +
  geom_histogram(bins = 22, fill = book_palette$fill, colour = "white",
                 linewidth = 0.25) +
  geom_vline(xintercept = median(asi_2010$gva_crore),
             colour = book_palette$ink, linetype = "dashed", linewidth = 0.7) +
  geom_vline(xintercept = mean(asi_2010$gva_crore),
             colour = book_palette$reject, linewidth = 0.9) +
  annotate("label", x = mean(asi_2010$gva_crore), y = Inf, vjust = 1.4,
           label = "mean", colour = book_palette$reject, fill = "white",
           size = 3.5, label.size = 0.3) +
  annotate("label", x = median(asi_2010$gva_crore), y = Inf, vjust = 3.2,
           label = "median", colour = book_palette$ink, fill = "white",
           size = 3.5, label.size = 0.3) +
  scale_x_continuous(labels = scales::comma) +
  labs(x = "gross value added (₹ crore)", y = "number of states") +
  theme_book()
```

How concentrated is it really?

```{r d2i-concentration}
top5 <- sort(asi_2010$gva_crore, decreasing = TRUE)[1:5]

round(100 * sum(top5) / sum(asi_2010$gva_crore))   # % of all value added
round(max(asi_2010$gva_crore) / median(asi_2010$gva_crore), 1)   # top state vs median
```

Five states hold about 59% of registered industrial value added. The largest
produces roughly thirteen times the median state.

::: {.intuition}
Now ask what "average industrial value added across states" is good for.

If you are asking how much industry India has in total, it is fine — the mean
times the number of states is the total, by construction.

If you are asking what industry looks like in a typical Indian state, it is
badly misleading. No state is near the mean. The distribution is two different
countries: a handful of industrial giants and a long tail of states doing
something else entirely.

If you are asking whether industrial growth is spreading, the mean can rise
while every state but one falls further behind. **The average can move in the
opposite direction to almost everybody in it.**
:::

::: {.keyidea}
Learn to ask two questions of every average you meet, including the ones you
compute yourself:

1. **What is the spread?** A mean with no measure of dispersion is half a
   sentence.
2. **Who is where?** Means are useless for questions about the bottom of a
   distribution — poverty, vulnerability, who is hurt by a shock. Those are
   questions about *tails*, and a mean is specifically the statistic that
   averages tails away.
:::

This is not a philosophical aside; it changes what you estimate. Whole
literatures in development economics exist because the mean was the wrong
target: poverty measures weight the poor rather than averaging over everyone,
vulnerability measures ask who is exposed to falling below a line, and
distributional analyses ask whether a policy that raised average income raised
*anyone's* income who was previously badly off.

We will come back to this repeatedly. When we fit a regression, we will be
modelling a conditional *mean*, and it will be worth asking each time what that
choice hides. When we get to binary explanatory variables in
the unit on binary explanatory variables, we will estimate an average
treatment effect, and the
word "average" in that phrase will deserve every bit of the suspicion you can
give it.

::: {.watchout}
None of this means averages are bad. They are the right tool for an enormous
number of questions, and this book spends most of its length on them for good
reason.

It means the average is an *answer to a specific question*, not a neutral
summary of the data. Know which question you are asking before you decide the
mean is the answer.
:::

## What can go wrong before any mathematics {#d2i-what-goes-wrong}

Inference assumes your sample carries information about the population you care
about. Three ways that assumption fails, none of which any formula will warn
you about.

**The sample is not from the population you mean.** The ASI covers
*registered* factories. Most Indian manufacturing employment is unregistered.
Conclusions about "Indian manufacturing" from ASI data are conclusions about a
formal sliver of it. The arithmetic will not object.

**The sample is not random.** Our class survey has 32 heights. They are not a
random sample of APU students, let alone of young Indians — they are the people
who took this course and turned up that day. You may compute a confidence
interval from them, and it will mean nothing about the university.

**The question is causal but the data is not.** Suppose states with more
industry have higher literacy. Does industry raise literacy? Does literacy
attract industry? Does something else drive both? The correlation is the same
in all three cases. No amount of statistical machinery distinguishes them
without an argument about *how the data came to be*.

::: {.intuition}
This last one is the reason this book is organised the way it is. From
the unit on two-sample tests onwards, we stop asking only "what is this
number?" and
start asking "what would have to be true for this number to mean what I want it
to mean?"

That question — not any particular estimator — is what separates a good applied
economist from someone who can run regressions.
:::

## What you should take from this unit {#d2i-summary}

::: {.keyidea}
1. **Description** summarises data you have. **Inference** reaches beyond it,
   and must report how far it has reached.
2. Greek letters ($\mu$, $\sigma$) are population parameters you never observe.
   Roman letters ($\bar{x}$, $s$) are sample statistics you compute. Inference
   is the bridge.
3. The bridge is built from two theorems, the Law of Large Numbers and the
   Central Limit Theorem, and it needs distributions first.
4. **The mean is one feature of a distribution, not a summary of it.** Indian
   industrial value added has a mean twice its median; five states hold 59% of
   it. Always ask about spread, and about who sits in the tails.
5. No formula can rescue a sample drawn from the wrong population, drawn
   non-randomly, or asked a causal question it cannot answer.
:::

Next we look at the distributions themselves — binomial, Poisson, normal, and
the difference between discrete and continuous. They are how we describe a
population. What they do *not* do, on their own, is let us infer anything from
a sample — and seeing exactly why is what makes the two theorems feel necessary
rather than arbitrary.
