# From Description to Inference {#description-to-inference}

```{r setup-01, include = FALSE}
source("_common.R")
library(ggplot2)
```

What do a gambler betting on dice, a doctor reading a patient's blood pressure,
and a pollster forecasting an election have in common?

All three are asking the same question: **what are the possible outcomes, and
how likely is each one?** The gambler wants to know how often a seven comes up.
The doctor wants to know whether this reading is unusual for a man of this age.
The pollster wants to know how far a survey of two thousand voters can be
trusted about ninety crore of them.

None of them can observe everything they care about. Each has a fragment, and
each must say something about the whole. That is the problem this book exists to
solve, and this unit is about why it is hard.

## What we can already do {#d2i-descriptive}

Start with what is already familiar.

Given a set of numbers, we know how to summarise them. We can ask where they
are **centred** — the mean, the median, the mode. We can ask how **spread out**
they are — the variance, the standard deviation, the range. In the language of
distributions, the first describes the **first moment** and the second the
**second moment**: where the mass sits, and how far it is scattered.

We also know how to *look* at data rather than only compute from it. A
histogram shows the frequency of numerical values and reveals symmetry,
skewness and outliers at a glance. A bar chart does the same for a discrete
variable. For a continuous variable, a smooth density curve captures the
overall shape.

Here is all of that on Indian data. The Annual Survey of Industries records
value added by registered factories in each state:

```{r d2i-desc}
asi <- read.csv("data/asi-state-gva.csv")
asi_2010 <- subset(asi, yr == 2010 & !is.na(gva))

# ASI reports value added in rupees lakh; 100 lakh = 1 crore
asi_2010$gva_crore <- asi_2010$gva / 100

c(states = nrow(asi_2010),
  mean   = mean(asi_2010$gva_crore),
  median = median(asi_2010$gva_crore),
  sd     = sd(asi_2010$gva_crore))
```

Nothing here is mysterious. We have taken `r nrow(subset(read.csv("data/asi-state-gva.csv"), yr == 2010 & !is.na(gva)))`
numbers and replaced them with four. That is **descriptive statistics**: the
art of summarising the data you have so that a human can hold it in mind.

::: {.keyidea}
Descriptive statistics is genuinely useful and genuinely limited.

Useful, because it is always the first thing you should do with data. Trends,
outliers, data-entry errors and interesting patterns all show up here, and no
amount of later sophistication rescues an analysis that skipped this step.

Limited, because **every statement it makes is about the data in front of
you** — and that is almost never what you actually want to know.
:::

## Where description runs out {#d2i-two-questions}

To see the limit sharply, compare two questions that look identical.

**First.** What is the average height in this class?

```{r d2i-class}
students <- read.csv("data/stats-class.csv")

mean(students$height, na.rm = TRUE)
```

About `r round(mean(read.csv("data/stats-class.csv")$height, na.rm = TRUE), 1)` cm.

Now: how confident are you in that number?

::: {.intuition}
Completely confident — and the question is almost meaningless. You measured
everyone. The average height of this class is not an estimate of anything; it
is a fact you computed. There is no margin of error. Asking "is it
statistically significant?" would be incoherent: significantly different from
what?
:::

**Second.** What is the average industrial value added of an Indian state?

```{r d2i-asi}
mean(asi_2010$gva_crore)
```

Identical arithmetic. Completely different situation — for two distinct
reasons, and it is worth separating them.

::: {.watchout}
**The ASI is a survey, not a census.** It does not visit every factory; it
samples them. Each state figure is already an estimate carrying sampling error,
before we compute anything.

**And what do you want the number *for*?** If you genuinely care only about
those 32 states in that one year, the mean is a fact, like the class average.
But nobody wants that. They want to say something about *Indian industry* —
next year too, unregistered firms too, under a different policy too. The moment
you want that, you have left description behind.
:::

**Inferential statistics** is the discipline of making that leap and being
honest about how far you have jumped.

::: {.keyidea}
Descriptive statistics summarises the data you have.

Inferential statistics uses the data you have to make claims about data you do
not have — and attaches a statement of how uncertain those claims are.

A number without an honest account of its uncertainty is not a finding. It is
an anecdote with decimal places.
:::

Before we can make the leap, we must be clear about what we are leaping
between. That requires the idea of a distribution.

## What a distribution actually is {#d2i-distribution}

A **distribution** is a pattern that emerges from repetition under randomness.
It tells you which outcomes are common, which are rare, and where values tend
to concentrate.

The key word is *repetition*. A single observation has no distribution. Roll a
die once and you get a four; that tells you almost nothing. To see the pattern
you must roll again and again, recording outcomes, until the shape appears.

```{r d2i-die}
set.seed(1)
rolls <- sample(1:6, size = 10000, replace = TRUE)

round(prop.table(table(rolls)), 3)
```

```{r d2i-die-plot, echo = FALSE, fig.height = 3.4, fig.cap = "Ten thousand rolls of a fair die. Each face turns up close to one-sixth of the time."}
ggplot(data.frame(x = factor(rolls)), aes(x)) +
  geom_bar(aes(y = after_stat(count / sum(count))),
           fill = book_palette$fill, width = 0.7) +
  geom_hline(yintercept = 1/6, colour = book_palette$reject,
             linetype = "dashed", linewidth = 0.7) +
  annotate("text", x = 6.4, y = 1/6, label = "1/6", vjust = -0.8,
           colour = book_palette$reject, size = 3.6) +
  labs(x = "outcome", y = "relative frequency") +
  theme_book()
```

Each face appears roughly one-sixth of the time. That pattern *is* the
distribution of a fair die, and it became visible only through repetition.

### Randomness is not optional {#d2i-degenerate}

Now take the randomness away. Imagine a die with a 3 on every face.

```{r d2i-degenerate}
rolls_fixed <- rep(3, 10000)

table(rolls_fixed)
sd(rolls_fixed)
```

Every observation is identical. No spread, no uncertainty, standard deviation
exactly zero. The distribution has collapsed onto a single point — what is
called a **degenerate** distribution.

::: {.keyidea}
**A distribution is not a property of a single observation. It is a property of
repeated observations generated by a random process.**

Where there is no randomness there is nothing to describe, nothing to be
uncertain about, and no statistics to do. Every technique in this book is a way
of handling variation. Remove the variation and the entire subject falls
silent.
:::

This is why statistics is built on **random variables** — quantities whose
value depends on the outcome of a random process. Height, income, rainfall, the
number on a die: all vary, and it is precisely their variation that we model.

### Expectation as a long-run average {#d2i-expectation}

If a random variable takes value $x_i$ with probability $p(x_i)$, its
**expected value** is

$$E[X] = \sum_i x_i \, p(x_i)$$

The name invites a misunderstanding. $E[X]$ is not the value you expect on any
given try — for a fair die it is 3.5, a number the die cannot produce. It is
the **long-run average** over many repetitions.

```{r d2i-expectation}
mean(rolls)                # average over 10,000 actual rolls
sum(1:6 * rep(1/6, 6))     # theoretical E[X]
```

::: {.intuition}
That the two agree is neither a coincidence nor trivial. It is the first
appearance of the Law of Large Numbers, which we will meet properly later: run
a random process enough times and the average of what you observe closes in on
the expectation you calculated on paper.

Notice how unremarkable that felt. We are about to need it for something much
less obvious.
:::

## Population, sample, and the assumption that licenses everything {#d2i-population}

Now the vocabulary of the leap.

The **population** is the entire collection you want to know about. Every
registered factory in India. Every agricultural labourer. Every household
eligible for a job guarantee scheme.

The **sample** is the part you actually observe.

A **parameter** is a number describing the population — the true mean $\mu$,
the true standard deviation $\sigma$. You never see these. They are written as
Greek letters as a permanent reminder that they are out of reach.

A **statistic** is a number computed from your sample — $\bar{x}$, $s$. You can
always see these. They are your only evidence.

::: {.keyidea}
The whole problem of inference, in one line:

$$\underbrace{\bar{x}}_{\text{you have this}} \quad\longrightarrow\quad \underbrace{\mu}_{\text{you want this}}$$

The arrow is not an equals sign. Making it trustworthy, and saying precisely
how much to trust it, is what the rest of this book is for.
:::

Why should a sample tell us anything at all about a population? Because of an
assumption we make — and it is worth stating explicitly, because everything
rests on it.

We assume the population has an underlying probability distribution, and that
the members of our sample are **independent and identically distributed**
(i.i.d.) draws from it.

- **Identically distributed**: every observation comes from the same
  distribution. No drift, no changing the rules partway through.
- **Independent**: knowing one observation tells you nothing about another.

::: {.keyidea}
It is this assumption — that the sample is drawn from the same distribution as
the population — that licenses inference at all.

If the sample came from a *different* distribution than the population, no
amount of arithmetic could bridge them. This is why sampling design matters so
much, and why a badly drawn sample cannot be repaired by clever statistics
afterwards. The formulas will still run. They will simply be answering a
question about a population you did not mean.
:::

## The pivot: distributions of statistics {#d2i-sampling-distribution}

Here is the conceptual step that makes inference possible, and it is the most
important idea in this unit.

Everything so far concerned the distribution of **raw observations**: repeat an
experiment, record individual outcomes, watch a pattern form. Now we shift
attention from raw observations to **summaries of observations**.

Instead of recording one die roll at a time, roll the die thirty times,
calculate the average, and record *only that average*. Then do the whole thing
again: a fresh thirty rolls, a fresh average. And again, hundreds of times.

The pattern formed by those averages is a **sampling distribution**.

::: {.keyidea}
The distribution of a **random variable** describes how raw observations vary.

A **sampling distribution** describes how a *statistic* — a sample mean, a
proportion, a standard deviation — varies across repeated samples of a fixed
size.

It answers the question inference actually turns on: *if I took another sample,
how different would my answer be?*
:::

### A worked example: a deck of cards {#d2i-cards}

Take a standard deck as the population. Score the cards A = 1, J = 11, Q = 12,
K = 13, and 2 through 10 at face value. Because we built this population
ourselves, we know its parameters exactly — a luxury we will never have again.

```{r d2i-deck}
deck <- rep(1:13, times = 4)

c(N = length(deck), mu = mean(deck), sigma = sd(deck))
```

The population mean is exactly $\mu = 7$.

Now draw a sample of 10 cards and take its mean:

```{r d2i-onesample}
set.seed(7)
mean(sample(deck, size = 10, replace = TRUE))
```

::: {.watchout}
Note `replace = TRUE`: each card is returned to the deck before the next draw.

That is not a technicality. It is what makes the draws **independent** — the
i.i.d. assumption from the previous section. Draw *without* replacement and
each card removed changes what remains, so the draws are dependent and the
sample mean varies slightly less than the theory predicts. (The adjustment is
called the finite population correction, and you can see it at work in the
exercises.)

Sampling with replacement is also the better model of what we usually mean.
When you survey 2,000 Indians you are not meaningfully depleting a population
of a billion.
:::

Not 7. Draw again and you would get something else again. **The sample mean is
itself a random variable** — it has its own distribution, because it changes
every time you draw.

So let us look at that distribution. Repeat the exercise a thousand times:

```{r d2i-sampdist}
set.seed(7)
sample_means <- replicate(1000, mean(sample(deck, size = 10, replace = TRUE)))

c(mean_of_sample_means = mean(sample_means),
  sd_of_sample_means   = sd(sample_means))
```

```{r d2i-sampdist-plot, echo = FALSE, fig.height = 3.8, fig.cap = "One thousand sample means, each from 10 cards. Individual samples miss the population mean constantly. Their average lands almost exactly on it."}
ggplot(data.frame(m = sample_means), aes(m)) +
  geom_histogram(bins = 32, fill = book_palette$fill, colour = "white",
                 linewidth = 0.2) +
  geom_vline(xintercept = 7, colour = book_palette$reject, linewidth = 0.9) +
  annotate("label", x = 7, y = Inf, vjust = 1.4, label = "population mean = 7",
           colour = book_palette$reject, fill = "white", size = 3.5,
           label.size = 0.3) +
  labs(x = "sample mean of 10 cards", y = "how often") +
  theme_book()
```

Two things happened, and both matter.

**The sample means are centred on the population mean.** No single sample got
it right, but their average is `r round(mean(sample_means), 3)` — essentially 7.
This is the result $E(\bar{X}) = \mu$: the sample mean is an **unbiased**
estimator. It is not right, but it is not *systematically* wrong, which is a
different and far more useful property.

**The sample means vary much less than the cards do.** The population standard
deviation is about `r round(sd(rep(1:13, 4)), 2)`; the sample means have a
standard deviation of about `r round(sd(sample_means), 2)`. Averaging has
cancelled out much of the noise.

### What happens as the sample grows {#d2i-samplesize}

That second observation is worth pushing on. What if we take bigger samples?

```{r d2i-nsizes}
set.seed(7)

sizes <- c(5, 10, 20, 40, 80)
results <- t(sapply(sizes, function(n) {
  m <- replicate(2000, mean(sample(deck, size = n, replace = TRUE)))
  c(n = n, mean = mean(m), sd = sd(m), theoretical_sd = sd(deck) / sqrt(n))
}))

round(results, 3)
```

```{r d2i-nsizes-plot, echo = FALSE, fig.height = 4.4, fig.cap = "Sampling distributions of the mean for growing sample sizes. The centre never moves. The spread collapses."}
set.seed(7)
dd <- do.call(rbind, lapply(c(5, 20, 80), function(n) {
  data.frame(n = paste("n =", n),
             m = replicate(2000, mean(sample(deck, n, replace = TRUE))))
}))
dd$n <- factor(dd$n, levels = paste("n =", c(5, 20, 80)))

ggplot(dd, aes(m)) +
  geom_histogram(bins = 40, fill = book_palette$fill, colour = "white",
                 linewidth = 0.15) +
  geom_vline(xintercept = 7, colour = book_palette$reject, linewidth = 0.8) +
  facet_wrap(~ n, ncol = 1, scales = "free_y") +
  labs(x = "sample mean", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank())
```

Every row is centred on 7 — the sample mean is unbiased at any sample size. But
the spread shrinks steadily, and in a completely predictable way. Compare the
last two columns: the observed standard deviation of the sample means matches
$\sigma/\sqrt{n}$ almost exactly.

::: {.keyidea}
Two results, which we will prove in a later unit:

$$E(\bar{X}) = \mu \qquad\qquad \mathrm{Var}(\bar{X}) = \frac{\sigma^2}{n}$$

The sample mean is centred on the population mean, and its spread falls as the
sample grows. The quantity $\sigma/\sqrt{n}$ is the **standard error**, and it
is the single most important number in applied statistics: it is how wrong you
should expect to be.
:::

::: {.watchout}
Note the square root. To halve your standard error you must **quadruple** your
sample. Precision is expensive, and it grows more expensive the more of it you
already have.

This is why surveys of two thousand people are common and surveys of two lakh
people are not: the second costs a hundred times as much for about seven times
the precision.
:::

::: {.intuition}
Look again at what just happened, because it is genuinely remarkable.

We began knowing nothing about how wrong a single sample would be. By studying
how the sample mean behaves *across many hypothetical samples*, we worked out
exactly how much it varies — and obtained a formula, $\sigma/\sqrt{n}$, that
can be computed from a single real sample.

In practice you only ever draw one sample. The sampling distribution tells you
what would have happened had you drawn many, and that is enough to attach an
honest margin of error to the one you have.
:::

### The same logic on Indian data {#d2i-asi-sampling}

Cards are a clean laboratory. Real data is messier, and the ASI figures let us
watch the same logic operate on something considerably less well behaved.

Treat the 32 state values as a population and sample from it:

```{r d2i-asi-sampling}
set.seed(11)

pop <- asi_2010$gva_crore
asi_means <- replicate(2000, mean(sample(pop, size = 10, replace = TRUE)))

c(population_mean      = mean(pop),
  mean_of_sample_means = mean(asi_means),
  population_sd        = sd(pop),
  se_of_sample_means   = sd(asi_means))
```

Once again the sample means centre on the population mean, and once again they
vary far less than the underlying data. The machinery does not care that
industrial value added looks nothing like a deck of cards.

Which brings us to the feature of this data we have so far ignored.

## The average is not the distribution {#d2i-average}

A habit worth forming now, before any technique gets in the way.

Almost all of inferential statistics is about averages. Nearly every method in
this book estimates a mean, compares two means, or fits a line through
conditional means. That focus is not arbitrary — means are well behaved, their
sampling distributions are tractable, and the two great theorems are statements
about them.

But the mean is only **one** feature of a distribution, and often not the one a
policy question cares about.

```{r d2i-skew}
c(mean   = mean(asi_2010$gva_crore),
  median = median(asi_2010$gva_crore))

mean(asi_2010$gva_crore) / median(asi_2010$gva_crore)
```

The mean is more than **twice** the median. When those two diverge that
sharply, the average has stopped describing a typical state.

```{r d2i-skew-plot, echo = FALSE, fig.height = 4.2, fig.cap = "Industrial value added across Indian states, 2010. The mean sits far to the right of most states, dragged there by a handful of large ones."}
ggplot(asi_2010, aes(gva_crore)) +
  geom_histogram(bins = 22, fill = book_palette$fill, colour = "white",
                 linewidth = 0.25) +
  geom_vline(xintercept = median(asi_2010$gva_crore),
             colour = book_palette$ink, linetype = "dashed", linewidth = 0.7) +
  geom_vline(xintercept = mean(asi_2010$gva_crore),
             colour = book_palette$reject, linewidth = 0.9) +
  annotate("label", x = mean(asi_2010$gva_crore), y = Inf, vjust = 1.4,
           label = "mean", colour = book_palette$reject, fill = "white",
           size = 3.5, label.size = 0.3) +
  annotate("label", x = median(asi_2010$gva_crore), y = Inf, vjust = 3.2,
           label = "median", colour = book_palette$ink, fill = "white",
           size = 3.5, label.size = 0.3) +
  scale_x_continuous(labels = scales::comma) +
  labs(x = "gross value added (₹ crore)", y = "number of states") +
  theme_book()
```

How concentrated is it?

```{r d2i-concentration}
top5 <- sort(asi_2010$gva_crore, decreasing = TRUE)[1:5]

round(100 * sum(top5) / sum(asi_2010$gva_crore))                 # % of all GVA
round(max(asi_2010$gva_crore) / median(asi_2010$gva_crore), 1)   # top vs median
```

Five states hold about 59% of registered industrial value added. The largest
produces roughly thirteen times the median state.

::: {.intuition}
Now ask what "average industrial value added across states" is good for.

Asking how much industry India has in total? Fine — the mean times the number
of states is the total, by construction.

Asking what industry looks like in a typical Indian state? Badly misleading. No
state sits near the mean. The distribution is really two countries: a few
industrial giants and a long tail of states doing something else.

Asking whether industrial growth is spreading? The mean can rise while every
state but one falls further behind. **An average can move in the opposite
direction to almost everybody inside it.**
:::

::: {.keyidea}
Ask two questions of every average you meet, including your own:

1. **What is the spread?** A mean with no measure of dispersion is half a
   sentence.
2. **Who is where?** Means are the wrong tool for questions about the bottom of
   a distribution — poverty, vulnerability, who is hurt by a shock. Those are
   questions about *tails*, and the mean is precisely the statistic that
   averages tails away.
:::

This is not a philosophical aside; it changes what you estimate. Whole
literatures in development economics exist because the mean was the wrong
target. Poverty measures weight the poor rather than averaging over everyone.
Vulnerability measures ask who is at risk of falling below a line.
Distributional analysis asks whether a policy that raised average income raised
the income of anyone who was badly off to begin with.

We will return to this repeatedly. When we fit a regression we will be
modelling a conditional *mean*, and it will be worth asking each time what that
choice conceals. When we reach binary explanatory variables and estimate an
average treatment effect, the word "average" in that phrase will deserve every
bit of suspicion you can bring to it.

::: {.watchout}
None of this makes averages bad. They are the right tool for a great many
questions, and this book spends most of its length on them for good reason.

It means an average is an **answer to a specific question**, not a neutral
summary of the data. Decide which question you are asking before deciding the
mean answers it.
:::

## What can go wrong before any mathematics {#d2i-what-goes-wrong}

Inference assumes your sample carries information about the population you care
about. Three ways that fails, none of which any formula will warn you about.

**The sample is not from the population you mean.** The ASI covers *registered*
factories. Most Indian manufacturing employment is unregistered. Conclusions
about "Indian manufacturing" drawn from ASI data are conclusions about a formal
sliver of it. The arithmetic will not object.

**The sample is not random.** Our class survey has 32 heights. They are not a
random sample of APU students, let alone of young Indians — they are the people
who took this course and turned up that day. You may compute a confidence
interval from them and it will mean nothing about the university.

**The question is causal but the data is not.** Suppose states with more
industry have higher literacy. Does industry raise literacy? Does literacy
attract industry? Does something else drive both? The correlation is identical
in all three cases, and no statistical machinery separates them without an
argument about *how the data came to exist*.

::: {.intuition}
That last failure is why this book is organised as it is. From the unit on
two-sample tests onwards, we stop asking only "what is this number?" and start
asking "what would have to be true for this number to mean what I want it to
mean?"

That question — not any particular estimator — is what separates a good applied
economist from someone who can run regressions.
:::

## Summary {#d2i-summary}

::: {.keyidea}
1. **Description** summarises data you have. **Inference** reaches beyond it and
   must report how far it has reached.
2. A **distribution** is a pattern emerging from repetition under randomness.
   No randomness, no distribution, no statistics.
3. Greek letters ($\mu$, $\sigma$) are population parameters you never observe.
   Roman letters ($\bar{x}$, $s$) are sample statistics you compute.
4. Inference is licensed by the **i.i.d. assumption** — that your sample is
   drawn from the same distribution as the population. A badly drawn sample
   cannot be repaired later.
5. A **sampling distribution** describes how a statistic varies across repeated
   samples. It is the pivot on which all inference turns.
6. $E(\bar{X}) = \mu$ and $\mathrm{Var}(\bar{X}) = \sigma^2/n$. The sample mean
   is unbiased, and its **standard error** $\sigma/\sqrt{n}$ falls with the
   square root of the sample size.
7. **The mean is one feature of a distribution, not a summary of it.** Indian
   industrial value added has a mean twice its median, with five states holding
   59% of the total. Always ask about spread, and about the tails.
:::

We have watched sample means centre on the population mean and tighten as
samples grow. We have not yet *proved* either claim, and we have said nothing
about the **shape** of a sampling distribution.

Those are the next steps. The distributions unit supplies the vocabulary —
binomial, Poisson, normal, discrete against continuous — for describing
populations in the first place. Then the **Law of Large Numbers** proves the
convergence we just watched happen, and the **Central Limit Theorem** delivers
the genuinely surprising result: the shape of a sampling distribution is
approximately normal *whatever the population looks like*. Not the deck of
cards, not the ASI, not anything. That is what turns "we are probably close"
into a number.
