# Towards Inferential Statistics {#towards-inference}

```{r setup-01, include = FALSE}
source("_common.R")
library(ggplot2)
```

You already know how to calculate the average of a sample.

In this unit, you will learn why that average is almost certainly wrong.

At first glance, this seems like a strange claim. After all, there is only one
way to calculate a sample mean, and if the arithmetic is correct, the answer
must be correct as well. Yet the number you obtain from a sample is almost
never exactly equal to the quantity you ultimately wish to know — the mean of
the population from which that sample was drawn.

Suppose you wish to estimate the average income of households in India. Since
surveying every household is impossible, you select a sample of 2,000
households and calculate its mean income. The calculation is straightforward,
but the result immediately raises a question. How close is this sample mean to
the true population mean?

The difficulty is that the sample you have drawn is only one of many that could
have been observed. Had a different 2,000 households been selected, the sample
mean would almost certainly have been different. Draw another sample, and the
answer changes again. Every statistic calculated from a sample therefore
reflects two things at once: the characteristics of the underlying population,
and the randomness introduced by the sampling process.

This simple observation marks the boundary between descriptive and inferential
statistics.

Until now, our objective has been to describe the data that we observe.
Measures of central tendency, such as the mean and the median, summarise where
observations are concentrated. Measures of dispersion, such as the variance and
standard deviation, describe how widely they are spread. Histograms, bar charts
and density plots reveal the overall shape of a distribution and help us
identify skewness, outliers and other important features. Together, these tools
provide a concise description of a dataset and are the natural starting point
of every empirical investigation.

Most empirical research, however, is concerned not with the sample itself but
with the population from which it was drawn. Economists survey a few thousand
households to understand millions. Polling agencies interview a sample of
voters to estimate national voting intentions. Medical researchers evaluate new
treatments on a limited number of patients before recommending them to the
wider population. In every case, the sample serves as evidence about a
population that cannot be observed directly.

The central problem of inferential statistics is therefore straightforward to
state: **how can we use the information contained in a sample to learn about
the population from which it came?**

Throughout this course we adopt the **frequentist** approach to answering this
question. Rather than treating the observed sample as unique, we imagine
drawing many samples from the same population and calculating the same
statistic repeatedly. Some sample means will be larger than others, some
smaller. Understanding how statistics vary across these hypothetical samples
allows us to assess how much confidence we should place in the estimate
computed from the one sample we actually observe.

Before we can understand why, we need to understand where the uncertainty comes
from. The answer lies in randomness.

## Randomness and variation {#u1-randomness}

::: {.question}
Where does the uncertainty come from?
:::

The uncertainty that surrounds every statistical estimate has a single source:
randomness.

If we were to repeat the same study with a different sample, we would almost
certainly obtain a different average. Inferential statistics is concerned with
understanding this variability. Without randomness, every sample would be
identical, every estimate would equal the population value, and there would be
no uncertainty to quantify.

The idea is easiest to see with a die.

Suppose we roll a fair die once.

```{r u1-die-one}
set.seed(1)
sample(1:6, size = 1)
```

The outcome is a one. Beyond that, there is little to say. A single observation
tells us almost nothing about the process that generated it. Was the die fair?
Would the next roll also be a one? Is one an unusually common outcome? One
observation cannot answer these questions.

Now imagine repeating the experiment ten thousand times.

```{r u1-die-many}
rolls <- sample(1:6, size = 10000, replace = TRUE)

round(prop.table(table(rolls)), 3)
```

Each individual roll remains unpredictable, but something remarkable happens
collectively. The outcomes no longer appear as isolated events. They form a
stable pattern. Every face occurs with almost the same frequency, close to
one-sixth of the time, exactly as we would expect from a fair die.

```{r u1-die-plot, echo = FALSE, fig.height = 3.2, fig.cap = "Ten thousand rolls of a fair die. No single roll was predictable. The pattern across all of them is."}
ggplot(data.frame(x = factor(rolls)), aes(x)) +
  geom_bar(aes(y = after_stat(count / sum(count))),
           fill = book_palette$fill, width = 0.7) +
  geom_hline(yintercept = 1/6, colour = book_palette$reject,
             linetype = "dashed", linewidth = 0.7) +
  annotate("text", x = 6.4, y = 1/6, label = "1/6", vjust = -0.8,
           colour = book_palette$reject, size = 3.6) +
  labs(x = "outcome", y = "share of rolls") +
  theme_book()
```

The important point is not that the die turned out to be fair. It is that a
pattern exists at all.

To see why that matters, imagine removing the randomness altogether.
Suppose every face of the die is labelled three.

```{r u1-die-loaded}
loaded <- rep(3, 10000)

table(loaded)
sd(loaded)
```

Every roll now produces exactly the same outcome. There is no variation, no
uncertainty and no pattern waiting to emerge through repetition. The average is
always three. The standard deviation is zero. Repeating the experiment a
million times teaches us nothing that we did not already know after the first
roll.

The contrast between the two dice illustrates a fundamental idea. Randomness
makes individual outcomes unpredictable, but repeated observations reveal
stable patterns. Statistics is built on this coexistence of uncertainty and
regularity. If outcomes never varied, there would be no need for inference; if
they varied without any underlying regularity, inference would be impossible.

Dice provide a simple illustration, but the same principle governs every
empirical investigation. Household surveys interview different families.
Election polls reach different voters. Clinical trials recruit different
patients. Agricultural experiments observe different harvests. Every study
could have produced a different dataset. The central challenge of inferential
statistics is to understand what can be learned from **one realisation of a
process that could have produced many others**.

## Describing variation {#u1-describing}

::: {.question}
Once variation exists, how do we describe it?
:::

Variation is what makes statistics necessary. The next question is equally
natural: how do we describe it?

A die has only six possible outcomes, so its behaviour is easy to summarise.
Real datasets are rather different. Consider the annual earnings of twenty
thousand Indian workers.

```{r u1-earnings}
workers <- read.csv("data/wages-india-synthetic.csv")

nrow(workers)

head(workers$annual_earnings)
```

This dataset is simulated, but it closely resembles the earnings distribution
observed in large household surveys such as the India Human Development Survey
(IHDS). Throughout this book, we will use it to illustrate statistical ideas
and methods.

Twenty thousand numbers are too many for anyone to understand by inspection. We
need ways of reducing the data without losing its essential features. The first
step is simply to look at it.

```{r u1-earnings-hist, fig.height = 3.6, fig.cap = "Annual earnings of 20,000 Indian workers."}
ggplot(workers, aes(annual_earnings / 1000)) +
  geom_histogram(bins = 60, fill = book_palette$fill, colour = "white",
                 linewidth = 0.15) +
  scale_x_continuous(labels = scales::comma) +
  labs(x = "Annual earnings (₹ thousand)",
       y = "Number of workers") +
  theme_book()
```

The histogram immediately reveals two important features. Most workers earn
relatively modest incomes, while a much smaller number earn substantially more.
The distribution is therefore concentrated on the left with a long tail
extending to the right — a pattern known as **right skewness**.

One picture already tells us a great deal, but we can describe the same
distribution more precisely using a few numerical summaries.

```{r u1-centre}
c(mean   = mean(workers$annual_earnings),
  median = median(workers$annual_earnings))
```

The mean annual earnings are more than twice the median. This difference
reflects the long right tail visible in the histogram. A relatively small
number of high earners pull the mean upwards, whereas the median simply divides
the distribution into two equal halves and is much less affected by extreme
values.

Variation also concerns how widely observations are dispersed.

```{r u1-spread}
c(sd  = sd(workers$annual_earnings),
  iqr = IQR(workers$annual_earnings))
```

The **standard deviation** measures the typical distance of observations from
the mean, while the **interquartile range** measures the spread of the middle
fifty per cent of the distribution. Together they indicate that earnings vary
considerably across workers.

Sometimes we are interested not only in the overall distribution but also in
how it differs across groups. A density curve provides a smooth representation
of the same data and makes such comparisons easier.

```{r u1-density, fig.height = 3.4, fig.cap = "Density curves for rural and urban workers."}
ggplot(workers, aes(annual_earnings / 1000, colour = residence)) +
  geom_density(linewidth = 0.8) +
  scale_colour_manual(values = c(Rural = book_palette$ink,
                                 Urban = book_palette$accent),
                      name = NULL) +
  coord_cartesian(xlim = c(0, 400)) +
  labs(x = "Annual earnings (₹ thousand)", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank(),
        legend.position = "top")
```

The entire urban distribution lies to the right of the rural distribution,
indicating that urban workers generally earn more than rural workers. The
density curve makes this comparison immediately visible without requiring us to
examine thousands of individual observations.

By this point, twenty thousand observations have been reduced to a handful of
informative summaries: a picture of the distribution, measures of its centre
and spread, and a comparison across groups. This is the purpose of
**descriptive statistics**. Rather than examining every observation
individually, we summarise the data in ways that preserve its most important
features.

Descriptive statistics is always the starting point of a statistical analysis.
Before asking whether a result is statistically significant or whether it can
be generalised beyond the data at hand, we must first understand what the
observed data look like. Only then can we ask the central question of this
unit: if we had collected a different sample, would these summaries have been
different?

## The limits of descriptive statistics {#u1-not-enough}

::: {.question}
How far can description take us?
:::

Descriptive statistics tells us everything we could want to know about the data
we have. The difficulty is that the data we have are rarely the data we
ultimately care about.

The twenty thousand workers in our dataset were not selected individually
because they were special. They are one sample drawn from a much larger
population of Indian workers. Had a different set of workers been surveyed, the
dataset would have been different — and so too would the summaries we
calculated in the previous section.

To see this, let us treat our dataset as the population and repeatedly draw
samples of two thousand workers.

```{r u1-three-samples}
set.seed(1)

s1 <- workers$annual_earnings[sample(nrow(workers), 2000)]
s2 <- workers$annual_earnings[sample(nrow(workers), 2000)]
s3 <- workers$annual_earnings[sample(nrow(workers), 2000)]
```

```{r u1-three-plot, echo = FALSE, fig.height = 4.4, fig.cap = "Three samples of two thousand workers drawn from the same population."}
three <- rbind(
  data.frame(sample = "Sample 1", x = s1),
  data.frame(sample = "Sample 2", x = s2),
  data.frame(sample = "Sample 3", x = s3))

means <- aggregate(x ~ sample, three, mean)

ggplot(three, aes(x / 1000)) +
  geom_histogram(bins = 45, fill = book_palette$fill, colour = "white",
                 linewidth = 0.12) +
  geom_vline(data = means, aes(xintercept = x / 1000),
             colour = book_palette$reject, linewidth = 0.8) +
  facet_wrap(~ sample, ncol = 1) +
  coord_cartesian(xlim = c(0, 500)) +
  labs(x = "Annual earnings (₹ thousand)", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank())
```

The three histograms look remarkably similar. Each has the same broad shape:
most workers earn relatively little, while a small number earn much more.
Nothing suggests that one sample is "better" than another.

Yet the numerical summaries are not identical.

```{r u1-three-means}
c(sample1 = mean(s1),
  sample2 = mean(s2),
  sample3 = mean(s3))
```

Each sample produces a different mean.

The same is true of the median.

```{r u1-three-medians}
c(sample1 = median(s1),
  sample2 = median(s2),
  sample3 = median(s3))
```

And of the standard deviation.

```{r u1-three-sds}
c(sample1 = sd(s1),
  sample2 = sd(s2),
  sample3 = sd(s3))
```

Every statistic changes because the sample itself has changed. None of the
calculations is incorrect, nor has anything gone wrong with the survey. The
differences arise simply because different individuals happened to be included.

The exercise above highlights a distinction that lies at the heart of
statistical inference. The **population** is the complete collection of units
about which we wish to draw conclusions, while the **sample** is the subset
that we happen to observe. Because different samples contain different
observations, the statistics computed from them also differ. This inherent
variability is known as **sampling variation**.

Sampling variation is neither a mistake nor evidence of poor data collection.
It is an unavoidable consequence of drawing inferences about a population from
a sample. Even when a survey is carefully designed and correctly implemented, a
different sample would generally produce different numerical summaries.

This observation also marks the boundary between descriptive and inferential
statistics. Descriptive statistics summarises the sample accurately: the sample
mean is the mean of the observed data, the sample median is its median, and so
on. What descriptive statistics cannot tell us is how closely these sample
summaries reflect the corresponding characteristics of the population, or how
much they would differ had another sample been drawn.

Inferential statistics addresses precisely these questions. By recognising that
the observed sample is only one of many that could have been selected, it
provides a framework for assessing the uncertainty associated with
sample-based conclusions and for drawing conclusions about the underlying
population.

## What if we repeated the survey? {#u1-frequentist}

::: {.question}
If our conclusions depend on the particular sample we happened to observe, how
can we learn anything about the population?
:::

The previous section ended with a simple but unsettling observation. A
different sample produces a different mean. The same is true of the median, the
standard deviation and every other statistic we compute.

One possibility is to collect another sample.

Imagine repeating the survey. A new team visits a different set of households,
following exactly the same sampling design. They compute the average annual
earnings and obtain a value that differs slightly from the first survey.

Now imagine repeating the exercise again.

And again.

Every new survey produces a new sample, and every sample produces a new mean.
None of these means is "correct" or "incorrect". Each is simply the result of
observing a different subset of the population.

Of course, conducting thousands of nationwide surveys would be prohibitively
expensive. Fortunately, we do not need to. Since we are treating our dataset as
the population for the purposes of this chapter, we can recreate this thought
experiment on a computer.

Suppose we repeatedly draw random samples of 2,000 workers from our population.
For each sample we calculate the mean annual earnings and store the result.
After repeating this process one thousand times, we have one thousand sample
means.

```{r u1-many-samples}
set.seed(1)

sample_means <- replicate(
  1000,
  mean(sample(workers$annual_earnings,
              size = 2000,
              replace = TRUE))
)
```

What do these one thousand means look like?

```{r u1-many-plot, echo = FALSE, fig.height = 3.6, fig.cap = "Distribution of the sample mean based on 1,000 samples of 2,000 workers."}
ggplot(data.frame(m = sample_means / 1000), aes(m)) +
  geom_histogram(bins = 45, fill = book_palette$fill, colour = "white",
                 linewidth = 0.15) +
  geom_vline(xintercept = mean(workers$annual_earnings) / 1000,
             colour = book_palette$reject, linewidth = 0.9) +
  annotate("label", x = mean(workers$annual_earnings) / 1000, y = Inf,
           vjust = 1.4, label = "Population mean", colour = book_palette$reject,
           fill = "white", size = 3.5, label.size = 0.3) +
  labs(x = "Sample mean (₹ thousand)", y = "Number of samples") +
  theme_book()
```

Several features stand out immediately.

First, the sample means are not all identical. Even though every sample was
drawn from the same population using the same procedure, each survey produced a
slightly different average.

Second, the variation is not arbitrary. Most of the sample means cluster within
a relatively narrow range, while only a few lie far from the centre. Extreme
values are possible, but they are uncommon.

Finally, the distribution itself has a remarkably regular shape. Although the
earnings of individual workers are highly skewed, the distribution of the
sample means is much more symmetric and concentrated.

The thousand means are not just a collection of numbers. They form a
distribution of their own. Notice what this distribution describes. It is not
the distribution of workers' earnings — that was the distribution we began
with. Instead, it is the distribution of the sample mean obtained by repeating
the survey many times. The object of interest has changed. We are no longer
studying individual workers; we are studying the behaviour of the sample mean.

This distinction lies at the heart of inferential statistics. The original
distribution describes the variability in earnings across workers. The new
distribution describes the variability in the sample mean across repeated
samples. It is this second source of variation that determines how much
confidence we should place in the average computed from a single survey.

The distribution formed by a statistic over repeated samples is called its
**sampling distribution**. Every statistic has a sampling distribution: the
mean, the median, the standard deviation, and many others. In this chapter we
focus on the sampling distribution of the sample mean because it provides the
foundation for much of statistical inference.

Two features of Figure \@ref(fig:u1-many-plot) are too striking to ignore.
First, the sample means are tightly concentrated around the population mean.
Second, despite the strong skewness of individual earnings, their sampling
distribution is almost bell-shaped. Neither feature is a coincidence. Both are
consequences of two fundamental results in probability theory: the Law of Large
Numbers and the Central Limit Theorem. Together, these results explain why a
single sample can provide reliable information about an entire population.

## Why do larger samples give more reliable answers? {#u1-lln}

::: {.question}
What determines how much they vary?
:::

Figure \@ref(fig:u1-many-plot) answered one important question. Repeating the
survey many times does not produce wildly different
sample means. Instead, the sample means cluster around the population mean.

A natural question follows. What determines how much they vary?

One possibility is the size of the survey itself. Intuition suggests that a
survey of twenty workers should produce averages that vary much more from one
survey to the next than a survey of two thousand workers. But how much
difference does sample size really make?

To find out, let us repeat the experiment from the previous section several
times. Instead of fixing the sample size at 2,000 workers, we consider surveys
of different sizes. For each sample size we draw one thousand independent
samples and compute the average annual earnings.

```{r u1-lln-sizes}
set.seed(3)

earn  <- workers$annual_earnings
sizes <- c(20, 100, 500, 2000)

by_size <- do.call(rbind, lapply(sizes, function(n) {
  data.frame(n = factor(paste("n =", n), levels = paste("n =", sizes)),
             m = replicate(4000, mean(sample(earn, n, replace = TRUE))))
}))
```

```{r u1-lln-plot, echo = FALSE, fig.height = 5.6, fig.cap = "Sampling distributions of the sample mean for different sample sizes."}
ggplot(by_size, aes(m / 1000)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 1.2,
                 fill = book_palette$fill, colour = "white", linewidth = 0.08) +
  geom_vline(xintercept = mean(earn) / 1000, colour = book_palette$reject,
             linewidth = 0.7) +
  facet_wrap(~ n, ncol = 1, scales = "free_y") +
  coord_cartesian(xlim = c(15, 105)) +
  labs(x = "Sample mean (₹ thousand)", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank())
```

The pattern is unmistakable.

When the sample contains only a small number of workers, the sample means are
widely dispersed. Some surveys produce averages well above the population mean,
while others fall well below it. The averages vary substantially from one
survey to the next.

As the sample size increases, the sampling distribution becomes progressively
narrower. The sample means remain centred at the population mean, but they
fluctuate much less. Large departures from the population mean become
increasingly uncommon, while averages close to the population mean become
increasingly likely.

```{r u1-lln-spread}
round(tapply(by_size$m, by_size$n, sd))
```

This behaviour reflects a simple idea. Every sample contains chance variation.
Some samples happen to include more high-income workers than average, while
others include more low-income workers. In a small sample these random
differences have a noticeable effect on the average. As the sample grows,
unusually high and unusually low observations increasingly offset one another,
and the sample mean settles closer to the population mean.

The behaviour we have just observed is not merely an empirical regularity. It
is one of the most fundamental results in probability theory and forms the
foundation of statistical inference. It is known as the **Law of Large
Numbers**.

In informal terms, the Law of Large Numbers states that as the sample size
increases, the sample mean gets closer and closer to the population mean. Large
samples do not eliminate randomness, but they reduce its influence. Although no
single survey is guaranteed to produce the correct answer, larger surveys are
increasingly likely to do so.

The Law of Large Numbers explains why surveys, opinion polls, clinical trials
and economic censuses can provide reliable information about populations
without observing every individual. The key is not that they observe everyone,
but that they observe enough.

::: {.realworld}
Real demonstrations of the Law of Large Numbers are surprisingly hard to find.
The law concerns what happens when the same random process is repeated a great
many times, and few processes in economic life are repeated identically enough,
often enough, for the convergence to be visible.

The cricket toss is an exception. Before every international match a coin is
tossed, and over nearly a century and a half some teams have accumulated
thousands of these tosses while others have played only a handful.

```{r u1-cricket}
toss <- read.csv("data/cricket-toss.csv")
keep <- c("team", "span", "matches", "pct_toss_won")

# the most and least fortunate teams, and the most experienced
rbind(head(toss[order(-toss$pct_toss_won), keep], 2),
      head(toss[order( toss$pct_toss_won), keep], 2),
      head(toss[keep], 3))
```

Greece has won 80% of its tosses and Africa XI barely one in six, while
England, after 2,124 matches, sits at 49.2%. Nobody is luckier than anybody
else. The teams with extreme records are simply the teams that have played few
matches.

That is the Law of Large Numbers, and the agreement with theory is close. If
the toss is fair, a team with $n$ matches should have a toss-win percentage
scattered around 50 with a standard deviation of $50/\sqrt{n}$ percentage
points. Restricting attention to teams above a given number of matches:

```{r u1-cricket-lln}
lln_toss <- t(sapply(c(0, 50, 100, 200, 500), function(m) {
  x <- toss[toss$matches >= m, ]
  c(min_matches = m,
    teams       = nrow(x),
    observed_sd = sd(x$pct_toss_won),
    predicted   = sqrt(mean((50 / sqrt(x$matches))^2)))
}))

round(lln_toss, 2)
```

The spread of toss-win percentages falls from 8.5 percentage points across all
teams to 0.9 among the ten most experienced — and at every step it tracks the
value the theory predicts. No survey was designed and no experiment was run.
The convergence is simply there in the record of the game.
:::

Yet Figure \@ref(fig:u1-lln-plot) reveals another feature that the Law of
Large Numbers does not explain. As the sample size increases, the sampling
distributions not only become narrower — they also become increasingly
bell-shaped. This is surprising because the underlying distribution of earnings
is strongly skewed.

Why should averages of skewed data look approximately normal?

The answer is provided by an even more remarkable result: the **Central Limit
Theorem**.

## The Central Limit Theorem {#u1-clt}

::: {.question}
Why do sample means look bell-shaped?
:::

The Law of Large Numbers explains why the sample mean moves closer to the
population mean as the sample size increases. It tells us where the sample mean
is centred, but not how it is distributed for any given sample size. Without
knowing that distribution, we cannot quantify the uncertainty associated with a
single survey.

The Central Limit Theorem answers precisely this question. Its remarkable
feature is its generality.

::: {.keyidea}
**Central Limit Theorem.** For sufficiently large samples, the sampling
distribution of the sample mean is approximately normal, regardless of the
shape of the population from which the sample is drawn.
:::

The theorem will be stated more precisely once we have developed the necessary
notation. For now, its essential idea is already visible in Figure
\@ref(fig:u1-lln-plot). Even when the population itself is far from
bell-shaped, the distribution of the sample mean becomes increasingly
bell-shaped as the sample size grows.

The strength of this result is easy to underestimate. The population may be
binary, highly skewed, or concentrated in only one part of its range. It makes
remarkably little difference. Average enough observations together, and the
distribution of those averages approaches the familiar bell-shaped curve.

### A simulation {-}

The easiest way to appreciate the theorem is to watch it happen. Figure
\@ref(fig:u1-clt-plot) compares the sampling distributions of the sample mean
for four populations with very different shapes, using sample sizes of one,
five and thirty.

```{r u1-clt-sim}
set.seed(3)

draw_mean <- function(parent, n) {
  switch(parent,
    "Bernoulli" = mean(rbinom(n, 1, 0.3)),
    "Poisson"   = mean(rpois(n, 2)),
    "Uniform"   = mean(runif(n)),
    "Earnings"  = mean(sample(earn, n, replace = TRUE)))
}

parents <- c("Bernoulli", "Poisson", "Uniform", "Earnings")
sizes   <- c(1, 5, 30)

clt <- do.call(rbind, lapply(parents, function(p)
  do.call(rbind, lapply(sizes, function(n)
    data.frame(parent = p,
               n      = factor(paste("n =", n), levels = paste("n =", sizes)),
               m      = replicate(4000, draw_mean(p, n)))))))

# Standardise within each panel, so all twelve are on one scale and the
# comparison is purely about shape.
clt$z <- ave(clt$m, clt$parent, clt$n,
             FUN = function(x) (x - mean(x)) / sd(x))
```

```{r u1-clt-plot, echo = FALSE, fig.height = 6, fig.cap = "Standardised sampling distributions of the sample mean for four populations, against the standard normal curve."}
clt$parent <- factor(clt$parent, levels = parents)

ggplot(clt, aes(z)) +
  geom_histogram(aes(y = after_stat(density)), bins = 45,
                 fill = book_palette$fill, colour = "white", linewidth = 0.08) +
  stat_function(fun = dnorm, colour = book_palette$reject, linewidth = 0.6) +
  facet_grid(parent ~ n) +
  coord_cartesian(xlim = c(-3.2, 3.2), ylim = c(0, 0.52)) +
  labs(x = "Standardised sample mean", y = NULL) +
  theme_book(base_size = 10) +
  theme(axis.text.y = element_blank(),
        strip.text.y = element_text(angle = 0, hjust = 0))
```

The first column shows the four underlying populations. None resembles a normal
distribution. By the third column, however, the distributions of the sample
means are all much closer to the familiar bell-shaped curve.

The theorem guarantees that this convergence eventually occurs, but it says
nothing about how quickly it happens. Populations that are already fairly
symmetric require only modest sample sizes, whereas heavily skewed populations
may require substantially larger samples before the normal approximation
becomes accurate. In Figure \@ref(fig:u1-clt-plot), the uniform population is
already close to normal when $n = 5$, while the earnings distribution remains
visibly asymmetric even when $n = 30$.

### An intuition {-}

Why should averaging observations produce a bell-shaped distribution?

A simple example provides the intuition. Roll a fair die once and every outcome
from one to six is equally likely. Now roll two dice and add the results.

```{r u1-dice-sum}
sums <- outer(1:6, 1:6, "+")

table(sums)
```

```{r u1-dice-plot, echo = FALSE, fig.height = 3, fig.cap = "The number of ways each total can be obtained when two dice are rolled."}
ds <- as.data.frame(table(sums)); names(ds) <- c("sum", "ways")
ds$sum <- as.numeric(as.character(ds$sum))

ggplot(ds, aes(sum, ways)) +
  geom_col(fill = book_palette$fill, width = 0.75) +
  geom_text(aes(label = ways), vjust = -0.5, size = 3.1,
            colour = book_palette$ink) +
  scale_x_continuous(breaks = 2:12) +
  labs(x = "Total of two dice", y = "Number of ways") +
  theme_book() +
  theme(axis.text.y = element_blank())
```

Seven can be obtained in six different ways:
$1+6,\; 2+5,\; 3+4,\; 4+3,\; 5+2,\; 6+1$. By contrast, a total of two can
be obtained in only one way. The middle values have many more possible
combinations than the extremes, so they occur more frequently.

The same intuition extends far beyond dice. To obtain an unusually large
average, most observations must themselves be unusually large. Likewise, an
unusually small average requires most observations to be unusually small. Both
situations are relatively rare. A value close to the middle, however, can arise
from countless combinations of larger and smaller observations offsetting one
another.

As more observations are averaged together, the number of ways to obtain values
near the centre grows much faster than the number of ways to obtain values at
the extremes. The result is the familiar bell-shaped distribution.

### How large is "large enough"? {-}

Many textbooks suggest that a sample size of at least thirty is sufficient for
the Central Limit Theorem to provide a good approximation. This is a useful
rule of thumb, but it is not a mathematical guarantee.

How quickly the approximation becomes accurate depends largely on the shape of
the population. Populations that are approximately symmetric require relatively
few observations, whereas heavily skewed populations often require much larger
samples.

```{r u1-clt-howbig}
set.seed(9)

skewness <- function(x) mean((x - mean(x))^3) / sd(x)^3

sizes <- c(5, 30, 100, 500)

sk <- c(population = skewness(earn),
        sapply(sizes, function(n)
          skewness(replicate(2000, mean(sample(earn, n, replace = TRUE))))))
names(sk)[-1] <- paste0("n=", sizes)

round(sk, 2)
```

```{r u1-clt-skewvals, echo = FALSE}
sk_pop <- round(mean((earn - mean(earn))^3) / sd(earn)^3, 1)
```

The earnings distribution used throughout this chapter has a skewness of
approximately `r sk_pop`. When the sample size is five, the sampling
distribution remains strongly right-skewed. Even at thirty observations, some
asymmetry is still visible. By one hundred observations, however, the
distribution is much closer to symmetric.

This matters because many economic variables — including incomes, firm sizes,
land holdings and city populations — are themselves highly skewed. For such
variables, the normal approximation may still be poor when the sample size is
only thirty. In these situations, larger samples or simulation-based checks are
often preferable to relying mechanically on a rule of thumb.

### Why this matters {-}

The Central Limit Theorem, together with the Law of Large Numbers, explains why
statistical inference is possible.

The Law of Large Numbers tells us that repeated samples become increasingly
centred on the population mean as the sample size grows. The Central Limit
Theorem tells us that the distribution of those sample means approaches a
normal distribution. Together, these results reveal that although different
samples produce different answers, those answers are not arbitrary. They vary
in systematic and predictable ways.

::: {.keyidea}
This insight lies at the heart of inferential statistics. We observe only one
sample, but by understanding how the same statistic would behave across many
hypothetical samples, we can quantify the uncertainty associated with the
sample we actually observe.
:::

So far we have developed these ideas largely through intuition, simulation and
visualisation. The next section introduces the notation and concepts needed to
express them more precisely and to build the methods of statistical inference
developed in the remainder of this book.

## A language for randomness {#u1-random-variables}

::: {.question}
Why do we need a mathematical language?
:::

Throughout this chapter we have described repeated sampling almost entirely in
plain English. We have spoken about rolling dice, drawing surveys, computing
averages and repeating experiments thousands of times. That language has helped
us build intuition, but it is no longer sufficient. To develop statistical
inference, we need a concise and precise way of expressing these ideas.

The first concept we need is that of a random variable.

### Random variables {-}

Some quantities are known before we observe them. If you measure your height
today, it is a fixed number. It may be unknown to you, but it is not changing
from one repetition to the next.

Other quantities are different. Roll a die and you do not know what number will
appear. Randomly select a worker from a population and you do not know that
person's earnings. Draw a sample of one hundred households and you do not know
what the sample mean will be until the sample has actually been collected.

Statisticians describe such quantities as **stochastic**. A stochastic quantity
is one whose value is determined by the outcome of a random process.

The mathematical object used to describe a stochastic quantity is called a
**random variable**. We usually denote random variables by capital letters such
as $X$, $Y$ and $Z$. For example, $X$ may denote the earnings of a randomly
selected worker, $Y$ the annual rainfall in a particular district, and $D$ the
outcome of rolling a fair die.

The sample mean deserves special attention. Every time we draw a different
sample, we obtain a different average. The sample mean is therefore itself a
random variable. Before observing the data, we denote it by $\bar{X}$.

::: {.keyidea}
This observation is fundamental to inferential statistics. Since $\bar{X}$ is
a random variable, it has a probability distribution. That distribution tells
us how the sample mean behaves across repeated samples. Once we understand its
centre and spread, we can quantify the uncertainty associated with the single
sample we actually observe.
:::

### Realisations {-}

Eventually every random variable takes a particular observed value.

Suppose $X$ represents the outcome of rolling a fair die. Before the die is
rolled, $X$ could take any value from one to six. After the die lands showing
four, we have observed one **realisation** of the random variable, written as
$x = 4$.

The same distinction applies to the sample mean. Before collecting a sample,
the sample mean is the random variable $\bar{X}$. After collecting the data and
computing the average, we observe one realisation of that random variable,
written as $\bar{x}$.

Throughout the remainder of this book, capital letters denote random variables,
while lower-case letters denote their observed values.

### The long-run average {-}

Every random variable has a centre. The centre is called its **expected value**
and is written as

$$E(X) = \sum_i x_i \, P(X = x_i).$$

The term *expected value* can be misleading. It is not the value we expect to
observe on any particular repetition. Rather, it is the average value the
random variable would take over a very large number of repetitions.

For a fair die,

$$E(X) = \frac{1+2+3+4+5+6}{6} = 3.5.$$

Of course, no single roll can ever produce 3.5. It is a theoretical average,
not an individual outcome.

This distinction becomes clearer when we compare theory with data.

```{r u1-expectation}
mean(rolls)                # average of 10,000 simulated rolls
sum(1:6 * rep(1/6, 6))     # theoretical expected value
```

The close agreement between these two numbers is exactly what the Law of Large
Numbers predicts. The expected value is calculated from the probability model,
while the sample average is calculated from observed data. As the number of
observations grows, the latter approaches the former.

When the random variable represents a randomly selected member of a population,
its expected value is simply the population mean. For this reason we denote
both by the same symbol,

$$E(X) = \mu.$$

### Measuring variation {-}

Knowing the centre of a distribution is not enough. Two random variables can
have the same expected value but very different amounts of variation.

The spread of a random variable around its expected value is measured by its
**variance**,

$$\operatorname{Var}(X) = E\big[(X - \mu)^2\big],$$

and the square root of the variance is the **standard deviation**,

$$\sigma = \sqrt{\operatorname{Var}(X)}.$$

The expected value tells us where a random variable is centred. The variance
and standard deviation tell us how much it fluctuates around that centre.

### Parameters and statistics {-}

Statistics is concerned with learning about populations from samples. It is
therefore essential to distinguish between quantities that describe a
population and quantities computed from a sample.

A **parameter** is a numerical quantity describing the population. The
population mean is written $\mu$, the population variance $\sigma^2$, and the
population standard deviation $\sigma$. These quantities are fixed — they do
not change — but they are usually unknown, because we rarely observe the entire
population.

A **statistic**, by contrast, is computed from a sample. The sample mean,
sample variance and sample standard deviation are all statistics. Unlike
parameters, statistics change from one sample to the next, because they depend
on which observations happen to be selected.

The entire objective of inferential statistics can now be expressed in a single
sentence:

::: {.keyidea}
Use statistics computed from a sample to learn about unknown parameters
describing the population.
:::

Everything that follows in this book — estimation, confidence intervals,
hypothesis testing and regression analysis — is an extension of this single
idea.

### A summary of notation {-}

The notation introduced in this section will be used throughout the book.

| Symbol | Meaning |
|:--|:--|
| $X,\ Y,\ Z$ | random variables — quantities not yet observed |
| $x,\ y,\ z$ | realisations — values actually observed |
| $\bar{X}$ | the sample mean, before the sample is drawn |
| $\bar{x}$ | the sample mean, once the sample is observed |
| $n$ | the sample size |
| $E(X)$ | expected value: the long-run average of $X$ |
| $\operatorname{Var}(X)$ | variance: the spread of $X$ about its expected value |

And, separating what describes the population from what is computed from a
sample:

| Quantity | Population (parameter) | Sample (statistic) |
|:--|:--|:--|
| Mean | $\mu$ | $\bar{x}$ |
| Variance | $\sigma^2$ | $s^2$ |
| Standard deviation | $\sigma$ | $s$ |

The left-hand column is fixed and unknown. The right-hand column is observed
and changes with every sample. Almost all of statistics is the effort to say
something about the first using only the second.

## Before any mathematics can help {#u1-what-goes-wrong}

::: {.question}
What must be true of the data before any of this applies?
:::

Everything in this unit assumes that the sample carries information about the
population we care about. That assumption is not guaranteed by any formula, and
when it fails, no amount of statistical sophistication will announce the
failure. The arithmetic runs perfectly on data that cannot answer the question.

Four ways it fails, none of which any calculation will warn you about.

**The sample is not from the population you mean.** The Annual Survey of
Industries covers *registered* factories. Most Indian manufacturing employment
is unregistered. Conclusions about "Indian manufacturing" drawn from that
source are conclusions about a formal sliver of it, however large the sample.

**The sample is not random.** The class survey in Section
\@ref(u1-describing) has thirty-two heights. They are not a random sample of
students at the university, still less of young Indians: they are the people
who took the course and attended that day. A confidence interval computed from
them is arithmetically correct and tells you nothing about anyone else.

**The data are measured badly.** Incomes reported from memory are systematically
understated. Rainfall attributed to a district may have been recorded at a
station a hundred kilometres away. National accounts are revised for years
after first publication. A sampling distribution describes how an estimate
varies across samples; it says nothing about whether the quantity was measured
correctly in the first place. Careful statistics performed on poor measurements
remain poor statistics.

**The question is causal but the data are not.** Suppose states with more
industry have higher literacy. Does industry raise literacy? Does literacy
attract industry? Does something else drive both? The correlation is identical
in all three cases, and no statistical procedure separates them without an
argument about how the data came to exist.

::: {.keyidea}
These are not edge cases. They are the ordinary condition of applied economics,
and recognising them is a skill quite separate from computation.

Good inference begins with good data and thoughtful study design, not with
clever mathematics.
:::

## Summary {#u1-summary}

::: {.keyidea}
1. **Randomness creates variation, and variation is the raw material of
   statistics.** Where nothing varies there is no distribution, no uncertainty,
   and nothing to infer.
2. **Descriptive statistics** summarises the data in hand. **Inferential
   statistics** uses that data to make claims about a population it did not
   observe, and reports how uncertain those claims are.
3. Different samples give different answers. That is **sampling variation** —
   unavoidable, and not a defect in the data.
4. A quantity whose value changes from one repetition to the next is a **random
   variable**, written $X$. Its observed value is a **realisation**, written
   $x$. The sample mean is $\bar{X}$ before the sample is drawn and $\bar{x}$
   after.
5. $E(X) = \sum_i x_i P(X = x_i)$ is the **expected value** — a long-run
   average, not a value you expect to see. For a population draw,
   $E(X) = \mu$.
6. $\operatorname{Var}(X) = E[(X - \mu)^2]$ measures spread, and
   $\sigma = \sqrt{\operatorname{Var}(X)}$ is the standard deviation.
7. **Parameters** ($\mu$, $\sigma^2$, $\sigma$) describe the population and
   are fixed but unknown. **Statistics** ($\bar{x}$, $s^2$, $s$) are computed
   from a sample and change with every sample. Inference uses the second to
   learn about the first.
8. The distribution of a statistic across repeated samples is its **sampling
   distribution**. Every statistic has one, not only the mean.
9. **Law of Large Numbers:** as $n$ grows, the sample mean becomes increasingly
   likely to lie close to $\mu$. It does not say any particular large sample is
   correct, and it offers no comfort to a gambler.
10. **Central Limit Theorem:** for sufficiently large $n$, the sampling
    distribution of $\bar{X}$ is approximately normal whatever the shape of the
    population. The rule $n \geq 30$ is a rule of thumb, not a guarantee — a
    heavily skewed population needs far more.
11. No formula warns you when the sample is drawn from the wrong population,
    drawn non-randomly, measured badly, or asked a causal question it cannot
    answer.
:::

Taken together, these amount to a single change of perspective. We stopped
asking what a particular sample says, and began asking how a statistic behaves
across the many samples that might have been drawn. Everything that follows —
estimation, confidence intervals, hypothesis testing, regression — rests on
that shift.

## Practice problems {#u1-practice}

::: {.yourturn}
These are for learning. Work each one before opening the solution; the answer
is worth very little if you have not first tried to produce it yourself.

The [question bank](#u1-question-bank) that follows is for testing yourself and
has no answers at all.
:::

### Concept checks {-}

Short questions. A sentence or two is enough.

**1.1** Why does repeating the same survey produce a different average each
time?

::: {.answer}
Because a different set of people is surveyed each time. Every average is
computed from whoever happened to be selected, and selection involves chance.
The population has not changed and no arithmetic error has occurred — only the
membership of the sample.
:::

**1.2** A die with a three painted on all six faces is rolled ten thousand
times. What is the distribution of the outcome? Why is this example in the
chapter at all?

::: {.answer}
There is no distribution to speak of: every outcome is three, and the standard
deviation is zero. The example is there to show that statistics requires
variation. Where outcomes never differ, there is nothing to describe, nothing
to be uncertain about, and no inference to perform.
:::

**1.3** Distinguish between $\bar{X}$ and $\bar{x}$.

::: {.answer}
$\bar{X}$ is the sample mean regarded as a random variable, before the sample
is drawn: it has a distribution, a centre and a spread. $\bar{x}$ is the single
number obtained once a particular sample has been collected — one realisation
of $\bar{X}$.
:::

**1.4** A survey of 400 households gives a mean monthly income of ₹18,000. Is
₹18,000 a parameter or a statistic? What is the parameter here?

::: {.answer}
₹18,000 is a statistic: it was computed from a sample and would change with a
different sample. The parameter is the mean monthly income of the whole
population, $\mu$, which is fixed but unknown.
:::

### Explain {-}

One paragraph each. These reward clarity rather than calculation.

**1.5** Explain to a friend who has never studied statistics why the Law of
Large Numbers does not mean that a large sample is guaranteed to give the right
answer.

::: {.answer}
The law says that as the sample grows, the sample mean becomes increasingly
likely to lie close to the population mean — not that it will equal it. Any
particular large sample may still be unlucky. What changes with sample size is
the *probability* of being far off, which shrinks; the possibility never
disappears. A large sample makes a badly wrong answer improbable, not
impossible.
:::

**1.6** A friend argues that because a coin has landed heads six times running,
tails is now more likely. Using the Law of Large Numbers, explain why this is
wrong.

::: {.answer}
The coin has no memory, and the law promises nothing of the kind. What
converges is the *proportion* of heads, and it converges by dilution rather
than by correction. The six extra heads are never cancelled out; they simply
become a smaller and smaller share of a growing total. The absolute surplus of
heads typically grows with the number of tosses even as the proportion
approaches one half.
:::

**1.7** Two researchers study earnings in the same district. One surveys 100
households, the other 2,500. Both report a mean. Which should be trusted more,
and what would you need to know before answering?

::: {.answer}
Other things equal, the larger survey has the smaller standard deviation of the sample mean — by a
factor of five, since $\sqrt{2500/100} = 5$. But this holds only if both
samples were drawn from the population of interest and drawn at random. A
carefully designed survey of 100 households is worth more than a badly drawn
survey of 2,500. Sample size governs precision; sampling design governs whether
precision is about the right quantity at all.
:::

### Numerical problems {-}

Work each of these with a pen first. Each ends by asking you to check your
answer by simulation — the point being that theory and simulation should agree,
and that when they do not, one of them is wrong.

**1.8** The earnings data used throughout this chapter has a population mean of
about ₹54,200 and a population standard deviation of about ₹100,100.

Suppose a survey team draws random samples of (a) 50, (b) 100, (c) 200 and
(d) 1,000 workers.

i. Find the expected value of the sample mean in each case.
ii. Find the standard deviation of the sample mean in each case.
iii. What happens to that standard deviation as the sample size increases, and why?
iv. Verify your answers by simulation.

::: {.answer}
The expected value of the sample mean equals the population mean whatever the
sample size, so $E(\bar{X}) = ₹54{,}200$ in all four cases. Sample size affects
precision, not location.

The standard deviation of the sample mean is $\sigma/\sqrt{n} = 100{,}100/\sqrt{n}$:

```{r u1-ex-se}
sigma <- 100100
n     <- c(50, 100, 200, 1000)

round(data.frame(n, expected = 54200, se = sigma / sqrt(n)))
```

It falls in proportion to $1/\sqrt{n}$. Each additional worker
contributes information, so the average over many of them is pulled more
tightly around the population value. Note that raising the sample from 50 to
1,000 — twenty times the data — reduces it only by a factor of
$\sqrt{20} \approx 4.5$.

Because we hold the whole population, the answer can be checked directly:

```{r u1-ex-se-sim}
set.seed(4)
earn <- read.csv("data/wages-india-synthetic.csv")$annual_earnings

sim <- sapply(n, function(k)
  sd(replicate(2000, mean(sample(earn, k, replace = TRUE)))))

round(data.frame(n, theoretical = sd(earn) / sqrt(n), simulated = sim))
```

The two columns agree closely. The small discrepancies are themselves sampling
variation: each simulated standard deviation is estimated from 2,000 samples rather
than infinitely many.
:::

**1.9** An auto-rickshaw driver's daily earnings vary from day to day, with a
mean of ₹800 and a standard deviation of ₹200. In a particular month the driver
works 30 days.

i. What are the mean and standard deviation of total earnings for the month?
ii. Approximate the probability that the month's total exceeds ₹25,000.
iii. Approximate the probability that it falls below ₹22,000.
iv. Approximate the probability that it lies between ₹22,000 and ₹25,000.
v. What assumption does this calculation require, and is it plausible here?
vi. Check your answers by simulating 10,000 months.

::: {.answer}
The monthly total is a sum of 30 daily earnings. Its mean is
$30 \times 800 = ₹24{,}000$, and because variances add for independent
quantities, its standard deviation is $200\sqrt{30} = ₹1{,}095$.

By the Central Limit Theorem the total is approximately normal, so:

```{r u1-ex-auto}
mu_day <- 800; sd_day <- 200; days <- 30

mu_tot <- mu_day * days
sd_tot <- sd_day * sqrt(days)

c(mean = mu_tot, sd = sd_tot,
  above_25000 = pnorm(25000, mu_tot, sd_tot, lower.tail = FALSE),
  below_22000 = pnorm(22000, mu_tot, sd_tot),
  between     = pnorm(25000, mu_tot, sd_tot) - pnorm(22000, mu_tot, sd_tot))
```

So roughly an 18% chance of clearing ₹25,000, a 3% chance of falling below
₹22,000, and a 79% chance of landing between the two.

**The assumption is independence.** The calculation treats each day's earnings
as unrelated to every other day's. That is questionable. A week of heavy
monsoon suppresses earnings on consecutive days; a festival week raises them
together. When days are positively correlated, the true standard deviation of
the monthly total is *larger* than $200\sqrt{30}$, and the probability of an
unusually bad month is correspondingly higher than this calculation suggests.

This is worth dwelling on. The arithmetic is not wrong; the assumption is. No
amount of care in the computation would reveal the problem.

Simulating under the stated assumption:

```{r u1-ex-auto-sim}
set.seed(5)
months <- replicate(10000, sum(rnorm(days, mu_day, sd_day)))

c(mean = mean(months), sd = sd(months),
  above_25000 = mean(months > 25000),
  below_22000 = mean(months < 22000),
  between     = mean(months >= 22000 & months <= 25000))
```

The simulated proportions match the normal approximation closely — as they
must, since the simulation was itself built on the assumption of independent
daily earnings. Simulation confirms the arithmetic; it cannot confirm the
assumption.
:::

### R exercises {-}

**1.10** Simulate the Central Limit Theorem starting from an exponential
distribution, which is strongly right-skewed. Draw 5,000 sample means for
sample sizes of 1, 5, 30 and 200 from `rexp(n, rate = 1)`, and plot the four
distributions. At which sample size does the result begin to look normal?

::: {.answer}
```{r u1-ex-clt, fig.height = 5, fig.cap = "Sampling distributions of the mean from an exponential population."}
set.seed(1)

sizes <- c(1, 5, 30, 200)
sim <- do.call(rbind, lapply(sizes, function(n)
  data.frame(n = factor(paste("n =", n), levels = paste("n =", sizes)),
             m = replicate(5000, mean(rexp(n, rate = 1))))))

ggplot(sim, aes(m)) +
  geom_histogram(bins = 50, fill = book_palette$fill, colour = "white",
                 linewidth = 0.1) +
  facet_wrap(~ n, scales = "free") +
  labs(x = "Sample mean", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank())
```

At $n = 1$ the picture is the exponential distribution itself — a sharp peak at
zero with a long right tail. By $n = 5$ it has a recognisable mound but remains
clearly skewed. By $n = 30$ it is close to symmetric, and by $n = 200$ it is
difficult to distinguish from a normal distribution by eye.

The exponential distribution has skewness 2, which is milder than the earnings
data used in this chapter. That is why $n = 30$ works reasonably well here and
less well there.
:::

**1.11** Using `data/cricket-toss.csv`, plot each team's toss-win percentage
against its number of matches. What shape does the scatter take, and why?

::: {.answer}
```{r u1-ex-cricket, fig.height = 3.8, fig.cap = "Toss-win percentage against number of matches played, with the theoretical limits of two standard deviations."}
toss <- read.csv("data/cricket-toss.csv")

band <- data.frame(n = 5:2200)
band$hi <- 50 + 2 * 50 / sqrt(band$n)
band$lo <- 50 - 2 * 50 / sqrt(band$n)

ggplot(toss, aes(matches, pct_toss_won)) +
  geom_ribbon(data = band, aes(x = n, ymin = lo, ymax = hi),
              inherit.aes = FALSE, fill = book_palette$fill, alpha = 0.45) +
  geom_hline(yintercept = 50, colour = book_palette$reject, linewidth = 0.6) +
  geom_point(alpha = 0.7, size = 1.6, colour = book_palette$ink) +
  scale_x_log10(labels = scales::comma) +
  labs(x = "Matches played (log scale)", y = "Toss-win percentage") +
  theme_book()
```

The scatter forms a funnel: wide on the left, where teams have played few
matches, and narrowing towards 50% as the number of matches grows. The shaded
band shows two standard deviations either side of 50, computed as
$50/\sqrt{n}$ — the same formula that governs the sample mean. Almost every
team falls inside it.

This is the Law of Large Numbers made visible. The teams with extreme records
are not lucky; they are simply the teams with few observations.
:::

## Question bank {#u1-question-bank}

::: {.watchout}
**No answers are given for this section, by design.** These questions are for
testing yourself once the practice problems are complete. There are no hints
and no worked solutions — exactly as in an examination.
:::

### Multiple choice {-}

**QB 1.1** Which of the following is a random variable? *(select all that
apply)*

a. The number showing after a die is rolled
b. The number showing on a die with a three painted on every face
c. The mean earnings of a sample of 500 workers, before the sample is drawn
d. The mean earnings of a sample of 500 workers, after the sample is drawn

**QB 1.2** A population has mean $\mu = 40$ and standard deviation
$\sigma = 12$. For samples of size 36, the standard deviation of the sample mean
is

a. 12
b. 2
c. 0.33
d. 6

**QB 1.3** Which statements about the Law of Large Numbers are correct?
*(select all that apply)*

a. As the sample grows, the sample mean becomes more likely to be close to the
   population mean
b. As the sample grows, the sample mean becomes certain to equal the population
   mean
c. It applies only when the population is normally distributed
d. It explains why a survey need not interview everyone

**QB 1.4** A researcher doubles the size of a survey from 500 to 1,000
households. The standard deviation of the sample mean falls by a factor of
approximately

a. 2
b. 4
c. 1.41
d. It does not change

**QB 1.5** The Central Limit Theorem states that

a. Large samples are normally distributed
b. The population becomes normal as the sample grows
c. The sampling distribution of the sample mean is approximately normal for
   sufficiently large samples
d. Any variable is approximately normal if enough observations are collected

**QB 1.6** Which of the following would make the normal approximation to the
sampling distribution *less* accurate at a given sample size? *(select all that
apply)*

a. A heavily right-skewed population
b. A symmetric population
c. A population with a small number of extreme outliers
d. A larger population standard deviation

**QB 1.7** A statistic differs from a parameter because

a. A statistic is always correct
b. A statistic is computed from a sample and changes from sample to sample
c. A parameter is computed from a sample
d. A parameter changes when a different sample is drawn

### Short reasoning {-}

**QB 1.8** A newspaper reports that average household income in a district is
₹32,000, based on a survey of 800 households, and describes this as "the income
of a typical household". Give two separate objections to that description.

**QB 1.9** An analyst has a sample of 50,000 observations and concludes that
because the sample is very large, the results must be reliable. Under what
circumstances would this reasoning be wrong?

**QB 1.10** Explain why the sampling distribution of the sample mean is
narrower than the distribution of the individual observations, using the idea
of offsetting variation rather than any formula.

### Numerical {-}

**QB 1.11** The weights of rice bags filled by a machine have mean 25 kg and
standard deviation 0.6 kg. A quality inspector weighs 36 bags.

(a) What is the expected value of the sample mean weight?
(b) What is its standard deviation?
(c) How many bags would need to be weighed to halve that standard deviation?

**QB 1.12** Daily wages of agricultural labourers in a district have mean ₹350
and standard deviation ₹90. For a random sample of 100 labourers, approximate
the probability that the sample mean wage

(a) exceeds ₹360, (b) falls below ₹340, (c) lies between ₹340 and ₹360.

**QB 1.13** A population is strongly right-skewed with mean 12 and standard
deviation 15. A student proposes to use the normal approximation for the
sampling distribution of the mean with a sample of 20 observations. Comment on
whether this is advisable, and describe how you would check.

## Looking ahead {#u1-looking-ahead}

We now understand why different samples produce different answers, and why
those answers follow predictable patterns.

Suppose you collect a sample of fifty households and find an average monthly
income of ₹24,000.

Is that evidence that the population average is ₹24,000?

Could the true average instead be ₹25,000? Or ₹23,500? How different would your
sample have to be before you concluded that the population itself is different?

These questions motivate the next unit.
