# Towards Inferential Statistics {#towards-inference}

```{r setup-01, include = FALSE}
source("_common.R")
library(ggplot2)
```

You already know how to calculate the average of a sample.

In this unit, you will learn why that average is almost certainly wrong.

At first glance, this seems like a strange claim. After all, there is only one
way to calculate a sample mean, and if the arithmetic is correct, the answer
must be correct as well. Yet the number you obtain from a sample is almost
never exactly equal to the quantity you ultimately wish to know — the mean of
the population from which that sample was drawn.

Suppose you wish to estimate the average income of households in India. Since
surveying every household is impossible, you select a sample of 2,000
households and calculate its mean income. The calculation is straightforward,
but the result immediately raises a question. How close is this sample mean to
the true population mean?

The difficulty is that the sample you have drawn is only one of many that could
have been observed. Had a different 2,000 households been selected, the sample
mean would almost certainly have been different. Draw another sample, and the
answer changes again. Every statistic calculated from a sample therefore
reflects two things at once: the characteristics of the underlying population,
and the randomness introduced by the sampling process.

This simple observation marks the boundary between descriptive and inferential
statistics.

Until now, our objective has been to describe the data that we observe.
Measures of central tendency, such as the mean and the median, summarise where
observations are concentrated. Measures of dispersion, such as the variance and
standard deviation, describe how widely they are spread. Histograms, bar charts
and density plots reveal the overall shape of a distribution and help us
identify skewness, outliers and other important features. Together, these tools
provide a concise description of a dataset and are the natural starting point
of every empirical investigation.

Most empirical research, however, is concerned not with the sample itself but
with the population from which it was drawn. Economists survey a few thousand
households to understand millions. Polling agencies interview a sample of
voters to estimate national voting intentions. Medical researchers evaluate new
treatments on a limited number of patients before recommending them to the
wider population. In every case, the sample serves as evidence about a
population that cannot be observed directly.

The central problem of inferential statistics is therefore straightforward to
state: **how can we use the information contained in a sample to learn about
the population from which it came?**

Throughout this course we adopt the **frequentist** approach to answering this
question. Rather than treating the observed sample as unique, we imagine
drawing many samples from the same population and calculating the same
statistic repeatedly. Some sample means will be larger than others, some
smaller. Understanding how statistics vary across these hypothetical samples
allows us to assess how much confidence we should place in the estimate
computed from the one sample we actually observe.

Before we can understand why, we need to understand where the uncertainty comes
from. The answer lies in randomness.

## Randomness and variation {#u1-randomness}

::: {.question}
Where does the uncertainty come from?
:::

The uncertainty that surrounds every statistical estimate has a single source:
randomness.

If we were to repeat the same study with a different sample, we would almost
certainly obtain a different average. Inferential statistics is concerned with
understanding this variability. Without randomness, every sample would be
identical, every estimate would equal the population value, and there would be
no uncertainty to quantify.

The idea is easiest to see with a die.

Suppose we roll a fair die once.

```{r u1-die-one}
set.seed(1)
sample(1:6, size = 1)
```

The outcome is a one. Beyond that, there is little to say. A single observation
tells us almost nothing about the process that generated it. Was the die fair?
Would the next roll also be a one? Is one an unusually common outcome? One
observation cannot answer these questions.

Now imagine repeating the experiment ten thousand times.

```{r u1-die-many}
rolls <- sample(1:6, size = 10000, replace = TRUE)

round(prop.table(table(rolls)), 3)
```

Each individual roll remains unpredictable, but something remarkable happens
collectively. The outcomes no longer appear as isolated events. They form a
stable pattern. Every face occurs with almost the same frequency, close to
one-sixth of the time, exactly as we would expect from a fair die.

```{r u1-die-plot, echo = FALSE, fig.height = 3.2, fig.cap = "Ten thousand rolls of a fair die. No single roll was predictable. The pattern across all of them is."}
ggplot(data.frame(x = factor(rolls)), aes(x)) +
  geom_bar(aes(y = after_stat(count / sum(count))),
           fill = book_palette$fill, width = 0.7) +
  geom_hline(yintercept = 1/6, colour = book_palette$reject,
             linetype = "dashed", linewidth = 0.7) +
  annotate("text", x = 6.4, y = 1/6, label = "1/6", vjust = -0.8,
           colour = book_palette$reject, size = 3.6) +
  labs(x = "outcome", y = "share of rolls") +
  theme_book()
```

The important point is not that the die turned out to be fair. It is that a
pattern exists at all.

To see why that matters, imagine removing the randomness altogether.
Suppose every face of the die is labelled three.

```{r u1-die-loaded}
loaded <- rep(3, 10000)

table(loaded)
sd(loaded)
```

Every roll now produces exactly the same outcome. There is no variation, no
uncertainty and no pattern waiting to emerge through repetition. The average is
always three. The standard deviation is zero. Repeating the experiment a
million times teaches us nothing that we did not already know after the first
roll.

The contrast between the two dice illustrates a fundamental idea. Randomness
makes individual outcomes unpredictable, but repeated observations reveal
stable patterns. Statistics is built on this coexistence of uncertainty and
regularity. If outcomes never varied, there would be no need for inference; if
they varied without any underlying regularity, inference would be impossible.

Dice provide a simple illustration, but the same principle governs every
empirical investigation. Household surveys interview different families.
Election polls reach different voters. Clinical trials recruit different
patients. Agricultural experiments observe different harvests. Every study
could have produced a different dataset. The central challenge of inferential
statistics is to understand what can be learned from **one realisation of a
process that could have produced many others**.

## Describing variation {#u1-describing}

::: {.question}
Once variation exists, how do we describe it?
:::

Variation is what makes statistics necessary. The next question is equally
natural: how do we describe it?

A die has only six possible outcomes, so its behaviour is easy to summarise.
Real datasets are rather different. Consider the annual earnings of twenty
thousand Indian workers.

```{r u1-earnings}
workers <- read.csv("data/wages-india-synthetic.csv")

nrow(workers)

head(workers$annual_earnings)
```

This dataset is simulated, but it closely resembles the earnings distribution
observed in large household surveys such as the India Human Development Survey
(IHDS). Throughout this book, we will use it to illustrate statistical ideas
and methods.

Twenty thousand numbers are too many for anyone to understand by inspection. We
need ways of reducing the data without losing its essential features. The first
step is simply to look at it.

```{r u1-earnings-hist, fig.height = 3.6, fig.cap = "Annual earnings of 20,000 Indian workers."}
ggplot(workers, aes(annual_earnings / 1000)) +
  geom_histogram(bins = 60, fill = book_palette$fill, colour = "white",
                 linewidth = 0.15) +
  scale_x_continuous(labels = scales::comma) +
  labs(x = "Annual earnings (₹ thousand)",
       y = "Number of workers") +
  theme_book()
```

The histogram immediately reveals two important features. Most workers earn
relatively modest incomes, while a much smaller number earn substantially more.
The distribution is therefore concentrated on the left with a long tail
extending to the right — a pattern known as **right skewness**.

One picture already tells us a great deal, but we can describe the same
distribution more precisely using a few numerical summaries.

```{r u1-centre}
c(mean   = mean(workers$annual_earnings),
  median = median(workers$annual_earnings))
```

The mean annual earnings are more than twice the median. This difference
reflects the long right tail visible in the histogram. A relatively small
number of high earners pull the mean upwards, whereas the median simply divides
the distribution into two equal halves and is much less affected by extreme
values.

Variation also concerns how widely observations are dispersed.

```{r u1-spread}
c(sd  = sd(workers$annual_earnings),
  iqr = IQR(workers$annual_earnings))
```

The **standard deviation** measures the typical distance of observations from
the mean, while the **interquartile range** measures the spread of the middle
fifty per cent of the distribution. Together they indicate that earnings vary
considerably across workers.

Sometimes we are interested not only in the overall distribution but also in
how it differs across groups. A density curve provides a smooth representation
of the same data and makes such comparisons easier.

```{r u1-density, fig.height = 3.4, fig.cap = "Density curves for rural and urban workers."}
ggplot(workers, aes(annual_earnings / 1000, colour = residence)) +
  geom_density(linewidth = 0.8) +
  scale_colour_manual(values = c(Rural = book_palette$ink,
                                 Urban = book_palette$accent),
                      name = NULL) +
  coord_cartesian(xlim = c(0, 400)) +
  labs(x = "Annual earnings (₹ thousand)", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank(),
        legend.position = "top")
```

The entire urban distribution lies to the right of the rural distribution,
indicating that urban workers generally earn more than rural workers. The
density curve makes this comparison immediately visible without requiring us to
examine thousands of individual observations.

By this point, twenty thousand observations have been reduced to a handful of
informative summaries: a picture of the distribution, measures of its centre
and spread, and a comparison across groups. This is the purpose of
**descriptive statistics**. Rather than examining every observation
individually, we summarise the data in ways that preserve its most important
features.

Descriptive statistics is always the starting point of a statistical analysis.
Before asking whether a result is statistically significant or whether it can
be generalised beyond the data at hand, we must first understand what the
observed data look like. Only then can we ask the central question of this
unit: if we had collected a different sample, would these summaries have been
different?

## The limits of descriptive statistics {#u1-not-enough}

::: {.question}
How far can description take us?
:::

Descriptive statistics tells us everything we could want to know about the data
we have. The difficulty is that the data we have are rarely the data we
ultimately care about.

The twenty thousand workers in our dataset were not selected individually
because they were special. They are one sample drawn from a much larger
population of Indian workers. Had a different set of workers been surveyed, the
dataset would have been different — and so too would the summaries we
calculated in the previous section.

To see this, let us treat our dataset as the population and repeatedly draw
samples of two thousand workers.

```{r u1-three-samples}
set.seed(1)

s1 <- workers$annual_earnings[sample(nrow(workers), 2000)]
s2 <- workers$annual_earnings[sample(nrow(workers), 2000)]
s3 <- workers$annual_earnings[sample(nrow(workers), 2000)]
```

```{r u1-three-plot, echo = FALSE, fig.height = 4.4, fig.cap = "Three samples of two thousand workers drawn from the same population."}
three <- rbind(
  data.frame(sample = "Sample 1", x = s1),
  data.frame(sample = "Sample 2", x = s2),
  data.frame(sample = "Sample 3", x = s3))

means <- aggregate(x ~ sample, three, mean)

ggplot(three, aes(x / 1000)) +
  geom_histogram(bins = 45, fill = book_palette$fill, colour = "white",
                 linewidth = 0.12) +
  geom_vline(data = means, aes(xintercept = x / 1000),
             colour = book_palette$reject, linewidth = 0.8) +
  facet_wrap(~ sample, ncol = 1) +
  coord_cartesian(xlim = c(0, 500)) +
  labs(x = "Annual earnings (₹ thousand)", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank())
```

The three histograms look remarkably similar. Each has the same broad shape:
most workers earn relatively little, while a small number earn much more.
Nothing suggests that one sample is "better" than another.

Yet the numerical summaries are not identical.

```{r u1-three-means}
c(sample1 = mean(s1),
  sample2 = mean(s2),
  sample3 = mean(s3))
```

Each sample produces a different mean.

The same is true of the median.

```{r u1-three-medians}
c(sample1 = median(s1),
  sample2 = median(s2),
  sample3 = median(s3))
```

And of the standard deviation.

```{r u1-three-sds}
c(sample1 = sd(s1),
  sample2 = sd(s2),
  sample3 = sd(s3))
```

Every statistic changes because the sample itself has changed. None of the
calculations is incorrect, nor has anything gone wrong with the survey. The
differences arise simply because different individuals happened to be included.

The exercise above highlights a distinction that lies at the heart of
statistical inference. The **population** is the complete collection of units
about which we wish to draw conclusions, while the **sample** is the subset
that we happen to observe. Because different samples contain different
observations, the statistics computed from them also differ. This inherent
variability is known as **sampling variation**.

Sampling variation is neither a mistake nor evidence of poor data collection.
It is an unavoidable consequence of drawing inferences about a population from
a sample. Even when a survey is carefully designed and correctly implemented, a
different sample would generally produce different numerical summaries.

This observation also marks the boundary between descriptive and inferential
statistics. Descriptive statistics summarises the sample accurately: the sample
mean is the mean of the observed data, the sample median is its median, and so
on. What descriptive statistics cannot tell us is how closely these sample
summaries reflect the corresponding characteristics of the population, or how
much they would differ had another sample been drawn.

Inferential statistics addresses precisely these questions. By recognising that
the observed sample is only one of many that could have been selected, it
provides a framework for assessing the uncertainty associated with
sample-based conclusions and for drawing conclusions about the underlying
population.

## What if we repeated the survey? {#u1-frequentist}

::: {.question}
If our conclusions depend on the particular sample we happened to observe, how
can we learn anything about the population?
:::

The previous section ended with a simple but unsettling observation. A
different sample produces a different mean. The same is true of the median, the
standard deviation and every other statistic we compute.

One possibility is to collect another sample.

Imagine repeating the survey. A new team visits a different set of households,
following exactly the same sampling design. They compute the average annual
earnings and obtain a value that differs slightly from the first survey.

Now imagine repeating the exercise again.

And again.

Every new survey produces a new sample, and every sample produces a new mean.
None of these means is "correct" or "incorrect". Each is simply the result of
observing a different subset of the population.

Of course, conducting thousands of nationwide surveys would be prohibitively
expensive. Fortunately, we do not need to. Since we are treating our dataset as
the population for the purposes of this chapter, we can recreate this thought
experiment on a computer.

Suppose we repeatedly draw random samples of 2,000 workers from our population.
For each sample we calculate the mean annual earnings and store the result.
After repeating this process one thousand times, we have one thousand sample
means.

```{r u1-many-samples}
set.seed(1)

sample_means <- replicate(
  1000,
  mean(sample(workers$annual_earnings,
              size = 2000,
              replace = TRUE))
)
```

What do these one thousand means look like?

```{r u1-many-plot, echo = FALSE, fig.height = 3.6, fig.cap = "Distribution of the sample mean based on 1,000 samples of 2,000 workers."}
ggplot(data.frame(m = sample_means / 1000), aes(m)) +
  geom_histogram(bins = 45, fill = book_palette$fill, colour = "white",
                 linewidth = 0.15) +
  geom_vline(xintercept = mean(workers$annual_earnings) / 1000,
             colour = book_palette$reject, linewidth = 0.9) +
  annotate("label", x = mean(workers$annual_earnings) / 1000, y = Inf,
           vjust = 1.4, label = "Population mean", colour = book_palette$reject,
           fill = "white", size = 3.5, label.size = 0.3) +
  labs(x = "Sample mean (₹ thousand)", y = "Number of samples") +
  theme_book()
```

Several features stand out immediately.

First, the sample means are not all identical. Even though every sample was
drawn from the same population using the same procedure, each survey produced a
slightly different average.

Second, the variation is not arbitrary. Most of the sample means cluster within
a relatively narrow range, while only a few lie far from the centre. Extreme
values are possible, but they are uncommon.

Finally, the distribution itself has a remarkably regular shape. Although the
earnings of individual workers are highly skewed, the distribution of the
sample means is much more symmetric and concentrated.

The thousand means are not just a collection of numbers. They form a
distribution of their own. Notice what this distribution describes. It is not
the distribution of workers' earnings — that was the distribution we began
with. Instead, it is the distribution of the sample mean obtained by repeating
the survey many times. The object of interest has changed. We are no longer
studying individual workers; we are studying the behaviour of the sample mean.

This distinction lies at the heart of inferential statistics. The original
distribution describes the variability in earnings across workers. The new
distribution describes the variability in the sample mean across repeated
samples. It is this second source of variation that determines how much
confidence we should place in the average computed from a single survey.

The distribution formed by a statistic over repeated samples is called its
**sampling distribution**. Every statistic has a sampling distribution: the
mean, the median, the standard deviation, and many others. In this chapter we
focus on the sampling distribution of the sample mean because it provides the
foundation for much of statistical inference.

Two features of Figure \@ref(fig:u1-many-plot) are too striking to ignore.
First, the sample means are tightly concentrated around the population mean.
Second, despite the strong skewness of individual earnings, their sampling
distribution is almost bell-shaped. Neither feature is a coincidence. Both are
consequences of two fundamental results in probability theory: the Law of Large
Numbers and the Central Limit Theorem. Together, these results explain why a
single sample can provide reliable information about an entire population.

## Why do larger samples give more reliable answers? {#u1-lln}

::: {.question}
What determines how much they vary?
:::

Figure \@ref(fig:u1-many-plot) answered one important question. Repeating the
survey many times does not produce wildly different
sample means. Instead, the sample means cluster around the population mean.

A natural question follows. What determines how much they vary?

One possibility is the size of the survey itself. Intuition suggests that a
survey of twenty workers should produce averages that vary much more from one
survey to the next than a survey of two thousand workers. But how much
difference does sample size really make?

To find out, let us repeat the experiment from the previous section several
times. Instead of fixing the sample size at 2,000 workers, we consider surveys
of different sizes. For each sample size we draw one thousand independent
samples and compute the average annual earnings.

```{r u1-lln-sizes}
set.seed(3)

earn  <- workers$annual_earnings
sizes <- c(20, 100, 500, 2000)

by_size <- do.call(rbind, lapply(sizes, function(n) {
  data.frame(n = factor(paste("n =", n), levels = paste("n =", sizes)),
             m = replicate(4000, mean(sample(earn, n, replace = TRUE))))
}))
```

```{r u1-lln-plot, echo = FALSE, fig.height = 5.6, fig.cap = "Sampling distributions of the sample mean for different sample sizes."}
ggplot(by_size, aes(m / 1000)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 1.2,
                 fill = book_palette$fill, colour = "white", linewidth = 0.08) +
  geom_vline(xintercept = mean(earn) / 1000, colour = book_palette$reject,
             linewidth = 0.7) +
  facet_wrap(~ n, ncol = 1, scales = "free_y") +
  coord_cartesian(xlim = c(15, 105)) +
  labs(x = "Sample mean (₹ thousand)", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank())
```

The pattern is unmistakable.

When the sample contains only a small number of workers, the sample means are
widely dispersed. Some surveys produce averages well above the population mean,
while others fall well below it. The averages vary substantially from one
survey to the next.

As the sample size increases, the sampling distribution becomes progressively
narrower. The sample means remain centred at the population mean, but they
fluctuate much less. Large departures from the population mean become
increasingly uncommon, while averages close to the population mean become
increasingly likely.

```{r u1-lln-spread}
round(tapply(by_size$m, by_size$n, sd))
```

This behaviour reflects a simple idea. Every sample contains chance variation.
Some samples happen to include more high-income workers than average, while
others include more low-income workers. In a small sample these random
differences have a noticeable effect on the average. As the sample grows,
unusually high and unusually low observations increasingly offset one another,
and the sample mean settles closer to the population mean.

The behaviour we have just observed is not merely an empirical regularity. It
is one of the most fundamental results in probability theory and forms the
foundation of statistical inference. It is known as the **Law of Large
Numbers**.

In informal terms, the Law of Large Numbers states that as the sample size
increases, the sample mean gets closer and closer to the population mean. Large
samples do not eliminate randomness, but they reduce its influence. Although no
single survey is guaranteed to produce the correct answer, larger surveys are
increasingly likely to do so.

The Law of Large Numbers explains why surveys, opinion polls, clinical trials
and economic censuses can provide reliable information about populations
without observing every individual. The key is not that they observe everyone,
but that they observe enough.

Yet Figure \@ref(fig:u1-lln-plot) reveals another feature that the Law of
Large Numbers does not explain. As the sample size increases, the sampling
distributions not only become narrower — they also become increasingly
bell-shaped. This is surprising because the underlying distribution of earnings
is strongly skewed.

Why should averages of skewed data look approximately normal?

The answer is provided by an even more remarkable result: the **Central Limit
Theorem**.

## The Central Limit Theorem {#u1-clt}

::: {.question}
Why do the sample means look bell-shaped?
:::

The Law of Large Numbers explains why the sample mean approaches the population
mean as the sample size increases. It does not, however, tell us how the sample
means are distributed for any given sample size. Without that distribution we
cannot quantify the uncertainty associated with a single survey.

The Central Limit Theorem answers precisely this question. The result is
surprisingly general.

::: {.keyidea}
**Central Limit Theorem.** For sufficiently large samples, the sampling
distribution of the sample mean is approximately normal, regardless of the
shape of the population from which the sample is drawn.

The result is stated precisely, in notation, once we have the vocabulary for
it later in this unit. Its content is already visible in
Figure \@ref(fig:u1-lln-plot): a strongly skewed population, yet sample means
that grow steadily more symmetric.
:::

The strength of this statement is easy to overlook. The underlying population
may be binary, highly skewed or heavily concentrated in one part of its range.
Yet the distribution of the sample mean becomes increasingly close to a normal
distribution as the sample size grows.

### A simulation {-}

The theorem is easiest to appreciate by simulation. Figure
\@ref(fig:d2i-clt-plot) compares the sampling distributions of the sample mean
for four populations with very different shapes, at sample sizes of one, five
and thirty.

```{r d2i-clt-sim}
set.seed(3)

draw_mean <- function(parent, n) {
  switch(parent,
    "Bernoulli"  = mean(rbinom(n, 1, 0.3)),
    "Poisson"    = mean(rpois(n, 2)),
    "Uniform"    = mean(runif(n)),
    "Earnings"   = mean(sample(earn, n, replace = TRUE)))
}

parents <- c("Bernoulli", "Poisson", "Uniform", "Earnings")
sizes   <- c(1, 5, 30)

clt <- do.call(rbind, lapply(parents, function(p)
  do.call(rbind, lapply(sizes, function(n)
    data.frame(parent = p,
               n      = factor(paste("n =", n), levels = paste("n =", sizes)),
               m      = replicate(4000, draw_mean(p, n)))))))

# Standardise within each panel, so all twelve are on one scale and the
# comparison is purely about shape.
clt$z <- ave(clt$m, clt$parent, clt$n,
             FUN = function(x) (x - mean(x)) / sd(x))

head(clt, 3)
```

```{r d2i-clt-plot, echo = FALSE, fig.height = 6, fig.cap = "Sampling distributions of the mean from four very different populations, standardised, against the standard normal curve in orange. At $n = 1$ you are looking at each population itself. By $n = 30$ all four track the curve — though look closely at the last row."}
clt$parent <- factor(clt$parent, levels = parents)

ggplot(clt, aes(z)) +
  geom_histogram(aes(y = after_stat(density)), bins = 45,
                 fill = book_palette$fill, colour = "white", linewidth = 0.08) +
  stat_function(fun = dnorm, colour = book_palette$reject, linewidth = 0.6) +
  facet_grid(parent ~ n) +
  coord_cartesian(xlim = c(-3.2, 3.2), ylim = c(0, 0.52)) +
  labs(x = "standardised sample mean", y = NULL) +
  theme_book(base_size = 10) +
  theme(axis.text.y = element_blank(),
        strip.text.y = element_text(angle = 0, hjust = 0))
```

The first column shows the four underlying populations. None resembles a normal
distribution. By the third column, however, the sampling distributions of the
sample means all closely resemble the familiar bell-shaped curve.

The theorem guarantees eventual convergence to normality, but it says nothing
about the rate at which convergence occurs. Populations that are already fairly
symmetric require only modest sample sizes, whereas heavily skewed populations
may require substantially larger samples before the normal approximation
becomes accurate. In the figure, the uniform population is close to normal by
$n = 5$; the earnings distribution is still visibly asymmetric at $n = 30$.

### An intuition {-}

::: {.intuition}
Roll one die: every face is equally likely. Now roll two and add them.

```{r d2i-dice-sum}
sums <- outer(1:6, 1:6, "+")
table(sums)
```

Seven can be made six ways — $1+6$, $2+5$, $3+4$, $4+3$, $5+2$, $6+1$. Two can
be made exactly one way. The middle values have more routes to them than the
extremes do, so they turn up more often.

The intuition extends far beyond dice. To obtain an extreme average, every
observation must be extreme in the same direction, which is rare. To obtain a
middling average, the observations need only offset one another, and there are
a great many ways for that to happen. Adding observations multiplies the routes
to the middle far faster than the routes to the extremes, and the resulting
shape is the familiar bell.
:::

```{r d2i-dice-plot, echo = FALSE, fig.height = 3.2, fig.cap = "The sum of two dice. Combinations pile up in the middle, and the flat distribution of a single die has already started to become a mound."}
ds <- as.data.frame(table(sums)); names(ds) <- c("sum", "ways")
ds$sum <- as.numeric(as.character(ds$sum))

ggplot(ds, aes(sum, ways)) +
  geom_col(fill = book_palette$fill, width = 0.75) +
  geom_text(aes(label = ways), vjust = -0.5, size = 3.1,
            colour = book_palette$ink) +
  scale_x_continuous(breaks = 2:12) +
  labs(x = "sum of two dice", y = "number of ways") +
  theme_book() +
  theme(axis.text.y = element_blank())
```

### How large is "large enough"? {-}

Textbooks often give $n \geq 30$. It is a useful rule of thumb, not a
mathematical guarantee.

How quickly the approximation takes hold depends on how skewed the population
is. A symmetric population requires very few observations; a heavily skewed one
requires many more.

```{r d2i-clt-howbig}
set.seed(9)

# A crude measure of how far a distribution is from symmetric
skewness <- function(x) mean((x - mean(x))^3) / sd(x)^3

sapply(c(5, 30, 100, 500), function(n)
  round(skewness(replicate(2000, mean(sample(earn, n, replace = TRUE)))), 2))
```

```{r u1-clt-popskew, echo = FALSE}
pop_skew <- round(mean((earn - mean(earn))^3) / sd(earn)^3, 1)
```

The earnings population has a skewness of about `r pop_skew`. At $n = 5$ the
sampling distribution is still visibly lopsided. By $n = 100$ it is close to
symmetric.

::: {.watchout}
Many economic variables — including incomes, firm sizes, land holdings and city
populations — are strongly right-skewed. For such variables, the normal
approximation may still be poor when the sample size is thirty. In these
settings, larger samples or simulation-based checks of the sampling
distribution are often preferable.
:::

### Why inference is possible {-}

Together, the Law of Large Numbers and the Central Limit Theorem explain why
statistical inference is possible. The Law of Large Numbers tells us that the
sample mean is centred on the population mean as the sample size grows. The
Central Limit Theorem tells us that the sampling distribution of the sample
mean is approximately normal. Together, these results tell us where the
sampling distribution is centred, how it behaves, and how uncertainty can be
quantified.

::: {.keyidea}
The central idea of inferential statistics is that although different samples
produce different answers, those differences are not arbitrary. They follow
systematic patterns that can be described mathematically. By understanding how
statistics behave across repeated samples, we can quantify the uncertainty
associated with the single sample we actually observe.
:::

The ideas developed so far are largely intuitive. To understand them more
deeply — and to express them precisely — we now introduce the notation and
concepts that statisticians use to describe random phenomena.


## A language for randomness {#u1-random-variables}

::: {.question}
How do we talk about a quantity that changes every time we look?
:::

Return to the sentence from the first section: every roll of a die produces a
different outcome, every survey contains different households, every poll
interviews different voters.

A quantity whose value changes from one repetition to the next is called a
**random variable**. The die roll is one. A worker's earnings, drawn at random
from the population, is another. So — and this is the step that matters — is
the *mean* of a sample, because it too changes every time the sample changes.

::: {.keyidea}
The sample mean is itself a random variable.

This single realisation is what makes inference possible. If $\bar{X}$ is a
random variable, then it has a distribution. If it has a distribution, that
distribution has a centre and a spread. And if we can find them, we can say how
wrong a typical sample mean is.
:::

### The long-run average {-}

If a random variable takes value $x_i$ with probability $p(x_i)$, its
**expected value** is

$$E[X] = \sum_i x_i \, p(x_i)$$

The name invites a misunderstanding. $E[X]$ is not the value you expect to see
on any given try. For a fair die it is 3.5 — a number the die cannot produce.
It is the **long-run average** over many repetitions.

```{r u1-expectation}
mean(rolls)                # average of our 10,000 actual rolls
sum(1:6 * rep(1/6, 6))     # E[X], computed on paper
```

::: {.intuition}
That these agree is neither a coincidence nor a triviality. It is the first
appearance of the Law of Large Numbers: run a random process enough times, and
the average of what you observe closes in on the number you calculated in
advance.

Notice how unremarkable that felt. We are about to need it for something far
less obvious.
:::

When the random variable is a draw from a population, its expected value *is*
the population mean, which is why both are written $\mu$. And the spread of a
random variable around its expectation is its **variance**, written $\sigma^2$,
with the standard deviation $\sigma$ its square root.

### Population and sample, in notation {-}

::: {.keyidea}
A **parameter** describes the population: the true mean $\mu$, the true
standard deviation $\sigma$. You never observe these. They are written as Greek
letters as a permanent reminder that they are out of reach.

A **statistic** is computed from your sample: $\bar{x}$, $s$. You can always
see these. They are your only evidence.

The whole problem of inference, in one line:

$$\underbrace{\bar{x}}_{\text{you have this}} \quad\longrightarrow\quad \underbrace{\mu}_{\text{you want this}}$$

The arrow is not an equals sign.
:::

### The assumption that licenses everything {-}

One more thing must be said before the machinery works, and it is usually said
too quickly.

We assume the population has an underlying distribution, and that the members
of our sample are **independent and identically distributed** draws from it.

- **Identically distributed**: every observation comes from the same
  distribution. No drift, no changing the rules partway through.
- **Independent**: knowing one observation tells you nothing about another.

::: {.watchout}
It is this assumption — that your sample is drawn from the same distribution as
the population you care about — that licenses inference at all.

If the sample came from a *different* distribution, no amount of arithmetic
could bridge them. This is why sampling design matters so much, and why a badly
drawn sample cannot be repaired by clever statistics afterwards. The formulas
will still run. They will simply be answering a question about a population you
did not mean.
:::

## A population where we know the answer {#u1-sampling-distribution}

::: {.question}
Can we check that any of this actually works?
:::

The earnings data has a drawback for learning. We treated it as a population,
but in a real study we would never see the population at all, and so could
never confirm that a sampling distribution behaves as claimed.

The remedy is to build a population small enough that its true mean can be
written down exactly, and then see whether the machinery recovers it.

Take a deck of cards, scored A = 1, J = 11, Q = 12, K = 13, and 2 through 10 at
face value.

```{r u1-deck}
deck <- rep(1:13, times = 4)

c(N = length(deck), mu = mean(deck), sigma = sd(deck))
```

The population mean is exactly $\mu = 7$. Now draw ten cards and average them:

```{r u1-onesample}
set.seed(7)
mean(sample(deck, size = 10, replace = TRUE))
```

::: {.watchout}
Note `replace = TRUE`: each card goes back before the next draw.

That is not a technicality. It is what makes the draws **independent**, the
assumption we just spent a page on. Draw *without* replacement and each card
removed changes what is left, so the draws are dependent and the sample mean
varies slightly less than theory predicts.

It is also the better model of what we usually mean. Surveying 2,000 people
does not meaningfully deplete a population of a billion.
:::

The result is not 7, and a second draw would give something else again. This is
sampling variation, exactly as observed with the earnings surveys. Repeating
the draw a thousand times gives the whole sampling distribution:

```{r u1-sampdist}
set.seed(7)
deck_means <- replicate(1000, mean(sample(deck, size = 10, replace = TRUE)))

c(mean_of_deck_means = mean(deck_means),
  sd_of_deck_means   = sd(deck_means))
```

```{r u1-sampdist-plot, echo = FALSE, fig.height = 3.6, fig.cap = "One thousand sample means, each from 10 cards. Individual samples miss the population mean constantly. Their average lands almost exactly on it."}
ggplot(data.frame(m = deck_means), aes(m)) +
  geom_histogram(bins = 32, fill = book_palette$fill, colour = "white",
                 linewidth = 0.2) +
  geom_vline(xintercept = 7, colour = book_palette$reject, linewidth = 0.9) +
  annotate("label", x = 7, y = Inf, vjust = 1.4, label = "population mean = 7",
           colour = book_palette$reject, fill = "white", size = 3.5,
           label.size = 0.3) +
  labs(x = "sample mean of 10 cards", y = "how often") +
  theme_book()
```

Two things happened, and both matter.

**The sample means are centred on the population mean.** No single sample got
it right, but their average is `r round(mean(deck_means), 3)` — essentially
7. The sample mean is an **unbiased** estimator: not right, but not
*systematically* wrong, which is a different and far more useful property.

**The sample means vary much less than the cards do.** The population standard
deviation is about `r round(sd(rep(1:13, 4)), 2)`; the sample means have a
standard deviation of about `r round(sd(deck_means), 2)`. Averaging has
cancelled out much of the noise.

### What happens as the sample grows {-}

```{r u1-nsizes}
set.seed(7)

sizes <- c(5, 10, 20, 40, 80)
results <- t(sapply(sizes, function(n) {
  m <- replicate(2000, mean(sample(deck, size = n, replace = TRUE)))
  c(n = n, mean = mean(m), sd = sd(m), theoretical_sd = sd(deck) / sqrt(n))
}))

round(results, 3)
```

```{r u1-nsizes-plot, echo = FALSE, fig.height = 4.4, fig.cap = "Sampling distributions of the mean for growing sample sizes. The centre never moves. The spread collapses."}
set.seed(7)
dd <- do.call(rbind, lapply(c(5, 20, 80), function(n) {
  data.frame(n = paste("n =", n),
             m = replicate(2000, mean(sample(deck, n, replace = TRUE))))
}))
dd$n <- factor(dd$n, levels = paste("n =", c(5, 20, 80)))

ggplot(dd, aes(m)) +
  geom_histogram(bins = 40, fill = book_palette$fill, colour = "white",
                 linewidth = 0.15) +
  geom_vline(xintercept = 7, colour = book_palette$reject, linewidth = 0.8) +
  facet_wrap(~ n, ncol = 1, scales = "free_y") +
  labs(x = "sample mean", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank())
```

Every row is centred on 7 — the sample mean is unbiased at any sample size. But
the spread shrinks steadily, and in a completely predictable way. Compare the
last two columns of the table: the observed standard deviation of the sample
means matches $\sigma/\sqrt{n}$ almost exactly.

::: {.keyidea}
Two results, both of which can be established directly:

$$E(\bar{X}) = \mu \qquad\qquad \mathrm{Var}(\bar{X}) = \frac{\sigma^2}{n}$$

The sample mean is centred on the population mean, and its spread falls as the
sample grows. The quantity $\sigma/\sqrt{n}$ is called the **standard error**
of the sample mean.
:::

### Where these results come from {-}

Neither result requires anything beyond the definition of the sample mean.

::: {.theory}
**The sample mean is centred on the population mean.**

$$
\begin{aligned}
E(\bar{X}) &= E\!\left(\frac{X_1 + X_2 + \cdots + X_n}{n}\right) \\[4pt]
&= \frac{1}{n}\Big(E(X_1) + E(X_2) + \cdots + E(X_n)\Big) \\[4pt]
&= \frac{1}{n}\underbrace{(\mu + \mu + \cdots + \mu)}_{n \text{ times}} \\[4pt]
&= \frac{n\mu}{n} \;=\; \mu
\end{aligned}
$$

Every observation is drawn from the same distribution, so each has expectation
$\mu$. Averaging $n$ of them leaves $\mu$ unchanged. A statistic with this
property is described as **unbiased**: it is not systematically too high or too
low, although any particular value of it will still differ from $\mu$.
:::

::: {.theory}
**Its spread falls as the sample grows.**

$$
\begin{aligned}
\mathrm{Var}(\bar{X}) &= \mathrm{Var}\!\left(\frac{1}{n}(X_1 + \cdots + X_n)\right) \\[4pt]
&= \frac{1}{n^2}\,\mathrm{Var}(X_1 + \cdots + X_n) \\[4pt]
&= \frac{1}{n^2}\underbrace{(\sigma^2 + \sigma^2 + \cdots + \sigma^2)}_{n \text{ times}} \\[4pt]
&= \frac{n\sigma^2}{n^2} \;=\; \frac{\sigma^2}{n}
\end{aligned}
$$

The second line uses $\mathrm{Var}(aX) = a^2\mathrm{Var}(X)$. The third uses
**independence**: the variance of a sum equals the sum of the variances only
when the observations do not influence one another. This is the single point at
which the independence assumption is required.
:::

Together these two results account for what Figure \@ref(fig:u1-lln-plot)
showed. The sampling distribution is centred on $\mu$ at every sample size,
and its width is $\sigma/\sqrt{n}$, which falls as $n$ grows. The Law of
Large Numbers is the consequence: a distribution centred on the right value,
whose spread tends to zero, must eventually concentrate on that value.

::: {.watchout}
Note the square root. To halve your standard error you must **quadruple** your
sample. Precision is expensive, and it grows more expensive the more of it you
already have.

This is why surveys of two thousand people are common and surveys of two lakh
are not: the second costs a hundred times as much for about seven times the
precision.
:::


## The average is not the distribution {#u1-average}

::: {.question}
And what does an average hide?
:::

Before leaving this unit, one qualification deserves attention.

Almost all of inferential statistics concerns averages. Most of the methods in
this book estimate a mean, compare two means, or fit a line through conditional
means. That emphasis is not arbitrary: means are well behaved, their sampling
distributions are tractable, and both of the results we have just studied are
statements about them.

The mean, however, is only one feature of a distribution, and frequently not
the feature that a policy question concerns.

```{r u1-skew}
c(mean   = mean(earn),
  median = median(earn))

mean(earn) / median(earn)
```

The mean is more than **twice** the median. When those two diverge that
sharply, the average has stopped describing a typical worker.

```{r u1-skew-plot, echo = FALSE, fig.height = 4, fig.cap = "Annual earnings of 20,000 Indian workers. The mean sits far to the right of most people, dragged there by a thin tail of high earners."}
ggplot(workers, aes(annual_earnings / 1000)) +
  geom_histogram(bins = 60, fill = book_palette$fill, colour = "white",
                 linewidth = 0.15) +
  geom_vline(xintercept = median(earn) / 1000,
             colour = book_palette$ink, linetype = "dashed", linewidth = 0.7) +
  geom_vline(xintercept = mean(earn) / 1000,
             colour = book_palette$reject, linewidth = 0.9) +
  annotate("label", x = mean(earn) / 1000, y = Inf, vjust = 1.4,
           label = "mean", colour = book_palette$reject, fill = "white",
           size = 3.5, label.size = 0.3) +
  annotate("label", x = median(earn) / 1000, y = Inf, vjust = 3.2,
           label = "median", colour = book_palette$ink, fill = "white",
           size = 3.5, label.size = 0.3) +
  coord_cartesian(xlim = c(0, 400)) +
  labs(x = "annual earnings (₹ thousand)", y = "number of workers") +
  theme_book()
```

How concentrated is it?

```{r u1-concentration}
top10 <- earn[earn >= quantile(earn, 0.90)]

round(100 * sum(top10) / sum(earn))              # % of all earnings, top tenth
round(quantile(earn, 0.90) / median(earn), 1)    # top-decile threshold vs median
```

The best-paid tenth of workers take a large share of all earnings, and the
threshold for entering that tenth is several times the median wage.

::: {.intuition}
Consider what "average earnings in India" is able to answer.

For total earnings it is exact: the mean multiplied by the number of workers is
the total, by construction.

For the earnings of a typical worker it is misleading. Very few people earn
anything close to the mean, and most earn considerably less.

For whether a policy has helped it can be actively deceptive. If the gains
accrue at the top, the mean may rise while the majority are worse off. **An
average can move in the opposite direction to almost everyone within it.**
:::

::: {.keyidea}
Ask two questions of every average you meet, including your own:

1. **What is the spread?** A mean with no measure of dispersion is half a
   sentence.
2. **Who is where?** Means are the wrong tool for questions about the bottom of
   a distribution — poverty, vulnerability, who is hurt by a shock. Those are
   questions about *tails*, and the mean is precisely the statistic that
   averages tails away.
:::

This is not a philosophical aside; it changes what you estimate. Whole
literatures in development economics exist because the mean was the wrong
target. Poverty measures weight the poor rather than averaging over everyone.
Vulnerability measures ask who is at risk of falling below a line.
Distributional analysis asks whether a policy that raised average income raised
the income of anyone who was badly off to begin with.

We will return to this repeatedly. When we fit a regression we will be
modelling a conditional *mean*, and it will be worth asking each time what that
choice conceals. When we reach binary explanatory variables and estimate an
average treatment effect, the word "average" in that phrase will deserve every
bit of suspicion you can bring to it.

::: {.watchout}
None of this makes averages bad. They are the right tool for a great many
questions, and this book spends most of its length on them for good reason.

It means an average is an **answer to a specific question**, not a neutral
summary of the data. Decide which question you are asking before deciding the
mean answers it.
:::

## What can go wrong before any mathematics {#u1-what-goes-wrong}

::: {.question}
When does none of this machinery help?
:::

Inference assumes your sample carries information about the population you care
about. Three ways that fails, none of which any formula will warn you about.

**The sample is not from the population you mean.** The ASI covers *registered*
factories. Most Indian manufacturing employment is unregistered. Conclusions
about "Indian manufacturing" drawn from ASI data are conclusions about a formal
sliver of it. The arithmetic will not object.

**The sample is not random.** Our class survey has 32 heights. They are not a
random sample of APU students, let alone of young Indians — they are the people
who took this course and turned up that day. You may compute a confidence
interval from them and it will mean nothing about the university.

**The question is causal but the data is not.** Suppose states with more
industry have higher literacy. Does industry raise literacy? Does literacy
attract industry? Does something else drive both? The correlation is identical
in all three cases, and no statistical machinery separates them without an
argument about *how the data came to exist*.

::: {.intuition}
That last failure is why this book is organised as it is. From the unit on
two-sample tests onwards, we stop asking only "what is this number?" and start
asking "what would have to be true for this number to mean what I want it to
mean?"

That question — not any particular estimator — is what separates a good applied
economist from someone who can run regressions.
:::

## Summary {#u1-summary}

The argument of this unit can be stated in a few sentences.

Observations vary, and they vary because the processes that generate them
involve chance. Descriptive statistics summarises the variation in the data we
have, accurately and completely. But the data we have is a sample, and a
different sample would have produced different summaries. That difference —
sampling variation — is not a defect in the data, and cannot be removed by
working more carefully.

The resolution is to change the object of study. Instead of asking what a
particular sample says, we ask how a statistic behaves across the many samples
that might have been drawn. The pattern formed by a statistic under repeated
sampling is its sampling distribution, and it is the foundation on which
inference rests.

Two results describe that pattern. The Law of Large Numbers establishes that
the sample mean approaches the population mean as the sample grows. The Central
Limit Theorem establishes that the sampling distribution of the sample mean is
approximately normal, whatever the shape of the population — although how large
a sample this requires depends on how skewed that population is.

::: {.keyidea}
Points worth carrying forward:

1. A distribution is a pattern that emerges from repetition under randomness.
   Where nothing varies, there is nothing for statistics to describe.
2. Greek letters ($\mu$, $\sigma$) denote population quantities, which are
   never observed. Roman letters ($\bar{x}$, $s$) denote quantities computed
   from a sample.
3. Inference relies on the sample being drawn from the same distribution as the
   population. A poorly drawn sample cannot be repaired by later analysis.
4. $E(\bar{X}) = \mu$ and $\mathrm{Var}(\bar{X}) = \sigma^2/n$. The sample
   mean is unbiased, and its standard error $\sigma/\sqrt{n}$ falls with the
   square root of the sample size — so halving it requires quadrupling the
   sample.
5. The mean is one feature of a distribution, not a summary of it. Indian
   earnings have a mean more than twice their median. Spread, and the position
   of those in the tails, are separate questions requiring separate tools.
:::

This unit opened by claiming that the average computed from a sample is almost
certainly wrong. It is — and we can now say how wrong, and with what
probability.

The next unit puts that knowledge to use. Knowing that the sample mean is
centred on $\mu$, has standard error $\sigma/\sqrt{n}$, and is approximately
normally distributed, a single sample can be turned into a range of plausible
values for the population mean, accompanied by a stated level of confidence.
That is the subject of estimation.
