# Towards Inferential Statistics {#towards-inference}

```{r setup-01, include = FALSE}
source("_common.R")
library(ggplot2)
```

You already know how to calculate the average of a sample.

In this unit, you will learn why that average is almost certainly wrong.

At first glance, this seems like a strange claim. After all, there is only one
way to calculate a sample mean, and if the arithmetic is correct, the answer
must be correct as well. Yet the number you obtain from a sample is almost
never exactly equal to the quantity you ultimately wish to know — the mean of
the population from which that sample was drawn.

Suppose you wish to estimate the average income of households in India. Since
surveying every household is impossible, you select a sample of 2,000
households and calculate its mean income. The calculation is straightforward,
but the result immediately raises a question. How close is this sample mean to
the true population mean?

The difficulty is that the sample you have drawn is only one of many that could
have been observed. Had a different 2,000 households been selected, the sample
mean would almost certainly have been different. Draw another sample, and the
answer changes again. Every statistic calculated from a sample therefore
reflects two things at once: the characteristics of the underlying population,
and the randomness introduced by the sampling process.

This simple observation marks the boundary between descriptive and inferential
statistics.

Until now, our objective has been to describe the data that we observe.
Measures of central tendency, such as the mean and the median, summarise where
observations are concentrated. Measures of dispersion, such as the variance and
standard deviation, describe how widely they are spread. Histograms, bar charts
and density plots reveal the overall shape of a distribution and help us
identify skewness, outliers and other important features. Together, these tools
provide a concise description of a dataset and are the natural starting point
of every empirical investigation.

Most empirical research, however, is concerned not with the sample itself but
with the population from which it was drawn. Economists survey a few thousand
households to understand millions. Polling agencies interview a sample of
voters to estimate national voting intentions. Medical researchers evaluate new
treatments on a limited number of patients before recommending them to the
wider population. In every case, the sample serves as evidence about a
population that cannot be observed directly.

The central problem of inferential statistics is therefore straightforward to
state: **how can we use the information contained in a sample to learn about
the population from which it came?**

Throughout this course we adopt the **frequentist** approach to answering this
question. Rather than treating the observed sample as unique, we imagine
drawing many samples from the same population and calculating the same
statistic repeatedly. Some sample means will be larger than others, some
smaller. Understanding how statistics vary across these hypothetical samples
allows us to assess how much confidence we should place in the estimate
computed from the one sample we actually observe.

Before we can understand why, we need to understand where the uncertainty comes
from. The answer lies in randomness.

## Randomness and variation {#u1-randomness}

::: {.question}
Where does the uncertainty come from?
:::

Statistics begins with an observation so ordinary that it is easy to walk past:
**identical experiments rarely produce identical outcomes.**

Roll a die.

```{r u1-die-one}
set.seed(1)
sample(1:6, size = 1)
```

A three. That is all. One roll tells you almost nothing — not whether the die
is fair, not what to expect next, not anything you could act on.

Now roll it ten thousand times and count how often each face appears.

```{r u1-die-many}
rolls <- sample(1:6, size = 10000, replace = TRUE)

round(prop.table(table(rolls)), 3)
```

```{r u1-die-plot, echo = FALSE, fig.height = 3.2, fig.cap = "Ten thousand rolls of a fair die. No single roll was predictable. The pattern across all of them is."}
ggplot(data.frame(x = factor(rolls)), aes(x)) +
  geom_bar(aes(y = after_stat(count / sum(count))),
           fill = book_palette$fill, width = 0.7) +
  geom_hline(yintercept = 1/6, colour = book_palette$reject,
             linetype = "dashed", linewidth = 0.7) +
  annotate("text", x = 6.4, y = 1/6, label = "1/6", vjust = -0.8,
           colour = book_palette$reject, size = 3.6) +
  labs(x = "outcome", y = "share of rolls") +
  theme_book()
```

Something has appeared that was not there before. Each face turns up close to
one-sixth of the time. No individual roll was predictable, yet the *pattern
across many rolls* is highly predictable — so much so that you would be
surprised if it came out otherwise.

That pattern is called a **distribution**. It is not a property of any single
roll. It only exists across repetition.

Now take the randomness away. Imagine a die with a 3 painted on all six faces.

```{r u1-die-loaded}
loaded <- rep(3, 10000)

table(loaded)
sd(loaded)
```

Ten thousand rolls, one outcome, spread of exactly zero. There is no pattern to
find, because there is nothing to be uncertain about. Every question you might
ask has the same answer. What would a histogram of this show? A single bar.
What is the average? Three. What is the chance of a three? One.

::: {.keyidea}
**Randomness creates variation. Variation is the raw material of statistics.**

Where outcomes vary, there is a pattern to describe, a spread to measure and an
uncertainty to quantify. Where they do not, there is nothing for statistics to
do.

Every technique in this book is a way of handling variation. Take the variation
away and the entire subject falls silent.
:::

This is not a fact about dice. Every roll of a die produces a different
outcome. Every household survey contains different households. Every election
poll interviews different voters. Every harvest yields a different quantity.
Statistics exists because the world is variable — and it is variable whether or
not we are looking.

## Describing variation {#u1-describing}

::: {.question}
Once variation exists, how do we describe it?
:::

A die has six outcomes and you can list them. Real data does not work like
that.

Here are the annual earnings of twenty thousand Indian workers.

```{r u1-earnings}
workers <- read.csv("data/wages-india-synthetic.csv")

nrow(workers)
head(workers$annual_earnings)
```

Twenty thousand numbers, no two households alike. Reading them one at a time
would take a day and teach you nothing. We need to compress.

::: {.watchout}
A word about this data. The Indian Human Development Survey records exactly
these variables for real households, and it is the natural source for a book
like this one. It cannot be redistributed, so the file used here is a
**simulation calibrated to IHDS-II** — the return to a year of education, the
gender gap, the urban premium and the shape of the earnings distribution all
match the real survey to within a rounding error.

Everything you learn from it transfers. But it is simulated, and you should
never be left guessing about that. See `data/README.md`.
:::

The first thing to do with any dataset is look at it.

```{r u1-earnings-hist, fig.height = 3.6, fig.cap = "Annual earnings of 20,000 Indian workers. Most people are bunched at the left; a thin tail runs a long way right."}
ggplot(workers, aes(annual_earnings / 1000)) +
  geom_histogram(bins = 60, fill = book_palette$fill, colour = "white",
                 linewidth = 0.15) +
  scale_x_continuous(labels = scales::comma) +
  labs(x = "annual earnings (₹ thousand)", y = "number of workers") +
  theme_book()
```

The shape tells you a great deal at a glance. Earnings are not spread evenly.
Most workers are packed into the left, and a small number stretch far to the
right. The distribution is **skewed**.

Now the summaries. Where is the centre?

```{r u1-centre}
c(mean   = mean(workers$annual_earnings),
  median = median(workers$annual_earnings))
```

Two answers, and they disagree — the mean is more than twice the median. The
**mean** balances every observation against the others, so a handful of very
high earners pull it upward. The **median** simply splits the group in half,
and does not care how far the top of the distribution extends.

How spread out are they?

```{r u1-spread}
c(sd  = sd(workers$annual_earnings),
  iqr = IQR(workers$annual_earnings))
```

The **standard deviation** measures the typical distance of an observation from
the mean. The **interquartile range** gives the width of the middle half. Both
say the same thing here: the spread is enormous relative to the centre.

And a smooth version of the shape, which makes comparison across groups easy:

```{r u1-density, fig.height = 3.4, fig.cap = "The same data as a density curve, split by where people live. The urban distribution sits to the right of the rural one throughout."}
ggplot(workers, aes(annual_earnings / 1000, colour = residence)) +
  geom_density(linewidth = 0.8) +
  scale_colour_manual(values = c(Rural = book_palette$ink,
                                 Urban = book_palette$accent), name = NULL) +
  coord_cartesian(xlim = c(0, 400)) +
  labs(x = "annual earnings (₹ thousand)", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank(), legend.position = "top")
```

Look at what has happened. Twenty thousand numbers have become a picture and
four figures: a centre, a spread, a shape, and a comparison between two groups.
Nothing has been invented and nothing has been assumed. We have simply replaced
tens of thousands of observations with a handful of carefully chosen summaries.

This is **descriptive statistics**: the business of summarising the data in
front of you so that a human mind can hold it.

It is genuinely useful and it is always the first thing to do. Trends,
outliers, data-entry errors and interesting patterns all surface here, and no
amount of later sophistication rescues an analysis that skipped this step.

## Why description isn't enough {#u1-not-enough}

::: {.question}
Why isn't description enough?
:::

Here is the question that changes everything.

Those twenty thousand workers were not chosen by anyone in particular. A survey
team went out, followed a sampling design, and ended up with these people
rather than others. Suppose they had gone out on a different week and found a
different twenty thousand.

**Would the histogram look the same?**

Let us find out. Treat the full file as the group we could have drawn from, and
take three separate samples of two thousand workers each — three parallel
universes in which the survey went slightly differently.

```{r u1-three-samples}
set.seed(1)

s1 <- workers$annual_earnings[sample(nrow(workers), 2000)]
s2 <- workers$annual_earnings[sample(nrow(workers), 2000)]
s3 <- workers$annual_earnings[sample(nrow(workers), 2000)]
```

```{r u1-three-plot, echo = FALSE, fig.height = 4.4, fig.cap = "Three surveys of two thousand workers each, drawn from the same group. Same shape. Not the same numbers."}
three <- rbind(
  data.frame(survey = "Survey 1", x = s1),
  data.frame(survey = "Survey 2", x = s2),
  data.frame(survey = "Survey 3", x = s3))

means <- aggregate(x ~ survey, three, mean)

ggplot(three, aes(x / 1000)) +
  geom_histogram(bins = 45, fill = book_palette$fill, colour = "white",
                 linewidth = 0.12) +
  geom_vline(data = means, aes(xintercept = x / 1000),
             colour = book_palette$reject, linewidth = 0.8) +
  facet_wrap(~ survey, ncol = 1) +
  coord_cartesian(xlim = c(0, 500)) +
  labs(x = "annual earnings (₹ thousand)", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank())
```

The shapes are recognisably the same. Skewed, bunched at the left, long tail
right. Nothing about the picture would lead you to say one of these surveys got
it wrong.

**Would the mean change?**

```{r u1-three-means}
c(survey1 = mean(s1), survey2 = mean(s2), survey3 = mean(s3))
```

Yes. Three surveys, three answers, differing by thousands of rupees.

**Would the median change?**

```{r u1-three-medians}
c(survey1 = median(s1), survey2 = median(s2), survey3 = median(s3))
```

Yes. And the standard deviation:

```{r u1-three-sds}
c(survey1 = sd(s1), survey2 = sd(s2), survey3 = sd(s3))
```

Yes again. Every summary we computed in the last section — every single one —
comes out differently depending on who happened to be surveyed.

So here is the question to sit with:

::: {.question}
Which one is the correct mean?
:::

None of them.

Each survey was conducted properly. Each calculation is arithmetically perfect.
And each gives a different answer. There is no error to find and nobody to
blame. The numbers differ because *the people differ*, and the people differ
because they were selected by a process with randomness in it.

::: {.keyidea}
Now the vocabulary, for ideas you have already met.

The **population** is the entire group you want to know about — every worker in
India, not the two thousand in your file.

The **sample** is the part you actually observed.

**Sampling variation** is what you have just watched: the fact that a statistic
computed from a sample changes when the sample changes.

Sampling variation is not a mistake. It is not caused by sloppy work and it
cannot be removed by being careful. It is a permanent feature of learning about
a large group by looking at a small part of it.
:::

This is why description is not enough. Descriptive statistics answers questions
about the data in front of you, perfectly and with certainty. But the question
you actually care about — *what are earnings like in India?* — is about the
population, and your sample is one of countless samples that might have turned
up instead.

Your average is almost certainly wrong because it is one draw from a process
that could have handed you a different answer.

## What if we repeated the survey? {#u1-frequentist}

::: {.question}
What happens if we repeat the sample?
:::

We saw three surveys give three answers. That is unsettling, but it also
contains a clue. Three is a small number. What if we could see more?

Imagine running the survey again. And again. And again. Not three times but a
thousand times, each with a fresh two thousand workers, each producing its own
mean. In reality this is impossible — nobody has the budget. On a computer it
takes a moment.

```{r u1-many-surveys}
set.seed(2)

many_means <- replicate(1000, mean(sample(workers$annual_earnings, 2000)))

head(round(many_means), 8)
```

A thousand surveys, a thousand different answers. Now stop looking at them one
at a time and look at all of them together.

```{r u1-many-plot, echo = FALSE, fig.height = 3.6, fig.cap = "A thousand surveys, each of two thousand workers. Individual surveys scatter. Collectively they form a tight, symmetric, entirely predictable pattern."}
ggplot(data.frame(m = many_means / 1000), aes(m)) +
  geom_histogram(bins = 45, fill = book_palette$fill, colour = "white",
                 linewidth = 0.15) +
  geom_vline(xintercept = mean(workers$annual_earnings) / 1000,
             colour = book_palette$reject, linewidth = 0.9) +
  annotate("label", x = mean(workers$annual_earnings) / 1000, y = Inf,
           vjust = 1.4, label = "true mean", colour = book_palette$reject,
           fill = "white", size = 3.5, label.size = 0.3) +
  labs(x = "survey mean (₹ thousand)", y = "number of surveys") +
  theme_book()
```

Look at that shape.

The individual earnings we started with were violently skewed — most people
bunched left, a few stretching far right. The *survey means* are nothing like
that. They are symmetric. They are tightly clustered. They pile up around the
true value and thin out smoothly on both sides. And notice how narrow the
spread is: individual earnings run from near zero into the millions, while the
survey means barely stray from one another.

::: {.intuition}
This is the turn the whole subject rests on, so it is worth saying slowly.

We stopped asking *what does my sample say?* and started asking *how do samples
behave?*

One sample mean, on its own, is a number with no error attached — you cannot
tell from it alone how far off it might be. But the *collection* of sample
means has a shape, and that shape is regular enough to describe. If we can
describe it, we can say how far a typical sample mean falls from the truth.

And if we can say that, then we can take our one real survey and attach an
honest margin of error to it — even though we will only ever run it once.
:::

That picture — the distribution of a statistic across repeated samples — has a
name, and we will come to it properly in a moment. For now the important thing
is not the name but the move: **from studying one sample to studying the
behaviour of samples.**

This way of reasoning is called the **frequentist** approach. It answers
questions about uncertainty by asking what would happen in the long run, over
many repetitions, even when only one repetition will ever actually occur.

## Where we are going {#u1-roadmap}

::: {.question}
Can we predict how those averages behave?
:::

We now have the question in its sharpest form. Survey means vary. Their
variation has a shape. If we could predict that shape *in advance* — without
running a thousand surveys, from the single sample we actually have — we could
say precisely how much to trust our answer.

That is exactly what the rest of this unit does, and it needs three things.

**Probability**, first, because we need a language for describing randomness.
We have been using words like "varies" and "typical" loosely. To get numbers we
need to be precise about what a random quantity is, what its long-run average
means, and how to measure its spread.

**The sampling distribution**, next, defined properly rather than merely drawn.
We will find that its centre and its spread can both be worked out on paper,
from facts about a single sample.

**Two theorems**, finally, which are the reason any of this works. The **Law of
Large Numbers** explains why sample averages close in on the truth as samples
grow. The **Central Limit Theorem** explains why the picture above was
bell-shaped — and, remarkably, why it would have been bell-shaped no matter
what the earnings distribution looked like.

Together they turn "your average is almost certainly wrong" into something far
more useful: *your average is wrong by about this much, with about this
probability.* That sentence is what separates an estimate from an anecdote, and
everything in the rest of this book is built from it.

## A language for randomness {#u1-random-variables}

::: {.question}
How do we talk about a quantity that changes every time we look?
:::

Return to the sentence from the first section: every roll of a die produces a
different outcome, every survey contains different households, every poll
interviews different voters.

A quantity whose value changes from one repetition to the next is called a
**random variable**. The die roll is one. A worker's earnings, drawn at random
from the population, is another. So — and this is the step that matters — is
the *mean* of a sample, because it too changes every time the sample changes.

::: {.keyidea}
The sample mean is itself a random variable.

This single realisation is what makes inference possible. If $\bar{X}$ is a
random variable, then it has a distribution. If it has a distribution, that
distribution has a centre and a spread. And if we can find them, we can say how
wrong a typical sample mean is.
:::

### The long-run average {-}

If a random variable takes value $x_i$ with probability $p(x_i)$, its
**expected value** is

$$E[X] = \sum_i x_i \, p(x_i)$$

The name invites a misunderstanding. $E[X]$ is not the value you expect to see
on any given try. For a fair die it is 3.5 — a number the die cannot produce.
It is the **long-run average** over many repetitions.

```{r u1-expectation}
mean(rolls)                # average of our 10,000 actual rolls
sum(1:6 * rep(1/6, 6))     # E[X], computed on paper
```

::: {.intuition}
That these agree is neither a coincidence nor a triviality. It is the first
appearance of the Law of Large Numbers: run a random process enough times, and
the average of what you observe closes in on the number you calculated in
advance.

Notice how unremarkable that felt. We are about to need it for something far
less obvious.
:::

When the random variable is a draw from a population, its expected value *is*
the population mean, which is why both are written $\mu$. And the spread of a
random variable around its expectation is its **variance**, written $\sigma^2$,
with the standard deviation $\sigma$ its square root.

### Population and sample, in notation {-}

::: {.keyidea}
A **parameter** describes the population: the true mean $\mu$, the true
standard deviation $\sigma$. You never observe these. They are written as Greek
letters as a permanent reminder that they are out of reach.

A **statistic** is computed from your sample: $\bar{x}$, $s$. You can always
see these. They are your only evidence.

The whole problem of inference, in one line:

$$\underbrace{\bar{x}}_{\text{you have this}} \quad\longrightarrow\quad \underbrace{\mu}_{\text{you want this}}$$

The arrow is not an equals sign.
:::

### The assumption that licenses everything {-}

One more thing must be said before the machinery works, and it is usually said
too quickly.

We assume the population has an underlying distribution, and that the members
of our sample are **independent and identically distributed** draws from it.

- **Identically distributed**: every observation comes from the same
  distribution. No drift, no changing the rules partway through.
- **Independent**: knowing one observation tells you nothing about another.

::: {.watchout}
It is this assumption — that your sample is drawn from the same distribution as
the population you care about — that licenses inference at all.

If the sample came from a *different* distribution, no amount of arithmetic
could bridge them. This is why sampling design matters so much, and why a badly
drawn sample cannot be repaired by clever statistics afterwards. The formulas
will still run. They will simply be answering a question about a population you
did not mean.
:::

## The sampling distribution {#u1-sampling-distribution}

::: {.question}
What exactly was that picture of a thousand survey means?
:::

Now we can name it properly.

The distribution of a **random variable** describes how raw observations vary —
how earnings differ from worker to worker.

A **sampling distribution** describes how a *statistic* varies across repeated
samples of a fixed size — how the survey mean differs from survey to survey.

::: {.keyidea}
These are two different distributions and confusing them is the most common
error in this material.

The earnings of 20,000 workers form one distribution: skewed, wide, running
from near zero into the millions.

The means of 1,000 surveys form quite another: symmetric, narrow, clustered
around the truth.

The second is not a summary of the first. It is a distribution of a *different
quantity* — not of people, but of surveys.
:::

### A worked example we can check {-}

Real data has a drawback for learning: we never know the true answer, so we
cannot tell how well the machinery is doing. So build a population where the
truth is known exactly.

Take a deck of cards, scored A = 1, J = 11, Q = 12, K = 13, and 2 through 10 at
face value.

```{r u1-deck}
deck <- rep(1:13, times = 4)

c(N = length(deck), mu = mean(deck), sigma = sd(deck))
```

The population mean is exactly $\mu = 7$. Now draw ten cards and average them:

```{r u1-onesample}
set.seed(7)
mean(sample(deck, size = 10, replace = TRUE))
```

::: {.watchout}
Note `replace = TRUE`: each card goes back before the next draw.

That is not a technicality. It is what makes the draws **independent**, the
assumption we just spent a page on. Draw *without* replacement and each card
removed changes what is left, so the draws are dependent and the sample mean
varies slightly less than theory predicts.

It is also the better model of what we usually mean. Surveying 2,000 people
does not meaningfully deplete a population of a billion.
:::

Not 7. Draw again and you would get something else again — sampling variation,
exactly as with the earnings surveys. So look at the whole sampling
distribution:

```{r u1-sampdist}
set.seed(7)
sample_means <- replicate(1000, mean(sample(deck, size = 10, replace = TRUE)))

c(mean_of_sample_means = mean(sample_means),
  sd_of_sample_means   = sd(sample_means))
```

```{r u1-sampdist-plot, echo = FALSE, fig.height = 3.6, fig.cap = "One thousand sample means, each from 10 cards. Individual samples miss the population mean constantly. Their average lands almost exactly on it."}
ggplot(data.frame(m = sample_means), aes(m)) +
  geom_histogram(bins = 32, fill = book_palette$fill, colour = "white",
                 linewidth = 0.2) +
  geom_vline(xintercept = 7, colour = book_palette$reject, linewidth = 0.9) +
  annotate("label", x = 7, y = Inf, vjust = 1.4, label = "population mean = 7",
           colour = book_palette$reject, fill = "white", size = 3.5,
           label.size = 0.3) +
  labs(x = "sample mean of 10 cards", y = "how often") +
  theme_book()
```

Two things happened, and both matter.

**The sample means are centred on the population mean.** No single sample got
it right, but their average is `r round(mean(sample_means), 3)` — essentially
7. The sample mean is an **unbiased** estimator: not right, but not
*systematically* wrong, which is a different and far more useful property.

**The sample means vary much less than the cards do.** The population standard
deviation is about `r round(sd(rep(1:13, 4)), 2)`; the sample means have a
standard deviation of about `r round(sd(sample_means), 2)`. Averaging has
cancelled out much of the noise.

### What happens as the sample grows {-}

```{r u1-nsizes}
set.seed(7)

sizes <- c(5, 10, 20, 40, 80)
results <- t(sapply(sizes, function(n) {
  m <- replicate(2000, mean(sample(deck, size = n, replace = TRUE)))
  c(n = n, mean = mean(m), sd = sd(m), theoretical_sd = sd(deck) / sqrt(n))
}))

round(results, 3)
```

```{r u1-nsizes-plot, echo = FALSE, fig.height = 4.4, fig.cap = "Sampling distributions of the mean for growing sample sizes. The centre never moves. The spread collapses."}
set.seed(7)
dd <- do.call(rbind, lapply(c(5, 20, 80), function(n) {
  data.frame(n = paste("n =", n),
             m = replicate(2000, mean(sample(deck, n, replace = TRUE))))
}))
dd$n <- factor(dd$n, levels = paste("n =", c(5, 20, 80)))

ggplot(dd, aes(m)) +
  geom_histogram(bins = 40, fill = book_palette$fill, colour = "white",
                 linewidth = 0.15) +
  geom_vline(xintercept = 7, colour = book_palette$reject, linewidth = 0.8) +
  facet_wrap(~ n, ncol = 1, scales = "free_y") +
  labs(x = "sample mean", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank())
```

Every row is centred on 7 — the sample mean is unbiased at any sample size. But
the spread shrinks steadily, and in a completely predictable way. Compare the
last two columns of the table: the observed standard deviation of the sample
means matches $\sigma/\sqrt{n}$ almost exactly.

::: {.keyidea}
Two results, which we prove in the next section:

$$E(\bar{X}) = \mu \qquad\qquad \mathrm{Var}(\bar{X}) = \frac{\sigma^2}{n}$$

The sample mean is centred on the population mean, and its spread falls as the
sample grows. The quantity $\sigma/\sqrt{n}$ is the **standard error**, and it
is the single most important number in applied statistics: it is how wrong you
should expect to be.
:::

::: {.watchout}
Note the square root. To halve your standard error you must **quadruple** your
sample. Precision is expensive, and it grows more expensive the more of it you
already have.

This is why surveys of two thousand people are common and surveys of two lakh
are not: the second costs a hundred times as much for about seven times the
precision.
:::


## The Law of Large Numbers {#u1-lln}

::: {.question}
Why does a bigger sample give a better answer?
:::

We have now seen twice, in simulation, that sample means cluster around the
population mean and tighten as the sample grows. The **Law of Large Numbers**
is the theorem that says this must happen.

::: {.keyidea}
**Law of Large Numbers.** Let $X_1, X_2, \ldots, X_n$ be independent draws from
a distribution with mean $\mu$. As $n$ grows, the probability that the sample
mean $\bar{X}_n$ lies close to $\mu$ approaches 1.

Informally: take a big enough sample and your average will be near the truth.
:::

### Watching it happen {-}

The cleanest way to see it is to compute a **running mean** — the average after
one observation, after two, after three, and so on — and watch where it goes.

```{r d2i-lln-run}
set.seed(20)

rolls_seq <- sample(1:6, size = 5000, replace = TRUE)
running   <- cumsum(rolls_seq) / seq_along(rolls_seq)

running[c(1, 5, 10, 100, 1000, 5000)]
```

The first few averages are all over the place. By a thousand rolls the running
mean has settled close to 3.5, and it stays there.

```{r d2i-lln-plot, echo = FALSE, fig.height = 3.6, fig.cap = "The running average of a fair die, over 5,000 rolls. Early on it swings wildly; it then converges on the expected value of 3.5 and stops moving."}
ggplot(data.frame(n = seq_along(running), m = running), aes(n, m)) +
  geom_hline(yintercept = 3.5, colour = book_palette$reject,
             linetype = "dashed", linewidth = 0.7) +
  geom_line(colour = book_palette$accent, linewidth = 0.5) +
  annotate("text", x = 4200, y = 3.5, label = "E[X] = 3.5", vjust = -0.7,
           colour = book_palette$reject, size = 3.6) +
  scale_x_log10(labels = scales::comma) +
  labs(x = "number of rolls (log scale)", y = "running average") +
  theme_book()
```

The same thing happens with the earnings data. Draw surveys of growing size and
watch how far the survey mean sits from the truth:

```{r u1-lln-earnings}
set.seed(20)

earn <- workers$annual_earnings

lln_earn <- t(sapply(c(10, 50, 100, 500, 2000, 10000), function(n) {
  errs <- replicate(400, abs(mean(sample(earn, n, replace = TRUE)) - mean(earn)))
  c(n = n, avg_error = mean(errs))
}))

round(lln_earn)
```

The average distance between a survey mean and the truth falls steadily towards
zero. Nothing about earnings being violently skewed prevents it — it only means
you need a bigger survey to get equally close.

### Why it works {-}

Two results do all the work, and both follow from the definition of the sample
mean.

::: {.theory}
**The sample mean is centred on the population mean.**

$$
\begin{aligned}
E(\bar{X}) &= E\!\left(\frac{X_1 + X_2 + \cdots + X_n}{n}\right) \\[4pt]
&= \frac{1}{n}\Big(E(X_1) + E(X_2) + \cdots + E(X_n)\Big) \\[4pt]
&= \frac{1}{n}\underbrace{(\mu + \mu + \cdots + \mu)}_{n \text{ times}} \\[4pt]
&= \frac{n\mu}{n} \;=\; \mu
\end{aligned}
$$

Every draw comes from the same distribution, so each has expectation $\mu$.
Averaging $n$ of them leaves $\mu$ untouched. This is what **unbiasedness**
means: the estimator is not systematically too high or too low.
:::

::: {.theory}
**Its spread falls as the sample grows.**

$$
\begin{aligned}
\mathrm{Var}(\bar{X}) &= \mathrm{Var}\!\left(\frac{1}{n}(X_1 + \cdots + X_n)\right) \\[4pt]
&= \frac{1}{n^2}\,\mathrm{Var}(X_1 + \cdots + X_n) \\[4pt]
&= \frac{1}{n^2}\underbrace{(\sigma^2 + \sigma^2 + \cdots + \sigma^2)}_{n \text{ times}} \\[4pt]
&= \frac{n\sigma^2}{n^2} \;=\; \frac{\sigma^2}{n}
\end{aligned}
$$

The step from the first line to the second uses $\mathrm{Var}(aX) = a^2\mathrm{Var}(X)$.
The step to the third uses **independence**: the variance of a sum equals the
sum of the variances only when the draws do not influence one another. This is
the single place where the independence assumption earns its keep.
:::

Put the two together. The sample mean sits on top of $\mu$, and the width of
its distribution is $\sigma/\sqrt{n}$, which shrinks to nothing as $n$ grows.
A distribution centred on the right answer with vanishing spread must
eventually be concentrated at the right answer. That is the Law of Large
Numbers.

### What it does not say {-}

::: {.watchout}
The LLN is the most frequently misapplied result in statistics, usually in the
following form: *"the coin has come up heads six times in a row, so tails is
due."*

It is not. The coin has no memory, and the LLN promises nothing of the kind.
What converges is the **proportion**, and it converges by being progressively
diluted, not by being corrected.

```{r d2i-gamblers}
set.seed(5)
flips <- rbinom(20000, 1, 0.5)

check <- t(sapply(c(10, 100, 1000, 20000), function(n) {
  h <- sum(flips[1:n])
  c(tosses = n, heads = h,
    excess_heads = h - n / 2,        # absolute gap from half
    proportion   = h / n)            # the thing that converges
}))
round(check, 3)
```

The proportion marches towards 0.5. The **absolute** surplus of heads does not
shrink — it typically grows. Nothing pushes the count back; the early
imbalance simply becomes a smaller and smaller share of the total.
:::

## The Central Limit Theorem {#u1-clt}

::: {.question}
Why was that picture of a thousand survey means bell-shaped?
:::

The Law of Large Numbers tells us *where* the sample mean ends up. It says
nothing about the **shape** of its distribution along the way — and without a
shape we cannot compute the probability of anything.

That is what the Central Limit Theorem supplies, and it supplies far more than
seems reasonable.

::: {.keyidea}
**Central Limit Theorem.** For a sufficiently large sample, the sampling
distribution of the mean is approximately normal — *whatever the shape of the
population it was drawn from*.

Standardised,

$$Z = \frac{\bar{X} - \mu}{\sigma / \sqrt{n}} \;\longrightarrow\; N(0, 1)$$
:::

Read that again, because the strength of the claim is easy to miss. The
population may be a coin flip taking only the values 0 and 1. It may be a count
with a long right tail. It may be Indian earnings, where a
small minority hold most of the total. **It does not matter.** Average
enough draws and the distribution of that average is bell-shaped.

### Watching it happen {-}

Four populations, chosen to look as unlike a normal distribution as possible.
For each we take samples of size 1, then 5, then 30, and plot the distribution
of the sample mean.

```{r d2i-clt-sim}
set.seed(3)

draw_mean <- function(parent, n) {
  switch(parent,
    "Bernoulli"  = mean(rbinom(n, 1, 0.3)),
    "Poisson"    = mean(rpois(n, 2)),
    "Uniform"    = mean(runif(n)),
    "Earnings"   = mean(sample(earn, n, replace = TRUE)))
}

parents <- c("Bernoulli", "Poisson", "Uniform", "Earnings")
sizes   <- c(1, 5, 30)

clt <- do.call(rbind, lapply(parents, function(p)
  do.call(rbind, lapply(sizes, function(n)
    data.frame(parent = p,
               n      = factor(paste("n =", n), levels = paste("n =", sizes)),
               m      = replicate(4000, draw_mean(p, n)))))))

# Standardise within each panel, so all twelve are on one scale and the
# comparison is purely about shape.
clt$z <- ave(clt$m, clt$parent, clt$n,
             FUN = function(x) (x - mean(x)) / sd(x))

head(clt, 3)
```

```{r d2i-clt-plot, echo = FALSE, fig.height = 6, fig.cap = "Sampling distributions of the mean from four very different populations, standardised, against the standard normal curve in orange. At $n = 1$ you are looking at each population itself. By $n = 30$ all four track the curve — though look closely at the last row."}
clt$parent <- factor(clt$parent, levels = parents)

ggplot(clt, aes(z)) +
  geom_histogram(aes(y = after_stat(density)), bins = 45,
                 fill = book_palette$fill, colour = "white", linewidth = 0.08) +
  stat_function(fun = dnorm, colour = book_palette$reject, linewidth = 0.6) +
  facet_grid(parent ~ n) +
  coord_cartesian(xlim = c(-3.2, 3.2), ylim = c(0, 0.52)) +
  labs(x = "standardised sample mean", y = NULL) +
  theme_book(base_size = 10) +
  theme(axis.text.y = element_blank(),
        strip.text.y = element_text(angle = 0, hjust = 0))
```

Read the grid left to right. The first column is each population itself — a
two-spike coin, a discrete count, a flat rectangle, a badly lopsided earnings
distribution. Not one of them is normal. By the third column all four are
tracking the orange curve.

Notice how differently they get there. The uniform population is nearly normal
by $n = 5$. The Bernoulli is still visibly stepped at $n = 30$, because a mean
of thirty zeros and ones can only take thirty-one distinct values. And the ASI
row — real Indian data — is still leaning right at $n = 30$. Convergence is
guaranteed; its speed is not.

### Why it works {-}

The intuition is easier than the proof, and it is worth having.

::: {.intuition}
Roll one die: every face is equally likely. Now roll two and add them.

```{r d2i-dice-sum}
sums <- outer(1:6, 1:6, "+")
table(sums)
```

Seven can be made six ways — $1+6$, $2+5$, $3+4$, $4+3$, $5+2$, $6+1$. Two can
be made exactly one way. The middle values have more routes to them than the
extremes do, so they turn up more often.

That is the whole idea. To get an extreme *average*, every draw has to be
extreme in the same direction, and that is rare. To get a middling average,
the draws only have to offset one another, and there are enormously many ways
for that to happen. Adding more draws multiplies the number of routes to the
middle far faster than the routes to the edges — and the shape that results is
the bell.
:::

```{r d2i-dice-plot, echo = FALSE, fig.height = 3.2, fig.cap = "The sum of two dice. Combinations pile up in the middle, and the flat distribution of a single die has already started to become a mound."}
ds <- as.data.frame(table(sums)); names(ds) <- c("sum", "ways")
ds$sum <- as.numeric(as.character(ds$sum))

ggplot(ds, aes(sum, ways)) +
  geom_col(fill = book_palette$fill, width = 0.75) +
  geom_text(aes(label = ways), vjust = -0.5, size = 3.1,
            colour = book_palette$ink) +
  scale_x_continuous(breaks = 2:12) +
  labs(x = "sum of two dice", y = "number of ways") +
  theme_book() +
  theme(axis.text.y = element_blank())
```

### How large is "large enough"? {-}

Textbooks often give $n \geq 30$. It is a serviceable rule and it is not a law.

How quickly the CLT takes hold depends on how skewed the population is. A
symmetric population needs very few observations; a badly lopsided one needs
many more.

```{r d2i-clt-howbig}
set.seed(9)

# A crude measure of how far a distribution is from symmetric
skewness <- function(x) mean((x - mean(x))^3) / sd(x)^3

sapply(c(5, 30, 100, 500), function(n)
  round(skewness(replicate(2000, mean(sample(earn, n, replace = TRUE)))), 2))
```

```{r u1-clt-popskew, echo = FALSE}
pop_skew <- round(mean((earn - mean(earn))^3) / sd(earn)^3, 1)
```

The earnings population has a skewness of about `r pop_skew`. At $n = 5$ the
sampling distribution is still visibly lopsided. By $n = 100$ it is close to
symmetric.

::: {.watchout}
Do not treat $n = 30$ as a guarantee. It is a rule of thumb calibrated on
mildly non-normal populations.

Economic data is frequently *not* mildly non-normal. Incomes, firm sizes, land
holdings and city populations all have long right tails, and for these the
approximation can still be poor at $n = 30$. When your variable is heavily
skewed, either use a larger sample or check the sampling distribution by
simulation, exactly as we have just done.
:::

### Why this is the foundation {-}

::: {.keyidea}
The Central Limit Theorem is what makes inference practical. Because of it we
can say, for a sample of reasonable size and *without knowing the shape of the
population*:

- the sample mean is centred on $\mu$ — from the LLN
- its spread is $\sigma/\sqrt{n}$ — the standard error
- its shape is approximately normal — from the CLT

Knowing the centre, the spread and the shape of a distribution is knowing the
distribution. That is enough to attach a probability to any statement about
$\bar{X}$, which is exactly what confidence intervals and hypothesis tests do.

Everything in the rest of this book rests on these three lines.
:::


## The average is not the distribution {#u1-average}

::: {.question}
And what does an average hide?
:::

A habit worth forming now, before any technique gets in the way.

Almost all of inferential statistics is about averages. Nearly every method in
this book estimates a mean, compares two means, or fits a line through
conditional means. That focus is not arbitrary — means are well behaved, their
sampling distributions are tractable, and the two great theorems are statements
about them.

But the mean is only **one** feature of a distribution, and often not the one a
policy question cares about.

```{r u1-skew}
c(mean   = mean(earn),
  median = median(earn))

mean(earn) / median(earn)
```

The mean is more than **twice** the median. When those two diverge that
sharply, the average has stopped describing a typical worker.

```{r u1-skew-plot, echo = FALSE, fig.height = 4, fig.cap = "Annual earnings of 20,000 Indian workers. The mean sits far to the right of most people, dragged there by a thin tail of high earners."}
ggplot(workers, aes(annual_earnings / 1000)) +
  geom_histogram(bins = 60, fill = book_palette$fill, colour = "white",
                 linewidth = 0.15) +
  geom_vline(xintercept = median(earn) / 1000,
             colour = book_palette$ink, linetype = "dashed", linewidth = 0.7) +
  geom_vline(xintercept = mean(earn) / 1000,
             colour = book_palette$reject, linewidth = 0.9) +
  annotate("label", x = mean(earn) / 1000, y = Inf, vjust = 1.4,
           label = "mean", colour = book_palette$reject, fill = "white",
           size = 3.5, label.size = 0.3) +
  annotate("label", x = median(earn) / 1000, y = Inf, vjust = 3.2,
           label = "median", colour = book_palette$ink, fill = "white",
           size = 3.5, label.size = 0.3) +
  coord_cartesian(xlim = c(0, 400)) +
  labs(x = "annual earnings (₹ thousand)", y = "number of workers") +
  theme_book()
```

How concentrated is it?

```{r u1-concentration}
top10 <- earn[earn >= quantile(earn, 0.90)]

round(100 * sum(top10) / sum(earn))              # % of all earnings, top tenth
round(quantile(earn, 0.90) / median(earn), 1)    # top-decile threshold vs median
```

The best-paid tenth of workers take a large share of all earnings, and the
threshold for entering that tenth is several times the median wage.

::: {.intuition}
Now ask what "average earnings in India" is good for.

Asking how much India earns in total? Fine — the mean times the number of
workers is the total, by construction.

Asking what a typical Indian worker earns? Badly misleading. Very few people
earn anything near the mean; most earn far less.

Asking whether a policy helped? The mean can rise while the majority are worse
off, if the gains land at the top. **An average can move in the opposite
direction to almost everybody inside it.**
:::

::: {.keyidea}
Ask two questions of every average you meet, including your own:

1. **What is the spread?** A mean with no measure of dispersion is half a
   sentence.
2. **Who is where?** Means are the wrong tool for questions about the bottom of
   a distribution — poverty, vulnerability, who is hurt by a shock. Those are
   questions about *tails*, and the mean is precisely the statistic that
   averages tails away.
:::

This is not a philosophical aside; it changes what you estimate. Whole
literatures in development economics exist because the mean was the wrong
target. Poverty measures weight the poor rather than averaging over everyone.
Vulnerability measures ask who is at risk of falling below a line.
Distributional analysis asks whether a policy that raised average income raised
the income of anyone who was badly off to begin with.

We will return to this repeatedly. When we fit a regression we will be
modelling a conditional *mean*, and it will be worth asking each time what that
choice conceals. When we reach binary explanatory variables and estimate an
average treatment effect, the word "average" in that phrase will deserve every
bit of suspicion you can bring to it.

::: {.watchout}
None of this makes averages bad. They are the right tool for a great many
questions, and this book spends most of its length on them for good reason.

It means an average is an **answer to a specific question**, not a neutral
summary of the data. Decide which question you are asking before deciding the
mean answers it.
:::

## What can go wrong before any mathematics {#u1-what-goes-wrong}

::: {.question}
When does none of this machinery help?
:::

Inference assumes your sample carries information about the population you care
about. Three ways that fails, none of which any formula will warn you about.

**The sample is not from the population you mean.** The ASI covers *registered*
factories. Most Indian manufacturing employment is unregistered. Conclusions
about "Indian manufacturing" drawn from ASI data are conclusions about a formal
sliver of it. The arithmetic will not object.

**The sample is not random.** Our class survey has 32 heights. They are not a
random sample of APU students, let alone of young Indians — they are the people
who took this course and turned up that day. You may compute a confidence
interval from them and it will mean nothing about the university.

**The question is causal but the data is not.** Suppose states with more
industry have higher literacy. Does industry raise literacy? Does literacy
attract industry? Does something else drive both? The correlation is identical
in all three cases, and no statistical machinery separates them without an
argument about *how the data came to exist*.

::: {.intuition}
That last failure is why this book is organised as it is. From the unit on
two-sample tests onwards, we stop asking only "what is this number?" and start
asking "what would have to be true for this number to mean what I want it to
mean?"

That question — not any particular estimator — is what separates a good applied
economist from someone who can run regressions.
:::

## Summary {#u1-summary}

::: {.keyidea}
1. **Description** summarises data you have. **Inference** reaches beyond it and
   must report how far it has reached.
2. A **distribution** is a pattern emerging from repetition under randomness.
   No randomness, no distribution, no statistics.
3. Greek letters ($\mu$, $\sigma$) are population parameters you never observe.
   Roman letters ($\bar{x}$, $s$) are sample statistics you compute.
4. Inference is licensed by the **i.i.d. assumption** — that your sample is
   drawn from the same distribution as the population. A badly drawn sample
   cannot be repaired later.
5. A **sampling distribution** describes how a statistic varies across repeated
   samples. It is the pivot on which all inference turns.
6. $E(\bar{X}) = \mu$ and $\mathrm{Var}(\bar{X}) = \sigma^2/n$. The sample mean
   is unbiased, and its **standard error** $\sigma/\sqrt{n}$ falls with the
   square root of the sample size.
7. The **Law of Large Numbers**: as $n$ grows, $\bar{X}$ converges on $\mu$.
   It says the *proportion* settles down, not that deviations get corrected —
   which is why it is no help to a gambler.
8. The **Central Limit Theorem**: for large enough $n$, the sampling
   distribution of $\bar{X}$ is approximately normal whatever the population
   looks like. "Large enough" depends on skewness, and $n = 30$ is a rule of
   thumb, not a guarantee — Indian industrial data is still lopsided there.
9. **The mean is one feature of a distribution, not a summary of it.** Indian
   earnings have a mean more than twice their median. Always ask about spread,
   and about who sits in the tails.
:::

We opened by claiming that the average you compute from a sample is almost
certainly wrong. It is — and we can now say precisely how wrong, and with what
probability. That is the whole of the difference between an anecdote and an
estimate.

The next unit puts the machinery to work. If we know the sample mean is centred
on $\mu$, spread by $\sigma/\sqrt{n}$ and approximately normal, we can turn
a single sample into a **range** of plausible values for the population mean,
with a stated level of confidence. That is estimation, and it is the first
thing this apparatus was built for.
