# Estimation {#estimation}

```{r setup-02, include = FALSE}
source("_common.R")
library(ggplot2)
```

## A population where we know the truth {#u2-known-truth}

::: {.question}
Can we see these ideas working in practice?
:::

The previous unit introduced a language for describing random phenomena. We
defined random variables, expected values and variances, and argued that the
sample mean behaves in remarkably predictable ways.

Those arguments were largely theoretical. With a real population we almost
never know the true mean or variance, which makes it difficult to check whether
the machinery behaves as claimed. If a sample mean comes out at ₹54,000, there
is nothing to compare it against.

To overcome this, we construct an artificial population whose properties are
known exactly. Because we know the truth in advance, we can compare it with
what repeated sampling actually produces.

### Constructing the population {-}

Take a deck of playing cards, scored A = 1, J = 11, Q = 12, K = 13, and 2
through 10 at face value.

```{r u2-deck}
deck <- rep(1:13, times = 4)

c(N = length(deck), mu = mean(deck), sigma = sd(deck))
```

This population contains 52 values. Its mean is exactly $\mu = 7$ and its
standard deviation is $\sigma = 3.78$. Both are known with certainty, which is
precisely what real data never offers.

### Drawing one sample {-}

Now draw ten cards and compute their average.

```{r u2-onesample}
set.seed(7)
mean(sample(deck, size = 10, replace = TRUE))
```

The result is not 7. Draw again and it would differ again. This is sampling
variation, exactly as we saw with the earnings surveys — but now we can see
precisely how far the sample has strayed, because we know where it should be.

::: {.watchout}
Note `replace = TRUE`: each card is returned to the deck before the next draw.

That is what makes the draws **independent**, the assumption discussed below.
Drawing without replacement changes what remains after each card is removed, so
the draws are dependent and the sample mean varies slightly less than theory
predicts.

It is also the better model of what we usually mean. Surveying 2,000 people
does not meaningfully deplete a population of a billion.
:::

### Drawing a thousand samples {-}

One sample tells us little. Repeating the exercise a thousand times gives the
whole sampling distribution.

```{r u2-sampdist}
set.seed(7)
deck_means <- replicate(1000, mean(sample(deck, size = 10, replace = TRUE)))

c(mean_of_sample_means = mean(deck_means),
  sd_of_sample_means   = sd(deck_means))
```

```{r u2-sampdist-plot, echo = FALSE, fig.height = 3.6, fig.cap = "One thousand sample means, each computed from 10 cards drawn from a population whose mean is exactly 7."}
ggplot(data.frame(m = deck_means), aes(m)) +
  geom_histogram(bins = 32, fill = book_palette$fill, colour = "white",
                 linewidth = 0.2) +
  geom_vline(xintercept = 7, colour = book_palette$reject, linewidth = 0.9) +
  annotate("label", x = 7, y = Inf, vjust = 1.4, label = "Population mean = 7",
           colour = book_palette$reject, fill = "white", size = 3.5,
           label.size = 0.3) +
  labs(x = "Sample mean of 10 cards", y = "Number of samples") +
  theme_book()
```

This is the sampling distribution discussed throughout the previous unit — but
for the first time we can hold it against the truth.

### Two observations {-}

Two features of this simulation are worth recording before we explain them.

**The average of the sample means is almost exactly the population mean.** No
individual sample produced 7, yet their average is
`r round(mean(deck_means), 3)` — essentially the population value.

**The sample means vary far less than the cards do.** The population standard
deviation is about `r round(sd(rep(1:13, 4)), 2)`; the standard deviation of
the sample means is about `r round(sd(deck_means), 2)`, roughly a third as
large. Averaging has cancelled out much of the variation.

Both observations have names, and both have explanations.

### What happens as the sample grows {-}

```{r u2-nsizes}
set.seed(7)

sizes <- c(5, 10, 20, 40, 80)
results <- t(sapply(sizes, function(n) {
  m <- replicate(2000, mean(sample(deck, size = n, replace = TRUE)))
  c(n = n, mean = mean(m), sd = sd(m), theoretical_sd = sd(deck) / sqrt(n))
}))

round(results, 3)
```

```{r u2-nsizes-plot, echo = FALSE, fig.height = 4.4, fig.cap = "Sampling distributions of the mean for growing sample sizes. The centre never moves. The spread collapses."}
set.seed(7)
dd <- do.call(rbind, lapply(c(5, 20, 80), function(n) {
  data.frame(n = paste("n =", n),
             m = replicate(2000, mean(sample(deck, n, replace = TRUE))))
}))
dd$n <- factor(dd$n, levels = paste("n =", c(5, 20, 80)))

ggplot(dd, aes(m)) +
  geom_histogram(bins = 40, fill = book_palette$fill, colour = "white",
                 linewidth = 0.15) +
  geom_vline(xintercept = 7, colour = book_palette$reject, linewidth = 0.8) +
  facet_wrap(~ n, ncol = 1, scales = "free_y") +
  labs(x = "Sample mean", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank())
```

Every row is centred on 7 — the sample mean is unbiased at any sample size. But
the spread shrinks steadily, and in a completely predictable way. Compare the
last two columns of the table: the observed standard deviation of the sample
means matches $\sigma/\sqrt{n}$ almost exactly.

::: {.keyidea}
Two results, both of which can be established directly:

$$E(\bar{X}) = \mu \qquad\qquad \mathrm{Var}(\bar{X}) = \frac{\sigma^2}{n}$$

The sample mean is centred on the population mean, and its spread falls as the
sample grows. The quantity $\sigma/\sqrt{n}$ is called the **standard error**
of the sample mean.
:::

### Why were the sample means centred on seven? {-}

Neither observation requires anything beyond the definition of the sample mean.

::: {.theory}

$$
\begin{aligned}
E(\bar{X}) &= E\!\left(\frac{X_1 + X_2 + \cdots + X_n}{n}\right) \\[4pt]
&= \frac{1}{n}\Big(E(X_1) + E(X_2) + \cdots + E(X_n)\Big) \\[4pt]
&= \frac{1}{n}\underbrace{(\mu + \mu + \cdots + \mu)}_{n \text{ times}} \\[4pt]
&= \frac{n\mu}{n} \;=\; \mu
\end{aligned}
$$

Every observation is drawn from the same distribution, so each has expectation
$\mu$. Averaging $n$ of them leaves $\mu$ unchanged. A statistic with this
property is described as **unbiased**: it is not systematically too high or too
low, although any particular value of it will still differ from $\mu$.
:::

### Why did the distribution become narrower? {-}

::: {.theory}

$$
\begin{aligned}
\mathrm{Var}(\bar{X}) &= \mathrm{Var}\!\left(\frac{1}{n}(X_1 + \cdots + X_n)\right) \\[4pt]
&= \frac{1}{n^2}\,\mathrm{Var}(X_1 + \cdots + X_n) \\[4pt]
&= \frac{1}{n^2}\underbrace{(\sigma^2 + \sigma^2 + \cdots + \sigma^2)}_{n \text{ times}} \\[4pt]
&= \frac{n\sigma^2}{n^2} \;=\; \frac{\sigma^2}{n}
\end{aligned}
$$

The second line uses $\mathrm{Var}(aX) = a^2\mathrm{Var}(X)$. The third uses
**independence**: the variance of a sum equals the sum of the variances only
when the observations do not influence one another. This is the single point at
which the independence assumption is required.
:::

These two results explain the behaviour observed in the previous unit. The
sample mean remains centred on the population mean regardless of sample size,
while its variability decreases as the sample size increases. Together they
provide the mathematical foundation for the Law of Large Numbers introduced in
Section \@ref(u1-lln).

### The price of precision {-}

Notice the square root. Doubling the sample size does not halve the
uncertainty. To reduce the standard error by half, the sample size must
increase fourfold.

This illustrates one of the fundamental realities of empirical research:
precision is expensive. As samples become larger, each additional observation
contributes a little less than the one before it. This is why surveys of a few
thousand households are common, while surveys of hundreds of thousands are
comparatively rare. Beyond a certain point, the additional precision is simply
not worth the additional cost.



## The assumptions behind inference {#u2-iid}

::: {.question}
What has to be true before a sample can tell us anything?
:::

Throughout the previous unit we assumed, without saying so, that our
observations were drawn from the population of interest. More precisely, we
assume that the observations are **independent and identically distributed**,
usually abbreviated to i.i.d.

**Identically distributed** means that every observation is drawn from the same
probability distribution. The process generating the data does not change from
one observation to the next.

**Independent** means that knowing the value of one observation provides no
information about the value of another.

Together, these assumptions ensure that every observation provides comparable
information about the same population.

::: {.watchout}
Their importance cannot be overstated. If the sample does not represent the
population we wish to study, no amount of mathematical sophistication can
recover the missing information. Statistical methods cannot repair a
fundamentally flawed sample.

Good inference begins not with mathematics, but with good data and thoughtful
study design.
:::
