# Estimation: Why the Sample Mean? {#estimation}

```{r setup-02, include = FALSE}
source("_common.R")
library(ggplot2)
```

## The problem of estimation {#u2-problem}

::: {.question}
We want to know something about a population we cannot observe. What exactly
are we going to do?
:::

The previous unit was about uncertainty. It established that repeated samples
produce different averages, that those averages have a sampling distribution of
their own, and that this distribution becomes increasingly concentrated as the
sample grows. All of that described how a statistic *behaves*.

It did not tell us what to *do*.

Consider the position of anyone conducting a real survey. There is a quantity
they want to know — the average annual earnings of Indian wage workers, the
proportion of households with piped water, the average yield per hectare of
paddy in Punjab. That quantity is a fixed number. It exists. It is simply not
observed, because nobody has surveyed all 1.4 billion people, and nobody ever
will.

What they have instead is a sample, and a decision to make: which number
computed from that sample should be put forward as the answer?

::: {.keyidea}
A **parameter** is a fixed, unknown feature of the population, such as $\mu$ or
$\sigma^2$.

An **estimator** is a rule for computing a guess at a parameter from a sample.
Because it is computed from a sample, it is a random variable, and we write it
with a capital letter: $\bar{X}$.

An **estimate** is the number that rule produces from the sample actually
drawn: $\bar{x} = 54{,}300$.
:::

The distinction between the last two is worth pausing on, because the words are
often used interchangeably in ordinary speech. An estimator is the rule: take
the observations, add them together and divide by $n$. An estimate is the
number produced after the rule has been applied to a particular sample.

The difference matters. An estimate can never be judged directly, because we
almost never know the true population mean. If a survey estimates average
monthly income to be ₹50,000, we do not know whether the truth is ₹49,800,
₹50,100 or ₹55,000. The truth is precisely what we are trying to discover.

The estimator, however, can be judged. Although we never learn whether one
particular estimate is close to the truth, we can ask how the estimator behaves
across many hypothetical samples. Does it systematically overestimate the
population mean? Does it systematically underestimate it? Does it produce very
different answers from one sample to the next? These are questions about the
procedure, not about any single outcome.

::: {.keyidea}
Since the parameter is never known, an estimator cannot be judged by any single
estimate. It is judged by how it behaves across the many samples that might
have been drawn.

Evaluating procedures rather than outcomes is the idea that drives the whole of
this unit.
:::

There are many possible estimators of the population mean. We could use the
sample median. We could take the midpoint of the largest and smallest
observations. We could even ignore the data entirely and always report ₹50,000.
Each is an estimator of $\mu$, because each provides a rule for producing an
estimate from a sample.

Why, then, has statistics settled almost universally on the sample mean?

That is the question for the rest of this unit.



## When can a sample represent a population? {#u2-iid}

::: {.question}
Before comparing estimators — can a sample tell us anything about a population
at all?
:::

There is an objection to deal with first. Everything above assumed that a
sample carries information about the population it came from. That is not
automatic. It depends on how the sample was collected, and it can fail badly.

Rather than assert this, we can watch it happen. Below is a population of
20,000 households, spread across 200 localities. Sixty of those localities are
urban, and incomes there are higher. Because we build the population
ourselves, we know its mean exactly.

```{r u2-pop}
set.seed(2)

n_loc <- 200                 # localities
per   <- 100                 # households in each

urban    <- rep(c(TRUE, FALSE), times = c(60, 140))
loc_mean <- 18000 + 9000 * urban + rnorm(n_loc, sd = 4000)

pop <- data.frame(
  locality = rep(1:n_loc, each = per),
  urban    = rep(urban,   each = per),
  income   = round(rnorm(n_loc * per, rep(loc_mean, each = per), sd = 5000))
)

mu    <- mean(pop$income)
sigma <- sd(pop$income)

c(mu = mu, sigma = sigma, se_if_random = sigma / sqrt(100))
```

The population mean is `r format(round(mean_pop <- mean(pop$income)), big.mark = ",")`.
Each survey below collects 100 households, and is repeated a thousand times.
Only the *method of collection* differs.

```{r u2-four-designs}
# Survey A: 100 households drawn at random from the whole population
# Survey B: 100 households drawn at random, but only from urban localities
# Survey C: five localities chosen at random, 20 households in each
# Survey D: five *urban* localities chosen at random, 20 households in each

cluster_sample <- function(pool) {
  chosen <- sample(pool, 5)
  unlist(lapply(chosen, function(l) sample(pop$income[pop$locality == l], 20)))
}

urban_pop <- pop$income[pop$urban]
urb_loc   <- which(urban)

set.seed(11)
designs <- list(
  A = replicate(1000, mean(sample(pop$income, 100))),
  B = replicate(1000, mean(sample(urban_pop,  100))),
  C = replicate(1000, mean(cluster_sample(1:n_loc))),
  D = replicate(1000, mean(cluster_sample(urb_loc)))
)

round(t(sapply(designs, function(x) c(centre = mean(x), spread = sd(x)))))
```

```{r u2-four-plot, echo = FALSE, fig.height = 4.6, fig.cap = "One thousand surveys under each design. The vertical line is the true population mean. Only the first design is both centred on it and as precise as theory promises."}
lab <- c(A = "A. Random sample",
         B = "B. Urban households only",
         C = "C. Five localities, 20 each",
         D = "D. Five urban localities, 20 each")
dd <- do.call(rbind, lapply(names(designs), function(k)
  data.frame(design = lab[k], m = designs[[k]])))
dd$design <- factor(dd$design, levels = lab)

ggplot(dd, aes(m)) +
  geom_histogram(bins = 45, fill = book_palette$fill, colour = "white",
                 linewidth = 0.15) +
  geom_vline(xintercept = mu, colour = book_palette$reject, linewidth = 0.8) +
  facet_wrap(~ design, ncol = 1, scales = "free_y") +
  scale_x_continuous(labels = function(x) format(x, big.mark = ",")) +
  labs(x = "Estimated mean monthly household income (₹)", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank())
```

Read the four rows carefully, because they fail in different ways.

**Survey A works.** The thousand estimates are centred on the population mean,
and their spread is almost exactly the $\sigma/\sqrt{n}$ predicted in the
previous unit.

**Survey B is wrong, and looks excellent.** Every estimate is far too high —
the surveyor never visits a village, so village incomes cannot appear in the
answer. Notice the second column: the spread is *smaller* than in Survey A.
Urban households resemble one another, so the estimates agree closely with each
other while agreeing on the wrong number. Consistency is not accuracy. This is
the more dangerous failure, because a narrow spread is easily mistaken for a
reliable result.

**Survey C is centred correctly but far too variable.** Choosing five
localities and surveying twenty households in each is cheap and common — an
interviewer travels to a place and works through it. But households in a
locality resemble one another, so the second household adds much less new
information than the first. The sample has 100 rows and nothing like 100
households' worth of information. The spread is over three times what
$\sigma/\sqrt{n}$ predicts, so a researcher using that formula would report
confidence they have not earned.

**Survey D fails both ways at once**, which is what most convenience samples
do in practice.

::: {.question}
What did the first survey have that the others did not?
:::

Two things, and each failure above corresponds to losing one of them.

::: {.keyidea}
**Identically distributed** — every observation is drawn from the same
distribution, and it is the distribution we want to learn about. Survey B
violated this: urban households are drawn from a different distribution than
the population of Indian households.

**Independent** — knowing the value of one observation tells us nothing about
another. Survey C violated this: neighbours share a locality, a labour market
and often an occupation.

Together the two are abbreviated **i.i.d.**
:::

These assumptions are not mathematical decoration. Sections
\@ref(u2-centred) and \@ref(u2-precision) derive the two properties that
justify using the sample mean, and each derivation uses exactly one of them —
in one identifiable line. Break the assumption and the property it supports
disappears, which is precisely what the four surveys show.

::: {.watchout}
No amount of mathematical sophistication repairs a badly collected sample. If
the households you surveyed are not drawn from the population you are asking
about, there is no formula that recovers the ones you missed.

Good inference begins with study design, not with statistics.
:::



## How can we judge an estimator? {#u2-known-truth}

::: {.question}
If $\mu$ is never known, how can we ever tell whether an estimator is any good?
:::

We now know what a sample must satisfy. The original question remains: of the
many rules available, why the sample mean?

The obstacle is the one identified at the start. To see whether an estimator
lands near the truth we would have to know the truth, and in any real
application we do not. If we survey Indian households we do not know true
average household income; if we sample firms we do not know true average
productivity. There is nothing to compare against.

The way around this is to build a laboratory. We construct one artificial
population whose properties are known exactly, and study repeated sampling from
it. Because the true mean is known in advance, every sample mean can be
measured against it. Nothing is being discovered about the world here — the
point is to watch a procedure work under conditions where its performance is
visible.

### The population {#u2-lab-pop -}

Take a deck of playing cards, scored A = 1, J = 11, Q = 12, K = 13, and 2
through 10 at face value.

```{r u2-deck}
deck <- rep(1:13, times = 4)

c(N = length(deck), mu = mean(deck), sigma = sd(deck))
```

This population contains 52 values. Its mean is exactly $\mu = 7$ and its
standard deviation is $\sigma = 3.78$. Both are known with certainty, which is
exactly what real data never offers.

### One sample from it {#u2-lab-one -}

Draw ten cards and compute their average.

```{r u2-onesample}
set.seed(7)
mean(sample(deck, size = 10, replace = TRUE))
```

The result is not 7. Draw again and it would differ again — sampling variation,
as before. What is new is that we can see how far this sample strayed, because
we know where it should have been.

::: {.watchout}
Note `replace = TRUE`: each card is returned to the deck before the next draw.

That is what makes the draws **independent**, in the sense of the previous
section. Drawing without replacement changes what remains after each card is
removed, so the draws are dependent and the sample mean varies slightly less
than theory predicts.

It is also the better model of what we usually mean. Surveying 2,000 people
does not meaningfully deplete a population of a billion.
:::

### A thousand samples {#u2-lab-many -}

One sample tells us little about a procedure. Repeating the exercise a thousand
times shows the whole sampling distribution — and, for the first time, shows it
alongside the truth it is supposed to be tracking.

```{r u2-sampdist}
set.seed(7)
deck_means <- replicate(1000, mean(sample(deck, size = 10, replace = TRUE)))

c(mean_of_sample_means = mean(deck_means),
  sd_of_sample_means   = sd(deck_means))
```

```{r u2-sampdist-plot, echo = FALSE, fig.height = 3.6, fig.cap = "One thousand sample means, each computed from 10 cards drawn from a population whose mean is exactly 7."}
ggplot(data.frame(m = deck_means), aes(m)) +
  geom_histogram(bins = 32, fill = book_palette$fill, colour = "white",
                 linewidth = 0.2) +
  geom_vline(xintercept = 7, colour = book_palette$reject, linewidth = 0.9) +
  annotate("label", x = 7, y = Inf, vjust = 1.4, label = "Population mean = 7",
           colour = book_palette$reject, fill = "white", size = 3.5,
           label.size = 0.3) +
  labs(x = "Sample mean of 10 cards", y = "Number of samples") +
  theme_book()
```

Two features of this picture are worth separating, and the next two sections
take them one at a time: where the distribution sits, and how wide it is.



## Is the sample mean centred on the truth? {#u2-centred}

::: {.question}
The sample means cluster around 7. Is that a coincidence of this simulation?
:::

No individual sample produced 7. Yet the average of the thousand sample means
is `r round(mean(deck_means), 3)` — essentially the population value.

It is not a coincidence, and it does not depend on the deck. It follows from
the definition of the sample mean and one line of algebra.

::: {.theory}

$$
\begin{aligned}
E(\bar{X}) &= E\!\left(\frac{X_1 + X_2 + \cdots + X_n}{n}\right) \\[4pt]
&= \frac{1}{n}\Big(E(X_1) + E(X_2) + \cdots + E(X_n)\Big) \\[4pt]
&= \frac{1}{n}\underbrace{(\mu + \mu + \cdots + \mu)}_{n \text{ times}} \\[4pt]
&= \frac{n\mu}{n} \;=\; \mu
\end{aligned}
$$

The third line is the one that matters. It replaces every $E(X_i)$ with the
same $\mu$, and it is entitled to do so only because the observations are
**identically distributed** — each drawn from the population we are asking
about. Averaging $n$ of them then leaves $\mu$ unchanged.
:::

Look back at Survey B. Its observations were identically distributed among
themselves, but not drawn from the population of interest, so the $\mu$
substituted in that third line was the urban mean rather than the national one.
The algebra was never wrong; it answered a question nobody had asked.

This is the first of the two properties promised earlier, and it now has a
name.

::: {.keyidea}
An estimator is **unbiased** if its expected value equals the parameter it
estimates. The sample mean is unbiased for the population mean:

$$E(\bar{X}) = \mu$$

Unbiasedness is a statement about the procedure, not about any particular
estimate. It says the rule is not systematically too high or too low. It does
not say your $\bar{x}$ is close to $\mu$ — every one of the thousand samples
above missed.
:::



## Why do larger samples fluctuate less? {#u2-precision}

::: {.question}
Unbiasedness cannot be the whole story. What does a larger sample actually buy?
:::

Unbiasedness alone is worth remarkably little. Consider the rule "take the
first observation and ignore the rest." It is perfectly unbiased —
$E(X_1) = \mu$ — and nobody would use it, because it never improves. Collect a
million households and it still reports one household's income.

So the second property has to concern spread. Run the deck experiment across a
range of sample sizes.

```{r u2-nsizes}
set.seed(7)

sizes <- c(5, 10, 20, 40, 80)
results <- t(sapply(sizes, function(n) {
  m <- replicate(2000, mean(sample(deck, size = n, replace = TRUE)))
  c(n = n, mean = mean(m), sd = sd(m), theoretical_sd = sd(deck) / sqrt(n))
}))

round(results, 3)
```

```{r u2-nsizes-plot, echo = FALSE, fig.height = 4.4, fig.cap = "Sampling distributions of the mean for growing sample sizes. The centre never moves. The spread collapses."}
set.seed(7)
dd <- do.call(rbind, lapply(c(5, 20, 80), function(n) {
  data.frame(n = paste("n =", n),
             m = replicate(2000, mean(sample(deck, n, replace = TRUE))))
}))
dd$n <- factor(dd$n, levels = paste("n =", c(5, 20, 80)))

ggplot(dd, aes(m)) +
  geom_histogram(bins = 40, fill = book_palette$fill, colour = "white",
                 linewidth = 0.15) +
  geom_vline(xintercept = 7, colour = book_palette$reject, linewidth = 0.8) +
  facet_wrap(~ n, ncol = 1, scales = "free_y") +
  labs(x = "Sample mean", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank())
```

The second column confirms the previous section: every sample size is centred
on 7, and even samples of five are centred correctly. Centring does not improve
with $n$, because it was never wrong.

The third column is what changes. At $n = 5$ the sample means have a standard
deviation of about `r round(results[1, "sd"], 2)`; at $n = 80$ it is about
`r round(results[5, "sd"], 2)`. Compare it against the fourth column: the
observed spread matches $\sigma/\sqrt{n}$ almost exactly at every sample size.

::: {.theory}

$$
\begin{aligned}
\mathrm{Var}(\bar{X}) &= \mathrm{Var}\!\left(\frac{1}{n}(X_1 + \cdots + X_n)\right) \\[4pt]
&= \frac{1}{n^2}\,\mathrm{Var}(X_1 + \cdots + X_n) \\[4pt]
&= \frac{1}{n^2}\underbrace{(\sigma^2 + \sigma^2 + \cdots + \sigma^2)}_{n \text{ times}} \\[4pt]
&= \frac{n\sigma^2}{n^2} \;=\; \frac{\sigma^2}{n}
\end{aligned}
$$

The second line uses $\mathrm{Var}(aX) = a^2\mathrm{Var}(X)$. The third line is
the one that matters: the variance of a sum equals the sum of the variances
only when the observations are **independent**. This is the single point in the
entire argument at which independence is required.
:::

Survey C is now explained. Its households were drawn from the right population,
so it stayed centred — but they were not independent, the third line above did
not hold, and the true spread was more than three times what $\sigma/\sqrt{n}$
claimed. A researcher who applied the formula anyway would have reported a
precision the data did not contain.

The square root of this variance is used so constantly that it has its own
name.

::: {.keyidea}
The **standard error** of the sample mean is the standard deviation of its
sampling distribution:

$$\mathrm{se}(\bar{X}) = \sqrt{\mathrm{Var}(\bar{X})} = \frac{\sigma}{\sqrt{n}}$$

It measures how much the sample mean moves from sample to sample. A small
standard error means that whatever answer we happened to get, a different
sample would have given something similar.
:::

The word *error* is unfortunate, and worth guarding against. It does not
measure a mistake, and it is not the distance between your estimate and the
truth — that distance is unknowable. Like unbiasedness, it is a property of the
procedure.



## Why does the sample mean become more reliable? {#u2-lln-explained}

::: {.question}
Two properties, established separately. What do they amount to together?
:::

We can now answer the question this unit opened with.

The sample mean is used almost universally because its sampling distribution
sits in the right place and tightens as the sample grows. It is centred on
$\mu$ at every sample size, and the width of that distribution,
$\sigma/\sqrt{n}$, falls towards zero as $n$ increases. A distribution centred
on the correct value and becoming ever narrower must concentrate on that value.
Large samples are therefore increasingly likely to produce averages close to
$\mu$.

That last sentence is the Law of Large Numbers, which Section \@ref(u1-lln)
introduced as a description of observed behaviour. It is no longer a
description. Averages settle down because the procedure that produces them is
unbiased and increasingly precise, and there is nothing else to the result.

Both halves are load-bearing, and it is worth seeing why neither would do
alone.

An estimator that was centred but whose spread never shrank would remain
permanently unreliable — the "use the first observation" rule from the previous
section, which stays honest and stays useless however much data you give it.

An estimator whose spread shrank around a value that was not $\mu$ would be
worse still. It would become *reliably wrong*, converging with increasing
confidence on the incorrect answer, and every additional observation would
harden the mistake rather than expose it. Survey B is exactly this: had the
urban-only surveyor collected ten thousand households instead of a hundred,
their estimates would have clustered even more tightly around a number that was
still ₹6,800 too high.

::: {.keyidea}
The sample mean earns its place through two properties together:

$$E(\bar{X}) = \mu \qquad\qquad \mathrm{Var}(\bar{X}) = \frac{\sigma^2}{n}$$

Centred on the truth, and increasingly precise. Neither is sufficient on its
own, and both depend on assumptions about how the sample was collected.
:::



## Why is precision expensive? {#u2-cost}

::: {.question}
If larger samples are better, why does anyone stop collecting data?
:::

One practical consequence deserves its own section, because it shapes how
survey work is actually done.

Return to the square root. Doubling the sample size does not halve the standard
error; it divides it by $\sqrt{2} \approx 1.41$. To halve the standard error,
the sample must grow *fourfold*.

```{r u2-cost, echo = FALSE, fig.height = 3.2, fig.cap = "The standard error of the sample mean against sample size. Almost all of the available precision is bought in the first few hundred observations."}
nn <- 25:3000
ggplot(data.frame(n = nn, se = 7974 / sqrt(nn)), aes(n, se)) +
  geom_line(colour = book_palette$fill, linewidth = 0.9) +
  labs(x = "Households surveyed", y = "Standard error (₹)") +
  theme_book()
```

Precision is therefore expensive, and grows more expensive the more of it you
buy. Each additional household costs the same to interview as the last, but
contributes less than the last. Going from 100 households to 400 halves the
standard error; going from 400 to 1,600 halves it again, at four times the
cost.

This is why national surveys run to a few thousand households rather than a few
hundred thousand. Past a certain point the additional precision is not worth
the additional money, and the money is better spent on the things no sample
size can fix — better questions, better interviewers, better coverage of the
households that are hardest to reach.

That last point is worth holding onto. The failures in Section
\@ref(u2-iid) were all failures of design, and none of them would have been
helped by a larger sample. Survey B with ten thousand urban households is still
wrong. When a result is disappointing, the instinct is to ask for more data;
often the honest answer is that more of the same data will not help.


## Summary {#u2-summary}

::: {.keyidea}
- A population has fixed but unknown **parameters**.
- An **estimator** is a rule; an **estimate** is the result of applying it to
  one sample.
- Since the truth is never observed, estimators are judged by their behaviour
  across repeated samples, not by any single estimate they produce.
- A sample speaks for a population only if it is drawn **independently** and
  from **the population of interest**. Neither assumption is decoration: break
  the first and the spread is far wider than any formula predicts, break the
  second and the estimates are centred on the wrong number while often looking
  more precise than the honest ones.
- Under random sampling the sample mean is **unbiased**: $E(\bar{X}) = \mu$.
- Its **standard error**, $\sigma/\sqrt{n}$, falls as $1/\sqrt{n}$ — so larger
  samples give more precise estimates, at a rising price per unit of precision.
- Centred and increasingly precise together are what make the sample mean
  trustworthy, and together they are the Law of Large Numbers.
:::

Everything in this unit has been about choosing a good estimator. The next unit
asks how to use that estimator to make statements about an unknown population
from a single observed sample.


## Looking ahead {#u2-looking-ahead}

We opened by asking why statistics has settled on the sample mean, and the
answer is now in hand. It is centred on the truth, its variability falls as the
sample grows, and larger samples therefore tend to produce estimates closer to
the population mean.

But notice what every argument in this unit relied on. We drew a thousand
samples and looked at how they scattered. We compared four survey designs by
running each a thousand times. The sampling distribution has been visible on
every page.

In practice it never is. Nobody draws a thousand samples. A survey is conducted
once, at considerable expense, and yields exactly one sample mean.

The sample mean has earned our trust as an estimator. But we see only one of
them. The next question is therefore unavoidable: what can a single sample tell
us about an unknown population?
