# Estimation {#estimation}

```{r setup-02, include = FALSE}
source("_common.R")
library(ggplot2)
```

## The problem of estimation {#u2-problem}

::: {.question}
We want to know something about a population we cannot observe. What exactly
are we going to do?
:::

The previous unit was about uncertainty. It established that repeated samples
produce different averages, that those averages have a sampling distribution of
their own, and that this distribution becomes increasingly concentrated as the
sample grows. All of that described how a statistic *behaves*.

It did not tell us what to *do*.

Consider the position of anyone conducting a real survey. There is a quantity
they want to know — the average annual earnings of Indian wage workers, the
proportion of households with piped water, the average yield per hectare of
paddy in Punjab. That quantity is a fixed number. It exists. It is simply not
observed, because nobody has surveyed all 1.4 billion people, and nobody ever
will.

What they have instead is a sample, and a decision to make: which number
computed from that sample should be put forward as the answer?

::: {.keyidea}
A **parameter** is a fixed, unknown feature of the population, such as $\mu$ or
$\sigma^2$.

An **estimator** is a rule for computing a guess at a parameter from a sample.
Because it is computed from a sample, it is a random variable, and we write it
with a capital letter: $\bar{X}$.

An **estimate** is the number that rule produces from the sample actually
drawn: $\bar{x} = 54{,}300$.
:::

The distinction between the last two is worth pausing on, because the words are
used loosely in ordinary speech. The estimator is the *procedure* — take the
observations, add them up, divide by $n$. The estimate is the *result*. The
procedure can be judged before any data are collected; the result cannot be
judged at all, because we never learn whether it was right.

That last point is the awkward one. If we cannot tell whether a particular
estimate is close to the truth, then the only thing we can sensibly ask about
is the procedure that produced it.

::: {.question}
So: why the sample mean?
:::

It is not the only available rule. We could use the median. We could use the
midpoint of the largest and smallest values. We could ignore the data entirely
and always answer ₹50,000, which costs nothing and is occasionally correct.
Each of these is an estimator of $\mu$ in the sense defined above.

The sample mean is used almost universally, and the reason is not that it is
obvious. It is that its behaviour across repeated samples has two specific
properties that the alternatives lack. The rest of this section establishes
what they are.



## What we assume about the sample {#u2-iid}

::: {.question}
What has to be true of a sample before it can tell us anything at all?
:::

Before studying how the sample mean behaves, we have to be explicit about
something the previous unit assumed silently: that the observations were drawn
from the population of interest, and drawn in a particular way.

The standard assumption is that observations are **independent and identically
distributed**, usually abbreviated to i.i.d.

**Identically distributed** means every observation is drawn from the same
probability distribution. The process generating the data does not change from
one observation to the next. If the first half of a survey is conducted in
cities and the second half in villages, this fails.

**Independent** means that knowing the value of one observation tells us
nothing about the value of another. If an interviewer surveys one household and
then its next-door neighbour, this fails: neighbours share a locality, a labour
market and often a caste.

Together they ensure that every observation carries comparable information
about the same population. We will see shortly exactly where each is needed —
they are not decorative, and each does a specific job in a specific line of a
specific proof.

::: {.watchout}
Their importance is hard to overstate. If a sample does not represent the
population we wish to study, no amount of mathematical sophistication recovers
the missing information. Statistical methods cannot repair a fundamentally
flawed sample.

Good inference begins not with mathematics, but with good data and thoughtful
study design.
:::



## One population where we know the truth {#u2-known-truth}

::: {.question}
How would we ever check that the sample mean is a good rule?
:::

There is an obvious difficulty in the way. The population mean is almost never
known. If we survey Indian households we do not know true average household
income; if we sample firms we do not know true average productivity. So we
cannot compare a sample mean against the truth and see how close it came —
which is precisely the comparison we need in order to judge the rule.

The way around this is to build a laboratory. We construct one artificial
population whose properties are known exactly, and study repeated sampling from
it. Because the true mean is known in advance, every sample mean can be
measured against it. Nothing is being discovered about the world here; the
purpose is to watch a procedure work under conditions where its performance is
visible.

### Constructing the population {-}

Take a deck of playing cards, scored A = 1, J = 11, Q = 12, K = 13, and 2
through 10 at face value.

```{r u2-deck}
deck <- rep(1:13, times = 4)

c(N = length(deck), mu = mean(deck), sigma = sd(deck))
```

This population contains 52 values. Its mean is exactly $\mu = 7$ and its
standard deviation is $\sigma = 3.78$. Both are known with certainty, which is
exactly what real data never offers.

### Drawing one sample {-}

Now draw ten cards and compute their average.

```{r u2-onesample}
set.seed(7)
mean(sample(deck, size = 10, replace = TRUE))
```

The result is not 7. Draw again and it would differ again — sampling variation,
as before. What is new is that we can see how far this sample strayed, because
we know where it should have been.

::: {.watchout}
Note `replace = TRUE`: each card is returned to the deck before the next draw.

That is what makes the draws **independent**, in the sense of the previous
section. Drawing without replacement changes what remains after each card is
removed, so the draws are dependent and the sample mean varies slightly less
than theory predicts.

It is also the better model of what we usually mean. Surveying 2,000 people
does not meaningfully deplete a population of a billion.
:::

### Drawing a thousand samples {-}

One sample tells us little about a procedure. Repeating the exercise a thousand
times shows the whole sampling distribution.

```{r u2-sampdist}
set.seed(7)
deck_means <- replicate(1000, mean(sample(deck, size = 10, replace = TRUE)))

c(mean_of_sample_means = mean(deck_means),
  sd_of_sample_means   = sd(deck_means))
```

```{r u2-sampdist-plot, echo = FALSE, fig.height = 3.6, fig.cap = "One thousand sample means, each computed from 10 cards drawn from a population whose mean is exactly 7."}
ggplot(data.frame(m = deck_means), aes(m)) +
  geom_histogram(bins = 32, fill = book_palette$fill, colour = "white",
                 linewidth = 0.2) +
  geom_vline(xintercept = 7, colour = book_palette$reject, linewidth = 0.9) +
  annotate("label", x = 7, y = Inf, vjust = 1.4, label = "Population mean = 7",
           colour = book_palette$reject, fill = "white", size = 3.5,
           label.size = 0.3) +
  labs(x = "Sample mean of 10 cards", y = "Number of samples") +
  theme_book()
```

### Three questions {-}

We now have something the previous unit never had: a sampling distribution and
the true value it is supposed to be telling us about. Three questions follow
naturally, and answering them is what justifies using the sample mean at all.

1. Is the sample mean centred on the population mean?
2. Why do larger samples fluctuate less?
3. Why does the sample mean become more reliable as the sample grows?

### Question 1: is the sample mean centred on the truth? {-}

No individual sample produced 7. Yet the average of the thousand sample means
is `r round(mean(deck_means), 3)` — essentially the population value.

Widen the experiment across several sample sizes and the pattern is
unmistakable.

```{r u2-nsizes}
set.seed(7)

sizes <- c(5, 10, 20, 40, 80)
results <- t(sapply(sizes, function(n) {
  m <- replicate(2000, mean(sample(deck, size = n, replace = TRUE)))
  c(n = n, mean = mean(m), sd = sd(m), theoretical_sd = sd(deck) / sqrt(n))
}))

round(results, 3)
```

```{r u2-nsizes-plot, echo = FALSE, fig.height = 4.4, fig.cap = "Sampling distributions of the mean for growing sample sizes. The centre never moves. The spread collapses."}
set.seed(7)
dd <- do.call(rbind, lapply(c(5, 20, 80), function(n) {
  data.frame(n = paste("n =", n),
             m = replicate(2000, mean(sample(deck, n, replace = TRUE))))
}))
dd$n <- factor(dd$n, levels = paste("n =", c(5, 20, 80)))

ggplot(dd, aes(m)) +
  geom_histogram(bins = 40, fill = book_palette$fill, colour = "white",
                 linewidth = 0.15) +
  geom_vline(xintercept = 7, colour = book_palette$reject, linewidth = 0.8) +
  facet_wrap(~ n, ncol = 1, scales = "free_y") +
  labs(x = "Sample mean", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank())
```

Every row is centred on 7, and the centring does not improve with sample size
because it was never wrong to begin with. Even samples of five are centred
correctly. This is not something the simulation discovered by luck.

::: {.theory}

$$
\begin{aligned}
E(\bar{X}) &= E\!\left(\frac{X_1 + X_2 + \cdots + X_n}{n}\right) \\[4pt]
&= \frac{1}{n}\Big(E(X_1) + E(X_2) + \cdots + E(X_n)\Big) \\[4pt]
&= \frac{1}{n}\underbrace{(\mu + \mu + \cdots + \mu)}_{n \text{ times}} \\[4pt]
&= \frac{n\mu}{n} \;=\; \mu
\end{aligned}
$$

The third line is where **identically distributed** is used: every observation
is drawn from the same distribution, so each has expectation $\mu$. Averaging
$n$ of them leaves $\mu$ unchanged.
:::

This is the first of the two properties promised earlier, and it now has a
name.

::: {.keyidea}
An estimator is **unbiased** if its expected value equals the parameter it
estimates. The sample mean is unbiased for the population mean:

$$E(\bar{X}) = \mu$$

Unbiasedness is a statement about the procedure, not about any particular
estimate. It says the rule is not systematically too high or too low. It does
not say your $\bar{x}$ is close to $\mu$ — every one of the thousand samples
above missed.
:::

### Question 2: why do larger samples fluctuate less? {-}

Unbiasedness alone is worth little. The useless estimator "take the first
observation and ignore the rest" is also unbiased, and nobody would use it.

What separates the sample mean is the second column of the simulation. At
$n = 5$ the sample means have a standard deviation of about
`r round(results[1, "sd"], 2)`; at $n = 80$ it is about
`r round(results[5, "sd"], 2)`. Compare the last two columns of that table: the
observed spread matches $\sigma/\sqrt{n}$ almost exactly at every sample size.

::: {.theory}

$$
\begin{aligned}
\mathrm{Var}(\bar{X}) &= \mathrm{Var}\!\left(\frac{1}{n}(X_1 + \cdots + X_n)\right) \\[4pt]
&= \frac{1}{n^2}\,\mathrm{Var}(X_1 + \cdots + X_n) \\[4pt]
&= \frac{1}{n^2}\underbrace{(\sigma^2 + \sigma^2 + \cdots + \sigma^2)}_{n \text{ times}} \\[4pt]
&= \frac{n\sigma^2}{n^2} \;=\; \frac{\sigma^2}{n}
\end{aligned}
$$

The second line uses $\mathrm{Var}(aX) = a^2\mathrm{Var}(X)$. The third uses
**independence**: the variance of a sum equals the sum of the variances only
when the observations do not influence one another. This is the single point at
which independence is required, and it is why the assumption matters so much in
practice.
:::

The square root of this quantity is used so constantly that it has its own
name.

::: {.keyidea}
The **standard error** of the sample mean is the standard deviation of its
sampling distribution:

$$\mathrm{se}(\bar{X}) = \sqrt{\mathrm{Var}(\bar{X})} = \frac{\sigma}{\sqrt{n}}$$

It measures how much the sample mean moves from sample to sample. A small
standard error means that whatever answer we happened to get, a different
sample would have given something similar.
:::

The word *error* is unfortunate, and worth guarding against. It does not
measure a mistake, and it is not the distance between your estimate and the
truth — that distance is unknowable. It is a property of the procedure, like
unbiasedness.

### Question 3: why does the sample mean become more reliable? {-}

The two results answer the third question together, and neither would suffice
alone.

The Law of Large Numbers was introduced in Section \@ref(u1-lln) as a
statement about behaviour: as samples grow, sample means settle down around the
population mean. We can now say why it holds. The sampling distribution of
$\bar{X}$ is centred on $\mu$ at every sample size, and its spread
$\sigma/\sqrt{n}$ shrinks towards zero as $n$ grows. A distribution centred on
the right value and becoming ever narrower must concentrate on that value.
Large samples are therefore increasingly likely to produce averages close to
$\mu$.

Notice that both halves are doing work. An estimator that was centred but whose
spread never shrank would stay permanently unreliable. An estimator whose
spread shrank towards a value that was not $\mu$ would become reliably wrong,
which is worse — it would converge confidently on the incorrect answer, and
more data would only strengthen the mistake.

### The price of precision {-}

Return to the square root, because it carries an uncomfortable practical
message. Doubling the sample size does not halve the standard error; it divides
it by $\sqrt{2} \approx 1.41$. To halve the standard error, the sample must
grow fourfold.

Precision is therefore expensive, and gets more expensive as you buy more of
it. Each additional observation contributes a little less than the one before.
This is why surveys of a few thousand households are common while surveys of
hundreds of thousands are rare: past a certain point the additional precision
is not worth the additional cost. It is also why the honest answer to "can we
just collect more data?" is usually "how much more, and who is paying?"
