# Estimation: Why the Sample Mean? {#estimation}

```{r setup-02, include = FALSE}
source("_common.R")
library(ggplot2)
```

## The problem of estimation {#u2-problem}

::: {.question}
We want to know something about a population we cannot observe. What exactly
are we going to do?
:::

The previous unit was about uncertainty. It established that repeated samples
produce different averages, that those averages have a sampling distribution of
their own, and that this distribution becomes increasingly concentrated as the
sample grows. All of that described how a statistic *behaves*.

It did not tell us what to *do*.

Consider the position of anyone conducting a real survey. There is a quantity
they want to know — the average annual earnings of Indian wage workers, the
proportion of households with piped water, the average yield per hectare of
paddy in Punjab. That quantity is a fixed number. It exists. It is simply not
observed, because nobody has surveyed all 1.4 billion people, and nobody ever
will.

What they have instead is a sample, and a decision to make: which number
computed from that sample should be put forward as the answer?

::: {.keyidea}
A **parameter** is a fixed, unknown feature of the population, such as $\mu$ or
$\sigma^2$.

An **estimator** is a rule for computing a guess at a parameter from a sample.
Because it is computed from a sample, it is a random variable, and we write it
with a capital letter: $\bar{X}$.

An **estimate** is the number that rule produces from the sample actually
drawn: $\bar{x} = 54{,}300$.
:::

The distinction between the last two is worth pausing on, because the words are
often used interchangeably in ordinary speech. An estimator is the rule: take
the observations, add them together and divide by $n$. An estimate is the
number produced after the rule has been applied to a particular sample.

The difference matters. An estimate can never be judged directly, because we
almost never know the true population mean. If a survey estimates average
monthly income to be ₹50,000, we do not know whether the truth is ₹49,800,
₹50,100 or ₹55,000. The truth is precisely what we are trying to discover.

The estimator, however, can be judged. Although we never learn whether one
particular estimate is close to the truth, we can ask how the estimator behaves
across many hypothetical samples. Does it systematically overestimate the
population mean? Does it systematically underestimate it? Does it produce very
different answers from one sample to the next? These are questions about the
procedure, not about any single outcome.

::: {.keyidea}
Since the parameter is never known, an estimator cannot be judged by any single
estimate. It is judged by how it behaves across the many samples that might
have been drawn.

Evaluating procedures rather than outcomes is the idea that drives the whole of
this unit.
:::

There are many possible estimators of the population mean. We could use the
sample median. We could take the midpoint of the largest and smallest
observations. We could even ignore the data entirely and always report ₹50,000.
Each is an estimator of $\mu$, because each provides a rule for producing an
estimate from a sample.

Why, then, has statistics settled almost universally on the sample mean?

That is the question for the rest of this unit.



## When can a sample represent a population? {#u2-iid}

::: {.question}
Before comparing estimators — can a sample tell us anything about a population
at all?
:::

There is an objection to deal with first. Everything above assumed that a
sample carries information about the population it came from. That is not
automatic. It depends on how the sample was collected, and it can fail badly.

Rather than assert this, we can watch it happen. Below is a population of
20,000 households, spread across 200 localities. Sixty of those localities are
urban, and incomes there are higher. Because we build the population
ourselves, we know its mean exactly.

```{r u2-pop}
set.seed(2)

n_loc <- 200                 # localities
per   <- 100                 # households in each

urban    <- rep(c(TRUE, FALSE), times = c(60, 140))
loc_mean <- 18000 + 9000 * urban + rnorm(n_loc, sd = 4000)

pop <- data.frame(
  locality = rep(1:n_loc, each = per),
  urban    = rep(urban,   each = per),
  income   = round(rnorm(n_loc * per, rep(loc_mean, each = per), sd = 5000))
)

mu    <- mean(pop$income)
sigma <- sd(pop$income)

c(mu = mu, sigma = sigma, se_if_random = sigma / sqrt(100))
```

The population mean is `r format(round(mean_pop <- mean(pop$income)), big.mark = ",")`.
Each survey below collects 100 households, and is repeated a thousand times.
Only the *method of collection* differs.

```{r u2-four-designs}
# Survey A: 100 households drawn at random from the whole population
# Survey B: 100 households drawn at random, but only from urban localities
# Survey C: five localities chosen at random, 20 households in each
# Survey D: five *urban* localities chosen at random, 20 households in each

cluster_sample <- function(pool) {
  chosen <- sample(pool, 5)
  unlist(lapply(chosen, function(l) sample(pop$income[pop$locality == l], 20)))
}

urban_pop <- pop$income[pop$urban]
urb_loc   <- which(urban)

set.seed(11)
designs <- list(
  A = replicate(1000, mean(sample(pop$income, 100))),
  B = replicate(1000, mean(sample(urban_pop,  100))),
  C = replicate(1000, mean(cluster_sample(1:n_loc))),
  D = replicate(1000, mean(cluster_sample(urb_loc)))
)

round(t(sapply(designs, function(x) c(centre = mean(x), spread = sd(x)))))
```

```{r u2-four-plot, echo = FALSE, fig.height = 4.6, fig.cap = "One thousand surveys under each design. The vertical line is the true population mean. Only the first design is both centred on it and as precise as theory promises."}
lab <- c(A = "A. Random sample",
         B = "B. Urban households only",
         C = "C. Five localities, 20 each",
         D = "D. Five urban localities, 20 each")
dd <- do.call(rbind, lapply(names(designs), function(k)
  data.frame(design = lab[k], m = designs[[k]])))
dd$design <- factor(dd$design, levels = lab)

ggplot(dd, aes(m)) +
  geom_histogram(bins = 45, fill = book_palette$fill, colour = "white",
                 linewidth = 0.15) +
  geom_vline(xintercept = mu, colour = book_palette$reject, linewidth = 0.8) +
  facet_wrap(~ design, ncol = 1, scales = "free_y") +
  scale_x_continuous(labels = function(x) format(x, big.mark = ",")) +
  labs(x = "Estimated mean monthly household income (₹)", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank())
```

Read the four rows carefully, because they fail in different ways.

**Survey A works.** The thousand estimates are centred on the population mean,
and their spread is almost exactly the $\sigma/\sqrt{n}$ predicted in the
previous unit.

**Survey B is wrong, and looks excellent.** Every estimate is far too high —
the surveyor never visits a village, so village incomes cannot appear in the
answer. Notice the second column: the spread is *smaller* than in Survey A.
Urban households resemble one another, so the estimates agree closely with each
other while agreeing on the wrong number. Consistency is not accuracy. This is
the more dangerous failure, because a narrow spread is easily mistaken for a
reliable result.

**Survey C is centred correctly but far too variable.** Choosing five
localities and surveying twenty households in each is cheap and common — an
interviewer travels to a place and works through it. But households in a
locality resemble one another, so the second household adds much less new
information than the first. The sample has 100 rows and nothing like 100
households' worth of information. The spread is over three times what
$\sigma/\sqrt{n}$ predicts, so a researcher using that formula would report
confidence they have not earned.

**Survey D fails both ways at once**, which is what most convenience samples
do in practice.

::: {.question}
What did the first survey have that the others did not?
:::

Two things, and each failure above corresponds to losing one of them.

::: {.keyidea}
**Identically distributed** — every observation is drawn from the same
distribution, and it is the distribution we want to learn about. Survey B
violated this: urban households are drawn from a different distribution than
the population of Indian households.

**Independent** — knowing the value of one observation tells us nothing about
another. Survey C violated this: neighbours share a locality, a labour market
and often an occupation.

Together the two are abbreviated **i.i.d.**
:::

These assumptions are not mathematical decoration. Sections
\@ref(u2-centred) and \@ref(u2-precision) derive the two properties that
justify using the sample mean, and each derivation uses exactly one of them —
in one identifiable line. Break the assumption and the property it supports
disappears, which is precisely what the four surveys show.

::: {.watchout}
No amount of mathematical sophistication repairs a badly collected sample. If
the households you surveyed are not drawn from the population you are asking
about, there is no formula that recovers the ones you missed.

Good inference begins with study design, not with statistics.
:::



## How can we judge an estimator? {#u2-known-truth}

::: {.question}
If $\mu$ is never known, how can we ever tell whether an estimator is any good?
:::

We now know what a sample must satisfy. The original question remains: of the
many rules available, why the sample mean?

The obstacle is the one identified at the start. To see whether an estimator
lands near the truth we would have to know the truth, and in any real
application we do not. If we survey Indian households we do not know true
average household income; if we sample firms we do not know true average
productivity. There is nothing to compare against.

The way around this is to build a laboratory. We construct one artificial
population whose properties are known exactly, and study repeated sampling from
it. Because the true mean is known in advance, every sample mean can be
measured against it. Nothing is being discovered about the world here — the
point is to watch a procedure work under conditions where its performance is
visible.

### The population {#u2-lab-pop -}

Take a deck of playing cards, scored A = 1, J = 11, Q = 12, K = 13, and 2
through 10 at face value.

```{r u2-deck}
deck <- rep(1:13, times = 4)

c(N = length(deck), mu = mean(deck), sigma = sd(deck))
```

This population contains 52 values. Its mean is exactly $\mu = 7$ and its
standard deviation is $\sigma = 3.78$. Both are known with certainty, which is
exactly what real data never offers.

### One sample from it {#u2-lab-one -}

Draw ten cards and compute their average.

```{r u2-onesample}
set.seed(7)
mean(sample(deck, size = 10, replace = TRUE))
```

The result is not 7. Draw again and it would differ again — sampling variation,
as before. What is new is that we can see how far this sample strayed, because
we know where it should have been.

::: {.watchout}
Note `replace = TRUE`: each card is returned to the deck before the next draw.

That is what makes the draws **independent**, in the sense of the previous
section. Drawing without replacement changes what remains after each card is
removed, so the draws are dependent and the sample mean varies slightly less
than theory predicts.

It is also the better model of what we usually mean. Surveying 2,000 people
does not meaningfully deplete a population of a billion.
:::

### A thousand samples {#u2-lab-many -}

One sample tells us little about a procedure. Repeating the exercise a thousand
times shows the whole sampling distribution — and, for the first time, shows it
alongside the truth it is supposed to be tracking.

```{r u2-sampdist}
set.seed(7)
deck_means <- replicate(1000, mean(sample(deck, size = 10, replace = TRUE)))

c(mean_of_sample_means = mean(deck_means),
  sd_of_sample_means   = sd(deck_means))
```

```{r u2-sampdist-plot, echo = FALSE, fig.height = 3.6, fig.cap = "One thousand sample means, each computed from 10 cards drawn from a population whose mean is exactly 7."}
ggplot(data.frame(m = deck_means), aes(m)) +
  geom_histogram(bins = 32, fill = book_palette$fill, colour = "white",
                 linewidth = 0.2) +
  geom_vline(xintercept = 7, colour = book_palette$reject, linewidth = 0.9) +
  annotate("label", x = 7, y = Inf, vjust = 1.4, label = "Population mean = 7",
           colour = book_palette$reject, fill = "white", size = 3.5,
           label.size = 0.3) +
  labs(x = "Sample mean of 10 cards", y = "Number of samples") +
  theme_book()
```

Two features of this picture are worth separating, and the next two sections
take them one at a time: where the distribution sits, and how wide it is.



## Is the sample mean centred on the truth? {#u2-centred}

::: {.question}
The sample means cluster around 7. Is that a coincidence of this simulation?
:::

No individual sample produced 7. Yet the average of the thousand sample means
is `r round(mean(deck_means), 3)` — essentially the population value.

It is not a coincidence, and it does not depend on the deck. It follows from
the definition of the sample mean and one line of algebra.

::: {.theory}

$$
\begin{aligned}
E(\bar{X}) &= E\!\left(\frac{X_1 + X_2 + \cdots + X_n}{n}\right) \\[4pt]
&= \frac{1}{n}\Big(E(X_1) + E(X_2) + \cdots + E(X_n)\Big) \\[4pt]
&= \frac{1}{n}\underbrace{(\mu + \mu + \cdots + \mu)}_{n \text{ times}} \\[4pt]
&= \frac{n\mu}{n} \;=\; \mu
\end{aligned}
$$

The third line is the one that matters. It replaces every $E(X_i)$ with the
same $\mu$, and it is entitled to do so only because the observations are
**identically distributed** — each drawn from the population we are asking
about. Averaging $n$ of them then leaves $\mu$ unchanged.
:::

Look back at Survey B. Its observations were identically distributed among
themselves, but not drawn from the population of interest, so the $\mu$
substituted in that third line was the urban mean rather than the national one.
The algebra was never wrong; it answered a question nobody had asked.

This is the first of the two properties promised earlier, and it now has a
name.

::: {.keyidea}
An estimator is **unbiased** if its expected value equals the parameter it
estimates. The sample mean is unbiased for the population mean:

$$E(\bar{X}) = \mu$$

Unbiasedness is a statement about the procedure, not about any particular
estimate. It says the rule is not systematically too high or too low. It does
not say your $\bar{x}$ is close to $\mu$ — every one of the thousand samples
above missed.
:::



## Why do larger samples fluctuate less? {#u2-precision}

::: {.question}
Unbiasedness cannot be the whole story. What does a larger sample actually buy?
:::

Unbiasedness alone is worth remarkably little. Consider the rule "take the
first observation and ignore the rest." It is perfectly unbiased —
$E(X_1) = \mu$ — and nobody would use it, because it never improves. Collect a
million households and it still reports one household's income.

So the second property has to concern spread. Run the deck experiment across a
range of sample sizes.

```{r u2-nsizes}
set.seed(7)

sizes <- c(5, 10, 20, 40, 80)
results <- t(sapply(sizes, function(n) {
  m <- replicate(2000, mean(sample(deck, size = n, replace = TRUE)))
  c(n = n, mean = mean(m), sd = sd(m), theoretical_sd = sd(deck) / sqrt(n))
}))

round(results, 3)
```

```{r u2-nsizes-plot, echo = FALSE, fig.height = 4.4, fig.cap = "Sampling distributions of the mean for growing sample sizes. The centre never moves. The spread collapses."}
set.seed(7)
dd <- do.call(rbind, lapply(c(5, 20, 80), function(n) {
  data.frame(n = paste("n =", n),
             m = replicate(2000, mean(sample(deck, n, replace = TRUE))))
}))
dd$n <- factor(dd$n, levels = paste("n =", c(5, 20, 80)))

ggplot(dd, aes(m)) +
  geom_histogram(bins = 40, fill = book_palette$fill, colour = "white",
                 linewidth = 0.15) +
  geom_vline(xintercept = 7, colour = book_palette$reject, linewidth = 0.8) +
  facet_wrap(~ n, ncol = 1, scales = "free_y") +
  labs(x = "Sample mean", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank())
```

The second column confirms the previous section: every sample size is centred
on 7, and even samples of five are centred correctly. Centring does not improve
with $n$, because it was never wrong.

The third column is what changes. At $n = 5$ the sample means have a standard
deviation of about `r round(results[1, "sd"], 2)`; at $n = 80$ it is about
`r round(results[5, "sd"], 2)`. Compare it against the fourth column: the
observed spread matches $\sigma/\sqrt{n}$ almost exactly at every sample size.

::: {.theory}

$$
\begin{aligned}
\mathrm{Var}(\bar{X}) &= \mathrm{Var}\!\left(\frac{1}{n}(X_1 + \cdots + X_n)\right) \\[4pt]
&= \frac{1}{n^2}\,\mathrm{Var}(X_1 + \cdots + X_n) \\[4pt]
&= \frac{1}{n^2}\underbrace{(\sigma^2 + \sigma^2 + \cdots + \sigma^2)}_{n \text{ times}} \\[4pt]
&= \frac{n\sigma^2}{n^2} \;=\; \frac{\sigma^2}{n}
\end{aligned}
$$

The second line uses $\mathrm{Var}(aX) = a^2\mathrm{Var}(X)$. The third line is
the one that matters: the variance of a sum equals the sum of the variances
only when the observations are **independent**. This is the single point in the
entire argument at which independence is required.
:::

Survey C is now explained. Its households were drawn from the right population,
so it stayed centred — but they were not independent, the third line above did
not hold, and the true spread was more than three times what $\sigma/\sqrt{n}$
claimed. A researcher who applied the formula anyway would have reported a
precision the data did not contain.

The square root of this variance is used so constantly that it has its own
name.

::: {.keyidea}
The **standard error** of the sample mean is the standard deviation of its
sampling distribution:

$$\mathrm{se}(\bar{X}) = \sqrt{\mathrm{Var}(\bar{X})} = \frac{\sigma}{\sqrt{n}}$$

It measures how much the sample mean moves from sample to sample. A small
standard error means that whatever answer we happened to get, a different
sample would have given something similar.
:::

The word *error* is unfortunate, and worth guarding against. It does not
measure a mistake, and it is not the distance between your estimate and the
truth — that distance is unknowable. Like unbiasedness, it is a property of the
procedure.



## Why does the sample mean become more reliable? {#u2-lln-explained}

::: {.question}
Two properties, established separately. What do they amount to together?
:::

We can now answer the question this unit opened with.

The sample mean is used almost universally because its sampling distribution
sits in the right place and tightens as the sample grows. It is centred on
$\mu$ at every sample size, and the width of that distribution,
$\sigma/\sqrt{n}$, falls towards zero as $n$ increases. A distribution centred
on the correct value and becoming ever narrower must concentrate on that value.
Large samples are therefore increasingly likely to produce averages close to
$\mu$.

That last sentence is the Law of Large Numbers, which Section \@ref(u1-lln)
introduced as a description of observed behaviour. It is no longer a
description. Averages settle down because the procedure that produces them is
unbiased and increasingly precise, and there is nothing else to the result.

Both halves are load-bearing, and it is worth seeing why neither would do
alone.

An estimator that was centred but whose spread never shrank would remain
permanently unreliable — the "use the first observation" rule from the previous
section, which stays honest and stays useless however much data you give it.

An estimator whose spread shrank around a value that was not $\mu$ would be
worse still. It would become *reliably wrong*, converging with increasing
confidence on the incorrect answer, and every additional observation would
harden the mistake rather than expose it. Survey B is exactly this: had the
urban-only surveyor collected ten thousand households instead of a hundred,
their estimates would have clustered even more tightly around a number that was
still ₹6,800 too high.

::: {.keyidea}
The sample mean earns its place through two properties together:

$$E(\bar{X}) = \mu \qquad\qquad \mathrm{Var}(\bar{X}) = \frac{\sigma^2}{n}$$

Centred on the truth, and increasingly precise. Neither is sufficient on its
own, and both depend on assumptions about how the sample was collected.
:::



## Why is precision expensive? {#u2-cost}

::: {.question}
If larger samples are better, why does anyone stop collecting data?
:::

One practical consequence deserves its own section, because it shapes how
survey work is actually done.

Return to the square root. Doubling the sample size does not halve the standard
error; it divides it by $\sqrt{2} \approx 1.41$. To halve the standard error,
the sample must grow *fourfold*.

```{r u2-cost, echo = FALSE, fig.height = 3.2, fig.cap = "The standard error of the sample mean against sample size. Almost all of the available precision is bought in the first few hundred observations."}
nn <- 25:3000
ggplot(data.frame(n = nn, se = 7974 / sqrt(nn)), aes(n, se)) +
  geom_line(colour = book_palette$fill, linewidth = 0.9) +
  labs(x = "Households surveyed", y = "Standard error (₹)") +
  theme_book()
```

Precision is therefore expensive, and grows more expensive the more of it you
buy. Each additional household costs the same to interview as the last, but
contributes less than the last. Going from 100 households to 400 halves the
standard error; going from 400 to 1,600 halves it again, at four times the
cost.

This is why national surveys run to a few thousand households rather than a few
hundred thousand. Past a certain point the additional precision is not worth
the additional money, and the money is better spent on the things no sample
size can fix — better questions, better interviewers, better coverage of the
households that are hardest to reach.

That last point is worth holding onto. The failures in Section
\@ref(u2-iid) were all failures of design, and none of them would have been
helped by a larger sample. Survey B with ten thousand urban households is still
wrong. When a result is disappointing, the instinct is to ask for more data;
often the honest answer is that more of the same data will not help.


## Summary {#u2-summary}

::: {.keyidea}
- A population has fixed but unknown **parameters**.
- An **estimator** is a rule; an **estimate** is the result of applying it to
  one sample.
- Since the truth is never observed, estimators are judged by their behaviour
  across repeated samples, not by any single estimate they produce.
- A sample speaks for a population only if it is drawn **independently** and
  from **the population of interest**. Neither assumption is decoration: break
  the first and the spread is far wider than any formula predicts, break the
  second and the estimates are centred on the wrong number while often looking
  more precise than the honest ones.
- Under random sampling the sample mean is **unbiased**: $E(\bar{X}) = \mu$.
- Its **standard error**, $\sigma/\sqrt{n}$, falls as $1/\sqrt{n}$ — so larger
  samples give more precise estimates, at a rising price per unit of precision.
- Centred and increasingly precise together are what make the sample mean
  trustworthy, and together they are the Law of Large Numbers.
:::

Everything in this unit has been about choosing a good estimator. The next unit
asks how to use that estimator to make statements about an unknown population
from a single observed sample.


## Practice problems {#u2-practice}

::: {.yourturn}
These are for learning. Work each one before opening the solution; the answer
is worth very little if you have not first tried to produce it yourself.

The [question bank](#u2-question-bank) that follows is for testing yourself and
has no answers at all.

Several problems treat `data/wages-india-synthetic.csv` as a *population* of
20,000 wage earners. This is artificial, and deliberately so — it is the same
device as the deck of cards, at realistic Indian magnitudes. The dataset is
simulated but calibrated to IHDS-II estimates, so the returns to schooling and
the urban premium in it are close to the real ones.
:::

### Concept checks {-}

Short questions. A sentence or two is enough.

**2.1** A survey of 800 households reports mean monthly income of ₹18,400.
Which part of that sentence is the estimator, and which is the estimate?

::: {.answer}
The estimator is the sample mean — the rule "add the 800 incomes and divide by
800." The estimate is ₹18,400, the number that rule returned from this
particular sample.

The estimator exists before any data are collected. The estimate exists only
afterwards.
:::

**2.2** Why can a statistician never check whether a particular estimate is
close to the truth?

::: {.answer}
Because checking would require knowing the population mean, and if the
population mean were known there would have been no reason to conduct the
survey. The truth is precisely what the estimate is trying to discover.

This is why the subject evaluates procedures rather than results.
:::

**2.3** Consider the estimator "ignore the data entirely and always report
₹50,000." Is it unbiased? What is its standard error?

::: {.answer}
Its standard error is zero — it returns the same number from every possible
sample, so it does not vary at all.

It is unbiased if and only if the population mean happens to be exactly
₹50,000, which we have no way of knowing. In general it is biased by whatever
amount $\mu$ differs from ₹50,000.

The problem is worth doing because it separates the two properties completely.
Perfect precision is compatible with being arbitrarily wrong, and a small
standard error is therefore no evidence that an estimate is any good.
:::

**2.4** In Section \@ref(u2-iid), Survey B produced estimates with a *smaller*
spread than Survey A. Why is that alarming rather than reassuring?

::: {.answer}
Because the small spread came from surveying households that resemble one
another, not from surveying more of India. The estimates agreed closely with
each other while agreeing on a number ₹6,800 above the truth.

A researcher who saw only their own survey would observe a tight, well-behaved
sample and conclude that the estimate was reliable. Spread measures agreement
among samples drawn the same way; it says nothing about whether that way was
the right one.
:::

**2.5** A report states: "mean earnings were ₹54,000 with a standard error of
₹10,000, so the true figure is within ₹10,000 of our estimate." What is wrong
with the second half of that sentence?

::: {.answer}
The standard error is not the distance between the estimate and the truth. That
distance is unknowable, since the truth is unobserved.

The standard error describes the *procedure*: it says that repeating this
survey would produce sample means with a standard deviation of about ₹10,000.
Any particular estimate may lie much closer to $\mu$ than that, or much further
from it.
:::

### Explain {-}

A paragraph each.

**2.6** Explain why unbiasedness alone is not a good enough reason to use an
estimator. Give an example of an unbiased estimator nobody would use.

::: {.answer}
Unbiasedness says the estimator is right *on average across repeated samples*.
It says nothing about how far any single sample strays, and a rule can be
centred correctly while being wildly variable.

The clearest example is "use the first observation and ignore the rest." Since
$E(X_1) = \mu$, it is exactly unbiased. It is also useless, because its
standard error is $\sigma$ regardless of sample size. Collecting a million
households does not improve it by any amount at all.

Usefulness requires a second property: the spread must fall as the sample
grows. Unbiasedness places the distribution correctly; the variance is what
makes the placement worth anything.
:::

**2.7** An interviewer travels to six villages and surveys thirty households in
each. Which i.i.d. assumption is threatened, which of the two derivations in
this unit stops working, and what happens to the results?

::: {.answer}
Independence is threatened. Households in a village share a labour market, a
cropping pattern, and often an occupation, so once the first few are surveyed
the rest carry much less new information.

The derivation that fails is the variance one. Its third line replaces
$\mathrm{Var}(X_1 + \cdots + X_n)$ with $\sigma^2 + \cdots + \sigma^2$, and
that step requires independence. Nothing goes wrong with $E(\bar{X}) = \mu$,
whose third line uses only identical distribution.

So the estimates remain centred on the population mean — the design is not
biased — but they vary far more from survey to survey than $\sigma/\sqrt{n}$
predicts. A researcher who applied that formula would report a precision the
data does not contain. In Survey C the true spread was more than three times
the formula's claim.
:::

**2.8** Explain why the Law of Large Numbers needs *both* properties of the
sample mean. Describe what would go wrong if each were missing in turn.

::: {.answer}
The Law of Large Numbers says large samples are increasingly likely to produce
averages close to $\mu$. That requires the sampling distribution both to sit in
the right place and to narrow.

Without shrinking variance, the estimator stays centred but never improves.
"Use the first observation" is centred on $\mu$ forever and never gets closer
to it, so no sample size makes it likely to land near the truth.

Without correct centring, shrinking variance actively makes matters worse. The
estimator becomes *reliably wrong* — it converges with increasing confidence on
the incorrect value, and every extra observation hardens the error rather than
exposing it. The urban-only survey is exactly this case: ten thousand urban
households would have produced a tighter cluster around a number still ₹6,800
too high.
:::

### Numerical problems {-}

Pen and paper. Calculators are fine; R is not needed until the next section.

**2.9** In the population of wage earners used below, annual earnings have
standard deviation $\sigma = ₹100{,}141$.

i. Find the standard error of the sample mean for $n = 100$, $n = 400$ and
   $n = 2{,}500$.
ii. How large a sample is needed for the standard error to fall below ₹5,000?
iii. A colleague proposes to improve an $n = 400$ survey by adding 100
     households. By how much does the standard error fall?

::: {.answer}
**(i)** $\mathrm{se} = \sigma/\sqrt{n}$:

$$\begin{aligned}
n = 100:&\quad \frac{100{,}141}{10} = ₹10{,}014 \\[4pt]
n = 400:&\quad \frac{100{,}141}{20} = ₹5{,}007 \\[4pt]
n = 2{,}500:&\quad \frac{100{,}141}{50} = ₹2{,}003
\end{aligned}$$

**(ii)** Require $100{,}141/\sqrt{n} < 5{,}000$, so
$\sqrt{n} > 20.03$ and $n > 401.1$. A sample of **402** suffices.

**(iii)** At $n = 500$ the standard error is $100{,}141/\sqrt{500} = ₹4{,}479$,
down from ₹5,007 — a fall of about ₹528, or roughly 11%, for 25% more
fieldwork. The returns are already thin at this sample size.
:::

**2.10** A researcher with $n$ observations proposes the estimator
$T = \tfrac{1}{2}(X_1 + X_n)$: average the first and last household only.

i. Show that $T$ is unbiased.
ii. Find $\mathrm{Var}(T)$.
iii. Using $\sigma = ₹100{,}141$, compare the standard error of $T$ with that of
     the sample mean of all 100 observations.
iv. What does this show about unbiasedness?

::: {.answer}
**(i)** $E(T) = \tfrac{1}{2}\big(E(X_1) + E(X_n)\big) = \tfrac{1}{2}(\mu + \mu)
= \mu$. Both observations are drawn from the same population, so each has
expectation $\mu$.

**(ii)** Using $\mathrm{Var}(aX) = a^2\mathrm{Var}(X)$ and independence,

$$\mathrm{Var}(T) = \tfrac{1}{4}\big(\sigma^2 + \sigma^2\big)
= \frac{\sigma^2}{2}$$

which is $\sigma^2/n$ with $n = 2$ — unsurprising, since $T$ *is* the sample
mean of two observations.

**(iii)** $\mathrm{se}(T) = 100{,}141/\sqrt{2} = ₹70{,}810$, against
$100{,}141/10 = ₹10{,}014$ for the full sample mean. $T$ is about seven times
more variable.

**(iv)** That unbiasedness is a low bar. $T$ is exactly as unbiased as
$\bar{X}$ and vastly worse, because it discards 98 of the 100 observations
collected. Choosing between unbiased estimators is decided on variance.
:::

**2.11** A survey costs ₹900 per household in fieldwork. The current design
surveys 400 households.

i. What does the current survey cost?
ii. What would it cost to halve the standard error?
iii. The funder instead offers to double the budget. By what factor does the
     standard error fall?

::: {.answer}
**(i)** $400 \times ₹900 = ₹360{,}000$.

**(ii)** Halving the standard error requires four times the sample, so 1,600
households at a cost of ₹1,440,000 — an extra ₹1,080,000 to buy one factor of
two.

**(iii)** Doubling the budget buys 800 households. The standard error falls by
a factor of $\sqrt{2} \approx 1.41$, a reduction of about 29%.

Twice the money does not buy twice the precision, and this gets worse the
larger the survey already is. It is often a better use of the second ₹360,000
to improve coverage of hard-to-reach households than to double the sample —
because the errors that arise from missing those households are not reduced by
sample size at all.
:::

### R exercises {-}

**2.12** Treat `data/wages-india-synthetic.csv` as a population of 20,000 wage
earners. Compute its true mean and standard deviation. Then draw 2,000 samples
of 100 workers each, and check both properties of the sample mean against
theory.

::: {.answer}
```{r u2-ex-pop}
wages <- read.csv("data/wages-india-synthetic.csv")
earn  <- wages$annual_earnings

mu_pop    <- mean(earn)
sigma_pop <- sd(earn)

set.seed(3)
sim <- replicate(2000, mean(sample(earn, 100, replace = TRUE)))

round(c(population_mean = mu_pop,
        centre_of_estimates = mean(sim),
        theoretical_se = sigma_pop / sqrt(100),
        observed_se = sd(sim)))
```

The estimates are centred on the population mean, and their spread is close to
$\sigma/\sqrt{n}$. Both properties hold.

This population is severely right-skewed — its mean is
`r format(round(mean(earn)), big.mark = ",")` against a median of
`r format(round(median(earn)), big.mark = ",")` — yet neither property needed
the population to be well behaved. Unbiasedness followed from each observation
having expectation $\mu$, and the variance result from independence. Shape
never entered either derivation.
:::

**2.13** Repeat the previous exercise sampling only from urban workers, as an
interviewer working exclusively in cities would. Report the centre and spread,
and say which assumption has been broken.

::: {.answer}
```{r u2-ex-urban}
urban_earn <- earn[wages$residence == "Urban"]

set.seed(3)
sim_urban <- replicate(2000, mean(sample(urban_earn, 100, replace = TRUE)))

round(c(population_mean = mu_pop,
        centre_of_estimates = mean(sim_urban),
        observed_se = sd(sim_urban)))
```

The estimates are centred near ₹102,000 against a true population mean of about
₹54,000 — almost double. The assumption broken is *identically distributed*:
urban workers are drawn from a different distribution than Indian wage earners
as a whole, so the $\mu$ appearing in the third line of the $E(\bar{X})$
derivation is the urban mean, not the national one.

Note also that no amount of extra data would help. The estimator is converging
efficiently on the wrong number.
:::

**2.14** Draw 2,000 sample means for $n = 25, 100, 400$ and $1{,}600$ from the
same population, and plot the observed standard error against the theoretical
$\sigma/\sqrt{n}$. Does the $1/\sqrt{n}$ rule hold?

::: {.answer}
```{r u2-ex-se, fig.height = 3.4, fig.cap = "Observed and theoretical standard errors for growing sample sizes."}
set.seed(3)
ns <- c(25, 100, 400, 1600)

se_tab <- data.frame(
  n = ns,
  observed = sapply(ns, function(n) sd(replicate(4000, mean(sample(earn, n, replace = TRUE))))),
  theory   = sigma_pop / sqrt(ns)
)

round(se_tab)

ggplot(se_tab, aes(n)) +
  geom_line(aes(y = theory), colour = book_palette$reject, linewidth = 0.9) +
  geom_point(aes(y = observed), size = 2.6, colour = book_palette$fill) +
  labs(x = "Sample size", y = "Standard error (₹)",
       subtitle = "Line: theoretical. Points: simulated.") +
  theme_book()
```

The points sit on the curve at every sample size, agreeing with theory to
within about one per cent. Quadrupling $n$ halves the standard error each time,
exactly as $1/\sqrt{n}$ requires.

Note the `replace = TRUE`, which matters more here than it did with the deck.
Drop it and the observed spread at $n = 1{,}600$ comes out roughly five per
cent *below* theory, because 1,600 is eight per cent of this population of
20,000 — sample a substantial fraction of a finite population without
replacement and the draws are no longer independent, so $\sigma/\sqrt{n}$
stops being right. Real surveys of India are nowhere near this fraction, which
is why the assumption is harmless in practice and not in a simulation.
:::

**2.15** Do the two properties depend on the population being well behaved?
Compare the sampling distribution of the mean at $n = 10$ and $n = 400$ for
this heavily skewed earnings population, and comment on what changes and what
does not.

::: {.answer}
```{r u2-ex-skew, fig.height = 3.8, fig.cap = "Sampling distribution of mean earnings at two sample sizes. The red line is the population mean."}
set.seed(3)
sk <- do.call(rbind, lapply(c(10, 400), function(n)
  data.frame(n = factor(paste("n =", n), levels = c("n = 10", "n = 400")),
             m = replicate(4000, mean(sample(earn, n, replace = TRUE))))))

ggplot(sk, aes(m)) +
  geom_histogram(bins = 60, fill = book_palette$fill, colour = "white",
                 linewidth = 0.1) +
  geom_vline(xintercept = mu_pop, colour = book_palette$reject,
             linewidth = 0.8) +
  facet_wrap(~ n, scales = "free") +
  scale_x_continuous(labels = function(x) format(x, big.mark = ",")) +
  labs(x = "Sample mean of annual earnings (₹)", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank())
```

Both panels are centred on the population mean. Unbiasedness holds at $n = 10$
just as at $n = 400$, and it required nothing of the population's shape — the
derivation used only that each observation has expectation $\mu$.

What changes is the *shape*. At $n = 10$ the sampling distribution inherits the
population's long right tail and is clearly skewed. At $n = 400$ it is close to
symmetric. That is the Central Limit Theorem from Section \@ref(u1-clt) at
work, and it is a reminder that "large enough" depends on the population: for
earnings data, $n = 30$ would not have been remotely sufficient.
:::

**2.16** Using the same population, compare three estimators of $\mu$ across
2,000 samples of 50: the sample mean, the sample median, and the midpoint of
the smallest and largest observations. Which would you use, and why?

::: {.answer}
```{r u2-ex-compare}
set.seed(3)
midrange <- function(x) (min(x) + max(x)) / 2

comp <- replicate(2000, {
  s <- sample(earn, 50, replace = TRUE)
  c(mean = mean(s), median = median(s), midrange = midrange(s))
})

round(cbind(centre = rowMeans(comp),
            bias   = rowMeans(comp) - mu_pop,
            spread = apply(comp, 1, sd)))
```

Only the sample mean is centred on $\mu$. The median sits far below it, because
this population is strongly right-skewed and the median estimates the
population *median*, not the mean — it is an excellent estimator of a different
parameter. The midrange is wildly variable, since it depends entirely on the
two most extreme observations drawn and discards the other 48.

The sample mean is the only one of the three that is unbiased for $\mu$ and
tightens dependably as the sample grows.

Worth noting, though: if the question of interest were "what does a typical
wage earner make?", the median would be the better summary. Choosing an
estimator requires first being clear about which parameter you actually want.
:::


## Question bank {#u2-question-bank}

::: {.watchout}
**No answers are given for this section, by design.** These questions are for
testing yourself once the practice problems are complete. There are no hints
and no worked solutions — exactly as in an examination.
:::

### Multiple choice {-}

**QB 2.1** Which of the following is an estimate rather than an estimator?

a. $\bar{X}$
b. The rule "take the median of the sample"
c. ₹18,400
d. $\sigma/\sqrt{n}$

**QB 2.2** An estimator is unbiased if

a. every estimate it produces equals the parameter
b. its expected value equals the parameter
c. its variance falls as the sample size grows
d. the sample is drawn at random

**QB 2.3** A population has $\sigma = 60$. For the standard error of the sample
mean to equal 2, the sample size must be

a. 30
b. 120
c. 900
d. 3,600

**QB 2.4** A survey samples only households with telephones. Which is the most
accurate description of the consequence?

a. The estimates will be centred correctly but too variable
b. The estimates will be centred on the wrong value, and a larger sample will
   not fix it
c. The estimates will be centred on the wrong value, but a larger sample will
   fix it
d. Nothing goes wrong provided the sample is large

**QB 2.5** The variance derivation $\mathrm{Var}(\bar{X}) = \sigma^2/n$ uses
independence at exactly one step. Which one?

a. Taking the constant $1/n$ outside the variance
b. Writing $\mathrm{Var}(aX) = a^2 \mathrm{Var}(X)$
c. Replacing the variance of the sum with the sum of the variances
d. Cancelling $n$ against $n^2$

**QB 2.6** Two surveys estimate the same quantity. Survey X has a standard
error of ₹4,000; Survey Y has a standard error of ₹1,000. Which statement
follows? *(select all that apply)*

a. Survey Y's estimate is closer to the truth
b. Survey Y's estimates vary less across repeated sampling
c. Survey Y used a larger sample, if both were drawn the same way
d. Survey Y is unbiased

**QB 2.7** Doubling a survey's sample size reduces its standard error by a
factor of

a. 4
b. 2
c. $\sqrt{2}$
d. It depends on $\sigma$

### Short reasoning {-}

**QB 2.8** A student writes: "the standard error tells us how much our estimate
differs from the population mean." Correct the statement, and explain in one
sentence why the original cannot be right.

**QB 2.9** Explain why an estimator with zero variance is not necessarily a
good estimator. Give an example.

**QB 2.10** A researcher collects 500 households by standing outside a
government hospital on a weekday morning and interviewing whoever passes.
Identify at least two distinct problems with this sample, and state for each
whether it affects the centre of the estimates, their spread, or both.

**QB 2.11** Two rules for estimating $\mu$ from $n$ observations are proposed:
the sample mean, and the average of the first $n/2$ observations. Both are
unbiased. Explain how you would decide between them, and what the answer is.

**QB 2.12** An economist reports that increasing a survey from 1,000 to 1,200
households "substantially improved the precision of our estimates." Evaluate
that claim quantitatively.

**QB 2.13** Section \@ref(u2-iid) showed a badly designed survey producing
estimates with a *smaller* spread than a well-designed one. Explain how this is
possible, and what it implies about using a reported standard error to judge
whether a study is trustworthy.

### Numerical {-}

**QB 2.14** A population of farm households has mean annual income
$\mu = ₹84{,}000$ and standard deviation $\sigma = ₹36{,}000$.

i. Find the standard error of the sample mean for a survey of 144 households.
ii. What sample size would be needed to reduce the standard error to ₹1,500?
iii. Fieldwork costs ₹1,200 per household. What is the additional cost of the
     sample in (ii) relative to the one in (i)?

**QB 2.15** For a population with variance $\sigma^2$, consider the estimator
$W = \tfrac{1}{3}X_1 + \tfrac{2}{3}X_2$ computed from two independent
observations.

i. Show that $W$ is unbiased.
ii. Find $\mathrm{Var}(W)$.
iii. Compare it with the variance of the ordinary sample mean of the same two
     observations. Which is preferable, and what general principle does this
     illustrate?

**QB 2.16** A state has 40 districts. A survey selects 4 districts at random and
interviews 250 households in each, reporting a standard error of
$\sigma/\sqrt{1000}$.

i. State the assumption this calculation requires.
ii. Explain why it is unlikely to hold here.
iii. Say whether the reported standard error is likely to be too large or too
     small, and why that direction is the dangerous one.

**QB 2.17** Using `data/wages-india-synthetic.csv` as a population, write R code
to estimate how large a sample is needed before the standard error of mean
annual earnings falls below ₹4,000. Verify your answer by simulation as well as
by formula, and explain any discrepancy between the two.


## Looking ahead {#u2-looking-ahead}

We opened by asking why statistics has settled on the sample mean, and the
answer is now in hand. It is centred on the truth, its variability falls as the
sample grows, and larger samples therefore tend to produce estimates closer to
the population mean.

But notice what every argument in this unit relied on. We drew a thousand
samples and looked at how they scattered. We compared four survey designs by
running each a thousand times. The sampling distribution has been visible on
every page.

In practice it never is. Nobody draws a thousand samples. A survey is conducted
once, at considerable expense, and yields exactly one sample mean.

The sample mean has earned our trust as an estimator. But we see only one of
them. The next question is therefore unavoidable: what can a single sample tell
us about an unknown population?
