# Point and Interval Estimation {#interval-estimation}

```{r setup-03, include = FALSE}
source("_common.R")
library(ggplot2)
```

## What can a single sample tell us? {#u3-single-sample}

::: {.question}
We have one sample and one number. What are we entitled to say?
:::

The previous unit established why the sample mean deserves our trust. It is
centred on the population mean, and its standard error $\sigma/\sqrt{n}$ falls
as the sample grows. Both results were demonstrated by drawing a thousand
samples and watching how they scattered.

A real survey draws one.

So we are left holding a single number — say a sample mean attendance of 26.1
classes — and the honest question is what may be claimed on the strength of it.
Reporting 26.1 alone is misleading, since we know it is almost certainly not
the population value. Reporting it with a standard error is honest but
awkward, because a standard error is a technical property of a procedure rather
than a statement about attendance.

What we want is a claim of the form: *the population mean lies between here and
there, and this is how confident we are of it.* Producing such claims is the
business of this unit.

Three things stand between us and that goal.

The first is that $\sigma/\sqrt{n}$ contains $\sigma$, a population quantity we
do not know. If we are estimating $\mu$ because it is unobserved, we can hardly
help ourselves to $\sigma$. So we will need to estimate that too.

The second is that many of the questions worth asking are not about averages at
all. What proportion of workers are in urban areas? What proportion of
households own land? These are proportions, not means, and we need estimators
for them as well.

The third is the step from a single number to a range, which requires deciding
what "confident" is going to mean.

::: {.keyidea}
Two kinds of answer run through this unit.

A **point estimate** is a single number offered as the best available guess at
a parameter: $\bar{x} = 26.1$.

An **interval estimate** is a range of values, together with a statement of how
often such ranges capture the truth: $[24.4,\ 27.8]$ with 95% confidence.

The first is what we have. The second is what we want.
:::

We take the missing point estimates first, then build the intervals on top of
them.



## How do we estimate the population variance? {#u3-variance}

::: {.question}
The standard error formula needs $\sigma$. Where is that supposed to come from?
:::

The natural move is to do to the variance what we did to the mean: define the
sample analogue of the population quantity and hope it behaves.

The population variance is the average squared distance from the population
mean,

$$\sigma^2 = \frac{1}{N}\sum_{i=1}^{N}(x_i - \mu)^2$$

so the obvious sample version replaces $\mu$ with $\bar{x}$ and $N$ with $n$:

$$\widetilde{s}^2 = \frac{1}{n}\sum_{i=1}^{n}(x_i - \bar{x})^2$$

This turns out to be systematically wrong, and we can see it happen. The
dataset `attendance-grades.csv` records 680 students at Azim Premji University.
Treat those 680 as a population, so its variance is known exactly, and sample
from it.

```{r u3-nminus1}
students <- read.csv("data/attendance-grades.csv")
att      <- students$attendance

sigma2_pop <- mean((att - mean(att))^2)      # divisor N, since this IS the population

set.seed(5)
n <- 10
sims <- replicate(20000, {
  x <- sample(att, n, replace = TRUE)
  c(divide_by_n     = mean((x - mean(x))^2),
    divide_by_n_1   = var(x))               # var() divides by n - 1
})

round(c(population_variance = sigma2_pop,
        avg_divide_by_n     = mean(sims[1, ]),
        avg_divide_by_n_1   = mean(sims[2, ])), 3)
```

Dividing by $n$ gives `r round(mean(sims[1, ]), 2)` against a true variance of
`r round(sigma2_pop, 2)` — too small, not occasionally but on average across
twenty thousand samples. Dividing by $n - 1$ gives `r round(mean(sims[2, ]), 2)`,
which lands on the population value.

::: {.question}
Why should dividing by a smaller number be the correction?
:::

The reason is that $\bar{x}$ is computed from the very observations whose
spread we are measuring.

The sample mean sits, by construction, in the middle of the sample. It is the
value that makes $\sum(x_i - \bar{x})^2$ as small as it can possibly be — no
other number, including the true $\mu$, gives a smaller total. So when we
measure the spread of the sample around its own mean, we are measuring it
around the most flattering point available, and we get an answer that is too
small.

::: {.theory}
Once $\bar{x}$ is known, the deviations are not free to take any values. They
must sum to zero:

$$\sum_{i=1}^{n}(x_i - \bar{x}) = 0$$

Given the first $n-1$ of them, the last is determined. Only $n - 1$ of the $n$
deviations carry independent information about the spread, and dividing by
$n - 1$ rather than $n$ is what corrects for it.

The quantity $n - 1$ is called the **degrees of freedom**: the number of
observations, less the number of quantities estimated from them along the way.
:::

::: {.keyidea}
The **sample variance** and **sample standard deviation** are

$$s^2 = \frac{1}{n-1}\sum_{i=1}^{n}(x_i - \bar{x})^2
\qquad\qquad s = \sqrt{s^2}$$

$s^2$ is an unbiased estimator of $\sigma^2$. In R they are `var()` and `sd()`,
both of which use $n - 1$.
:::

::: {.watchout}
$s$ is *not* an unbiased estimator of $\sigma$, even though $s^2$ is unbiased
for $\sigma^2$. Taking a square root does not preserve unbiasedness, and $s$
runs slightly low on average.

The bias is small, it shrinks quickly with $n$, and correcting it is not worth
the trouble in practice. But it is worth knowing that "unbiased" attaches to a
specific estimator of a specific parameter, and does not survive being passed
through an arbitrary function.
:::

### When the population mean is known {-}

Occasionally $\mu$ is known even though $\sigma^2$ is not — a machine
calibrated to a specification, or a process with a target value. Then no
degrees of freedom are spent estimating the centre, nothing is flattered, and
the divisor is $n$:

$$\widehat{\sigma}^2 = \frac{1}{n}\sum_{i=1}^{n}(x_i - \mu)^2$$

Note that this uses $\mu$, not $\bar{x}$. Since $\bar{x}$ minimises the sum of
squared deviations, the total measured from $\mu$ is *larger* than the total
measured from $\bar{x}$ — which is precisely why it can be divided by the
larger $n$ and still come out right.

::: {.intuition}
The rule is easier to remember as a question than as two formulas. How many
quantities did I have to estimate from this sample before I could measure its
spread?

Estimated the centre from the data: divide by $n - 1$. Knew the centre already:
divide by $n$.
:::

### Estimating the standard error {-}

We now have what Section \@ref(u3-single-sample) said was missing. The standard
error of the sample mean is $\sigma/\sqrt{n}$, and with $\sigma$ unknown we
substitute $s$:

$$\widehat{\mathrm{se}}(\bar{X}) = \frac{s}{\sqrt{n}}$$

This is what every piece of software reports as "the standard error", and it is
an estimate, with its own sampling variability. That substitution looks
harmless and is not, as Section \@ref(u3-t) shows.



## How do we estimate a proportion? {#u3-proportion}

::: {.question}
What proportion of Indian wage earners work in urban areas? That is not an
average. Do we need new machinery?
:::

Many important quantities are proportions. The share of households below the
poverty line, the share of children enrolled in school, the share of a
workforce that is female, the unemployment rate. None of these is obviously an
average, and it would be reasonable to expect a separate theory.

There is none, and the reason is a small trick.

Write down, for each person in the sample, a variable that equals 1 if they
have the characteristic and 0 if they do not.

```{r u3-indicator}
wages <- read.csv("data/wages-india-synthetic.csv")

is_urban <- as.numeric(wages$residence == "Urban")
head(is_urban, 20)
```

The *average* of that column is the *proportion* of people with the
characteristic. Adding up a column of ones and zeros counts the ones, and
dividing by $n$ turns the count into a share.

```{r u3-phat}
c(mean_of_indicator = mean(is_urban),
  proportion_urban  = sum(wages$residence == "Urban") / nrow(wages))
```

::: {.keyidea}
A proportion is a mean. The **sample proportion**

$$\hat{p} = \frac{x}{n}$$

where $x$ is the number of observations with the characteristic, is exactly the
sample mean of a variable coded 1 and 0.

Everything established about $\bar{X}$ therefore applies to $\hat{p}$
unchanged. It is unbiased, its variability falls with $\sqrt{n}$, and the
Central Limit Theorem makes it approximately normal in large samples.
:::

This is worth pausing on, because it is the first time in the book that a
result has been obtained for free. We did not prove anything new. We noticed
that a question we had not met was a question we had already answered, wearing
different clothes.

### How much does a sample proportion vary? {-}

One thing does simplify. For a general population we need $\sigma$ from
somewhere. For a 0/1 variable, $\sigma$ is determined by $p$ itself.

::: {.theory}
Let $X_i$ equal 1 with probability $p$ and 0 with probability $1-p$. Then

$$E(X_i) = 1 \cdot p + 0 \cdot (1-p) = p$$

so the population mean of the indicator *is* the population proportion. For the
variance, note that $X_i^2 = X_i$, since $0^2 = 0$ and $1^2 = 1$. Hence
$E(X_i^2) = p$ and

$$\mathrm{Var}(X_i) = E(X_i^2) - \big[E(X_i)\big]^2 = p - p^2 = p(1-p)$$

Applying $\mathrm{Var}(\bar{X}) = \sigma^2/n$ from the previous unit,

$$\mathrm{Var}(\hat{p}) = \frac{p(1-p)}{n}
\qquad\qquad
\mathrm{se}(\hat{p}) = \sqrt{\frac{p(1-p)}{n}}$$
:::

In practice $p$ is unknown — it is what we are estimating — so we substitute
$\hat{p}$:

$$\widehat{\mathrm{se}}(\hat{p}) = \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

```{r u3-se-phat}
set.seed(8)
sample_urban <- sample(is_urban, 500)

p_hat <- mean(sample_urban)
se_p  <- sqrt(p_hat * (1 - p_hat) / 500)

round(c(p_hat = p_hat, se = se_p, true_p = mean(is_urban)), 4)
```

::: {.intuition}
The quantity $p(1-p)$ is largest at $p = 0.5$ and falls away towards zero at
either end.

That has a sensible meaning. A population evenly split is the hardest to pin
down, because samples can land either way. A population where 99% share a
characteristic is easy — almost every sample says 99%. Rare and near-universal
things are estimated precisely; genuinely contested ones are not.

The consequence for survey design is that $p = 0.5$ is the worst case, and a
sample large enough for a 50-50 split is large enough for anything.
:::



## From a point estimate to an interval {#u3-ci}

::: {.question}
We have a number and a standard error. How do we turn them into a range?
:::

Suppose a survey of 40 students gives a mean attendance of 26.4 classes, and we
know from long experience that $\sigma = 5.45$. The standard error is
$5.45/\sqrt{40} = 0.86$.

We know from the Central Limit Theorem that $\bar{X}$ is approximately normal
around $\mu$ with that standard error. And we know a fact about normal
distributions: about 95% of the probability lies within 1.96 standard
deviations of the centre.

Put those together. Before the sample is drawn,

$$P\left(\mu - 1.96\frac{\sigma}{\sqrt{n}}
\;\le\; \bar{X} \;\le\;
\mu + 1.96\frac{\sigma}{\sqrt{n}}\right) \approx 0.95$$

The sample mean lands within 1.96 standard errors of the population mean 95% of
the time. Now rearrange the inequality so that $\mu$ sits in the middle rather
than $\bar{X}$ — the algebra is nothing more than moving terms across, but the
statement it produces is the one we want.

::: {.keyidea}
A **confidence interval** for $\mu$, when $\sigma$ is known, is

$$\bar{x} \;\pm\; z_{\alpha/2}\,\frac{\sigma}{\sqrt{n}}$$

The quantity $z_{\alpha/2}\,\sigma/\sqrt{n}$ is the **margin of error**. The
confidence level is $1 - \alpha$.

| Confidence | $\alpha$ | $z_{\alpha/2}$ |
|:--|:--|:--|
| 90% | 0.10 | 1.645 |
| 95% | 0.05 | 1.960 |
| 99% | 0.01 | 2.576 |

In R these come from `qnorm(1 - alpha/2)` — there is no need for a printed
table.
:::

```{r u3-first-ci}
x_bar <- 26.4
sigma <- 5.45
n     <- 40

for (level in c(0.90, 0.95, 0.99)) {
  z  <- qnorm(1 - (1 - level) / 2)
  ci <- x_bar + c(-1, 1) * z * sigma / sqrt(n)
  cat(sprintf("%2.0f%%  z = %.3f   [%.2f, %.2f]   width %.2f\n",
              100 * level, z, ci[1], ci[2], diff(ci)))
}
```

Higher confidence buys a wider interval, and this is the central trade-off of
the whole method. A 99% interval is more likely to contain $\mu$ precisely
because it commits to less. An interval certain to contain the truth would run
from minus infinity to infinity and tell us nothing at all.

### What does 95% actually refer to? {-}

The interval either contains $\mu$ or it does not. Nothing is random about it
once the sample is drawn — $\mu$ is a fixed number, and so are the two endpoints
sitting in front of us.

The 95% describes the **procedure**. If we drew sample after sample and
constructed an interval each time, about 95% of those intervals would contain
$\mu$. Since we cannot know whether ours is among them, we report the interval
and its coverage rate together.

We can watch this happen. Treat the 680 students as a population, so $\mu$ is
known, and construct 100 intervals from 100 samples.

```{r u3-coverage, fig.height = 4.2, fig.cap = "One hundred 95% confidence intervals, each from a fresh sample of 40 students. The vertical line is the true population mean."}
mu_pop    <- mean(att)
sigma_pop <- sqrt(mean((att - mean(att))^2))

set.seed(1)
ci_sim <- t(replicate(100, {
  x <- sample(att, 40, replace = TRUE)
  mean(x) + c(-1, 1) * 1.96 * sigma_pop / sqrt(40)
}))

covers <- ci_sim[, 1] <= mu_pop & ci_sim[, 2] >= mu_pop

d <- data.frame(id = 1:100, lo = ci_sim[, 1], hi = ci_sim[, 2],
                covers = ifelse(covers, "Contains μ", "Misses μ"))

ggplot(d, aes(y = id, xmin = lo, xmax = hi, colour = covers)) +
  geom_errorbarh(height = 0, linewidth = 0.45) +
  geom_vline(xintercept = mu_pop, linewidth = 0.7, colour = "grey20") +
  scale_colour_manual(values = c(book_palette$fill, book_palette$reject),
                      name = NULL) +
  labs(x = "Mean attendance", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank(), legend.position = "top")
```

Of these 100 intervals, `r sum(covers)` contain the population mean and
`r sum(!covers)` do not. The misses are not mistakes. They are the 5% the method
openly budgets for, and a procedure that never missed would be one that had
stopped saying anything.

Note also that the misses are not near-misses in any special sense — they are
simply the samples whose means happened to fall furthest from $\mu$. Nothing
about such a sample looks wrong from the inside. A researcher holding one of
those five would see a perfectly ordinary set of forty students.



## What if $\sigma$ is unknown? {#u3-t}

::: {.question}
Every interval so far used $\sigma$. We never know it. What breaks?
:::

Section \@ref(u3-variance) suggested substituting $s$ for $\sigma$, which looks
like a small matter. It is not, and the reason is worth taking seriously.

When we use $\sigma$, the only random quantity in the interval is $\bar{x}$.
When we use $s$, *both* the centre and the width are estimated from the same
sample. An unlucky sample can be unrepresentative in its spread as well as its
centre — and a sample that happens to have a small $s$ produces an interval
that is both narrow and badly placed. Those two errors compound rather than
cancel.

The effect is easy to measure. Take a normal population, draw samples of ten,
and build intervals using $s$ but the normal critical value 1.96.

```{r u3-t-coverage}
set.seed(6)
n <- 10

cover <- replicate(50000, {
  x  <- rnorm(n, mean = 26, sd = 5.45)
  se <- sd(x) / sqrt(n)
  c(using_1.96      = abs(mean(x) - 26) < 1.96 * se,
    using_t_critical = abs(mean(x) - 26) < qt(0.975, df = n - 1) * se)
})

round(rowMeans(cover), 4)
```

The interval advertised as 95% delivers about
`r round(100 * rowMeans(cover)[1], 1)`%. One statement in twenty was supposed to
be wrong; nearly one in twelve is. Using the $t$ critical value instead restores
it to `r round(100 * rowMeans(cover)[2], 1)`%.

The fix is to widen the interval by exactly enough to compensate, using a
distribution with heavier tails than the normal.

::: {.keyidea}
When $\sigma$ is unknown, the confidence interval for $\mu$ is

$$\bar{x} \;\pm\; t_{n-1,\,\alpha/2}\,\frac{s}{\sqrt{n}}$$

where $t_{n-1,\,\alpha/2}$ comes from the **$t$ distribution** with $n-1$
degrees of freedom — the same $n-1$ that appeared in $s^2$, and for the same
reason.

In R: `qt(1 - alpha/2, df = n - 1)`.
:::

The $t$ distribution is symmetric and bell-shaped like the normal, but with
more probability in the tails, which is what makes the interval wider. How much
wider depends on the degrees of freedom.

```{r u3-t-table}
ns <- c(5, 10, 20, 30, 60, 120)
round(data.frame(n = ns, df = ns - 1,
                 t_critical = qt(0.975, ns - 1),
                 z_critical = 1.96), 3)
```

The correction is substantial in small samples and negligible in large ones. By
$n = 120$ the two differ in the second decimal place, which is why the
distinction stops mattering once samples run to a few hundred — and why it
matters a great deal when a study has thirty observations.

::: {.watchout}
The $t$ distribution corrects for *estimating* $\sigma$. It does not correct
for a non-normal population.

Repeat the coverage experiment on the real attendance data rather than a normal
population, and at $n = 10$ the $t$ interval achieves about 90% rather than
95%. The remaining shortfall is the Central Limit Theorem not having done its
work yet at that sample size.

Two separate assumptions are in play, and $t$ addresses only one of them.
:::



## How large a sample do we need? {#u3-samplesize}

::: {.question}
A survey has not yet been run. How many households should it cover?
:::

Everything so far took the sample as given. In practice the question usually
arrives earlier, from someone who has to budget for the fieldwork.

The margin of error for a mean is

$$E = z_{\alpha/2}\,\frac{\sigma}{\sqrt{n}}$$

which can be solved for $n$:

$$n = \left(\frac{z_{\alpha/2}\,\sigma}{E}\right)^{2}$$

The squaring is the $\sqrt{n}$ rule from the previous unit, arriving from the
other direction: halving the margin of error quadruples the sample.

For a proportion the same rearrangement gives

$$n = \frac{z_{\alpha/2}^{2}\;p(1-p)}{E^{2}}$$

with one difficulty. It requires $p$, which is what the survey exists to find
out. The standard resolution is to use the worst case. Since $p(1-p)$ is
maximised at $p = 0.5$, setting $p = 0.5$ gives a sample size large enough
whatever the truth turns out to be.

```{r u3-samplesize}
sample_size_prop <- function(margin, level = 0.95, p = 0.5) {
  z <- qnorm(1 - (1 - level) / 2)
  ceiling(z^2 * p * (1 - p) / margin^2)
}

data.frame(margin_of_error = c("5%", "3%", "2%", "1%"),
           n_at_95 = sapply(c(0.05, 0.03, 0.02, 0.01), sample_size_prop),
           n_at_90 = sapply(c(0.05, 0.03, 0.02, 0.01), sample_size_prop,
                            level = 0.90))
```

These numbers explain something otherwise puzzling about published surveys. A
national opinion poll covering 1.4 billion people typically samples one or two
thousand — and the table shows why that is sufficient rather than negligent.
The population size does not appear anywhere in the formula. What determines
precision is the number of people asked, not the fraction of the country they
represent.

::: {.watchout}
This calculation answers a narrow question: how large a sample is needed for
the standard error to reach a target, *assuming the sample is drawn properly*.

It says nothing about the failures of Section \@ref(u2-iid), and the sample
size it returns will not repair any of them. A survey of two thousand urban
households is not a survey of India, however carefully the two thousand was
computed.
:::

### Intervals for a proportion {-}

Collecting the pieces, an interval for a population proportion is

$$\hat{p} \;\pm\; z_{\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

Note that $z$ appears here rather than $t$. The $t$ correction exists because
$\sigma$ was estimated separately from $\mu$; for a proportion, $\hat{p}$
determines the standard error on its own, so there is no second estimate to
account for.

Suppose that in a random sample of 1,000 secondary school teachers in
Karnataka, 518 are women.

```{r u3-prop-ci}
x <- 518
n <- 1000

p_hat <- x / n
se    <- sqrt(p_hat * (1 - p_hat) / n)

round(c(p_hat = p_hat, se = se,
        lower = p_hat - 1.96 * se,
        upper = p_hat + 1.96 * se), 4)
```

So we report that between 48.7% and 54.9% of secondary school teachers in
Karnataka are women, with 95% confidence.

That interval contains 0.5, which is worth noticing. The sample proportion is
above half, but the data are also consistent with women being a minority of the
profession. A headline reading "most secondary school teachers are women" would
be going beyond what a sample of a thousand can support.



## When only one direction matters {#u3-onesided}

::: {.question}
Sometimes we care about one end only. Why pay for the other?
:::

A confidence interval spends half its error budget at each end. Often only one
end is of interest.

A regulator checking whether a pollutant exceeds a limit does not care how low
it might be. A funder asking whether attendance is at least 24 classes does not
care how high it might be. In these cases the whole of $\alpha$ can be spent on
the side that matters, which buys a tighter bound.

::: {.keyidea}
A **one-sided confidence bound** uses $z_{\alpha}$ rather than $z_{\alpha/2}$:

$$\text{upper bound:}\quad \bar{x} + z_{\alpha}\frac{\sigma}{\sqrt{n}}
\qquad\qquad
\text{lower bound:}\quad \bar{x} - z_{\alpha}\frac{\sigma}{\sqrt{n}}$$

At 95% confidence, $z_{\alpha} = 1.645$ rather than 1.960.
:::

```{r u3-onesided}
x_bar <- 26.4; sigma <- 5.45; n <- 40

round(c(two_sided_lower = x_bar - qnorm(0.975) * sigma / sqrt(n),
        two_sided_upper = x_bar + qnorm(0.975) * sigma / sqrt(n),
        one_sided_upper = x_bar + qnorm(0.950) * sigma / sqrt(n),
        one_sided_lower = x_bar - qnorm(0.950) * sigma / sqrt(n)), 3)
```

The one-sided bound is closer to the estimate than the two-sided one at the
same confidence level. Nothing has been gained for free: we have given up any
claim about the other direction entirely.

::: {.watchout}
The choice between one-sided and two-sided must be made **before** looking at
the data, on the basis of what the question is.

Computing a two-sided interval, noticing it just fails to exclude some value of
interest, and switching to a one-sided bound to obtain a tighter result is not
a calculation. It is a way of arranging to be right, and the stated confidence
level no longer describes what was done.
:::



## What a confidence interval does not mean {#u3-misreadings}

::: {.question}
Four sentences that sound right. Which are, and why not?
:::

Confidence intervals are misread more often than any other object in
introductory statistics, and the misreadings are systematic enough to be worth
naming.

**"There is a 95% probability that $\mu$ lies in $[24.4,\ 27.8]$."**

This is the most common, and the most understandable. But $\mu$ is a fixed
number and the interval is a fixed pair of numbers; either the one is inside
the other or it is not. There is nothing left to assign a probability to. The
randomness lived in the sampling, and it was spent when the sample was drawn.
What is 95% is the *procedure's* long-run success rate.

**"95% of the population lies within the interval."**

The interval describes the plausible location of a population *mean*, not the
spread of individuals. In the attendance data the 95% interval is under two
classes wide, while individual students range across the whole register. An
interval for $\mu$ narrows towards a point as $n$ grows; the spread of the
population does not budge.

**"The two intervals overlap, so the groups are not different."**

Tempting, and wrong often enough to matter. Two intervals can overlap while the
difference between the groups is nevertheless larger than sampling variation
would explain. Comparing groups requires an interval for the *difference*,
which is the subject of a later unit — not an eyeball comparison of two
separate intervals.

**"A wider interval means the estimate is worse."**

It means the estimate is less precise, which is not the same thing. A wide
interval from a well-designed survey is an honest report of genuine
uncertainty. A narrow interval from a badly designed one, as Survey B in
Section \@ref(u2-iid) demonstrated, is a confident statement about the wrong
population. Width measures precision; nothing in the arithmetic measures
whether the sample was any good.

::: {.keyidea}
A confidence interval reports what the data can support, and no more. It cannot
detect a badly drawn sample, it says nothing about individuals, and its
confidence level is a property of the method rather than of the particular
numbers in front of you.
:::



## Summary {#u3-summary}

::: {.keyidea}
- The **sample variance** $s^2 = \frac{1}{n-1}\sum(x_i - \bar{x})^2$ is
  unbiased for $\sigma^2$. The divisor is $n-1$ because $\bar{x}$ was estimated
  from the same data, leaving $n-1$ free deviations. If $\mu$ is known, the
  divisor is $n$ and the deviations are taken from $\mu$.
- A **proportion is a mean** — of a variable coded 1 and 0 — so every result
  about $\bar{X}$ transfers to $\hat{p} = x/n$ without new proof.
- $\mathrm{se}(\hat{p}) = \sqrt{p(1-p)/n}$, largest at $p = 0.5$. Evenly split
  populations are the hardest to estimate; near-universal ones are easy.
- A **confidence interval** is $\bar{x} \pm z_{\alpha/2}\,\sigma/\sqrt{n}$ when
  $\sigma$ is known, and $\bar{x} \pm t_{n-1,\,\alpha/2}\,s/\sqrt{n}$ when it is
  not. For a proportion, $\hat{p} \pm z_{\alpha/2}\sqrt{\hat{p}(1-\hat{p})/n}$.
- The $t$ distribution compensates for estimating $\sigma$ from the same
  sample. It does not compensate for a non-normal population — two different
  assumptions, one correction.
- The confidence level is a property of the **procedure**: across repeated
  samples, that proportion of intervals contains the parameter. A particular
  interval either contains it or does not.
- Higher confidence means a wider interval. Halving the margin of error
  quadruples the required sample, and $n = (z_{\alpha/2}\sigma/E)^2$ makes that
  explicit before any fieldwork is commissioned.
- What determines precision is how many people were asked, not what fraction of
  the population they are. Population size appears in none of these formulas.
:::

Everything in this unit produced a *range* of values consistent with the data.
The next unit asks a sharper question: given a specific claim about the
population, is our sample consistent with it?


## Practice problems {#u3-practice}

::: {.yourturn}
These are for learning. Work each one before opening the solution.

The [question bank](#u3-question-bank) that follows has no answers at all.
:::

### Concept checks {-}

**3.1** Why does the sample variance divide by $n - 1$ rather than $n$?

::: {.answer}
Because the deviations are measured from $\bar{x}$, which was itself computed
from the same observations. The sample mean is the value that makes
$\sum(x_i - \bar{x})^2$ as small as possible, so measuring spread around it
gives an answer that is too small.

Formally, the deviations must sum to zero, so only $n - 1$ of them are free to
vary. Dividing by $n - 1$ corrects the shortfall exactly, making $s^2$
unbiased for $\sigma^2$.
:::

**3.2** A dataset records whether each of 900 households owns a bicycle. Why is
no new theory needed to estimate the proportion that do?

::: {.answer}
Because coding "owns" as 1 and "does not own" as 0 makes the proportion the
mean of that column. Summing ones counts the owners, and dividing by 900 turns
the count into a share.

So $\hat{p}$ *is* a sample mean, and everything already established —
unbiasedness, standard error falling with $\sqrt{n}$, approximate normality by
the Central Limit Theorem — applies without modification.
:::

**3.3** Explain why a 99% confidence interval is wider than a 95% one.

::: {.answer}
Higher confidence requires the interval to succeed more often across repeated
samples, and the only way to succeed more often is to claim less. Widening the
interval makes it easier to capture $\mu$.

The trade-off is unavoidable. An interval guaranteed to contain the truth would
have to span every possible value, and would carry no information.
:::

**3.4** For which value of $p$ is a proportion hardest to estimate precisely,
and why does that make sense?

::: {.answer}
$p = 0.5$, since $p(1-p)$ is maximised there.

It makes sense because an evenly split population is the one where samples can
most easily land either way. If 99% of households have electricity, almost any
sample says so. If 50% do, the sample proportion has real room to move.

This is why survey planners use $p = 0.5$ when computing sample sizes: it is
the worst case, so it is safe whatever the truth turns out to be.
:::

**3.5** A student reports a 95% interval of $[24.4, 27.8]$ and writes: "so 95%
of students attend between 24.4 and 27.8 classes." What has gone wrong?

::: {.answer}
The interval is about the *population mean*, not about individual students.

It says the average attendance is plausibly between 24.4 and 27.8. Individual
students vary far more widely — the standard deviation is about 5.5 classes,
so many students lie well outside that range.

Note also that the interval narrows as $n$ grows, whereas the spread of
students does not change at all. The two quantities behave differently, which
is a sign they are measuring different things.
:::

### Explain {-}

**3.6** Explain what the "95%" in a 95% confidence interval refers to, and why
it cannot refer to the particular interval you have computed.

::: {.answer}
It refers to the procedure. If samples were drawn repeatedly and an interval
constructed from each by this method, about 95% of those intervals would
contain the population parameter.

It cannot refer to the interval in front of you because nothing about it is
random any more. The population mean is a fixed number, and once the sample is
drawn, the two endpoints are fixed numbers too. Either $\mu$ is between them or
it is not; there is no probability left in the situation.

The randomness lived in which sample was drawn, and that was spent. The 95% is
a statement about the method's track record, which is why we report the
interval and its confidence level together — the level describes how much such
statements can generally be trusted.
:::

**3.7** Why is the $t$ distribution needed at all, and why does it stop
mattering in large samples?

::: {.answer}
When $\sigma$ is known, the only quantity estimated from the sample is
$\bar{x}$. When $\sigma$ is unknown, $s$ is estimated from the same data, so
both the centre and the width of the interval are subject to sampling error.
A sample with an unusually small $s$ produces an interval that is narrow *and*
poorly positioned, and those errors compound. Using 1.96 anyway gives about 92%
coverage at $n = 10$ instead of 95%.

The $t$ distribution has heavier tails, widening the interval by exactly enough
to restore the advertised coverage.

It stops mattering in large samples because $s$ becomes an increasingly
reliable estimate of $\sigma$. The critical value falls from 2.776 at $n = 5$
to 1.984 at $n = 100$, approaching 1.96 as the degrees of freedom grow. Once
$s$ is essentially known, there is nothing left to correct for.
:::

**3.8** A colleague computes a two-sided 95% interval, notices that it barely
includes zero, and recomputes it as a one-sided bound so that zero falls
outside. Explain the problem.

::: {.answer}
The confidence level describes a procedure fixed in advance. Once the choice
between one-sided and two-sided is made *after* seeing where the data fell, the
procedure being followed is no longer the one whose 95% property was
established.

The real procedure here is "compute a two-sided interval, and if it fails to
give the answer I want, switch." That procedure covers the truth far less than
95% of the time, because it has two chances to produce the conclusion sought.

One-sided bounds are legitimate when the question is genuinely one-directional
— a regulator who cares only about exceeding a limit. What makes them
illegitimate is choosing the direction from the data.
:::

### Numerical problems {-}

**3.9** A sample of 10 professors reported the following hours worked in a
week: 48, 22, 19, 65, 72, 37, 55, 60, 49, 28.

i. Estimate the population mean.
ii. Estimate the population variance and standard deviation.
iii. Estimate them again given that the population mean is known to be 46 hours.
iv. Why does (iii) use a different divisor?

::: {.answer}
**(i)** $\bar{x} = 455/10 = 45.5$ hours.

**(ii)** With $\mu$ unknown, deviations are taken from $\bar{x} = 45.5$ and
$\sum(x_i - \bar{x})^2 = 3034.5$:

$$s^2 = \frac{3034.5}{9} = 337.17 \qquad s = 18.36$$

**(iii)** With $\mu = 46$ known, deviations are taken from 46 and
$\sum(x_i - \mu)^2 = 3037$:

$$\widehat{\sigma}^2 = \frac{3037}{10} = 303.7 \qquad \widehat{\sigma} = 17.43$$

**(iv)** In (ii) the centre was estimated from the sample, costing one degree
of freedom. In (iii) it was known, so none was spent.

Notice that the sum of squares is *larger* in (iii) — 3037 against 3034.5 —
because $\bar{x}$ minimises that sum and $\mu$ is not $\bar{x}$. The larger
total divided by the larger $n$ is what keeps both estimates unbiased.
:::

**3.10** The average life of a sample of 10 tyres was 28,400 km. Lifetimes are
normally distributed with $\sigma = 3{,}300$ km.

i. Construct 90%, 95% and 99% confidence intervals for the mean life.
ii. Find a value that, with 95% confidence, is larger than the population mean.
iii. Find a value that, with 99% confidence, is smaller than the population
     mean.

::: {.answer}
$\sigma/\sqrt{n} = 3300/\sqrt{10} = 1043.6$.

**(i)**

$$\begin{aligned}
90\%:&\quad 28400 \pm 1.645(1043.6) = [26683,\ 30117] \\
95\%:&\quad 28400 \pm 1.960(1043.6) = [26355,\ 30445] \\
99\%:&\quad 28400 \pm 2.576(1043.6) = [25712,\ 31088]
\end{aligned}$$

**(ii)** A one-sided upper bound at 95% uses $z_{0.05} = 1.645$:
$28400 + 1.645(1043.6) = 30{,}117$ km.

**(iii)** A one-sided lower bound at 99% uses $z_{0.01} = 2.326$:
$28400 - 2.326(1043.6) = 25{,}973$ km.

Note that (ii) uses 1.645 rather than 1.960: the whole 5% sits in one tail,
because we are making a claim in one direction only.
:::

**3.11** In a random sample of 400 death certificates of college students, 98
recorded a motorcycle accident.

i. Estimate the proportion of such deaths due to motorcycle accidents.
ii. Estimate the standard error of that estimate.
iii. Construct a 95% confidence interval.

::: {.answer}
**(i)** $\hat{p} = 98/400 = 0.245$.

**(ii)**

$$\widehat{\mathrm{se}}(\hat{p}) = \sqrt{\frac{0.245 \times 0.755}{400}}
= \sqrt{0.000462} = 0.0215$$

**(iii)** $0.245 \pm 1.96(0.0215) = [0.203,\ 0.287]$.

So between about 20% and 29%, with 95% confidence.
:::

**3.12** In a random sample of 1,000 secondary school teachers in Karnataka,
518 identify as women.

i. Construct a 95% confidence interval for the population proportion.
ii. Construct a 90% upper confidence bound.
iii. Can the state conclude that women are a majority of the profession?

::: {.answer}
$\hat{p} = 0.518$ and
$\widehat{\mathrm{se}} = \sqrt{0.518 \times 0.482 / 1000} = 0.0158$.

**(i)** $0.518 \pm 1.96(0.0158) = [0.487,\ 0.549]$.

**(ii)** $0.518 + 1.282(0.0158) = 0.538$.

**(iii)** No. The 95% interval contains 0.5, so the data are also consistent
with women being slightly under half of secondary school teachers. The point
estimate is above half, but a sample of a thousand cannot separate 51.8% from
50% with confidence.

This is a good illustration of why the interval is reported rather than the
point estimate alone. "51.8% of teachers are women" and "between 48.7% and
54.9% are" support very different headlines.
:::

**3.13** A state wants to estimate the proportion of households with a
functioning water connection to within 2 percentage points at 95% confidence.

i. How many households must be surveyed?
ii. A pilot suggests the proportion is near 0.85. How many are needed now?
iii. Why is (i) larger, and when should it still be preferred?

::: {.answer}
**(i)** With no prior information, use the worst case $p = 0.5$:

$$n = \frac{1.96^2 (0.5)(0.5)}{0.02^2} = \frac{0.9604}{0.0004} = 2401$$

**(ii)** With $p = 0.85$, $p(1-p) = 0.1275$:

$$n = \frac{1.96^2 (0.1275)}{0.0004} = 1225$$

Roughly half as many.

**(iii)** Because $p(1-p)$ is largest at 0.5, so that choice always gives the
biggest requirement and therefore a safe one.

It should still be preferred when the pilot is thin, or drawn from a
non-representative area, or when the true proportion might differ across the
state. If the real figure turns out to be 0.5 and the survey was sized for
0.85, the margin of error will exceed the 2 points promised.
:::

### R exercises {-}

**3.14** Using `data/attendance-grades.csv`, compute point estimates of the
mean, variance and standard deviation of attendance, semester GPA and final
score. Then construct 95% confidence intervals for each mean.

::: {.answer}
```{r u3-ex-points}
students <- read.csv("data/attendance-grades.csv")
vars     <- c("attendance", "sem_gpa", "final_score")

est <- t(sapply(vars, function(v) {
  x  <- students[[v]]
  n  <- length(x)
  se <- sd(x) / sqrt(n)
  tc <- qt(0.975, df = n - 1)
  c(mean = mean(x), var = var(x), sd = sd(x), se = se,
    lower = mean(x) - tc * se, upper = mean(x) + tc * se)
}))

round(est, 3)
```

The intervals are narrow because $n = 680$ is large. Note that `var()` and
`sd()` already use the $n - 1$ divisor, so no manual correction is needed.
:::

**3.15** Verify the $n-1$ correction yourself. Treat attendance as a
population, draw 20,000 samples of 8, and compare the average of
$\frac{1}{n}\sum(x_i - \bar{x})^2$ with the average of
$\frac{1}{n-1}\sum(x_i - \bar{x})^2$.

::: {.answer}
```{r u3-ex-nminus1}
att <- students$attendance
sigma2 <- mean((att - mean(att))^2)

set.seed(21)
n <- 8
out <- replicate(20000, {
  x <- sample(att, n, replace = TRUE)
  c(over_n = mean((x - mean(x))^2), over_n_1 = var(x))
})

round(c(population_variance = sigma2,
        avg_over_n          = mean(out[1, ]),
        avg_over_n_1        = mean(out[2, ]),
        predicted_over_n    = sigma2 * (n - 1) / n), 3)
```

Dividing by $n$ understates the population variance, and it does so by exactly
the predicted factor $(n-1)/n = 7/8$. Dividing by $n - 1$ recovers the right
answer on average.

The bias is not small at this sample size — about 12% — which is why the
correction is built into `var()` rather than left to the user.
:::

**3.16** Treat `wages-india-synthetic.csv` as a population and let $p$ be the
proportion of workers in urban areas. Draw 1,000 samples of 400, construct a
95% interval from each, and count how many contain the true $p$.

::: {.answer}
```{r u3-ex-prop-coverage}
wages    <- read.csv("data/wages-india-synthetic.csv")
is_urban <- as.numeric(wages$residence == "Urban")
p_true   <- mean(is_urban)

set.seed(4)
ci <- t(replicate(1000, {
  s  <- sample(is_urban, 400, replace = TRUE)
  ph <- mean(s)
  ph + c(-1, 1) * 1.96 * sqrt(ph * (1 - ph) / 400)
}))

c(true_p = p_true,
  coverage = mean(ci[, 1] <= p_true & ci[, 2] >= p_true))
```

Close to the advertised 95%. The method delivers what it promises for a
proportion of this size at this sample size.

It would perform less well for a very rare characteristic — with $p = 0.01$ and
$n = 400$ the expected number of successes is 4, the normal approximation is
poor, and coverage falls short. A common rule of thumb requires both $np$ and
$n(1-p)$ to be at least 10.
:::

**3.17** Show that the $t$ correction matters. On a normal population with
$\mu = 50$ and $\sigma = 12$, compare the coverage of intervals built with 1.96
against those built with the $t$ critical value, for $n = 5, 10, 30$ and 100.

::: {.answer}
```{r u3-ex-t}
set.seed(30)
cover <- t(sapply(c(5, 10, 30, 100), function(n) {
  r <- replicate(20000, {
    x  <- rnorm(n, 50, 12)
    se <- sd(x) / sqrt(n)
    c(abs(mean(x) - 50) < 1.96 * se,
      abs(mean(x) - 50) < qt(0.975, n - 1) * se)
  })
  c(n = n, using_z = mean(r[1, ]), using_t = mean(r[2, ]))
}))

round(cover, 4)
```

The $t$ intervals hold their 95% at every sample size. The $z$ intervals fall
well short when $n$ is small, and the gap closes as $n$ grows — by $n = 100$
the two are nearly indistinguishable.

This is the whole case for the $t$ distribution in one table, and also the
reason nobody worries about it in large samples.
:::

**3.18** How does interval width respond to sample size and confidence level?
Using the attendance population, plot the width of the confidence interval for
the mean against $n$, for the 90%, 95% and 99% levels.

::: {.answer}
```{r u3-ex-width, fig.height = 3.4, fig.cap = "Confidence interval width against sample size, at three confidence levels."}
sigma_att <- sqrt(mean((att - mean(att))^2))

grid <- expand.grid(n = seq(10, 500, by = 5),
                    level = c(0.90, 0.95, 0.99))
grid$width <- 2 * qnorm(1 - (1 - grid$level) / 2) * sigma_att / sqrt(grid$n)
grid$level <- factor(paste0(100 * grid$level, "%"))

ggplot(grid, aes(n, width, colour = level)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Sample size", y = "Interval width (classes)", colour = NULL) +
  theme_book() +
  theme(legend.position = "top")
```

Width falls as $1/\sqrt{n}$, steeply at first and then hardly at all. Most of
the available precision is bought in the first hundred observations.

The three curves never cross, and the gap between them narrows in absolute
terms as $n$ grows. Choosing 99% over 95% is expensive in a small survey and
nearly free in a large one.
:::


## Question bank {#u3-question-bank}

::: {.watchout}
**No answers are given for this section, by design.** These are for testing
yourself once the practice problems are complete.
:::

### Multiple choice {-}

**QB 3.1** The divisor $n - 1$ in the sample variance is used because

a. it makes the variance larger, which is safer
b. $\bar{x}$ was estimated from the same data, leaving $n-1$ free deviations
c. the population is finite
d. the sample standard deviation would otherwise be negative

**QB 3.2** A 95% confidence interval for $\mu$ is $[112, 128]$. Which statement
is correct?

a. 95% of the population lies between 112 and 128
b. There is a 95% probability that $\mu$ lies between 112 and 128
c. 95% of intervals built this way would contain $\mu$
d. The sample mean has a 95% chance of lying between 112 and 128

**QB 3.3** Holding everything else fixed, moving from 95% to 99% confidence

a. narrows the interval
b. widens the interval
c. leaves the width unchanged but shifts the centre
d. requires a larger sample

**QB 3.4** For a sample proportion, the standard error is largest when
$\hat{p}$ is closest to

a. 0
b. 0.25
c. 0.5
d. 1

**QB 3.5** A survey of 600 households finds 30% own a two-wheeler. The
approximate 95% margin of error is

a. 0.4 percentage points
b. 1.9 percentage points
c. 3.7 percentage points
d. 7.5 percentage points

**QB 3.6** The $t$ distribution is used instead of the normal when

a. the sample is small
b. $\sigma$ is estimated from the sample
c. the population is skewed
d. the confidence level exceeds 95%

**QB 3.7** To halve the margin of error of a survey, the sample size must be
multiplied by

a. $\sqrt{2}$
b. 2
c. 4
d. It depends on the confidence level

**QB 3.8** Which of the following would *not* narrow a confidence interval for
$\mu$? *(select all that apply)*

a. Increasing the sample size
b. Lowering the confidence level
c. Sampling from a population with smaller $\sigma$
d. Sampling only from a subgroup that happens to be homogeneous

### Short reasoning {-}

**QB 3.9** A report gives a 95% confidence interval for mean household income of
[₹18,200, ₹19,600] and concludes: "almost all households in this district earn
between ₹18,200 and ₹19,600." Identify the error and state what the interval
does say.

**QB 3.10** Explain why the population size does not appear in the formula for
the margin of error, and why this surprises people.

**QB 3.11** Two districts are surveyed. District A's 95% interval for mean
income is [₹19,000, ₹23,000]; District B's is [₹22,000, ₹26,000]. A journalist
writes that the districts do not differ significantly because the intervals
overlap. Evaluate.

**QB 3.12** A researcher has a sample of 12 observations from a strongly
skewed population and constructs a $t$ interval. State two distinct assumptions
being relied on, and say which one the $t$ distribution addresses.

**QB 3.13** Explain why $s$ is not an unbiased estimator of $\sigma$ even
though $s^2$ is unbiased for $\sigma^2$, and why this is rarely mentioned in
practice.

**QB 3.14** A polling organisation reports a margin of error of ±3 points but
sampled only households with landline telephones. Explain what the ±3 does and
does not cover.

### Numerical {-}

**QB 3.15** A sample of 16 bags of rice has mean weight 24.6 kg and sample
standard deviation 1.2 kg.

i. Construct a 95% confidence interval for the mean weight.
ii. Construct a 99% confidence interval.
iii. The label claims 25 kg. Comment.
iv. How would your answer to (i) change if the standard deviation of 1.2 kg
    were known to be the population value?

**QB 3.16** In a random sample of 250 farm households in a district, 145
reported using irrigation.

i. Estimate the proportion and its standard error.
ii. Construct a 90% confidence interval.
iii. Construct a 95% lower confidence bound.
iv. A scheme guarantees that at least half of households have irrigation. Is
    the sample consistent with that?

**QB 3.17** A state government wishes to estimate mean annual household
expenditure to within ₹500 at 95% confidence. Previous rounds suggest
$\sigma \approx ₹9{,}000$.

i. What sample size is required?
ii. Fieldwork costs ₹1,100 per household. What is the budget?
iii. The budget is capped at ₹15 lakh. What margin of error is achievable?

**QB 3.18** Using `data/wages-india-synthetic.csv`, write R code to construct
95% confidence intervals for mean annual earnings separately for urban and
rural workers, using samples of 300 from each. Report both intervals, state
whether they overlap, and explain carefully what that does and does not tell
you about the difference between the two groups.


## Looking ahead {#u3-looking-ahead}

We can now take one sample and produce a range of values consistent with it.
That is a considerable advance on a single number, and for many purposes it is
the whole answer — a survey estimate reported with its margin of error is doing
exactly this.

But notice the shape of what we have been asking. We began with the data and
worked outwards to whatever range they would support.

Many questions arrive the other way round. A government claims average
attendance meets a target. A manufacturer claims a mean lifetime. A previous
round of a survey reported a figure, and we want to know whether things have
changed. In each case a specific value is proposed *before* the data are
examined, and the question is whether the sample is consistent with it.

That is a different question, and it needs its own machinery — though, as we
will see, one built from precisely the same parts.
