# Hypothesis Testing: One Sample {#one-sample}

```{r setup-11, include = FALSE}
source("_common.R")
library(ggplot2)
```

In the previous unit we learned that every estimate comes with uncertainty. A
confidence interval tells us which population values are plausible given our
sample.

But many real statistical questions begin with a specific claim. Has average
monsoon rainfall changed? Is the unemployment rate below 5%? Does a new drug
lower blood pressure?

Rather than asking *which values are plausible*, these questions ask *is this
particular claim consistent with the data?* Answering that is the purpose of
hypothesis testing.

Consider a claim made about a population. You collect a sample. Your estimate
is not exactly what the claim predicted.

Should you conclude that the claim is wrong?

Or is this simply the kind of difference that appears because every random
sample is slightly different?

That question is the starting point for hypothesis testing, and it is the
engine behind much of modern statistics.

## The problem, before any formulas {#os-problem}

The APU administration claims the average height of all students is 165 cm.

You are sceptical. You could settle it by measuring all 3,500 students, but that
costs weeks of work. So instead you measure 36 students chosen at random, and
you find their average height is **168 cm**.

The claim said 165. Your sample says 168. Is the administration wrong?

::: {.watchout}
The tempting answer is "yes, obviously — 168 is not 165." That answer is wrong,
and understanding *why* it is wrong is the whole chapter.

You did not measure everyone. You measured 36 people. Even if the true average
really were exactly 165, your particular 36 students would almost never average
exactly 165. Some samples come out high, some low. **A gap between the sample
and the claim is guaranteed to exist even when the claim is perfectly true.**
:::

Let us actually watch that happen.

We obviously do not know whether the administration is correct. But imagine,
just for a moment, that they are. If the true average really were 165 cm, with
a standard deviation of 8 cm, what kinds of sample means would we expect to
see? Draw a sample of 36 students, ten separate times, and look.

```{r os-sim-ten}
set.seed(42)

ten_samples <- replicate(10, mean(rnorm(n = 36, mean = 165, sd = 8)))
round(ten_samples, 2)
```

Every one of those numbers came from a population whose mean is *exactly* 165.
Not one of them is 165. Several are further from 165 than a full centimetre.

So the real question is not "is 168 different from 165?" It obviously is. The
real question is:

::: {.intuition}
**If the administration's claim were true, how often would a sample of 36
students come out as far from 165 as ours did?**

If the answer is "quite often" — then our sample is unremarkable, and we have no
grounds to call anyone a liar.

If the answer is "almost never" — then either something very unusual happened
to us, or the administration's claim is not correct. In statistics we usually
regard the second explanation as the more convincing one.
:::

We can answer that question by brute force before we do any algebra at all.
Let's draw not ten samples but fifty thousand, still assuming the claim is true,
and count how many stray at least 3 cm away from 165 (as ours did).

```{r os-sim-many}
set.seed(42)

many_means <- replicate(50000, mean(rnorm(n = 36, mean = 165, sd = 8)))

# How far did our real sample fall from the claim?
observed_gap <- abs(168 - 165)

mean(abs(many_means - 165) >= observed_gap)
```

About `r round(100 * mean(abs(many_means - 165) >= observed_gap), 1)`% of the
time. So if the administration is telling the truth, a sample like ours turns
up roughly once in forty tries.

We have now answered the question using simulation alone, and without naming a
single technical term. If the claim were true, a sample mean at least 3 cm away
from 165 would appear about once in forty surveys.

That makes our sample unusual. But how unusual is unusual *enough*? Once in
forty is rare; would once in ten have been rare enough? Once in a hundred? The
rest of this chapter builds a systematic way of answering exactly that — and by
the end of it you will be able to obtain the number above exactly, in one line,
with no simulation at all.

```{r os-sim-plot, echo = FALSE, fig.cap = "Fifty thousand sample means, each from 36 students, drawn from a population whose true mean is 165 cm (dashed line). The solid line marks our observed sample mean of 168 cm. Only about 2.4% of the simulated samples fall at least this far from the claimed value in either direction, which is what makes our sample surprising if the claim is true."}
df <- data.frame(m = many_means)

ggplot(df, aes(m)) +
  geom_histogram(bins = 60, fill = book_palette$fill, colour = "white",
                 linewidth = 0.2) +
  geom_vline(xintercept = 165, colour = book_palette$ink,
             linetype = "dashed", linewidth = 0.7) +
  geom_vline(xintercept = 168, colour = book_palette$reject, linewidth = 1) +
  annotate("label", x = 168, y = Inf, vjust = 1.4, label = "our sample: 168",
           colour = book_palette$reject, fill = "white", size = 3.6,
           label.size = 0.3) +
  annotate("label", x = 165, y = Inf, vjust = 3.4, label = "the claim: 165",
           colour = book_palette$ink, fill = "white", size = 3.6,
           label.size = 0.3) +
  labs(x = "sample mean height (cm)", y = "how often") +
  theme_book()
```

## Naming the two hypotheses {#os-hypotheses}

::: {.question}
We have the reasoning. What are the two positions we are reasoning between?
:::

The previous section did the whole of a hypothesis test without naming
anything. We assumed the administration's claim, asked how often a sample would
stray as far from it as ours did, and found the answer was about once in forty.

Let us give the two positions in that argument their names.

We start from a sample $x_1, x_2, \ldots, x_n$ drawn from a population with
unknown mean $\mu$ and standard deviation $\sigma$. The claim under scrutiny is
the **null hypothesis**:

$$H_0 : \mu = 165$$

The position we would move to if the claim collapses is the **alternative
hypothesis**:

$$H_A : \mu \neq 165$$

### Why begin by assuming the claim is true? {-}

At first this looks backwards. If we suspect the claim is wrong, why start by
supposing it is right?

Because the assumption is what gives us something to measure against. Notice
what we actually did in the previous section: we could only simulate those
fifty thousand samples *because* we had a specific value of $\mu$ to simulate
from. Had we begun with "the mean is not 165," there would have been nothing to
draw samples from — not 165 covers 164, and 170, and 3,000, and each would give
a different answer.

So we grant the claim, temporarily, and see where it leads. If our sample would
be quite ordinary under it, we have no reason to doubt it. If our sample would
be extraordinarily unlikely under it, then the claim itself becomes hard to
believe.

::: {.keyidea}
The null hypothesis is always the *boring* option — nothing is going on, the
claim stands, the difference is zero. We never *prove* it. We only ever fail to
knock it down.

This is why we say "**fail to reject** $H_0$" and never "accept $H_0$." A court
that acquits has not proven innocence; it has failed to prove guilt.
:::

### Which hypothesis does the claim go in? {#os-which-hypothesis}

::: {.question}
A claim arrives. Does it become $H_0$ or $H_A$?
:::

This is one of the most common places for a beginner to go wrong, and the way
out is not to memorise a rule but to ask a question first.

Before writing anything down, ask yourself:

> What result would convince me?

Suppose a report claims that the mean CAT score of applicants is *more than*
499. What would persuade you? Almost everyone answers the same way: a sample
mean comfortably **above** 499. A sample averaging 470 would not support the
report, and neither would one averaging exactly 499.

That answer has already determined the test. The thing you would find
convincing — a mean above 499 — is the thing you are gathering evidence *for*,
and a test only ever produces evidence *against* $H_0$. So the claim has to sit
in the alternative:

$$H_0 : \mu \leq 499 \qquad H_A : \mu > 499$$

The direction follows too. You would be convinced by large values, so it is
large values that must count against the null, and the test looks in the upper
tail.

::: {.keyidea}
Two questions settle every case.

**What am I trying to gather evidence for?** That goes in $H_A$. A test can
only accumulate evidence against the null, so whatever you hope to establish
must be what survives when the null falls.

**Which direction would count as evidence?** That fixes the inequality in
$H_A$, and with it the tail the test looks in.
:::

Examination questions usually signal this in their phrasing, and the signal is
worth recognising.

Table: (\#tab:os-claim-table) How the usual phrasings map onto the two questions above.

| The question asks | The claim becomes | Because |
|:--|:--|:--|
| "Is there evidence to **support** the claim?" | $H_A$ | The claim is what you are gathering evidence for. |
| "Can you **reject** the claim?" | $H_0$ | The claim is what you are challenging, and you are asking whether the data can knock it down. |

### Where the equality goes {-}

One structural rule holds in every case, and it will save you from a good deal
of trouble:

::: {.keyidea}
**The equality always belongs in the null hypothesis.**

$H_0$ takes $=$, $\leq$ or $\geq$. $H_A$ takes $\neq$, $<$ or $>$, and never
equality.
:::

The reason is the one from earlier in this section. Every test begins by
assuming one specific value of $\mu$ and computing how likely our sample would
be under it. Only a hypothesis containing an equality supplies such a value —
$\mu \leq 499$ can be tested at its boundary, 499, whereas $\mu > 499$ names no
particular number to compute with.

::: {.watchout}
A claim phrased with a *strict* inequality — "the mean is more than 499" —
therefore cannot go into $H_0$ as stated. If you are asked for evidence
supporting it, the pair is

$$H_0 : \mu \leq 499 \qquad H_A : \mu > 499$$

Writing $H_0: \mu \geq 499$ is the common error. It puts the claim in the null
*and* points the test in the wrong direction, so a sample mean **above** 499 —
evidence *for* the claim — ends up producing a $p$-value near 1 and a "fail to
reject." The arithmetic is fine; the conclusion is backwards.
:::

### The CAT report, worked through {-}

A report claims the mean CAT score of applicants is more than 499. A sample of
100 applicants has mean 502, with $\sigma = 10.6$. Is there enough evidence at
$\alpha = 0.01$ to support the report?

We settled the hypotheses above by asking what would convince us:

$$H_0 : \mu \leq 499 \qquad H_A : \mu > 499$$

Right-tailed, so the critical value is $Z_{0.01} = 2.33$ — positive, not
negative.

Before computing anything, check the direction. The sample mean is 502 and the
claim is "more than 499." The sample points the *same way* as the claim, so
whatever the arithmetic returns, it must come out favourable to the report. If
it does not, the tail has been pointed the wrong way.

```{r os-claim-worked}
xbar_c <- 502; mu_c <- 499; sigma_c <- 10.6; n_c <- 100

z_c <- (xbar_c - mu_c) / (sigma_c / sqrt(n_c))
c(z_calc = z_c,
  z_crit = qnorm(0.99),
  p_value = pnorm(z_c, lower.tail = FALSE))   # upper tail: H_A is ">"
```

$Z_{calc} = 2.83 > 2.33$, and $p = 0.0023 < 0.01$. We **reject $H_0$**, which
means the data **support** the report's claim — and the answer points the way
the sanity check said it would.

::: {.intuition}
Make that check a habit, and make it before you trust any $p$-value rather than
after.

Ask which way the sample points relative to the claim, and decide what answer
you should therefore expect. If the arithmetic hands back something else — a
$p$-value of 0.998 and "no support" when the sample sits above the claimed
value — the mistake is almost always in $H_A$, not in the calculation.
:::


## How far is too far? {#os-critical}

Write $\mu_0$ for the value the null claims (here, 165). We will reject $H_0$
when the sample mean lands too far from $\mu_0$ — that is, when

$$|\bar{x} - \mu_0| \geq c$$

for some cutoff $c$. Everything now hinges on choosing $c$.

::: {.theory}
We choose $c$ so that, **if the null were true**, a gap that large would occur
with probability only $\alpha$ — a small risk we decide in advance, typically
0.05:

$$P\big[|\bar{x} - \mu_0| \geq c\big] = \alpha$$

From the Central Limit Theorem (Chapter \@ref(one-sample)) we know $\bar{x}$ is
normally distributed with mean $\mu$ and standard deviation $\sigma/\sqrt{n}$.
Dividing through by that standard error turns the left-hand side into a standard
normal:

$$
\begin{aligned}
P\left[\frac{|\bar{x} - \mu_0|}{\sigma/\sqrt{n}} \geq \frac{c}{\sigma/\sqrt{n}}\right] &= \alpha \\[4pt]
P\left[|Z| \geq \frac{c}{\sigma/\sqrt{n}}\right] &= \alpha
\end{aligned}
$$

The event $|Z| \ge k$ splits into two symmetric tails, each carrying half the
probability:

$$
2 \cdot P\left[Z \geq \frac{c}{\sigma/\sqrt{n}}\right] = \alpha
\qquad\Longrightarrow\qquad
P\left[Z \geq \frac{c}{\sigma/\sqrt{n}}\right] = \frac{\alpha}{2}
$$

By definition $P[Z \geq Z_{\alpha/2}] = \alpha/2$, so the bracketed quantity
*is* $Z_{\alpha/2}$, and therefore

$$\boxed{\ c = Z_{\alpha/2}\,\frac{\sigma}{\sqrt{n}}\ }$$
:::

Substituting that $c$ back into the rejection rule and dividing both sides by
the standard error gives the form you will actually use:

$$
\frac{|\bar{x} - \mu_0|}{\sigma/\sqrt{n}} \geq Z_{\alpha/2}
$$

The quantity on the left is the **test statistic**, written $Z_{calc}$. The
value on the right is the **critical value**. The rule is simply: *if the
statistic exceeds the critical value, reject the null.*

::: {.inr}
Those critical values are the ones you have been reading off a printed table.
There is no need — `qnorm()` gives you any of them to as many decimals as you
like.

```{r os-crit-z}
alpha <- c(0.10, 0.05, 0.01)

# Two-tailed: we want the value leaving alpha/2 in the upper tail
qnorm(1 - alpha / 2)
```

Those three numbers are the 1.645, 1.96 and 2.58 you have memorised. Now you
know where they come from, and you will never need the table again.
:::

## The worked example, both ways {#os-worked}

Back to the students. We have $\bar{x} = 168$, $\mu_0 = 165$, $\sigma = 8$,
$n = 36$, and we will work at $\alpha = 0.05$.

### On paper {-}

**Step 1 — Hypotheses.**

$$H_0 : \mu = 165 \qquad H_A : \mu \neq 165$$

**Step 2 — Critical value.** Two-tailed at $\alpha = 0.05$, so
$Z_{0.025} = 1.96$.

**Step 3 — Test statistic.**

$$
Z_{calc} = \frac{\bar{x} - \mu_0}{\sigma/\sqrt{n}}
= \frac{168 - 165}{8/\sqrt{36}}
= \frac{3}{8/6}
= \frac{3}{1.333}
= 2.25
$$

**Step 4 — Decision.** $|2.25| > 1.96$, so we **reject** $H_0$ at the 5% level.

**Step 5 — The $p$-value.**

$$
P(|Z| > 2.25) = 2 \times P(Z > 2.25) = 2 \times 0.01222 = 0.0244
$$

If the administration's claim were true, only 2.4% of samples would sit as far
from 165 as ours. That is unlikely enough to reject the claim.

### The same thing in R {-}

Now watch every one of those numbers reappear.

```{r os-byhand-r}
xbar  <- 168
mu0   <- 165
sigma <- 8
n     <- 36

se     <- sigma / sqrt(n)          # standard error
z_calc <- (xbar - mu0) / se        # the test statistic
z_crit <- qnorm(1 - 0.05 / 2)      # the critical value
p_val  <- 2 * pnorm(-abs(z_calc))  # the p-value

c(se = se, z_calc = z_calc, z_crit = z_crit, p_value = p_val)
```

The standard error is `r round(8/6, 4)`, the test statistic is
`r round((168-165)/(8/6), 3)`, and the $p$-value is
`r round(2*pnorm(-2.25), 4)` — exactly the hand calculation, and near enough
exactly the `r round(mean(abs(many_means - 165) >= 3), 4)` we got by brute-force
simulation at the start of the chapter.

::: {.intuition}
Notice what just happened. The simulation, the algebra, and the R code are three
descriptions of one idea. The formula is not a rule handed down from above — it
is a shortcut for the counting exercise we did by hand in Section
\@ref(os-problem).
:::

And here is the picture the numbers are describing:

```{r os-reject-plot, fig.cap = "The rejection region at $\\alpha = 0.05$ (shaded), with our test statistic of 2.25 falling inside it."}
plot_rejection(stat = 2.25, crit = 1.96, tails = "two",
               stat_label = "Z = 2.25")
```

### Letting a package do it {-}

For a real dataset rather than summary statistics, `z.test()` from the **BSDA**
package does the whole thing at once. Since our problem quoted only summary
statistics, we first build a sample consistent with them:

```{r os-ztest}
library(BSDA)

heights <- make_sample(n = 36, mean = 168, sd = 8)

z.test(heights, mu = 165, sigma.x = 8, conf.level = 0.95)
```

Read that output against the hand calculation: `z = 2.25`, `p-value = 0.0244`,
and a confidence interval that we are about to explain.

::: {.watchout}
`z.test()` needs `sigma.x` — the **population** standard deviation. If you do
not know it (and in real work you almost never do), you must not use a
$Z$-test. Use `t.test()` instead; see Section \@ref(os-ttest).
:::

## What a $p$-value actually says {#os-pvalue}

The $p$-value is the single most misquoted number in empirical work, so let us
be exact.

::: {.keyidea}
The $p$-value is the probability of seeing a sample **at least as extreme as
ours**, *given that the null hypothesis is true*.

It is **not** the probability that the null is true. It is **not** the
probability that we made a mistake. It is a statement about the data, computed
under an assumption about the world — not a statement about the world.
:::

Small $p$-value means the data would be surprising if the null held, which is
evidence against the null. Large $p$-value means the data are unremarkable
under the null, so we have no case.

Table: (\#tab:os-pval-table) Reading a $p$-value.

| $p$-value | Read it as | At $\alpha = 0.05$ |
|:--|:--|:--|
| 0.60 | Completely ordinary under the null. | Fail to reject |
| 0.18 | A bit high, but this happens roughly 1 time in 6. | Fail to reject |
| 0.049 | Would happen about 1 time in 20. | Reject |
| 0.0001 | Would happen 1 time in 10,000. Strong evidence. | Reject |

## The same test, viewed as an interval {#os-ci}

::: {.question}
Unit \@ref(interval-estimation) built confidence intervals; this unit builds
tests. Are they two methods, or one?
:::

We now have two procedures that use the same sample, the same standard error
and the same critical value. It would be strange if they were unrelated, and
they are not.

Start by computing the interval for our students, exactly as in the previous
unit.

```{r os-ci-calc}
xbar + c(-1, 1) * qnorm(0.975) * (8 / sqrt(36))
```

The 95% interval is $[165.39,\ 170.61]$. The claimed value of 165 lies
**outside** it — the same verdict the test reached, and not a coincidence.

The reason is visible in the algebra. We fail to reject $H_0$ when the test
statistic is small enough, that is when

$$\left|\frac{\bar{x} - \mu_0}{\sigma/\sqrt{n}}\right| \le z_{\alpha/2}$$

Multiplying through by $\sigma/\sqrt{n}$ and rearranging for $\mu_0$ gives

$$\bar{x} - z_{\alpha/2}\frac{\sigma}{\sqrt{n}}
\;\le\; \mu_0 \;\leq\;
\bar{x} + z_{\alpha/2}\frac{\sigma}{\sqrt{n}}$$

which is the confidence interval. The condition for *not rejecting a claim* and
the formula for the *interval* are the same inequality, solved for different
things.

### Testing every possible claim at once {-}

That equivalence can be checked rather than taken on trust. Rather than testing
the single claim $\mu_0 = 165$, test every value on a fine grid and collect the
ones that survive.

```{r os-ci-equiv}
se  <- 8 / sqrt(36)
grid <- seq(160, 176, by = 0.001)

not_rejected <- grid[abs(xbar - grid) / se <= qnorm(0.975)]

surviving <- range(not_rejected)
interval  <- xbar + c(-1, 1) * qnorm(0.975) * se

rbind(claims_that_survive_the_test = surviving,
      confidence_interval          = interval,
      difference                   = surviving - interval)
```

The two agree to within a thousandth of a centimetre, which is the spacing of
the grid — test a finer grid and the discrepancy shrinks with it. The set of
claims a test would not reject *is* the confidence interval. Nothing was
assumed to make this happen; it follows from the two being one inequality
solved for different unknowns.

```{r os-ci-pvalue-plot, echo = FALSE, fig.height = 3.4, fig.cap = "The p-value for every possible claimed value of mu. Claims above the 0.05 line survive the test, and the range of surviving claims is exactly the 95% confidence interval."}
pv <- data.frame(mu0 = seq(162, 174, length.out = 800))
pv$p <- 2 * (1 - pnorm(abs(xbar - pv$mu0) / se))
lo <- xbar - qnorm(0.975) * se
hi <- xbar + qnorm(0.975) * se

ggplot(pv, aes(mu0, p)) +
  geom_ribbon(data = subset(pv, mu0 >= lo & mu0 <= hi),
              aes(ymin = 0.05, ymax = p), fill = book_palette$fill,
              alpha = 0.85) +
  geom_line(linewidth = 0.8, colour = book_palette$ink) +
  geom_hline(yintercept = 0.05, linetype = "dashed",
             colour = book_palette$reject) +
  annotate("segment", x = c(lo, hi), xend = c(lo, hi), y = 0, yend = 0.05,
           colour = book_palette$reject, linewidth = 0.5) +
  annotate("text", x = 165, y = 0.44, label = "claims that survive",
           size = 3.5, colour = "grey25") +
  annotate("text", x = 172.6, y = 0.075, label = "α = 0.05",
           size = 3.3, colour = book_palette$reject) +
  scale_x_continuous(breaks = c(164, round(lo, 1), 168, round(hi, 1), 174)) +
  labs(x = expression(paste("claimed value  ", mu[0])), y = "p-value") +
  theme_book()
```

The curve peaks at $\mu_0 = \bar{x} = 168$, where the claim matches the data
exactly and the $p$-value is 1. It falls away in both directions, crossing 0.05
precisely at the two endpoints of the interval.

::: {.keyidea}
A $(1-\alpha)$ confidence interval is exactly the set of values $\mu_0$ that a
two-tailed test at level $\alpha$ would **fail** to reject.

Test and interval are one object viewed from two sides. If the interval misses
the claimed value, the test rejects it. Always.
:::

### So why keep both? {-}

Because they carry different information, and the figure above shows what each
one throws away.

A confidence interval answers infinitely many tests at once, and reports the
answer in the units of the problem. Handed $[165.39,\ 170.61]$, a reader can
evaluate any claim about average height — 165, 166, 172 — without recomputing
anything. A test answers one question, chosen in advance.

A test, however, reports *how far* rather than merely *in or out*. Both 165 and
160 fall outside our interval, and the interval says the same thing about both:
rejected. The $p$-values do not agree — 0.024 for 165, and effectively zero for
160. One claim is marginally implausible and the other is not remotely
survivable, and the interval alone does not distinguish them.

::: {.intuition}
Report the interval. It contains the test, states its answer in the units the
reader cares about, and does not tempt anyone into treating a single threshold
as a verdict.

Report the $p$-value alongside it when one specific claim is genuinely at
issue, since that is where the interval is silent about degree.
:::

Everything Section \@ref(u3-misreadings) said about reading a confidence
interval applies unchanged here. In particular, the 95% still describes the
procedure and not this interval, and an interval that excludes a claimed value
has not proved the claim false — it has reported that the data are unlikely
under it.

## When σ is unknown: the t-test {#os-ttest}

Everything so far assumed we knew the population standard deviation $\sigma$.
We almost never do. Replacing it with the sample standard deviation $s$ adds a
second source of uncertainty, and the statistic

$$t_{calc} = \frac{\bar{x} - \mu_0}{s/\sqrt{n}}$$

no longer follows a normal distribution. It follows Student's $t$ with
$n - 1$ **degrees of freedom** — a distribution with fatter tails, which demands
more evidence before it will let you reject.

```{r os-t-vs-z, echo = FALSE, fig.cap = "Student's $t$ against the standard normal. With few degrees of freedom the tails are noticeably heavier, so critical values sit further out."}
grid <- expand.grid(x = seq(-4, 4, length.out = 400),
                    dist = c("t, df = 5", "t, df = 30", "normal"))
grid$y <- with(grid, ifelse(dist == "normal", dnorm(x),
                     ifelse(dist == "t, df = 5", dt(x, 5), dt(x, 30))))

ggplot(grid, aes(x, y, colour = dist, linewidth = dist)) +
  geom_line() +
  scale_colour_manual(values = c("t, df = 5" = book_palette$reject,
                                 "t, df = 30" = book_palette$accent,
                                 "normal" = book_palette$ink), name = NULL) +
  scale_linewidth_manual(values = c("t, df = 5" = 0.8, "t, df = 30" = 0.8,
                                    "normal" = 1.1), guide = "none") +
  labs(x = NULL, y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank(), legend.position = "top")
```

### Settling the question with the actual class {-}

We opened this chapter with a made-up sample of 36 students. We can now do the
real thing. `stats-class.csv` holds the survey this class filled in — names,
ages, heights, gender:

```{r os-class-data}
students <- read.csv("data/stats-class.csv")

str(students)
```

Thirty-three rows, but look at the last one. One student left age and height
blank, and another did not give a gender:

```{r os-class-na}
students[!complete.cases(students), ]
```

::: {.watchout}
**A blank is not always an `NA`.** In a numeric column an empty field is read
as `NA`, but in a text column it is read as the empty string `""` — and `""`
is a perfectly ordinary value as far as `is.na()`, `complete.cases()` and
`na.rm` are concerned. A gap in a text column can therefore hide in plain
sight.

If a dataset you did not prepare yourself has blanks, say so explicitly when
you read it:

```{r os-na-strings, eval = FALSE}
read.csv("somefile.csv", na.strings = c("", "NA"))
```

The files in this book are already written so that plain `read.csv()` does the
right thing.
:::

::: {.watchout}
`mean()` returns `NA` if even one value is missing — it refuses to guess what
you meant. Pass `na.rm = TRUE` to drop the gaps, and always know how many you
dropped.

```{r os-class-narm}
mean(students$height)                  # NA — one height is missing
mean(students$height, na.rm = TRUE)    # 32 students
```
:::

Now the test. We do **not** know the population standard deviation of student
heights, so this is a $t$-test, not a $Z$-test:

```{r os-class-test}
heights_real <- students$height[!is.na(students$height)]

c(n = length(heights_real),
  mean = mean(heights_real),
  sd = sd(heights_real))

t.test(heights_real, mu = 165)
```

```{r os-class-stash, include = FALSE}
ct <- t.test(heights_real, mu = 165)
```

$t = `r round(ct$statistic, 2)`$ on `r ct$parameter` degrees of freedom, with a
$p$-value of `r round(ct$p.value, 4)`. We **reject** the claim that the average
is 165 cm. The 95% interval,
$[`r round(ct$conf.int[1], 1)`,\ `r round(ct$conf.int[2], 1)`]$, excludes 165 —
the same verdict, as always.

::: {.intuition}
Notice how much narrower this is than the textbook version of the problem. The
opening example assumed we somehow knew $\sigma = 8$. Here we had to estimate
it from 32 people, and the $t$ distribution charges us for that uncertainty
with a wider interval and a higher bar to clear.

Also notice what the data is not: 32 students in one classroom are not a random
sample of all APU students. The arithmetic is correct; whether it answers the
question about *the university* is a separate matter, and a harder one.
:::

### Worked example {-}

The dean estimates that full-time faculty teach 11.0 classroom hours per week.
You sample eight faculty members. Can you reject the dean's claim at
$\alpha = 0.10$? At $\alpha = 0.01$?

```{r os-t-data}
hours <- c(11.8, 8.6, 10.5, 7.9, 6.4, 10.4, 10.0, 8.2)

c(n = length(hours), mean = mean(hours), sd = sd(hours))
```

**On paper.** With $\bar{x} = 9.225$, $s = 1.749$, $n = 8$ and $df = 7$:

$$
t_{calc} = \frac{9.225 - 11}{1.749/\sqrt{8}}
= \frac{-1.775}{0.618}
= -2.87
$$

The critical values are $t_{0.05,\,7} = 1.895$ and $t_{0.005,\,7} = 3.499$.

- At $\alpha = 0.10$: $2.87 > 1.895$ $\Rightarrow$ **reject** $H_0$.
- At $\alpha = 0.01$: $2.87 < 3.499$ $\Rightarrow$ **fail to reject** $H_0$.

**In R.** Both the pieces and the packaged version:

```{r os-t-manual}
t_calc <- (mean(hours) - 11) / (sd(hours) / sqrt(length(hours)))
t_crit <- qt(c(0.95, 0.995), df = 7)     # 10% and 1%, two-tailed

c(t_calc = t_calc, crit_10pct = t_crit[1], crit_1pct = t_crit[2])
```

```{r os-t-test}
t.test(hours, mu = 11, conf.level = 0.90)
```

The $p$-value is `r round(t.test(hours, mu = 11)$p.value, 4)` — below 0.10 but
above 0.01, exactly matching the two different verdicts we reached by hand.

::: {.intuition}
The same data can be "significant" at one level and not at another. Nothing has
changed about the world between those two lines — only how much risk of a false
alarm we were willing to tolerate. This is why reporting the $p$-value itself
is far more informative than reporting a bare "significant / not significant."
:::

## One-tailed tests {#os-onetailed}

Sometimes only one direction matters. A firm claims its new curing technique
gets nicotine content *below* 1.5 mg; we only care if the true mean is lower.

$$H_0 : \mu \geq 1.5 \qquad H_A : \mu < 1.5$$

All the probability we are willing to risk now goes into a single tail, so the
critical value moves *inward*: $-Z_{0.05} = -1.645$ rather than $\pm 1.96$.

```{r os-onetail}
qnorm(0.05)             # left-tailed critical value at 5%
qnorm(0.95)             # right-tailed critical value at 5%
```

In R, add the `alternative` argument:

```{r os-onetail-test}
nicotine <- make_sample(n = 20, mean = 1.42, sd = 0.7)

z.test(nicotine, mu = 1.5, sigma.x = 0.7, alternative = "less",
       conf.level = 0.95)
```

$Z_{calc} = -0.51$, which is not below $-1.645$, so we fail to reject. The
firm's claim is not supported — a sample mean of 1.42 would arise about 30% of
the time even if the true mean were 1.5.

### The one-sided interval looks backwards {-}

This trips up nearly everyone. For the left-tailed test above, the confidence
interval is

$$\left(-\infty,\ \bar{x} + Z_\alpha \frac{\sigma}{\sqrt{n}}\right)$$

An *upper* bound — for a *left*-tailed test. Why?

::: {.theory}
Start from the rejection rule and solve for $\mu_0$ rather than for $\bar{x}$:

$$
\begin{aligned}
\frac{\bar{x} - \mu_0}{\sigma/\sqrt{n}} &< -Z_\alpha \\[4pt]
\bar{x} - \mu_0 &< -Z_\alpha \frac{\sigma}{\sqrt{n}} \\[4pt]
\mu_0 &> \bar{x} + Z_\alpha \frac{\sigma}{\sqrt{n}}
\end{aligned}
$$

The inequality flipped when $\mu_0$ moved across. So we reject exactly when
$\mu_0$ is **too large**, and the values we *fail* to reject are those with
$\mu \le \bar{x} + Z_\alpha\,\sigma/\sqrt{n}$.
:::

::: {.intuition}
The rejection region is defined in **$Z$-space**; the confidence interval lives
in **parameter space**. Solving the inequality for $\mu$ reverses the direction.

Put plainly: a left-tailed test rejects when the sample mean comes out too
small. So the values of $\mu$ it rules out are the ones that are too *large*
relative to what we saw. That leaves an upper bound.
:::

## Reference tables {#os-reference}

Table: (\#tab:os-ref-z) One-sample tests at a glance. Use $Z$ and `z.test()` when $\sigma$ is known; use $t$ with $df = n-1$ and `t.test()` when it is not.

| Test | Hypotheses | Reject when | R argument |
|:--|:--|:--|:--|
| Two-tailed | $H_0: \mu = \mu_0$ vs $H_A: \mu \neq \mu_0$ | $\lvert Z_{calc} \rvert > Z_{\alpha/2}$ | (default) |
| Left-tailed | $H_0: \mu \geq \mu_0$ vs $H_A: \mu < \mu_0$ | $Z_{calc} < -Z_{\alpha}$ | `alternative = "less"` |
| Right-tailed | $H_0: \mu \leq \mu_0$ vs $H_A: \mu > \mu_0$ | $Z_{calc} > Z_{\alpha}$ | `alternative = "greater"` |

Table: (\#tab:os-ref-crit) Standard normal critical values. Every figure below is `qnorm()` evaluated inline, not copied from a printed table.

| $\alpha$ | Two-tailed, $Z_{\alpha/2}$ | One-tailed, $Z_{\alpha}$ |
|:--|:--|:--|
| 10% | `r round(qnorm(0.95), 3)` | `r round(qnorm(0.90), 3)` |
| 5%  | `r round(qnorm(0.975), 3)` | `r round(qnorm(0.95), 3)` |
| 1%  | `r round(qnorm(0.995), 3)` | `r round(qnorm(0.99), 3)` |

## Summary {#os-summary}

::: {.keyidea}
1. A gap between a sample and a claim proves nothing on its own — sampling
   noise guarantees some gap. The test asks whether the gap is *bigger than
   noise*.
2. $\displaystyle Z_{calc} = \frac{\bar{x}-\mu_0}{\sigma/\sqrt{n}}$ when
   $\sigma$ is known; $\displaystyle t_{calc} = \frac{\bar{x}-\mu_0}{s/\sqrt{n}}$
   with $df = n-1$ when it is not.
3. Reject when the statistic beats the critical value, or equivalently when the
   $p$-value falls below $\alpha$. The two rules never disagree.
4. A confidence interval is the set of null values a test would *not* reject.
   Interval and test are the same statement.
5. A $p$-value is $P(\text{data} \mid H_0)$, never $P(H_0 \mid \text{data})$.
6. The equals sign always lives in $H_0$. "Can you reject the claim?" puts the
   claim in $H_0$; "is there evidence supporting the claim?" puts it in $H_A$.
7. `qnorm()` and `qt()` replace every printed table you own.
:::



## Practice problems {#os-exercises}

::: {.yourturn}
Work each one on paper first, then check yourself in R. The solutions are one
click away, which makes them worth nothing unless you try first.

These are for learning, so the workings are shown in full. The
[question bank](#os-question-bank) that follows is for testing yourself, and
has no answers.
:::

**11.1** A consumer research organisation states that the mean caffeine content
per 12-ounce bottle of a caffeinated soft drink is 37.7 mg. A random sample of
36 bottles has a mean of 35.28 mg. The population standard deviation is 10.8 mg.
At $\alpha = 0.01$, can you reject the organisation's claim?

::: {.answer}
$H_0: \mu = 37.7$ against $H_A: \mu \neq 37.7$. Two-tailed at $\alpha = 0.01$,
so the critical value is $Z_{0.005} = 2.58$.

$$
Z_{calc} = \frac{35.28 - 37.7}{10.8/\sqrt{36}} = \frac{-2.42}{1.8} = -1.34
$$

```{r os-ex1}
caffeine <- make_sample(n = 36, mean = 35.28, sd = 10.8)
z.test(caffeine, mu = 37.7, sigma.x = 10.8, conf.level = 0.99)
```

$|-1.34| < 2.58$ and the $p$-value of 0.179 is far above 0.01, so we **fail to
reject**. If the true mean really were 37.7 mg, a sample mean this far away
would turn up about 18% of the time — thoroughly unremarkable.
:::

**11.2** Historical data show household water use is normally distributed with
mean 360 gallons and standard deviation 40 gallons per day. A sample of 200
households averages 374 gallons. Are these data consistent with the historical
distribution at the 5% level? What is the $p$-value?

::: {.answer}
$$
Z_{calc} = \frac{374 - 360}{40/\sqrt{200}} = \frac{14}{2.828} = 4.95
$$

```{r os-ex2}
water <- make_sample(n = 200, mean = 374, sd = 40)
z.test(water, mu = 360, sigma.x = 40, conf.level = 0.95)
```

$4.95 \gg 1.96$, and the $p$-value is about
`r signif(2 * pnorm(-4.95), 2)` — under one in a million. **Reject $H_0$.**
Water use has moved away from the historical mean.
:::

**11.3** A study claims the average monthly salary of full professors in India
is ₹87,800. Students at one university sample ten salaries (in thousands):

```{r os-ex3-data, echo = FALSE, comment = NA}
cat("91.0, 79.8, 102.0, 93.5, 82.0, 88.6, 90.0, 98.6, 101.0, 84.0")
```

They believe their professors earn *more*. Test at the 10% and 5% levels.

::: {.answer}
The word "more" makes this **right-tailed**:
$H_0: \mu \leq 87.8$ against $H_A: \mu > 87.8$. Since $\sigma$ is unknown and
$n = 10$, use a $t$-test with $df = 9$.

```{r os-ex3}
salary <- c(91.0, 79.8, 102.0, 93.5, 82.0, 88.6, 90.0, 98.6, 101.0, 84.0)

c(mean = mean(salary), sd = sd(salary))
t.test(salary, mu = 87.8, alternative = "greater")
```

$t_{calc} = `r round((mean(c(91.0,79.8,102.0,93.5,82.0,88.6,90.0,98.6,101.0,84.0)) - 87.8) / (sd(c(91.0,79.8,102.0,93.5,82.0,88.6,90.0,98.6,101.0,84.0))/sqrt(10)), 3)`$,
against critical values $t_{0.10,\,9} = `r round(qt(0.90, 9), 3)`$ and
$t_{0.05,\,9} = `r round(qt(0.95, 9), 3)`$.

The statistic falls short of both. **Fail to reject** at either level — there is
not enough evidence that these professors are paid above the national figure.
:::

**11.4** A Department of Transportation claims mean wait time is *at most* 6
minutes. A sample of 34 locations has mean 10.3 minutes, standard deviation 8.0.
Test at 5% and 1%.

::: {.answer}
"At most" makes this right-tailed: $H_0: \mu \leq 6$ against $H_A: \mu > 6$,
$t$-test with $df = 33$.

```{r os-ex4}
waits <- make_sample(n = 34, mean = 10.3, sd = 8.0)
t.test(waits, mu = 6, alternative = "greater")
```

$t_{calc} = 3.14$, well beyond $t_{0.05,\,33} = `r round(qt(0.95, 33), 3)`$ and
$t_{0.01,\,33} = `r round(qt(0.99, 33), 3)`$. **Reject at both levels** — wait
times are longer than claimed.
:::

**11.5** `marriage-ages.csv` records the ages of both partners in ten couples.
Is there evidence that one partner tends to be older? Test at the 5% level.

*Hint: the two columns are not independent samples — they are ten couples. What
is the one number per couple that captures the question?*

::: {.answer}
The two columns are **paired**: each row is one couple. Treating them as two
independent samples throws away that pairing and inflates the standard error.
The right move is to reduce each couple to a single number — the age gap — and
run a **one-sample** test on it against zero.

```{r os-ex5}
couples <- read.csv("data/marriage-ages.csv")

gap <- couples$partner1 - couples$partner2
gap

c(mean = mean(gap), sd = sd(gap), n = length(gap))
```

```{r os-ex5-test}
t.test(gap, mu = 0)
```

The mean gap is 2.5 years, $t = 3.0$ on 9 degrees of freedom, $p = 0.015$.
Since $0.015 < 0.05$, we **reject** $H_0$: partner 1 is reliably older.

R will do the pairing for you if you ask:

```{r os-ex5-paired}
t.test(couples$partner1, couples$partner2, paired = TRUE)
```

Identical output — `paired = TRUE` computes exactly the differences we formed
by hand. Compare what happens if you forget:

```{r os-ex5-wrong}
t.test(couples$partner1, couples$partner2, paired = FALSE)$p.value
```

```{r os-ex5-stash, include = FALSE}
p_unpaired <- t.test(couples$partner1, couples$partner2, paired = FALSE)$p.value
p_paired   <- t.test(couples$partner1, couples$partner2, paired = TRUE)$p.value
```

The same ten couples, a $p$-value of `r round(p_unpaired, 3)` instead of
`r round(p_paired, 3)`, and the opposite conclusion. **Pairing is not a
detail.**
:::

**11.6** *(No computation.)* A classmate reports "$p = 0.03$, so there is a 3%
chance the null hypothesis is true." Explain what is wrong, and state what
$p = 0.03$ does mean.

::: {.answer}
The $p$-value is computed **assuming the null is true**. It cannot then also be
the probability that the null is true — that would be circular. Formally, your
classmate has swapped $P(\text{data} \mid H_0)$ for $P(H_0 \mid \text{data})$,
which are different quantities entirely.

The correct reading: *if* $H_0$ were true, samples at least as extreme as this
one would arise 3% of the time. Since that is unusual, we treat it as evidence
against $H_0$ — but the statement is about how surprising the data are, not
about how probable the hypothesis is.
:::

## Question bank {#os-question-bank}

::: {.watchout}
**No solutions are given for this section, by design.** These questions are
drawn from past quizzes and exams. If you can work them without checking an
answer key, you are ready; if you cannot, the section to reread is named
beside each one.

They run from easiest to hardest. Do them in order.
:::

### Multiple choice {-}

**QB 11.1** The level of significance, $\alpha$, is the risk of *(select all
that apply)*

a. Rejecting the null hypothesis when it is true
b. Rejecting the null hypothesis when the alternative hypothesis is true
c. Rejecting the alternative hypothesis when the null hypothesis is true
d. Rejecting the alternative hypothesis when it is true

*See Section \@ref(os-critical).*

**QB 11.2** An economist sets the alternative hypothesis that the average CGPA
of undergraduate students is less than 8. What type of test should be used, and
what is the critical value at $\alpha = 5\%$?

a. One-tailed test and $+1.96$
b. One-tailed test and $-1.96$
c. One-tailed test and $+1.645$
d. One-tailed test and $-1.645$

*See Section \@ref(os-onetailed).*

**QB 11.3** The margin of error for a confidence interval is
$\pm Z_{\alpha/2}\,\sigma/\sqrt{n}$. Which statements are correct? *(select all
that apply)*

a. Increasing the sample size will decrease the margin of error
b. The margin of error decreases as $\alpha$ rises
c. The larger the margin of error, the less confident one is about the estimate
d. The margin of error increases with the population standard deviation $\sigma$

*See Section \@ref(os-ci).*

**QB 11.4** A 95% confidence interval for mean student height is
$(165.41,\ 170.95)$. You now compute a 99% interval from the same data. Which
statements are likely correct? *(select all that apply)*

a. The 90%, 95% and 99% intervals are the same
b. The 99% interval is $(164.54,\ 171.83)$
c. The 99% interval is $(165.86,\ 170.51)$
d. The answer cannot be determined from the information given

*See Section \@ref(os-ci). Ask yourself which way an interval must move when you demand more confidence.*

**QB 11.5** A company tests the null hypothesis that the average time to accept
a job offer is at most 7 days. The test concludes that the average exceeds 7
days. It is later learned that the true mean is exactly 7 days. What error
occurred?

a. Type II error
b. Type I error
c. A correct decision was made
d. Cannot be determined from the information provided

### Short reasoning {-}

**QB 11.6** "A high $p$-value is evidence that the null hypothesis is true."
Do you agree? Why or why not?

*See Section \@ref(os-pvalue).*

**QB 11.7** Two economists, Amit and Anand, collect data on Indian agricultural
wages to test the effect of MNREGA. Amit concludes that wages after the policy
are higher than before, when in fact wages have not changed. Anand concludes
that wages before and after are the same — also incorrectly.

It is claimed that Amit has committed a Type I error and Anand a Type II error.
Do you agree? Why or why not?

### Numerical {-}

**QB 11.8** A researcher wants to estimate the average weekly study hours of
university students. The population standard deviation is 6 hours. What sample
size is needed to estimate the mean with a margin of error of 1 hour at 90%
confidence?

a. $n = 98$
b. $n = 64$
c. $n = 121$
d. $n = 82$

**QB 11.9** The numbers of people taking the morning university shuttle on 16
randomly chosen days are:

```{r qb-shuttle, echo = FALSE, comment = NA}
cat("27, 46, 35, 33, 29, 45, 28, 24, 30, 40, 38, 35, 41, 27, 38, 29")
```

The administration claims the daily average is 30. Taking $\sigma = 7$:

a. Construct a 99% confidence interval for the mean number of riders.
b. Use that interval to test the administration's claim.
c. Would your conclusion change at the 95% level? Explain *why* before you
   compute it.

*Part (c) is the one that separates understanding from arithmetic.*


## Looking ahead {#os-looking-ahead}

Every test in this unit compared a sample against a *claimed* number — a
figure from a government report, a manufacturer's specification, a previous
year's estimate. That number had to come from somewhere outside the data.

Most questions worth asking do not arrive in that form.

Did the households that received the transfer end up better off than those
that did not? Do girls in the scheme attend school more than girls outside it?
Did the districts that got the road see wages rise more than the districts
that did not? In each case there is no external claim to test against. There
are two groups, each with its own sample mean, and the question is whether the
difference between them is larger than sampling variation alone would produce.

That is the subject of the next unit. It is also where the difficulty changes
character. Deciding whether two groups differ is a statistical problem, and we
will solve it. Deciding whether one group differs *because* of the treatment is
a question about how the groups came to be formed — and no test statistic can
answer it.
