# Hypothesis Testing: One Sample {#one-sample}

```{r setup-11, include = FALSE}
source("_common.R")
library(ggplot2)
```

In the previous unit we learned that every estimate comes with uncertainty. A
confidence interval tells us which population values are plausible given our
sample.

But many real statistical questions begin with a specific claim. Has average
monsoon rainfall changed? Is the unemployment rate below 5%? Does a new drug
lower blood pressure?

Rather than asking *which values are plausible*, these questions ask *is this
particular claim consistent with the data?* Answering that is the purpose of
hypothesis testing.

Consider a claim made about a population. You collect a sample. Your estimate
is not exactly what the claim predicted.

Should you conclude that the claim is wrong?

Or is this simply the kind of difference that appears because every random
sample is slightly different?

That question is the starting point for hypothesis testing, and it is the
engine behind much of modern statistics.

## The problem, before any formulas {#os-problem}

The APU administration claims the average height of all students is 165 cm.

You are sceptical. You could settle it by measuring all 3,500 students, but that
costs weeks of work. So instead you measure 36 students chosen at random, and
you find their average height is **168 cm**.

The claim said 165. Your sample says 168. Is the administration wrong?

::: {.watchout}
The tempting answer is "yes, obviously — 168 is not 165." That answer is wrong,
and understanding *why* it is wrong is the whole chapter.

You did not measure everyone. You measured 36 people. Even if the true average
really were exactly 165, your particular 36 students would almost never average
exactly 165. Some samples come out high, some low. **A gap between the sample
and the claim is guaranteed to exist even when the claim is perfectly true.**
:::

Let us actually watch that happen.

We obviously do not know whether the administration is correct. But imagine,
just for a moment, that they are. If the true average really were 165 cm, with
a standard deviation of 8 cm, what kinds of sample means would we expect to
see? Draw a sample of 36 students, ten separate times, and look.

```{r os-sim-ten}
set.seed(42)

ten_samples <- replicate(10, mean(rnorm(n = 36, mean = 165, sd = 8)))
round(ten_samples, 2)
```

Every one of those numbers came from a population whose mean is *exactly* 165.
Not one of them is 165. Several are further from 165 than a full centimetre.

So the real question is not "is 168 different from 165?" It obviously is. The
real question is:

::: {.intuition}
**If the administration's claim were true, how often would a sample of 36
students come out as far from 165 as ours did?**

If the answer is "quite often" — then our sample is unremarkable, and we have no
grounds to call anyone a liar.

If the answer is "almost never" — then either something very unusual happened
to us, or the administration's claim is not correct. In statistics we usually
regard the second explanation as the more convincing one.
:::

We can answer that question by brute force before we do any algebra at all.
Let's draw not ten samples but fifty thousand, still assuming the claim is true,
and count how many stray at least 3 cm away from 165 (as ours did).

```{r os-sim-many}
set.seed(42)

many_means <- replicate(50000, mean(rnorm(n = 36, mean = 165, sd = 8)))

# How far did our real sample fall from the claim?
observed_gap <- abs(168 - 165)

mean(abs(many_means - 165) >= observed_gap)
```

About `r round(100 * mean(abs(many_means - 165) >= observed_gap), 1)`% of the
time. So if the administration is telling the truth, a sample like ours turns
up roughly once in forty tries.

We have now answered the question using simulation alone, and without naming a
single technical term. If the claim were true, a sample mean at least 3 cm away
from 165 would appear about once in forty surveys.

That makes our sample unusual. But how unusual is unusual *enough*? Once in
forty is rare; would once in ten have been rare enough? Once in a hundred? The
rest of this chapter builds a systematic way of answering exactly that — and by
the end of it you will be able to obtain the number above exactly, in one line,
with no simulation at all.

```{r os-sim-plot, echo = FALSE, fig.cap = "Fifty thousand sample means, each from 36 students, drawn from a population whose true mean is 165 cm (dashed line). The solid line marks our observed sample mean of 168 cm. Only about 2.4% of the simulated samples fall at least this far from the claimed value in either direction, which is what makes our sample surprising if the claim is true."}
df <- data.frame(m = many_means)

ggplot(df, aes(m)) +
  geom_histogram(bins = 60, fill = book_palette$fill, colour = "white",
                 linewidth = 0.2) +
  geom_vline(xintercept = 165, colour = book_palette$ink,
             linetype = "dashed", linewidth = 0.7) +
  geom_vline(xintercept = 168, colour = book_palette$reject, linewidth = 1) +
  annotate("label", x = 168, y = Inf, vjust = 1.4, label = "our sample: 168",
           colour = book_palette$reject, fill = "white", size = 3.6,
           label.size = 0.3) +
  annotate("label", x = 165, y = Inf, vjust = 3.4, label = "the claim: 165",
           colour = book_palette$ink, fill = "white", size = 3.6,
           label.size = 0.3) +
  labs(x = "sample mean height (cm)", y = "how often") +
  theme_book()
```

## Naming the two hypotheses {#os-hypotheses}

::: {.question}
We have the reasoning. What are the two positions we are reasoning between?
:::

The previous section did the whole of a hypothesis test without naming
anything. We assumed the administration's claim, asked how often a sample would
stray as far from it as ours did, and found the answer was about once in forty.

Let us give the two positions in that argument their names.

We start from a sample $x_1, x_2, \ldots, x_n$ drawn from a population with
unknown mean $\mu$ and standard deviation $\sigma$. The claim under scrutiny is
the **null hypothesis**:

$$H_0 : \mu = 165$$

The position we would move to if the claim collapses is the **alternative
hypothesis**:

$$H_A : \mu \neq 165$$

### Why begin by assuming the claim is true? {-}

At first this looks backwards. If we suspect the claim is wrong, why start by
supposing it is right?

Because the assumption is what gives us something to measure against. Notice
what we actually did in the previous section: we could only simulate those
fifty thousand samples *because* we had a specific value of $\mu$ to simulate
from. Had we begun with "the mean is not 165," there would have been nothing to
draw samples from — not 165 covers 164, and 170, and 3,000, and each would give
a different answer.

So we grant the claim, temporarily, and see where it leads. If our sample would
be quite ordinary under it, we have no reason to doubt it. If our sample would
be extraordinarily unlikely under it, then the claim itself becomes hard to
believe.

::: {.keyidea}
The null hypothesis is always the *boring* option — nothing is going on, the
claim stands, the difference is zero. We never *prove* it. We only ever fail to
knock it down.

This is why we say "**fail to reject** $H_0$" and never "accept $H_0$." A court
that acquits has not proven innocence; it has failed to prove guilt.
:::

### Which hypothesis does the claim go in? {#os-which-hypothesis}

::: {.question}
A claim arrives. Does it become $H_0$ or $H_A$?
:::

This is one of the most common places for a beginner to go wrong, and the way
out is not to memorise a rule but to ask a question first.

Before writing anything down, ask yourself:

> What result would convince me?

Suppose a report claims that the mean CAT score of applicants is *more than*
499. What would persuade you? Almost everyone answers the same way: a sample
mean comfortably **above** 499. A sample averaging 470 would not support the
report, and neither would one averaging exactly 499.

That answer has already determined the test. The thing you would find
convincing — a mean above 499 — is the thing you are gathering evidence *for*,
and a test only ever produces evidence *against* $H_0$. So the claim has to sit
in the alternative:

$$H_0 : \mu \leq 499 \qquad H_A : \mu > 499$$

The direction follows too. You would be convinced by large values, so it is
large values that must count against the null, and the test looks in the upper
tail.

::: {.keyidea}
Two questions settle every case.

**What am I trying to gather evidence for?** That goes in $H_A$. A test can
only accumulate evidence against the null, so whatever you hope to establish
must be what survives when the null falls.

**Which direction would count as evidence?** That fixes the inequality in
$H_A$, and with it the tail the test looks in.
:::

Examination questions usually signal this in their phrasing, and the signal is
worth recognising.

Table: (\#tab:os-claim-table) How the usual phrasings map onto the two questions above.

| The question asks | The claim becomes | Because |
|:--|:--|:--|
| "Is there evidence to **support** the claim?" | $H_A$ | The claim is what you are gathering evidence for. |
| "Can you **reject** the claim?" | $H_0$ | The claim is what you are challenging, and you are asking whether the data can knock it down. |

### Where the equality goes {-}

One structural rule holds in every case, and it will save you from a good deal
of trouble:

::: {.keyidea}
**The equality always belongs in the null hypothesis.**

$H_0$ takes $=$, $\leq$ or $\geq$. $H_A$ takes $\neq$, $<$ or $>$, and never
equality.
:::

The reason is the one from earlier in this section. Every test begins by
assuming one specific value of $\mu$ and computing how likely our sample would
be under it. Only a hypothesis containing an equality supplies such a value —
$\mu \leq 499$ can be tested at its boundary, 499, whereas $\mu > 499$ names no
particular number to compute with.

::: {.watchout}
A claim phrased with a *strict* inequality — "the mean is more than 499" —
therefore cannot go into $H_0$ as stated. If you are asked for evidence
supporting it, the pair is

$$H_0 : \mu \leq 499 \qquad H_A : \mu > 499$$

Writing $H_0: \mu \geq 499$ is the common error. It puts the claim in the null
*and* points the test in the wrong direction, so a sample mean **above** 499 —
evidence *for* the claim — ends up producing a $p$-value near 1 and a "fail to
reject." The arithmetic is fine; the conclusion is backwards.
:::

### The CAT report, worked through {-}

A report claims the mean CAT score of applicants is more than 499. A sample of
100 applicants has mean 502, with $\sigma = 10.6$. Is there enough evidence at
$\alpha = 0.01$ to support the report?

We settled the hypotheses above by asking what would convince us:

$$H_0 : \mu \leq 499 \qquad H_A : \mu > 499$$

Right-tailed, so the critical value is $Z_{0.01} = 2.33$ — positive, not
negative.

Before computing anything, check the direction. The sample mean is 502 and the
claim is "more than 499." The sample points the *same way* as the claim, so
whatever the arithmetic returns, it must come out favourable to the report. If
it does not, the tail has been pointed the wrong way.

```{r os-claim-worked}
xbar_c <- 502; mu_c <- 499; sigma_c <- 10.6; n_c <- 100

z_c <- (xbar_c - mu_c) / (sigma_c / sqrt(n_c))
c(z_calc = z_c,
  z_crit = qnorm(0.99),
  p_value = pnorm(z_c, lower.tail = FALSE))   # upper tail: H_A is ">"
```

$Z_{calc} = 2.83 > 2.33$, and $p = 0.0023 < 0.01$. We **reject $H_0$**, which
means the data **support** the report's claim — and the answer points the way
the sanity check said it would.

::: {.intuition}
Make that check a habit, and make it before you trust any $p$-value rather than
after.

Ask which way the sample points relative to the claim, and decide what answer
you should therefore expect. If the arithmetic hands back something else — a
$p$-value of 0.998 and "no support" when the sample sits above the claimed
value — the mistake is almost always in $H_A$, not in the calculation.
:::


## What a p-value actually says {#os-pvalue}

::: {.question}
We already computed a probability by simulation. What exactly was it?
:::

Back in Section \@ref(os-problem) we simulated fifty thousand samples from a
population whose mean really was 165 cm. Only about
`r round(100 * mean(abs(many_means - 165) >= observed_gap), 1)`% of those
samples produced a sample mean at least as far from 165 as the one we actually
observed.

Without realising it, we had already computed a **$p$-value**.

::: {.keyidea}
A **$p$-value** is the probability of obtaining a sample at least as extreme as
the one we observed, *assuming the null hypothesis is true*.
:::

Notice the order carefully.

1. We begin by assuming the null hypothesis is true.
2. We ask what kinds of samples we would expect under that assumption.
3. We compare our observed sample with those possibilities.
4. The $p$-value is the proportion of those hypothetical samples at least as
   extreme as ours.

In the height example the $p$-value was about
`r round(mean(abs(many_means - 165) >= observed_gap), 3)`. If the true
population mean really were 165 cm, a sample as unusual as ours would occur
only about
`r round(100 * mean(abs(many_means - 165) >= observed_gap), 1)`% of the time.
That makes the observed sample unusual under the null hypothesis, which is why
we would reject it at the 5% significance level.

The $p$-value does **not** tell us whether the null hypothesis is true. It
tells us how surprising our data would be **if** the null hypothesis were true.
This distinction is one of the most important ideas in statistics.

A small $p$-value means the observed data would be difficult to explain if the
null hypothesis were correct. That is evidence *against* the null.

A large $p$-value means the observed data are quite ordinary under the null
hypothesis, so we have no convincing reason to reject it. Notice what that does
*not* mean: it does not prove the null hypothesis true. It means only that our
data are compatible with it.

Table: (\#tab:os-pval-table) Reading a $p$-value.

| $p$-value | Read it as | At $\alpha = 0.05$ |
|:--|:--|:--|
| 0.60 | A result like this is very common if the null hypothesis is true. | Fail to reject $H_0$ |
| 0.18 | This would occur roughly once every six samples. Not unusual. | Fail to reject $H_0$ |
| 0.049 | This would occur about once every twenty samples. Unusual enough to reject at the 5% level. | Reject $H_0$ |
| 0.0001 | This would occur about once every ten thousand samples. Extremely unusual under the null. | Reject $H_0$ |

### Four ways to read it backwards {-}

$p$-values are misquoted more often than any other number in empirical work,
and the errors are predictable enough to be worth naming.

::: {.watchout}
**"The $p$-value is the probability that the null hypothesis is true."** False.
The null hypothesis is *assumed* true in order to compute the $p$-value at all.
A quantity computed under an assumption cannot also be the probability that the
assumption holds.

**"The $p$-value is the probability that we have made the wrong decision."**
False. It measures how unusual the observed data are under the null, not the
probability that our conclusion is incorrect.

**"A large $p$-value proves the null hypothesis."** False. It means the data do
not provide strong evidence against it, which is a much weaker statement.

**"A very small $p$-value means the effect is important."** False. Statistical
significance and practical importance are different ideas. A tiny effect
produces a very small $p$-value if the sample is large enough.
:::

A useful way to hold the idea:

> A $p$-value measures how surprising the data are, not how believable the null
> hypothesis is.


## How far is too far? {#os-critical}

::: {.question}
A gap between the sample and the claim is guaranteed. How large a gap is large
enough to act on?
:::

By now we know the basic logic. We begin by assuming the null hypothesis is
true and ask: if that assumption were correct, how unusual would our sample be?

Section \@ref(os-problem) answered that by simulation. Sampling repeatedly from
a population whose mean really was 165 cm, a sample mean at least 3 cm away
turned up about
`r round(100 * mean(abs(many_means - 165) >= observed_gap), 1)`% of the time.

That raises a natural question. **How unusual is unusual enough?**

We need a rule telling us when the difference between the sample and the null
is large enough to reject the claim.

Write $\mu_0$ for the value the null claims, so that $H_0: \mu = \mu_0$. If the
sample mean lies close to $\mu_0$, we treat the difference as ordinary sampling
variation and keep the null. If it lies far enough away, we reject. That is, we
reject whenever

$$|\bar{x} - \mu_0| \geq c$$

for some cutoff $c$ we have yet to determine. The entire problem has reduced to
a single question: **how should we choose $c$?**

### Choosing the cutoff {-}

We choose it *before* collecting any data, by deciding how much risk we are
willing to run of rejecting a null hypothesis that is in fact true.

::: {.keyidea}
That risk is the **significance level**, written $\alpha$. It is the
probability of rejecting a true null hypothesis, and it is chosen in advance.
Common choices are 0.05 and 0.01.
:::

So $c$ is chosen so that, if the null really is true, only a proportion
$\alpha$ of all possible samples would fall beyond it:

$$P\big[\,|\bar{X} - \mu_0| \geq c\,\big] = \alpha$$

Everything needed to solve that equation is already in hand from the previous
units.

::: {.theory}
When the null is true, the Central Limit Theorem tells us the sampling
distribution of the sample mean is approximately normal, centred on $\mu_0$
with standard deviation $\sigma/\sqrt{n}$:

$$\bar{X} \;\sim\; N\!\left(\mu_0,\; \frac{\sigma^2}{n}\right)$$

Following the usual convention, the second argument is the *variance*, so the
standard deviation is its square root, $\sigma/\sqrt{n}$ — the standard error.

Standardising, as in Section \@ref(u3-ci), removes the units and the sample
size together:

$$Z = \frac{\bar{X} - \mu_0}{\sigma/\sqrt{n}} \;\sim\; N(0, 1)$$

Dividing both sides of the probability statement by the standard error
therefore gives

$$P\left[\,|Z| \geq \frac{c}{\sigma/\sqrt{n}}\,\right] = \alpha$$

The standard normal is symmetric, so that probability is split equally between
the two tails:

$$P\left[\,Z \geq \frac{c}{\sigma/\sqrt{n}}\,\right] = \frac{\alpha}{2}$$

The point leaving an upper-tail probability of $\alpha/2$ is the **critical
value**, written $Z_{\alpha/2}$. Since $P[Z \geq Z_{\alpha/2}] = \alpha/2$ by
definition, the bracketed quantity *is* $Z_{\alpha/2}$:

$$\frac{c}{\sigma/\sqrt{n}} = Z_{\alpha/2}
\qquad\Longrightarrow\qquad
c = Z_{\alpha/2}\,\frac{\sigma}{\sqrt{n}}$$
:::

Substituting that cutoff back into the rejection rule, and dividing both sides
by the standard error, produces the form used in practice:

$$\boxed{\;\frac{|\bar{x} - \mu_0|}{\sigma/\sqrt{n}} \;\geq\; Z_{\alpha/2}\;}$$

The quantity on the left is the **test statistic**, usually written
$Z_{calc}$. The quantity on the right is the critical value.

### The decision rule {-}

1. Compute the test statistic.
2. Compare it with the critical value.
3. If the statistic is larger, reject the null hypothesis.

Notice what has happened. The complicated question

> Is my sample unusually far from the null hypothesis?

has become the much simpler question

> Is my test statistic larger than the critical value?

That substitution is the whole point of standardising. Every one-sample problem
about means — whatever the units, whatever the sample size — now arrives at the
same comparison against the same handful of numbers.

::: {.inr}
Those familiar numbers — 1.645, 1.96 and 2.576 — are simply critical values of
the standard normal distribution. There is no need to consult a printed table;
`qnorm()` returns them to as many decimals as you like.

```{r os-crit-z}
alpha <- c(0.10, 0.05, 0.01)

# Two-tailed: we want the value leaving alpha/2 in the upper tail
qnorm(1 - alpha / 2)
```

These are exactly the values defining the rejection regions for the most
commonly used significance levels. They are not arbitrary quantities to
memorise. They are the points leaving 10%, 5% and 1% of the probability split
equally between the two tails of the standard normal distribution.
:::


## The worked example, both ways {#os-worked}

Back to the students. We have $\bar{x} = 168$, $\mu_0 = 165$, $\sigma = 8$,
$n = 36$, and we will work at $\alpha = 0.05$.

### On paper {-}

**Step 1 — Hypotheses.**

$$H_0 : \mu = 165 \qquad H_A : \mu \neq 165$$

**Step 2 — Critical value.** Two-tailed at $\alpha = 0.05$, so
$Z_{0.025} = 1.96$.

**Step 3 — Test statistic.**

$$
Z_{calc} = \frac{\bar{x} - \mu_0}{\sigma/\sqrt{n}}
= \frac{168 - 165}{8/\sqrt{36}}
= \frac{3}{8/6}
= \frac{3}{1.333}
= 2.25
$$

**Step 4 — Decision.** $|2.25| > 1.96$, so we **reject** $H_0$ at the 5% level.

**Step 5 — The $p$-value.**

$$
P(|Z| > 2.25) = 2 \times P(Z > 2.25) = 2 \times 0.01222 = 0.0244
$$

If the administration's claim were true, only 2.4% of samples would sit as far
from 165 as ours. That is unlikely enough to reject the claim.

### The same thing in R {-}

Now watch every one of those numbers reappear.

```{r os-byhand-r}
xbar  <- 168
mu0   <- 165
sigma <- 8
n     <- 36

se     <- sigma / sqrt(n)          # standard error
z_calc <- (xbar - mu0) / se        # the test statistic
z_crit <- qnorm(1 - 0.05 / 2)      # the critical value
p_val  <- 2 * pnorm(-abs(z_calc))  # the p-value

c(se = se, z_calc = z_calc, z_crit = z_crit, p_value = p_val)
```

The standard error is `r round(8/6, 4)`, the test statistic is
`r round((168-165)/(8/6), 3)`, and the $p$-value is
`r round(2*pnorm(-2.25), 4)` — exactly the hand calculation, and near enough
exactly the `r round(mean(abs(many_means - 165) >= 3), 4)` we got by brute-force
simulation at the start of the chapter.

::: {.intuition}
Notice what just happened. The simulation, the algebra, and the R code are three
descriptions of one idea. The formula is not a rule handed down from above — it
is a shortcut for the counting exercise we did by hand in Section
\@ref(os-problem).
:::

And here is the picture the numbers are describing:

```{r os-reject-plot, fig.cap = "The rejection region at $\\alpha = 0.05$ (shaded), with our test statistic of 2.25 falling inside it."}
plot_rejection(stat = 2.25, crit = 1.96, tails = "two",
               stat_label = "Z = 2.25")
```

### Letting a package do it {-}

For a real dataset rather than summary statistics, `z.test()` from the **BSDA**
package does the whole thing at once. Since our problem quoted only summary
statistics, we first build a sample consistent with them:

```{r os-ztest}
library(BSDA)

heights <- make_sample(n = 36, mean = 168, sd = 8)

z.test(heights, mu = 165, sigma.x = 8, conf.level = 0.95)
```

Read that output against the hand calculation: `z = 2.25`, `p-value = 0.0244`,
and a confidence interval that we are about to explain.

::: {.watchout}
`z.test()` needs `sigma.x` — the **population** standard deviation. If you do
not know it (and in real work you almost never do), you must not use a
$Z$-test. Use `t.test()` instead; see Section \@ref(os-ttest).
:::

## The same test, viewed as an interval {#os-ci}

::: {.question}
Unit \@ref(interval-estimation) built confidence intervals; this unit builds
tests. Are they two methods, or one?
:::

We now have two procedures that use the same sample, the same standard error
and the same critical value. It would be strange if they were unrelated, and
they are not.

Start by computing the interval for our students, exactly as in the previous
unit.

```{r os-ci-calc}
xbar + c(-1, 1) * qnorm(0.975) * (8 / sqrt(36))
```

The 95% interval is $[165.39,\ 170.61]$. The claimed value of 165 lies
**outside** it — the same verdict the test reached, and not a coincidence.

The reason is visible in the algebra. We fail to reject $H_0$ when the test
statistic is small enough, that is when

$$\left|\frac{\bar{x} - \mu_0}{\sigma/\sqrt{n}}\right| \le z_{\alpha/2}$$

Multiplying through by $\sigma/\sqrt{n}$ and rearranging for $\mu_0$ gives

$$\bar{x} - z_{\alpha/2}\frac{\sigma}{\sqrt{n}}
\;\le\; \mu_0 \;\leq\;
\bar{x} + z_{\alpha/2}\frac{\sigma}{\sqrt{n}}$$

which is the confidence interval.

::: {.keyidea}
A confidence interval and a hypothesis test are the same inequality, solved for
different unknowns.

The test fixes $\mu_0$ and asks which sample means are acceptable. The interval
fixes $\bar{x}$ and asks which values of $\mu_0$ are acceptable.
:::

### Testing every possible claim at once {-}

That equivalence can be checked rather than taken on trust. Rather than testing
the single claim $\mu_0 = 165$, test every value on a fine grid and collect the
ones that survive.

```{r os-ci-equiv}
se  <- 8 / sqrt(36)
grid <- seq(160, 176, by = 0.001)

not_rejected <- grid[abs(xbar - grid) / se <= qnorm(0.975)]

surviving <- range(not_rejected)
interval  <- xbar + c(-1, 1) * qnorm(0.975) * se

rbind(claims_that_survive_the_test = surviving,
      confidence_interval          = interval,
      difference                   = surviving - interval)
```

The two agree to within a thousandth of a centimetre, which is the spacing of
the grid — test a finer grid and the discrepancy shrinks with it. The set of
claims a test would not reject *is* the confidence interval. Nothing was
assumed to make this happen; it follows from the two being one inequality
solved for different unknowns.

```{r os-ci-pvalue-plot, echo = FALSE, fig.height = 3.4, fig.cap = "The p-value for every possible claimed value of mu. Claims above the 0.05 line survive the test, and the range of surviving claims is exactly the 95% confidence interval."}
pv <- data.frame(mu0 = seq(162, 174, length.out = 800))
pv$p <- 2 * (1 - pnorm(abs(xbar - pv$mu0) / se))
lo <- xbar - qnorm(0.975) * se
hi <- xbar + qnorm(0.975) * se

ggplot(pv, aes(mu0, p)) +
  geom_ribbon(data = subset(pv, mu0 >= lo & mu0 <= hi),
              aes(ymin = 0.05, ymax = p), fill = book_palette$fill,
              alpha = 0.85) +
  geom_line(linewidth = 0.8, colour = book_palette$ink) +
  geom_hline(yintercept = 0.05, linetype = "dashed",
             colour = book_palette$reject) +
  annotate("segment", x = c(lo, hi), xend = c(lo, hi), y = 0, yend = 0.05,
           colour = book_palette$reject, linewidth = 0.5) +
  annotate("text", x = 165, y = 0.44, label = "claims that survive",
           size = 3.5, colour = "grey25") +
  annotate("text", x = 172.6, y = 0.075, label = "α = 0.05",
           size = 3.3, colour = book_palette$reject) +
  scale_x_continuous(breaks = c(164, round(lo, 1), 168, round(hi, 1), 174)) +
  labs(x = expression(paste("claimed value  ", mu[0])), y = "p-value") +
  theme_book()
```

The curve reaches its maximum at $\mu_0 = \bar{x} = 168$, and it is worth being
clear about why the $p$-value is exactly 1 there.

A $p$-value asks how often a sample would fall at least as far from the claim as
ours did. If the claim *is* our sample mean, then our sample sits at distance
zero from it — and every conceivable sample is at least that far away. The
proportion is therefore 1. A claim that matches the data exactly leaves nothing
to explain.

As the claimed mean moves away from the observed mean in either direction, the
sample becomes increasingly surprising under that claim, and the $p$-value
falls steadily — crossing 0.05 precisely at the two endpoints of the
interval.

::: {.keyidea}
A $(1-\alpha)$ confidence interval is exactly the set of values $\mu_0$ that a
two-tailed test at level $\alpha$ would **fail** to reject.

Test and interval are one object viewed from two sides. If the interval misses
the claimed value, the test rejects it. Always.
:::

### So why keep both? {-}

Because they carry different information, and the figure above shows what each
one throws away.

A confidence interval answers infinitely many tests at once, and reports the
answer in the units of the problem. Handed $[165.39,\ 170.61]$, a reader can
evaluate any claim about average height — 165, 166, 172 — without recomputing
anything. A test answers one question, chosen in advance.

This is why confidence intervals are generally preferred in scientific papers.
A single interval communicates the estimate, its uncertainty, and the outcome
of every two-sided hypothesis test at that significance level, all at once.

A test, however, reports *how strongly*. Both 165 and 160 fall outside our
interval, and the interval says the same thing about each: not plausible. The
$p$-values do not agree — 0.024 for 165, and effectively zero for 160. One
claim is marginally implausible; the other is not remotely survivable.

Put precisely: the interval tells us **which claims remain plausible**, and the
$p$-value tells us **how strongly the data disagree with one particular
claim**.

::: {.intuition}
Report the interval. It contains the test, states its answer in the units the
reader cares about, and does not tempt anyone into treating a single threshold
as a verdict.

Report the $p$-value alongside it when one specific claim is genuinely at
issue, since that is where the interval is silent about degree.
:::

Everything Section \@ref(u3-misreadings) said about reading a confidence
interval applies unchanged here. In particular, the 95% still describes the
procedure and not this interval, and an interval that excludes a claimed value
has not proved the claim false — it has reported that the data are unlikely
under it.

## When σ is unknown: the t-test {#os-ttest}

Everything so far assumed we knew the population standard deviation $\sigma$.
We almost never do.

Everything else stays the same. We still compare the sample mean with the value
the null hypothesis claims, and we still ask how surprising that gap would be if
the claim were true. The only change is that we must first estimate $\sigma$
from the sample itself.

That estimate is not free. Replacing $\sigma$ with the sample standard deviation
$s$ adds a second source of uncertainty, and the statistic

$$t_{calc} = \frac{\bar{x} - \mu_0}{s/\sqrt{n}}$$

no longer follows a normal distribution. It follows Student's $t$ with
$n - 1$ **degrees of freedom** — a distribution with fatter tails, which demands
more evidence before it will let you reject.

```{r os-t-vs-z, echo = FALSE, fig.cap = "Student's t against the standard normal. Fewer degrees of freedom put more probability in the tails, so critical values sit further out and more evidence is needed to reject. By 30 degrees of freedom the two are already hard to tell apart."}
xs <- seq(-4.2, 4.2, length.out = 500)
lv <- c("Normal", "t, df = 30", "t, df = 10", "t, df = 5")

grid <- rbind(
  data.frame(x = xs, y = dnorm(xs),      dist = "Normal"),
  data.frame(x = xs, y = dt(xs, df = 30), dist = "t, df = 30"),
  data.frame(x = xs, y = dt(xs, df = 10), dist = "t, df = 10"),
  data.frame(x = xs, y = dt(xs, df = 5),  dist = "t, df = 5")
)
grid$dist <- factor(grid$dist, levels = lv)

ggplot(grid, aes(x, y, colour = dist, linewidth = dist)) +
  geom_line() +
  scale_colour_manual(values = c("grey45", "#2f855a", book_palette$accent,
                                 book_palette$reject), name = NULL) +
  scale_linewidth_manual(values = c(1.4, 0.75, 0.75, 0.75), guide = "none") +
  annotate("text", x = 2.9, y = 0.30, size = 3.3, colour = "grey30",
           hjust = 0, lineheight = 0.95,
           label = "As the sample grows,\nthe t distribution\napproaches the normal") +
  coord_cartesian(xlim = c(-4.2, 5.6)) +
  labs(x = NULL, y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank(), legend.position = "top")
```

### Settling the question with the actual class {-}

We opened this chapter with a made-up sample of 36 students. We can now do the
real thing. `stats-class.csv` holds the survey this class filled in — names,
ages, heights, gender:

```{r os-class-data}
students <- read.csv("data/stats-class.csv")

str(students)
```

Thirty-three rows, but look at the last one. One student left age and height
blank, and another did not give a gender:

```{r os-class-na}
students[!complete.cases(students), ]
```

::: {.watchout}
**Working with real data.** What follows is not part of hypothesis testing. It
is the sort of thing that has to be dealt with the moment you leave textbook
datasets behind, and this is the first point in the book where you have.

**A blank is not always an `NA`.** In a numeric column an empty field is read
as `NA`, but in a text column it is read as the empty string `""` — and `""`
is a perfectly ordinary value as far as `is.na()`, `complete.cases()` and
`na.rm` are concerned. A gap in a text column can therefore hide in plain
sight.

If a dataset you did not prepare yourself has blanks, say so explicitly when
you read it:

```{r os-na-strings, eval = FALSE}
read.csv("somefile.csv", na.strings = c("", "NA"))
```

The files in this book are already written so that plain `read.csv()` does the
right thing.
:::

::: {.watchout}
`mean()` returns `NA` if even one value is missing — it refuses to guess what
you meant. Pass `na.rm = TRUE` to drop the gaps, and always know how many you
dropped.

```{r os-class-narm}
mean(students$height)                  # NA — one height is missing
mean(students$height, na.rm = TRUE)    # 32 students
```
:::

Now the test. We do **not** know the population standard deviation of student
heights, so this is a $t$-test, not a $Z$-test:

```{r os-class-test}
heights_real <- students$height[!is.na(students$height)]

c(n = length(heights_real),
  mean = mean(heights_real),
  sd = sd(heights_real))

t.test(heights_real, mu = 165)
```

```{r os-class-stash, include = FALSE}
ct <- t.test(heights_real, mu = 165)
```

$t = `r round(ct$statistic, 2)`$ on `r ct$parameter` degrees of freedom, with a
$p$-value of `r round(ct$p.value, 4)`. We **reject** the claim that the average
is 165 cm. The 95% interval,
$[`r round(ct$conf.int[1], 1)`,\ `r round(ct$conf.int[2], 1)`]$, excludes 165 —
the same verdict, as always.

::: {.intuition}
Notice how much narrower this is than the textbook version of the problem. The
opening example assumed we somehow knew $\sigma = 8$. Here we had to estimate
it from 32 people, and the $t$ distribution charges us for that uncertainty
with a wider interval and a higher bar to clear.

**Statistical validity is not sampling validity.** Notice what the data is not:
32 students in one classroom are not a random sample of all APU students.

Every number above is correctly computed. The $t$ statistic, the degrees of
freedom, the $p$-value and the interval are all exactly right *for this sample*.
Whether they answer a question about **the university** is a separate matter,
and a harder one — and no amount of care with the arithmetic addresses it. The
formulas cannot tell you which population you were sampling from; only the
design can.
:::

### Worked example {-}

The dean estimates that full-time faculty teach 11.0 classroom hours per week.
You sample eight faculty members. Can you reject the dean's claim at
$\alpha = 0.10$? At $\alpha = 0.01$?

```{r os-t-data}
hours <- c(11.8, 8.6, 10.5, 7.9, 6.4, 10.4, 10.0, 8.2)

c(n = length(hours), mean = mean(hours), sd = sd(hours))
```

::: {.yourturn}
Before calculating anything, form an expectation.

The claim is 11.0 hours and the sample averages 9.2 — a gap of nearly two
hours. But there are only eight faculty members, and their hours range from 6.4
to 11.8, so the sample is both small and quite spread out.

Do you expect to reject the claim at the 10% level? At the 1%? Commit to an
answer before reading on, and then see whether the arithmetic agrees.
:::

**On paper.** With $\bar{x} = 9.225$, $s = 1.749$, $n = 8$ and $df = 7$:

$$
t_{calc} = \frac{9.225 - 11}{1.749/\sqrt{8}}
= \frac{-1.775}{0.618}
= -2.87
$$

The critical values are $t_{0.05,\,7} = 1.895$ and $t_{0.005,\,7} = 3.499$.

- At $\alpha = 0.10$: $2.87 > 1.895$ $\Rightarrow$ **reject** $H_0$.
- At $\alpha = 0.01$: $2.87 < 3.499$ $\Rightarrow$ **fail to reject** $H_0$.

**In R.** Both the pieces and the packaged version:

```{r os-t-manual}
t_calc <- (mean(hours) - 11) / (sd(hours) / sqrt(length(hours)))
t_crit <- qt(c(0.95, 0.995), df = 7)     # 10% and 1%, two-tailed

c(t_calc = t_calc, crit_10pct = t_crit[1], crit_1pct = t_crit[2])
```

```{r os-t-test}
t.test(hours, mu = 11, conf.level = 0.90)
```

The $p$-value is `r round(t.test(hours, mu = 11)$p.value, 4)` — below 0.10 but
above 0.01, exactly matching the two different verdicts we reached by hand.

::: {.intuition}
The same data can be "significant" at one level and not at another. Nothing has
changed about the world between those two lines — only how much risk of a false
alarm we were willing to tolerate. This is why reporting the $p$-value itself
is far more informative than reporting a bare "significant / not significant."
:::

## One-tailed tests {#os-onetailed}

::: {.question}
Sometimes only one direction would worry us. Why spend half the error budget on
the other?
:::

A paper mill discharges effluent into a river. The state pollution control
board sets a limit on biochemical oxygen demand of 30 mg per litre, and the
mill claims its average discharge is *below* that limit.

Only one direction matters here. Nobody objects if the effluent is far cleaner
than required; the concern is entirely about exceeding the limit. So the claim
being tested is one-sided:

$$H_0 : \mu \geq 30 \qquad H_A : \mu < 30$$

All the probability we are willing to risk now goes into a single tail, so the
critical value moves *inward*: $-Z_{0.05} = -1.645$ rather than $\pm 1.96$.

```{r os-onetail}
qnorm(0.05)             # left-tailed critical value at 5%
qnorm(0.95)             # right-tailed critical value at 5%
```

In R, add the `alternative` argument:

```{r os-onetail-test}
bod <- make_sample(n = 20, mean = 29.2, sd = 7)

z.test(bod, mu = 30, sigma.x = 7, alternative = "less", conf.level = 0.95)
```

$Z_{calc} = -0.51$, which is not below $-1.645$, so we fail to reject. The
mill's claim is not supported by this sample — a mean of 29.2 would arise about
30% of the time even if the true average were exactly at the limit.

Two details in that output are worth noticing. The interval has an *upper*
limit of 31.77 for a *left*-tailed test, which looks wrong and is not; that is
the subject of the last part of this section. And `z.test()` prints `NA` for
the lower limit where it means $-\infty$, whereas `t.test()` prints `-Inf`
below. The two packages simply report the same thing differently.

### Unknown σ: the one-sided t-test {-}

Nothing about one-tailed testing is special to the $Z$-test. When $\sigma$ is
unknown, the same `alternative` argument does the same job.

Return to the faculty teaching hours. The dean claims 11.0 hours per week, and
suppose we specifically suspect the true figure is *lower* — we are not
interested in the possibility that faculty teach more.

$$H_0 : \mu \geq 11 \qquad H_A : \mu < 11$$

```{r os-onetail-t}
t.test(hours, mu = 11, alternative = "less", conf.level = 0.90)
```

Notice that nothing fundamental has changed.

- The hypotheses are written exactly as before.
- The direction comes from `alternative`, not from a different formula.
- The $p$-value is one-sided — 0.012 here, half the two-sided 0.024 we computed
  earlier.
- The confidence interval is one-sided, running from $-\infty$ to 10.10.
- The only substantive change is the reference distribution: Student's $t$ with
  7 degrees of freedom in place of the normal.

::: {.keyidea}
There are not six different procedures to learn. There are two — the $Z$-test
when $\sigma$ is known and the $t$-test when it is not — and each can be
pointed in any of three directions.

Everything else stays the same.
:::

### The one-sided interval looks backwards {-}

This trips up nearly everyone. For a left-tailed test, the confidence interval
is

$$\left(-\infty,\ \bar{x} + Z_\alpha \frac{\sigma}{\sqrt{n}}\right)$$

An *upper* bound — for a *left*-tailed test. Why?

::: {.theory}
Start from the rejection rule and solve for $\mu_0$ rather than for $\bar{x}$:

$$
\begin{aligned}
\frac{\bar{x} - \mu_0}{\sigma/\sqrt{n}} &< -Z_\alpha \\[4pt]
\bar{x} - \mu_0 &< -Z_\alpha \frac{\sigma}{\sqrt{n}} \\[4pt]
\mu_0 &> \bar{x} + Z_\alpha \frac{\sigma}{\sqrt{n}}
\end{aligned}
$$

The inequality flipped when $\mu_0$ moved across. So we reject exactly when
$\mu_0$ is **too large**, and the values we *fail* to reject are those with
$\mu \le \bar{x} + Z_\alpha\,\sigma/\sqrt{n}$.

The result is identical for the $t$-test. Replace $Z_\alpha$ with
$t_{\alpha,\,n-1}$ and $\sigma$ with $s$; not a single step of the argument
changes. That is why the interval above ran to 10.10 rather than starting
there.
:::

::: {.intuition}
The rejection region is defined in **$Z$-space**; the confidence interval lives
in **parameter space**. Solving the inequality for $\mu$ reverses the direction.

Put plainly: a left-tailed test rejects when the sample mean comes out too
small. So the values of $\mu$ it rules out are the ones that are too *large*
relative to what we saw. That leaves an upper bound.
:::

## Two ways to be wrong {#os-errors}

::: {.question}
We have spent this unit controlling one kind of mistake. What about the other?
:::

Every test so far has been built around $\alpha$, the risk of rejecting a claim
that happens to be true. We chose it in advance, usually 0.05, and everything
followed.

But rejecting a true claim is only one way to get the answer wrong. We could also
fail to reject a false one — conclude that a drug does nothing when it works,
that a programme has no effect when it does.

That second error has had no name and no symbol so far. It gets both now, and it
turns out to matter at least as much as the first.

### The two errors {-}

::: {.keyidea}
A **Type I error** is rejecting $H_0$ when $H_0$ is true. A false alarm.

$$\alpha = P(\text{reject } H_0 \mid H_0 \text{ true})$$

A **Type II error** is failing to reject $H_0$ when $H_A$ is true. A missed
detection.

$$\beta = P(\text{fail to reject } H_0 \mid H_A \text{ true})$$
:::

Think of a smoke alarm.

Sometimes there is no fire, and the alarm goes off anyway — a false alarm, and a
Type I error. Setting $\alpha = 0.05$ says we will tolerate this about five times
in a hundred when nothing is wrong.

Sometimes there *is* a fire and the alarm stays silent. That is a Type II error,
and it is the one that burns the house down.

Table: (\#tab:os-errors-table) The four possible outcomes of a hypothesis test.

| | **Fail to reject $H_0$** | **Reject $H_0$** |
|:--|:--|:--|
| **$H_0$ is true** | Correct | **Type I error** ($\alpha$) |
| **$H_A$ is true** | **Type II error** ($\beta$) | Correct |

Two cells are mistakes, and they are not the same mistake. A test cannot make
both at once — which of them is even possible depends on a fact about the world
that we do not know.

### The umpire's call {-}

Cricket's Decision Review System is a hypothesis test, and a well-designed one.

$$H_0 : \text{the on-field umpire was right}
\qquad
H_A : \text{the on-field umpire was wrong}$$

A Type I error overturns a correct decision. A Type II error lets a wrong
decision stand.

::: {.intuition}
The system's "umpire's call" rule — where marginal ball-tracking evidence leaves
the original decision standing — is exactly a choice to keep $\alpha$ small.

Overturning requires clear evidence, not a coin-flip's worth. The designers
decided that wrongly overturning a correct umpire is a worse error than
occasionally failing to correct a wrong one, and they built that judgement into
the threshold.

Every choice of $\alpha$ is a judgement of the same kind, whether or not it is
made deliberately.
:::

### Why errors happen at all {-}

Both errors have the same root cause: the sampling distribution under $H_0$ and
the sampling distribution under $H_A$ overlap.

If a sample mean lands in the region where the two overlap, it is consistent with
either explanation, and no procedure can tell them apart with certainty.

```{r os-power-plot, echo = FALSE, fig.height = 3.8, cache = TRUE, fig.cap = "Two sampling distributions: the left curve holds if the null is true, the right one if the true mean is 2. The vertical line is the critical value, 1.645. Alpha is the sliver of the null distribution beyond it -- rejecting when the null is true. Beta is the much larger part of the alternative distribution below it -- failing to reject when the alternative is true. Power is what remains."}
xg <- seq(-3.5, 5.5, length.out = 800)
crit <- qnorm(0.95)

d0 <- data.frame(x = xg, y = dnorm(xg, 0, 1))
d1 <- data.frame(x = xg, y = dnorm(xg, 2, 1))

ggplot() +
  geom_area(data = subset(d1, x <= crit), aes(x, y),
            fill = book_palette$reject, alpha = 0.45) +
  geom_area(data = subset(d0, x >= crit), aes(x, y),
            fill = "#2f855a", alpha = 0.55) +
  geom_line(data = d0, aes(x, y), colour = "grey25", linewidth = 0.8) +
  geom_line(data = d1, aes(x, y), colour = book_palette$accent, linewidth = 0.8) +
  geom_vline(xintercept = crit, linewidth = 0.6, colour = "grey15") +
  annotate("text", x = -0.9, y = 0.30, label = "H[0]~true", parse = TRUE,
           colour = "grey25", size = 3.4) +
  annotate("text", x = 3.4, y = 0.30, label = "mu==2", parse = TRUE,
           colour = book_palette$accent, size = 3.4) +
  annotate("text", x = 2.45, y = 0.022, label = "alpha", parse = TRUE,
           colour = "#2f855a", size = 4) +
  annotate("text", x = 0.95, y = 0.055, label = "beta", parse = TRUE,
           colour = book_palette$reject, size = 4) +
  annotate("text", x = crit, y = 0.43, label = "critical value", size = 3.2,
           colour = "grey15", hjust = -0.08) +
  labs(x = "Sample mean (standard errors from the null)", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank(), panel.grid.minor = element_blank())
```

Notice the shape of the trade-off. Moving the critical value to the right shrinks
the green area and enlarges the red one: fewer false alarms, more missed
detections. Moving it left does the reverse.

::: {.watchout}
**$\alpha$ and $\beta$ cannot both be made small by choosing a threshold.**

The threshold only slides the boundary between them. Reducing $\alpha$ from 0.05
to 0.01 buys fewer false alarms at the direct cost of more missed real effects.

The only way to reduce both at once is to separate the two distributions — which
means a larger sample, a bigger effect, or less noise. That is a matter of study
design, not of choosing a number.
:::

### Power {-}

::: {.keyidea}
The **power** of a test is the probability of detecting an effect that is really
there:

$$\text{Power} = 1 - \beta = P(\text{reject } H_0 \mid H_A \text{ true})$$

$\alpha$ is chosen. Power is a *consequence* — of the sample size, the size of
the effect, and the noise in the data.
:::

Unlike $\alpha$, power cannot be stated without saying *how large* the effect is.
A test that easily detects a large effect may be hopeless against a small one, so
"the power of this test" is always shorthand for "the power against an effect of
this particular size."

### Working it out {-}

A new drug is believed to lower blood pressure. We test

$$H_0 : \mu \leq 0 \qquad H_A : \mu > 0$$

where $\mu$ is the average reduction. Suppose $\sigma = 4$, the trial has
$n = 16$ patients, and we work at $\alpha = 0.05$.

**Step 1 — the rejection rule.** One-sided at 5%, so the critical value is
$z_{0.05} = 1.645$, and we reject when $Z > 1.645$.

**Step 2 — suppose the drug works.** Say the true average reduction is
$\mu_1 = 2$ points. The standard error is

$$\mathrm{se} = \frac{\sigma}{\sqrt{n}} = \frac{4}{\sqrt{16}} = 1$$

**Step 3 — where the statistic now sits.** Under $H_0$ the statistic
$Z = \bar{X}/\mathrm{se}$ is centred on zero. If the truth is $\mu_1 = 2$, it is
centred instead on

$$\delta = \frac{\mu_1 - \mu_0}{\mathrm{se}} = \frac{2 - 0}{1} = 2$$

**Step 4 — the probability of missing it.** We fail to reject when
$Z \leq 1.645$. Under the alternative, $Z$ is centred on 2, so

$$\beta = P(Z \leq 1.645 \mid \mu = 2) = P(Z - 2 \leq -0.355) = \Phi(-0.355)$$

```{r os-power-calc, cache = TRUE}
se    <- 4 / sqrt(16)
delta <- (2 - 0) / se
crit  <- qnorm(0.95)

c(critical_value = crit,
  shift          = delta,
  beta           = pnorm(crit - delta),
  power          = 1 - pnorm(crit - delta))
```

So $\beta = 0.36$ and the power is **0.64**.

Read that carefully. The drug genuinely works — it lowers blood pressure by two
points on average — and this trial has a better than one-in-three chance of
failing to notice.

::: {.watchout}
A trial like this is **underpowered**. If it returns "no significant effect",
that tells us very little: we already knew the study would miss a real
two-point effect about 36% of the time.

This is the sharpest version of a point made in Section \@ref(os-pvalue). Failing
to reject $H_0$ is not evidence that $H_0$ is true, and when power is low it is
barely evidence of anything at all.
:::

### What determines power {-}

Four things, and it is worth seeing which of them are under our control.

```{r os-power-table, cache = TRUE}
power_at <- function(n, effect = 2, sigma = 4, alpha = 0.05) {
  se <- sigma / sqrt(n)
  round(1 - pnorm(qnorm(1 - alpha) - effect / se), 3)
}

data.frame(n            = c(16, 25, 50, 100, 200),
           power_vs_2   = sapply(c(16, 25, 50, 100, 200), power_at),
           power_vs_1   = sapply(c(16, 25, 50, 100, 200), power_at, effect = 1))
```

::: {.keyidea}
Power rises when

- the **effect is larger** — big effects are easy to find;
- the **sample is larger** — the two distributions pull apart;
- the **noise is smaller** — the distributions are narrower;
- **$\alpha$ is larger** — a lower bar to clear, at the cost of more false
  alarms.

The first is a fact about the world. The last is a choice with a price. Only the
middle two are improvements in the ordinary sense.
:::

The table shows the practical consequence. Against a two-point effect this trial
needs about fifty patients for respectable power. Against a one-point effect,
even two hundred patients leave it detecting the effect barely two-thirds of the
time.

::: {.intuition}
This is why power should be computed *before* a study, not after.

Deciding how many patients to recruit means asking: what is the smallest effect
worth detecting, and how many observations does it take to have a fair chance of
detecting it? A study that cannot answer its question is not worth running, and
that can be known in advance.

Computing power after a null result is a different and largely futile exercise.
The honest statement is the one the design already implied.
:::


## Reference tables {#os-reference}

Table: (\#tab:os-ref-z) Every one-sample test of a mean in this unit. Use `z.test()` when $\sigma$ is known and `t.test()` when it is not; the argument below selects the direction. There are only two procedures, each run three ways.

| $\sigma$ | Direction | Hypotheses | Reject when | R argument |
|:--|:--|:--|:--|:--|
| Known | Two-sided | $H_0: \mu = \mu_0$ vs $H_A: \mu \neq \mu_0$ | $\lvert Z_{calc} \rvert > Z_{\alpha/2}$ | `alternative = "two.sided"` |
| Known | Left | $H_0: \mu \geq \mu_0$ vs $H_A: \mu < \mu_0$ | $Z_{calc} < -Z_{\alpha}$ | `alternative = "less"` |
| Known | Right | $H_0: \mu \leq \mu_0$ vs $H_A: \mu > \mu_0$ | $Z_{calc} > Z_{\alpha}$ | `alternative = "greater"` |
| Unknown | Two-sided | $H_0: \mu = \mu_0$ vs $H_A: \mu \neq \mu_0$ | $\lvert t_{calc} \rvert > t_{\alpha/2,\,n-1}$ | `alternative = "two.sided"` |
| Unknown | Left | $H_0: \mu \geq \mu_0$ vs $H_A: \mu < \mu_0$ | $t_{calc} < -t_{\alpha,\,n-1}$ | `alternative = "less"` |
| Unknown | Right | $H_0: \mu \leq \mu_0$ vs $H_A: \mu > \mu_0$ | $t_{calc} > t_{\alpha,\,n-1}$ | `alternative = "greater"` |

Read down the first column rather than across all six rows. The distribution
changes when $\sigma$ becomes unknown; the direction changes where the
rejection region sits. Nothing else differs.

Table: (\#tab:os-ref-crit) Standard normal critical values. Every figure below is `qnorm()` evaluated inline, not copied from a printed table.

| $\alpha$ | Two-tailed, $Z_{\alpha/2}$ | One-tailed, $Z_{\alpha}$ |
|:--|:--|:--|
| 10% | `r round(qnorm(0.95), 3)` | `r round(qnorm(0.90), 3)` |
| 5%  | `r round(qnorm(0.975), 3)` | `r round(qnorm(0.95), 3)` |
| 1%  | `r round(qnorm(0.995), 3)` | `r round(qnorm(0.99), 3)` |

## Summary {#os-summary}

::: {.keyidea}
1. A gap between a sample and a claim proves nothing on its own — sampling
   noise guarantees some gap. The test asks whether the gap is *bigger than
   noise*.
2. $\displaystyle Z_{calc} = \frac{\bar{x}-\mu_0}{\sigma/\sqrt{n}}$ when
   $\sigma$ is known; $\displaystyle t_{calc} = \frac{\bar{x}-\mu_0}{s/\sqrt{n}}$
   with $df = n-1$ when it is not.
3. Reject when the statistic beats the critical value, or equivalently when the
   $p$-value falls below $\alpha$. The two rules never disagree.
4. A confidence interval is the set of null values a test would *not* reject.
   Interval and test are the same statement.
5. A $p$-value is $P(\text{data} \mid H_0)$, never $P(H_0 \mid \text{data})$.
6. The equals sign always lives in $H_0$. "Can you reject the claim?" puts the
   claim in $H_0$; "is there evidence supporting the claim?" puts it in $H_A$.
7. A **Type I error** rejects a true null, with probability $\alpha$. A **Type
   II error** fails to reject a false one, with probability $\beta$. Lowering
   one raises the other; only better data lowers both.
8. **Power** $= 1 - \beta$ is the chance of detecting an effect that is really
   there. It is not chosen but inherited — from the sample size, the size of the
   effect and the noise. A trial with 64% power misses a real effect more than
   once in three.
9. `qnorm()` and `qt()` replace every printed table you own.
:::



## Practice problems {#os-exercises}

::: {.yourturn}
Work each one on paper first, then check yourself in R. The solutions are one
click away, which makes them worth nothing unless you try first.

These are for learning, so the workings are shown in full. The
[question bank](#os-question-bank) that follows is for testing yourself, and
has no answers.
:::

**11.1** A consumer research organisation states that the mean caffeine content
per 12-ounce bottle of a caffeinated soft drink is 37.7 mg. A random sample of
36 bottles has a mean of 35.28 mg. The population standard deviation is 10.8 mg.
At $\alpha = 0.01$, can you reject the organisation's claim?

::: {.answer}
$H_0: \mu = 37.7$ against $H_A: \mu \neq 37.7$. Two-tailed at $\alpha = 0.01$,
so the critical value is $Z_{0.005} = 2.58$.

$$
Z_{calc} = \frac{35.28 - 37.7}{10.8/\sqrt{36}} = \frac{-2.42}{1.8} = -1.34
$$

```{r os-ex1}
caffeine <- make_sample(n = 36, mean = 35.28, sd = 10.8)
z.test(caffeine, mu = 37.7, sigma.x = 10.8, conf.level = 0.99)
```

$|-1.34| < 2.58$ and the $p$-value of 0.179 is far above 0.01, so we **fail to
reject**. If the true mean really were 37.7 mg, a sample mean this far away
would turn up about 18% of the time — thoroughly unremarkable.
:::

**11.2** Historical data show household water use is normally distributed with
mean 360 gallons and standard deviation 40 gallons per day. A sample of 200
households averages 374 gallons. Are these data consistent with the historical
distribution at the 5% level? What is the $p$-value?

::: {.answer}
$$
Z_{calc} = \frac{374 - 360}{40/\sqrt{200}} = \frac{14}{2.828} = 4.95
$$

```{r os-ex2}
water <- make_sample(n = 200, mean = 374, sd = 40)
z.test(water, mu = 360, sigma.x = 40, conf.level = 0.95)
```

$4.95 \gg 1.96$, and the $p$-value is about
`r signif(2 * pnorm(-4.95), 2)` — under one in a million. **Reject $H_0$.**
Water use has moved away from the historical mean.
:::

**11.3** A study claims the average monthly salary of full professors in India
is ₹87,800. Students at one university sample ten salaries (in thousands):

```{r os-ex3-data, echo = FALSE, comment = NA}
cat("91.0, 79.8, 102.0, 93.5, 82.0, 88.6, 90.0, 98.6, 101.0, 84.0")
```

They believe their professors earn *more*. Test at the 10% and 5% levels.

::: {.answer}
The word "more" makes this **right-tailed**:
$H_0: \mu \leq 87.8$ against $H_A: \mu > 87.8$. Since $\sigma$ is unknown and
$n = 10$, use a $t$-test with $df = 9$.

```{r os-ex3}
salary <- c(91.0, 79.8, 102.0, 93.5, 82.0, 88.6, 90.0, 98.6, 101.0, 84.0)

c(mean = mean(salary), sd = sd(salary))
t.test(salary, mu = 87.8, alternative = "greater")
```

$t_{calc} = `r round((mean(c(91.0,79.8,102.0,93.5,82.0,88.6,90.0,98.6,101.0,84.0)) - 87.8) / (sd(c(91.0,79.8,102.0,93.5,82.0,88.6,90.0,98.6,101.0,84.0))/sqrt(10)), 3)`$,
against critical values $t_{0.10,\,9} = `r round(qt(0.90, 9), 3)`$ and
$t_{0.05,\,9} = `r round(qt(0.95, 9), 3)`$.

The statistic falls short of both. **Fail to reject** at either level — there is
not enough evidence that these professors are paid above the national figure.
:::

**11.4** A Department of Transportation claims mean wait time is *at most* 6
minutes. A sample of 34 locations has mean 10.3 minutes, standard deviation 8.0.
Test at 5% and 1%.

::: {.answer}
"At most" makes this right-tailed: $H_0: \mu \leq 6$ against $H_A: \mu > 6$,
$t$-test with $df = 33$.

```{r os-ex4}
waits <- make_sample(n = 34, mean = 10.3, sd = 8.0)
t.test(waits, mu = 6, alternative = "greater")
```

$t_{calc} = 3.14$, well beyond $t_{0.05,\,33} = `r round(qt(0.95, 33), 3)`$ and
$t_{0.01,\,33} = `r round(qt(0.99, 33), 3)`$. **Reject at both levels** — wait
times are longer than claimed.
:::

**11.5** `marriage-ages.csv` records the ages of both partners in ten couples.
Is there evidence that one partner tends to be older? Test at the 5% level.

*Hint: the two columns are not independent samples — they are ten couples. What
is the one number per couple that captures the question?*

::: {.answer}
The two columns are **paired**: each row is one couple. Treating them as two
independent samples throws away that pairing and inflates the standard error.
The right move is to reduce each couple to a single number — the age gap — and
run a **one-sample** test on it against zero.

```{r os-ex5}
couples <- read.csv("data/marriage-ages.csv")

gap <- couples$partner1 - couples$partner2
gap

c(mean = mean(gap), sd = sd(gap), n = length(gap))
```

```{r os-ex5-test}
t.test(gap, mu = 0)
```

The mean gap is 2.5 years, $t = 3.0$ on 9 degrees of freedom, $p = 0.015$.
Since $0.015 < 0.05$, we **reject** $H_0$: partner 1 is reliably older.

R will do the pairing for you if you ask:

```{r os-ex5-paired}
t.test(couples$partner1, couples$partner2, paired = TRUE)
```

Identical output — `paired = TRUE` computes exactly the differences we formed
by hand. Compare what happens if you forget:

```{r os-ex5-wrong}
t.test(couples$partner1, couples$partner2, paired = FALSE)$p.value
```

```{r os-ex5-stash, include = FALSE}
p_unpaired <- t.test(couples$partner1, couples$partner2, paired = FALSE)$p.value
p_paired   <- t.test(couples$partner1, couples$partner2, paired = TRUE)$p.value
```

The same ten couples, a $p$-value of `r round(p_unpaired, 3)` instead of
`r round(p_paired, 3)`, and the opposite conclusion. **Pairing is not a
detail.**
:::

**11.6** *(No computation.)* A classmate reports "$p = 0.03$, so there is a 3%
chance the null hypothesis is true." Explain what is wrong, and state what
$p = 0.03$ does mean.

::: {.answer}
The $p$-value is computed **assuming the null is true**. It cannot then also be
the probability that the null is true — that would be circular. Formally, your
classmate has swapped $P(\text{data} \mid H_0)$ for $P(H_0 \mid \text{data})$,
which are different quantities entirely.

The correct reading: *if* $H_0$ were true, samples at least as extreme as this
one would arise 3% of the time. Since that is unusual, we treat it as evidence
against $H_0$ — but the statement is about how surprising the data are, not
about how probable the hypothesis is.
:::

### Type I and Type II errors {-}

**4.15** A district administration tests whether a new teaching programme raised
average test scores. State, in the language of the problem, what a Type I error
and a Type II error would be, and say which you would consider more serious.

::: {.answer}
A **Type I error** is concluding the programme raised scores when it did not.
The district would scale up an ineffective programme, spending money and
teaching time that could have gone elsewhere, and would believe a problem had
been solved when it had not.

A **Type II error** is concluding the programme did not work when it did. A
genuinely effective intervention would be abandoned.

Which is worse is not a statistical question. If the programme is cheap and the
alternative is doing nothing, a missed effect is the costlier error and a larger
$\alpha$ might be defensible. If scaling up is expensive and irreversible, the
false alarm is worse.

The point is that $\alpha$ encodes such a judgement whether or not anyone makes
it deliberately. Choosing 0.05 out of habit is still choosing.
:::

**4.16** Explain why setting $\alpha = 0.001$ does not make a test "more
accurate".

::: {.answer}
It makes the test less likely to raise a false alarm and *more* likely to miss a
real effect. Lowering $\alpha$ moves the critical value further out, which
shrinks the rejection region — and the rejection region is where real effects get
detected too.

Accuracy in any useful sense involves both errors. Reducing one by moving the
threshold always increases the other, because the threshold is the boundary
between them.

Both fall together only when the two sampling distributions move apart, which
requires a larger sample, a bigger effect or less noise. That is a matter of
design, not of choosing a smaller number.
:::

**4.17** A trial with $\sigma = 4$ and $n = 16$ tests $H_0: \mu \leq 0$ against
$H_A: \mu > 0$ at $\alpha = 0.05$. The true effect is $\mu_1 = 2$.

i. Compute the critical value and the standard error.
ii. Compute $\beta$ and the power.
iii. Recompute the power at $\alpha = 0.01$.
iv. What does the comparison show?

::: {.answer}
**(i)** One-sided at 5%: $z_{0.05} = 1.645$. Standard error
$= 4/\sqrt{16} = 1$.

**(ii)** Under the alternative the statistic is centred on
$\delta = (2-0)/1 = 2$, so

$$\beta = P(Z \leq 1.645 - 2) = \Phi(-0.355) = 0.361$$

and the power is $1 - 0.361 = 0.639$.

**(iii)** At $\alpha = 0.01$ the critical value is $z_{0.01} = 2.326$, so
$\beta = \Phi(2.326 - 2) = \Phi(0.326) = 0.628$ and the power falls to **0.372**.

**(iv)** Tightening $\alpha$ from 5% to 1% cut the chance of detecting a real
two-point effect from 64% to 37%. The stricter test is not the better test; it
has simply traded one error for the other. Nothing about the drug changed.
:::

**4.18** How large must the trial in the previous question be to detect a
two-point effect with 80% power?

::: {.answer}
Power of 80% means $\beta = 0.20$, and the critical value must sit $z_{0.20}$
below the alternative's centre. Since $z_{0.20} = 0.842$, we need

$$\delta = z_{\alpha} + z_{\beta} = 1.645 + 0.842 = 2.487$$

With $\delta = (\mu_1 - \mu_0)\sqrt{n}/\sigma$,

$$\sqrt{n} = \frac{2.487 \times 4}{2} = 4.97
\qquad\Longrightarrow\qquad
n = 24.7$$

so 25 patients. Checking against the table in Section \@ref(os-errors), $n = 25$
gives power 0.804.

Note the shape of this calculation: it is done *before* the trial, and it
requires naming the smallest effect worth detecting. That is a judgement about
medicine, not statistics, and it has to be made either way.
:::

**4.19** Plot the power of the trial in question 4.17 against sample size, for
true effects of one point and two points.

::: {.answer}
```{r os-ex-power, fig.height = 3.4, fig.cap = "Power against sample size for two effect sizes. Small effects need far more data."}
power_at <- function(n, effect, sigma = 4, alpha = 0.05) {
  1 - pnorm(qnorm(1 - alpha) - effect / (sigma / sqrt(n)))
}

grid <- expand.grid(n = seq(5, 250, by = 5), effect = c(1, 2))
grid$power <- power_at(grid$n, grid$effect)
grid$effect <- factor(grid$effect, labels = c("1 point", "2 points"))

ggplot(grid, aes(n, power, colour = effect)) +
  geom_hline(yintercept = 0.8, linetype = "dashed", colour = "grey45") +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = c(book_palette$reject, book_palette$accent),
                      name = "True effect") +
  labs(x = "Sample size", y = "Power") +
  theme_book() +
  theme(legend.position = "top")
```

Both curves rise towards 1, and the dashed line marks the conventional 80%
target. Detecting a two-point effect takes about 25 patients; detecting a
one-point effect takes about 100.

Halving the effect roughly quadruples the sample needed, which is the $\sqrt{n}$
rule appearing once more: precision improves with the square root of the sample,
so detecting an effect half the size requires four times the data.
:::

## Question bank {#os-question-bank}

::: {.watchout}
**No solutions are given for this section, by design.** These questions are
drawn from past quizzes and exams. If you can work them without checking an
answer key, you are ready; if you cannot, the section to reread is named
beside each one.

They run from easiest to hardest. Do them in order.
:::

### Multiple choice {-}

**QB 11.1** The level of significance, $\alpha$, is the risk of *(select all
that apply)*

a. Rejecting the null hypothesis when it is true
b. Rejecting the null hypothesis when the alternative hypothesis is true
c. Rejecting the alternative hypothesis when the null hypothesis is true
d. Rejecting the alternative hypothesis when it is true

*See Section \@ref(os-critical).*

**QB 11.2** An economist sets the alternative hypothesis that the average CGPA
of undergraduate students is less than 8. What type of test should be used, and
what is the critical value at $\alpha = 5\%$?

a. One-tailed test and $+1.96$
b. One-tailed test and $-1.96$
c. One-tailed test and $+1.645$
d. One-tailed test and $-1.645$

*See Section \@ref(os-onetailed).*

**QB 11.3** The margin of error for a confidence interval is
$\pm Z_{\alpha/2}\,\sigma/\sqrt{n}$. Which statements are correct? *(select all
that apply)*

a. Increasing the sample size will decrease the margin of error
b. The margin of error decreases as $\alpha$ rises
c. The larger the margin of error, the less confident one is about the estimate
d. The margin of error increases with the population standard deviation $\sigma$

*See Section \@ref(os-ci).*

**QB 11.4** A 95% confidence interval for mean student height is
$(165.41,\ 170.95)$. You now compute a 99% interval from the same data. Which
statements are likely correct? *(select all that apply)*

a. The 90%, 95% and 99% intervals are the same
b. The 99% interval is $(164.54,\ 171.83)$
c. The 99% interval is $(165.86,\ 170.51)$
d. The answer cannot be determined from the information given

*See Section \@ref(os-ci). Ask yourself which way an interval must move when you demand more confidence.*

**QB 11.5** A company tests the null hypothesis that the average time to accept
a job offer is at most 7 days. The test concludes that the average exceeds 7
days. It is later learned that the true mean is exactly 7 days. What error
occurred?

a. Type II error
b. Type I error
c. A correct decision was made
d. Cannot be determined from the information provided

### Short reasoning {-}

**QB 11.6** "A high $p$-value is evidence that the null hypothesis is true."
Do you agree? Why or why not?

*See Section \@ref(os-pvalue).*

**QB 11.7** Two economists, Amit and Anand, collect data on Indian agricultural
wages to test the effect of MNREGA. Amit concludes that wages after the policy
are higher than before, when in fact wages have not changed. Anand concludes
that wages before and after are the same — also incorrectly.

It is claimed that Amit has committed a Type I error and Anand a Type II error.
Do you agree? Why or why not?

### Numerical {-}

**QB 11.8** A researcher wants to estimate the average weekly study hours of
university students. The population standard deviation is 6 hours. What sample
size is needed to estimate the mean with a margin of error of 1 hour at 90%
confidence?

a. $n = 98$
b. $n = 64$
c. $n = 121$
d. $n = 82$

**QB 11.9** The numbers of people taking the morning university shuttle on 16
randomly chosen days are:

```{r qb-shuttle, echo = FALSE, comment = NA}
cat("27, 46, 35, 33, 29, 45, 28, 24, 30, 40, 38, 35, 41, 27, 38, 29")
```

The administration claims the daily average is 30. Taking $\sigma = 7$:

a. Construct a 99% confidence interval for the mean number of riders.
b. Use that interval to test the administration's claim.
c. Would your conclusion change at the 95% level? Explain *why* before you
   compute it.

*Part (c) is the one that separates understanding from arithmetic.*


### Errors and power {-}

**QB 4.18** A Type II error is

a. rejecting a true null hypothesis
b. failing to reject a false null hypothesis
c. choosing the wrong significance level
d. computing the $p$-value incorrectly

**QB 4.19** Reducing $\alpha$ from 0.05 to 0.01, holding everything else fixed,

a. reduces both error probabilities
b. reduces Type I error and increases Type II error
c. increases Type I error and reduces Type II error
d. leaves power unchanged

**QB 4.20** The power of a test is

a. $\alpha$
b. $1 - \alpha$
c. $\beta$
d. $1 - \beta$

**QB 4.21** Power increases when *(select all that apply)*

a. the sample size grows
b. the true effect is larger
c. the population standard deviation is larger
d. $\alpha$ is made smaller

**QB 4.22** A study reports "no significant effect" and had 30% power against
the effect it was looking for. The most accurate conclusion is

a. there is no effect
b. the effect is small
c. the study was unlikely to detect the effect even if it existed
d. the null hypothesis has been confirmed

**QB 4.23** In the Decision Review System, "umpire's call" on a marginal
ball-tracking decision reflects a choice to

a. keep $\alpha$ small
b. keep $\beta$ small
c. maximise power
d. eliminate both errors

**QB 4.24** Explain why power cannot be stated for a test without also stating
an effect size, whereas $\alpha$ can.

**QB 4.25** A colleague computes the power of a study after it has returned a
null result, using the effect size the study itself estimated. Explain what is
wrong with this.

**QB 4.26** A quality inspector tests whether a machine's mean fill weight has
drifted from 500 g. With $\sigma = 8$ g, $n = 25$ and $\alpha = 0.05$ two-sided:

i. State the critical values.
ii. Compute the power against a true mean of 503 g.
iii. Compute the power against a true mean of 505 g.
iv. How large a sample would give 90% power against a 3 g drift?

**QB 4.27** Two studies test the same hypothesis. Study A has 40 observations,
Study B has 400. Both report $p = 0.04$.

i. Which provides stronger evidence that the effect is *large*?
ii. Which had more power against a given effect size?
iii. Explain why these two questions have different answers.

## Looking ahead {#os-looking-ahead}

We have now completed the basic toolkit of statistical inference for a single
mean. Given one sample, we can

- estimate an unknown population mean,
- quantify the uncertainty around that estimate with a confidence interval, and
- test a specific claim about the population with a hypothesis test.

The last two, as Section \@ref(os-ci) showed, are the same object seen from two
sides.

Every test in this unit compared a sample against a *claimed* number — a
figure from a government report, a manufacturer's specification, a previous
year's estimate. That number had to come from somewhere outside the data.

Most questions worth asking do not arrive in that form.

Did the households that received the transfer end up better off than those
that did not? Do girls in the scheme attend school more than girls outside it?
Did the districts that got the road see wages rise more than the districts
that did not? In each case there is no external claim to test against. There
are two groups, each with its own sample mean, and the question is whether the
difference between them is larger than sampling variation alone would produce.

That is the subject of the next unit. It is also where the difficulty changes
character. Deciding whether two groups differ is a statistical problem, and we
will solve it. Deciding whether one group differs *because* of the treatment is
a question about how the groups came to be formed — and no test statistic can
answer it.
