# Hypothesis Testing: Two Samples {#two-sample}

```{r setup-05, include = FALSE}
source("_common.R")
library(ggplot2)
library(BSDA)
```

Every test so far compared a sample against a number that came from outside the
data — a government figure, a manufacturer's specification, a dean's estimate.

Most questions worth asking are not shaped like that. They compare two groups.

Do men and women earn different wages for the same work? Do students from
government schools perform differently from those from private schools? Does
the high-yielding variety of wheat out-yield the conventional one? Did the
villages that received the road see incomes rise more than those that did not?

In each case there is no external claim to test against. There are two groups,
each with its own sample mean, and the question is whether the difference
between them is larger than sampling variation alone would produce.

This unit builds that test. It also brings us to a distinction that will occupy
the rest of the book: the difference between establishing that two groups
differ, and establishing that something *caused* them to differ.

## Comparing two groups {#ts-motivation}

::: {.question}
Two groups, two sample means, and they are not equal. What have we learned?
:::

Suppose a claim is made that two populations have the same mean. Male and
female students at a university consume the same number of cigarettes on
average; conventional and high-yielding wheat produce the same yield per
hectare; two districts have the same average household income.

We could settle it by measuring everyone in both populations. That is expensive
in exactly the way Unit \@ref(estimation) described, and nobody does it.

So we draw a sample from each group and compute both sample means. Unlike the
one-sample case, we now have *two* numbers, each carrying its own sampling
variation.

Suppose they differ. Have we shown the populations differ?

::: {.watchout}
No — and for the same reason as in Unit \@ref(one-sample), only more so.

Each sample mean would miss its population mean even if we had measured
perfectly. Now there are two of them, each free to land high or low. A
difference between two sample means is *guaranteed* to appear even when the two
populations are identical.
:::

So the question is the one we have asked all along, adapted:

> If the two populations really had the same mean, how often would two samples
> come out as far apart as ours did?

If that would happen often, we have learned nothing about the populations. If it
would almost never happen, then either something unusual happened to us or the
two populations are not the same.

Everything that follows is machinery for answering that question. And as
before, the machinery needs a sampling distribution — this time of a
*difference*.

## What does a difference of means do across samples? {#ts-sampling}

::: {.question}
We know how $\bar{X}$ behaves. How does $\bar{X} - \bar{Y}$ behave?
:::

Take a sample $x_1, \ldots, x_n$ from a population with mean $\mu_x$ and
standard deviation $\sigma_x$, and an independent sample $y_1, \ldots, y_m$
from a population with mean $\mu_y$ and standard deviation $\sigma_y$. The
sample sizes need not be equal.

The null hypothesis is that the two populations have the same mean:

$$H_0 : \mu_x = \mu_y \qquad H_A : \mu_x \neq \mu_y$$

It is more useful to write the same thing as a statement about the difference:

$$H_0 : \mu_x - \mu_y = 0$$

because that turns a comparison of two quantities into a claim about one, and
we already know how to test a claim about one quantity.

The quantity we will study is therefore $\bar{X} - \bar{Y}$. Two things need
establishing: where its distribution sits, and how wide it is.

::: {.theory}
**Where it sits.** Expectation passes through a difference:

$$E[\bar{X} - \bar{Y}] = E[\bar{X}] - E[\bar{Y}] = \mu_x - \mu_y$$

So the difference of sample means is an unbiased estimator of the difference of
population means. Under $H_0$ it is centred on zero.

**How wide it is.** Variance does *not* pass through a difference in the same
way. For independent quantities,

$$
\begin{aligned}
\mathrm{Var}[\bar{X} - \bar{Y}] &= \mathrm{Var}[\bar{X}] + \mathrm{Var}[-\bar{Y}] \\[4pt]
&= \mathrm{Var}[\bar{X}] + (-1)^2\mathrm{Var}[\bar{Y}] \\[4pt]
&= \frac{\sigma_x^2}{n} + \frac{\sigma_y^2}{m}
\end{aligned}
$$

using $\mathrm{Var}(aX) = a^2\mathrm{Var}(X)$ from Unit \@ref(estimation), with
$a = -1$. Taking the square root gives the standard error of the difference:

$$\mathrm{se}(\bar{X} - \bar{Y}) = \sqrt{\frac{\sigma_x^2}{n} + \frac{\sigma_y^2}{m}}$$

Independence is needed for the first line, exactly as it was in Unit
\@ref(estimation). It is what allows the variance of a sum to be the sum of the
variances.
:::

::: {.keyidea}
**The variances add, even though the means subtract.**

This is the single most important line in the unit. Squaring $-1$ gives $+1$, so
subtracting a second noisy quantity makes the result *more* variable, not less.

Comparing two groups is therefore harder than measuring one. Both estimates
wobble, and both wobbles end up in the answer.
:::

That has a practical consequence worth seeing now rather than later. Suppose two
groups each have $\sigma = 10$ and each is sampled $n = 100$ times. The standard
error of either mean alone is $10/\sqrt{100} = 1$. The standard error of their
difference is

$$\sqrt{\frac{100}{100} + \frac{100}{100}} = \sqrt{2} \approx 1.41$$

about 40% larger. Detecting a difference between two groups needs more data
than estimating either group's mean to the same precision.

## Testing the difference when σ is known {#ts-known}

::: {.question}
We have a sampling distribution. What is the test?
:::

The argument is the one from Section \@ref(os-critical), with the new standard
error in place of the old one. We reject when the observed difference is too
large:

$$|\bar{x} - \bar{y}| \geq c$$

and we choose $c$ so that, if the null were true, a gap that large would arise
with probability only $\alpha$.

::: {.theory}
Standardising the difference gives a quantity with no unknowns in it:

$$Z = \frac{(\bar{X} - \bar{Y}) - (\mu_x - \mu_y)}{\sqrt{\dfrac{\sigma_x^2}{n} + \dfrac{\sigma_y^2}{m}}} \;\sim\; N(0,1)$$

Under $H_0$ the term $(\mu_x - \mu_y)$ is zero and drops out. Setting

$$P\big[\,|\bar{X} - \bar{Y}| \geq c\,\big] = \alpha$$

and dividing through by the standard error,

$$P\left[\,|Z| \geq \frac{c}{\sqrt{\sigma_x^2/n + \sigma_y^2/m}}\,\right] = \alpha$$

The two tails carry $\alpha/2$ each, so the bracketed quantity is
$Z_{\alpha/2}$, giving

$$c = Z_{\alpha/2}\sqrt{\frac{\sigma_x^2}{n} + \frac{\sigma_y^2}{m}}$$
:::

::: {.keyidea}
The test statistic and the interval for the difference are

$$Z_{calc} = \frac{\bar{x} - \bar{y}}{\sqrt{\dfrac{\sigma_x^2}{n} + \dfrac{\sigma_y^2}{m}}}
\qquad\qquad
(\bar{x} - \bar{y}) \pm Z_{\alpha/2}\sqrt{\frac{\sigma_x^2}{n} + \frac{\sigma_y^2}{m}}$$

Reject $H_0$ when $|Z_{calc}| > Z_{\alpha/2}$, which happens exactly when the
interval excludes zero — the same equivalence as Section \@ref(os-ci), now
about a difference rather than a mean.
:::

### Conventional against high-yielding wheat {-}

Twelve plots are sown with conventional seed and fourteen with a high-yielding
variety. The conventional plots average 45.2 kg per hectare and the HYV plots
48.6 kg. From long agronomic experience the standard deviations are known:
$\sigma_x = 0.8$ for conventional and $\sigma_y = 1.0$ for HYV.

Are these data consistent, at the 5% level, with the two varieties yielding the
same on average?

```{r ts-wheat}
xbar <- 45.2; sx <- 0.8; n <- 12      # conventional
ybar <- 48.6; sy <- 1.0; m <- 14      # high-yielding

se_diff <- sqrt(sx^2 / n + sy^2 / m)
z_calc  <- (xbar - ybar) / se_diff

c(difference = xbar - ybar,
  se         = se_diff,
  z_calc     = z_calc,
  p_value    = 2 * pnorm(-abs(z_calc)))
```

$Z_{calc} = -9.63$, far beyond $-1.96$, and the $p$-value is smaller than any
conventional threshold. We reject the claim that the two varieties yield the
same.

The interval says how much the varieties differ, which is the more useful
statement:

```{r ts-wheat-ci}
(xbar - ybar) + c(-1, 1) * qnorm(0.975) * se_diff
```

The HYV seed yields between 2.7 and 4.1 kg per hectare more than the
conventional variety, with 95% confidence. Note that the interval lies entirely
below zero and so excludes it — the same verdict the test reached.

::: {.intuition}
Report the interval, not just the rejection. "The varieties differ
significantly" is a much weaker statement than "the HYV variety yields 2.7 to
4.1 kg per hectare more," and a farmer deciding what to sow needs the second.
:::

## When σ is unknown {#ts-unknown}

::: {.question}
As always, we do not know $\sigma_x$ or $\sigma_y$. Now there are two of them.
:::

In practice both population standard deviations must be estimated from the
samples, using $s_x$ and $s_y$. As in Section \@ref(os-ttest), that costs
something, and what it costs depends on the situation. Three cases arise, and
they differ only in what goes underneath the square root.

### Large samples {-}

If both samples are reasonably large — a common rule of thumb is at least 20
each — then $s_x$ and $s_y$ are reliable enough that substituting them changes
little, and the normal distribution remains a good approximation:

$$Z_{calc} = \frac{\bar{x} - \bar{y}}{\sqrt{\dfrac{s_x^2}{n} + \dfrac{s_y^2}{m}}}$$

This is the same statistic with hats on the standard deviations.

### Equal variances: the pooled estimator {-}

If we are willing to assume the two populations share a common standard
deviation, $\sigma_x = \sigma_y = \sigma$, then both samples are estimating the
*same* quantity and it is wasteful to estimate it twice. We combine them.

::: {.keyidea}
The **pooled variance estimator** is a weighted average of the two sample
variances, each weighted by its degrees of freedom:

$$s_p^2 = \frac{(n-1)s_x^2 + (m-1)s_y^2}{n + m - 2}$$

and the test statistic becomes

$$t_{calc} = \frac{\bar{x} - \bar{y}}{\sqrt{s_p^2\left(\dfrac{1}{n} + \dfrac{1}{m}\right)}}$$

compared against $t$ with $n + m - 2$ degrees of freedom.
:::

The degrees of freedom are worth pausing on. We began with $n + m$
observations, and spent one estimating each of the two sample means, leaving
$n + m - 2$. The same accounting as Section \@ref(u3-variance), applied twice.

Pooling buys precision — $n + m - 2$ degrees of freedom rather than something
closer to the smaller sample's — but only if the assumption holds. If one group
really is more variable than the other, the pooled estimate is a weighted
average of two different things, and the test is wrong in a way the output will
not reveal.

### Unequal variances: Welch's test {-}

The safer option makes no assumption about the variances. It uses the separate
estimates,

$$t_{calc} = \frac{\bar{x} - \bar{y}}{\sqrt{\dfrac{s_x^2}{n} + \dfrac{s_y^2}{m}}}$$

and adjusts the degrees of freedom downwards to compensate for the extra
uncertainty. The adjustment is not a whole number and is not worth memorising;
R computes it.

::: {.watchout}
**R's `t.test()` uses Welch by default.** If you have derived a pooled
statistic by hand and R disagrees with you, this is usually why.

To pool, pass `var.equal = TRUE` explicitly.

Modern practice is to prefer Welch. It costs very little when the variances
really are equal, and it protects you when they are not — and you can never
verify equal variances, only fail to reject the claim that they are equal.
:::

### Do North and South Indian students differ in height? {-}

`student-survey-height.csv` records 100 students, each classified by region.

```{r ts-height-desc}
heights <- read.csv("data/student-survey-height.csv")

aggregate(height ~ region, data = heights,
          function(x) round(c(n = length(x), mean = mean(x), sd = sd(x)), 2))
```

The two sample means differ by about 1.5 cm, and the two sample standard
deviations are not equal — 7.3 against 9.9. That is a case for Welch.

```{r ts-height-test}
t.test(height ~ region, data = heights)
```

$t = 0.83$ on 74.4 degrees of freedom, with $p = 0.41$. We fail to reject. The
95% interval for the difference runs from $-2.07$ to $5.04$ cm, and it
**contains zero** — the two statements agree, as they must.

So a 1.5 cm gap in the samples is entirely unremarkable. With these sample
sizes and this much variation within each group, differences of that size turn
up constantly when the populations are identical.

::: {.deeper}
Compare the degrees of freedom the two approaches give here.

```{r ts-df-compare}
c(welch  = t.test(height ~ region, data = heights)$parameter,
  pooled = t.test(height ~ region, data = heights, var.equal = TRUE)$parameter)
```

Pooling claims 98 degrees of freedom; Welch allows only 74.4. The difference is
the price of not assuming the two groups are equally variable — and here, where
one sample's standard deviation is a third larger than the other's, it is a
price worth paying.

Both give the same verdict in this case. They do not always.
:::

## When two samples are not two samples {#ts-paired}

::: {.question}
Ten couples. Two columns of ages. Is that two samples?
:::

`marriage-ages.csv` holds the ages of both partners in ten married couples.

```{r ts-marriage}
couples <- read.csv("data/marriage-ages.csv")
couples
```

The question is whether partner 1 is on average older than partner 2. There are
two columns of ten numbers, so it is tempting to run the test we have just
learned.

```{r ts-marriage-wrong}
t.test(couples$partner1, couples$partner2)
```

Nothing at all: $t = 0.52$, $p = 0.61$. On this evidence the two are the same
age on average.

Now look at the data again. In eight of the ten couples partner 1 is the older
one, and in the remaining two the partners are exactly the same age. Not once is
partner 2 older. A procedure that looks at that and reports $p = 0.61$ has
thrown something away.

What it threw away is the pairing. The two-sample test treats the twenty numbers
as two unrelated heaps, and asks whether one heap averages higher than the
other. But the ages within a couple are strongly related — couples marry at
similar ages to each other, and the couples differ hugely among themselves,
from the mid-thirties to the seventies. That between-couple variation is
enormous, it is identical in both columns, and it swamps the small, consistent
within-couple gap.

Seen properly, the structure is obvious.

```{r ts-marriage-plot, echo = FALSE, fig.height = 3.8, fig.cap = "Ten couples, ordered by the age of partner 2. Each segment joins the two partners in one couple. The couples span nearly forty years between them; within any couple the gap is at most eight. Two couples are the same age, and show as a red ring around a blue centre."}
cp <- couples[order(couples$partner2), ]
cp$id <- seq_len(nrow(cp))

# Partner 1 is drawn first and larger: in the two couples of equal age the two
# points coincide exactly, and this leaves a visible ring rather than one dot
# silently covering the other.
long <- rbind(
  data.frame(id = cp$id, age = cp$partner1, who = "Partner 1", sz = 3.6),
  data.frame(id = cp$id, age = cp$partner2, who = "Partner 2", sz = 1.9)
)

ggplot(long, aes(age, id)) +
  geom_segment(data = cp, aes(x = partner2, xend = partner1, y = id, yend = id),
               colour = book_palette$soft, linewidth = 1.6,
               inherit.aes = FALSE) +
  geom_point(aes(colour = who, size = sz)) +
  scale_size_identity() +
  scale_colour_manual(values = c("Partner 1" = book_palette$reject,
                                 "Partner 2" = book_palette$accent),
                      name = NULL) +
  scale_y_continuous(breaks = NULL) +
  labs(x = "Age (years)", y = "Couple") +
  theme_book() +
  theme(legend.position = "top", panel.grid.major.y = element_blank())
```

The figure explains both results at once. Vertically, the couples are spread
across nearly forty years — that is the variation the unpaired test measures the
gap against, and it is overwhelming. Horizontally, within each couple, the red
point is at or to the right of the blue one every single time — in two cases
exactly on top of it — and never more than eight years away. That small, utterly consistent gap is what the paired
test measures.

The fix, then, is to compare each partner with *their own* partner, by working
with the differences.

```{r ts-marriage-diff}
d <- couples$partner1 - couples$partner2
d

c(mean_difference = mean(d), sd = sd(d), n = length(d))
```

Every difference is zero or positive. Now the between-couple variation has
cancelled out entirely — it was common to both columns, so subtracting removes
it — and what remains is the quantity of interest.

::: {.keyidea}
A **paired** design produces one number per unit, not two samples. Take the
differences and run a *one-sample* test on them, against $H_0: \mu_d = 0$.

In R, `paired = TRUE`.
:::

```{r ts-marriage-paired}
t.test(couples$partner1, couples$partner2, paired = TRUE)
```

$t = 3.0$ on 9 degrees of freedom, $p = 0.015$. Partner 1 is older by between
0.6 and 4.4 years, with 95% confidence.

::: {.watchout}
The same ten couples gave $p = 0.61$ and $p = 0.015$. The data did not change.
Only whether the analysis respected the design.

Pairing arises whenever the two measurements belong to the same unit: before and
after a training programme on the same workers, the same plots under two
treatments in successive seasons, the same households surveyed in two rounds. In
every such case, treating the columns as independent samples discards the
comparison the design was built to make.
:::

## A difference is not an effect {#ts-causal}

::: {.question}
The test says two groups differ. Does it say why?
:::

We can now establish that two groups differ by more than sampling variation.
That is a genuine achievement, and it is not what most interesting questions
actually ask.

Consider the attendance data from this course. Split the 680 students at the
median attendance and compare their final examination scores.

```{r ts-attend}
students <- read.csv("data/attendance-grades.csv")
students$group <- ifelse(students$attendance >= median(students$attendance),
                         "High attendance", "Low attendance")

aggregate(final_score ~ group, data = students,
          function(x) round(c(n = length(x), mean = mean(x)), 2))

t.test(final_score ~ group, data = students)
```

Students who attended more scored higher, and $p = 0.011$ rejects the claim
that the two groups have the same mean.

It is tempting to conclude that attending class raises examination scores by
about a point. Two separate problems stand in the way.

### The difference is significant and trivial {-}

The interval for the difference runs from about 0.21 to 1.59 marks out of 100.

A gap that small would not change anyone's grade, and would not justify any
policy. Yet it is "statistically significant", because 680 students is a large
enough sample to detect very small differences reliably.

::: {.keyidea}
Statistical significance answers "is this difference bigger than noise?"

It does not answer "is this difference big enough to matter?" That is a question
about the subject, not about the statistics, and no $p$-value speaks to it.

Always read the interval, in the units of the problem, before deciding whether
a result is important.
:::

### The groups were not assigned {-}

The deeper problem is that nobody allocated students to high and low
attendance. They chose, and the reasons they chose are not in the dataset.

A student who attends regularly may be more motivated, better prepared, less
burdened by paid work or family obligations, or living closer to campus. Any of
those could raise examination scores on its own. The two groups differ in
attendance *and* in everything that attendance is correlated with, and the test
cannot separate them.

::: {.watchout}
The two-sample test compares two groups as they arrive. It has nothing to say
about *why* they came to be different.

If the groups formed themselves, a difference in outcomes may reflect the
treatment, or the reasons for selection, or both — and the arithmetic is
identical either way. This is the problem of **selection**, and it is the
central obstacle in all of empirical economics.
:::

It is worth being precise about what has and has not been shown here.

We have shown that students who attended more scored slightly higher. That is a
fact about the world and the test establishes it properly. We have not shown
that attending more *causes* higher scores, and no amount of additional care
with the $t$ statistic would help.

What would help is a different design. If attendance had been assigned by
something unrelated to the students themselves — a lottery, a timetable clash,
a bus route — then the groups would be comparable in expectation and the
difference could be read as an effect. Creating such comparisons, or finding
them in the world, is the subject of the second part of this book.

::: {.intuition}
The question "do these groups differ?" is a statistical one, and this unit
answers it.

The question "does this treatment work?" is a question about how the groups were
formed. It is answered by the design of the study, not by the choice of test.
:::

## Summary {#ts-summary}

::: {.keyidea}
- Two groups produce two sample means, and a difference between them is
  guaranteed to appear even when the populations are identical. The question is
  whether it is larger than sampling variation would produce.
- $E[\bar{X} - \bar{Y}] = \mu_x - \mu_y$: the difference of sample means is
  unbiased for the difference of population means.
- **The variances add even though the means subtract**, so
  $\mathrm{se}(\bar{X} - \bar{Y}) = \sqrt{\sigma_x^2/n + \sigma_y^2/m}$.
  Comparing two groups is noisier than measuring one, and needs more data.
- With $\sigma$ known the test is a $Z$-test on the difference; the interval is
  $(\bar{x} - \bar{y}) \pm Z_{\alpha/2}\,\mathrm{se}$, and it excludes zero
  exactly when the test rejects.
- With $\sigma$ unknown, three cases differ only under the square root: large
  samples use $s_x$ and $s_y$ with the normal; **pooling** assumes equal
  variances and buys $n + m - 2$ degrees of freedom; **Welch** assumes nothing
  and adjusts the degrees of freedom down. R's `t.test()` uses Welch unless you
  pass `var.equal = TRUE`, and Welch is the safer default.
- A **paired** design gives one number per unit. Take differences and run a
  one-sample test. Treating paired data as two independent samples discards the
  comparison the design was built to make — in the marriage data it turned
  $p = 0.015$ into $p = 0.61$.
- A significant difference need not be an important one. Read the interval in
  the units of the problem.
- A difference between two groups that formed themselves is not an effect. The
  test compares the groups as they arrive and is silent about why they differ.
:::

## Practice problems {#ts-practice}

::: {.yourturn}
Work each one before opening the solution. The
[question bank](#ts-question-bank) that follows has no answers.
:::

### Concept checks {-}

**5.1** Why is the standard error of a difference of means larger than the
standard error of either mean on its own?

::: {.answer}
Because the two sample means each carry sampling variation, and both end up in
the difference. Formally, $\mathrm{Var}(aX) = a^2\mathrm{Var}(X)$ with $a = -1$
gives $(-1)^2 = +1$, so the variances add:

$$\mathrm{Var}[\bar{X} - \bar{Y}] = \frac{\sigma_x^2}{n} + \frac{\sigma_y^2}{m}$$

Subtracting a noisy quantity does not cancel noise; it adds more. This is why
detecting a difference between two groups takes more data than estimating one
group's mean to the same precision.
:::

**5.2** A study reports that two groups differ with $p = 0.003$. A colleague
concludes the difference is large. What is wrong?

::: {.answer}
The $p$-value measures how incompatible the data are with the claim of no
difference. It says nothing about the size of the difference.

A very small $p$-value can come from a tiny difference measured in a large
sample. The size of the difference is reported by the confidence interval, in
the units of the problem, and that is what has to be inspected before calling a
result large or important.
:::

**5.3** Why does R's `t.test()` report fractional degrees of freedom?

::: {.answer}
Because it performs Welch's test by default, which does not assume the two
populations have equal variances. The correction for that extra uncertainty
adjusts the degrees of freedom downwards, and the adjustment is generally not a
whole number.

The pooled test, obtained with `var.equal = TRUE`, uses $n + m - 2$ degrees of
freedom, which always is.
:::

**5.4** The same twenty households are surveyed before and after a cash
transfer. Why is a two-sample test the wrong tool?

::: {.answer}
Because the two columns are not independent samples — each household appears in
both. The design produces one number per household, the change.

Households differ enormously among themselves, and that variation is present
identically in both columns. A two-sample test treats it as noise, and it
swamps the change we are trying to detect. Taking differences within households
removes it, leaving only the quantity of interest.
:::

**5.5** Villages that received a road show higher incomes than villages that
did not. State two distinct reasons this need not mean the road raised incomes.

::: {.answer}
First, the villages may have differed beforehand. Roads are not built at random
— they tend to reach places that are more accessible, more populous, more
politically connected or already better off. Those same features raise incomes
on their own.

Second, the two groups may differ in things that changed alongside the road:
electrification, a market, a school. The road is correlated with a bundle of
improvements, and a comparison of means attributes the whole bundle to it.

In both cases the arithmetic is correct and the causal reading is not.
:::

### Explain {-}

**5.6** Explain the trade-off between the pooled and Welch tests, and why
modern practice prefers Welch.

::: {.answer}
Pooling assumes the two populations share a standard deviation. If true, both
samples estimate the same quantity, they can be combined, and the resulting
test has $n + m - 2$ degrees of freedom — more than Welch allows, so more
power.

If false, the pooled estimate averages two different quantities and the test is
wrong in a way nothing in the output reveals. Welch avoids the assumption
entirely, at the cost of fewer degrees of freedom.

The reason to prefer Welch is asymmetric risk. When the variances really are
equal, Welch loses very little. When they are not, pooling misleads. And the
assumption cannot be verified — a test of equal variances can only fail to
reject, which is not the same as establishing it.
:::

**5.7** A researcher tests whether men and women in a firm are paid differently,
finds $p = 0.02$, and concludes the firm discriminates. Evaluate.

::: {.answer}
The test establishes that average pay differs by more than sampling variation
would explain. That is not the same as discrimination.

Men and women in the firm may differ in role, seniority, years of experience,
hours, or field of training, and any of those could account for a pay gap
without anyone being treated differently for the same work. The comparison of
raw means attributes the whole difference to sex because sex is the only
variable in the analysis.

Establishing discrimination requires comparing like with like — people doing
the same work with the same experience — which is a question of design and, in
later units, of holding other variables constant. The two-sample test is not
capable of it.
:::

### Numerical problems {-}

**5.8** Two irrigation methods are trialled. Drip irrigation on 15 plots yields
a mean of 38.4 quintals per hectare; flood irrigation on 18 plots yields 35.9.
Population standard deviations are known to be 3.2 and 4.1 respectively.

i. Test at the 5% level whether the mean yields differ.
ii. Construct a 95% confidence interval for the difference.
iii. Does the interval agree with the test?

::: {.answer}
**(i)** The standard error of the difference is

$$\sqrt{\frac{3.2^2}{15} + \frac{4.1^2}{18}} = \sqrt{0.6827 + 0.9339}
= \sqrt{1.6166} = 1.271$$

$$Z_{calc} = \frac{38.4 - 35.9}{1.271} = \frac{2.5}{1.271} = 1.967$$

Since $1.967 > 1.96$, we reject $H_0$ at the 5% level — but only just.

**(ii)** $2.5 \pm 1.96(1.271) = 2.5 \pm 2.491 = [0.009,\ 4.991]$.

**(iii)** Yes. The interval excludes zero, and the test rejects. Both do so by
the narrowest possible margin, which is the same fact stated twice: the interval
endpoint sits at 0.009 because $Z_{calc}$ sits at 1.967.

A result this marginal should be reported as marginal. Nothing about it
justifies confidence that drip irrigation is better.
:::

**5.9** Two samples give $n = 10$, $s_x = 4$ and $m = 14$, $s_y = 6$.

i. Compute the pooled variance estimate.
ii. How many degrees of freedom does the pooled test use?
iii. Why is it not $n + m$?

::: {.answer}
**(i)**

$$s_p^2 = \frac{(10-1)16 + (14-1)36}{10 + 14 - 2}
= \frac{144 + 468}{22} = \frac{612}{22} = 27.82$$

so $s_p = 5.27$, lying between 4 and 6 and closer to 6 because the larger
sample carries more weight.

**(ii)** $n + m - 2 = 22$.

**(iii)** Because two quantities were estimated from the data before the spread
could be measured — one sample mean in each group. Each costs a degree of
freedom, exactly as in Section \@ref(u3-variance), where estimating a single
mean cost one.
:::

**5.10** A programme is evaluated by comparing 200 participants with 200
non-participants. Participants' mean income is ₹4,200 higher, with a standard
error of ₹1,400 for the difference.

i. Test whether the difference is zero at the 5% level.
ii. Construct a 95% interval.
iii. The programme was open to anyone who applied. What can be concluded about
     its effect?

::: {.answer}
**(i)** $Z_{calc} = 4200/1400 = 3.0$, which exceeds 1.96. We reject the claim
that the two groups have the same mean income.

**(ii)** $4200 \pm 1.96(1400) = [₹1{,}456,\ ₹6{,}944]$.

**(iii)** Nothing about its effect. Because the programme was open to
applicants, participants selected themselves. People who apply to income
programmes may be more motivated, better informed, or better connected — or,
equally plausibly, more desperate and worse off to begin with. The direction of
the bias is not even clear.

The interval is a correct statement about the difference between two groups of
people. It is not an estimate of what the programme did to anybody.
:::

### R exercises {-}

**5.11** Using `data/student-survey-sleep.csv`, test whether urban and rural
students report different average hours of sleep. Report the interval and say
what you conclude.

::: {.answer}
```{r ts-ex-sleep}
sleep <- read.csv("data/student-survey-sleep.csv")

aggregate(sleep ~ residence, data = sleep,
          function(x) round(c(n = length(x), mean = mean(x), sd = sd(x)), 2))

t.test(sleep ~ residence, data = sleep)
```

The two group means are close and the interval for the difference contains
zero, so we fail to reject. There is no evidence here that urban and rural
students sleep different amounts on average.

Note what "no evidence of a difference" is not: evidence of no difference. With
these sample sizes, a real difference of well under an hour would be difficult
to detect at all, and the interval says as much by being wide.
:::

**5.12** Test whether social media use differs between North and South Indian
students in the same dataset, using both Welch and the pooled test. Do they
agree, and which would you report?

::: {.answer}
```{r ts-ex-pooled}
rbind(
  welch  = unlist(t.test(socialmediahours ~ region, data = sleep)[c("statistic", "parameter", "p.value")]),
  pooled = unlist(t.test(socialmediahours ~ region, data = sleep, var.equal = TRUE)[c("statistic", "parameter", "p.value")])
)
```

The two agree closely here, which is what happens when the sample standard
deviations are similar — the pooled and separate estimates of the variance are
then nearly the same number.

Report Welch. It gives the same answer when the assumption holds and a correct
one when it does not, so there is no situation in which pooling is the safer
choice.
:::

**5.13** Take the wage data as a population and draw independent samples of 200
urban and 200 rural workers. Test whether mean annual earnings differ, and
report the interval. Then say what the interval does and does not establish.

::: {.answer}
```{r ts-ex-wages}
wages <- read.csv("data/wages-india-synthetic.csv")

set.seed(9)
urban <- sample(wages$annual_earnings[wages$residence == "Urban"], 200)
rural <- sample(wages$annual_earnings[wages$residence == "Rural"], 200)

t.test(urban, rural)
```

The difference is large and the interval is nowhere near zero, so urban workers
in this population earn substantially more than rural workers.

What the interval establishes is the size of that gap. What it does not
establish is that moving a worker from a village to a city would raise their
earnings by that amount. Urban and rural workers differ in education,
occupation and industry, and the sample says nothing about which part of the gap
belongs to the place and which to the people.
:::

**5.14** Simulate the point that variances add. Draw two independent samples of
100 from a population with $\sigma = 10$, and compare the observed standard
deviation of $\bar{x}$ with that of $\bar{x} - \bar{y}$ across 5,000
repetitions. Check both against theory.

::: {.answer}
```{r ts-ex-varadd}
set.seed(4)
sims <- replicate(5000, {
  x <- rnorm(100, mean = 50, sd = 10)
  y <- rnorm(100, mean = 50, sd = 10)
  c(one_mean = mean(x), difference = mean(x) - mean(y))
})

round(c(sd_of_one_mean       = sd(sims["one_mean", ]),
        theory_one_mean      = 10 / sqrt(100),
        sd_of_difference     = sd(sims["difference", ]),
        theory_difference    = sqrt(10^2/100 + 10^2/100)), 4)
```

The difference is about $\sqrt{2}$ times as variable as a single mean, exactly
as $\sqrt{\sigma^2/n + \sigma^2/m}$ predicts. Note also that the difference is
centred on zero: both populations have mean 50, and the estimator is unbiased
for a difference of zero.
:::

**5.15** Demonstrate the cost of ignoring pairing. Using `marriage-ages.csv`,
report the paired and unpaired $p$-values, and explain the gap by comparing the
standard deviation of the differences with the standard deviations of the two
columns.

::: {.answer}
```{r ts-ex-paired}
couples <- read.csv("data/marriage-ages.csv")

c(paired   = t.test(couples$partner1, couples$partner2, paired = TRUE)$p.value,
  unpaired = t.test(couples$partner1, couples$partner2)$p.value)

round(c(sd_partner1  = sd(couples$partner1),
        sd_partner2  = sd(couples$partner2),
        sd_of_diffs  = sd(couples$partner1 - couples$partner2)), 3)
```

The two columns each have a standard deviation of about 11 years, because the
couples range from their mid-thirties to their seventies. The differences have a
standard deviation of about 2.6 years.

That is the whole explanation. The unpaired test measures the small gap against
the huge between-couple variation and sees nothing. The paired test removes that
variation — it is common to both columns and cancels on subtraction — and
measures the gap against what is left.
:::

## Question bank {#ts-question-bank}

::: {.watchout}
**No answers are given for this section, by design.**
:::

### Multiple choice {-}

**QB 5.1** The standard error of $\bar{X} - \bar{Y}$ for independent samples is

a. $\dfrac{\sigma_x}{\sqrt{n}} - \dfrac{\sigma_y}{\sqrt{m}}$
b. $\sqrt{\dfrac{\sigma_x^2}{n} + \dfrac{\sigma_y^2}{m}}$
c. $\sqrt{\dfrac{\sigma_x^2}{n} - \dfrac{\sigma_y^2}{m}}$
d. $\dfrac{\sigma_x + \sigma_y}{\sqrt{n + m}}$

**QB 5.2** The pooled two-sample $t$-test uses how many degrees of freedom?

a. $n + m$
b. $n + m - 1$
c. $n + m - 2$
d. $\min(n, m) - 1$

**QB 5.3** A 95% confidence interval for $\mu_x - \mu_y$ is $[-1.2,\ 4.8]$. At
the 5% level we

a. reject $H_0: \mu_x = \mu_y$
b. fail to reject $H_0: \mu_x = \mu_y$
c. cannot tell without the $p$-value
d. conclude the two means are equal

**QB 5.4** `t.test(x, y)` in R by default

a. assumes equal variances
b. performs Welch's test
c. performs a paired test
d. performs a one-tailed test

**QB 5.5** Which of the following are paired designs? *(select all that apply)*

a. Blood pressure of 40 patients before and after a drug
b. Test scores of 40 boys and 45 girls
c. Yields on 20 plots under one seed and 20 different plots under another
d. Household expenditure in the same 200 households in 2019 and 2023

**QB 5.6** Ignoring pairing in a genuinely paired design usually

a. makes it easier to reject the null
b. makes it harder to reject the null
c. has no effect on the $p$-value
d. reverses the sign of the difference

**QB 5.7** Two groups differ with $p < 0.001$ in a sample of 50,000. This
implies

a. the difference is large
b. the difference is important
c. the difference is unlikely to be sampling variation alone
d. one group caused the other's outcome

### Short reasoning {-}

**QB 5.8** Explain why subtracting one sample mean from another increases
variability rather than reducing it, and give the algebraic reason.

**QB 5.9** A researcher has two samples with $s_x = 2$ and $s_y = 9$, and uses
the pooled test. State the assumption being made, why it is doubtful here, and
what the consequence is.

**QB 5.10** A newspaper reports that graduates of a coaching institute score 40
marks higher on average than non-attendees, with a very small $p$-value.
Explain what the comparison can and cannot support.

**QB 5.11** In a paired design, the two columns are strongly positively
correlated. Explain, without algebra, why this makes the paired test more
powerful than the unpaired one.

**QB 5.12** A trial assigns a new fertiliser to plots by lottery. Explain why
the difference in mean yields can be read as an effect here, when the same
comparison in Section \@ref(ts-causal) could not.

**QB 5.13** Two studies estimate the same difference. One reports
$[0.1,\ 9.9]$ and the other $[4.8,\ 5.2]$. Both exclude zero. Explain what
distinguishes them and which is more useful.

### Numerical {-}

**QB 5.14** Government school students score a mean of 62.4 with $s = 11.2$ in a
sample of 45. Private school students score 66.1 with $s = 9.8$ in a sample of
52.

i. Compute the standard error of the difference.
ii. Compute the test statistic.
iii. Test at the 5% level whether the means differ.
iv. Construct a 95% confidence interval, and state whether it agrees with (iii).

**QB 5.15** A sample of 8 plots under a new seed yields a mean of 41.2 quintals
with $s = 3.1$. A sample of 10 plots under the old seed yields 37.8 with
$s = 2.9$.

i. Compute the pooled variance estimate and $s_p$.
ii. Compute the pooled test statistic and its degrees of freedom.
iii. Test at the 5% level whether the new seed yields more, stating your
     hypotheses carefully.
iv. Would Welch's test use more or fewer degrees of freedom here, and why?

**QB 5.16** Twelve workers are timed on a task before and after training. The
mean reduction is 4.2 minutes with a standard deviation of differences of 5.1
minutes.

i. State the hypotheses for a test that training reduced completion time.
ii. Compute the test statistic and its degrees of freedom.
iii. Test at the 5% level.
iv. Explain why the standard deviation of the *differences* is the right
    quantity here rather than the standard deviations of the two rounds.

**QB 5.17** Using `data/attendance-grades.csv`, write R code to compare
cumulative GPA between students above and below the median attendance. Report
the interval, then write a short paragraph stating precisely what your result
establishes and what it does not.

## Looking ahead {#ts-looking-ahead}

We can now compare two groups properly. We can also say exactly what such a
comparison does not deliver, which turns out to be the more important lesson.

Two limitations remain, and both point the same way.

The first is that comparing two groups uses the data wastefully. Splitting the
680 students at the median attendance threw away everything we knew about *how
much* each student attended, reducing a rich variable to a coin flip. A student
who attended 32 classes and one who attended 20 were treated as identical.

The second is selection. The attendance groups differed in attendance and in
everything correlated with it, and a two-sample test cannot separate them
because it knows about nothing else.

Both problems have the same solution. Instead of splitting a variable in two
and comparing averages, we can model the relationship between the whole of one
variable and the whole of another — and, crucially, do so while holding other
variables fixed.

That is regression, and it occupies the rest of this book.
