# Hypothesis Testing: Two Samples {#two-sample}

```{r setup-05, include = FALSE}
source("_common.R")
library(ggplot2)
library(BSDA)
```

Every test so far compared a sample against a number that came from outside the
data — a government figure, a manufacturer's specification, a dean's estimate.

Most questions worth asking are not shaped like that. They compare two groups.

Do men and women earn different wages for the same work? Do students from
government schools perform differently from those from private schools? Does
the high-yielding variety of wheat out-yield the conventional one? Did the
villages that received the road see incomes rise more than those that did not?

In each case there is no external claim to test against. There are two groups,
each with its own sample mean, and the question is whether the difference
between them is larger than sampling variation alone would produce.

This unit builds that test. It also brings us to a distinction that will occupy
the rest of the book: the difference between establishing that two groups
differ, and establishing that something *caused* them to differ.

## Comparing two groups {#ts-motivation}

::: {.question}
Two groups, two sample means, and they are not equal. What have we learned?
:::

Suppose a claim is made that two populations have the same mean. Male and
female students at a university consume the same number of cigarettes on
average; conventional and high-yielding wheat produce the same yield per
hectare; two districts have the same average household income.

We could settle it by measuring everyone in both populations. That is expensive
in exactly the way Unit \@ref(estimation) described, and nobody does it.

So we draw a sample from each group and compute both sample means. Unlike the
one-sample case, we now have *two* numbers, each carrying its own sampling
variation.

Suppose they differ. Have we shown the populations differ?

::: {.watchout}
No — and for the same reason as in Unit \@ref(one-sample), only more so.

Each sample mean would miss its population mean even if we had measured
perfectly. Now there are two of them, each free to land high or low. A
difference between two sample means is *guaranteed* to appear even when the two
populations are identical.
:::

So the question is the one we have asked all along, adapted:

> If the two populations really had the same mean, how often would two samples
> come out as far apart as ours did?

If that would happen often, we have learned nothing about the populations. If it
would almost never happen, then either something unusual happened to us or the
two populations are not the same.

Everything that follows is machinery for answering that question. And as
before, the machinery needs a sampling distribution — this time of a
*difference*.

## What does a difference of means do across samples? {#ts-sampling}

::: {.question}
We know how $\bar{X}$ behaves. How does $\bar{X} - \bar{Y}$ behave?
:::

Take a sample $x_1, \ldots, x_n$ from a population with mean $\mu_x$ and
standard deviation $\sigma_x$, and an independent sample $y_1, \ldots, y_m$
from a population with mean $\mu_y$ and standard deviation $\sigma_y$. The
sample sizes need not be equal.

The null hypothesis is that the two populations have the same mean:

$$H_0 : \mu_x = \mu_y \qquad H_A : \mu_x \neq \mu_y$$

It is more useful to write the same thing as a statement about the difference:

$$H_0 : \mu_x - \mu_y = 0$$

because that turns a comparison of two quantities into a claim about one, and
we already know how to test a claim about one quantity.

The quantity we will study is therefore $\bar{X} - \bar{Y}$. Two things need
establishing: where its distribution sits, and how wide it is.

::: {.theory}
**Where it sits.** Expectation passes through a difference:

$$E[\bar{X} - \bar{Y}] = E[\bar{X}] - E[\bar{Y}] = \mu_x - \mu_y$$

So the difference of sample means is an unbiased estimator of the difference of
population means. Under $H_0$ it is centred on zero.

**How wide it is.** Variance does *not* pass through a difference in the same
way. For independent quantities,

$$
\begin{aligned}
\mathrm{Var}[\bar{X} - \bar{Y}] &= \mathrm{Var}[\bar{X}] + \mathrm{Var}[-\bar{Y}] \\[4pt]
&= \mathrm{Var}[\bar{X}] + (-1)^2\mathrm{Var}[\bar{Y}] \\[4pt]
&= \frac{\sigma_x^2}{n} + \frac{\sigma_y^2}{m}
\end{aligned}
$$

using $\mathrm{Var}(aX) = a^2\mathrm{Var}(X)$ from Unit \@ref(estimation), with
$a = -1$. Taking the square root gives the standard error of the difference:

$$\mathrm{se}(\bar{X} - \bar{Y}) = \sqrt{\frac{\sigma_x^2}{n} + \frac{\sigma_y^2}{m}}$$

Independence is needed for the first line, exactly as it was in Unit
\@ref(estimation). It is what allows the variance of a sum to be the sum of the
variances.
:::

::: {.keyidea}
**The variances add, even though the means subtract.**

This is the single most important line in the unit. Squaring $-1$ gives $+1$, so
subtracting a second noisy quantity makes the result *more* variable, not less.

Comparing two groups is therefore harder than measuring one. Both estimates
wobble, and both wobbles end up in the answer.
:::

That has a practical consequence worth seeing now rather than later. Suppose two
groups each have $\sigma = 10$ and each is sampled $n = 100$ times. The standard
error of either mean alone is $10/\sqrt{100} = 1$. The standard error of their
difference is

$$\sqrt{\frac{100}{100} + \frac{100}{100}} = \sqrt{2} \approx 1.41$$

about 40% larger. Detecting a difference between two groups needs more data
than estimating either group's mean to the same precision.

## Testing the difference when σ is known {#ts-known}

::: {.question}
We have a sampling distribution. What is the test?
:::

The argument is the one from Section \@ref(os-critical), with the new standard
error in place of the old one. We reject when the observed difference is too
large:

$$|\bar{x} - \bar{y}| \geq c$$

and we choose $c$ so that, if the null were true, a gap that large would arise
with probability only $\alpha$.

::: {.theory}
Standardising the difference gives a quantity with no unknowns in it:

$$Z = \frac{(\bar{X} - \bar{Y}) - (\mu_x - \mu_y)}{\sqrt{\dfrac{\sigma_x^2}{n} + \dfrac{\sigma_y^2}{m}}} \;\sim\; N(0,1)$$

Under $H_0$ the term $(\mu_x - \mu_y)$ is zero and drops out. Setting

$$P\big[\,|\bar{X} - \bar{Y}| \geq c\,\big] = \alpha$$

and dividing through by the standard error,

$$P\left[\,|Z| \geq \frac{c}{\sqrt{\sigma_x^2/n + \sigma_y^2/m}}\,\right] = \alpha$$

The two tails carry $\alpha/2$ each, so the bracketed quantity is
$Z_{\alpha/2}$, giving

$$c = Z_{\alpha/2}\sqrt{\frac{\sigma_x^2}{n} + \frac{\sigma_y^2}{m}}$$
:::

::: {.keyidea}
The test statistic and the interval for the difference are

$$Z_{calc} = \frac{\bar{x} - \bar{y}}{\sqrt{\dfrac{\sigma_x^2}{n} + \dfrac{\sigma_y^2}{m}}}
\qquad\qquad
(\bar{x} - \bar{y}) \pm Z_{\alpha/2}\sqrt{\frac{\sigma_x^2}{n} + \frac{\sigma_y^2}{m}}$$

Reject $H_0$ when $|Z_{calc}| > Z_{\alpha/2}$, which happens exactly when the
interval excludes zero — the same equivalence as Section \@ref(os-ci), now
about a difference rather than a mean.
:::

### Conventional against high-yielding wheat {-}

Twelve plots are sown with conventional seed and fourteen with a high-yielding
variety. The conventional plots average 45.2 kg per hectare and the HYV plots
48.6 kg. From long agronomic experience the standard deviations are known:
$\sigma_x = 0.8$ for conventional and $\sigma_y = 1.0$ for HYV.

Are these data consistent, at the 5% level, with the two varieties yielding the
same on average?

```{r ts-wheat}
xbar <- 45.2; sx <- 0.8; n <- 12      # conventional
ybar <- 48.6; sy <- 1.0; m <- 14      # high-yielding

se_diff <- sqrt(sx^2 / n + sy^2 / m)
z_calc  <- (xbar - ybar) / se_diff

c(difference = xbar - ybar,
  se         = se_diff,
  z_calc     = z_calc,
  p_value    = 2 * pnorm(-abs(z_calc)))
```

$Z_{calc} = -9.63$, far beyond $-1.96$, and the $p$-value is smaller than any
conventional threshold. We reject the claim that the two varieties yield the
same.

The interval says how much the varieties differ, which is the more useful
statement:

```{r ts-wheat-ci}
(xbar - ybar) + c(-1, 1) * qnorm(0.975) * se_diff
```

The HYV seed yields between 2.7 and 4.1 kg per hectare more than the
conventional variety, with 95% confidence. Note that the interval lies entirely
below zero and so excludes it — the same verdict the test reached.

::: {.intuition}
Report the interval, not just the rejection. "The varieties differ
significantly" is a much weaker statement than "the HYV variety yields 2.7 to
4.1 kg per hectare more," and a farmer deciding what to sow needs the second.
:::

## When σ is unknown {#ts-unknown}

::: {.question}
As always, we do not know $\sigma_x$ or $\sigma_y$. Now there are two of them.
:::

In practice both population standard deviations must be estimated from the
samples, using $s_x$ and $s_y$. As in Section \@ref(os-ttest), that costs
something, and what it costs depends on the situation. Three cases arise, and
they differ only in what goes underneath the square root.

### Large samples {-}

If both samples are reasonably large — a common rule of thumb is at least 20
each — then $s_x$ and $s_y$ are reliable enough that substituting them changes
little, and the normal distribution remains a good approximation:

$$Z_{calc} = \frac{\bar{x} - \bar{y}}{\sqrt{\dfrac{s_x^2}{n} + \dfrac{s_y^2}{m}}}$$

This is the same statistic with hats on the standard deviations.

### Equal variances: the pooled estimator {-}

If we are willing to assume the two populations share a common standard
deviation, $\sigma_x = \sigma_y = \sigma$, then both samples are estimating the
*same* quantity and it is wasteful to estimate it twice. We combine them.

::: {.keyidea}
The **pooled variance estimator** is a weighted average of the two sample
variances, each weighted by its degrees of freedom:

$$s_p^2 = \frac{(n-1)s_x^2 + (m-1)s_y^2}{n + m - 2}$$

and the test statistic becomes

$$t_{calc} = \frac{\bar{x} - \bar{y}}{\sqrt{s_p^2\left(\dfrac{1}{n} + \dfrac{1}{m}\right)}}$$

compared against $t$ with $n + m - 2$ degrees of freedom.
:::

The degrees of freedom are worth pausing on. We began with $n + m$
observations, and spent one estimating each of the two sample means, leaving
$n + m - 2$. The same accounting as Section \@ref(u3-variance), applied twice.

Pooling buys precision — $n + m - 2$ degrees of freedom rather than something
closer to the smaller sample's — but only if the assumption holds. If one group
really is more variable than the other, the pooled estimate is a weighted
average of two different things, and the test is wrong in a way the output will
not reveal.

### Unequal variances: Welch's test {-}

The safer option makes no assumption about the variances. It uses the separate
estimates,

$$t_{calc} = \frac{\bar{x} - \bar{y}}{\sqrt{\dfrac{s_x^2}{n} + \dfrac{s_y^2}{m}}}$$

and adjusts the degrees of freedom downwards to compensate for the extra
uncertainty. The adjustment is not a whole number and is not worth memorising;
R computes it.

::: {.watchout}
**R's `t.test()` uses Welch by default.** If you have derived a pooled
statistic by hand and R disagrees with you, this is usually why.

To pool, pass `var.equal = TRUE` explicitly.

Modern practice is to prefer Welch. It costs very little when the variances
really are equal, and it protects you when they are not — and you can never
verify equal variances, only fail to reject the claim that they are equal.
:::

### Do North and South Indian students differ in height? {-}

`student-survey-height.csv` records 100 students, each classified by region.

```{r ts-height-desc}
heights <- read.csv("data/student-survey-height.csv")

aggregate(height ~ region, data = heights,
          function(x) round(c(n = length(x), mean = mean(x), sd = sd(x)), 2))
```

The two sample means differ by about 1.5 cm, and the two sample standard
deviations are not equal — 7.3 against 9.9. That is a case for Welch.

```{r ts-height-test}
t.test(height ~ region, data = heights)
```

$t = 0.83$ on 74.4 degrees of freedom, with $p = 0.41$. We fail to reject. The
95% interval for the difference runs from $-2.07$ to $5.04$ cm, and it
**contains zero** — the two statements agree, as they must.

So a 1.5 cm gap in the samples is entirely unremarkable. With these sample
sizes and this much variation within each group, differences of that size turn
up constantly when the populations are identical.

::: {.deeper}
Compare the degrees of freedom the two approaches give here.

```{r ts-df-compare}
c(welch  = t.test(height ~ region, data = heights)$parameter,
  pooled = t.test(height ~ region, data = heights, var.equal = TRUE)$parameter)
```

Pooling claims 98 degrees of freedom; Welch allows only 74.4. The difference is
the price of not assuming the two groups are equally variable — and here, where
one sample's standard deviation is a third larger than the other's, it is a
price worth paying.

Both give the same verdict in this case. They do not always.
:::

## When two samples are not really two samples {#ts-paired}

::: {.question}
Ten couples. Two columns of ages. Is that two samples?
:::

Suppose we record the ages of both partners in ten married couples.

```{r ts-marriage}
couples <- read.csv("data/marriage-ages.csv")
couples
```

Our question is simple: **is partner 1, on average, older than partner 2?**

At first glance this looks like a two-sample problem. There are two columns.
There are ten observations in each. So why not perform the test we have just
learned?

```{r ts-marriage-wrong}
t.test(couples$partner1, couples$partner2)
```

The test finds essentially nothing: $t = 0.52$ with $p = 0.61$. On this evidence
the two are the same age on average, and the interval for the difference runs
from $-7.5$ to $12.5$ years.

### Something is wrong {-}

Look carefully at the data. In every couple, which partner is older?

| Couple | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 | 10 |
|:--|:--|:--|:--|:--|:--|:--|:--|:--|:--|:--|
| Older | P1 | P1 | P1 | P1 | P1 | P1 | same | P1 | P1 | same |

Partner 1 is older in eight couples. In the other two the partners are exactly
the same age. **Partner 2 is never older.**

How can a statistical test look at that and return $p = 0.61$?

The answer is that we analysed the wrong problem.

### What the two-sample test assumed {-}

The two-sample $t$-test assumes the observations in one sample have nothing to
do with the observations in the other. It treats the data as two unrelated
collections of people:

$$\text{Partner 1:}\quad 36 \quad 72 \quad 37 \quad 36 \quad 51 \quad \ldots$$
$$\text{Partner 2:}\quad 35 \quad 67 \quad 33 \quad 35 \quad 50 \quad \ldots$$

But they are not unrelated. Each observation belongs to a *married couple*, and
within a couple the two ages are naturally linked. Someone who marries at 70 is
likely to have a spouse around 70; someone who marries at 35 is likely to have a
spouse around 35. The ages within a couple are strongly correlated.

### Two sources of variation {-}

There are two quite different sources of variation in this dataset, and only one
of them is of any interest.

**Variation between couples.** Some couples are in their thirties, some in
their fifties, one in their seventies. These differences are enormous.

**Variation within couples.** Partners usually differ by only a few years. This
is the variation we actually care about.

The two-sample test mixes them together. The between-couple variation dominates
the calculation and hides the small but remarkably consistent within-couple
differences.

Seen properly, the structure is obvious.

```{r ts-marriage-plot, echo = FALSE, fig.height = 3.8, fig.cap = "Ten couples, ordered by the age of partner 2. Each segment joins the two partners in one couple. The couples span nearly forty years between them; within any couple the gap is at most eight. Two couples are the same age, and show as a red ring around a blue centre."}
cp <- couples[order(couples$partner2), ]
cp$id <- seq_len(nrow(cp))

# Partner 1 is drawn first and larger: in the two couples of equal age the two
# points coincide exactly, and this leaves a visible ring rather than one dot
# silently covering the other.
long <- rbind(
  data.frame(id = cp$id, age = cp$partner1, who = "Partner 1", sz = 3.6),
  data.frame(id = cp$id, age = cp$partner2, who = "Partner 2", sz = 1.9)
)

ggplot(long, aes(age, id)) +
  geom_segment(data = cp, aes(x = partner2, xend = partner1, y = id, yend = id),
               colour = book_palette$soft, linewidth = 1.6,
               inherit.aes = FALSE) +
  geom_point(aes(colour = who, size = sz)) +
  scale_size_identity() +
  scale_colour_manual(values = c("Partner 1" = book_palette$reject,
                                 "Partner 2" = book_palette$accent),
                      name = NULL) +
  scale_y_continuous(breaks = NULL) +
  labs(x = "Age (years)", y = "Couple") +
  theme_book() +
  theme(legend.position = "top", panel.grid.major.y = element_blank())
```

The figure explains both results at once. Vertically, the couples are spread
across nearly forty years — that is the variation the unpaired test measures the
gap against, and it is overwhelming. Horizontally, within each couple, the red
point is at or to the right of the blue one every single time, and never more
than eight years away. That small, utterly consistent gap is what a paired test
measures.

### Removing what we do not care about {-}

Instead of comparing all the partner 1s with all the partner 2s, compare each
person with *their own* partner. Compute the difference for every couple.

```{r ts-marriage-diff}
d <- couples$partner1 - couples$partner2
d
```

Now something worth pausing on has happened. The ages of the couples have
disappeared completely.

The couple aged 72 and 67 contributes $72 - 67 = 5$. The couple aged 36 and 35
contributes $36 - 35 = 1$. It no longer matters whether a couple is young or
old — subtracting removes the between-couple variation automatically, because it
was common to both columns.

Only the within-couple difference remains, and that is exactly the quantity we
wanted to study.

### The statistical model {-}

Let

$$d_i = x_i - y_i$$

where $x_i$ is the age of partner 1 in couple $i$ and $y_i$ is the age of
partner 2 in the *same* couple.

Instead of two samples, we now have one sample of differences
$d_1, d_2, \ldots, d_n$. Our parameter of interest is

$$\mu_d = E(d)$$

the population mean difference.

### Setting up the hypotheses {-}

If we simply want to know whether the average ages differ,

$$H_0 : \mu_d = 0 \qquad H_A : \mu_d \neq 0$$

If we specifically suspect partner 1 is older, the alternative is one-sided, as
in Section \@ref(os-onetailed):

$$H_0 : \mu_d \leq 0 \qquad H_A : \mu_d > 0$$

::: {.keyidea}
Notice that the hypotheses are no longer about two population means. We are not
testing $H_0: \mu_1 = \mu_2$.

We are testing whether the average *difference* equals zero — a claim about a
single parameter, which is why the machinery of Unit \@ref(one-sample) applies
unchanged.
:::

### Estimating the mean difference {-}

The sample mean difference is

$$\bar{d} = \frac{1}{n}\sum_{i=1}^{n} d_i$$

and the sample standard deviation of the differences is the ordinary one from
Section \@ref(u3-variance), applied to the $d_i$:

$$s_d = \sqrt{\frac{\sum_{i=1}^{n}(d_i - \bar{d})^2}{n-1}}$$

```{r ts-marriage-stats}
c(n = length(d), d_bar = mean(d), s_d = sd(d), se = sd(d) / sqrt(length(d)))
```

So $\bar{d} = 2.5$ years and $s_d = 2.635$ years.

### The test statistic {-}

Under the null hypothesis $\mu_d = 0$,

$$t = \frac{\bar{d} - 0}{s_d / \sqrt{n}}$$

This is exactly the one-sample $t$ statistic from Section \@ref(os-ttest), with
$d$ in place of $x$. Substituting,

$$t = \frac{2.5}{2.635/\sqrt{10}} = \frac{2.5}{0.833} = 3.00$$

and under $H_0$ this follows $t_{n-1}$, so with $n = 10$ we have $df = 9$.

```{r ts-marriage-byhand}
t_calc <- mean(d) / (sd(d) / sqrt(length(d)))

c(t_calc  = t_calc,
  df      = length(d) - 1,
  t_crit  = qt(0.975, df = length(d) - 1),
  p_value = 2 * pt(-abs(t_calc), df = length(d) - 1))
```

Since $3.00 > 2.262$, we reject $H_0$ at the 5% level. The two-sided $p$-value
is 0.015.

### Confidence interval {-}

A 95% interval for the mean difference is

$$\bar{d} \pm t_{\alpha/2,\,n-1}\,\frac{s_d}{\sqrt{n}}$$

```{r ts-marriage-ci}
mean(d) + c(-1, 1) * qt(0.975, df = 9) * sd(d) / sqrt(10)
```

Partner 1 is between about 0.6 and 4.4 years older, on average. The interval
excludes zero, agreeing with the test as always.

### Doing it in R {-}

Everything above can be had two ways. Either compute the differences yourself
and run a one-sample test,

```{r ts-marriage-onesample}
t.test(d, mu = 0)
```

or let R take the differences for you:

```{r ts-marriage-paired}
t.test(couples$partner1, couples$partner2, paired = TRUE)
```

The two are the same test, because `paired = TRUE` does nothing more than
subtract the columns and run the one-sample test on the result.

### Why did the answer change so much? {-}

Notice what changed, and what did not.

Table: (\#tab:ts-paired-compare) The same ten couples, analysed twice.

| Analysis | Treats the observations as | $p$-value |
|:--|:--|--:|
| Two-sample $t$-test | Twenty unrelated people | 0.61 |
| Paired $t$-test | Ten differences within couples | 0.015 |

The data never changed. The same ten couples were analysed twice. The only
difference was whether the analysis respected the way the data were collected.

::: {.theory}
The algebra shows exactly what pairing removes. Write each age as

$$x_i = \mu + \alpha_i + \delta_i \qquad\qquad y_i = \mu + \alpha_i + \varepsilon_i$$

where $\mu$ is the overall average age, $\alpha_i$ captures whether couple $i$
is generally young or old, and $\delta_i$ and $\varepsilon_i$ are the remaining
individual departures.

The term $\alpha_i$ is what makes the columns so variable: it ranges across
nearly forty years in these data. Crucially, it is *the same* $\alpha_i$ in both
expressions, because both partners belong to the same couple.

Taking differences,

$$
\begin{aligned}
d_i &= x_i - y_i \\[4pt]
&= (\mu + \alpha_i + \delta_i) - (\mu + \alpha_i + \varepsilon_i) \\[4pt]
&= \delta_i - \varepsilon_i
\end{aligned}
$$

Both $\mu$ and $\alpha_i$ cancel. The large between-couple effect disappears
entirely, and the differences carry only the within-couple departures.

That is why the paired test is so much more powerful. It is not measuring a
smaller quantity more cleverly; it is measuring the same quantity against a much
smaller standard error.
:::

The standard deviations make the point numerically:

```{r ts-marriage-sds}
round(c(sd_partner1 = sd(couples$partner1),
        sd_partner2 = sd(couples$partner2),
        sd_of_diffs = sd(d)), 3)
```

Each column varies by about eleven years. The differences vary by 2.6. The
unpaired test measured a 2.5-year gap against eleven years of noise and saw
nothing; the paired test measures it against 2.6 and finds it clearly.

### When should we use a paired test? {-}

A paired test is appropriate whenever the two measurements belong to the same
unit. Typical cases:

- before and after measurements on the same patients;
- yields from the same plots under two fertilisers in successive seasons;
- examination scores of the same students before and after a training programme;
- the same households interviewed in two survey rounds;
- twins, or matched pairs, receiving different treatments.

The common feature is that every observation in one column has a natural partner
in the other. The analysis should therefore focus on the difference within each
pair, not on the two columns separately.

::: {.watchout}
The reverse error is also possible. If the two columns are *not* paired — 45
government school students and 52 private school students, say — then there is
no natural partner for anyone, `paired = TRUE` will pair them by row number,
which is meaningless, and R will simply refuse if the lengths differ.

Pairing is a fact about how the data were collected, not an option to try.
:::

::: {.keyidea}
**A paired $t$-test is not a special kind of two-sample $t$-test. It is a
one-sample $t$-test applied to the differences between paired observations.**

If you remember one sentence from this section, it should be that one.
:::

Table: (\#tab:ts-paired-vs-one) A paired test is a one-sample test wearing different clothes.

| | One-sample $t$-test | Paired $t$-test |
|:--|:--|:--|
| Data | $x_1, \ldots, x_n$ | $d_1, \ldots, d_n$ where $d_i = x_i - y_i$ |
| Parameter | $\mu$ | $\mu_d$ |
| Null | $H_0: \mu = \mu_0$ | $H_0: \mu_d = 0$ |
| Statistic | $t = \dfrac{\bar{x} - \mu_0}{s/\sqrt{n}}$ | $t = \dfrac{\bar{d}}{s_d/\sqrt{n}}$ |
| Degrees of freedom | $n - 1$ | $n - 1$ |

Read across the rows. There is no new mathematics here at all — only a new first
step, in which two columns become one.


## A difference is not an effect {#ts-causal}

::: {.question}
The test says two groups differ. Does it say why?
:::

We can now establish that two groups differ by more than sampling variation.
That is a genuine achievement, and it is not what most interesting questions
actually ask.

Consider the attendance data from this course. Split the 680 students at the
median attendance and compare their final examination scores.

```{r ts-attend}
students <- read.csv("data/attendance-grades.csv")
students$group <- ifelse(students$attendance >= median(students$attendance),
                         "High attendance", "Low attendance")

aggregate(final_score ~ group, data = students,
          function(x) round(c(n = length(x), mean = mean(x)), 2))

t.test(final_score ~ group, data = students)
```

Students who attended more scored higher, and $p = 0.011$ rejects the claim
that the two groups have the same mean.

It is tempting to conclude that attending class raises examination scores by
about a point. Two separate problems stand in the way.

### The difference is significant and trivial {-}

The interval for the difference runs from about 0.21 to 1.59 marks out of 100.

A gap that small would not change anyone's grade, and would not justify any
policy. Yet it is "statistically significant", because 680 students is a large
enough sample to detect very small differences reliably.

::: {.keyidea}
Statistical significance answers "is this difference bigger than noise?"

It does not answer "is this difference big enough to matter?" That is a question
about the subject, not about the statistics, and no $p$-value speaks to it.

Always read the interval, in the units of the problem, before deciding whether
a result is important.
:::

### The groups were not assigned {-}

The deeper problem is that nobody allocated students to high and low
attendance. They chose, and the reasons they chose are not in the dataset.

A student who attends regularly may be more motivated, better prepared, less
burdened by paid work or family obligations, or living closer to campus. Any of
those could raise examination scores on its own. The two groups differ in
attendance *and* in everything that attendance is correlated with, and the test
cannot separate them.

::: {.watchout}
The two-sample test compares two groups as they arrive. It has nothing to say
about *why* they came to be different.

If the groups formed themselves, a difference in outcomes may reflect the
treatment, or the reasons for selection, or both — and the arithmetic is
identical either way. This is the problem of **selection**, and it is the
central obstacle in all of empirical economics.
:::

It is worth being precise about what has and has not been shown here.

We have shown that students who attended more scored slightly higher. That is a
fact about the world and the test establishes it properly. We have not shown
that attending more *causes* higher scores, and no amount of additional care
with the $t$ statistic would help.

What would help is a different design. If attendance had been assigned by
something unrelated to the students themselves — a lottery, a timetable clash,
a bus route — then the groups would be comparable in expectation and the
difference could be read as an effect. Creating such comparisons, or finding
them in the world, is the subject of the second part of this book.

::: {.intuition}
The question "do these groups differ?" is a statistical one, and this unit
answers it.

The question "does this treatment work?" is a question about how the groups were
formed. It is answered by the design of the study, not by the choice of test.
:::

## Summary {#ts-summary}

::: {.keyidea}
- Two groups produce two sample means, and a difference between them is
  guaranteed to appear even when the populations are identical. The question is
  whether it is larger than sampling variation would produce.
- $E[\bar{X} - \bar{Y}] = \mu_x - \mu_y$: the difference of sample means is
  unbiased for the difference of population means.
- **The variances add even though the means subtract**, so
  $\mathrm{se}(\bar{X} - \bar{Y}) = \sqrt{\sigma_x^2/n + \sigma_y^2/m}$.
  Comparing two groups is noisier than measuring one, and needs more data.
- With $\sigma$ known the test is a $Z$-test on the difference; the interval is
  $(\bar{x} - \bar{y}) \pm Z_{\alpha/2}\,\mathrm{se}$, and it excludes zero
  exactly when the test rejects.
- With $\sigma$ unknown, three cases differ only under the square root: large
  samples use $s_x$ and $s_y$ with the normal; **pooling** assumes equal
  variances and buys $n + m - 2$ degrees of freedom; **Welch** assumes nothing
  and adjusts the degrees of freedom down. R's `t.test()` uses Welch unless you
  pass `var.equal = TRUE`, and Welch is the safer default.
- A **paired** design gives one number per unit. Take differences and run a
  one-sample test. Treating paired data as two independent samples discards the
  comparison the design was built to make — in the marriage data it turned
  $p = 0.015$ into $p = 0.61$.
- A significant difference need not be an important one. Read the interval in
  the units of the problem.
- A difference between two groups that formed themselves is not an effect. The
  test compares the groups as they arrive and is silent about why they differ.
:::

## Practice problems {#ts-practice}

::: {.yourturn}
Work each one before opening the solution. The
[question bank](#ts-question-bank) that follows has no answers.
:::

### Concept checks {-}

**5.1** Why is the standard error of a difference of means larger than the
standard error of either mean on its own?

::: {.answer}
Because the two sample means each carry sampling variation, and both end up in
the difference. Formally, $\mathrm{Var}(aX) = a^2\mathrm{Var}(X)$ with $a = -1$
gives $(-1)^2 = +1$, so the variances add:

$$\mathrm{Var}[\bar{X} - \bar{Y}] = \frac{\sigma_x^2}{n} + \frac{\sigma_y^2}{m}$$

Subtracting a noisy quantity does not cancel noise; it adds more. This is why
detecting a difference between two groups takes more data than estimating one
group's mean to the same precision.
:::

**5.2** A study reports that two groups differ with $p = 0.003$. A colleague
concludes the difference is large. What is wrong?

::: {.answer}
The $p$-value measures how incompatible the data are with the claim of no
difference. It says nothing about the size of the difference.

A very small $p$-value can come from a tiny difference measured in a large
sample. The size of the difference is reported by the confidence interval, in
the units of the problem, and that is what has to be inspected before calling a
result large or important.
:::

**5.3** Why does R's `t.test()` report fractional degrees of freedom?

::: {.answer}
Because it performs Welch's test by default, which does not assume the two
populations have equal variances. The correction for that extra uncertainty
adjusts the degrees of freedom downwards, and the adjustment is generally not a
whole number.

The pooled test, obtained with `var.equal = TRUE`, uses $n + m - 2$ degrees of
freedom, which always is.
:::

**5.4** The same twenty households are surveyed before and after a cash
transfer. Why is a two-sample test the wrong tool?

::: {.answer}
Because the two columns are not independent samples — each household appears in
both. The design produces one number per household, the change.

Households differ enormously among themselves, and that variation is present
identically in both columns. A two-sample test treats it as noise, and it
swamps the change we are trying to detect. Taking differences within households
removes it, leaving only the quantity of interest.
:::

**5.5** Villages that received a road show higher incomes than villages that
did not. State two distinct reasons this need not mean the road raised incomes.

::: {.answer}
First, the villages may have differed beforehand. Roads are not built at random
— they tend to reach places that are more accessible, more populous, more
politically connected or already better off. Those same features raise incomes
on their own.

Second, the two groups may differ in things that changed alongside the road:
electrification, a market, a school. The road is correlated with a bundle of
improvements, and a comparison of means attributes the whole bundle to it.

In both cases the arithmetic is correct and the causal reading is not.
:::

### Explain {-}

**5.6** Explain the trade-off between the pooled and Welch tests, and why
modern practice prefers Welch.

::: {.answer}
Pooling assumes the two populations share a standard deviation. If true, both
samples estimate the same quantity, they can be combined, and the resulting
test has $n + m - 2$ degrees of freedom — more than Welch allows, so more
power.

If false, the pooled estimate averages two different quantities and the test is
wrong in a way nothing in the output reveals. Welch avoids the assumption
entirely, at the cost of fewer degrees of freedom.

The reason to prefer Welch is asymmetric risk. When the variances really are
equal, Welch loses very little. When they are not, pooling misleads. And the
assumption cannot be verified — a test of equal variances can only fail to
reject, which is not the same as establishing it.
:::

**5.7** A researcher tests whether men and women in a firm are paid differently,
finds $p = 0.02$, and concludes the firm discriminates. Evaluate.

::: {.answer}
The test establishes that average pay differs by more than sampling variation
would explain. That is not the same as discrimination.

Men and women in the firm may differ in role, seniority, years of experience,
hours, or field of training, and any of those could account for a pay gap
without anyone being treated differently for the same work. The comparison of
raw means attributes the whole difference to sex because sex is the only
variable in the analysis.

Establishing discrimination requires comparing like with like — people doing
the same work with the same experience — which is a question of design and, in
later units, of holding other variables constant. The two-sample test is not
capable of it.
:::

### Numerical problems {-}

**5.8** Two irrigation methods are trialled. Drip irrigation on 15 plots yields
a mean of 38.4 quintals per hectare; flood irrigation on 18 plots yields 35.9.
Population standard deviations are known to be 3.2 and 4.1 respectively.

i. Test at the 5% level whether the mean yields differ.
ii. Construct a 95% confidence interval for the difference.
iii. Does the interval agree with the test?

::: {.answer}
**(i)** The standard error of the difference is

$$\sqrt{\frac{3.2^2}{15} + \frac{4.1^2}{18}} = \sqrt{0.6827 + 0.9339}
= \sqrt{1.6166} = 1.271$$

$$Z_{calc} = \frac{38.4 - 35.9}{1.271} = \frac{2.5}{1.271} = 1.967$$

Since $1.967 > 1.96$, we reject $H_0$ at the 5% level — but only just.

**(ii)** $2.5 \pm 1.96(1.271) = 2.5 \pm 2.491 = [0.009,\ 4.991]$.

**(iii)** Yes. The interval excludes zero, and the test rejects. Both do so by
the narrowest possible margin, which is the same fact stated twice: the interval
endpoint sits at 0.009 because $Z_{calc}$ sits at 1.967.

A result this marginal should be reported as marginal. Nothing about it
justifies confidence that drip irrigation is better.
:::

**5.9** Two samples give $n = 10$, $s_x = 4$ and $m = 14$, $s_y = 6$.

i. Compute the pooled variance estimate.
ii. How many degrees of freedom does the pooled test use?
iii. Why is it not $n + m$?

::: {.answer}
**(i)**

$$s_p^2 = \frac{(10-1)16 + (14-1)36}{10 + 14 - 2}
= \frac{144 + 468}{22} = \frac{612}{22} = 27.82$$

so $s_p = 5.27$, lying between 4 and 6 and closer to 6 because the larger
sample carries more weight.

**(ii)** $n + m - 2 = 22$.

**(iii)** Because two quantities were estimated from the data before the spread
could be measured — one sample mean in each group. Each costs a degree of
freedom, exactly as in Section \@ref(u3-variance), where estimating a single
mean cost one.
:::

**5.10** A programme is evaluated by comparing 200 participants with 200
non-participants. Participants' mean income is ₹4,200 higher, with a standard
error of ₹1,400 for the difference.

i. Test whether the difference is zero at the 5% level.
ii. Construct a 95% interval.
iii. The programme was open to anyone who applied. What can be concluded about
     its effect?

::: {.answer}
**(i)** $Z_{calc} = 4200/1400 = 3.0$, which exceeds 1.96. We reject the claim
that the two groups have the same mean income.

**(ii)** $4200 \pm 1.96(1400) = [₹1{,}456,\ ₹6{,}944]$.

**(iii)** Nothing about its effect. Because the programme was open to
applicants, participants selected themselves. People who apply to income
programmes may be more motivated, better informed, or better connected — or,
equally plausibly, more desperate and worse off to begin with. The direction of
the bias is not even clear.

The interval is a correct statement about the difference between two groups of
people. It is not an estimate of what the programme did to anybody.
:::

### R exercises {-}

**5.11** Using `data/student-survey-sleep.csv`, test whether urban and rural
students report different average hours of sleep. Report the interval and say
what you conclude.

::: {.answer}
```{r ts-ex-sleep}
sleep <- read.csv("data/student-survey-sleep.csv")

aggregate(sleep ~ residence, data = sleep,
          function(x) round(c(n = length(x), mean = mean(x), sd = sd(x)), 2))

t.test(sleep ~ residence, data = sleep)
```

The two group means are close and the interval for the difference contains
zero, so we fail to reject. There is no evidence here that urban and rural
students sleep different amounts on average.

Note what "no evidence of a difference" is not: evidence of no difference. With
these sample sizes, a real difference of well under an hour would be difficult
to detect at all, and the interval says as much by being wide.
:::

**5.12** Test whether social media use differs between North and South Indian
students in the same dataset, using both Welch and the pooled test. Do they
agree, and which would you report?

::: {.answer}
```{r ts-ex-pooled}
rbind(
  welch  = unlist(t.test(socialmediahours ~ region, data = sleep)[c("statistic", "parameter", "p.value")]),
  pooled = unlist(t.test(socialmediahours ~ region, data = sleep, var.equal = TRUE)[c("statistic", "parameter", "p.value")])
)
```

The two agree closely here, which is what happens when the sample standard
deviations are similar — the pooled and separate estimates of the variance are
then nearly the same number.

Report Welch. It gives the same answer when the assumption holds and a correct
one when it does not, so there is no situation in which pooling is the safer
choice.
:::

**5.13** Take the wage data as a population and draw independent samples of 200
urban and 200 rural workers. Test whether mean annual earnings differ, and
report the interval. Then say what the interval does and does not establish.

::: {.answer}
```{r ts-ex-wages}
wages <- read.csv("data/wages-india-synthetic.csv")

set.seed(9)
urban <- sample(wages$annual_earnings[wages$residence == "Urban"], 200)
rural <- sample(wages$annual_earnings[wages$residence == "Rural"], 200)

t.test(urban, rural)
```

The difference is large and the interval is nowhere near zero, so urban workers
in this population earn substantially more than rural workers.

What the interval establishes is the size of that gap. What it does not
establish is that moving a worker from a village to a city would raise their
earnings by that amount. Urban and rural workers differ in education,
occupation and industry, and the sample says nothing about which part of the gap
belongs to the place and which to the people.
:::

**5.14** Simulate the point that variances add. Draw two independent samples of
100 from a population with $\sigma = 10$, and compare the observed standard
deviation of $\bar{x}$ with that of $\bar{x} - \bar{y}$ across 5,000
repetitions. Check both against theory.

::: {.answer}
```{r ts-ex-varadd}
set.seed(4)
sims <- replicate(5000, {
  x <- rnorm(100, mean = 50, sd = 10)
  y <- rnorm(100, mean = 50, sd = 10)
  c(one_mean = mean(x), difference = mean(x) - mean(y))
})

round(c(sd_of_one_mean       = sd(sims["one_mean", ]),
        theory_one_mean      = 10 / sqrt(100),
        sd_of_difference     = sd(sims["difference", ]),
        theory_difference    = sqrt(10^2/100 + 10^2/100)), 4)
```

The difference is about $\sqrt{2}$ times as variable as a single mean, exactly
as $\sqrt{\sigma^2/n + \sigma^2/m}$ predicts. Note also that the difference is
centred on zero: both populations have mean 50, and the estimator is unbiased
for a difference of zero.
:::

**5.15** Demonstrate the cost of ignoring pairing. Using `marriage-ages.csv`,
report the paired and unpaired $p$-values, and explain the gap by comparing the
standard deviation of the differences with the standard deviations of the two
columns.

::: {.answer}
```{r ts-ex-paired}
couples <- read.csv("data/marriage-ages.csv")

c(paired   = t.test(couples$partner1, couples$partner2, paired = TRUE)$p.value,
  unpaired = t.test(couples$partner1, couples$partner2)$p.value)

round(c(sd_partner1  = sd(couples$partner1),
        sd_partner2  = sd(couples$partner2),
        sd_of_diffs  = sd(couples$partner1 - couples$partner2)), 3)
```

The two columns each have a standard deviation of about 11 years, because the
couples range from their mid-thirties to their seventies. The differences have a
standard deviation of about 2.6 years.

That is the whole explanation. The unpaired test measures the small gap against
the huge between-couple variation and sees nothing. The paired test removes that
variation — it is common to both columns and cancels on subtraction — and
measures the gap against what is left.
:::

### Paired designs {-}

**5.16** A paired test on $n$ pairs has $n - 1$ degrees of freedom, not
$2n - 2$. Why, when there are $2n$ numbers in the dataset?

::: {.answer}
Because the analysis is not performed on the $2n$ numbers. It is performed on
the $n$ differences.

Taking differences reduces $2n$ observations to $n$, one per pair. From those
$n$ we estimate one quantity, $\bar{d}$, before we can measure their spread, so
one degree of freedom is spent and $n - 1$ remain — exactly as in the one-sample
case of Section \@ref(u3-variance).

The $2n - 2$ of the pooled two-sample test is right for a design in which the
$2n$ observations really are $2n$ independent pieces of information. In a paired
design they are not: knowing one member of a pair tells you a great deal about
the other.
:::

**5.17** What feature of the data determines whether pairing was worth doing?

::: {.answer}
The correlation between the two columns.

Pairing removes whatever is common to both members of a pair. If the two
measurements are strongly related — the same household in two rounds, the same
plot in two seasons — then a great deal is common, differencing removes it, and
the standard error falls sharply.

If the two columns happen to be unrelated, there is nothing common to remove.
Differencing then leaves the variability essentially unchanged, and the paired
test performs about as well as the unpaired one, losing only the few degrees of
freedom.

So pairing rarely costs much and can gain a great deal. What determines which
happens is not a choice; it is a property of the design.
:::

**5.18** A researcher measures productivity at the same 30 firms before and
after a policy change, then runs a two-sample $t$-test. Describe what will
happen to the $p$-value and why.

::: {.answer}
It will be too large — often dramatically so — and the researcher will
under-report a real effect.

Firms differ enormously among themselves in productivity: a large firm and a
small one may differ by a factor of ten, while the policy might move any single
firm by a few per cent. In a two-sample test that between-firm variation is
counted as noise, and it swamps the change being looked for.

The pairing is exactly what removes it. Each firm's before-measurement acts as
its own comparison, so the firm's general productivity level cancels and only
its change remains. Writing $x_i = \mu + \alpha_i + \delta_i$ and
$y_i = \mu + \alpha_i + \varepsilon_i$, the large firm effect $\alpha_i$ appears
in both and disappears on subtraction.

The error is conservative rather than anti-conservative, which is why it often
goes unnoticed: the researcher simply concludes there is no effect.
:::

**5.19** A new irrigation technique is trialled on eight plots. Yields in
quintals per hectare before and after are

| Plot | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 |
|:--|:--|:--|:--|:--|:--|:--|:--|:--|
| Before | 32 | 28 | 35 | 30 | 33 | 29 | 31 | 34 |
| After | 35 | 33 | 37 | 36 | 37 | 30 | 36 | 40 |

i. Explain why this is a paired design.
ii. Compute the differences, $\bar{d}$ and $s_d$.
iii. Test at the 5% level whether the technique changed yields.
iv. Construct a 95% confidence interval for the mean change.

::: {.answer}
**(i)** Each plot is measured twice. The two columns are not eight plots and
eight other plots; they are the same eight plots under two conditions, so every
observation in one column has a natural partner in the other. Plots differ in
soil, slope and drainage, and those differences are present identically in both
rounds.

**(ii)** Differences (after $-$ before): $3, 5, 2, 6, 4, 1, 5, 6$.

$$\bar{d} = \frac{32}{8} = 4.0$$

Deviations from 4: $-1, 1, -2, 2, 0, -3, 1, 2$, whose squares sum to 24, so

$$s_d^2 = \frac{24}{7} = 3.4286 \qquad s_d = 1.852$$

and the standard error is $1.852/\sqrt{8} = 0.655$.

**(iii)**

$$t = \frac{4.0}{0.655} = 6.11$$

with $df = 7$. The critical value is $t_{0.025,\,7} = 2.365$, and
$6.11 > 2.365$, so we reject $H_0: \mu_d = 0$. The $p$-value is 0.0005.

**(iv)** $4.0 \pm 2.365(0.655) = [2.45,\ 5.55]$ quintals per hectare.

Note the interval is in the units of the problem and excludes zero, agreeing
with the test. Note also that the estimated gain of 4 quintals per hectare is
large in agronomic terms as well as statistically detectable — the two
questions are separate, and here both answers happen to be favourable.
:::

**5.20** Seven workers are timed on a task, in minutes, before and after
training.

| Worker | 1 | 2 | 3 | 4 | 5 | 6 | 7 |
|:--|:--|:--|:--|:--|:--|:--|:--|
| Before | 14 | 12 | 15 | 13 | 16 | 11 | 14 |
| After | 11 | 10 | 13 | 12 | 13 | 11 | 12 |

i. State the hypotheses for a test that training *reduced* completion time.
ii. Compute the test statistic and its degrees of freedom.
iii. Test at the 5% level.
iv. Construct the appropriate one-sided 95% bound and interpret it.

::: {.answer}
**(i)** Working with $d_i = \text{before}_i - \text{after}_i$, so that a
reduction in time gives a positive difference:

$$H_0 : \mu_d \leq 0 \qquad H_A : \mu_d > 0$$

The equality lives in $H_0$, as always, and the claim we want evidence *for*
sits in $H_A$.

**(ii)** Differences: $3, 2, 2, 1, 3, 0, 2$. So $\bar{d} = 13/7 = 1.857$ and
$s_d = 1.069$, giving a standard error of $1.069/\sqrt{7} = 0.404$ and

$$t = \frac{1.857}{0.404} = 4.60$$

with $df = 6$.

**(iii)** Right-tailed at 5%, so the critical value is $t_{0.05,\,6} = 1.943$ —
not 2.447. Since $4.60 > 1.943$ we reject $H_0$; the one-sided $p$-value is
0.0019.

**(iv)** A one-sided lower bound spends all of $\alpha$ in one tail:

$$\bar{d} - t_{0.05,\,6}\,\frac{s_d}{\sqrt{n}} = 1.857 - 1.943(0.404) = 1.07$$

so the interval is $[1.07,\ \infty)$. With 95% confidence, training reduced
completion time by at least 1.07 minutes. We have deliberately given up any
statement about how large the reduction might be, because the question only
asked whether there was one.
:::

**5.21** Using `marriage-ages.csv`, verify by direct computation that
`paired = TRUE` does nothing more than take differences. Then run the one-sided
test of whether partner 1 is older.

::: {.answer}
```{r ts-ex-paired-equiv}
couples <- read.csv("data/marriage-ages.csv")
d <- couples$partner1 - couples$partner2

# Three routes to the same number
c(by_hand   = mean(d) / (sd(d) / sqrt(length(d))),
  one_sample = t.test(d, mu = 0)$statistic,
  paired     = t.test(couples$partner1, couples$partner2, paired = TRUE)$statistic)
```

Identical, as they must be: `paired = TRUE` subtracts the columns and calls the
one-sample routine.

```{r ts-ex-paired-onesided}
t.test(couples$partner1, couples$partner2,
       paired = TRUE, alternative = "greater")
```

The one-sided $p$-value is half the two-sided one, 0.0075 against 0.015, and the
interval is now a lower bound: partner 1 is older by at least 1.02 years.

The one-sided test is defensible here only because the direction was decided
before looking — we asked whether partner 1 is older, not whether the two
differ. Choosing the direction after seeing that eight of ten differences were
positive would not be a test at all.
:::

**5.22** Show that the benefit of pairing depends on the correlation between the
two columns. Simulate pairs of measurements with correlation 0, 0.5 and 0.9, a
true difference of 4, and $n = 10$, and compare how often each test rejects.

::: {.answer}
```{r ts-ex-paired-power}
set.seed(7)

power_at <- function(rho, reps = 4000) {
  out <- replicate(reps, {
    a <- rnorm(10)
    b <- rho * a + sqrt(1 - rho^2) * rnorm(10)
    x <- 50 + 10 * a
    y <- 50 + 10 * b + 4                       # true difference is 4
    c(paired   = t.test(x, y, paired = TRUE)$p.value < 0.05,
      unpaired = t.test(x, y)$p.value           < 0.05,
      se_paired   = sd(x - y) / sqrt(10),
      se_unpaired = sqrt(var(x)/10 + var(y)/10))
  })
  c(rho = rho, rowMeans(out))
}

round(t(sapply(c(0, 0.5, 0.9), power_at)), 3)
```

Read the two standard error columns first, because they explain everything else.
The paired standard error falls from about 4.35 to 1.37 as the correlation
rises. The unpaired standard error sits near 4.40 throughout — it is computed
from the two columns separately and so cannot notice that they are related at
all.

The rejection rates follow. At $\rho = 0$ the two tests are equivalent, and the
paired one is marginally worse because it has 9 degrees of freedom against 18.
At $\rho = 0.9$ the paired test rejects about 72% of the time and the unpaired
test almost never does.

That last figure deserves a moment. The unpaired test does not merely fail to
improve as the correlation rises; it gets *worse*. With strongly correlated
columns the observed difference is tightly concentrated near its true value of
4, while the standard error the unpaired test uses stays at 4.4 — so its
statistic sits reliably near $4/4.4 \approx 0.9$ and almost never strays past
the critical value. Ignoring a strong pairing produces a test that is
consistently, confidently unable to see anything.
:::


## Question bank {#ts-question-bank}

::: {.watchout}
**No answers are given for this section, by design.**
:::

### Multiple choice {-}

**QB 5.1** The standard error of $\bar{X} - \bar{Y}$ for independent samples is

a. $\dfrac{\sigma_x}{\sqrt{n}} - \dfrac{\sigma_y}{\sqrt{m}}$
b. $\sqrt{\dfrac{\sigma_x^2}{n} + \dfrac{\sigma_y^2}{m}}$
c. $\sqrt{\dfrac{\sigma_x^2}{n} - \dfrac{\sigma_y^2}{m}}$
d. $\dfrac{\sigma_x + \sigma_y}{\sqrt{n + m}}$

**QB 5.2** The pooled two-sample $t$-test uses how many degrees of freedom?

a. $n + m$
b. $n + m - 1$
c. $n + m - 2$
d. $\min(n, m) - 1$

**QB 5.3** A 95% confidence interval for $\mu_x - \mu_y$ is $[-1.2,\ 4.8]$. At
the 5% level we

a. reject $H_0: \mu_x = \mu_y$
b. fail to reject $H_0: \mu_x = \mu_y$
c. cannot tell without the $p$-value
d. conclude the two means are equal

**QB 5.4** `t.test(x, y)` in R by default

a. assumes equal variances
b. performs Welch's test
c. performs a paired test
d. performs a one-tailed test

**QB 5.5** Which of the following are paired designs? *(select all that apply)*

a. Blood pressure of 40 patients before and after a drug
b. Test scores of 40 boys and 45 girls
c. Yields on 20 plots under one seed and 20 different plots under another
d. Household expenditure in the same 200 households in 2019 and 2023

**QB 5.6** Ignoring pairing in a genuinely paired design usually

a. makes it easier to reject the null
b. makes it harder to reject the null
c. has no effect on the $p$-value
d. reverses the sign of the difference

**QB 5.7** Two groups differ with $p < 0.001$ in a sample of 50,000. This
implies

a. the difference is large
b. the difference is important
c. the difference is unlikely to be sampling variation alone
d. one group caused the other's outcome

### Short reasoning {-}

**QB 5.8** Explain why subtracting one sample mean from another increases
variability rather than reducing it, and give the algebraic reason.

**QB 5.9** A researcher has two samples with $s_x = 2$ and $s_y = 9$, and uses
the pooled test. State the assumption being made, why it is doubtful here, and
what the consequence is.

**QB 5.10** A newspaper reports that graduates of a coaching institute score 40
marks higher on average than non-attendees, with a very small $p$-value.
Explain what the comparison can and cannot support.

**QB 5.11** In a paired design, the two columns are strongly positively
correlated. Explain, without algebra, why this makes the paired test more
powerful than the unpaired one.

**QB 5.12** A trial assigns a new fertiliser to plots by lottery. Explain why
the difference in mean yields can be read as an effect here, when the same
comparison in Section \@ref(ts-causal) could not.

**QB 5.13** Two studies estimate the same difference. One reports
$[0.1,\ 9.9]$ and the other $[4.8,\ 5.2]$. Both exclude zero. Explain what
distinguishes them and which is more useful.

### Numerical {-}

**QB 5.14** Government school students score a mean of 62.4 with $s = 11.2$ in a
sample of 45. Private school students score 66.1 with $s = 9.8$ in a sample of
52.

i. Compute the standard error of the difference.
ii. Compute the test statistic.
iii. Test at the 5% level whether the means differ.
iv. Construct a 95% confidence interval, and state whether it agrees with (iii).

**QB 5.15** A sample of 8 plots under a new seed yields a mean of 41.2 quintals
with $s = 3.1$. A sample of 10 plots under the old seed yields 37.8 with
$s = 2.9$.

i. Compute the pooled variance estimate and $s_p$.
ii. Compute the pooled test statistic and its degrees of freedom.
iii. Test at the 5% level whether the new seed yields more, stating your
     hypotheses carefully.
iv. Would Welch's test use more or fewer degrees of freedom here, and why?

**QB 5.16** Twelve workers are timed on a task before and after training. The
mean reduction is 4.2 minutes with a standard deviation of differences of 5.1
minutes.

i. State the hypotheses for a test that training reduced completion time.
ii. Compute the test statistic and its degrees of freedom.
iii. Test at the 5% level.
iv. Explain why the standard deviation of the *differences* is the right
    quantity here rather than the standard deviations of the two rounds.

**QB 5.17** Using `data/attendance-grades.csv`, write R code to compare
cumulative GPA between students above and below the median attendance. Report
the interval, then write a short paragraph stating precisely what your result
establishes and what it does not.

### Paired designs {-}

**QB 5.18** A paired $t$-test on 15 pairs has how many degrees of freedom?

a. 13
b. 14
c. 28
d. 30

**QB 5.19** The paired $t$-test is equivalent to

a. a two-sample $t$-test with `var.equal = TRUE`
b. a one-sample $t$-test on the differences
c. Welch's test with adjusted degrees of freedom
d. a $Z$-test on the difference of means

**QB 5.20** Pairing gives the largest gain in precision when

a. the two sample sizes are equal
b. the two columns are strongly correlated
c. the two columns are independent
d. the sample is large

**QB 5.21** Which analysis is appropriate? *(match each design to a test)*

i. Yields on 20 plots under seed A and on 20 *different* plots under seed B
ii. Yields on the same 20 plots in two consecutive seasons, one seed each
iii. Blood pressure of 40 patients, half given a drug and half a placebo
iv. Blood pressure of 40 patients, each measured before and after a drug

a. Two-sample $t$-test
b. Paired $t$-test

**QB 5.22** A researcher analyses genuinely paired data as two independent
samples. The most likely consequence is

a. a $p$-value that is too small
b. a $p$-value that is too large
c. a confidence interval that is too narrow
d. no change, since the sample means are the same either way

### Paired reasoning {-}

**QB 5.23** Explain why the paired test uses the standard deviation of the
differences rather than the standard deviations of the two columns, and what
this achieves.

**QB 5.24** A colleague argues that pairing should always be used because it is
"more powerful". Under what circumstances is that claim false, and what is the
cost of pairing when the two columns are unrelated?

**QB 5.25** A study reports before-and-after measurements on the same 12
villages and finds a significant improvement. A critic replies that there was
no comparison group, so nothing has been shown. Explain what the paired test
does establish here and what the critic is right about.

### Paired numerical {-}

**QB 5.26** Nine households report monthly food expenditure, in rupees, before
and after a ration card reform.

| Household | Before | After |
|--:|--:|--:|
| 1 | 4200 | 4050 |
| 2 | 3800 | 3750 |
| 3 | 5100 | 4800 |
| 4 | 4600 | 4500 |
| 5 | 3900 | 3700 |
| 6 | 4400 | 4250 |
| 7 | 5000 | 4650 |
| 8 | 4100 | 4050 |
| 9 | 4700 | 4400 |

i. Compute the differences, $\bar{d}$ and $s_d$.
ii. Test at the 5% level whether expenditure changed.
iii. Construct a 95% confidence interval for the mean change.
iv. State one reason the change should not be attributed to the reform.

**QB 5.27** A paired study on $n = 16$ pairs gives $\bar{d} = 3.5$ and
$s_d = 6.4$.

i. Compute the test statistic and its degrees of freedom.
ii. Test at the 5% and 1% levels.
iii. How large would $n$ have to be for the same $\bar{d}$ and $s_d$ to reject
     at the 1% level?

**QB 5.28** Two analyses of the same twelve paired observations report
$s_d = 2.1$ and, from the two columns separately, $s_x = 9.4$ and $s_y = 9.1$.

i. Compute the standard error the paired test would use.
ii. Compute the standard error an unpaired test would use.
iii. Explain the ratio between them in terms of what pairing removes.

**QB 5.29** Using `data/student-survey-sleep.csv`, write R code that
deliberately performs a paired test on two variables that are *not* paired in
any meaningful sense. Report what R does, and explain why the output is
meaningless even though no error is raised.


## Looking ahead {#ts-looking-ahead}

We can now compare two groups properly. We can also say exactly what such a
comparison does not deliver, which turns out to be the more important lesson.

Two limitations remain, and both point the same way.

The first is that comparing two groups uses the data wastefully. Splitting the
680 students at the median attendance threw away everything we knew about *how
much* each student attended, reducing a rich variable to a coin flip. A student
who attended 32 classes and one who attended 20 were treated as identical.

The second is selection. The attendance groups differed in attendance and in
everything correlated with it, and a two-sample test cannot separate them
because it knows about nothing else.

Both problems have the same solution. Instead of splitting a variable in two
and comparing averages, we can model the relationship between the whole of one
variable and the whole of another — and, crucially, do so while holding other
variables fixed.

That is regression, and it occupies the rest of this book.
