# Simple Regression {#simple-regression}

```{r setup-06, include = FALSE}
source("_common.R")
library(ggplot2)
```

Throughout this book we have asked whether differences between groups are
larger than we would expect from chance alone. Regression asks a broader
question: **how is one variable related to another?**

That shift matters because economic theory usually makes qualitative
predictions — higher prices reduce demand, more education increases earnings —
and qualitative predictions are hard to act on. Regression lets us estimate *how
much* one variable moves with another, express it in units someone can use, and
attach a measure of uncertainty to the answer.

In this unit we build it out of ideas we already have: variation, sampling and
statistical inference. Nothing genuinely new is required. What changes is the
object of study, from a mean or a difference of means to a relationship.

The result is the central tool of **econometrics** — the field concerned with
measuring economic relationships from data — though the tool itself is used
across the social sciences and beyond.

## From comparing groups to modelling relationships {#sr-motivation}

::: {.question}
We can compare two groups. Why is that not enough?
:::

Unit \@ref(two-sample) compared students above and below the median attendance.
Those who attended more had higher average grades, by more than sampling
variation alone would explain.

But knowing that two groups differ is only the beginning. We still do not know
how grades change *as* attendance changes, or whether attendance itself is
responsible for the difference.

That second gap is the deeper one, and it is worth seeing clearly.

### The experiment we cannot run {-}

What we would really like is a thought experiment. Take one student and observe
them twice: once attending 20 classes and once attending 30, with everything
else about them unchanged — the same ability, the same motivation, the same
teacher, the same week. The difference in their grades would isolate the effect
of attendance itself.

That experiment is impossible. A student cannot attend both 20 and 30 classes.
Instead, we observe *different* students who differ in many ways at once. The
challenge of empirical economics is to separate the effect of attendance from
all the other differences between those students.

::: {.watchout}
Simple regression does not solve that problem. Section \@ref(sr-causal) returns
to this point in detail. If a reader finishes this unit believing that regression
automatically uncovers causal effects, then this book has failed them.

What regression does is change the question we ask of the data, and build the
machinery that later units extend towards the thought experiment above.
:::

### The question has changed {-}

Instead of asking whether students with high attendance differ from those with
low attendance, we now ask whether grades tend to rise as attendance rises.

Attendance is no longer a group label. It becomes a **numerical variable whose
entire range carries information.**

That is the conceptual step from a $t$-test to a regression, and everything
technical in this unit follows from it.

### The price of the old question {-}

To compare two groups, Unit \@ref(two-sample) took a variable recording how many
classes each of 680 students attended — a number running from 2 to 32 — and
replaced it with a label saying "high" or "low". A student who attended 32
classes and one who attended 28 were treated as identical, while one who
attended 28 and one who attended 27 were placed in opposite groups.

We can quantify the price of that simplification.

```{r sr-waste}
students <- read.csv("data/attendance-grades.csv")
students$high <- students$attendance >= median(students$attendance)

c(split_t      = abs(t.test(sem_gpa ~ high, data = students)$statistic),
  regression_t = summary(lm(sem_gpa ~ attendance, data = students))$coefficients[2, 3])
```

The same data. The same outcome. Almost the same question. Splitting attendance
into two groups produces a $t$ statistic of 11.8; using the full attendance
variable produces one of 17.6.

The relationship was always present — the median split simply discarded much of
the evidence for it.

::: {.keyidea}
Regression changes the question we ask, and in doing so improves on a two-sample
comparison in two ways.

- It uses the **full** explanatory variable rather than reducing it to two
  groups.
- It estimates a **slope** — the expected change in $y$ for a one-unit increase
  in $x$ — instead of merely comparing two averages.
:::

In the next unit we will add more explanatory variables, allowing us to compare
observations while holding other measured characteristics fixed. That brings us
closer to the counterfactual comparison we would like to make, although — as we
shall see — it does not eliminate the problem entirely.

## The population regression function {#sr-prf}

::: {.question}
What is the thing we are trying to estimate?
:::

Before fitting anything, it is worth being clear about what a regression is
looking for. It is not looking for a line through a cloud of points. It is
looking for the way the *average* of $y$ moves with $x$.

Group the students by attendance and compute the average semester GPA within
each group.

```{r sr-condmeans}
students$band <- cut(students$attendance,
                     breaks = c(0, 20, 24, 27, 30, 33),
                     labels = c("2–20", "21–24", "25–27", "28–30", "31–32"))

aggregate(sem_gpa ~ band, data = students,
          function(x) round(c(n = length(x), mean_gpa = mean(x)), 2))
```

Two things stand out, and both matter.

**Within each band, grades vary enormously.** Students who attended the same
number of classes did not get the same grade. Attendance plainly does not
determine grades.

**Across bands, the averages rise steadily** — from 4.38 to 7.67. The
*conditional mean* of GPA increases with attendance, even though individual
grades scatter widely around it.

That second pattern is what regression is about.

::: {.keyidea}
$E(y \mid x)$ — the average of $y$ among all members of the population with a
given value of $x$ — is called the **conditional mean** of $y$ given $x$.

Traced out across all values of $x$, it forms the **population regression
function** (PRF).

Regression estimates the conditional mean. It does not claim to predict
individuals, and the scatter around the line is not a defect in the method: it
is the part of $y$ that $x$ does not account for.
:::

```{r sr-prf-plot, echo = FALSE, fig.height = 3.8, fig.cap = "Semester GPA against attendance for 680 students. The points scatter widely at every level of attendance. The band averages, in orange, rise steadily -- and it is those averages that a regression line estimates."}
band_means <- aggregate(cbind(sem_gpa, attendance) ~ band, data = students, mean)

ggplot(students, aes(attendance, sem_gpa)) +
  geom_point(alpha = 0.18, colour = book_palette$ink, size = 1.1) +
  geom_point(data = band_means, aes(attendance, sem_gpa),
             colour = book_palette$reject, size = 3.4) +
  geom_line(data = band_means, aes(attendance, sem_gpa),
            colour = book_palette$reject, linewidth = 0.9) +
  labs(x = "Classes attended", y = "Semester GPA") +
  theme_book()
```

Notice that the orange points rise but not in a perfectly straight line. The
first band sits well below the others, and the second and third are almost
level. Nothing forced the conditional mean to be linear, and here it is not
exactly so.

This matters for what we are about to do. A *linear* regression describes the
conditional mean with a straight line, which is an approximation to the pattern
above rather than a description of it. That approximation is a modelling choice,
and one worth examining rather than assuming — Section \@ref(sr-algebra) returns
to how well it does.

## The model {#sr-model}

::: {.question}
How do we write down "grades depend on attendance, but not only on attendance"?
:::

Economic theory says grades depend on attendance. It also says grades depend on
effort, ability, teaching quality, health, and how many hours a student must
work for pay. Only the first is in our dataset.

The **simple linear regression model** handles this by dividing $y$ into a part
explained by $x$ and a part that is not:

$$y = \beta_0 + \beta_1 x + u$$

- $y$ is the **dependent** variable — semester GPA.
- $x$ is the **independent** or explanatory variable — attendance.
- $u$ is the **error** or disturbance term, holding everything else that affects
  $y$. It is not a mistake; it is the rest of the world.
- $\beta_0$ is the **intercept**, and is rarely of interest in itself.
- $\beta_1$ is the **slope**, and is almost always what we care about. It says
  how much $y$ changes when $x$ rises by one unit.

For the grades example,

$$\textit{GPA} = \beta_0 + \beta_1\,\textit{attendance} + u$$

and $\beta_1$ is the change in GPA associated with attending one more class.

### What must be true of the error term {-}

The model as written is empty until we say something about $u$. Two assumptions
do the work.

**The error has mean zero:** $E(u) = 0$.

This costs nothing. If the unobserved factors had a non-zero average, we could
absorb that average into the intercept and start again.

::: {.deeper}
Suppose $u = e + c$ where $c$ is some non-zero constant — a fixed effect of, say,
generous marking. Then

$$
\begin{aligned}
y &= \beta_0 + \beta_1 x + (e + c) \\[4pt]
&= (\beta_0 + c) + \beta_1 x + e \\[4pt]
&= \beta_0^{*} + \beta_1 x + e
\end{aligned}
$$

with $E(e) = 0$. The intercept has changed and the slope has not. Since $\beta_1$
is what we want, assuming $E(u) = 0$ restricts nothing of importance.
:::

**The error is unrelated to $x$:** $E(u \mid x) = E(u) = 0$.

This one is not free at all. It says that the average of the unobserved factors
is the *same* at every value of $x$.

::: {.keyidea}
The **zero conditional mean assumption**, $E(u \mid x) = 0$, is the assumption on
which everything rests.

For the grades example it says: students who attend more classes are, on
average, no more and no less able, motivated, or well-taught than students who
attend fewer.

If that fails, $\beta_1$ does not measure the effect of attendance. It measures
the effect of attendance plus whatever else differs between high and low
attenders.
:::

Stated in that concrete form, the assumption is clearly doubtful here. Students
who attend regularly are plausibly more motivated, and motivation raises grades
on its own. We cannot check this, because effort is not in the dataset — that is
what "unobserved" means.

Section \@ref(sr-causal) returns to what follows. For now we proceed, keeping in
mind that we are estimating a conditional mean, and that whether it is an
*effect* is a separate question.

## From the population to a sample {#sr-srf}

::: {.question}
The PRF describes a population we cannot see. What do we actually have?
:::

The parameters $\beta_0$ and $\beta_1$ belong to the population. They are fixed,
unknown numbers — exactly like $\mu$ in Unit \@ref(estimation) — and we have a
sample.

From the sample we compute estimates $\hat{\beta}_0$ and $\hat{\beta}_1$, and
these define the **sample regression function**:

$$\hat{y}_i = \hat{\beta}_0 + \hat{\beta}_1 x_i$$

::: {.keyidea}
| Population | Sample |
|:--|:--|
| $y = \beta_0 + \beta_1 x + u$ | $y_i = \hat{\beta}_0 + \hat{\beta}_1 x_i + \hat{u}_i$ |
| $E(y \mid x) = \beta_0 + \beta_1 x$ (the PRF) | $\hat{y}_i = \hat{\beta}_0 + \hat{\beta}_1 x_i$ (the SRF) |
| $u$ is the **error**, and is unobservable | $\hat{u}_i$ is the **residual**, and is computed |
| $\beta_1$ is a parameter | $\hat{\beta}_1$ is an estimator, with a sampling distribution |

The distinction between $u$ and $\hat{u}$ matters. The error is the deviation
from the true line, which nobody sees. The residual is the deviation from our
fitted line, which we can print.
:::

Every sample gives a different line. So $\hat{\beta}_1$ is a random variable
with a sampling distribution, and every question from Units
\@ref(estimation)–\@ref(one-sample) applies to it: is it unbiased, what is its
standard error, and how do we test a claim about it.

## Choosing a line: ordinary least squares {#sr-ols}

::: {.question}
Infinitely many lines pass through a scatter of points. Which one?
:::

A residual measures how badly the line misses observation $i$:

$$\hat{u}_i = y_i - \hat{y}_i = y_i - \hat{\beta}_0 - \hat{\beta}_1 x_i$$

We want the residuals small. The obvious criterion is to make their sum as close
to zero as possible.

That criterion fails, and it is worth seeing why.

::: {.watchout}
Consider two lines. The first misses four points by $+1, -1, +1, -1$. The second
misses them by $+8, -8, +8, -8$.

Both have residuals summing to exactly zero. By that criterion the two lines are
equally good, which is absurd — the second is dreadful.

The problem is that positive and negative misses cancel. A criterion that lets
errors offset one another can be satisfied by a line that is nowhere near the
data.
:::

The fix is the one from Section \@ref(u1-describing): square the deviations
first, so that they cannot cancel and large misses count for more than small
ones. We choose $\hat{\beta}_0$ and $\hat{\beta}_1$ to minimise

$$L = \sum_{i=1}^{n} \hat{u}_i^2 = \sum_{i=1}^{n}\left(y_i - \hat{\beta}_0 - \hat{\beta}_1 x_i\right)^2$$

This is the **method of ordinary least squares**, or OLS.

::: {.theory}
Minimising $L$ means setting both partial derivatives to zero.

**With respect to the intercept:**

$$
\begin{aligned}
\frac{\partial L}{\partial \hat{\beta}_0}
&= \sum_{i=1}^{n} -2\left(y_i - \hat{\beta}_0 - \hat{\beta}_1 x_i\right) = 0 \\[4pt]
0 &= \sum_{i=1}^{n} y_i - n\hat{\beta}_0 - \hat{\beta}_1\sum_{i=1}^{n} x_i
   && \text{divide by } -2 \text{ and expand} \\[4pt]
0 &= n\bar{y} - n\hat{\beta}_0 - \hat{\beta}_1 n\bar{x}
   && \text{since } \textstyle\sum x_i = n\bar{x} \\[4pt]
\hat{\beta}_0 &= \bar{y} - \hat{\beta}_1\bar{x}
   && \text{divide by } n \text{ and rearrange}
\end{aligned}
$$

**With respect to the slope:**

$$
\begin{aligned}
\frac{\partial L}{\partial \hat{\beta}_1}
&= \sum_{i=1}^{n} -2x_i\left(y_i - \hat{\beta}_0 - \hat{\beta}_1 x_i\right) = 0 \\[4pt]
0 &= \sum x_i y_i - \hat{\beta}_0\sum x_i - \hat{\beta}_1 \sum x_i^2 \\[4pt]
0 &= \sum x_i y_i - \left(\bar{y} - \hat{\beta}_1\bar{x}\right)n\bar{x} - \hat{\beta}_1\sum x_i^2
   && \text{substitute } \hat{\beta}_0 \\[4pt]
0 &= \sum x_i y_i - n\bar{x}\bar{y} - \hat{\beta}_1\left(\sum x_i^2 - n\bar{x}^2\right) \\[4pt]
\hat{\beta}_1 &= \frac{\sum x_i y_i - n\bar{x}\bar{y}}{\sum x_i^2 - n\bar{x}^2}
\end{aligned}
$$

Both parts of that fraction can be written more meaningfully. Expanding the
numerator,

$$
\begin{aligned}
\sum (x_i - \bar{x})(y_i - \bar{y})
&= \sum x_i y_i - \bar{y}\sum x_i - \bar{x}\sum y_i + n\bar{x}\bar{y} \\[4pt]
&= \sum x_i y_i - n\bar{x}\bar{y}
\end{aligned}
$$

and the denominator,

$$\sum (x_i - \bar{x})^2 = \sum x_i^2 - 2\bar{x}\sum x_i + n\bar{x}^2
= \sum x_i^2 - n\bar{x}^2$$

so that

$$\hat{\beta}_1 = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}{\sum_{i=1}^{n}(x_i - \bar{x})^2}$$
:::

::: {.keyidea}
The OLS estimators are

$$\hat{\beta}_1 = \frac{\sum(x_i - \bar{x})(y_i - \bar{y})}{\sum(x_i - \bar{x})^2}
\qquad\qquad
\hat{\beta}_0 = \bar{y} - \hat{\beta}_1\bar{x}$$

The slope is the **sample covariance of $x$ and $y$ divided by the sample
variance of $x$**. It answers: when $x$ is above its mean, is $y$ also above
its mean, and by how much per unit?
:::

::: {.intuition}
That reading explains several things at once.

If $x$ and $y$ move together, the numerator is positive and so is the slope. If
they move oppositely, it is negative. If they are unrelated, the products cancel
and the slope is near zero.

And the denominator explains why $x$ must vary. If every student attended the
same number of classes, $\sum(x_i - \bar{x})^2 = 0$ and the slope is undefined —
there is no way to learn how GPA responds to attendance from data in which
attendance never changes. This is the same requirement as Section
\@ref(u1-randomness): no variation, nothing to learn.
:::

### Doing it by hand and in R {-}

```{r sr-byhand}
x <- students$attendance
y <- students$sem_gpa

b1 <- sum((x - mean(x)) * (y - mean(y))) / sum((x - mean(x))^2)
b0 <- mean(y) - b1 * mean(x)

c(b0 = b0, b1 = b1)
```

```{r sr-lm}
fit <- lm(sem_gpa ~ attendance, data = students)
coef(fit)
```

Identical. `lm()` solves the same two equations.

So $\hat{\beta}_1 = 0.189$: each additional class attended is associated with
about 0.19 of a GPA point. Ten more classes, roughly 1.9 points — which on a
ten-point scale is a great deal.

```{r sr-fit-plot, echo = FALSE, fig.height = 3.8, fig.cap = "The fitted line, with the vertical distances it misses by shown for a handful of students. Ordinary least squares chooses the line making the sum of those squared distances as small as possible."}
set.seed(3)
shown <- students[sample(nrow(students), 25), ]
shown$pred <- predict(fit, newdata = shown)

ggplot(students, aes(attendance, sem_gpa)) +
  geom_point(alpha = 0.13, colour = book_palette$ink, size = 1.1) +
  geom_segment(data = shown, aes(xend = attendance, y = sem_gpa, yend = pred),
               colour = book_palette$reject, linewidth = 0.45) +
  geom_point(data = shown, colour = book_palette$reject, size = 1.5) +
  geom_abline(intercept = coef(fit)[1], slope = coef(fit)[2],
              colour = book_palette$accent, linewidth = 1) +
  labs(x = "Classes attended", y = "Semester GPA") +
  theme_book()
```

## What the fitted line guarantees {#sr-algebra}

::: {.question}
Having minimised the squares, what do we get for free?
:::

Three algebraic facts follow directly from the two equations we just solved.
They hold for *every* OLS regression, in every dataset, whether or not the model
is any good.

::: {.keyidea}
1. **The residuals sum to zero**: $\sum \hat{u}_i = 0$. This is the first
   normal equation, restated.
2. **The residuals are uncorrelated with $x$**: $\sum x_i \hat{u}_i = 0$.
   Whatever pattern is left over, it is not linearly related to $x$ — OLS has
   already extracted all of that.
3. **The line passes through the means**: $\bar{y} = \hat{\beta}_0 +
   \hat{\beta}_1\bar{x}$. The point $(\bar{x}, \bar{y})$ always lies on the
   fitted line.
:::

```{r sr-algebra-check}
u_hat <- residuals(fit)

c(sum_residuals   = sum(u_hat),
  sum_x_times_u   = sum(x * u_hat),
  predicted_at_xbar = coef(fit)[1] + coef(fit)[2] * mean(x),
  mean_of_y         = mean(y))
```

The first two are zero up to floating-point error, and the last two agree.

::: {.watchout}
These are *algebraic* properties, not evidence. They hold by construction, so
finding them in your output tells you the arithmetic worked and nothing more.

In particular, $\sum x_i \hat{u}_i = 0$ is true whether or not
$E(u \mid x) = 0$. The first is forced by the estimator; the second is an
assumption about the world that no computation can verify.
:::

### How much does the line explain? {-}

Decompose the variation in $y$. Each observation's distance from $\bar{y}$
splits into a part the line accounts for and a part it does not:

$$\underbrace{\sum(y_i - \bar{y})^2}_{\text{SST}}
= \underbrace{\sum(\hat{y}_i - \bar{y})^2}_{\text{SSE}}
+ \underbrace{\sum \hat{u}_i^2}_{\text{SSR}}$$

with SST the total sum of squares, SSE the explained sum of squares, and SSR the
residual sum of squares. The cross-term vanishes precisely because of property 2
above.

::: {.keyidea}
The **coefficient of determination** is the share of the variation in $y$ that
the regression accounts for:

$$R^2 = \frac{\text{SSE}}{\text{SST}} = 1 - \frac{\text{SSR}}{\text{SST}}$$

It lies between 0 and 1. In a simple regression it is exactly the square of the
correlation between $x$ and $y$.
:::

```{r sr-r2}
sst <- sum((y - mean(y))^2)
ssr <- sum(u_hat^2)

c(r_squared      = 1 - ssr / sst,
  from_summary   = summary(fit)$r.squared,
  cor_squared    = cor(x, y)^2)
```

So attendance accounts for about 31% of the variation in semester GPA. The other
69% is everything else — ability, effort, subject, luck.

::: {.watchout}
A low $R^2$ does not mean a regression is useless, and a high one does not mean
it is right.

$R^2$ measures how tightly the points cluster around the line. It says nothing
about whether $\beta_1$ is estimated accurately, and nothing whatever about
whether the relationship is causal. A regression of grades on shoe size might
achieve a respectable $R^2$ in a sample of children, where both rise with age.

In economics, $R^2$ values of 0.1 or 0.2 are common and unremarkable. Human
behaviour is not tidy.
:::

## How good is the estimate? {#sr-properties}

::: {.question}
$\hat{\beta}_1$ came from one sample. What would it do across many?
:::

We now ask about $\hat{\beta}_1$ exactly what Unit \@ref(estimation) asked about
$\bar{X}$: is it centred on the truth, and how much does it vary?

The answers require assumptions, which are worth listing plainly.

::: {.keyidea}
**The assumptions of simple regression.**

**SLR1 — linear in parameters.** The population model is
$y = \beta_0 + \beta_1 x + u$.

**SLR2 — random sampling.** The $n$ observations are a random sample from that
population.

**SLR3 — variation in $x$.** The $x_i$ are not all equal.

**SLR4 — zero conditional mean.** $E(u \mid x) = 0$.

**SLR5 — homoskedasticity.** $\mathrm{Var}(u \mid x) = \sigma^2$, the same at
every value of $x$.

The first four give unbiasedness. The fifth is needed only for the variance
formula.
:::

### Unbiasedness {-}

::: {.theory}
Write the slope in deviation form and substitute the model. With
$\text{SST}_x = \sum(x_i - \bar{x})^2$,

$$
\begin{aligned}
\hat{\beta}_1 &= \frac{\sum(x_i - \bar{x})y_i}{\text{SST}_x}
   && \text{since } \textstyle\sum(x_i-\bar{x})\bar{y} = 0 \\[4pt]
&= \frac{\sum(x_i - \bar{x})(\beta_0 + \beta_1 x_i + u_i)}{\text{SST}_x}
   && \text{substituting SLR1} \\[4pt]
&= \beta_1 + \frac{\sum(x_i - \bar{x})u_i}{\text{SST}_x}
\end{aligned}
$$

because $\sum(x_i - \bar{x}) = 0$ kills the $\beta_0$ term and
$\sum(x_i - \bar{x})x_i = \text{SST}_x$ leaves $\beta_1$.

That last line is worth reading carefully. **The estimate equals the truth plus
a term built entirely from the errors.** Taking expectations conditional on the
$x$'s, and using SLR4 to set $E(u_i \mid x) = 0$,

$$E(\hat{\beta}_1) = \beta_1 + \frac{\sum(x_i - \bar{x})E(u_i \mid x)}{\text{SST}_x} = \beta_1$$
:::

So OLS is unbiased: across repeated samples, $\hat{\beta}_1$ is centred on
$\beta_1$. And the derivation shows exactly where that can fail. If the errors
are systematically larger at larger $x$ — if high attenders are more able — then
$E(u_i \mid x) \neq 0$, the second term does not vanish, and $\hat{\beta}_1$ is
biased by precisely that amount.

### The variance of the slope {-}

::: {.theory}
From the same expression, and treating the $x$'s as fixed,

$$
\begin{aligned}
\mathrm{Var}(\hat{\beta}_1)
&= \mathrm{Var}\!\left(\frac{\sum(x_i - \bar{x})u_i}{\text{SST}_x}\right) \\[4pt]
&= \frac{1}{\text{SST}_x^2}\sum(x_i - \bar{x})^2\,\mathrm{Var}(u_i)
   && \text{independence} \\[4pt]
&= \frac{\sigma^2 \,\text{SST}_x}{\text{SST}_x^2}
   && \text{homoskedasticity, SLR5} \\[4pt]
&= \frac{\sigma^2}{\text{SST}_x}
\end{aligned}
$$
:::

::: {.keyidea}
$$\mathrm{Var}(\hat{\beta}_1) = \frac{\sigma^2}{\sum(x_i - \bar{x})^2}$$

The slope is estimated more precisely when the errors are small (small
$\sigma^2$) and when $x$ varies a lot (large $\text{SST}_x$).
:::

::: {.intuition}
The second half is the useful insight. To learn how $y$ responds to $x$, you
need to observe $x$ over a wide range.

Comparing students who attended 27 classes with students who attended 28 tells
you very little. Comparing those who attended 5 with those who attended 30 tells
you a great deal. Spread in the explanatory variable is *information*, and this
is why the median split of Section \@ref(sr-motivation) performed so much worse
— it discarded most of the spread.
:::

Since $\sigma^2$ is unknown, it is estimated from the residuals. Two parameters
were estimated on the way, so the divisor is $n - 2$:

$$\hat{\sigma}^2 = \frac{\sum \hat{u}_i^2}{n - 2}
\qquad\qquad
\mathrm{se}(\hat{\beta}_1) = \frac{\hat{\sigma}}{\sqrt{\sum(x_i - \bar{x})^2}}$$

The $n - 2$ is the same accounting as Section \@ref(u3-variance), where
estimating one mean cost one degree of freedom. Here we estimated an intercept
and a slope.

```{r sr-se-byhand}
n <- length(y)
sigma_hat <- sqrt(sum(u_hat^2) / (n - 2))
se_b1     <- sigma_hat / sqrt(sum((x - mean(x))^2))

c(sigma_hat = sigma_hat, se_b1 = se_b1,
  se_from_lm = summary(fit)$coefficients[2, 2])
```

## Inference about the slope {#sr-inference}

::: {.question}
Is the slope we found distinguishable from zero?
:::

$\hat{\beta}_1$ is an estimator with a standard error, so everything in Unit
\@ref(one-sample) applies without modification.

The hypothesis of interest is almost always that $x$ has no effect at all:

$$H_0 : \beta_1 = 0 \qquad H_A : \beta_1 \neq 0$$

and the test statistic has the form it always has — estimate minus hypothesised
value, over standard error:

$$t = \frac{\hat{\beta}_1 - 0}{\mathrm{se}(\hat{\beta}_1)}$$

compared against $t$ with $n - 2$ degrees of freedom.

```{r sr-inference}
summary(fit)$coefficients
```

$\hat{\beta}_1 = 0.189$ with a standard error of 0.0107, giving $t = 17.6$ on
678 degrees of freedom and a $p$-value indistinguishable from zero. We reject
the claim that attendance is unrelated to GPA.

The confidence interval is the more informative report:

```{r sr-confint}
confint(fit, level = 0.95)
```

Each additional class is associated with between 0.168 and 0.210 of a GPA point,
with 95% confidence. The interval excludes zero, agreeing with the test as it
must.

::: {.watchout}
"Statistically significant" is a claim about distinguishability from zero, not
about magnitude. With 680 observations, very small slopes clear the bar easily.

Always ask what the slope means in the units of the problem. Here 0.19 GPA
points per class is large. In the regression of final examination score on
attendance in the same dataset, the slope is 0.117 marks per class and equally
"significant" — and 0.117 marks is nothing at all.
:::

## What regression does not fix {#sr-causal}

::: {.question}
We have a slope, a standard error, and a tiny $p$-value. Do we have an effect?
:::

No, and it is worth being exact about why.

Everything in this unit estimates a conditional mean. We have established, with
considerable precision, that students who attend more classes have higher
average GPAs. That is a fact about the population.

The claim that attending more classes *raises* a student's GPA is a different
claim, and it requires SLR4: $E(u \mid x) = 0$. Section \@ref(sr-model) put that
assumption in concrete terms — students who attend more must be no more
motivated or able, on average, than students who attend less.

Nobody believes that.

::: {.keyidea}
When $E(u \mid x) \neq 0$, the unbiasedness derivation shows exactly what
happens. We had

$$\hat{\beta}_1 = \beta_1 + \frac{\sum(x_i - \bar{x})u_i}{\text{SST}_x}$$

and the second term no longer has expectation zero. Its expectation is the
**bias**, and its sign is the sign of the relationship between $x$ and the
omitted factors.

If motivated students both attend more *and* score higher for reasons unrelated
to attendance, then $\hat{\beta}_1$ overstates the effect of attendance. This is
**omitted variable bias**.
:::

Notice what this does and does not affect. The estimate is still the best
description of the conditional mean. The standard error is still right. The
$p$-value is still right. What is wrong is the *interpretation* — and no
diagnostic in the regression output will tell you so, because the problem lies
in a variable that is not in the data.

Two routes lead out, and both come later.

The first is to measure the confounders and hold them fixed, which is multiple
regression, the subject of the next unit. It works only for confounders you can
observe.

The second is to find a source of variation in $x$ that has nothing to do with
the unobservables — a lottery, a rule, an accident of administration. That is
the subject of the second part of this book.

::: {.intuition}
Regression is a description of how averages move together. Turning it into a
statement about causes requires an argument about where the variation in $x$
came from, and that argument is made by the researcher, not by the software.
:::

## Summary {#sr-summary}

::: {.keyidea}
- Regression uses the whole of an explanatory variable rather than splitting it
  in two, reports a slope in interpretable units, and can be extended to hold
  other variables fixed. On the same data, the median split gave $t = 11.8$ and
  the regression $t = 17.6$.
- The **population regression function** is the conditional mean
  $E(y \mid x) = \beta_0 + \beta_1 x$. Regression estimates how the average of
  $y$ moves with $x$; the scatter around it is the part $x$ does not account
  for.
- The model $y = \beta_0 + \beta_1 x + u$ splits $y$ into an explained part and
  an error containing everything else. $E(u) = 0$ costs nothing;
  $E(u \mid x) = 0$ costs everything.
- **OLS** minimises $\sum \hat{u}_i^2$. Minimising $\sum \hat{u}_i$ instead
  fails, because positive and negative misses cancel and a terrible line can
  satisfy it.
- $\hat{\beta}_1 = \dfrac{\sum(x_i-\bar{x})(y_i-\bar{y})}{\sum(x_i-\bar{x})^2}$,
  the sample covariance over the sample variance of $x$, and
  $\hat{\beta}_0 = \bar{y} - \hat{\beta}_1\bar{x}$.
- By construction $\sum\hat{u}_i = 0$, $\sum x_i\hat{u}_i = 0$, and the line
  passes through $(\bar{x}, \bar{y})$. These are algebraic and are not evidence
  of anything.
- $R^2 = 1 - \text{SSR}/\text{SST}$ is the share of variation explained. It
  measures fit, not correctness, and says nothing about causality.
- Under SLR1–SLR4, $E(\hat{\beta}_1) = \beta_1$. Under SLR5 as well,
  $\mathrm{Var}(\hat{\beta}_1) = \sigma^2/\sum(x_i - \bar{x})^2$ — so spread in
  $x$ buys precision. $\hat{\sigma}^2 = \sum\hat{u}_i^2/(n-2)$.
- Testing $H_0: \beta_1 = 0$ uses $t = \hat{\beta}_1/\mathrm{se}(\hat{\beta}_1)$
  on $n - 2$ degrees of freedom, exactly as in Unit \@ref(one-sample).
- $\hat{\beta}_1 = \beta_1 + \sum(x_i-\bar{x})u_i/\text{SST}_x$. When
  $E(u\mid x) \neq 0$ the second term is **omitted variable bias**, and nothing
  in the output reveals it.
:::

## Practice problems {#sr-practice}

::: {.yourturn}
Work each one before opening the solution. The
[question bank](#sr-question-bank) that follows has no answers.
:::

### Concept checks {-}

**6.1** Why does regression estimate a conditional mean rather than predict
individual outcomes?

::: {.answer}
Because $y$ is not determined by $x$. At any given attendance level, students
obtain a wide range of grades, because grades also depend on effort, ability and
much else — all of which sit in the error term.

What is stable across the population is the *average* of $y$ at each value of
$x$. That is what $E(y \mid x)$ describes and what the regression line
estimates.

The scatter around the line is therefore not a failure of the method. It is the
part of $y$ that $x$ does not account for, and a model claiming otherwise would
be claiming that attendance alone determines grades.
:::

**6.2** Why is minimising the sum of residuals a bad criterion?

::: {.answer}
Because positive and negative residuals cancel, so a line that misses badly in
both directions can have residuals summing to zero.

Misses of $+8, -8, +8, -8$ sum to zero, and so do misses of $+1, -1, +1, -1$. By
that criterion the two lines are equally good, though the first is far worse.

Squaring removes the cancellation and makes large misses count more than small
ones, which is what we want from a measure of fit.
:::

**6.3** What goes wrong if every observation has the same value of $x$?

::: {.answer}
The slope is undefined. Its denominator is $\sum(x_i - \bar{x})^2$, which is
zero when every $x_i$ equals $\bar{x}$, and the estimate would require dividing
by zero.

The statistical content is more interesting than the algebra. If attendance
never varies, the data contain no information about how GPA responds to
attendance — there is nothing to compare. This is assumption SLR3, and it is the
same point as Section \@ref(u1-randomness): where nothing varies, nothing can be
learned.
:::

**6.4** Distinguish the error $u_i$ from the residual $\hat{u}_i$.

::: {.answer}
The error is the deviation of $y_i$ from the *population* regression line:
$u_i = y_i - \beta_0 - \beta_1 x_i$. It involves the true parameters and is
therefore unobservable.

The residual is the deviation from our *fitted* line:
$\hat{u}_i = y_i - \hat{\beta}_0 - \hat{\beta}_1 x_i$. It is computed from the
data and can be printed.

The relationship mirrors $\mu$ against $\bar{x}$. Residuals are estimates of
errors, and confusing the two leads to claims like "the errors are uncorrelated
with $x$, so SLR4 holds" — which is wrong, because the *residuals* are
uncorrelated with $x$ by construction, whatever the errors do.
:::

**6.5** A regression reports $R^2 = 0.04$ and a slope significant at the 1%
level. Is that a contradiction?

::: {.answer}
No. They measure different things.

$R^2$ says the regression accounts for 4% of the variation in $y$ — the points
scatter widely around the line. The $p$-value says the slope is distinguishable
from zero, which depends on the standard error and therefore on the sample size.

With a large sample, a real but small relationship produces exactly this
pattern: a reliably estimated slope explaining little of the variation. Both
statements are true and neither is more important; they answer "how much does
$x$ account for?" and "is there a relationship at all?"
:::

### Explain {-}

**6.6** Explain why $\mathrm{Var}(\hat{\beta}_1)$ falls as the spread of $x$
increases, and what this implies for research design.

::: {.answer}
The formula is $\mathrm{Var}(\hat{\beta}_1) = \sigma^2/\sum(x_i - \bar{x})^2$,
so spread in $x$ appears in the denominator and larger spread means a smaller
variance.

The intuition is that a slope is a rate of change, and rates of change are
measured by observing change. If every student attended between 27 and 29
classes, we would be trying to infer the effect of attendance from almost no
variation in attendance, and small errors in $y$ would translate into wild
swings in the fitted slope. Observing attendance from 5 to 30 pins the line down.

For design, this says that variation in the explanatory variable is valuable and
should be sought rather than avoided. A study of the effect of class size that
only observes classes of 39 to 41 students will struggle no matter how many
schools it covers. It also explains why the median split did worse than the
regression: collapsing attendance to two categories discards most of its spread.
:::

**6.7** State the zero conditional mean assumption for a regression of earnings
on years of schooling, in concrete terms, and say whether you believe it.

::: {.answer}
The model is $\textit{earnings} = \beta_0 + \beta_1\,\textit{schooling} + u$,
where $u$ holds everything else affecting earnings — ability, family
connections, school quality, motivation, region.

$E(u \mid \textit{schooling}) = 0$ says that people with more schooling are, on
average, no more able, no better connected, and no better taught than people
with less.

That is not believable. Schooling is not distributed at random: children of
wealthier and better-connected families obtain more of it, and those advantages
raise earnings on their own. So $u$ is positively related to schooling, and
$\hat{\beta}_1$ overstates the return to education.

This does not make the regression useless. It correctly describes how average
earnings differ across schooling levels. It simply does not measure what an
extra year of schooling would do to a given person.
:::

### Numerical problems {-}

**6.8** For five plots, fertiliser applied ($x$, kg) and yield ($y$, quintals):

| $x$ | 2 | 4 | 6 | 8 | 10 |
|:--|:--|:--|:--|:--|:--|
| $y$ | 4 | 7 | 8 | 12 | 14 |

i. Compute $\bar{x}$, $\bar{y}$, $\hat{\beta}_1$ and $\hat{\beta}_0$.
ii. Write the fitted line and interpret the slope.
iii. Compute the fitted values and residuals, and verify they sum to zero.
iv. Compute $R^2$.

::: {.answer}
**(i)** $\bar{x} = 6$, $\bar{y} = 9$.

| $x_i$ | $y_i$ | $x_i - \bar{x}$ | $y_i - \bar{y}$ | product | $(x_i-\bar{x})^2$ |
|--:|--:|--:|--:|--:|--:|
| 2 | 4 | $-4$ | $-5$ | 20 | 16 |
| 4 | 7 | $-2$ | $-2$ | 4 | 4 |
| 6 | 8 | 0 | $-1$ | 0 | 0 |
| 8 | 12 | 2 | 3 | 6 | 4 |
| 10 | 14 | 4 | 5 | 20 | 16 |
| | | | | **50** | **40** |

$$\hat{\beta}_1 = \frac{50}{40} = 1.25
\qquad
\hat{\beta}_0 = 9 - 1.25(6) = 1.5$$

**(ii)** $\hat{y} = 1.5 + 1.25x$. Each additional kilogram of fertiliser is
associated with 1.25 more quintals of yield.

**(iii)** Fitted values: $4.0, 6.5, 9.0, 11.5, 14.0$. Residuals:
$0, 0.5, -1, 0.5, 0$, which sum to zero as they must.

**(iv)** $\text{SSR} = 0 + 0.25 + 1 + 0.25 + 0 = 1.5$ and
$\text{SST} = 25 + 4 + 1 + 9 + 25 = 64$, so

$$R^2 = 1 - \frac{1.5}{64} = 0.977$$

An unusually tight fit, because these are made-up numbers. Real data rarely
behaves so well.
:::

**6.9** Using the previous question's results:

i. Compute $\hat{\sigma}^2$ and $\hat{\sigma}$.
ii. Compute $\mathrm{se}(\hat{\beta}_1)$.
iii. Test $H_0: \beta_1 = 0$ at the 5% level.
iv. Construct a 95% confidence interval for $\beta_1$.

::: {.answer}
**(i)** With $n = 5$ and two parameters estimated,

$$\hat{\sigma}^2 = \frac{\text{SSR}}{n-2} = \frac{1.5}{3} = 0.5
\qquad \hat{\sigma} = 0.707$$

**(ii)** $\mathrm{se}(\hat{\beta}_1) = \dfrac{0.707}{\sqrt{40}} = 0.1118$.

**(iii)** $t = 1.25/0.1118 = 11.18$ on $n - 2 = 3$ degrees of freedom. The
critical value is $t_{0.025,\,3} = 3.182$, and $11.18 > 3.182$, so we reject
$H_0$.

**(iv)** $1.25 \pm 3.182(0.1118) = [0.894,\ 1.606]$.

Note how wide that interval is despite the excellent fit — three degrees of
freedom buy very little. The critical value of 3.182 rather than 1.96 is doing
most of the damage.
:::

**6.10** A regression of household expenditure on income in a sample of 200
gives $\hat{\beta}_1 = 0.62$ with $\mathrm{se}(\hat{\beta}_1) = 0.05$.

i. Test $H_0: \beta_1 = 0$.
ii. Test $H_0: \beta_1 = 1$, and explain why an economist might care about that
    hypothesis specifically.
iii. Construct a 95% interval and interpret it.

::: {.answer}
**(i)** $t = 0.62/0.05 = 12.4$ on 198 degrees of freedom. Reject overwhelmingly.

**(ii)** $t = (0.62 - 1)/0.05 = -7.6$, which also rejects.

An economist cares because $\beta_1$ is the marginal propensity to consume, and
$\beta_1 = 1$ would mean households spend every additional rupee of income. The
test says they do not: expenditure rises by about 62 paise per additional rupee,
and the remainder is saved.

This is worth noticing as a matter of method. The interesting null hypothesis is
not always zero. Testing against a value that theory suggests is often far more
informative.

**(iii)** $0.62 \pm 1.96(0.05) = [0.522,\ 0.718]$. The interval excludes both 0
and 1, so the data are consistent with a marginal propensity to consume of
roughly two-thirds and with neither extreme.
:::

### R exercises {-}

**6.11** Using `data/car-mileage.csv`, regress fuel efficiency on weight. Report
the slope with its interval, interpret it in the units of the data, and plot the
fit.

::: {.answer}
```{r sr-ex-cars, fig.height = 3.4, fig.cap = "Fuel efficiency against weight for 74 cars, with the fitted line."}
cars <- read.csv("data/car-mileage.csv")

fit_cars <- lm(mileage_kmpl ~ weight_kg, data = cars)
summary(fit_cars)$coefficients
confint(fit_cars)["weight_kg", ]

ggplot(cars, aes(weight_kg, mileage_kmpl)) +
  geom_point(colour = book_palette$ink, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, colour = book_palette$accent,
              fill = book_palette$fill) +
  labs(x = "Weight (kg)", y = "Fuel efficiency (km per litre)") +
  theme_book()
```

The slope is $-0.00601$ km per litre per kilogram, which is awkward to read at
that scale. Multiply by 100: each additional 100 kg costs about 0.6 km per
litre, with an interval of roughly 0.50 to 0.70.

$R^2$ is 0.65, much higher than the attendance regression, because weight is a
genuinely dominant determinant of fuel consumption in a way that attendance is
not of grades.
:::

**6.12** Verify the algebra. For the car regression, check that the residuals
sum to zero, that they are uncorrelated with weight, and that the line passes
through the point of means.

::: {.answer}
```{r sr-ex-algebra}
u <- residuals(fit_cars)
w <- cars$weight_kg

round(c(sum_u        = sum(u),
        sum_w_u      = sum(w * u),
        cor_w_u      = cor(w, u),
        fitted_at_wbar = coef(fit_cars)[1] + coef(fit_cars)[2] * mean(w),
        mean_mileage   = mean(cars$mileage_kmpl)), 6)
```

All three hold. The first two are zero to within floating-point error, and the
last two agree exactly.

None of this is evidence that the model is correct. These identities are forced
by the two equations OLS solves, and they would hold just as exactly if we
regressed mileage on something absurd.
:::

**6.13** Demonstrate that OLS is unbiased, and that spread in $x$ buys
precision. Simulate from a known model with $\beta_1 = 2$, using $x$ spread
widely in one case and narrowly in the other.

::: {.answer}
```{r sr-ex-unbiased}
simulate_slope <- function(x_sd, reps = 3000) {
  b1 <- replicate(reps, {
    x <- rnorm(50, mean = 20, sd = x_sd)
    y <- 5 + 2 * x + rnorm(50, sd = 3)          # truth: beta_1 = 2
    coef(lm(y ~ x))[2]
  })
  c(x_sd = x_sd, mean_b1 = mean(b1), sd_b1 = sd(b1))
}

set.seed(11)
round(t(sapply(c(1, 5), simulate_slope)), 4)
```

Both rows are centred on 2, so OLS is unbiased at either spread — that is what
SLR1–SLR4 deliver, and the spread of $x$ has nothing to do with it.

What the spread changes is precision. With $x$ five times as spread out, the
standard deviation of $\hat{\beta}_1$ falls by roughly a factor of five, exactly
as $\sigma^2/\sum(x_i - \bar{x})^2$ predicts.
:::

**6.14** Show what omitted variable bias looks like. Simulate data where $y$
depends on both $x$ and a variable $z$ that is correlated with $x$, then
estimate a regression that omits $z$.

::: {.answer}
```{r sr-ex-ovb}
set.seed(2)

biased <- replicate(3000, {
  z <- rnorm(200)                       # unobserved: motivation, say
  x <- 10 + 2 * z + rnorm(200)          # x is correlated with z
  y <- 5 + 1 * x + 3 * z + rnorm(200)   # truth: the effect of x is 1
  c(omitting_z  = coef(lm(y ~ x))[2],
    including_z = coef(lm(y ~ x + z))[2])
})

round(rowMeans(biased), 4)
```

The true effect of $x$ on $y$ is 1. Omitting $z$ returns about 2.2 — more than
double.

The reason is exactly the term in the unbiasedness derivation. With $z$ left
out, it sits inside the error, and since $z$ is correlated with $x$ the errors
are correlated with $x$ too, so $E(u \mid x) \neq 0$ and the bias does not
vanish. Including $z$ recovers 1.

Note that the biased regression is not badly behaved in any visible way. Its
residuals still sum to zero, its standard errors are still computed correctly,
and its $p$-value is still tiny. It is confidently wrong, and nothing in the
output says so.
:::

## Question bank {#sr-question-bank}

::: {.watchout}
**No answers are given for this section, by design.**
:::

### Multiple choice {-}

**QB 6.1** In the model $y = \beta_0 + \beta_1 x + u$, the term $u$ represents

a. measurement error in $y$
b. all factors other than $x$ that affect $y$
c. the residual from the fitted line
d. the difference between $\hat{\beta}_1$ and $\beta_1$

**QB 6.2** The OLS slope estimator equals

a. the correlation between $x$ and $y$
b. the sample covariance of $x$ and $y$ divided by the sample variance of $y$
c. the sample covariance of $x$ and $y$ divided by the sample variance of $x$
d. the ratio of the two sample means

**QB 6.3** Which of the following hold by construction in every OLS regression?
*(select all that apply)*

a. $\sum \hat{u}_i = 0$
b. $\sum x_i \hat{u}_i = 0$
c. $E(u \mid x) = 0$
d. The line passes through $(\bar{x}, \bar{y})$

**QB 6.4** In a simple regression, $\hat{\sigma}^2$ divides the sum of squared
residuals by

a. $n$
b. $n - 1$
c. $n - 2$
d. $n - 3$

**QB 6.5** $\mathrm{Var}(\hat{\beta}_1)$ decreases when

a. $\sigma^2$ increases
b. the spread of $x$ increases
c. the spread of $y$ increases
d. $R^2$ decreases

**QB 6.6** A regression of wages on schooling gives $R^2 = 0.11$. This means

a. the model is misspecified
b. schooling accounts for 11% of the variation in wages
c. 11% of the coefficient is reliable
d. schooling causes 11% of wage differences

**QB 6.7** Omitted variable bias arises when the omitted variable

a. is correlated with $y$ only
b. is correlated with $x$ only
c. is correlated with both $x$ and $y$
d. has a large variance

### Short reasoning {-}

**QB 6.8** Explain why $\sum x_i\hat{u}_i = 0$ is not evidence that
$E(u \mid x) = 0$.

**QB 6.9** A student reports that their regression "explains 95% of the
variation, so the model must be correct." Evaluate.

**QB 6.10** Why does the divisor in $\hat{\sigma}^2$ change from $n-1$ in Unit
\@ref(interval-estimation) to $n-2$ here?

**QB 6.11** A researcher regresses district crop yield on rainfall and finds a
strong positive slope. Give two reasons the slope may not measure the effect of
rainfall on yield.

**QB 6.12** In the fertiliser example the slope was 1.25 quintals per kilogram.
Explain what would happen to that number, and to its standard error, if
fertiliser were measured in grams instead.

**QB 6.13** Explain why a regression can have a correctly computed standard
error, a valid $p$-value, and a completely misleading interpretation.

### Numerical {-}

**QB 6.14** For six observations, $\sum(x_i - \bar{x})(y_i - \bar{y}) = 84$,
$\sum(x_i - \bar{x})^2 = 56$, $\bar{x} = 9$, $\bar{y} = 21$ and
$\sum \hat{u}_i^2 = 12$.

i. Compute $\hat{\beta}_1$ and $\hat{\beta}_0$.
ii. Compute $\hat{\sigma}^2$ and $\mathrm{se}(\hat{\beta}_1)$.
iii. Test $H_0: \beta_1 = 0$ at the 5% level.
iv. Construct a 95% confidence interval for $\beta_1$.

**QB 6.15** A regression on 30 observations gives $\hat{\beta}_1 = 4.8$,
$\mathrm{se}(\hat{\beta}_1) = 1.5$, $\text{SST} = 900$ and $\text{SSR} = 400$.

i. Compute $R^2$.
ii. Test $H_0: \beta_1 = 0$ at the 1% level.
iii. Test $H_0: \beta_1 = 8$ at the 5% level.
iv. Compute $\hat{\sigma}^2$.

**QB 6.16** Yields and rainfall for five districts:

| Rainfall (cm) | 60 | 80 | 100 | 120 | 140 |
|--:|--:|--:|--:|--:|--:|
| Yield (quintals) | 18 | 23 | 24 | 30 | 35 |

i. Compute the OLS estimates.
ii. Compute all five residuals and verify they sum to zero.
iii. Compute $R^2$.
iv. Predict yield at 110 cm, and say why predicting at 300 cm would be unwise.

**QB 6.17** Using `data/india-production.csv`, write R code to regress log
output on log labour. Report the slope with its confidence interval, interpret
it as an elasticity, and state what assumption would have to hold for it to
represent the effect of labour on output.

## Looking ahead {#sr-looking-ahead}

Simple regression gave us a slope, a standard error, and a test — and left us
exactly where Unit \@ref(two-sample) did, unable to separate the effect of $x$
from the effects of everything correlated with $x$.

The problem was never the arithmetic. It is that the model contains one
explanatory variable, and the world contains many.

If motivated students both attend more classes and score higher for reasons
unrelated to attendance, the remedy is to measure motivation and account for it
— to ask how GPA varies with attendance *among students of similar motivation*.
That requires a model with more than one explanatory variable on the right-hand
side.

Building it is the subject of the next unit, and it turns out to require
surprisingly little new machinery: the same least squares criterion, the same
normal equations, the same $t$ statistics. What changes is the interpretation of
a coefficient, which becomes considerably more subtle — and considerably more
useful.
