# The Simple Regression Model {#simple-regression}

```{r setup-06, include = FALSE}
source("_common.R")
library(ggplot2)
```

Throughout this book we have asked whether differences between groups are
larger than we would expect from chance alone. Regression asks a broader
question: **how is one variable related to another?**

That shift matters because economic theory usually makes qualitative
predictions — higher prices reduce demand, more education increases earnings —
and qualitative predictions are hard to act on. Regression lets us estimate *how
much* one variable moves with another, express it in units someone can use, and
attach a measure of uncertainty to the answer.

Almost every empirical question in economics has this shape. How do wages change
with education? How does consumption change with income? How does crop yield
change with rainfall? How does a student's GPA change with attendance?

Answering such questions turns out to be surprisingly difficult, and three
problems arise immediately.

**Other things matter.** Attendance is not the only determinant of GPA, and
education is not the only determinant of wages. The relationship is never exact,
so we need a way of describing a pattern that holds on average while individuals
depart from it.

**We do not know the functional form.** The relationship between two variables
might be a straight line, or a curve, or something that rises and then flattens.
Economic theory rarely specifies which.

**An observed relationship need not be causal.** Even if attendance and grades
move together, attendance may not be responsible for the difference.

::: {.keyidea}
The rest of this unit develops a model for studying relationships between
variables. Along the way it answers four questions.

1. What relationship are we trying to describe?
2. Why do individuals not lie on that relationship?
3. Under what conditions can we interpret it causally?
4. How do we estimate it from a sample?

Each section that follows answers one of them.
:::

We build all of it from ideas we already have — variation, sampling and
statistical inference. Nothing genuinely new is required; what changes is the
object of study, from a mean or a difference of means to a relationship.

The result is the central tool of **econometrics** — the field concerned with
measuring economic relationships from data — though the tool itself is used
across the social sciences and beyond.

## From comparing groups to modelling relationships {#sr-motivation}

::: {.question}
We can compare two groups. Why is that not enough?
:::

Unit \@ref(two-sample) compared students above and below the median attendance.
Those who attended more had higher average grades, by more than sampling
variation alone would explain.

But knowing that two groups differ is only the beginning. We still do not know
how grades change *as* attendance changes, or whether attendance itself is
responsible for the difference.

That second gap is the deeper one, and it is worth seeing clearly.

### The experiment we cannot run {-}

What we would really like is a thought experiment. Take one student and observe
them twice: once attending 20 classes and once attending 30, with everything
else about them unchanged — the same ability, the same motivation, the same
teacher, the same week. The difference in their grades would isolate the effect
of attendance itself.

That experiment is impossible. A student cannot attend both 20 and 30 classes.
Instead, we observe *different* students who differ in many ways at once. The
challenge of empirical economics is to separate the effect of attendance from
all the other differences between those students.

::: {.watchout}
Simple regression does not solve that problem. Section \@ref(ols-causal) returns
to this point in detail. If a reader finishes this unit believing that regression
automatically uncovers causal effects, then this book has failed them.

What regression does is change the question we ask of the data, and build the
machinery that later units extend towards the thought experiment above.
:::

### The question has changed {-}

Instead of asking whether students with high attendance differ from those with
low attendance, we now ask whether grades tend to rise as attendance rises.

Attendance is no longer a group label. It becomes a **numerical variable whose
entire range carries information.**

That is the conceptual step from a $t$-test to a regression, and everything
technical in this unit follows from it.

### The price of the old question {-}

To compare two groups, Unit \@ref(two-sample) took a variable recording how many
classes each of 680 students attended — a number running from 2 to 32 — and
replaced it with a label saying "high" or "low". A student who attended 32
classes and one who attended 28 were treated as identical, while one who
attended 28 and one who attended 27 were placed in opposite groups.

We can quantify the price of that simplification.

```{r sr-waste}
students <- read.csv("data/attendance-grades.csv")
students$high <- students$attendance >= median(students$attendance)

c(split_t      = abs(t.test(sem_gpa ~ high, data = students)$statistic),
  regression_t = summary(lm(sem_gpa ~ attendance, data = students))$coefficients[2, 3])
```

The same data. The same outcome. Almost the same question. Splitting attendance
into two groups produces a $t$ statistic of 11.8; using the full attendance
variable produces one of 17.6.

The relationship was always present — the median split simply discarded much of
the evidence for it.

::: {.keyidea}
Regression does not merely give us a better answer. It asks a different question
— and in doing so improves on a two-sample comparison in two ways.

- It uses the **full** explanatory variable rather than reducing it to two
  groups.
- It estimates a **slope** — the expected change in $y$ for a one-unit increase
  in $x$ — instead of merely comparing two averages.
:::

In the next unit we will add more explanatory variables, allowing us to compare
observations while holding other measured characteristics fixed. That brings us
closer to the counterfactual comparison we would like to make, although — as we
shall see — it does not eliminate the problem entirely.

## The population regression function {#sr-prf}

::: {.question}
What exactly is regression trying to estimate?
:::

The previous section argued that regression asks a different question from a
two-sample comparison. The natural next question is: what is the object we are
trying to estimate?

It is tempting to think that regression is trying to draw a straight line
through a cloud of points. That is not quite right.

Start with the individual outcomes, because they are noisier than students
expect.

Forty-nine students in our data attended exactly 25 classes. They did not
receive the same semester GPA.

```{r sr-at25}
at25 <- students[students$attendance == 25, ]

round(c(n = nrow(at25), average_gpa = mean(at25$sem_gpa),
        lowest = min(at25$sem_gpa), highest = max(at25$sem_gpa)), 3)
```

One scored 1.58 and another 9.08, and forty-seven others lay in between — all of
them having attended the same twenty-five classes. Attendance alone plainly does
not determine anyone's grade.

So a line through the individual points cannot be what we are looking for. There
is no line that passes near 1.58 and 9.08 at the same value of $x$.

Instead of looking at every individual student, suppose we ask a different
question:

> Among students who attended 25 classes, what is the average GPA?

Now repeat the exercise for students who attended 26 classes, then 27, then 28,
and so on.

Those averages are called **conditional means**, because they average GPA
conditional on a particular attendance level. We write the conditional mean as

$$E(Y \mid X = x)$$

which simply means "the average value of $Y$ among all members of the population
whose $X$ equals $x$."

The table below approximates these conditional means by grouping attendance into
bands.

```{r sr-condmeans}
students$band <- cut(students$attendance,
                     breaks = c(0, 20, 24, 27, 30, 33),
                     labels = c("2–20", "21–24", "25–27", "28–30", "31–32"))

aggregate(sem_gpa ~ band, data = students,
          function(x) round(c(n = length(x), mean_gpa = mean(x)), 2))
```

Two patterns immediately emerge.

First, students with the same attendance still obtain very different grades.
Attendance is therefore not enough to predict an individual's GPA.

Second, the average GPA rises steadily as attendance increases — from 4.38 to
7.67. Individual outcomes are noisy, but the centre of the distribution shifts
upwards.

Regression is interested in this second pattern.

If we computed the conditional mean for every possible attendance level and
plotted those averages, we would obtain a curve describing how the average GPA
changes with attendance throughout the population.

::: {.keyidea}
That curve is called the **population regression function** (PRF):

$$\text{PRF} \;=\; E(Y \mid X = x)$$

The population regression function is therefore not a fitted line. It exists
whether or not we ever estimate it. It is simply the true relationship between
attendance and the population's average GPA.
:::

Figure \@ref(fig:sr-prf-plot) illustrates the idea. The grey points are
individual students. The orange points are estimates of the conditional mean
within each attendance band. Regression is concerned with those averages, not
with explaining every individual observation.

```{r sr-prf-plot, echo = FALSE, fig.height = 3.8, fig.cap = "Semester GPA against attendance for 680 students. Grey points are individual students; orange points are the average GPA within each attendance band. Regression describes the orange points."}
band_means <- aggregate(cbind(sem_gpa, attendance) ~ band, data = students, mean)

ggplot(students, aes(attendance, sem_gpa)) +
  geom_point(alpha = 0.18, colour = book_palette$ink, size = 1.1) +
  geom_point(data = band_means, aes(attendance, sem_gpa),
             colour = book_palette$reject, size = 3.4) +
  geom_line(data = band_means, aes(attendance, sem_gpa),
            colour = book_palette$reject, linewidth = 0.9) +
  labs(x = "Classes attended", y = "Semester GPA") +
  theme_book()
```

Each conditional mean $E(Y \mid X = x)$ is a function of $x$, so the PRF is also
called the **conditional expectation function** (CEF):

$$E(Y \mid X = x) = f(x)$$

::: {.watchout}
The CEF tells us how the *average* value of $y$ changes with $x$. It does **not**
say that $y = \beta_0 + \beta_1 x$ for every member of the population.

Individual students depart from the average for their attendance level, often by
a great deal. Confusing the conditional mean with the individual outcome is the
most common misreading of a regression.
:::

Notice that the orange points do not lie exactly on a straight line. Nothing in
the definition of the population regression function requires linearity. The
conditional mean could be curved, flat over some ranges, or highly nonlinear.

Only now do we make a modelling choice.

The PRF is the object we want to estimate. Unfortunately it is an unknown
*function*, and functions can take infinitely many shapes. To make estimation
possible at all, we approximate it with a simple mathematical form.

In this unit we assume that the population regression function is approximately
linear,

$$E(Y \mid X = x) = \beta_0 + \beta_1 x$$

because a straight line is simple to estimate, easy to interpret, and often
provides a useful approximation to more complicated relationships.

Whether a straight line is a good approximation is an empirical question rather
than a mathematical one. Section \@ref(ols-fit) returns to how well it
performs for these data.

The population regression function gives us the target. Two questions remain.

First, why do individual students differ from the conditional mean? Second, how
can we estimate the population regression function when we observe only a sample
rather than the entire population?

The next two sections answer those questions.

## From averages to observations {#sr-model}

::: {.question}
The PRF describes averages. Individual students are not averages. How do we
write down the difference?
:::

Return to the forty-nine students who attended exactly 25 classes. Their
conditional mean was 6.10, and two of them scored 1.58 and 9.08.

The conditional expectation function accounts for the 6.10. It has nothing to say
about why one student was four and a half points below it and another three
points above.

So an individual's grade is the conditional mean *plus* a departure from it. Give
that departure a name.

::: {.keyidea}
For each individual $i$,

$$y_i = E(Y \mid X = x_i) + u_i$$

where $u_i$ is the **error term** or **stochastic disturbance** — the amount by
which individual $i$ departs from the average for their value of $x$.

Substituting the linear conditional mean from Section \@ref(sr-prf) gives the
**simple linear regression model**:

$$y_i = \beta_0 + \beta_1 x_i + u_i$$
:::

This is not a new model. It is simply another way of writing the population
regression function once we give the unexplained part a name.

The two students above become

$$1.58 = \beta_0 + \beta_1(25) + u_1
\qquad\qquad
9.08 = \beta_0 + \beta_1(25) + u_2$$

with a large negative $u_1$ and a large positive $u_2$. Both students attended
the same twenty-five classes, so the first two terms are identical for each of
them; everything that distinguishes them sits in $u$.

We now have a model. Before asking how to estimate it, we should be clear about
what that last term contains.

## The error term {#sr-error}

::: {.question}
What is $u$?
:::

Students meet the word "error" and assume something has gone wrong. Nothing has.

For the grades example, $u$ contains everything other than attendance that
affects a student's GPA:

- how hard they worked outside class,
- how well prepared they were when the course began,
- the quality of teaching they happened to receive,
- whether they were ill during the examinations,
- whether they held a job that semester,
- how much they happened to like the subject,
- and luck on the day.

None of these is a mistake, and none of them is measurement noise. They are real
influences on real outcomes that our model does not include — because they are
not in the dataset, and in some cases could not be.

::: {.keyidea}
**The error term is not a mistake and not measurement noise. It is the rest of
the world.**
:::

That is why $u$ is sometimes called the **unobserved** variable, and why
econometricians spend so much of their time worrying about it. It is the part of
the story we did not measure.

Whether those omitted influences are related to attendance is the central
question of causal inference.

### Systematic and stochastic {-}

The model divides $y$ into two parts, and it is worth naming them.

$$y_i = \underbrace{\beta_0 + \beta_1 x_i}_{\text{systematic}} + \underbrace{u_i}_{\text{stochastic}}$$

The **systematic** or deterministic part, $\beta_0 + \beta_1 x$, is what
attendance accounts for. It is identical for every student with the same
attendance.

The **stochastic** part, $u$, is everything else. It differs from person to
person and is not observed.

### What the two parameters mean {-}

In the linear PRF $E(Y \mid X = x) = \beta_0 + \beta_1 x$:

$\beta_0$ is the conditional mean when $x = 0$ — the average $y$ among
individuals with no $x$ at all. It is called the **intercept**, and it is rarely
central to the analysis; sometimes $x = 0$ does not even occur in the data.

$\beta_1$ is the amount by which the conditional mean changes when $x$ increases
by one unit, $\partial y / \partial x$. It is called the **slope**, and it is
almost always the parameter of interest.

The remaining names are worth having in one place. $y$ is the **dependent**
variable — semester GPA. $x$ is the **independent** variable, also called the
explanatory or control variable — attendance. $u$ is the **error** or disturbance
term.

For the grades example,

$$\textit{GPA} = \beta_0 + \beta_1\,\textit{attendance} + u$$

and $\beta_1$ is the change in GPA associated with attending one more class.

Notice the phrase *associated with*. Whether $\beta_1$ is the change attendance
would *cause* is a separate question, and it is the subject of the next section.

## When does β₁ measure an effect? {#sr-assumptions}

::: {.question}
We have a model and a slope. What has to be true for that slope to mean what we
want it to mean?
:::

The model so far is a decomposition. It says $y$ can be split into a part that
moves with $x$ and a part that does not, which is true of any two variables
whatever and therefore not yet useful.

To read $\beta_1$ as the effect of $x$ on $y$, we need an assumption about how
$u$ and $x$ are related. Two assumptions are usually stated; only the second
restricts the world.

### The error averages zero {-}

$$E(u) = 0$$

This costs nothing. If the unobserved factors had a non-zero average, we could
absorb that average into the intercept and start again.

::: {.deeper}
Suppose $u = e + c$ where $c$ is some non-zero constant — a fixed effect of, say,
generous marking. Then

$$
\begin{aligned}
y &= \beta_0 + \beta_1 x + (e + c) \\[4pt]
&= (\beta_0 + c) + \beta_1 x + e \\[4pt]
&= \beta_0^{*} + \beta_1 x + e
\end{aligned}
$$

with $E(e) = 0$. The intercept has changed and the slope has not. Since $\beta_1$
is what we want, assuming $E(u) = 0$ restricts nothing of importance.
:::

### The error is unrelated to x {-}

We now ask a simple question. Among students with the same attendance, what is
the average value of everything contained in $u$?

The assumption is that this average is the *same* at every value of $x$, and
equal to the overall average of zero:

$$E(u \mid x) = E(u) = 0$$

In that case $u$ is said to be **mean independent** of $x$. Taken together with
$E(u) = 0$, this is the **zero conditional mean assumption**.

::: {.keyidea}
$$E(u \mid x) = 0$$

This is the key assumption for interpreting the slope as the effect of $x$, and
the only way to judge it is to state it in the language of the problem.

For the grades example it says

$$E(\textit{effort} \mid \textit{attendance} = \text{low})
= E(\textit{effort} \mid \textit{attendance} = \text{high}) = 0$$

— that students who attend more classes are, on average, no more and no less
able, motivated, or well-taught than students who attend fewer.
:::

Note carefully what is *not* required. The assumption does not say that students
have identical effort — obviously they do not. It says only that effort does not
systematically differ with attendance. Equal averages, not equal individuals.

Stated that concretely, the assumption is clearly doubtful here. Students who
attend regularly are plausibly working harder in other ways too, and effort
raises grades on its own.

We cannot check it. Effort is not in the dataset, which is precisely what
"unobserved" means. There is no diagnostic, no test, and no amount of additional
data that settles it — only an argument about how the data came to look the way
it does.

::: {.watchout}
When $E(u \mid x) \neq 0$, $\beta_1$ does not measure the effect of $x$. It
measures the effect of $x$ plus the influence of whatever else differs between
high and low values of $x$.

This is **omitted variable bias**. Section \@ref(ols-causal) returns to it once we
have the algebra to show exactly how large it is and in which direction it runs.
:::

With the model written down and its assumptions stated, we can turn to the
practical question: how do we estimate $\beta_0$ and $\beta_1$ from data?

## From the population to a sample {#sr-srf}

::: {.question}
The PRF describes a population we cannot see. What do we actually have?
:::

The population regression function is a population object. So are $\beta_0$ and
$\beta_1$: they are fixed, unknown numbers describing every member of the
population, exactly like $\mu$ in Unit \@ref(estimation).

We never observe the population. What we observe is a sample of $y$ values with
their corresponding $x$ values.

We therefore estimate. From the sample we compute $\hat{\beta}_0$ and $\hat{\beta}_1$, and these define the
**sample regression function**:

$$\hat{y}_i = \hat{\beta}_0 + \hat{\beta}_1 x_i$$

where $\hat{y}$ estimates $E(Y \mid X = x)$, and $\hat{\beta}_0$ and
$\hat{\beta}_1$ estimate $\beta_0$ and $\beta_1$.

### Every sample gives a different line {-}

Suppose two sections of thirty students each are surveyed, and a line is fitted to
each. Which one is the population regression function?

```{r sr-two-samples}
set.seed(12)
shuffle <- sample(nrow(students))

section_a <- students[shuffle[1:30], ]
section_b <- students[shuffle[31:60], ]

round(c(section_a   = coef(lm(sem_gpa ~ attendance, section_a))[2],
        section_b   = coef(lm(sem_gpa ~ attendance, section_b))[2],
        whole_class = coef(lm(sem_gpa ~ attendance, students))[2]), 4)
```

```{r sr-two-samples-plot, echo = FALSE, fig.height = 3.8, fig.cap = "Two sections of thirty students, each with its own fitted line, against the line fitted to all 680. Neither section's line is the population regression function, and from a single section there would be no way to tell how far off it was."}
lines_df <- data.frame(
  which = c("Section A (n = 30)", "Section B (n = 30)", "All 680 students"),
  int   = c(coef(lm(sem_gpa ~ attendance, section_a))[1],
            coef(lm(sem_gpa ~ attendance, section_b))[1],
            coef(lm(sem_gpa ~ attendance, students))[1]),
  slope = c(coef(lm(sem_gpa ~ attendance, section_a))[2],
            coef(lm(sem_gpa ~ attendance, section_b))[2],
            coef(lm(sem_gpa ~ attendance, students))[2])
)
lines_df$which <- factor(lines_df$which, levels = lines_df$which)

ggplot(students, aes(attendance, sem_gpa)) +
  geom_point(alpha = 0.10, colour = book_palette$ink, size = 1.1) +
  geom_abline(data = lines_df, aes(intercept = int, slope = slope,
                                   colour = which, linewidth = which)) +
  scale_colour_manual(values = c(book_palette$reject, book_palette$accent,
                                 "grey30"), name = NULL) +
  scale_linewidth_manual(values = c(0.9, 0.9, 1.4), guide = "none") +
  labs(x = "Classes attended", y = "Semester GPA") +
  theme_book() +
  theme(legend.position = "top")
```

There is no way to tell. Section A's slope is 0.300 and Section B's is 0.162 —
one is nearly twice the other, and both come from the same class. A third section
would give a third answer.

This is the situation of Unit \@ref(estimation) in a new costume. The estimate
varies from sample to sample, and our task is to understand how much and why.
Notice too that the two sections of thirty scatter far more widely around the
full-sample line than two halves of a large sample would — which is the
$\sqrt{n}$ lesson of Unit \@ref(estimation) arriving in a new form.

::: {.keyidea}
| Population (PRF) | Sample (SRF) |
|:--|:--|
| $y = \beta_0 + \beta_1 x + u$ | $y_i = \hat{\beta}_0 + \hat{\beta}_1 x_i + \hat{u}_i$ |
| $y = E(Y \mid X = x) + u$ | $y_i = \hat{y}_i + \hat{u}_i$ |
| $\beta_0$ is the intercept | $\hat{\beta}_0$ estimates $\beta_0$ |
| $\beta_1$ is the slope | $\hat{\beta}_1$ estimates $\beta_1$ |
| **$u$ is the error, and is unobservable** | **$\hat{u}$ is the residual, and is computed** |

The distinction between $u$ and $\hat{u}$ is the one to hold onto. The error is
the deviation from the true line, which nobody sees. The residual is the
deviation from our fitted line, which we can print.
:::

Because every sample gives a different line, $\hat{\beta}_1$ is a random
variable with a sampling distribution — and every question from Units
\@ref(estimation) to \@ref(one-sample) applies to it. Is it unbiased? What is its
standard error? How do we test a claim about it?

Before any of that, though, we need to decide *which* line to fit.

## Which line should we fit? {#sr-ols}

::: {.question}
Infinitely many lines pass through a scatter of points. Which one?
:::

Every sample admits infinitely many possible lines. We therefore need a rule
telling us which one deserves to be called *the* regression line.

A residual measures how badly a candidate line misses observation $i$:

$$\hat{u}_i = y_i - \hat{y}_i = y_i - \hat{\beta}_0 - \hat{\beta}_1 x_i$$

We want the residuals small. The obvious criterion is to make their sum as close
to zero as possible.

That criterion fails, and it is worth seeing why.

::: {.watchout}
Consider two lines. The first misses four points by $+1, -1, +1, -1$. The second
misses them by $+8, -8, +8, -8$.

Both have residuals summing to exactly zero. By that criterion the two lines are
equally good, which is absurd — the second is dreadful.

The problem is that positive and negative misses cancel. A criterion that lets
errors offset one another can be satisfied by a line that is nowhere near the
data.
:::

The fix is the one from Section \@ref(u1-describing): square the deviations
first, so that they cannot cancel and large misses count for more than small
ones. We choose $\hat{\beta}_0$ and $\hat{\beta}_1$ to minimise

$$L = \sum_{i=1}^{n} \hat{u}_i^2 = \sum_{i=1}^{n}\left(y_i - \hat{\beta}_0 - \hat{\beta}_1 x_i\right)^2$$

This is the **method of ordinary least squares**, or OLS.

Notice that the criterion depends only on quantities we can observe. It is built
from residuals, not errors — we could not minimise $\sum u_i^2$ even if we wanted
to, because nobody sees $u$.

We now have a criterion. Finding the line that satisfies it is a matter of
calculus, and it is where the next unit begins.

::: {.keyidea}
Regression chooses the line making the sum of *squared* vertical distances from
the points to the line as small as possible.

Squaring does two things. It stops positive and negative misses from cancelling,
and it makes one large miss count for more than several small ones.
:::

## Summary {#sr-summary}

::: {.keyidea}
- Regression asks how one variable is related to another rather than whether two
  groups differ. Splitting attendance at the median gave $t = 11.8$; using the
  whole variable gave $t = 17.6$.
- Individuals with the same $x$ obtain very different $y$. Forty-nine students
  attended exactly 25 classes and scored between 1.58 and 9.08, so no line can
  pass through the individual points.
- The **population regression function** is the conditional mean
  $E(Y \mid X = x)$. It is not a fitted line, and it exists whether or not we
  ever estimate it.
- Since each conditional mean is a function of $x$, the PRF is also the
  **conditional expectation function**. It says how the *average* of $y$ moves
  with $x$, not what any individual obtains.
- Taking the CEF to be linear is a **modelling choice**, made after the target is
  defined rather than before it.
- Writing $y_i = E(Y \mid X = x_i) + u_i$ and substituting the linear CEF gives
  $y_i = \beta_0 + \beta_1 x_i + u_i$ — the PRF with a name attached to the part
  it does not describe, not a new assumption.
- The **error term** is not a mistake and not measurement noise. It is the rest
  of the world: effort, ability, teaching, health, luck.
- $E(u) = 0$ costs nothing, since a non-zero average can be absorbed into the
  intercept. $E(u \mid x) = 0$ costs everything, cannot be tested, and is what
  separates a description from an effect.
- $\beta_0$ and $\beta_1$ are population parameters; $\hat{\beta}_0$ and
  $\hat{\beta}_1$ are sample estimates defining the **sample regression
  function**. Two sections of thirty students gave slopes of 0.300 and 0.162.
- The **error** $u$ is the deviation from the true line and is unobservable. The
  **residual** $\hat{u}$ is the deviation from the fitted line and can be
  printed.
- Minimising $\sum \hat{u}_i$ fails because positive and negative misses cancel.
  **Ordinary least squares** minimises $\sum \hat{u}_i^2$.
:::

## Practice problems {#sr-practice}

::: {.yourturn}
Work each one before opening the solution. The
[question bank](#sr-question-bank) that follows has no answers.
:::

### Concept checks {-}

**6.1** Why does regression estimate a conditional mean rather than predict
individual outcomes?

::: {.answer}
Because $y$ is not determined by $x$. At any given attendance level, students
obtain a wide range of grades, because grades also depend on effort, ability and
much else — all of which sit in the error term.

What is stable across the population is the *average* of $y$ at each value of
$x$. That is what $E(y \mid x)$ describes and what the regression line
estimates.

The scatter around the line is therefore not a failure of the method. It is the
part of $y$ that $x$ does not account for, and a model claiming otherwise would
be claiming that attendance alone determines grades.
:::

**6.2** Why is minimising the sum of residuals a bad criterion?

::: {.answer}
Because positive and negative residuals cancel, so a line that misses badly in
both directions can have residuals summing to zero.

Misses of $+8, -8, +8, -8$ sum to zero, and so do misses of $+1, -1, +1, -1$. By
that criterion the two lines are equally good, though the first is far worse.

Squaring removes the cancellation and makes large misses count more than small
ones, which is what we want from a measure of fit.
:::

**6.3** A student writes: "the regression line shows what GPA a student who
attended 25 classes will get." Correct the statement.

::: {.answer}
The line gives the *average* GPA among students who attended 25 classes, which in
these data is 6.10. It does not give what any particular such student will get.

Forty-nine students attended exactly 25 classes, scoring between 1.58 and 9.08. A
prediction for an individual would be wrong for almost all of them.

The confusion is between the conditional mean and the individual outcome — which
is the difference the error term exists to describe.
:::

### Explain {-}

**6.4** State the zero conditional mean assumption for a regression of earnings
on years of schooling, in concrete terms, and say whether you believe it.

::: {.answer}
The model is $\textit{earnings} = \beta_0 + \beta_1\,\textit{schooling} + u$,
where $u$ holds everything else affecting earnings — ability, family
connections, school quality, motivation, region.

$E(u \mid \textit{schooling}) = 0$ says that people with more schooling are, on
average, no more able, no better connected, and no better taught than people
with less.

That is not believable. Schooling is not distributed at random: children of
wealthier and better-connected families obtain more of it, and those advantages
raise earnings on their own. So $u$ is positively related to schooling, and
$\hat{\beta}_1$ overstates the return to education.

This does not make the regression useless. It correctly describes how average
earnings differ across schooling levels. It simply does not measure what an
extra year of schooling would do to a given person.
:::

**6.5** Distinguish the error $u_i$ from the residual $\hat{u}_i$.

::: {.answer}
The error is the deviation of $y_i$ from the *population* regression line:
$u_i = y_i - \beta_0 - \beta_1 x_i$. It involves the true parameters and is
therefore unobservable.

The residual is the deviation from our *fitted* line:
$\hat{u}_i = y_i - \hat{\beta}_0 - \hat{\beta}_1 x_i$. It is computed from the
data and can be printed.

The relationship mirrors $\mu$ against $\bar{x}$. Residuals are estimates of
errors, and confusing the two leads to claims like "the errors are uncorrelated
with $x$, so SLR4 holds" — which is wrong, because the *residuals* are
uncorrelated with $x$ by construction, whatever the errors do.
:::

**6.6** Explain why $y_i = \beta_0 + \beta_1 x_i + u_i$ is not an extra
assumption once the conditional expectation function has been defined.

::: {.answer}
Because $u_i$ is *defined* as the gap between what individual $i$ obtained and
the average for their value of $x$:

$$u_i = y_i - E(Y \mid X = x_i)$$

Rearranged, that reads $y_i = E(Y \mid X = x_i) + u_i$, which is an identity. It
says only that an observation equals its conditional mean plus however far it
sits from that mean.

Substituting the linear CEF gives $y_i = \beta_0 + \beta_1 x_i + u_i$. The
genuine assumption was made earlier, when we chose to treat the conditional mean
as linear. Writing the model down adds nothing but a name for the remainder.
:::

**6.7** Two sections of the same class produce fitted slopes of 0.300 and 0.162.
A student concludes that one section must contain an arithmetic error. Evaluate.

::: {.answer}
No error is implied. Each section is a different sample from the same population,
and every sample produces a different sample regression function — exactly as
every sample produced a different $\bar{x}$ in Unit \@ref(estimation).

Neither slope is $\beta_1$. Both estimate it, and with thirty students each they
estimate it imprecisely. The whole class of 680 gives 0.189, which is also an
estimate rather than the truth.

The useful response is not to hunt for a mistake but to ask how much
$\hat{\beta}_1$ varies from sample to sample — a question about its sampling
distribution, and the subject of the next unit.
:::

### Numerical and conceptual {-}

**6.8** A population has these conditional means of $y$ given $x$:

| $x$ | 1 | 2 | 3 | 4 |
|:--|:--|:--|:--|:--|
| $E(Y \mid X = x)$ | 5 | 9 | 13 | 17 |

i. Is the population regression function linear here?
ii. Write down $\beta_0$ and $\beta_1$.
iii. An individual with $x = 3$ has $y = 11$. What is their $u_i$?
iv. What is $E(u \mid X = 3)$, and why?

::: {.answer}
**(i)** Yes. The conditional mean rises by exactly 4 for each unit increase in
$x$, so one straight line passes through all four points.

**(ii)** $\beta_1 = 4$. Working back from $E(Y \mid X = 1) = 5$ gives
$\beta_0 = 5 - 4 = 1$, so the PRF is $E(Y \mid X = x) = 1 + 4x$.

**(iii)** $u_i = y_i - E(Y \mid X = 3) = 11 - 13 = -2$. This individual sits two
units below the average for their value of $x$.

**(iv)** Zero. The conditional mean *is* the average of $y$ among all individuals
with $X = 3$, so deviations from it must average to zero within that group. That
is why $E(u \mid x) = 0$ is natural when the linear CEF is correct.
:::

**6.9** Explain why a regression of $y$ on $x$ cannot be estimated when every
observation has the same value of $x$ — using the definition of the conditional
mean rather than any formula.

::: {.answer}
The PRF describes how the average of $y$ changes *as $x$ changes*. If $x$ never
changes, there is exactly one conditional mean to observe and no information
about any other value.

Concretely: if all 680 students had attended 25 classes, we could compute the
average GPA of students attending 25 classes and nothing else. Whether GPA would
be higher at 30 classes is unanswerable from such data — not because the
arithmetic breaks down, but because the comparison the question requires was
never observed.

Variation in the explanatory variable is not a technical requirement. It is the
evidence.
:::

## Question bank {#sr-question-bank}

::: {.watchout}
**No answers are given for this section, by design.**
:::

### Multiple choice {-}

**QB 6.1** In the model $y = \beta_0 + \beta_1 x + u$, the term $u$ represents

a. measurement error in $y$
b. all factors other than $x$ that affect $y$
c. the residual from the fitted line
d. the difference between $\hat{\beta}_1$ and $\beta_1$

**QB 6.2** The population regression function is

a. the line fitted by ordinary least squares
b. the conditional mean $E(Y \mid X = x)$
c. the average of $y$ in the sample
d. the line that minimises squared residuals

**QB 6.3** $E(u \mid x) = 0$ says that

a. the residuals sum to zero
b. the unobserved factors average zero at every value of $x$
c. $x$ and $y$ are uncorrelated
d. the model is linear

**QB 6.4** Which of the following are unobservable? *(select all that apply)*

a. $u_i$
b. $\hat{u}_i$
c. $\beta_1$
d. $\hat{\beta}_1$

**QB 6.5** Minimising $\sum \hat{u}_i$ rather than $\sum \hat{u}_i^2$ fails
because

a. the sum is always negative
b. positive and negative residuals cancel
c. it cannot be differentiated
d. it requires a larger sample

**QB 6.6** Treating the conditional expectation function as linear is

a. implied by the definition of the PRF
b. a modelling choice
c. required for $E(u) = 0$
d. a consequence of random sampling

### Short reasoning {-}

**QB 6.7** Explain the difference between the population regression function and
the sample regression function, and say which one changes when a new sample is
drawn.

**QB 6.8** A colleague says the error term captures "measurement error in the
data". Explain what is wrong with that, and give three examples of what $u$
actually contains in a regression of wages on years of schooling.

**QB 6.9** Why is $E(u) = 0$ described as costing nothing, while
$E(u \mid x) = 0$ is the assumption on which everything rests?

**QB 6.10** State the zero conditional mean assumption in words for a regression
of crop yield on fertiliser use, and give one concrete reason it might fail.

**QB 6.11** A researcher regresses district crop yield on rainfall and finds a
strong positive slope. Give two reasons the slope may not measure the effect of
rainfall on yield.

**QB 6.12** Explain why the least squares criterion penalises one residual of 8
more heavily than eight residuals of 1.

### Numerical {-}

**QB 6.13** A population's conditional means are $E(Y \mid X = x) = 12 - 0.5x$.

i. State $\beta_0$ and $\beta_1$.
ii. What is the average $y$ among individuals with $x = 10$?
iii. An individual with $x = 10$ has $y = 9$. Find $u_i$.
iv. Interpret $\beta_1$ in words.

**QB 6.14** Four individuals with $x = 6$ have $y$ values 14, 18, 11 and 21.

i. Estimate $E(Y \mid X = 6)$.
ii. Compute the four implied values of $u_i$.
iii. Verify that they sum to zero, and explain why that had to happen.
iv. Explain why this does *not* demonstrate that $E(u \mid x) = 0$ in the
    population.

## Looking ahead {#sr-looking-ahead}

We have a target, a model and a criterion.

The target is the population regression function $E(Y \mid X = x)$, which we have
chosen to approximate with a straight line. The model
$y_i = \beta_0 + \beta_1 x_i + u_i$ adds a name for the part of $y$ that
attendance does not explain. The criterion says to pick the line minimising the
sum of squared residuals.

What we do not yet have is the line.

The next unit finds it. Two derivatives set to zero give $\hat{\beta}_0$ and
$\hat{\beta}_1$ in closed form, and the formulas turn out to say something
interpretable: the slope is the covariance of $x$ and $y$ divided by the variance
of $x$.

With estimates in hand we can then ask what Units \@ref(estimation) to
\@ref(one-sample) taught us to ask of any estimator. Is it centred on the truth?
How much does it vary from sample to sample? How do we test a claim about it? And
how much of the variation in $y$ has it actually accounted for?
