# Ordinary Least Squares {#ols}

```{r setup-07, include = FALSE}
source("_common.R")
library(ggplot2)
students <- read.csv("data/attendance-grades.csv")
```

The previous unit ended one step short of an answer. We had the target — the
population regression function — a model for how individuals depart from it, and
a criterion for choosing between candidate lines: minimise the sum of squared
residuals.

This unit carries that out. It derives the two estimators, establishes what they
guarantee and what they do not, works out how precise they are, and tests claims
about them.

::: {.keyidea}
The unit answers five questions.

1. Which line minimises the sum of squared residuals?
2. What does that line guarantee, and what does it merely appear to guarantee?
3. How much of the variation in $y$ has it accounted for?
4. Is $\hat{\beta}_1$ centred on $\beta_1$, and how much does it vary?
5. What happens when the errors are not equally spread?
6. Is the slope we found distinguishable from zero?
7. What does the slope mean, and in what units?
:::

## Deriving the estimators {#ols-derive}

::: {.question}
Which values of $\hat{\beta}_0$ and $\hat{\beta}_1$ make the sum of squared
residuals as small as possible?
:::

Recall the criterion from Section \@ref(sr-ols). We choose $\hat{\beta}_0$ and
$\hat{\beta}_1$ to minimise

$$L = \sum_{i=1}^{n} \hat{u}_i^2
= \sum_{i=1}^{n}\left(y_i - \hat{\beta}_0 - \hat{\beta}_1 x_i\right)^2$$

This is a minimisation problem in two variables, and it is solved the usual way:
differentiate with respect to each, set both derivatives to zero, and solve the
pair of equations that results. Those two equations are called the **normal
equations**.

::: {.theory}
Minimising $L$ means setting both partial derivatives to zero.

**With respect to the intercept:**

$$
\begin{aligned}
\frac{\partial L}{\partial \hat{\beta}_0}
&= \sum_{i=1}^{n} -2\left(y_i - \hat{\beta}_0 - \hat{\beta}_1 x_i\right) = 0 \\[4pt]
0 &= \sum_{i=1}^{n} y_i - n\hat{\beta}_0 - \hat{\beta}_1\sum_{i=1}^{n} x_i
   && \text{divide by } -2 \text{ and expand} \\[4pt]
0 &= n\bar{y} - n\hat{\beta}_0 - \hat{\beta}_1 n\bar{x}
   && \text{since } \textstyle\sum x_i = n\bar{x} \\[4pt]
\hat{\beta}_0 &= \bar{y} - \hat{\beta}_1\bar{x}
   && \text{divide by } n \text{ and rearrange}
\end{aligned}
$$

**With respect to the slope:**

$$
\begin{aligned}
\frac{\partial L}{\partial \hat{\beta}_1}
&= \sum_{i=1}^{n} -2x_i\left(y_i - \hat{\beta}_0 - \hat{\beta}_1 x_i\right) = 0 \\[4pt]
0 &= \sum x_i y_i - \hat{\beta}_0\sum x_i - \hat{\beta}_1 \sum x_i^2 \\[4pt]
0 &= \sum x_i y_i - \left(\bar{y} - \hat{\beta}_1\bar{x}\right)n\bar{x} - \hat{\beta}_1\sum x_i^2
   && \text{substitute } \hat{\beta}_0 \\[4pt]
0 &= \sum x_i y_i - n\bar{x}\bar{y} - \hat{\beta}_1\left(\sum x_i^2 - n\bar{x}^2\right) \\[4pt]
\hat{\beta}_1 &= \frac{\sum x_i y_i - n\bar{x}\bar{y}}{\sum x_i^2 - n\bar{x}^2}
\end{aligned}
$$

Both parts of that fraction can be written more meaningfully. Expanding the
numerator,

$$
\begin{aligned}
\sum (x_i - \bar{x})(y_i - \bar{y})
&= \sum x_i y_i - \bar{y}\sum x_i - \bar{x}\sum y_i + n\bar{x}\bar{y} \\[4pt]
&= \sum x_i y_i - n\bar{x}\bar{y}
\end{aligned}
$$

and the denominator,

$$\sum (x_i - \bar{x})^2 = \sum x_i^2 - 2\bar{x}\sum x_i + n\bar{x}^2
= \sum x_i^2 - n\bar{x}^2$$

so that

$$\hat{\beta}_1 = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}{\sum_{i=1}^{n}(x_i - \bar{x})^2}$$
:::

::: {.keyidea}
The OLS estimators are

$$\hat{\beta}_1 = \frac{\sum(x_i - \bar{x})(y_i - \bar{y})}{\sum(x_i - \bar{x})^2}
\qquad\qquad
\hat{\beta}_0 = \bar{y} - \hat{\beta}_1\bar{x}$$

The slope is the **sample covariance of $x$ and $y$ divided by the sample
variance of $x$**. It answers: when $x$ is above its mean, is $y$ also above
its mean, and by how much per unit?
:::

::: {.intuition}
That reading explains several things at once.

If $x$ and $y$ move together, the numerator is positive and so is the slope. If
they move oppositely, it is negative. If they are unrelated, the products cancel
and the slope is near zero.

And the denominator explains why $x$ must vary. If every student attended the
same number of classes, $\sum(x_i - \bar{x})^2 = 0$ and the slope is undefined —
there is no way to learn how GPA responds to attendance from data in which
attendance never changes. This is the same requirement as Section
\@ref(u1-randomness): no variation, nothing to learn.
:::

### Doing it by hand and in R {-}

```{r sr-byhand}
x <- students$attendance
y <- students$sem_gpa

b1 <- sum((x - mean(x)) * (y - mean(y))) / sum((x - mean(x))^2)
b0 <- mean(y) - b1 * mean(x)

c(b0 = b0, b1 = b1)
```

```{r sr-lm}
fit <- lm(sem_gpa ~ attendance, data = students)
coef(fit)
```

Identical. `lm()` solves the same two equations.

So $\hat{\beta}_1 = 0.189$: each additional class attended is associated with
about 0.19 of a GPA point. Ten more classes, roughly 1.9 points — which on a
ten-point scale is a great deal.

```{r sr-fit-plot, echo = FALSE, fig.height = 3.8, fig.cap = "The fitted line, with the vertical distances it misses by shown for a handful of students. Ordinary least squares chooses the line making the sum of those squared distances as small as possible."}
set.seed(3)
shown <- students[sample(nrow(students), 25), ]
shown$pred <- predict(fit, newdata = shown)

ggplot(students, aes(attendance, sem_gpa)) +
  geom_point(alpha = 0.13, colour = book_palette$ink, size = 1.1) +
  geom_segment(data = shown, aes(xend = attendance, y = sem_gpa, yend = pred),
               colour = book_palette$reject, linewidth = 0.45) +
  geom_point(data = shown, colour = book_palette$reject, size = 1.5) +
  geom_abline(intercept = coef(fit)[1], slope = coef(fit)[2],
              colour = book_palette$accent, linewidth = 1) +
  labs(x = "Classes attended", y = "Semester GPA") +
  theme_book()
```

## What the fitted line guarantees {#ols-fit}

::: {.question}
Having minimised the squares, what do we get for free?
:::

Three algebraic facts follow directly from the two equations we just solved.
They hold for *every* OLS regression, in every dataset, whether or not the model
is any good.

::: {.keyidea}
1. **The residuals sum to zero**: $\sum \hat{u}_i = 0$. This is the first
   normal equation, restated.
2. **The residuals are uncorrelated with $x$**: $\sum x_i \hat{u}_i = 0$.
   Whatever pattern is left over, it is not linearly related to $x$ — OLS has
   already extracted all of that.
3. **The line passes through the means**: $\bar{y} = \hat{\beta}_0 +
   \hat{\beta}_1\bar{x}$. The point $(\bar{x}, \bar{y})$ always lies on the
   fitted line.
:::

```{r sr-algebra-check}
u_hat <- residuals(fit)

c(sum_residuals   = sum(u_hat),
  sum_x_times_u   = sum(x * u_hat),
  predicted_at_xbar = coef(fit)[1] + coef(fit)[2] * mean(x),
  mean_of_y         = mean(y))
```

The first two are zero up to floating-point error, and the last two agree.

::: {.watchout}
These are *algebraic* properties, not evidence. They hold by construction, so
finding them in your output tells you the arithmetic worked and nothing more.

In particular, $\sum x_i \hat{u}_i = 0$ is true whether or not
$E(u \mid x) = 0$. The first is forced by the estimator; the second is an
assumption about the world that no computation can verify.
:::

### How much of the variation has the line accounted for? {-}

The fitted line runs through the data, but how much of what we see does it
actually capture? Answering that begins with a single observation, and requires
no algebra at all.

Take one student — number 554 in the data, who attended 9 classes and finished
with a semester GPA of 0.8. The class average is 6.50, so this student sits a
long way below it.

That gap can be split in two.

Part of it is anticipated by the line. This student attended only 9 classes, and
the line predicts a GPA of 3.26 for such a student — already well below average.
The rest is not anticipated at all: the line said 3.26 and the student scored
0.8.

```{r ols-decomp-plot, echo = FALSE, fig.height = 4.4, fig.cap = "One student's distance from the class average, split in two. The green segment is the total gap between this student and the mean. The blue segment is the part the fitted line accounts for -- this student attended few classes, and the line predicts a low grade. The orange segment is what the line does not account for. The two pieces add exactly to the whole."}
i     <- 554
xi    <- students$attendance[i]
yi    <- students$sem_gpa[i]
yhat  <- fitted(fit)[i]
ybar  <- mean(students$sem_gpa)

ggplot(students, aes(attendance, sem_gpa)) +
  geom_point(alpha = 0.12, colour = book_palette$ink, size = 1) +
  geom_hline(yintercept = ybar, linetype = "dashed",
             colour = "grey35", linewidth = 0.6) +
  geom_abline(intercept = coef(fit)[1], slope = coef(fit)[2],
              colour = "grey25", linewidth = 0.9) +
  # total, offset to the left so all three are visible at once
  annotate("segment", x = xi - 1.4, xend = xi - 1.4, y = ybar, yend = yi,
           colour = "#2f855a", linewidth = 1.5) +
  # accounted for by the line, then the residual, stacked at the observation
  annotate("segment", x = xi, xend = xi, y = ybar, yend = yhat,
           colour = book_palette$accent, linewidth = 1.8) +
  annotate("segment", x = xi, xend = xi, y = yhat, yend = yi,
           colour = book_palette$reject, linewidth = 1.8) +
  geom_point(x = xi, y = yi, colour = book_palette$ink, size = 2.8) +
  annotate("text", x = xi - 1.9, y = (ybar + yi)/2, hjust = 1, size = 3.3,
           colour = "#2f855a", label = "total\n5.70") +
  annotate("text", x = xi + 0.7, y = (ybar + yhat)/2, hjust = 0, size = 3.3,
           colour = book_palette$accent, label = "accounted for\nby the line  3.24") +
  annotate("text", x = xi + 0.7, y = (yhat + yi)/2, hjust = 0, size = 3.3,
           colour = book_palette$reject, label = "residual\n2.46") +
  annotate("text", x = 20, y = ybar + 0.42, hjust = 0.5, size = 3.2,
           colour = "grey35", label = "class average") +
  annotate("text", x = 30, y = 8.6, hjust = 1, size = 3.2,
           colour = "grey25", label = "fitted line") +
  labs(x = "Classes attended", y = "Semester GPA") +
  theme_book()
```

Read the three segments and the arithmetic is immediate:

$$\underbrace{(y_i - \bar{y})}_{\text{total}}
= \underbrace{(\hat{y}_i - \bar{y})}_{\text{accounted for}}
+ \underbrace{(y_i - \hat{y}_i)}_{\text{residual}}$$

For this student, $-5.70 = -3.24 + (-2.46)$.

::: {.keyidea}
Every observation's distance from the mean splits into two parts: the part the
fitted line accounts for, and the part it does not.

$$\text{total} = \text{accounted for by the line} + \text{residual}$$

That identity is not an approximation. Adding $\hat{y}_i$ and subtracting it
again changes nothing, so it holds exactly, for every observation, in every
dataset.
:::

### From one observation to all of them {-}

Now do the same for all 680 students. Square each distance so that departures
above and below the mean do not cancel — the same reason we squared residuals in
Section \@ref(sr-ols) — and add:

$$\sum(y_i - \bar{y})^2 = \sum(\hat{y}_i - \bar{y})^2 + \sum(y_i - \hat{y}_i)^2$$

The cross-term that ought to appear when squaring a sum vanishes exactly, and it
does so because of property 2 above: the residuals are uncorrelated with $x$, and
therefore with $\hat{y}$, which is a linear function of $x$. That algebraic
identity is what makes this decomposition work at all.

Only now are the names worth having.

::: {.keyidea}
| Quantity | Name | What it measures |
|:--|:--|:--|
| $\sum(y_i - \bar{y})^2$ | **SST**, total sum of squares | All the variation in $y$ |
| $\sum(\hat{y}_i - \bar{y})^2$ | **SSE**, explained sum of squares | The part the line accounts for |
| $\sum \hat{u}_i^2$ | **SSR**, residual sum of squares | The part it does not |

$$\text{SST} = \text{SSE} + \text{SSR}$$
:::

::: {.watchout}
"Explained" is conventional and overstates the case. The line has *accounted
for* that variation in an arithmetic sense; whether it has explained anything
depends on whether the model is appropriate and the relationship causal.

Nothing in SSE establishes that attendance produces grades. A regression of
grades on shoe size would produce a perfectly respectable SSE in a sample of
children, where both rise with age.
:::

### The coefficient of determination {-}

Dividing through by SST turns the decomposition into shares.

::: {.keyidea}
$$R^2 = \frac{\text{SSE}}{\text{SST}} = 1 - \frac{\text{SSR}}{\text{SST}}$$

$R^2$ is the proportion of the variation in $y$ accounted for by the regression.
It lies between 0 and 1, and in a simple regression it is exactly the square of
the correlation between $x$ and $y$.
:::

```{r ols-r2}
sst <- sum((y - mean(y))^2)
sse <- sum((fitted(fit) - mean(y))^2)
ssr <- sum(u_hat^2)

round(c(SST = sst, SSE = sse, SSR = ssr, SSE_plus_SSR = sse + ssr), 2)

c(r_squared    = sse / sst,
  one_minus    = 1 - ssr / sst,
  from_summary = summary(fit)$r.squared,
  cor_squared  = cor(x, y)^2)
```

SSE and SSR add to SST exactly, and all four routes to $R^2$ agree.

So attendance accounts for about 31% of the variation in semester GPA. The other
69% is everything else — ability, effort, subject, luck — which is to say, the
error term of Section \@ref(sr-error) measured as a share.

::: {.watchout}
A low $R^2$ does not mean a regression is useless, and a high one does not mean
it is right.

$R^2$ measures how tightly the points cluster around the line. It says nothing
about whether $\beta_1$ is estimated accurately, and nothing whatever about
whether the relationship is causal.

In economics, $R^2$ values of 0.1 or 0.2 are common and unremarkable. Human
behaviour is not tidy, and a model that accounted for most of the variation in
something like earnings would be more suspicious than impressive.
:::

## Every sample gives a different line {#ols-sampling}

::: {.question}
We have one sample and one line. What would a different sample have given?
:::

Section \@ref(sr-srf) showed two sections of thirty students producing slopes of
0.300 and 0.162. That was a demonstration with two samples. Let us do it
properly.

Draw forty students at random from the 680, fit a line, record the slope. Then
do it again. And again, five hundred times.

```{r ols-many-lines, cache = TRUE, fig.height = 4.2, fig.cap = "One hundred fitted lines, each from a different random sample of forty students, in grey. The line fitted to all 680 students is in orange. Every sample gives a different answer."}
set.seed(1)

draw_line <- function(n = 40) {
  samp <- students[sample(nrow(students), n, replace = TRUE), ]
  coef(lm(sem_gpa ~ attendance, data = samp))
}

many <- t(replicate(100, draw_line()))
colnames(many) <- c("intercept", "slope")

ggplot(students, aes(attendance, sem_gpa)) +
  geom_point(alpha = 0.08, colour = book_palette$ink, size = 0.9) +
  geom_abline(data = as.data.frame(many),
              aes(intercept = intercept, slope = slope),
              colour = "grey55", linewidth = 0.3, alpha = 0.6) +
  geom_abline(intercept = coef(fit)[1], slope = coef(fit)[2],
              colour = book_palette$reject, linewidth = 1.2) +
  labs(x = "Classes attended", y = "Semester GPA") +
  theme_book()
```

Every grey line came from the same population. None of them is the population
regression function, and a researcher holding any one of them would have no way
of knowing how far off it was.

::: {.keyidea}
This is the situation of Unit \@ref(estimation), with a slope in place of a mean.

$\hat{\beta}_1$ is computed from a sample, so it is a **random variable**. It has
a distribution across the samples that might have been drawn, and that
distribution is what tells us how much to trust the one estimate we have.
:::

### Looking only at the slopes {-}

Discard the lines and keep the numbers. Five hundred samples give five hundred
slopes.

```{r ols-slope-dist, cache = TRUE, fig.height = 3.4, fig.cap = "The sampling distribution of the slope: 5,000 slopes, each from a fresh sample of forty students. The orange line marks the slope from all 680 students. The distribution is centred on it and roughly symmetric."}
set.seed(1)
slopes <- replicate(5000, draw_line()["attendance"])

ggplot(data.frame(slope = slopes), aes(slope)) +
  geom_histogram(bins = 45, fill = book_palette$fill, colour = "white",
                 linewidth = 0.2) +
  geom_vline(xintercept = coef(fit)[2], colour = book_palette$reject,
             linewidth = 1) +
  labs(x = expression(hat(beta)[1]), y = "Number of samples") +
  theme_book()
```

That histogram is the **sampling distribution of $\hat{\beta}_1$**, and it is the
central object of this section. Two features of it matter.

```{r ols-slope-summary}
round(c(centre_of_distribution = mean(slopes),
        spread_of_distribution = sd(slopes),
        slope_from_all_680     = coef(fit)[2]), 4)
```

**It is centred on the right place.** The average of the five thousand slopes is
0.190, against 0.189 from the full data. No individual sample got it exactly
right, and the average of them did.

**It has a spread**, and that spread is `r round(sd(slopes), 4)`.

::: {.keyidea}
The **standard error** of $\hat{\beta}_1$ is the standard deviation of its
sampling distribution.

That is the same definition as in Unit \@ref(estimation), where the standard
error of $\bar{X}$ was the standard deviation of *its* sampling distribution. The
estimator has changed; the idea has not.
:::

The parallel is worth setting out in full, because everything that follows in
this unit is the second column of a table whose first column the reader already
knows.

Table: (\#tab:ols-parallel) Regression inference is the machinery of Units 2 to 4 applied to a different estimator.

| Unit \@ref(estimation)–\@ref(one-sample) | This unit |
|:--|:--|
| One sample gives one $\bar{x}$ | One sample gives one $\hat{\beta}_1$ |
| Repeat the sampling | Repeat the sampling |
| Sampling distribution of $\bar{X}$ | Sampling distribution of $\hat{\beta}_1$ |
| $E(\bar{X}) = \mu$ | $E(\hat{\beta}_1) = \beta_1$ |
| $\mathrm{Var}(\bar{X}) = \sigma^2/n$ | $\mathrm{Var}(\hat{\beta}_1) = \;?$ |
| $\mathrm{se}(\bar{X}) = \sigma/\sqrt{n}$ | $\mathrm{se}(\hat{\beta}_1) = \;?$ |
| Confidence interval for $\mu$ | Confidence interval for $\beta_1$ |
| $t$-test of $H_0: \mu = \mu_0$ | $t$-test of $H_0: \beta_1 = 0$ |

Nothing in the right-hand column is a new idea. What remains is to fill in the
two question marks.

### What makes the distribution narrow? {-}

Before deriving anything, we can find out by experiment. Three things plausibly
matter: how many students we draw, how spread out their attendance is, and how
noisy grades are around the line.

Take the first two in turn.

```{r ols-sd-by-n, cache = TRUE}
set.seed(1)
sd_at <- function(n) sd(replicate(2000, draw_line(n)["attendance"]))

round(sapply(c(20, 40, 80, 160, 320), sd_at), 4)
```

Larger samples give a tighter distribution, and the pattern is familiar: each
quadrupling of $n$ roughly halves the spread. That is the $\sqrt{n}$ rule of Unit
\@ref(estimation), arriving again.

Now hold the sample size fixed at forty and change the *range of attendance*
instead. Restrict the pool to students who attended between 24 and 30 classes,
so that $x$ varies far less.

```{r ols-sd-by-spread, cache = TRUE}
narrow <- students[students$attendance >= 24 & students$attendance <= 30, ]

set.seed(2)
sd_narrow <- sd(replicate(2000, {
  samp <- narrow[sample(nrow(narrow), 40, replace = TRUE), ]
  coef(lm(sem_gpa ~ attendance, data = samp))[2]
}))

round(c(sd_x_full    = sd(students$attendance),
        sd_x_narrow  = sd(narrow$attendance),
        sd_b1_full   = sd_at(40),
        sd_b1_narrow = sd_narrow), 4)
```

This is the striking one. The sample size is identical. What changed is that
attendance now varies by 1.9 classes instead of 5.5 — and the slope estimate
became about two and a half times more variable.

::: {.intuition}
A slope is a rate of change, and rates of change are measured by observing
change.

Comparing students who attended 27 classes with students who attended 28 tells
you very little about how attendance matters. Comparing those who attended 5 with
those who attended 30 tells you a great deal. Spread in the explanatory variable
is *information*, and a study that observes $x$ over a narrow range will estimate
its effect poorly however many observations it collects.

This is also why the median split of Section \@ref(sr-motivation) performed so
badly. It discarded most of the spread in attendance.
:::

The third determinant, the noise in $y$ around the line, needs no simulation to
believe: if grades were entirely determined by attendance, every sample would
recover the same line exactly.

::: {.keyidea}
Three things make the sampling distribution of $\hat{\beta}_1$ narrow.

1. A **large sample**.
2. **Wide variation in $x$**.
3. **Little noise** in $y$ around the line.

We have found all three by experiment. The next two sections establish them
algebraically, which requires being explicit about what we are assuming.
:::

## Why does OLS get the right answer on average? {#ols-unbiased}

::: {.question}
The histogram was centred on the truth. Was that luck?
:::

The histogram in Figure \@ref(fig:ols-slope-dist) had two striking features.

First, the slopes varied from sample to sample — sometimes a good deal.

Second, they fluctuated *around* the full-sample value rather than sitting
systematically above or below it. The average of five thousand slopes was 0.190
against 0.189 from all 680 students.

Was that simply good fortune with this population, or does OLS always behave
this way?

Showing that it always does requires the first four assumptions of the simple
linear regression model.

::: {.keyidea}
**SLR1** The population model is $y = \beta_0 + \beta_1 x + u$.
**SLR2** The observations are a random sample from that population.
**SLR3** The $x_i$ are not all equal.
**SLR4** $E(u \mid x) = 0$.

Nothing about the *spread* of the errors is assumed yet.
:::

::: {.theory}
Write the slope in deviation form and substitute the model. With
$\text{SST}_x = \sum(x_i - \bar{x})^2$,

$$
\begin{aligned}
\hat{\beta}_1 &= \frac{\sum(x_i - \bar{x})y_i}{\text{SST}_x}
   && \text{since } \textstyle\sum(x_i-\bar{x})\bar{y} = 0 \\[4pt]
&= \frac{\sum(x_i - \bar{x})(\beta_0 + \beta_1 x_i + u_i)}{\text{SST}_x}
   && \text{substituting SLR1} \\[4pt]
&= \beta_1 + \frac{\sum(x_i - \bar{x})u_i}{\text{SST}_x}
\end{aligned}
$$

because $\sum(x_i - \bar{x}) = 0$ kills the $\beta_0$ term and
$\sum(x_i - \bar{x})x_i = \text{SST}_x$ leaves $\beta_1$.
:::

That last line deserves to be pulled out and looked at on its own.

::: {.keyidea}
$$\hat{\beta}_1 \;=\; \underbrace{\beta_1}_{\text{truth}}
\;+\; \underbrace{\frac{\sum(x_i - \bar{x})u_i}{\text{SST}_x}}_{\text{sampling error}}$$

**Estimate = truth + sampling error.**
:::

The equation separates two quite different things.

The first term, $\beta_1$, is the quantity we are trying to estimate. It is a
fixed feature of the population and it is identical in every sample.

The second term is the price of observing only one sample. It depends entirely
on the unobserved errors $u_i$ — on which particular students happened to be
drawn and how far each of them sat from the line. It is different in every
sample, and it is the whole reason the grey lines in Figure
\@ref(fig:ols-many-lines) fan out.

So the question of whether OLS is centred correctly comes down to one thing: does
that second term average out to zero?

Taking expectations conditional on the $x$'s, and using SLR4 to set
$E(u_i \mid x) = 0$,

$$E(\hat{\beta}_1) = \beta_1 + \frac{\sum(x_i - \bar{x})E(u_i \mid x)}{\text{SST}_x} = \beta_1$$

It does. Across repeated samples $\hat{\beta}_1$ is centred on $\beta_1$, which
is the first question mark in Table \@ref(tab:ols-parallel) filled in, and what
the histogram was showing.

::: {.intuition}
The proof is worth having for a second reason. It shows that unbiasedness hangs
on exactly one place — the second term — and on exactly one condition, that the
average error does not change with $x$.

If students who attend more classes also happen to be more motivated, then the
errors are systematically larger at larger $x$. The second term no longer
vanishes, and the estimator is biased by precisely its expectation.

That single term is why omitted variables matter, and Section
\@ref(ols-causal) returns to it with this equation in hand.
:::

## How precise is OLS? {#ols-variance}

::: {.question}
The histogram had a width as well as a centre. Can we know that width from one
sample?
:::

The sampling distribution in Section \@ref(ols-sampling) had a standard deviation
of `r round(sd(slopes), 4)`. That number is exactly what we want: it says how far
a slope from a single sample of forty is liable to fall from the truth.

But look at how we obtained it. We drew five thousand samples.

Nobody does that. A survey is run once, at considerable expense, and yields one
sample and one slope. The standard deviation of the distribution those samples
came from is not something we can watch.

::: {.keyidea}
So the question is this. Can we recover, from a *single* sample, the spread of a
distribution we could otherwise see only by drawing thousands?

Remarkably, yes.
:::

### The formula {-}

The width of the sampling distribution turns out to depend on two things, and
both are things we can measure.

$$\mathrm{Var}(\hat{\beta}_1) = \frac{\sigma^2}{\sum(x_i - \bar{x})^2}$$

The **numerator** is the noise: how far individual observations scatter around
the line. Noisier relationships are harder to pin down.

The **denominator** is the total variation in $x$ — the same sum of squares we
met in Section \@ref(ols-fit), computed on the explanatory variable rather than
the outcome. It measures how much the data actually tell us about $x$.

Since $\sigma^2$ is unknown, it is estimated from the residuals. Two parameters
were estimated on the way, so the divisor is $n-2$ — the same accounting as
Section \@ref(u3-variance), where estimating one mean cost one degree of freedom:

$$\hat{\sigma}^2 = \frac{\sum \hat{u}_i^2}{n - 2}
\qquad\qquad
\mathrm{se}(\hat{\beta}_1) = \frac{\hat{\sigma}}{\sqrt{\sum(x_i - \bar{x})^2}}$$

```{r ols-se-byhand, cache = TRUE}
n <- length(y)
sigma_hat <- sqrt(sum(u_hat^2) / (n - 2))
se_b1     <- sigma_hat / sqrt(sum((x - mean(x))^2))

c(sigma_hat = sigma_hat, se_b1 = se_b1,
  se_from_lm = summary(fit)$coefficients[2, 2])
```

### Does it work? {-}

A formula claiming to reconstruct a distribution we never observe deserves to be
checked rather than believed.

Build a population where the assumptions hold exactly — a genuinely linear
relationship, errors of constant variance — then draw from it repeatedly and
compare what the formula says against what the samples actually do.

```{r ols-formula-check, cache = TRUE}
set.seed(4)
x_fixed <- runif(40, 2, 32)                    # forty students, attendance 2 to 32

check <- replicate(5000, {
  y_sim <- 1.56 + 0.189 * x_fixed + rnorm(40, sd = 1.53)
  m <- lm(y_sim ~ x_fixed)
  c(slope = coef(m)[2], se = summary(m)$coefficients[2, 2])
})

round(c(simulated_sd    = sd(check["slope.x_fixed", ]),
        average_formula = mean(check["se", ]),
        exact_formula   = 1.53 / sqrt(sum((x_fixed - mean(x_fixed))^2))), 4)
```

The three agree to three decimal places. The formula, computed from one sample of
forty, recovers the spread that five thousand samples reveal.

::: {.keyidea}
In other words: the standard error printed by `lm()` is an estimate of the
standard deviation of the sampling distribution we simulated in Section
\@ref(ols-sampling).

That sentence connects the whole unit. The histogram is the thing; the standard
error is our estimate of how wide it is, computed without ever drawing it.
:::

### Why the denominator matters {-}

Most readers remember that larger samples give smaller standard errors. Far fewer
remember the other half, which is at least as important.

Look at the denominator again and ask what happens in extreme cases.

**What if every student attended exactly 25 classes?** Then $x_i = \bar{x}$ for
everyone, $\sum(x_i - \bar{x})^2 = 0$, and the variance is undefined. There is no
slope to estimate — we have observed grades at one attendance level and nothing
about any other. This is assumption SLR3, arriving as a consequence rather than a
stipulation.

**What if attendance varies only between 24 and 26?** The denominator is small
and the standard error is large. We are trying to infer a rate of change from
almost no change.

**What if attendance ranges from 2 to 32?** The denominator is large and the
slope is pinned down tightly.

The following figure makes the point with two datasets built from the *same* true
line and the *same* error variance. Only the spread of $x$ differs.

```{r ols-spread-plot, echo = FALSE, fig.height = 3.6, cache = TRUE, fig.cap = "Two samples of forty from the same model: the same true slope of 0.189, the same noise, and the same sample size. Only the range of attendance differs. On the left the line is pinned down; on the right, where attendance varies by two classes rather than thirty, the same data are consistent with wildly different slopes."}
set.seed(6)
b0 <- 1.56; b1 <- 0.189; sg <- 1.53; nn <- 40

xw <- seq(2, 32, length.out = nn);  yw <- b0 + b1*xw + rnorm(nn, sd = sg)
xn <- seq(24, 26, length.out = nn); yn <- b0 + b1*xn + rnorm(nn, sd = sg)

sp <- rbind(
  data.frame(x = xw, y = yw, panel = "Attendance 2 to 32   (se = 0.031)"),
  data.frame(x = xn, y = yn, panel = "Attendance 24 to 26   (se = 0.329)")
)
sp$panel <- factor(sp$panel, levels = unique(sp$panel))

ggplot(sp, aes(x, y)) +
  geom_point(alpha = 0.65, colour = book_palette$ink, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, colour = book_palette$accent,
              fill = book_palette$fill, linewidth = 0.9) +
  geom_abline(intercept = b0, slope = b1, colour = book_palette$reject,
              linetype = "dashed", linewidth = 0.7) +
  facet_wrap(~ panel, scales = "free_x") +
  labs(x = "Classes attended", y = "Semester GPA") +
  theme_book()
```

The dashed orange line is the truth in both panels, and it is the same line.

On the left the fitted line sits almost on top of it: an estimated slope of 0.178
against a true 0.189. On the right the fitted slope is 0.360 — nearly twice the
truth — and the shaded band around it is so wide as to be useless.

Note that the two panels have different horizontal scales, so compare the fitted
lines to the dashed one within each panel rather than by eye across them.

Nothing about the relationship changed. Only the range over which we observed
it.

::: {.intuition}
A slope is a rate of change, and rates of change are measured by observing
change.

Comparing students who attended 24 classes with students who attended 26 tells
you very little about how attendance matters. Comparing those who attended 5 with
those who attended 30 tells you a great deal.

Spread in the explanatory variable is *information*. A study that observes $x$
over a narrow range will estimate its effect poorly however many observations it
collects — which is also why the median split of Section \@ref(sr-motivation)
did so badly.
:::

### Where the formula comes from {-}

Having seen that the formula works and what its parts mean, we can derive it.

One further assumption is needed, and it is worth seeing why. The derivation adds
up the contributions of the individual errors, and to add them we must know how
variable each one is. So far we have said nothing about that at all.

::: {.keyidea}
**SLR5 — homoskedasticity.** $\mathrm{Var}(u \mid x) = \sigma^2$, the same at
every value of $x$.

This is what allows a single number $\sigma^2$ to stand in for the variability of
every error. It played no part in unbiasedness, and Section \@ref(ols-hetero)
examines what happens when it fails.
:::

::: {.theory}
Start from the expression that gave unbiasedness, and treat the $x$'s as fixed:

$$
\begin{aligned}
\mathrm{Var}(\hat{\beta}_1)
&= \mathrm{Var}\!\left(\beta_1 + \frac{\sum(x_i - \bar{x})u_i}{\text{SST}_x}\right) \\[4pt]
&= \mathrm{Var}\!\left(\frac{\sum(x_i - \bar{x})u_i}{\text{SST}_x}\right)
   && \beta_1 \text{ is a constant} \\[4pt]
&= \frac{1}{\text{SST}_x^2}\sum(x_i - \bar{x})^2\,\mathrm{Var}(u_i)
   && \text{independence} \\[4pt]
&= \frac{\sigma^2 \,\text{SST}_x}{\text{SST}_x^2}
   && \text{homoskedasticity, SLR5} \\[4pt]
&= \frac{\sigma^2}{\text{SST}_x}
\end{aligned}
$$
:::

::: {.keyidea}
**Why is the slope estimated more precisely?**

The standard error becomes smaller when

- the relationship is **less noisy** — small $\sigma$;
- the sample is **larger** — more terms in $\sum(x_i - \bar{x})^2$;
- the explanatory variable **varies over a wider range** — bigger terms in that
  same sum.

The last two both enter through the denominator, which is why $n$ and the spread
of $x$ are not really two separate considerations. What matters is the total
variation in $x$, and either more observations or a wider range will increase it.
:::

Those are the three findings of Section \@ref(ols-sampling), now derived rather
than observed. The second question mark in Table \@ref(tab:ols-parallel) is
filled in.

## When the error variance is not constant {#ols-hetero}

::: {.question}
SLR5 assumed the errors have the same variance at every value of $x$. Why would
they?
:::

The variance formula of the previous section rested on **homoskedasticity**:

$$\mathrm{Var}(u \mid x) = \sigma^2 \quad \text{for every } x$$

Read that as a claim about the world. It says the unobserved factors are just as
spread out among students who attended 5 classes as among those who attended 30,
and just as spread out among poor households as among rich ones.

Often they are not.

::: {.intuition}
Consider annual earnings against years of schooling.

Someone with two years of schooling has little scope to be unusual. Their
earnings are constrained from above by the work available to them, and there is
not much room between the floor and that ceiling.

Someone with sixteen years of schooling might be a clerk, a teacher, a doctor or
a company director. The same qualification is consistent with an enormous range
of outcomes, because what happens after school depends on far more than school.

So the spread of the errors grows with $x$. This is **heteroskedasticity**, and in
economic data it is the normal case rather than the exception.
:::

### The gap we already saw {-}

Section \@ref(ols-variance) checked the standard error formula against a
simulation and found it exact — on data built so that every assumption held.

On the real attendance data the agreement is good rather than exact. The formula
gives a standard error of about 0.046 for a sample of forty, while resampling
gives about 0.053.

That gap is the subject of this section. The formula's assumptions hold only
approximately in real data: Section \@ref(sr-prf) already showed the conditional
mean of GPA is not exactly linear, and the error variance is not quite constant
either. A formula is only as good as the assumptions underneath it, which is
worth remembering whenever software prints a standard error to four decimals.

### Seeing it {-}

Regress annual earnings on years of education for the 20,000 wage earners, and
look at how the residuals are spread.

```{r ols-het-see}
wages <- read.csv("data/wages-india-synthetic.csv")
fit_w <- lm(annual_earnings ~ education, data = wages)

u_w <- residuals(fit_w)
f_w <- fitted(fit_w)

quartile <- cut(f_w, quantile(f_w, 0:4/4), include.lowest = TRUE)
round(tapply(u_w, quartile, sd))
```

The residuals in the top quarter of fitted values are more than three times as
spread out as those in the bottom quarter.

```{r ols-het-plot, echo = FALSE, fig.height = 3.6, fig.cap = "Residuals from a regression of annual earnings on years of schooling, shown as a boxplot at each level of education. The boxes widen steadily to the right: errors are far more variable where schooling is high. Homoskedasticity would appear as boxes of constant height. A few extreme residuals lie above the range plotted."}
res_df <- data.frame(education = wages$education, resid = u_w)

ggplot(res_df, aes(factor(education), resid)) +
  geom_hline(yintercept = 0, colour = book_palette$reject, linewidth = 0.6) +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.25,
               fill = book_palette$fill, colour = "grey35",
               linewidth = 0.35) +
  coord_cartesian(ylim = c(-2e5, 4e5)) +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
  labs(x = "Years of education", y = "Residual (₹)") +
  theme_book()
```

::: {.keyidea}
The **residual plot** — residuals against fitted values — is the standard
diagnostic.

Homoskedasticity appears as a band of roughly constant width. Heteroskedasticity
appears as a fan, widening or narrowing across the plot.

Look at it every time. It costs one line of code and catches a great deal.
:::

Education takes only whole-number values, so a plain scatter of residuals against
fitted values collapses into a series of vertical stripes. Drawing a box at each
level of education shows the same thing more legibly.

Formal tests exist — the Breusch–Pagan and White tests both regress squared
residuals on the explanatory variables and ask whether anything is there. They
are useful, but a plot usually settles the matter faster and tells you the shape
of the problem rather than merely its presence.

### What breaks, and what does not {-}

This is the part worth being careful about, because heteroskedasticity is often
described as though it invalidated everything.

::: {.keyidea}
**Unbiasedness survives.** The proof of $E(\hat{\beta}_1) = \beta_1$ used SLR1 to
SLR4 and never mentioned SLR5. The estimate is still centred on the truth.

**The variance formula fails.** $\mathrm{Var}(\hat{\beta}_1) = \sigma^2/\text{SST}_x$
was derived by replacing every $\mathrm{Var}(u_i)$ with a common $\sigma^2$. If
there is no common $\sigma^2$, that step is invalid.

So the estimate is fine and the *reported precision of the estimate* is wrong —
which means the standard errors, the $t$ statistics, the $p$-values and the
confidence intervals are all unreliable.
:::

How wrong? We can find out, by simulating from a model where we know both the
truth and the severity of the problem.

```{r ols-het-sim, cache = TRUE}
robust_se <- function(m) {
  X <- model.matrix(m); u <- residuals(m)
  n <- nrow(X); k <- ncol(X)
  XtXi <- solve(crossprod(X))
  V <- XtXi %*% (t(X) %*% diag(u^2) %*% X) %*% XtXi * n / (n - k)
  sqrt(diag(V))[2]
}

set.seed(8)
out <- replicate(4000, {
  x <- rexp(60, rate = 1/5)              # skewed x, as economic data often is
  y <- 2 + 1 * x + rnorm(60, sd = x)     # sd of the error grows with x
  m <- lm(y ~ x)
  c(b1 = coef(m)[2], classical = coef(summary(m))[2, 2], robust = robust_se(m))
})

round(c(actual_sd_of_b1  = sd(out[1, ]),
        avg_classical_se = mean(out[2, ]),
        avg_robust_se    = mean(out[3, ])), 4)
```

The true slope is 1, and across four thousand samples $\hat{\beta}_1$ has a
standard deviation of 0.43. That is the actual variability of the estimator.

The classical standard error reports 0.18 — less than half of it. Software that
prints 0.18 is claiming a precision the estimator does not have.

The consequence for confidence intervals is severe.

```{r ols-het-coverage}
round(c(classical = mean(abs(out[1, ] - 1) < qt(0.975, 58) * out[2, ]),
        robust    = mean(abs(out[1, ] - 1) < qt(0.975, 58) * out[3, ])), 3)
```

An interval advertised as 95% contains the true slope 59% of the time.

### Robust standard errors {-}

The fix does not touch $\hat{\beta}_1$, which needs no fixing. It replaces the
formula for its variance with one that does not assume a common $\sigma^2$,
using each observation's own squared residual in place of a single estimate.

These are called **heteroskedasticity-robust standard errors**, or often just
robust standard errors.

In the simulation above they report 0.33 against a true 0.43, and lift coverage
from 59% to 85%. A large improvement, and — worth noticing — not a complete
repair at this sample size and this severity.

```{r ols-het-robust}
round(c(classical_se = coef(summary(fit_w))[2, 2],
        robust_se    = robust_se(fit_w)), 2)
```

For the earnings regression the robust standard error is 175 against a classical
137 — a quarter larger again. Any conclusion resting on the difference between
those two numbers was resting on an assumption the residual plot had already
refused.

::: {.watchout}
Three things robust standard errors do not do.

They do not fix bias. There was none to fix; heteroskedasticity does not bias
$\hat{\beta}_1$.

They do not fix a wrong functional form. A fan-shaped residual plot sometimes
means the relationship is not linear rather than that the variance is
non-constant, and robust standard errors would paper over that.

They do not work miracles in small samples, as the simulation shows. They are an
improvement, not a guarantee.
:::

### Sometimes the variable is the problem {-}

Robust standard errors accept the heteroskedasticity and correct for it. The
other response is to ask why it is there.

Earnings are a highly skewed variable, and skewed variables often produce
fan-shaped residuals simply because their upper tail is long. Taking logarithms
compresses that tail.

```{r ols-het-logs}
fit_log <- lm(log(annual_earnings) ~ education, data = wages)

q_log <- cut(fitted(fit_log), quantile(fitted(fit_log), 0:4/4),
             include.lowest = TRUE)
round(tapply(residuals(fit_log), q_log, sd), 4)
```

The spread is now essentially identical across quartiles — a ratio of 1.01
against 3.61 in levels. The heteroskedasticity was not a deep feature of the
world; it was a consequence of measuring earnings in rupees rather than in
proportional terms.

Section \@ref(ols-forms) takes up what the log form means for interpretation.
For now, note that a residual plot which fans out is sometimes telling you about
the variance and sometimes telling you that the variable wants transforming.

::: {.intuition}
Modern practice is to report robust standard errors by default in
cross-sectional work.

The reasoning is the same asymmetric-risk argument as Welch's test in Section
\@ref(ts-unknown). When the errors really are homoskedastic, robust standard
errors cost very little. When they are not, classical standard errors mislead —
and you can never establish homoskedasticity, only fail to reject it.
:::

## Inference about the slope {#ols-inference}

::: {.question}
Is the slope we found distinguishable from zero?
:::

$\hat{\beta}_1$ is an estimator with a standard error, so everything in Unit
\@ref(one-sample) applies without modification.

The hypothesis of interest is almost always that $x$ has no effect at all:

$$H_0 : \beta_1 = 0 \qquad H_A : \beta_1 \neq 0$$

and the test statistic has the form it always has — estimate minus hypothesised
value, over standard error:

$$t = \frac{\hat{\beta}_1 - 0}{\mathrm{se}(\hat{\beta}_1)}$$

compared against $t$ with $n - 2$ degrees of freedom.

```{r sr-inference}
summary(fit)$coefficients
```

$\hat{\beta}_1 = 0.189$ with a standard error of 0.0107, giving $t = 17.6$ on
678 degrees of freedom and a $p$-value indistinguishable from zero. We reject
the claim that attendance is unrelated to GPA.

The confidence interval is the more informative report:

```{r sr-confint}
confint(fit, level = 0.95)
```

Each additional class is associated with between 0.168 and 0.210 of a GPA point,
with 95% confidence. The interval excludes zero, agreeing with the test as it
must.

::: {.watchout}
"Statistically significant" is a claim about distinguishability from zero, not
about magnitude. With 680 observations, very small slopes clear the bar easily.

Always ask what the slope means in the units of the problem. Here 0.19 GPA
points per class is large. In the regression of final examination score on
attendance in the same dataset, the slope is 0.117 marks per class and equally
"significant" — and 0.117 marks is nothing at all.
:::

## Functional forms {#ols-forms}

::: {.question}
A slope of $-0.006$ km per litre per kilogram. Is that a lot?
:::

The regression of fuel efficiency on weight in Section \@ref(ols-practice)
returned a slope of $-0.00601$. Nothing about that number tells you whether the
effect is large, because it depends entirely on the units chosen. Measure weight
in tonnes and the slope becomes $-6.01$. Measure efficiency in litres per
hundred kilometres and the sign flips.

Two problems follow, and one transformation solves both.

**Units are arbitrary.** A coefficient in kilograms and one in tonnes describe
the same relationship, but only one of them is easy to talk about. We would
prefer a measure that does not change when someone changes their mind about units.

**Straight lines are not always the right shape.** Section \@ref(sr-prf) showed
that the conditional mean of GPA against attendance was not exactly linear, and
took a line as an approximation. Sometimes a better approximation is available at
no cost, by transforming a variable before fitting.

Taking logarithms addresses both.

### The four forms {-}

Either variable may be entered in levels or in logarithms, giving four
combinations. The interpretation of $\beta_1$ changes in each.

Table: (\#tab:ols-forms) The four functional forms and what the slope means in each.

| Model | Equation | Interpretation of $\beta_1$ |
|:--|:--|:--|
| Level–level | $y = \beta_0 + \beta_1 x + u$ | A one-unit rise in $x$ changes $y$ by $\beta_1$ **units** |
| Log–level | $\ln y = \beta_0 + \beta_1 x + u$ | A one-unit rise in $x$ changes $y$ by $100\beta_1$ **per cent** |
| Level–log | $y = \beta_0 + \beta_1 \ln x + u$ | A 1% rise in $x$ changes $y$ by $\beta_1/100$ **units** |
| Log–log | $\ln y = \beta_0 + \beta_1 \ln x + u$ | A 1% rise in $x$ changes $y$ by $\beta_1$ **per cent** |

The last is an **elasticity**, and it is unit-free: it does not matter whether
weight is in kilograms or tonnes, or price in rupees or lakhs.

::: {.theory}
All four interpretations follow from one fact of calculus. For small changes,

$$\Delta \ln y \;\approx\; \frac{\Delta y}{y}$$

that is, a change in the logarithm is approximately a *proportional* change in
the variable. Multiplying by 100 turns it into a percentage.

**Log–level.** Differentiating $\ln y = \beta_0 + \beta_1 x$ with respect to $x$,

$$\frac{1}{y}\frac{dy}{dx} = \beta_1
\qquad\Longrightarrow\qquad
\frac{dy/y}{dx} = \beta_1$$

so a one-unit rise in $x$ raises $y$ by a proportion $\beta_1$, which is
$100\beta_1$ per cent.

**Level–log.** From $y = \beta_0 + \beta_1 \ln x$,

$$dy = \beta_1 \, d(\ln x) = \beta_1 \frac{dx}{x}$$

A 1% rise in $x$ means $dx/x = 0.01$, so $dy = 0.01\beta_1 = \beta_1/100$ units.

**Log–log.** From $\ln y = \beta_0 + \beta_1 \ln x$,

$$\frac{dy/y}{dx/x} = \beta_1$$

which is the definition of an elasticity: the ratio of the proportional change in
$y$ to the proportional change in $x$.
:::

### Fuel efficiency, in elasticity form {-}

```{r ols-forms-cars}
cars <- read.csv("data/car-mileage.csv")

level_level <- lm(mileage_kmpl ~ weight_kg, data = cars)
log_log     <- lm(log(mileage_kmpl) ~ log(weight_kg), data = cars)

round(rbind(level_level = c(slope = coef(level_level)[2],
                            r_squared = summary(level_level)$r.squared),
            log_log     = c(slope = coef(log_log)[2],
                            r_squared = summary(log_log)$r.squared)), 4)
```

The log–log slope is $-0.825$: **a car that is 1% heavier is about 0.83% less
fuel-efficient.** That is a sentence anyone can evaluate, and it would be
unchanged if weight were recorded in tonnes.

The fit also improves, from $R^2 = 0.65$ to $0.71$. That is a modest gain rather
than a dramatic one, and it is worth being clear that the case for the log form
here rests mainly on interpretation rather than on fit. Still, the direction is
not an accident: a constant *proportional* penalty for weight describes cars
better than the constant absolute penalty the level–level model imposed.

```{r ols-forms-plot, echo = FALSE, fig.height = 3.4, fig.cap = "The same 74 cars in levels and in logarithms. Both look broadly linear; the logarithmic version is a little tighter around the line, which is the improvement in R-squared from 0.65 to 0.71. The gain here is real but modest -- the case for the log form rests mainly on interpretation."}
d1 <- data.frame(x = cars$weight_kg, y = cars$mileage_kmpl, panel = "Levels")
d2 <- data.frame(x = log(cars$weight_kg), y = log(cars$mileage_kmpl),
                 panel = "Logarithms")
both <- rbind(d1, d2)

ggplot(both, aes(x, y)) +
  geom_point(alpha = 0.55, colour = book_palette$ink, size = 1.3) +
  geom_smooth(method = "lm", se = FALSE, colour = book_palette$accent,
              linewidth = 0.9) +
  facet_wrap(~ panel, scales = "free") +
  labs(x = NULL, y = NULL) +
  theme_book()
```

::: {.watchout}
This is where the warning about linearity belongs.

A great deal of weak empirical work consists of fitting a straight line to a
relationship that is not straight, and reporting the slope as though it described
the whole range. The residual plot of Section \@ref(ols-hetero) catches this too:
a curved band of residuals means the functional form is wrong, not that the
variance is non-constant.

Fitting a line is a choice. Check it.
:::

### The return to schooling {-}

The log–level form is the workhorse of labour economics, because wages are
naturally discussed in percentages.

```{r ols-forms-wages}
wages <- read.csv("data/wages-india-synthetic.csv")

fit_w <- lm(log(annual_earnings) ~ education, data = wages)
round(coef(summary(fit_w)), 5)
```

The slope of 0.0942 says that **one more year of schooling is associated with
about 9.4% higher annual earnings.** Multiply by 100 and read as a percentage;
that is the whole rule.

::: {.watchout}
Two cautions on that number.

It is a simple regression, so it carries every confounder discussed in Section
\@ref(ols-causal). People with more schooling differ in family background and
much else, and the coefficient absorbs all of it. Adding controls in the next
unit lowers it appreciably.

And the percentage reading is an approximation that degrades for large changes.
For a coefficient of 0.0942 the exact figure is
$100(e^{0.0942} - 1) = `r round(100*(exp(0.0942)-1), 2)`$%, which is close
enough. For a coefficient of 0.7 the approximation says 70% and the truth is
$`r round(100*(exp(0.7)-1), 0)`$%. Above about 0.2, use the exact expression.
:::

### Choosing a form {-}

::: {.keyidea}
Three considerations, in order of importance.

**What is the natural way to talk about the variable?** Wages, prices and GDP are
discussed in percentages, so logs. Test scores, class attendance and family size
are discussed in units, so levels.

**Does the relationship look multiplicative?** If a proportional change in $x$
plausibly produces a proportional change in $y$ — as with weight and fuel use —
logs will fit better and mean more.

**Does the scatter suggest it?** Skewed variables often straighten out in logs,
and heteroskedasticity often shrinks at the same time, since logs compress the
upper tail.

What should *not* drive the choice is which version gives the larger $R^2$ or the
smaller $p$-value. Trying transformations until something is significant produces
findings that will not replicate.
:::

::: {.watchout}
**Logarithms are undefined at zero and for negative numbers.**

Earnings data routinely contain zeros — people who did not work. Taking logs
silently deletes them, and since those are rarely a random subset of the sample,
the remaining regression answers a different question from the one asked.

```{r ols-forms-zero}
sum(wages$annual_earnings <= 0)
```

There are none here, which is a property of this simulated dataset rather than of
earnings data in general. With real data, check before transforming and say what
you did.
:::

## What regression does not fix {#ols-causal}

::: {.question}
We have a slope, a standard error, and a tiny $p$-value. Do we have an effect?
:::

No, and it is worth being exact about why.

Everything in this unit estimates a conditional mean. We have established, with
considerable precision, that students who attend more classes have higher
average GPAs. That is a fact about the population.

The claim that attending more classes *raises* a student's GPA is a different
claim, and it requires SLR4: $E(u \mid x) = 0$. Section \@ref(sr-assumptions) put that
assumption in concrete terms — students who attend more must be no more
motivated or able, on average, than students who attend less.

Nobody believes that.

::: {.keyidea}
When $E(u \mid x) \neq 0$, the unbiasedness derivation shows exactly what
happens. We had

$$\hat{\beta}_1 = \beta_1 + \frac{\sum(x_i - \bar{x})u_i}{\text{SST}_x}$$

and the second term no longer has expectation zero. Its expectation is the
**bias**, and its sign is the sign of the relationship between $x$ and the
omitted factors.

If motivated students both attend more *and* score higher for reasons unrelated
to attendance, then $\hat{\beta}_1$ overstates the effect of attendance. This is
**omitted variable bias**.
:::

Notice what this does and does not affect. The estimate is still the best
description of the conditional mean. The standard error is still right. The
$p$-value is still right. What is wrong is the *interpretation* — and no
diagnostic in the regression output will tell you so, because the problem lies
in a variable that is not in the data.

Two routes lead out, and both come later.

The first is to measure the confounders and hold them fixed, which is multiple
regression, the subject of the next unit. It works only for confounders you can
observe.

The second is to find a source of variation in $x$ that has nothing to do with
the unobservables — a lottery, a rule, an accident of administration. That is
the subject of the second part of this book.

::: {.intuition}
Regression is a description of how averages move together. Turning it into a
statement about causes requires an argument about where the variation in $x$
came from, and that argument is made by the researcher, not by the software.
:::

## Summary {#ols-summary}

::: {.keyidea}
- **OLS** minimises $\sum \hat{u}_i^2$. Setting both partial derivatives to zero
  gives the normal equations, whose solution is
  $\hat{\beta}_1 = \sum(x_i-\bar{x})(y_i-\bar{y}) / \sum(x_i-\bar{x})^2$ and
  $\hat{\beta}_0 = \bar{y} - \hat{\beta}_1\bar{x}$.
- The slope is the **sample covariance of $x$ and $y$ over the sample variance of
  $x$**. That reading explains its sign, and it explains why $x$ must vary: with
  no variation in $x$ the denominator is zero.
- By construction $\sum\hat{u}_i = 0$, $\sum x_i\hat{u}_i = 0$, and the fitted
  line passes through $(\bar{x}, \bar{y})$. These are algebraic identities and
  are not evidence about the model.
- $\text{SST} = \text{SSE} + \text{SSR}$, and
  $R^2 = \text{SSE}/\text{SST} = 1 - \text{SSR}/\text{SST}$ is the share of the
  variation in $y$ the regression accounts for. In a simple regression it equals
  the squared correlation between $x$ and $y$.
- $R^2$ measures fit, not correctness. Attendance accounts for 31% of the
  variation in semester GPA; values of 0.1 or 0.2 are unremarkable in economics.
- Under SLR1–SLR4, $E(\hat{\beta}_1) = \beta_1$. The derivation runs through
  $\hat{\beta}_1 = \beta_1 + \sum(x_i-\bar{x})u_i/\text{SST}_x$, which is the
  most useful line in the unit.
- Adding SLR5, $\mathrm{Var}(\hat{\beta}_1) = \sigma^2/\sum(x_i-\bar{x})^2$. Small
  errors and widely spread $x$ both buy precision, which is why spread in the
  explanatory variable is worth seeking out.
- $\hat{\sigma}^2 = \sum\hat{u}_i^2/(n-2)$, the divisor reflecting the two
  parameters estimated on the way.
- Testing $H_0: \beta_1 = 0$ uses
  $t = \hat{\beta}_1/\mathrm{se}(\hat{\beta}_1)$ on $n-2$ degrees of freedom,
  exactly as in Unit \@ref(one-sample). The interval is the more informative
  report.
- When $E(u \mid x) \neq 0$ the second term in the unbiasedness expression does
  not vanish. That is **omitted variable bias**, and nothing in the regression
  output reveals it.
:::

## Practice problems {#ols-practice}

::: {.yourturn}
Work each one before opening the solution. The
[question bank](#ols-question-bank) that follows has no answers.
:::

### Numerical problems {-}

**7.1** For five plots, fertiliser applied ($x$, kg) and yield ($y$, quintals):

| $x$ | 2 | 4 | 6 | 8 | 10 |
|:--|:--|:--|:--|:--|:--|
| $y$ | 4 | 7 | 8 | 12 | 14 |

i. Compute $\bar{x}$, $\bar{y}$, $\hat{\beta}_1$ and $\hat{\beta}_0$.
ii. Write the fitted line and interpret the slope.
iii. Compute the fitted values and residuals, and verify they sum to zero.
iv. Compute $R^2$.

::: {.answer}
**(i)** $\bar{x} = 6$, $\bar{y} = 9$.

| $x_i$ | $y_i$ | $x_i - \bar{x}$ | $y_i - \bar{y}$ | product | $(x_i-\bar{x})^2$ |
|--:|--:|--:|--:|--:|--:|
| 2 | 4 | $-4$ | $-5$ | 20 | 16 |
| 4 | 7 | $-2$ | $-2$ | 4 | 4 |
| 6 | 8 | 0 | $-1$ | 0 | 0 |
| 8 | 12 | 2 | 3 | 6 | 4 |
| 10 | 14 | 4 | 5 | 20 | 16 |
| | | | | **50** | **40** |

$$\hat{\beta}_1 = \frac{50}{40} = 1.25
\qquad
\hat{\beta}_0 = 9 - 1.25(6) = 1.5$$

**(ii)** $\hat{y} = 1.5 + 1.25x$. Each additional kilogram of fertiliser is
associated with 1.25 more quintals of yield.

**(iii)** Fitted values: $4.0, 6.5, 9.0, 11.5, 14.0$. Residuals:
$0, 0.5, -1, 0.5, 0$, which sum to zero as they must.

**(iv)** $\text{SSR} = 0 + 0.25 + 1 + 0.25 + 0 = 1.5$ and
$\text{SST} = 25 + 4 + 1 + 9 + 25 = 64$, so

$$R^2 = 1 - \frac{1.5}{64} = 0.977$$

An unusually tight fit, because these are made-up numbers. Real data rarely
behaves so well.
:::

**7.2** Using the previous question's results:

i. Compute $\hat{\sigma}^2$ and $\hat{\sigma}$.
ii. Compute $\mathrm{se}(\hat{\beta}_1)$.
iii. Test $H_0: \beta_1 = 0$ at the 5% level.
iv. Construct a 95% confidence interval for $\beta_1$.

::: {.answer}
**(i)** With $n = 5$ and two parameters estimated,

$$\hat{\sigma}^2 = \frac{\text{SSR}}{n-2} = \frac{1.5}{3} = 0.5
\qquad \hat{\sigma} = 0.707$$

**(ii)** $\mathrm{se}(\hat{\beta}_1) = \dfrac{0.707}{\sqrt{40}} = 0.1118$.

**(iii)** $t = 1.25/0.1118 = 11.18$ on $n - 2 = 3$ degrees of freedom. The
critical value is $t_{0.025,\,3} = 3.182$, and $11.18 > 3.182$, so we reject
$H_0$.

**(iv)** $1.25 \pm 3.182(0.1118) = [0.894,\ 1.606]$.

Note how wide that interval is despite the excellent fit — three degrees of
freedom buy very little. The critical value of 3.182 rather than 1.96 is doing
most of the damage.
:::

**7.3** A regression of household expenditure on income in a sample of 200
gives $\hat{\beta}_1 = 0.62$ with $\mathrm{se}(\hat{\beta}_1) = 0.05$.

i. Test $H_0: \beta_1 = 0$.
ii. Test $H_0: \beta_1 = 1$, and explain why an economist might care about that
    hypothesis specifically.
iii. Construct a 95% interval and interpret it.

::: {.answer}
**(i)** $t = 0.62/0.05 = 12.4$ on 198 degrees of freedom. Reject overwhelmingly.

**(ii)** $t = (0.62 - 1)/0.05 = -7.6$, which also rejects.

An economist cares because $\beta_1$ is the marginal propensity to consume, and
$\beta_1 = 1$ would mean households spend every additional rupee of income. The
test says they do not: expenditure rises by about 62 paise per additional rupee,
and the remainder is saved.

This is worth noticing as a matter of method. The interesting null hypothesis is
not always zero. Testing against a value that theory suggests is often far more
informative.

**(iii)** $0.62 \pm 1.96(0.05) = [0.522,\ 0.718]$. The interval excludes both 0
and 1, so the data are consistent with a marginal propensity to consume of
roughly two-thirds and with neither extreme.
:::

### Concept checks {-}

**7.4** What goes wrong if every observation has the same value of $x$?

::: {.answer}
The slope is undefined. Its denominator is $\sum(x_i - \bar{x})^2$, which is
zero when every $x_i$ equals $\bar{x}$, and the estimate would require dividing
by zero.

The statistical content is more interesting than the algebra. If attendance
never varies, the data contain no information about how GPA responds to
attendance — there is nothing to compare. This is assumption SLR3, and it is the
same point as Section \@ref(u1-randomness): where nothing varies, nothing can be
learned.
:::

**7.5** A regression reports $R^2 = 0.04$ and a slope significant at the 1%
level. Is that a contradiction?

::: {.answer}
No. They measure different things.

$R^2$ says the regression accounts for 4% of the variation in $y$ — the points
scatter widely around the line. The $p$-value says the slope is distinguishable
from zero, which depends on the standard error and therefore on the sample size.

With a large sample, a real but small relationship produces exactly this
pattern: a reliably estimated slope explaining little of the variation. Both
statements are true and neither is more important; they answer "how much does
$x$ account for?" and "is there a relationship at all?"
:::

**7.6** Explain why $\mathrm{Var}(\hat{\beta}_1)$ falls as the spread of $x$
increases, and what this implies for research design.

::: {.answer}
The formula is $\mathrm{Var}(\hat{\beta}_1) = \sigma^2/\sum(x_i - \bar{x})^2$,
so spread in $x$ appears in the denominator and larger spread means a smaller
variance.

The intuition is that a slope is a rate of change, and rates of change are
measured by observing change. If every student attended between 27 and 29
classes, we would be trying to infer the effect of attendance from almost no
variation in attendance, and small errors in $y$ would translate into wild
swings in the fitted slope. Observing attendance from 5 to 30 pins the line down.

For design, this says that variation in the explanatory variable is valuable and
should be sought rather than avoided. A study of the effect of class size that
only observes classes of 39 to 41 students will struggle no matter how many
schools it covers. It also explains why the median split did worse than the
regression: collapsing attendance to two categories discards most of its spread.
:::

### R exercises {-}

**7.7** Using `data/car-mileage.csv`, regress fuel efficiency on weight. Report
the slope with its interval, interpret it in the units of the data, and plot the
fit.

::: {.answer}
```{r sr-ex-cars, fig.height = 3.4, fig.cap = "Fuel efficiency against weight for 74 cars, with the fitted line."}
cars <- read.csv("data/car-mileage.csv")

fit_cars <- lm(mileage_kmpl ~ weight_kg, data = cars)
summary(fit_cars)$coefficients
confint(fit_cars)["weight_kg", ]

ggplot(cars, aes(weight_kg, mileage_kmpl)) +
  geom_point(colour = book_palette$ink, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, colour = book_palette$accent,
              fill = book_palette$fill) +
  labs(x = "Weight (kg)", y = "Fuel efficiency (km per litre)") +
  theme_book()
```

The slope is $-0.00601$ km per litre per kilogram, which is awkward to read at
that scale. Multiply by 100: each additional 100 kg costs about 0.6 km per
litre, with an interval of roughly 0.50 to 0.70.

$R^2$ is 0.65, much higher than the attendance regression, because weight is a
genuinely dominant determinant of fuel consumption in a way that attendance is
not of grades.
:::

**7.8** Verify the algebra. For the car regression, check that the residuals
sum to zero, that they are uncorrelated with weight, and that the line passes
through the point of means.

::: {.answer}
```{r sr-ex-algebra}
u <- residuals(fit_cars)
w <- cars$weight_kg

round(c(sum_u        = sum(u),
        sum_w_u      = sum(w * u),
        cor_w_u      = cor(w, u),
        fitted_at_wbar = coef(fit_cars)[1] + coef(fit_cars)[2] * mean(w),
        mean_mileage   = mean(cars$mileage_kmpl)), 6)
```

All three hold. The first two are zero to within floating-point error, and the
last two agree exactly.

None of this is evidence that the model is correct. These identities are forced
by the two equations OLS solves, and they would hold just as exactly if we
regressed mileage on something absurd.
:::

**7.9** Demonstrate that OLS is unbiased, and that spread in $x$ buys
precision. Simulate from a known model with $\beta_1 = 2$, using $x$ spread
widely in one case and narrowly in the other.

::: {.answer}
```{r sr-ex-unbiased, cache = TRUE}
simulate_slope <- function(x_sd, reps = 3000) {
  b1 <- replicate(reps, {
    x <- rnorm(50, mean = 20, sd = x_sd)
    y <- 5 + 2 * x + rnorm(50, sd = 3)          # truth: beta_1 = 2
    coef(lm(y ~ x))[2]
  })
  c(x_sd = x_sd, mean_b1 = mean(b1), sd_b1 = sd(b1))
}

set.seed(11)
round(t(sapply(c(1, 5), simulate_slope)), 4)
```

Both rows are centred on 2, so OLS is unbiased at either spread — that is what
SLR1–SLR4 deliver, and the spread of $x$ has nothing to do with it.

What the spread changes is precision. With $x$ five times as spread out, the
standard deviation of $\hat{\beta}_1$ falls by roughly a factor of five, exactly
as $\sigma^2/\sum(x_i - \bar{x})^2$ predicts.
:::

**7.10** Show what omitted variable bias looks like. Simulate data where $y$
depends on both $x$ and a variable $z$ that is correlated with $x$, then
estimate a regression that omits $z$.

::: {.answer}
```{r sr-ex-ovb, cache = TRUE}
set.seed(2)

biased <- replicate(3000, {
  z <- rnorm(200)                       # unobserved: motivation, say
  x <- 10 + 2 * z + rnorm(200)          # x is correlated with z
  y <- 5 + 1 * x + 3 * z + rnorm(200)   # truth: the effect of x is 1
  c(omitting_z  = coef(lm(y ~ x))[2],
    including_z = coef(lm(y ~ x + z))[2])
})

round(rowMeans(biased), 4)
```

The true effect of $x$ on $y$ is 1. Omitting $z$ returns about 2.2 — more than
double.

The reason is exactly the term in the unbiasedness derivation. With $z$ left
out, it sits inside the error, and since $z$ is correlated with $x$ the errors
are correlated with $x$ too, so $E(u \mid x) \neq 0$ and the bias does not
vanish. Including $z$ recovers 1.

Note that the biased regression is not badly behaved in any visible way. Its
residuals still sum to zero, its standard errors are still computed correctly,
and its $p$-value is still tiny. It is confidently wrong, and nothing in the
output says so.
:::

## Question bank {#ols-question-bank}

::: {.watchout}
**No answers are given for this section, by design.**
:::

### Multiple choice {-}

**QB 7.1** The OLS slope estimator equals

a. the correlation between $x$ and $y$
b. the sample covariance of $x$ and $y$ divided by the sample variance of $y$
c. the sample covariance of $x$ and $y$ divided by the sample variance of $x$
d. the ratio of the two sample means

**QB 7.2** Which of the following hold by construction in every OLS regression?
*(select all that apply)*

a. $\sum \hat{u}_i = 0$
b. $\sum x_i \hat{u}_i = 0$
c. $E(u \mid x) = 0$
d. The line passes through $(\bar{x}, \bar{y})$

**QB 7.3** In a simple regression, $\hat{\sigma}^2$ divides the sum of squared
residuals by

a. $n$
b. $n - 1$
c. $n - 2$
d. $n - 3$

**QB 7.4** $\mathrm{Var}(\hat{\beta}_1)$ decreases when

a. $\sigma^2$ increases
b. the spread of $x$ increases
c. the spread of $y$ increases
d. $R^2$ decreases

**QB 7.5** A regression of wages on schooling gives $R^2 = 0.11$. This means

a. the model is misspecified
b. schooling accounts for 11% of the variation in wages
c. 11% of the coefficient is reliable
d. schooling causes 11% of wage differences

**QB 7.6** Omitted variable bias arises when the omitted variable

a. is correlated with $y$ only
b. is correlated with $x$ only
c. is correlated with both $x$ and $y$
d. has a large variance

### Short reasoning {-}

**QB 7.7** Explain why $\sum x_i\hat{u}_i = 0$ is not evidence that
$E(u \mid x) = 0$.

**QB 7.8** A student reports that their regression "explains 95% of the
variation, so the model must be correct." Evaluate.

**QB 7.9** Why does the divisor in $\hat{\sigma}^2$ change from $n-1$ in Unit
\@ref(interval-estimation) to $n-2$ here?

**QB 7.10** In the fertiliser example the slope was 1.25 quintals per kilogram.
Explain what would happen to that number, and to its standard error, if
fertiliser were measured in grams instead.

**QB 7.11** Explain why a regression can have a correctly computed standard
error, a valid $p$-value, and a completely misleading interpretation.

**QB 7.12** For six observations, $\sum(x_i - \bar{x})(y_i - \bar{y}) = 84$,
$\sum(x_i - \bar{x})^2 = 56$, $\bar{x} = 9$, $\bar{y} = 21$ and
$\sum \hat{u}_i^2 = 12$.

i. Compute $\hat{\beta}_1$ and $\hat{\beta}_0$.
ii. Compute $\hat{\sigma}^2$ and $\mathrm{se}(\hat{\beta}_1)$.
iii. Test $H_0: \beta_1 = 0$ at the 5% level.
iv. Construct a 95% confidence interval for $\beta_1$.

### Numerical {-}

**QB 7.13** A regression on 30 observations gives $\hat{\beta}_1 = 4.8$,
$\mathrm{se}(\hat{\beta}_1) = 1.5$, $\text{SST} = 900$ and $\text{SSR} = 400$.

i. Compute $R^2$.
ii. Test $H_0: \beta_1 = 0$ at the 1% level.
iii. Test $H_0: \beta_1 = 8$ at the 5% level.
iv. Compute $\hat{\sigma}^2$.

**QB 7.14** Yields and rainfall for five districts:

| Rainfall (cm) | 60 | 80 | 100 | 120 | 140 |
|--:|--:|--:|--:|--:|--:|
| Yield (quintals) | 18 | 23 | 24 | 30 | 35 |

i. Compute the OLS estimates.
ii. Compute all five residuals and verify they sum to zero.
iii. Compute $R^2$.
iv. Predict yield at 110 cm, and say why predicting at 300 cm would be unwise.

**QB 7.15** Using `data/india-production.csv`, write R code to regress log
output on log labour. Report the slope with its confidence interval, interpret
it as an elasticity, and state what assumption would have to hold for it to
represent the effect of labour on output.

## Looking ahead {#ols-looking-ahead}

We can now fit a line, say how precisely we have fitted it, and test claims about
its slope. What we cannot do is separate the effect of $x$ from the effects of
everything correlated with $x$.

The unbiasedness derivation showed exactly where the problem sits. We had

$$\hat{\beta}_1 = \beta_1 + \frac{\sum(x_i - \bar{x})u_i}{\text{SST}_x}$$

and if the unobserved factors move with $x$, the second term does not average
away. Motivated students attend more classes *and* score higher for reasons of
their own, so $\hat{\beta}_1$ carries both influences and reports their sum.

The remedy is to stop leaving those factors unobserved. If motivation, prior
preparation and teaching quality can be measured, they can be included — and the
question becomes how GPA varies with attendance *among students who are alike in
those respects*.

That requires more than one explanatory variable, and it is the subject of the
next unit. Remarkably little new machinery is needed: the same least squares
criterion, the same normal equations, the same $t$ statistics. What changes is
the interpretation of a coefficient, which becomes both more subtle and far more
useful.
