# Multiple Regression: Holding Other Things Equal {#multiple-regression}

```{r setup-08, include = FALSE}
source("_common.R")
library(ggplot2)
```

Unit \@ref(ols) ended with two students. One attended thirty classes, the other
twenty, and the first earned the higher grade. The difficulty was that the two
students differed in many other ways as well. One may have been better prepared,
more motivated, more interested in the subject, or simply worked harder outside
class. The regression slope faithfully described the difference in their average
grades, but it could not tell us how much of that difference was due to
attendance itself.

A natural response is to measure some of those other differences and compare
students who are alike in every observed respect except attendance.

That is the idea behind multiple regression.

Conceptually, very little changes. We still model the conditional mean, fit the
model by least squares, and measure uncertainty with standard errors, confidence
intervals and hypothesis tests. The machinery is familiar.

What changes is the interpretation.

In a simple regression, the slope describes how average outcomes change as one
variable changes. In a multiple regression, each coefficient describes how the
outcome changes as one variable changes while the others are held fixed. That
seemingly small change is what makes multiple regression one of the most useful
tools in empirical economics.

It is also where many misunderstandings begin.

What does it really mean to "hold other things equal"? How does regression
compare observations that are never exactly identical? When does controlling for
additional variables reduce bias, and when can it make matters worse? Most
importantly, does adding more variables bring us any closer to the causal
comparison we wanted in Unit \@ref(ols)?

Those are the questions this unit takes up.

## Why one variable is not enough {#mlr-motivation}

::: {.question}
The simple regression of grades on attendance was estimated precisely. What is
missing?
:::

Here is the regression Unit \@ref(ols) ended with.

```{r mlr-simple}
students <- read.csv("data/attendance-grades.csv")

simple <- lm(sem_gpa ~ attendance, data = students)
round(coef(summary(simple)), 4)
```

A student who attends one more class has a semester GPA higher by 0.189 points
on average. The standard error is small, the $t$ statistic is 17.6, and the
$p$-value is beyond the range at which anyone bothers to read further.

Nothing about this output is wrong. The arithmetic is correct, the assumptions
of Unit \@ref(ols) may well hold well enough for the standard error to be
trusted, and the relationship is unmistakably there.

The problem is elsewhere. The number 0.189 answers a question we did not ask.

### What the slope is comparing {-}

The slope compares students who attended more classes with students who attended
fewer. Those two groups differ in attendance. They also differ in everything else
that comes bundled with attendance.

Our data record two such things. Every student has a **prior cumulative GPA**,
measuring how they performed before this semester began, and an **admission
score** from the entrance examination.

```{r mlr-bundled}
round(cor(students[, c("attendance", "sem_gpa", "cum_gpa", "admission_score")]), 3)
```

Students who attend more also arrive with higher prior GPAs — the correlation is
0.43 — and prior GPA is itself strongly related to this semester's grade.

So when we compare a student who attended thirty classes with one who attended
twenty, we are not comparing two otherwise identical students. On average, the
first also had a better academic record before the semester started. The
difference in their grades reflects both.

::: {.watchout}
In a simple regression, every variable we fail to measure remains in $u$, and
every one of them that moves with $x$ contributes to the slope.

The estimate is not measuring attendance. It is measuring attendance *together
with whatever travels alongside it*.
:::

### Asking a narrower question {-}

The question worth asking is narrower than the one the simple regression
answers.

> Among students who arrived with the *same* prior GPA and the *same* admission
> score, how much higher is the semester grade of one who attends an extra class?

This restricts the comparison. Instead of comparing all students to all students,
it compares students who are alike in prior record and differ in attendance.

Multiple regression is the tool that makes that comparison, and the rest of this
unit is about what it does and what it cannot do.

## The model and what a coefficient means {#mlr-model}

::: {.question}
We add a second explanatory variable. What does $\beta_1$ now measure?
:::

The population model with two explanatory variables is

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + u$$

and with $k$ of them,

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_k x_k + u$$

Everything from Unit \@ref(simple-regression) carries over. There is still a
population regression function that we do not observe, still an error term
collecting everything left out, and still a sample from which we estimate.

Take the conditional expectation of both sides. Under the zero conditional mean
assumption, which Section \@ref(mlr-assumptions) restates,

$$E[y \mid x_1, x_2] = \beta_0 + \beta_1 x_1 + \beta_2 x_2$$

The population regression function is now a **plane** rather than a line: for
every pair $(x_1, x_2)$ it returns the average value of $y$ among all members of
the population with those characteristics.

### The partial derivative {-}

Differentiate that expression with respect to each explanatory variable in turn.

$$\frac{\partial E[y \mid x_1, x_2]}{\partial x_1} = \beta_1
\qquad\qquad
\frac{\partial E[y \mid x_1, x_2]}{\partial x_2} = \beta_2$$

A partial derivative changes one argument and leaves the others where they are.
That is the whole of the interpretation.

::: {.keyidea}
$\beta_1$ is the change in the average value of $y$ associated with a one-unit
increase in $x_1$, **holding $x_2$ fixed**.

This is the *ceteris paribus* interpretation, and it is the reason multiple
regression exists. In a simple regression, $x_2$ was free to move whenever $x_1$
moved, and the slope absorbed the consequence.
:::

### Doing it {-}

In R, additional explanatory variables are added to the formula with `+`.

```{r mlr-fit}
full <- lm(sem_gpa ~ attendance + cum_gpa + admission_score, data = students)
round(coef(summary(full)), 4)
```

Read the second row. Among students with the same prior cumulative GPA and the
same admission score, one extra class attended is associated with a semester GPA
higher by **0.136** points.

Set that beside the simple regression.

```{r mlr-compare}
round(c(simple   = coef(simple)["attendance"],
        multiple = coef(full)["attendance"]), 4)
```

The coefficient has fallen from 0.189 to 0.136, a drop of about 28%.

::: {.intuition}
Roughly a quarter of what the simple regression attributed to attendance was
never attendance at all. It was the prior academic record that students bring
with them, arriving in the slope because it moves together with how often they
turn up.

Comparing students who are alike in prior record removes that part of the
difference. What is left is a narrower and more honest comparison.
:::

The other two coefficients read the same way. Among students who attended the
same number of classes and scored the same at admission, one point of prior
cumulative GPA is associated with 0.575 points of semester GPA. Prior record
matters, and it matters a great deal.

### A second example {-}

The same reading applies to earnings. This is a synthetic dataset of 20,000
Indian wage earners, built to resemble the structure of household survey data.

```{r mlr-wages}
wages <- read.csv("data/wages-india-synthetic.csv")

wages$male  <- as.integer(wages$sex == "Male")
wages$urban <- as.integer(wages$residence == "Urban")

w_simple <- lm(annual_earnings ~ education, data = wages)
w_full   <- lm(annual_earnings ~ education + age + male + urban, data = wages)

round(coef(summary(w_full)), 1)
```

The variables `male` and `urban` take only the values 0 and 1. A coefficient on
such a variable is a difference between two groups rather than a rate of change —
men earn ₹32,000 more per year than women who are otherwise comparable in
education, age and residence. Unit 9 treats these variables properly; for now,
read them as group differences.

```{r mlr-wages-compare}
round(c(simple   = coef(w_simple)["education"],
        multiple = coef(w_full)["education"]), 1)
```

An extra year of schooling is associated with ₹5,386 of additional annual
earnings when nothing else is accounted for, and ₹4,281 when age, sex and
residence are held fixed. About a fifth of the apparent return to education was
the fact that the educated in this population are disproportionately urban and
male, and urban and male workers earn more for reasons that have nothing to do
with their schooling.

::: {.watchout}
"Holding fixed" describes what the estimator does with the data. It does not
describe an experiment anyone performed.

Nobody took a rural woman and made her urban and male while keeping her
education constant. The regression compares people who already differ in
education among those who happen to be alike in the other measured
characteristics. Whether that comparison answers a causal question is the subject
of Section \@ref(mlr-causal).
:::

## What "holding constant" actually does {#mlr-fwl}

::: {.question}
The interpretation says "holding $x_2$ fixed". But every observation has its own
$x_2$, and nothing was held anywhere. What is the estimator really doing?
:::

This is the question students most often leave a course without having answered.
The phrase "holding constant" sounds like a description of an experiment, and it
is not. It is a description of an algebraic operation, and the operation can be
carried out in front of us.

The answer is a result usually called the **Frisch–Waugh–Lovell theorem**, and
it is easier to state as a recipe than as a theorem.

::: {.keyidea}
To obtain $\hat{\beta}_1$ from the multiple regression of $y$ on $x_1$ and $x_2$:

1. Regress $x_1$ on $x_2$, and keep the residuals $\hat{r}_1$.
2. Regress $y$ on $\hat{r}_1$ alone.

The slope from that second regression is exactly $\hat{\beta}_1$.
:::

The first step is where the work happens. Regressing $x_1$ on $x_2$ splits $x_1$
into two parts: the part that can be predicted from $x_2$, and the part that
cannot. The residual $\hat{r}_1$ is the second part — what remains of $x_1$ once
everything $x_2$ can account for has been removed.

By construction that residual is uncorrelated with $x_2$. Using it in place of
$x_1$ means the variation we are exploiting no longer moves with $x_2$ at all.

### Watching it happen {-}

Take attendance, and remove from it everything that prior GPA and admission score
can explain.

```{r mlr-fwl-step1}
step1 <- lm(attendance ~ cum_gpa + admission_score, data = students)
r1    <- residuals(step1)

round(c(r2_of_step1      = summary(step1)$r.squared,
        cor_with_cumgpa  = cor(r1, students$cum_gpa),
        cor_with_admscr  = cor(r1, students$admission_score)), 4)
```

Prior GPA and admission score together account for 29% of the variation in
attendance. The residual carries the remaining 71%, and it is uncorrelated with
both — to machine precision, because that is what least squares guarantees.

Now regress semester GPA on that residual alone.

```{r mlr-fwl-step2}
step2 <- lm(students$sem_gpa ~ r1)

noquote(format(c(from_two_steps   = unname(coef(step2)[2]),
                 from_lm_directly = unname(coef(full)["attendance"])),
               digits = 10))
```

The same number, to ten significant figures.

::: {.intuition}
This is what "holding prior record constant" means, stated concretely.

Two students with the same prior GPA and the same admission score would be
predicted to attend the same number of classes. If one of them attends more than
that prediction and the other less, the difference between them is unrelated to
prior record — it is attendance and nothing else.

The multiple regression coefficient uses only those departures. It discards the
part of attendance that could have been anticipated from the other variables, and
asks what the leftover part is worth.
:::

### Two consequences worth noticing {-}

**The coefficient changes because the variation changes.** The simple regression
used all the variation in attendance. The multiple regression uses only the part
orthogonal to the controls. Those are different questions asked of different
variation, and there is no reason for them to give the same answer.

**Some of the variation is thrown away.** The residual $\hat{r}_1$ has less
spread than attendance itself. Section \@ref(ols-variance) established that the
precision of a slope depends on the spread of the explanatory variable, so
discarding variation must cost something. Section \@ref(mlr-assumptions) returns
to the bill.

::: {.deeper}
**The general case, in matrix form.**

With $k$ explanatory variables, writing out $k+1$ normal equations and solving
them by hand is unpleasant. Matrix notation does the same work in three lines.

Collect the observations into an $n \times 1$ vector $Y$ and an $n \times (k+1)$
matrix $X$ whose first column is all ones and whose remaining columns are the
explanatory variables. The model is

$$Y = X\beta + u$$

The sum of squared residuals is $(Y - X\beta)'(Y - X\beta)$. Differentiating with
respect to $\beta$ and setting the result to zero,

$$-2X'Y + 2X'X\beta = 0
\qquad\Longrightarrow\qquad
X'X\hat{\beta} = X'Y$$

which is the whole system of normal equations in one line. Provided $X'X$ can be
inverted,

$$\hat{\beta} = (X'X)^{-1}X'Y$$

The variance follows the same way. Substituting $Y = X\beta + u$ gives
$\hat{\beta} = \beta + (X'X)^{-1}X'u$, and under homoskedasticity,

$$\mathrm{Var}(\hat{\beta}) = \sigma^2 (X'X)^{-1}$$

We can check the formula directly.

```{r mlr-matrix}
X <- cbind(1, students$attendance, students$cum_gpa, students$admission_score)
y <- students$sem_gpa

b_hat <- solve(t(X) %*% X) %*% t(X) %*% y
round(drop(b_hat), 5)
```

The same four numbers `lm()` reported. Note where the requirement of Section
\@ref(mlr-assumptions) appears: if any column of $X$ is an exact linear
combination of the others, $X'X$ has no inverse and there is no $\hat{\beta}$ to
compute.
:::

## The assumptions {#mlr-assumptions}

::: {.question}
Unit \@ref(ols) listed six assumptions. Which of them change?
:::

Almost none of them. The classical linear regression model of Section
\@ref(ols-unbiased) carries over with the explanatory variable replaced by the
whole set of explanatory variables, and one genuinely new requirement added.

::: {.theory}
**CLRM1 — linear in parameters.**
$y = \beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k + u$.

**CLRM2 — random sampling.** The $n$ observations are a random sample from that
population.

**CLRM3 — variation in the explanatory variables.** No $x_j$ takes the same value
for every observation.

**CLRM4 — zero conditional mean.** $E(u \mid x_1, \ldots, x_k) = 0$.

**CLRM5 — homoskedasticity.**
$\mathrm{Var}(u \mid x_1, \ldots, x_k) = \sigma^2$.

**CLRM6 — normality.** The errors are normally distributed. As before, this is
needed only for exact small-sample inference.

**CLRM7 — no perfect collinearity.** No explanatory variable is an exact linear
combination of the others.
:::

Under CLRM1–CLRM4 the OLS estimators are unbiased: $E(\hat{\beta}_j) = \beta_j$
for every $j$. The proof is the one from Section \@ref(ols-unbiased) written in
matrix form, and nothing about it is new.

Two of these deserve more than a restatement.

### Zero conditional mean, now harder and easier at once {-}

CLRM4 has become **harder** to state, because it must hold conditional on all of
the explanatory variables at once. And it has become **easier** to satisfy,
because $u$ now contains less. Prior GPA and admission score have moved out of the
error term and into the model, so they can no longer cause it to fail.

::: {.keyidea}
This is the precise sense in which multiple regression is progress.

It does not make the zero conditional mean assumption true. It moves specific,
nameable threats out of $u$ and into the regression, where they are controlled
rather than assumed away.

Whatever remains in $u$ — motivation, ability, family circumstances — is exactly
as dangerous as it was.
:::

### No perfect collinearity {-}

The new assumption rules out an exact linear relationship among the explanatory
variables. It is not a claim about the world but a requirement for the arithmetic
to have an answer at all.

Our data contain a violation, and it is instructive.

```{r mlr-collinear-show}
head(students$attendance + students$classes_missed)
```

The course had 32 classes, so attendance and classes missed always sum to 32.
Knowing one is knowing the other. Ask R to use both.

```{r mlr-collinear-fit}
coef(lm(sem_gpa ~ attendance + classes_missed, data = students))
```

R returns `NA` for the second coefficient. It has not failed; it is reporting
that the question has no answer.

::: {.intuition}
The coefficient on attendance is supposed to answer: what happens to the grade
when attendance rises by one, holding classes missed fixed?

Nothing happens, because nothing *can* happen. Attending one more class means
missing one fewer. There is no student for whom one changed and the other did
not, so the data contain no information about that comparison, and no amount of
cleverness can extract it.
:::

### The cost of near-collinearity {-}

Perfect collinearity is rare and announces itself immediately. The practically
important case is the one just short of it: explanatory variables that are
strongly but not exactly related.

This is where the price of Section \@ref(mlr-fwl) is paid. The variance of a
coefficient in a multiple regression is

$$\mathrm{Var}(\hat{\beta}_j)
= \frac{\sigma^2}{\sum_{i=1}^{n}(x_{ij} - \bar{x}_j)^2 \, (1 - R_j^2)}$$

where $R_j^2$ is the $R^2$ from regressing $x_j$ on all the *other* explanatory
variables. Compare it with the simple regression formula of Section
\@ref(ols-variance): everything is the same except the extra factor $(1-R_j^2)$
in the denominator.

::: {.keyidea}
$R_j^2$ is the share of $x_j$ that the other regressors can already explain — the
share the partialling-out step of Section \@ref(mlr-fwl) removes.

When $R_j^2$ is near zero, the control costs almost nothing. When it approaches
one, $(1-R_j^2)$ approaches zero and the variance explodes.

**Adding a control that is closely related to the variable of interest buys
credibility with precision.**
:::

For attendance, $R_j^2$ was 0.29, so the denominator is multiplied by 0.71 — a
modest price, and one worth paying. The estimate of $\sigma^2$ also changes, since
$k+1$ parameters have now been estimated rather than two:

$$\hat{\sigma}^2 = \frac{\sum_{i=1}^{n}\hat{u}_i^2}{n - k - 1}$$

## Testing one coefficient {#mlr-t}

::: {.question}
Does the machinery of Unit \@ref(ols) still work?
:::

It does, unchanged apart from the degrees of freedom. To test whether a single
population coefficient equals zero,

$$H_0 : \beta_j = 0
\qquad\qquad
t = \frac{\hat{\beta}_j}{\mathrm{se}(\hat{\beta}_j)}
\;\sim\; t_{\,n-k-1}$$

and confidence intervals follow as before.

```{r mlr-t-test}
round(coef(summary(full)), 4)
round(confint(full), 4)
```

Every coefficient is significant at any conventional level, and the interval for
attendance runs from 0.116 to 0.157.

::: {.watchout}
The null tested is $\beta_j = 0$ **in the model as specified**.

Rejecting it says attendance is related to grades among students alike in prior
GPA and admission score. It says nothing about a model with different controls,
where the coefficient, its standard error and its significance may all differ.

A $t$ statistic is a statement about one column of one regression.
:::

## Testing several coefficients at once {#mlr-f}

::: {.question}
We can test one coefficient. Why not just do it repeatedly?
:::

Suppose we want to know whether prior record matters at all — whether prior GPA
and admission score, taken together, belong in the model.

$$H_0 : \beta_2 = 0 \;\text{ and }\; \beta_3 = 0$$

The obvious approach is two $t$-tests. It does not work, and seeing why is more
useful than the test that replaces it.

### Why separate tests fail {-}

Each $t$-test asks about one coefficient while the others stay in the model. The
joint null asks about both coefficients simultaneously. These are different
claims, and they can disagree.

Here is a small constructed example where the disagreement is stark. Two
explanatory variables both genuinely affect $y$, and they are highly correlated
with each other.

```{r mlr-f-why}
set.seed(11)
n  <- 60
z  <- rnorm(n)
x1 <- z + rnorm(n, sd = 0.25)
x2 <- z + rnorm(n, sd = 0.25)
y  <- 1 + 0.8 * x1 + 0.8 * x2 + rnorm(n, sd = 2)

round(cor(x1, x2), 3)
round(coef(summary(lm(y ~ x1 + x2))), 4)
```

Neither variable is close to significant. The $p$-values are 0.20 and 0.30, and a
reader taking each test at face value would drop both.

Now test them together.

```{r mlr-f-why2}
anova(lm(y ~ 1), lm(y ~ x1 + x2))
```

Jointly, they are overwhelmingly significant.

::: {.intuition}
There is no contradiction, and the resolution is Section \@ref(mlr-fwl).

The $t$-test on $x_1$ uses only the part of $x_1$ that is unrelated to $x_2$.
Since the two are correlated at 0.91, almost nothing is left, and that remainder
tells us very little. The same is true of $x_2$.

The data cannot say which of the two is responsible. They are perfectly clear
that one of them is.

Two individually insignificant coefficients therefore do not license dropping
both. What can be said about them separately, and what can be said about them
together, are different things.
:::

### The restricted and unrestricted models {-}

The joint test compares two models.

The **unrestricted** model is the one we believe might be appropriate:

$$y = \beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k + u$$

The **restricted** model is what remains after imposing the null. If the null sets
$q$ coefficients to zero, the restricted model simply omits those $q$ variables.

Imposing a restriction can only make the fit worse: the unrestricted model was
free to choose those coefficients and chose them to minimise the sum of squared
residuals. So $\text{SSR}_r \geq \text{SSR}_{ur}$ always.

::: {.keyidea}
The question is therefore not *whether* the fit got worse, but **whether it got
worse by more than dropping irrelevant variables would**.

If the restricted variables truly have zero coefficients, the deterioration
should be no larger than sampling noise would produce.
:::

### Building the statistic {-}

Two quantities are needed, and both are sums of squared residuals.

The **deterioration** is $\text{SSR}_r - \text{SSR}_{ur}$. Larger values are
evidence against the null. But it must be judged relative to something, for two
reasons: dropping more variables must be expected to hurt more, and the raw size
depends on how noisy the data are.

So divide the deterioration by $q$, the number of restrictions, and divide the
remaining unexplained variation by its degrees of freedom, $n-k-1$. The ratio is

$$F = \frac{\big(\text{SSR}_r - \text{SSR}_{ur}\big)\big/ q}
{\text{SSR}_{ur}\big/(n-k-1)}$$

::: {.keyidea}
Read the two pieces separately.

The numerator is **the variation the restricted variables account for, per
restriction**.

The denominator is **the variation nothing accounts for, per degree of freedom**
— an estimate of $\sigma^2$.

$F$ asks whether the variables we are testing explain more per variable than
noise does. Under the null, they explain nothing systematic, so both pieces
estimate the same thing and the ratio averages about one.
:::

An equivalent form uses $R^2$, which is often more convenient because the sums of
squares depend on the units of $y$ and $R^2$ does not:

$$F = \frac{\big(R^2_{ur} - R^2_{r}\big)\big/q}{\big(1 - R^2_{ur}\big)\big/(n-k-1)}$$

### Why an F distribution {-}

Every test in this book compares a statistic to the distribution it would follow
if the null were true. So what distribution does this ratio follow?

The reasoning is one we have used before. In Section \@ref(u3-variance) a sample
variance built from normal errors, scaled appropriately, followed a chi-squared
distribution with degrees of freedom equal to the number of independent squared
deviations. That is exactly what both parts of $F$ are.

::: {.theory}
Under CLRM1–CLRM6, the numerator and denominator are each a sum of squared
normal deviations, scaled by $\sigma^2$:

$$\frac{\text{SSR}_r - \text{SSR}_{ur}}{\sigma^2} \sim \chi^2_q
\qquad\qquad
\frac{\text{SSR}_{ur}}{\sigma^2} \sim \chi^2_{\,n-k-1}$$

They are independent, and dividing each by its degrees of freedom and taking the
ratio gives, by definition, an **$F$ distribution** with $q$ and $n-k-1$ degrees
of freedom.

The unknown $\sigma^2$ appears in both and cancels — the same cancellation that
made a $t$ statistic computable in Section \@ref(u3-t).
:::

That is the formal answer. The useful one is shorter: **an $F$ distribution is
what the ratio of two independent variance estimates looks like when both are
estimating the same thing.** Under the null they are, which is why $F$ averages
about one and why large values are evidence against it.

The distribution is skewed rather than symmetric, and heavily so when $q$ is
small. Most of its mass sits below one — a ratio can fall no lower than zero but
has no ceiling — which is why the average and the typical value are not the same
number, and why the critical value of 3.12 lies so far out.

Since only large values count as evidence, the test is **always one-tailed on the
right**, whatever the alternative.

We can watch this happen. Generate data in which two variables genuinely have no
effect, compute $F$ thousands of times, and compare the results with the
distribution the theory predicts.

```{r mlr-f-dist, cache = TRUE}
set.seed(3)
xa <- rnorm(80); xb <- rnorm(80); xc <- rnorm(80)

f_stats <- replicate(4000, {
  yy <- 2 + 1.5 * xa + rnorm(80)                 # xb and xc have no effect
  anova(lm(yy ~ xa), lm(yy ~ xa + xb + xc))$F[2]
})

round(rbind(simulated = quantile(f_stats, c(0.50, 0.90, 0.95, 0.99)),
            theory    = qf(c(0.50, 0.90, 0.95, 0.99), df1 = 2, df2 = 76)), 3)
```

```{r mlr-f-dist-plot, echo = FALSE, fig.height = 3.4, cache = TRUE, fig.cap = "Four thousand F statistics computed from data in which the two tested variables have no effect whatever, against the F distribution with 2 and 76 degrees of freedom. The shaded region beyond the 5% critical value is where the test rejects -- and since the null is true here, every rejection in it is a Type I error."}
crit <- qf(0.95, 2, 76)
curve_df <- data.frame(x = seq(0, 8, length.out = 600))
curve_df$y <- df(curve_df$x, 2, 76)

ggplot() +
  geom_histogram(aes(x = f_stats, y = after_stat(density)), bins = 70,
                 fill = book_palette$fill, colour = "white", linewidth = 0.15) +
  geom_area(data = subset(curve_df, x >= crit), aes(x, y),
            fill = book_palette$reject, alpha = 0.5) +
  geom_line(data = curve_df, aes(x, y), colour = book_palette$ink,
            linewidth = 0.8) +
  geom_vline(xintercept = crit, colour = book_palette$reject, linewidth = 0.6) +
  annotate("text", x = crit, y = 0.55, label = "5% critical value",
           hjust = -0.08, size = 3.2, colour = book_palette$reject) +
  coord_cartesian(xlim = c(0, 8)) +
  labs(x = "F statistic", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank())
```

The simulated quantiles match the theoretical ones closely, and the test rejects
a true null about 5% of the time, which is what a 5% test is supposed to do.

### Does prior record belong in the model? {-}

Back to the question that opened this section.

```{r mlr-f-grades}
restricted   <- lm(sem_gpa ~ attendance, data = students)
unrestricted <- lm(sem_gpa ~ attendance + cum_gpa + admission_score,
                   data = students)

anova(restricted, unrestricted)
```

By hand, from the two sums of squared residuals:

```{r mlr-f-byhand}
ssr_r  <- sum(residuals(restricted)^2)
ssr_ur <- sum(residuals(unrestricted)^2)
q <- 2; n <- nrow(students); k <- 3

f_stat <- ((ssr_r - ssr_ur) / q) / (ssr_ur / (n - k - 1))

round(c(ssr_restricted = ssr_r, ssr_unrestricted = ssr_ur,
        F = f_stat, critical_5pct = qf(0.95, q, n - k - 1),
        p_value = 1 - pf(f_stat, q, n - k - 1)), 4)
```

$F = 168.8$ against a critical value of 3.01. Prior GPA and admission score
belong in the model, jointly and emphatically.

### The F statistic R prints by itself {-}

Every regression summary ends with a line reporting an $F$ statistic. It is this
same test applied to one particular null: that **all** the slope coefficients are
zero at once.

```{r mlr-f-overall}
summary(full)$fstatistic
```

The restricted model in that comparison has no explanatory variables at all — it
predicts $\bar{y}$ for everyone — so $R^2_r = 0$ and the formula reduces to

$$F = \frac{R^2/k}{(1-R^2)/(n-k-1)}$$

::: {.watchout}
This is the least interesting test in the output.

It asks whether the model explains anything whatever. With reasonable data and
sensible variables it is almost always rejected, and rejecting it establishes
very little. It is worth noticing chiefly when it *fails* to reject.
:::

### A joint test that does not reject {-}

Not every set of variables earns its place. The data also record which year of
the programme each student is in.

```{r mlr-f-years}
with_years <- lm(sem_gpa ~ attendance + cum_gpa + admission_score + year1 + year2,
                 data = students)

anova(unrestricted, with_years)
```

$F = 1.02$ with a $p$-value of 0.36. Once attendance and prior record are
accounted for, year of study adds nothing detectable.

::: {.watchout}
Failing to reject is not proof that year of study is irrelevant, for the reason
given in Section \@ref(os-errors): a test that does not reject may simply lack the
power to detect a small effect.

The honest statement is that these data provide no evidence that year of study
matters once the other variables are held fixed.
:::

## Does this give us causality? {#mlr-causal}

::: {.question}
We have controlled for prior record. Is 0.136 the effect of attending class?
:::

Not necessarily, and the reason is worth setting out precisely rather than as a
caution.

### Where bias comes from {-}

Suppose the true population model is

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + u$$

but we omit $x_2$ and estimate

$$y = \alpha_0 + \alpha_1 x_1 + e$$

What does $\hat{\alpha}_1$ estimate? Write the relationship between the two
explanatory variables as a regression of its own,

$$x_2 = \delta_0 + \delta_1 x_1 + \eta$$

and substitute into the true model:

::: {.theory}
$$
\begin{aligned}
y &= \beta_0 + \beta_1 x_1 + \beta_2(\delta_0 + \delta_1 x_1 + \eta) + u \\[4pt]
  &= (\beta_0 + \beta_2\delta_0) + (\beta_1 + \beta_2\delta_1)x_1
     + (\beta_2\eta + u)
\end{aligned}
$$

Matching this against $y = \alpha_0 + \alpha_1 x_1 + e$ gives

$$\alpha_1 = \beta_1 + \beta_2\delta_1$$

So the short regression estimates the true coefficient **plus a bias term**:

$$\text{bias} = \beta_2 \delta_1
= \big(\text{effect of } x_2 \text{ on } y\big)
\times \big(\text{relationship between } x_2 \text{ and } x_1\big)$$
:::

This is **omitted variable bias**, and the formula says something immediately
useful. The bias is a product of two things, so it vanishes if *either* is zero.
An omitted variable is harmless unless it both affects $y$ and moves with $x_1$.

::: {.keyidea}
The sign of the bias is the product of the two signs.

| $\beta_2$ | $\delta_1$ | Bias in $\hat{\alpha}_1$ |
|:--|:--|:--|
| $+$ | $+$ | Upward |
| $+$ | $-$ | Downward |
| $-$ | $+$ | Downward |
| $-$ | $-$ | Upward |

This can often be signed from knowledge of the subject even when the variable
cannot be measured, which makes it possible to say which way an estimate is
likely to be wrong.
:::

### Checking the arithmetic {-}

The decomposition is exact in a sample, not merely a statement about
expectations. We can verify it on the grades data.

```{r mlr-ovb-check}
long <- coef(unrestricted)

d_cum <- coef(lm(cum_gpa ~ attendance, data = students))[2]
d_adm <- coef(lm(admission_score ~ attendance, data = students))[2]

round(c(short_regression = coef(restricted)["attendance"],
        long_coefficient = long["attendance"],
        bias_via_cum_gpa = long["cum_gpa"] * d_cum,
        bias_via_adm     = long["admission_score"] * d_adm,
        reassembled      = long["attendance"] +
                           long["cum_gpa"] * d_cum +
                           long["admission_score"] * d_adm), 6)
```

The simple regression coefficient reassembles exactly: 0.136 of attendance, plus
0.062 arriving through prior GPA, less a small amount through admission score.
The gap between 0.189 and 0.136 is not a mystery. It is a quantity we can name
and compute.

::: {.intuition}
Prior GPA biased the simple regression *upward* because both signs were positive:
better students attend more ($\delta_1 > 0$) and better students earn higher
grades ($\beta_2 > 0$).

Admission score worked the other way. Its coefficient is positive, but in these
data students with higher admission scores attend slightly *fewer* classes, so
its contribution is negative and small.
:::

### What the controls did and did not do {-}

The controls removed the bias from prior GPA and admission score. They did
nothing about anything else.

::: {.watchout}
Motivation is not in the model. Neither is interest in the subject, nor health,
nor how far a student lives from campus, nor whether they hold a job.

Every one of these plausibly affects grades, and every one plausibly moves with
attendance. Each contributes its own $\beta_2\delta_1$ to the estimate, and the
formula does not care that we cannot measure it.
:::

So 0.136 is a better estimate than 0.189 in a specific and limited sense: two
named sources of bias have been removed. It is not the causal effect of
attendance, and the regression output cannot tell us how far from it we are.

::: {.keyidea}
Multiple regression controls for what is **measured**. Omitted variable bias is
caused by what is **not**.

Adding controls narrows the gap between the estimate and the causal effect
without ever closing it, and nothing in the output reports how much remains.
:::

### The other three ways CLRM4 fails {-}

Omission is the most common threat, and it is not the only one. The zero
conditional mean assumption also fails when:

- **The functional form is wrong.** If earnings rise with experience and then
  flatten, a model linear in experience is misspecified, and the misspecification
  lands in $u$.
- **Variables are measured with error.** A variable recorded imperfectly is not
  the variable the model refers to, and the difference goes into the error term.
- **$y$ and $x$ are determined together.** Regressing quantity on price estimates
  neither the demand curve nor the supply curve, since the observed pairs are
  where the two meet.

Each has its own remedy, and this book returns to them.

::: {.intuition}
It is worth being clear about what has and has not changed since Unit
\@ref(ols).

Multiple regression is a real advance. It is the reason economists can report
returns to education that are not simply the fact that educated people differ
from uneducated ones in a dozen other ways.

It is also not enough. The comparison it makes is between observations alike in
the variables we happened to measure, and the variables we most want to hold
fixed — ability, motivation, the quality of a school — are usually the ones no
survey records.

Closing that gap requires a different kind of idea, and the second half of this
book is about where such ideas come from.
:::

## Summary {#mlr-summary}

::: {.keyidea}
- The **multiple regression model** is
  $y = \beta_0 + \beta_1x_1 + \cdots + \beta_kx_k + u$. Taking conditional
  expectations, $E[y \mid x_1,\ldots,x_k]$ is the population regression function,
  now a plane rather than a line.
- Each $\beta_j$ is a **partial derivative**: the change in average $y$ for a
  one-unit rise in $x_j$, holding the other explanatory variables fixed. This is
  the *ceteris paribus* interpretation, and it is why multiple regression exists.
- "Holding fixed" is an operation on the data, not an experiment. By the
  **Frisch–Waugh–Lovell** result, $\hat{\beta}_j$ can be obtained by regressing
  $x_j$ on the other regressors, keeping the residuals, and regressing $y$ on
  those residuals alone.
- It follows that the coefficient uses only the variation in $x_j$ that the other
  variables cannot explain. Controls remove variation, and removing variation
  costs precision.
- The assumptions are those of Unit \@ref(ols) with the conditioning extended to
  all regressors, plus **CLRM7, no perfect collinearity**. A violation is not a
  statistical problem but an arithmetic one: $X'X$ cannot be inverted, and R
  returns `NA`.
- $\mathrm{Var}(\hat{\beta}_j) = \sigma^2 / [\sum(x_{ij}-\bar{x}_j)^2(1-R_j^2)]$,
  where $R_j^2$ comes from regressing $x_j$ on the other regressors. The larger
  $R_j^2$, the more the control costs. Also
  $\hat{\sigma}^2 = \sum \hat{u}_i^2/(n-k-1)$.
- Testing a single coefficient is unchanged:
  $t = \hat{\beta}_j/\mathrm{se}(\hat{\beta}_j)$ on $n-k-1$ degrees of freedom.
- Testing several at once requires the **F test**, which compares a restricted
  model against an unrestricted one:
  $F = [(\text{SSR}_r - \text{SSR}_{ur})/q] \big/ [\text{SSR}_{ur}/(n-k-1)]$. It
  is always a right-tailed test.
- Separate $t$-tests are not a substitute. Correlated regressors can be
  individually insignificant and jointly overwhelming, because each $t$-test uses
  only the part of its variable that the others cannot explain.
- $F$ is the ratio of two independent variance estimates, each divided by its
  degrees of freedom. Under the null both estimate $\sigma^2$, so $F$ averages
  about one. Its distribution is right-skewed, so most of the mass lies below
  that average.
- **Omitted variable bias** is $\beta_2\delta_1$ — the effect of the omitted
  variable on $y$, times its relationship with the included one. It vanishes if
  either factor is zero, and its sign is the product of their signs.
- Controlling for measured variables removes their contribution to the bias and
  nothing else. Multiple regression narrows the gap to a causal effect without
  closing it, and reports nothing about what remains.
:::

## Practice problems {#mlr-practice}

::: {.yourturn}
Work each one before opening the solution. The
[question bank](#mlr-question-bank) that follows has no answers.
:::

### Numerical problems {-}

**8.1** A researcher estimates the effect of fertiliser on rice yield across
farms, first alone and then controlling for irrigation:

$$\widehat{\text{yield}} = 12.4 + 0.85\,\text{fert}
\qquad\qquad
\widehat{\text{yield}} = 9.1 + 0.52\,\text{fert} + 6.3\,\text{irrig}$$

where `irrig` is hectares under irrigation.

i. Interpret 0.52 precisely.
ii. Using the omitted variable bias formula, what must be the sign of the
    relationship between fertiliser use and irrigation?
iii. Compute the implied slope from a regression of irrigation on fertiliser.

::: {.answer}
**(i)** Among farms with the same irrigated area, one additional unit of
fertiliser is associated with 0.52 more quintals of yield on average.

**(ii)** The short regression gives 0.85 and the long one 0.52, so the bias is
$0.85 - 0.52 = +0.33$. Since $\beta_2 = 6.3 > 0$, the bias formula
$\beta_2\delta_1 = 0.33$ requires $\delta_1 > 0$. Farms using more fertiliser
also have more irrigation — plausible, since both are investments made by
better-resourced farms.

**(iii)** $\delta_1 = 0.33 / 6.3 = 0.052$ hectares of irrigation per unit of
fertiliser.
:::

**8.2** A regression of household consumption on income and household size gives
$R^2 = 0.68$ with $n = 240$.

i. State the null hypothesis of the overall $F$ test and compute the statistic.
ii. The 5% critical value with 2 and 237 degrees of freedom is 3.03. What do you
    conclude?

::: {.answer}
**(i)** $H_0: \beta_1 = \beta_2 = 0$; both slopes are zero. With $k = 2$ and
$n - k - 1 = 237$,

$$F = \frac{0.68/2}{(1-0.68)/237} = \frac{0.34}{0.00135} = 251.8$$

**(ii)** $251.8 \gg 3.03$, so we reject decisively. Income and household size
together explain a substantial share of variation in consumption. As Section
\@ref(mlr-f) notes, this is a low bar and rejecting it establishes little beyond
the model not being empty.
:::

**8.3** An unrestricted model with 5 explanatory variables and $n = 90$ has
$\text{SSR}_{ur} = 148$. Dropping three of them gives $\text{SSR}_r = 176$.

i. Compute the $F$ statistic.
ii. With 3 and 84 degrees of freedom the 5% critical value is 2.71. Conclude.
iii. Why must $\text{SSR}_r$ exceed $\text{SSR}_{ur}$?

::: {.answer}
**(i)** $q = 3$, $n - k - 1 = 90 - 5 - 1 = 84$.

$$F = \frac{(176 - 148)/3}{148/84} = \frac{9.333}{1.762} = 5.30$$

**(ii)** $5.30 > 2.71$: reject. At least one of the three dropped variables has a
non-zero coefficient.

**(iii)** The unrestricted model can always reproduce the restricted one by
setting those three coefficients to zero, and least squares chooses the values
that minimise SSR. So the unrestricted SSR can never be larger.
:::

**8.4** In a wage regression, $R_j^2$ from regressing years of education on the
other explanatory variables is 0.10 in one specification and 0.90 in another.
Holding $\sigma^2$ and $\sum(x_{ij}-\bar{x}_j)^2$ fixed, by what factor does the
standard error of the education coefficient differ between the two?

::: {.answer}
The variance is proportional to $1/(1-R_j^2)$, so the standard error is
proportional to $1/\sqrt{1-R_j^2}$.

$$\frac{\sqrt{1/(1-0.90)}}{\sqrt{1/(1-0.10)}} = \sqrt{\frac{0.90}{0.10}} = 3$$

The standard error is three times larger in the second specification. The point
estimate may be no better; the reported uncertainty is three times worse.
:::

### Concept checks {-}

**8.5** A student says: "The coefficient on attendance fell from 0.189 to 0.136
when I added controls, so the controls made the estimate worse." Respond.

::: {.answer}
The two numbers answer different questions, so neither is a corrupted version of
the other. 0.189 compares all students to all students; 0.136 compares students
alike in prior record.

If the goal is to describe how grades vary with attendance in the population,
0.189 is the correct answer to that question. If the goal is to isolate what
attendance contributes, 0.136 is closer, because one identified source of
contamination has been removed.

"Worse" presumes a target. The change is evidence that prior record was
contributing to the simple slope — which is exactly what we suspected.
:::

**8.6** Explain why a variable that is strongly correlated with $y$ can be
insignificant in a multiple regression.

::: {.answer}
By Section \@ref(mlr-fwl), the coefficient uses only the part of the variable that
the other regressors cannot explain. If the variable is closely related to the
other regressors, that remaining part is small, and a small amount of variation
supports only an imprecise estimate.

The variable may be strongly related to $y$ while contributing nothing the other
variables have not already contributed. That is a statement about what the
variable adds, not about whether it matters.
:::

**8.7** A regression includes both `attendance` and `classes_missed`, and R
reports `NA` for the second. Explain what has happened and what should be done.

::: {.answer}
The two sum to 32 for every student, so one is an exact linear function of the
other and CLRM7 fails. There is no observation for which attendance changes while
classes missed does not, so the *ceteris paribus* comparison the coefficient
refers to does not exist in the data. $X'X$ is singular and cannot be inverted.

R drops one variable and reports `NA` rather than failing. The remedy is to
include only one of them: they carry identical information.
:::

**8.8** Two variables are individually insignificant at 5% but jointly
significant at 1%. What should be concluded, and what should not?

::: {.answer}
Conclude that at least one of the two has a non-zero coefficient, and that the
data cannot say which — a signature of strongly correlated regressors.

Do not conclude that both are irrelevant, and do not drop both on the strength of
the two $t$-tests. Do not conclude that both matter either: the joint test is
silent about which.
:::

### R exercises {-}

**8.9** Using `attendance-grades.csv`, verify the Frisch–Waugh–Lovell result for
`cum_gpa` rather than `attendance`: regress `cum_gpa` on the other two
regressors, keep the residuals, regress `sem_gpa` on them, and compare the slope
with the multiple regression coefficient.

::: {.answer}
```{r mlr-ex-fwl, eval = FALSE}
students <- read.csv("data/attendance-grades.csv")
full <- lm(sem_gpa ~ attendance + cum_gpa + admission_score, data = students)

r <- residuals(lm(cum_gpa ~ attendance + admission_score, data = students))

c(two_step = coef(lm(students$sem_gpa ~ r))[2],
  direct   = coef(full)["cum_gpa"])
```

The two agree to machine precision. The result holds for every coefficient, not
just the one of interest — each is the slope from a regression on the part of its
own variable that the others cannot explain.
:::

**8.10** Using `wages-india-synthetic.csv`, test whether sex and residence
jointly belong in a regression of annual earnings on education and age. Report
the $F$ statistic, its degrees of freedom and your conclusion.

::: {.answer}
```{r mlr-ex-f, eval = FALSE}
wages <- read.csv("data/wages-india-synthetic.csv")
wages$male  <- as.integer(wages$sex == "Male")
wages$urban <- as.integer(wages$residence == "Urban")

r  <- lm(annual_earnings ~ education + age, data = wages)
ur <- lm(annual_earnings ~ education + age + male + urban, data = wages)

anova(r, ur)
```

$F = 900.9$ on 2 and 19,995 degrees of freedom, with a $p$-value below any
printable threshold. Sex and residence belong in the model.

That they belong does not make the model causal. It says earnings differ
systematically by sex and residence among people of the same education and age —
a fact about the labour market, not an explanation of it.
:::

**8.11** Show by simulation that omitted variable bias behaves as the formula
predicts. Generate $x_2$ correlated with $x_1$, construct
$y = 1 + 2x_1 + 3x_2 + u$, estimate the short regression 2,000 times, and compare
the average $\hat{\alpha}_1$ with $\beta_1 + \beta_2\delta_1$.

::: {.answer}
```{r mlr-ex-ovb, eval = FALSE}
set.seed(20)
n <- 200

results <- replicate(2000, {
  x1 <- rnorm(n)
  x2 <- 0.5 * x1 + rnorm(n)              # delta_1 = 0.5 by construction
  y  <- 1 + 2 * x1 + 3 * x2 + rnorm(n, sd = 2)
  c(short = coef(lm(y ~ x1))[2], long = coef(lm(y ~ x1 + x2))[2])
})

c(mean_short = mean(results["short.x1", ]),
  predicted  = 2 + 3 * 0.5,
  mean_long  = mean(results["long.x1", ]))
```

The short regression averages about 3.5, matching
$\beta_1 + \beta_2\delta_1 = 2 + 3(0.5)$. The long regression averages about 2,
the true value.

The bias does not shrink with sample size. Increasing $n$ to 20,000 leaves it
where it is — bias is not sampling error, and no amount of data removes it.
:::

**8.12** Investigate the cost of a redundant control. Regress `sem_gpa` on
`attendance` alone, then add `cum_gpa`, then add a variable constructed to be
almost identical to `attendance`. Compare the standard errors and explain.

::: {.answer}
```{r mlr-ex-collin, eval = FALSE}
students <- read.csv("data/attendance-grades.csv")
set.seed(5)
students$attendance_copy <- students$attendance + rnorm(nrow(students), sd = 0.4)

se <- function(m) round(coef(summary(m))["attendance", 2], 4)

c(alone      = se(lm(sem_gpa ~ attendance, data = students)),
  plus_cum   = se(lm(sem_gpa ~ attendance + cum_gpa, data = students)),
  plus_copy  = se(lm(sem_gpa ~ attendance + attendance_copy, data = students)))
```

Adding `cum_gpa` barely moves the standard error and removes a real source of
bias. Adding the near-copy inflates it enormously while contributing no
information at all: $R_j^2$ for attendance is now close to one, so $(1-R_j^2)$ is
close to zero.

This is the practical content of Section \@ref(mlr-assumptions). Controls are not
free, and a control that duplicates the variable of interest is all cost.
:::

## Question bank {#mlr-question-bank}

### Multiple choice {-}

**QB 8.1** In the model $y = \beta_0 + \beta_1x_1 + \beta_2x_2 + u$, the
coefficient $\beta_1$ measures

a. the total change in $y$ when $x_1$ rises by one unit
b. the change in average $y$ when $x_1$ rises by one unit, holding $x_2$ fixed
c. the correlation between $x_1$ and $y$
d. the share of variation in $y$ explained by $x_1$

**QB 8.2** The degrees of freedom for a $t$-test on a single coefficient in a
regression with $n = 150$ and four explanatory variables is

a. 146 &nbsp;&nbsp; b. 145 &nbsp;&nbsp; c. 148 &nbsp;&nbsp; d. 149

**QB 8.3** Perfect collinearity between two explanatory variables means

a. the OLS estimates are biased
b. the OLS estimates have large standard errors
c. the OLS estimates cannot be computed at all
d. the $R^2$ will be exactly one

**QB 8.4** An omitted variable causes no bias if

a. it is uncorrelated with the dependent variable
b. it is uncorrelated with the included explanatory variables
c. its coefficient is large
d. the sample is large

**QB 8.5** The $F$ test for exclusion restrictions is always

a. two-tailed
b. left-tailed
c. right-tailed
d. tailed according to the alternative hypothesis

### Short reasoning {-}

**QB 8.6** Explain, without using the phrase "holding constant", what the
Frisch–Waugh–Lovell result says a multiple regression coefficient is computed
from.

**QB 8.7** A colleague argues that since adding controls always reduces
precision, one should include as few as possible. Assess this.

**QB 8.8** Why does the numerator of the $F$ statistic divide by $q$ and the
denominator by $n-k-1$? What would go wrong if neither division were performed?

**QB 8.9** In a study of the effect of an irrigation scheme on farm incomes, the
researcher controls for landholding, soil quality and district. Name two
variables likely still omitted, and use the bias formula to argue in which
direction each would push the estimate.

**QB 8.10** A regression of infant mortality on the number of doctors per
thousand people across Indian districts returns a *positive* coefficient. Explain
using the omitted variable bias formula, and suggest a control.

**QB 8.11** The overall $F$ statistic in a regression is significant but no
individual coefficient is. What does this suggest about the explanatory
variables, and what would you do next?

**QB 8.12** Explain why an omitted variable that strongly affects $y$ but is
unrelated to the explanatory variable of interest costs precision but causes no
bias.

### Numerical {-}

**QB 8.13** A researcher estimates $\widehat{\text{wage}} = 4200 + 380\,educ$ and
then $\widehat{\text{wage}} = 3100 + 240\,educ + 95\,exper$. Compute the implied
regression slope of experience on education, and comment on its sign.

**QB 8.14** An unrestricted model with $n = 120$, $k = 6$ has $R^2_{ur} = 0.54$.
Imposing four exclusion restrictions gives $R^2_r = 0.41$. Compute the $F$
statistic and state its degrees of freedom.

**QB 8.15** A regression with $n = 500$ and $k = 3$ reports
$\sum \hat{u}_i^2 = 892$. Compute $\hat{\sigma}^2$ and $\hat{\sigma}$, and state
why the divisor is not $n$.

**QB 8.16** For a regression of yield on fertiliser, irrigation and labour, the
$R^2$ from regressing fertiliser on the other two is 0.76. By what percentage is
the standard error on fertiliser inflated relative to a hypothetical case in
which fertiliser were uncorrelated with the other regressors?

## Looking ahead {#mlr-looking-ahead}

This unit took the assumptions and the variables as given. Both deserve
questioning.

The model has assumed throughout that a one-unit rise in $x$ has the same
consequence wherever it occurs — that the tenth year of schooling is worth what
the second was, and that the effect of experience never flattens. Economics
rarely believes this. The next unit shows how logarithms, squares and interaction
terms let a linear model describe relationships that are not linear at all, at no
cost to any of the machinery developed here.

The unit also treated the choice of controls as settled. It is not. Some
variables belong in a regression, some are harmless, and some make the estimate
worse than leaving them out — a possibility that surprises most readers, since
the instinct is that controlling for more is always safer. Distinguishing good
controls from bad ones is the second theme of the next unit, and it depends on
exactly the bias formula derived here.
