# Multiple Regression: Holding Other Things Equal {#multiple-regression}

```{r setup-08, include = FALSE}
source("_common.R")
library(ggplot2)
```

Unit \@ref(ols) ended with two students. One attended thirty classes, the other
twenty, and the first earned the higher grade. The difficulty was that the two
students differed in many other ways as well. One may have been better prepared,
more motivated, more interested in the subject, or simply worked harder outside
class. The regression slope faithfully described the difference in their average
grades, but it could not tell us how much of that difference was due to
attendance itself.

A natural response is to measure some of those other differences and compare
students who are alike in every observed respect except attendance.

That is the idea behind multiple regression.

Conceptually, very little changes. We still model the conditional mean, fit the
model by least squares, and measure uncertainty with standard errors, confidence
intervals and hypothesis tests. The machinery is familiar.

What changes is the interpretation.

In a simple regression, the slope describes how average outcomes change as one
variable changes. In a multiple regression, each coefficient describes how the
outcome changes as one variable changes while the others are held fixed. That
seemingly small change is what makes multiple regression one of the most useful
tools in empirical economics.

It is also where many misunderstandings begin.

What does it really mean to "hold other things equal"? How does regression
compare observations that are never exactly identical? When does controlling for
additional variables reduce bias, and when can it make matters worse? Most
importantly, does adding more variables bring us any closer to the causal
comparison we wanted in Unit \@ref(ols)?

Those are the questions this unit takes up.

## Why one variable is not enough {#mlr-motivation}

::: {.question}
The simple regression of grades on attendance was estimated precisely. What is
missing?
:::

Unit \@ref(ols) ended with a simple regression of semester GPA on attendance.

```{r mlr-simple}
students <- read.csv("data/attendance-grades.csv")

simple <- lm(sem_gpa ~ attendance, data = students)
round(coef(summary(simple)), 4)
```

The estimate is clear. Students who attend one additional class earn, on
average, a semester GPA that is 0.189 points higher. The standard error is
small, the confidence interval is narrow, and there is overwhelming evidence
that attendance and grades are associated.

The question is whether this is the comparison we wanted to make.

Suppose we compare two students. One attended thirty classes and the other
twenty. Their attendance differs by ten classes, but that is unlikely to be the
only difference between them. The student who attended more may also have
entered the semester better prepared, have stronger study habits, or simply be
more academically able.

Our data contain measures of two such characteristics: each student's cumulative
GPA before the semester began, and their admission score.

```{r mlr-bundled}
round(cor(students[, c("attendance", "sem_gpa",
                       "cum_gpa", "admission_score")]), 3)
```

One number immediately stands out. Attendance and prior GPA have a correlation
of 0.43. Students who attend more classes also tend to have performed better
before this semester even began.

Prior GPA, in turn, is strongly related to semester GPA.

So when the simple regression compares students with different attendance, it is
also comparing students with different academic histories. Part of the estimated
relationship may reflect attendance, and part may simply reflect the fact that
students who were already performing well continue to do so.

This is exactly the problem discussed at the end of Unit \@ref(ols). Variables
that affect the outcome and move with attendance remain in the error term, and
their influence becomes mixed with the coefficient on attendance.

The comparison we really want is more specific.

Instead of asking,

> Do students who attend more classes earn higher grades?

we would like to ask,

> Among students with the same prior GPA and the same admission score, do those
> who attend more classes earn higher grades?

That is a different comparison. We are no longer comparing every student with
every other student. We are comparing students who are similar in their observed
academic background and differ primarily in attendance.

Simple regression assumes that every relevant difference between students is
either unrelated to attendance or hidden inside the error term. Multiple
regression takes the first step towards relaxing that assumption by bringing
some of those differences into the model.

## Holding other things equal {#mlr-model}

::: {.question}
We add a second explanatory variable. What does $\beta_1$ now measure?
:::

The motivation for multiple regression was straightforward. We wanted to compare
students who differed in attendance but were otherwise similar.

To do that we introduce additional explanatory variables.

With two explanatory variables, the population model is

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + u$$

and with $k$ explanatory variables,

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_k x_k + u$$

Nothing fundamental has changed. There is still a population regression function
that we cannot observe, an error term containing everything left out of the
model, and a sample from which we estimate the unknown coefficients.

What has changed is the meaning of a coefficient.

### A different comparison {-}

In the simple regression of Unit \@ref(ols), the coefficient on attendance
compared students who attended different numbers of classes.

Now the comparison is narrower.

The coefficient on attendance compares students who attended different numbers
of classes but who had the same values of the other explanatory variables.

That is the meaning of holding other things equal.

Mathematically,

$$\frac{\partial E(Y \mid X_1, X_2)}{\partial X_1} = \beta_1$$

The notation looks new, but the idea is not. A partial derivative asks how the
conditional mean changes when one explanatory variable changes while the others
remain fixed.

::: {.keyidea}
$\beta_1$ is the change in the average value of $Y$ associated with a one-unit
increase in $X_1$, holding all the other explanatory variables constant.

This *ceteris paribus* interpretation is the entire reason for using multiple
regression.
:::

### An example {-}

Adding explanatory variables in R is as simple as separating them with `+`.

```{r mlr-fit}
full <- lm(sem_gpa ~ attendance + cum_gpa + admission_score,
           data = students)

round(coef(summary(full)), 4)
```

Compare the coefficient on attendance with the simple regression.

```{r mlr-compare}
round(c(simple   = coef(simple)["attendance"],
        multiple = coef(full)["attendance"]), 4)
```

The estimated association falls from 0.189 to 0.136.

The interpretation has changed. In the simple regression, the comparison was
between students who differed in attendance. In the multiple regression, the
comparison is between students who have the same prior GPA and the same
admission score, but who differ in attendance.

Some of the relationship that the simple regression attributed to attendance is
now explained by differences in students' previous academic performance. The
coefficient on attendance therefore becomes smaller.

::: {.intuition}
Notice that nothing has happened to the data. The students are the same. Only
the comparison has changed.
:::

### Reading the other coefficients {-}

Every coefficient is interpreted in exactly the same way.

The coefficient on cumulative GPA says that among students with the same
attendance and the same admission score, a one-point increase in prior GPA is
associated with an increase of about 0.575 points in semester GPA.

The coefficient on admission score compares students with the same attendance
and the same prior GPA.

Each coefficient asks what happens when one explanatory variable changes while
all the others remain fixed.

### Another example {-}

The same interpretation applies outside education. This is a synthetic dataset
of 20,000 Indian wage earners, built to resemble the structure of household
survey data.

```{r mlr-wages}
wages <- read.csv("data/wages-india-synthetic.csv")

wages$male  <- as.integer(wages$sex == "Male")
wages$urban <- as.integer(wages$residence == "Urban")

w_full <- lm(annual_earnings ~ education + age + male + urban,
             data = wages)

round(coef(summary(w_full)), 1)
```

The coefficient on education is interpreted as the association between an
additional year of schooling and annual earnings holding age, sex and place of
residence constant.

The variables `male` and `urban` take only the values 0 and 1, so their
coefficients compare two groups rather than describing the effect of a one-unit
increase. We return to indicator variables in the next unit.

### What "holding fixed" does not mean {-}

The phrase *holding other things fixed* often gives the wrong impression.

Regression does not perform an experiment.

Nobody took the same student, kept their prior GPA unchanged and forced them to
attend one more class. Nor did anyone move a rural worker to an urban area while
leaving everything else unchanged.

::: {.watchout}
The comparison is entirely statistical. The regression compares observations
that already exist and are similar in the variables included in the model.

Whether that comparison is enough to recover a causal effect depends on whether
the important differences between individuals have been measured.
:::

That question returns in Section \@ref(mlr-causal).

## What does "holding constant" actually mean? {#mlr-fwl}

::: {.question}
Nothing in the data was held anywhere. So what does the regression do?
:::

The phrase *holding other variables constant* is used constantly in economics.
It is also one of the least understood.

Nothing in the data was actually held fixed. Every student has a different
attendance, a different prior GPA and a different admission score. Nobody reran
the semester while changing only one variable.

So what does the regression actually do?

### Predicting attendance {-}

The answer can be described as a simple two-step procedure.

1. Predict attendance using the control variables.
2. Keep only what cannot be predicted.
3. Ask whether that remaining variation predicts grades.

Take the first step slowly, because everything rests on it.

Imagine predicting every student's attendance using only their previous GPA and
admission score. Some students attend exactly as much as we would expect. Some
attend much more. Some attend much less.

Those unexpected differences are the residuals.

```{r mlr-fwl-step1}
step1 <- lm(attendance ~ cum_gpa + admission_score, data = students)
r1    <- residuals(step1)

round(c(r2_of_step1      = summary(step1)$r.squared,
        cor_with_cumgpa  = cor(r1, students$cum_gpa),
        cor_with_admscr  = cor(r1, students$admission_score)), 4)
```

Prior GPA and admission score together account for 29% of the variation in
attendance. The residual carries the rest, and it is uncorrelated with both — to
machine precision, because that is what least squares guarantees.

That last point is the one that matters. Whatever these residuals contain, they
contain nothing about prior GPA or admission score.

### Using only what could not be predicted {-}

Now regress semester GPA on that residual alone.

```{r mlr-fwl-step2}
step2 <- lm(students$sem_gpa ~ r1)

noquote(format(c(from_two_steps   = unname(coef(step2)[2]),
                 from_lm_directly = unname(coef(full)["attendance"])),
               digits = 10))
```

The same number, to ten significant figures.

This result is known as the **Frisch–Waugh–Lovell theorem**, and it holds for
every coefficient in every multiple regression, not just this one.

::: {.keyidea}
Multiple regression does not compare students with the same attendance.

It compares students who differ in the part of attendance that cannot already be
explained by the other variables.
:::

```{r mlr-fwl-plot, echo = FALSE, fig.width = 9, fig.height = 3.1, out.width = "100%", cache = TRUE, fig.cap = "The two steps, and what they change. Panel 1 predicts attendance from prior GPA and admission score; the orange line is where a student would sit if the prediction were exact, and the vertical distances from it are the residuals. Panel 2 regresses GPA on those residuals alone. Panel 3 regresses GPA on attendance itself. The horizontal axes in panels 2 and 3 are drawn to the same width, so the two slopes can be compared directly."}
pred_att <- fitted(step1)

lab <- c("1. Predict attendance from the controls",
         "2. GPA against unexpected attendance",
         "3. GPA against all attendance")

dd <- rbind(
  data.frame(x = pred_att,            y = students$attendance, panel = lab[1]),
  data.frame(x = r1,                  y = students$sem_gpa,    panel = lab[2]),
  data.frame(x = students$attendance, y = students$sem_gpa,    panel = lab[3]))
dd$panel <- factor(dd$panel, levels = lab)

## Panels 2 and 3 are given x axes of identical width so that the difference
## between the two slopes is visible rather than merely asserted.
pad <- rbind(
  data.frame(x = c(-16, 16), y = c(0, 10), panel = lab[2]),
  data.frame(x = c(  2, 34), y = c(0, 10), panel = lab[3]))
pad$panel <- factor(pad$panel, levels = lab)

fit_lines <- rbind(
  data.frame(panel = lab[2], sl = coef(step2)[2],
             int = mean(students$sem_gpa) - coef(step2)[2] * mean(r1)),
  data.frame(panel = lab[3], sl = coef(simple)[2], int = coef(simple)[1]))
fit_lines$panel <- factor(fit_lines$panel, levels = lab)

ident <- data.frame(panel = factor(lab[1], levels = lab))

ggplot(dd, aes(x, y)) +
  geom_blank(data = pad) +
  geom_point(alpha = 0.25, colour = book_palette$ink, size = 0.9) +
  geom_abline(data = ident, aes(intercept = 0, slope = 1),
              colour = book_palette$reject, linewidth = 0.7) +
  geom_abline(data = fit_lines, aes(intercept = int, slope = sl),
              colour = book_palette$accent, linewidth = 0.9) +
  facet_wrap(~ panel, scales = "free") +
  labs(x = NULL, y = NULL) +
  theme_book(base_size = 10)
```

::: {.intuition}
Two students with the same prior GPA and admission score are predicted to attend
roughly the same number of classes.

If one attends more than predicted and the other less, that difference is
unrelated to the observed controls.

Multiple regression asks whether those unexpected differences in attendance are
associated with differences in grades.
:::

### Why the coefficient changes {-}

The coefficient changes because the variation changes.

The simple regression used all the variation in attendance — the third panel of
the figure. The multiple regression uses only the part that prior record cannot
account for — the second panel. Those are different questions asked of different
variation, and there is no reason for them to give the same answer.

This is worth separating from the more familiar account, which says that
controls remove bias. They may well do so, and Section \@ref(mlr-causal) takes
that up. But the mechanism is not that something contaminating has been
subtracted from the answer. It is that the estimator is now working with a
different set of comparisons.

### Why the standard error usually grows {-}

The residual $\hat{r}_1$ has less spread than attendance itself: a standard
deviation of 4.6 classes against 5.5.

```{r mlr-fwl-spread}
round(c(sd_of_attendance = sd(students$attendance),
        sd_of_residual   = sd(r1)), 3)
```

Section \@ref(ols-variance) established that the precision of a slope depends on
the spread of the explanatory variable. Discarding variation must therefore cost
something, and Section \@ref(mlr-assumptions) works out the bill.

That is the trade the whole method rests on. A control buys a narrower, more
credible comparison, and pays for it in precision.

## The assumptions of multiple regression {#mlr-assumptions}

::: {.question}
One explanatory variable has become several. Does that require a completely new
set of assumptions?
:::

Surprisingly, very little changes.

Everything that made OLS work in Unit \@ref(ols) still matters. We still need a
model that is linear in the parameters, a random sample, a conditional mean that
is correctly specified, and — if we want the classical standard errors —
constant error variance.

The only genuinely new assumption is one that could not arise before: the
explanatory variables cannot perfectly explain one another.

::: {.theory}
**CLRM1 — linear in parameters.**
$y = \beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k + u$.

**CLRM2 — random sampling.** The $n$ observations are a random sample from that
population.

**CLRM3 — variation in the explanatory variables.** No $x_j$ takes the same value
for every observation.

**CLRM4 — zero conditional mean.** $E(u \mid x_1, \ldots, x_k) = 0$.

**CLRM5 — homoskedasticity.**
$\mathrm{Var}(u \mid x_1, \ldots, x_k) = \sigma^2$.

**CLRM6 — normality.** The errors are normally distributed. As before, this is
needed only for exact small-sample inference.

**CLRM7 — no perfect collinearity.** No explanatory variable is an exact linear
combination of the others.
:::

Under CLRM1–CLRM4, OLS remains unbiased, exactly as before. The proof is
identical to Section \@ref(ols-unbiased), written in matrix notation rather than
scalar notation; Section \@ref(mlr-matrix-note) gives the compact version.

Three assumptions deserve comment.

### Zero conditional mean {-}

The assumption has not changed. The error term must still be unrelated to the
explanatory variables,

$$E(u \mid x_1, \ldots, x_k) = 0$$

What has changed is what remains inside $u$.

In the simple regression, prior GPA and admission score were omitted variables.
Because they were correlated with attendance, they threatened the assumption
directly.

In the multiple regression they are no longer omitted. They have become
explanatory variables.

::: {.keyidea}
That is the sense in which multiple regression improves on simple regression. It
does not make the assumption true. It removes some of the reasons why it might
fail.
:::

Whatever is still omitted — motivation, ability, family background — remains
inside $u$, and can still bias the coefficients if it is related to the
explanatory variables.

### No perfect collinearity {-}

This assumption is not about economics at all. It is a requirement for the
regression to be mathematically identifiable.

Our data contain a violation, and it is instructive.

```{r mlr-collinear-show}
head(students$attendance + students$classes_missed)
```

The course had 32 classes, so attendance and classes missed always sum to 32.
Knowing one is knowing the other. Ask R to use both.

```{r mlr-collinear-fit}
coef(lm(sem_gpa ~ attendance + classes_missed, data = students))
```

R returns `NA` for the second coefficient. It has not failed; it is reporting
that the question has no answer.

::: {.intuition}
The coefficient on attendance is supposed to answer: what happens to the grade
when attendance rises by one, holding classes missed fixed?

Nothing happens, because nothing *can* happen. Attending one more class means
missing one fewer. There is no student for whom one changed and the other did
not, so the data contain no information about that comparison, and no amount of
cleverness can extract it.
:::

### When controls tell us the same thing {-}

Perfect collinearity is rare. The important case is when two explanatory
variables contain much of the same information.

Attendance and prior GPA are not identical. But students with high prior GPAs
also tend to attend more classes. Part of the variation in attendance can
therefore already be predicted from prior GPA.

That is exactly what Section \@ref(mlr-fwl) removed during the residualisation
step.

The more predictable attendance becomes, the less independent variation remains
for estimating its coefficient. Less variation means less information. Less
information means larger standard errors.

The variance formula puts a number on it:

$$\mathrm{Var}(\hat{\beta}_j)
= \frac{\sigma^2}{\sum_{i=1}^{n}(x_{ij} - \bar{x}_j)^2 \, (1 - R_j^2)}$$

Compare it with the simple regression formula of Section \@ref(ols-variance):
everything is the same except the extra factor $(1-R_j^2)$ in the denominator.

::: {.keyidea}
$R_j^2$ is not the $R^2$ from the regression you care about.

It is the $R^2$ from regressing one explanatory variable on all the others. It
tells us how much of that variable is already predictable before the regression
even begins.

If $R_j^2 = 0$, the variable contributes completely new information.

If $R_j^2 = 0.95$, almost everything it contains was already available from the
remaining explanatory variables. OLS is then trying to estimate a coefficient
using only the tiny amount of variation left over.

That is why the standard error grows.
:::

About 29% of attendance can already be predicted from prior GPA and admission
score. The coefficient therefore relies on the remaining 71% — the unexpected
part of attendance. There is still plenty of independent variation left, so the
increase in the standard error is modest.

As before, estimating additional coefficients consumes degrees of freedom, so
the residual variance is now divided by $n-k-1$.

### The trade {-}

Multiple regression always involves a trade-off.

Every additional control makes the comparison more credible by removing another
potential source of bias.

But every additional control also consumes variation, making the coefficient
less precise.

Good empirical work is largely about deciding which controls are worth that
price.

## Testing one coefficient {#mlr-t}

::: {.question}
Does the machinery of Unit \@ref(ols) still work?
:::

It does, unchanged apart from the degrees of freedom. To test whether a single
population coefficient equals zero,

$$H_0 : \beta_j = 0
\qquad\qquad
t = \frac{\hat{\beta}_j}{\mathrm{se}(\hat{\beta}_j)}
\;\sim\; t_{\,n-k-1}$$

and confidence intervals follow as before.

```{r mlr-t-test}
round(coef(summary(full)), 4)
round(confint(full), 4)
```

Every coefficient is significant at any conventional level, and the interval for
attendance runs from 0.116 to 0.157.

::: {.watchout}
The null tested is $\beta_j = 0$ **in the model as specified**.

Rejecting it says attendance is related to grades among students alike in prior
GPA and admission score. It says nothing about a model with different controls,
where the coefficient, its standard error and its significance may all differ.

A $t$ statistic is a statement about one column of one regression.
:::

## Testing several coefficients at once {#mlr-f}

::: {.question}
We can test one coefficient. Why not just do it repeatedly?
:::

Suppose we want to know whether prior record matters at all — whether prior GPA
and admission score, taken together, belong in the model.

$$H_0 : \beta_2 = 0 \;\text{ and }\; \beta_3 = 0$$

The obvious approach is two $t$-tests. It does not work, and seeing why is more
useful than the test that replaces it.

### Why separate tests fail {-}

Each $t$-test asks about one coefficient while the others stay in the model. The
joint null asks about both coefficients simultaneously. These are different
claims, and they can disagree.

Here is a small constructed example where the disagreement is stark. Two
explanatory variables both genuinely affect $y$, and they are highly correlated
with each other.

```{r mlr-f-why}
set.seed(11)
n  <- 60
z  <- rnorm(n)
x1 <- z + rnorm(n, sd = 0.25)
x2 <- z + rnorm(n, sd = 0.25)
y  <- 1 + 0.8 * x1 + 0.8 * x2 + rnorm(n, sd = 2)

round(cor(x1, x2), 3)
round(coef(summary(lm(y ~ x1 + x2))), 4)
```

Neither variable is close to significant. The $p$-values are 0.20 and 0.30, and a
reader taking each test at face value would drop both.

Now test them together.

```{r mlr-f-why2}
anova(lm(y ~ 1), lm(y ~ x1 + x2))
```

Jointly, they are overwhelmingly significant.

::: {.intuition}
There is no contradiction, and the resolution is Section \@ref(mlr-fwl).

The $t$-test on $x_1$ uses only the part of $x_1$ that is unrelated to $x_2$.
Since the two are correlated at 0.91, almost nothing is left, and that remainder
tells us very little. The same is true of $x_2$.

The data cannot say which of the two is responsible. They are perfectly clear
that one of them is.

Two individually insignificant coefficients therefore do not license dropping
both. What can be said about them separately, and what can be said about them
together, are different things.
:::

### The restricted and unrestricted models {-}

The joint test compares two models.

The **unrestricted** model is the one we believe might be appropriate:

$$y = \beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k + u$$

The **restricted** model is what remains after imposing the null. If the null sets
$q$ coefficients to zero, the restricted model simply omits those $q$ variables.

Imposing a restriction can only make the fit worse: the unrestricted model was
free to choose those coefficients and chose them to minimise the sum of squared
residuals. So $\text{SSR}_r \geq \text{SSR}_{ur}$ always.

::: {.keyidea}
The question is therefore not *whether* the fit got worse, but **whether it got
worse by more than dropping irrelevant variables would**.

If the restricted variables truly have zero coefficients, the deterioration
should be no larger than sampling noise would produce.
:::

### Building the statistic {-}

Two quantities are needed, and both are sums of squared residuals.

The **deterioration** is $\text{SSR}_r - \text{SSR}_{ur}$. Larger values are
evidence against the null. But it must be judged relative to something, for two
reasons: dropping more variables must be expected to hurt more, and the raw size
depends on how noisy the data are.

So divide the deterioration by $q$, the number of restrictions, and divide the
remaining unexplained variation by its degrees of freedom, $n-k-1$. The ratio is

$$F = \frac{\big(\text{SSR}_r - \text{SSR}_{ur}\big)\big/ q}
{\text{SSR}_{ur}\big/(n-k-1)}$$

::: {.keyidea}
Read the two pieces separately.

The numerator is **the variation the restricted variables account for, per
restriction**.

The denominator is **the variation nothing accounts for, per degree of freedom**
— an estimate of $\sigma^2$.

$F$ asks whether the variables we are testing explain more per variable than
noise does. Under the null, they explain nothing systematic, so both pieces
estimate the same thing and the ratio averages about one.
:::

An equivalent form uses $R^2$, which is often more convenient because the sums of
squares depend on the units of $y$ and $R^2$ does not:

$$F = \frac{\big(R^2_{ur} - R^2_{r}\big)\big/q}{\big(1 - R^2_{ur}\big)\big/(n-k-1)}$$

### Why an F distribution {-}

Every test in this book compares a statistic to the distribution it would follow
if the null were true. So what distribution does this ratio follow?

The reasoning is one we have used before. In Section \@ref(u3-variance) a sample
variance built from normal errors, scaled appropriately, followed a chi-squared
distribution with degrees of freedom equal to the number of independent squared
deviations. That is exactly what both parts of $F$ are.

::: {.theory}
Under CLRM1–CLRM6, the numerator and denominator are each a sum of squared
normal deviations, scaled by $\sigma^2$:

$$\frac{\text{SSR}_r - \text{SSR}_{ur}}{\sigma^2} \sim \chi^2_q
\qquad\qquad
\frac{\text{SSR}_{ur}}{\sigma^2} \sim \chi^2_{\,n-k-1}$$

They are independent, and dividing each by its degrees of freedom and taking the
ratio gives, by definition, an **$F$ distribution** with $q$ and $n-k-1$ degrees
of freedom.

The unknown $\sigma^2$ appears in both and cancels — the same cancellation that
made a $t$ statistic computable in Section \@ref(u3-t).
:::

That is the formal answer. The useful one is shorter: **an $F$ distribution is
what the ratio of two independent variance estimates looks like when both are
estimating the same thing.** Under the null they are, which is why $F$ averages
about one and why large values are evidence against it.

The distribution is skewed rather than symmetric, and heavily so when $q$ is
small. Most of its mass sits below one — a ratio can fall no lower than zero but
has no ceiling — which is why the average and the typical value are not the same
number, and why the critical value of 3.12 lies so far out.

Since only large values count as evidence, the test is **always one-tailed on the
right**, whatever the alternative.

We can watch this happen. Generate data in which two variables genuinely have no
effect, compute $F$ thousands of times, and compare the results with the
distribution the theory predicts.

```{r mlr-f-dist, cache = TRUE}
set.seed(3)
xa <- rnorm(80); xb <- rnorm(80); xc <- rnorm(80)

f_stats <- replicate(4000, {
  yy <- 2 + 1.5 * xa + rnorm(80)                 # xb and xc have no effect
  anova(lm(yy ~ xa), lm(yy ~ xa + xb + xc))$F[2]
})

round(rbind(simulated = quantile(f_stats, c(0.50, 0.90, 0.95, 0.99)),
            theory    = qf(c(0.50, 0.90, 0.95, 0.99), df1 = 2, df2 = 76)), 3)
```

```{r mlr-f-dist-plot, echo = FALSE, fig.height = 3.4, cache = TRUE, fig.cap = "Four thousand F statistics computed from data in which the two tested variables have no effect whatever, against the F distribution with 2 and 76 degrees of freedom. The shaded region beyond the 5% critical value is where the test rejects -- and since the null is true here, every rejection in it is a Type I error."}
crit <- qf(0.95, 2, 76)
curve_df <- data.frame(x = seq(0, 8, length.out = 600))
curve_df$y <- df(curve_df$x, 2, 76)

ggplot() +
  geom_histogram(aes(x = f_stats, y = after_stat(density)), bins = 70,
                 fill = book_palette$fill, colour = "white", linewidth = 0.15) +
  geom_area(data = subset(curve_df, x >= crit), aes(x, y),
            fill = book_palette$reject, alpha = 0.5) +
  geom_line(data = curve_df, aes(x, y), colour = book_palette$ink,
            linewidth = 0.8) +
  geom_vline(xintercept = crit, colour = book_palette$reject, linewidth = 0.6) +
  annotate("text", x = crit, y = 0.55, label = "5% critical value",
           hjust = -0.08, size = 3.2, colour = book_palette$reject) +
  coord_cartesian(xlim = c(0, 8)) +
  labs(x = "F statistic", y = NULL) +
  theme_book() +
  theme(axis.text.y = element_blank())
```

The simulated quantiles match the theoretical ones closely, and the test rejects
a true null about 5% of the time, which is what a 5% test is supposed to do.

### Does prior record belong in the model? {-}

Back to the question that opened this section.

```{r mlr-f-grades}
restricted   <- lm(sem_gpa ~ attendance, data = students)
unrestricted <- lm(sem_gpa ~ attendance + cum_gpa + admission_score,
                   data = students)

anova(restricted, unrestricted)
```

By hand, from the two sums of squared residuals:

```{r mlr-f-byhand}
ssr_r  <- sum(residuals(restricted)^2)
ssr_ur <- sum(residuals(unrestricted)^2)
q <- 2; n <- nrow(students); k <- 3

f_stat <- ((ssr_r - ssr_ur) / q) / (ssr_ur / (n - k - 1))

round(c(ssr_restricted = ssr_r, ssr_unrestricted = ssr_ur,
        F = f_stat, critical_5pct = qf(0.95, q, n - k - 1),
        p_value = 1 - pf(f_stat, q, n - k - 1)), 4)
```

$F = 168.8$ against a critical value of 3.01. Prior GPA and admission score
belong in the model, jointly and emphatically.

### The F statistic R prints by itself {-}

Every regression summary ends with a line reporting an $F$ statistic. It is this
same test applied to one particular null: that **all** the slope coefficients are
zero at once.

```{r mlr-f-overall}
summary(full)$fstatistic
```

The restricted model in that comparison has no explanatory variables at all — it
predicts $\bar{y}$ for everyone — so $R^2_r = 0$ and the formula reduces to

$$F = \frac{R^2/k}{(1-R^2)/(n-k-1)}$$

::: {.watchout}
This is the least interesting test in the output.

It asks whether the model explains anything whatever. With reasonable data and
sensible variables it is almost always rejected, and rejecting it establishes
very little. It is worth noticing chiefly when it *fails* to reject.
:::

### A joint test that does not reject {-}

Not every set of variables earns its place. The data also record which year of
the programme each student is in.

```{r mlr-f-years}
with_years <- lm(sem_gpa ~ attendance + cum_gpa + admission_score + year1 + year2,
                 data = students)

anova(unrestricted, with_years)
```

$F = 1.02$ with a $p$-value of 0.36. Once attendance and prior record are
accounted for, year of study adds nothing detectable.

::: {.watchout}
Failing to reject is not proof that year of study is irrelevant, for the reason
given in Section \@ref(os-errors): a test that does not reject may simply lack the
power to detect a small effect.

The honest statement is that these data provide no evidence that year of study
matters once the other variables are held fixed.
:::

## Does this give us causality? {#mlr-causal}

::: {.question}
We have controlled for prior record. Is 0.136 the effect of attending class?
:::

Not necessarily, and the reason is worth setting out precisely rather than as a
caution.

### Where bias comes from {-}

Suppose the true population model is

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + u$$

but we omit $x_2$ and estimate

$$y = \alpha_0 + \alpha_1 x_1 + e$$

What does $\hat{\alpha}_1$ estimate? Write the relationship between the two
explanatory variables as a regression of its own,

$$x_2 = \delta_0 + \delta_1 x_1 + \eta$$

and substitute into the true model:

::: {.theory}
$$
\begin{aligned}
y &= \beta_0 + \beta_1 x_1 + \beta_2(\delta_0 + \delta_1 x_1 + \eta) + u \\[4pt]
  &= (\beta_0 + \beta_2\delta_0) + (\beta_1 + \beta_2\delta_1)x_1
     + (\beta_2\eta + u)
\end{aligned}
$$

Matching this against $y = \alpha_0 + \alpha_1 x_1 + e$ gives

$$\alpha_1 = \beta_1 + \beta_2\delta_1$$

So the short regression estimates the true coefficient **plus a bias term**:

$$\text{bias} = \beta_2 \delta_1
= \big(\text{effect of } x_2 \text{ on } y\big)
\times \big(\text{relationship between } x_2 \text{ and } x_1\big)$$
:::

This is **omitted variable bias**, and the formula says something immediately
useful. The bias is a product of two things, so it vanishes if *either* is zero.
An omitted variable is harmless unless it both affects $y$ and moves with $x_1$.

::: {.keyidea}
The sign of the bias is the product of the two signs.

| $\beta_2$ | $\delta_1$ | Bias in $\hat{\alpha}_1$ |
|:--|:--|:--|
| $+$ | $+$ | Upward |
| $+$ | $-$ | Downward |
| $-$ | $+$ | Downward |
| $-$ | $-$ | Upward |

This can often be signed from knowledge of the subject even when the variable
cannot be measured, which makes it possible to say which way an estimate is
likely to be wrong.
:::

### Checking the arithmetic {-}

The decomposition is exact in a sample, not merely a statement about
expectations. We can verify it on the grades data.

```{r mlr-ovb-check}
long <- coef(unrestricted)

d_cum <- coef(lm(cum_gpa ~ attendance, data = students))[2]
d_adm <- coef(lm(admission_score ~ attendance, data = students))[2]

round(c(short_regression = coef(restricted)["attendance"],
        long_coefficient = long["attendance"],
        bias_via_cum_gpa = long["cum_gpa"] * d_cum,
        bias_via_adm     = long["admission_score"] * d_adm,
        reassembled      = long["attendance"] +
                           long["cum_gpa"] * d_cum +
                           long["admission_score"] * d_adm), 6)
```

The simple regression coefficient reassembles exactly: 0.136 of attendance, plus
0.062 arriving through prior GPA, less a small amount through admission score.
The gap between 0.189 and 0.136 is not a mystery. It is a quantity we can name
and compute.

::: {.intuition}
Prior GPA biased the simple regression *upward* because both signs were positive:
better students attend more ($\delta_1 > 0$) and better students earn higher
grades ($\beta_2 > 0$).

Admission score worked the other way. Its coefficient is positive, but in these
data students with higher admission scores attend slightly *fewer* classes, so
its contribution is negative and small.
:::

### What the controls did and did not do {-}

The controls removed the bias from prior GPA and admission score. They did
nothing about anything else.

::: {.watchout}
Motivation is not in the model. Neither is interest in the subject, nor health,
nor how far a student lives from campus, nor whether they hold a job.

Every one of these plausibly affects grades, and every one plausibly moves with
attendance. Each contributes its own $\beta_2\delta_1$ to the estimate, and the
formula does not care that we cannot measure it.
:::

So 0.136 is a better estimate than 0.189 in a specific and limited sense: two
named sources of bias have been removed. It is not the causal effect of
attendance, and the regression output cannot tell us how far from it we are.

::: {.keyidea}
Multiple regression controls for what is **measured**. Omitted variable bias is
caused by what is **not**.

Adding controls narrows the gap between the estimate and the causal effect
without ever closing it, and nothing in the output reports how much remains.
:::

### The other three ways CLRM4 fails {-}

Omission is the most common threat, and it is not the only one. The zero
conditional mean assumption also fails when:

- **The functional form is wrong.** If earnings rise with experience and then
  flatten, a model linear in experience is misspecified, and the misspecification
  lands in $u$.
- **Variables are measured with error.** A variable recorded imperfectly is not
  the variable the model refers to, and the difference goes into the error term.
- **$y$ and $x$ are determined together.** Regressing quantity on price estimates
  neither the demand curve nor the supply curve, since the observed pairs are
  where the two meet.

Each has its own remedy, and this book returns to them.

::: {.intuition}
It is worth being clear about what has and has not changed since Unit
\@ref(ols).

Multiple regression is a real advance. It is the reason economists can report
returns to education that are not simply the fact that educated people differ
from uneducated ones in a dozen other ways.

It is also not enough. The comparison it makes is between observations alike in
the variables we happened to measure, and the variables we most want to hold
fixed — ability, motivation, the quality of a school — are usually the ones no
survey records.

Closing that gap requires a different kind of idea, and the second half of this
book is about where such ideas come from.
:::

## Summary {#mlr-summary}

::: {.keyidea}
- The **multiple regression model** is
  $y = \beta_0 + \beta_1x_1 + \cdots + \beta_kx_k + u$. Taking conditional
  expectations, $E[y \mid x_1,\ldots,x_k]$ is the population regression function,
  now a plane rather than a line.
- Each $\beta_j$ is a **partial derivative**: the change in average $y$ for a
  one-unit rise in $x_j$, holding the other explanatory variables fixed. This is
  the *ceteris paribus* interpretation, and it is why multiple regression exists.
- "Holding fixed" is an operation on the data, not an experiment. By the
  **Frisch–Waugh–Lovell** result, $\hat{\beta}_j$ can be obtained by regressing
  $x_j$ on the other regressors, keeping the residuals, and regressing $y$ on
  those residuals alone.
- It follows that the coefficient uses only the variation in $x_j$ that the other
  variables cannot explain. Controls remove variation, and removing variation
  costs precision.
- The assumptions are those of Unit \@ref(ols) with the conditioning extended to
  all regressors, plus **CLRM7, no perfect collinearity**. A violation is not a
  statistical problem but an arithmetic one: the comparison the coefficient
  refers to does not exist in the data, and R returns `NA`.
- $\mathrm{Var}(\hat{\beta}_j) = \sigma^2 / [\sum(x_{ij}-\bar{x}_j)^2(1-R_j^2)]$,
  where $R_j^2$ comes from regressing $x_j$ on the other regressors. The larger
  $R_j^2$, the more the control costs. Also
  $\hat{\sigma}^2 = \sum \hat{u}_i^2/(n-k-1)$.
- Testing a single coefficient is unchanged:
  $t = \hat{\beta}_j/\mathrm{se}(\hat{\beta}_j)$ on $n-k-1$ degrees of freedom.
- Testing several at once requires the **F test**, which compares a restricted
  model against an unrestricted one:
  $F = [(\text{SSR}_r - \text{SSR}_{ur})/q] \big/ [\text{SSR}_{ur}/(n-k-1)]$. It
  is always a right-tailed test.
- Separate $t$-tests are not a substitute. Correlated regressors can be
  individually insignificant and jointly overwhelming, because each $t$-test uses
  only the part of its variable that the others cannot explain.
- $F$ is the ratio of two independent variance estimates, each divided by its
  degrees of freedom. Under the null both estimate $\sigma^2$, so $F$ averages
  about one. Its distribution is right-skewed, so most of the mass lies below
  that average.
- **Omitted variable bias** is $\beta_2\delta_1$ — the effect of the omitted
  variable on $y$, times its relationship with the included one. It vanishes if
  either factor is zero, and its sign is the product of their signs.
- Controlling for measured variables removes their contribution to the bias and
  nothing else. Multiple regression narrows the gap to a causal effect without
  closing it, and reports nothing about what remains.
:::

## Practice problems {#mlr-practice}

::: {.yourturn}
Work each one before opening the solution. The
[question bank](#mlr-question-bank) that follows has no answers.
:::

### Numerical problems {-}

**8.1** A researcher estimates the effect of fertiliser on rice yield across
farms, first alone and then controlling for irrigation:

$$\widehat{\text{yield}} = 12.4 + 0.85\,\text{fert}
\qquad\qquad
\widehat{\text{yield}} = 9.1 + 0.52\,\text{fert} + 6.3\,\text{irrig}$$

where `irrig` is hectares under irrigation.

i. Interpret 0.52 precisely.
ii. Using the omitted variable bias formula, what must be the sign of the
    relationship between fertiliser use and irrigation?
iii. Compute the implied slope from a regression of irrigation on fertiliser.

::: {.answer}
**(i)** Among farms with the same irrigated area, one additional unit of
fertiliser is associated with 0.52 more quintals of yield on average.

**(ii)** The short regression gives 0.85 and the long one 0.52, so the bias is
$0.85 - 0.52 = +0.33$. Since $\beta_2 = 6.3 > 0$, the bias formula
$\beta_2\delta_1 = 0.33$ requires $\delta_1 > 0$. Farms using more fertiliser
also have more irrigation — plausible, since both are investments made by
better-resourced farms.

**(iii)** $\delta_1 = 0.33 / 6.3 = 0.052$ hectares of irrigation per unit of
fertiliser.
:::

**8.2** A regression of household consumption on income and household size gives
$R^2 = 0.68$ with $n = 240$.

i. State the null hypothesis of the overall $F$ test and compute the statistic.
ii. The 5% critical value with 2 and 237 degrees of freedom is 3.03. What do you
    conclude?

::: {.answer}
**(i)** $H_0: \beta_1 = \beta_2 = 0$; both slopes are zero. With $k = 2$ and
$n - k - 1 = 237$,

$$F = \frac{0.68/2}{(1-0.68)/237} = \frac{0.34}{0.00135} = 251.8$$

**(ii)** $251.8 \gg 3.03$, so we reject decisively. Income and household size
together explain a substantial share of variation in consumption. As Section
\@ref(mlr-f) notes, this is a low bar and rejecting it establishes little beyond
the model not being empty.
:::

**8.3** An unrestricted model with 5 explanatory variables and $n = 90$ has
$\text{SSR}_{ur} = 148$. Dropping three of them gives $\text{SSR}_r = 176$.

i. Compute the $F$ statistic.
ii. With 3 and 84 degrees of freedom the 5% critical value is 2.71. Conclude.
iii. Why must $\text{SSR}_r$ exceed $\text{SSR}_{ur}$?

::: {.answer}
**(i)** $q = 3$, $n - k - 1 = 90 - 5 - 1 = 84$.

$$F = \frac{(176 - 148)/3}{148/84} = \frac{9.333}{1.762} = 5.30$$

**(ii)** $5.30 > 2.71$: reject. At least one of the three dropped variables has a
non-zero coefficient.

**(iii)** The unrestricted model can always reproduce the restricted one by
setting those three coefficients to zero, and least squares chooses the values
that minimise SSR. So the unrestricted SSR can never be larger.
:::

**8.4** In a wage regression, $R_j^2$ from regressing years of education on the
other explanatory variables is 0.10 in one specification and 0.90 in another.
Holding $\sigma^2$ and $\sum(x_{ij}-\bar{x}_j)^2$ fixed, by what factor does the
standard error of the education coefficient differ between the two?

::: {.answer}
The variance is proportional to $1/(1-R_j^2)$, so the standard error is
proportional to $1/\sqrt{1-R_j^2}$.

$$\frac{\sqrt{1/(1-0.90)}}{\sqrt{1/(1-0.10)}} = \sqrt{\frac{0.90}{0.10}} = 3$$

The standard error is three times larger in the second specification. The point
estimate may be no better; the reported uncertainty is three times worse.
:::

### Concept checks {-}

**8.5** A student says: "The coefficient on attendance fell from 0.189 to 0.136
when I added controls, so the controls made the estimate worse." Respond.

::: {.answer}
The two numbers answer different questions, so neither is a corrupted version of
the other. 0.189 compares all students to all students; 0.136 compares students
alike in prior record.

If the goal is to describe how grades vary with attendance in the population,
0.189 is the correct answer to that question. If the goal is to isolate what
attendance contributes, 0.136 is closer, because one identified source of
contamination has been removed.

"Worse" presumes a target. The change is evidence that prior record was
contributing to the simple slope — which is exactly what we suspected.
:::

**8.6** Explain why a variable that is strongly correlated with $y$ can be
insignificant in a multiple regression.

::: {.answer}
By Section \@ref(mlr-fwl), the coefficient uses only the part of the variable that
the other regressors cannot explain. If the variable is closely related to the
other regressors, that remaining part is small, and a small amount of variation
supports only an imprecise estimate.

The variable may be strongly related to $y$ while contributing nothing the other
variables have not already contributed. That is a statement about what the
variable adds, not about whether it matters.
:::

**8.7** A regression includes both `attendance` and `classes_missed`, and R
reports `NA` for the second. Explain what has happened and what should be done.

::: {.answer}
The two sum to 32 for every student, so one is an exact linear function of the
other and CLRM7 fails. There is no observation for which attendance changes while
classes missed does not, so the *ceteris paribus* comparison the coefficient
refers to does not exist in the data.

R drops one variable and reports `NA` rather than failing. The remedy is to
include only one of them: they carry identical information.
:::

**8.8** Two variables are individually insignificant at 5% but jointly
significant at 1%. What should be concluded, and what should not?

::: {.answer}
Conclude that at least one of the two has a non-zero coefficient, and that the
data cannot say which — a signature of strongly correlated regressors.

Do not conclude that both are irrelevant, and do not drop both on the strength of
the two $t$-tests. Do not conclude that both matter either: the joint test is
silent about which.
:::

### R exercises {-}

**8.9** Using `attendance-grades.csv`, verify the Frisch–Waugh–Lovell result for
`cum_gpa` rather than `attendance`: regress `cum_gpa` on the other two
regressors, keep the residuals, regress `sem_gpa` on them, and compare the slope
with the multiple regression coefficient.

::: {.answer}
```{r mlr-ex-fwl, eval = FALSE}
students <- read.csv("data/attendance-grades.csv")
full <- lm(sem_gpa ~ attendance + cum_gpa + admission_score, data = students)

r <- residuals(lm(cum_gpa ~ attendance + admission_score, data = students))

c(two_step = coef(lm(students$sem_gpa ~ r))[2],
  direct   = coef(full)["cum_gpa"])
```

The two agree to machine precision. The result holds for every coefficient, not
just the one of interest — each is the slope from a regression on the part of its
own variable that the others cannot explain.
:::

**8.10** Using `wages-india-synthetic.csv`, test whether sex and residence
jointly belong in a regression of annual earnings on education and age. Report
the $F$ statistic, its degrees of freedom and your conclusion.

::: {.answer}
```{r mlr-ex-f, eval = FALSE}
wages <- read.csv("data/wages-india-synthetic.csv")
wages$male  <- as.integer(wages$sex == "Male")
wages$urban <- as.integer(wages$residence == "Urban")

r  <- lm(annual_earnings ~ education + age, data = wages)
ur <- lm(annual_earnings ~ education + age + male + urban, data = wages)

anova(r, ur)
```

$F = 900.9$ on 2 and 19,995 degrees of freedom, with a $p$-value below any
printable threshold. Sex and residence belong in the model.

That they belong does not make the model causal. It says earnings differ
systematically by sex and residence among people of the same education and age —
a fact about the labour market, not an explanation of it.
:::

**8.11** Show by simulation that omitted variable bias behaves as the formula
predicts. Generate $x_2$ correlated with $x_1$, construct
$y = 1 + 2x_1 + 3x_2 + u$, estimate the short regression 2,000 times, and compare
the average $\hat{\alpha}_1$ with $\beta_1 + \beta_2\delta_1$.

::: {.answer}
```{r mlr-ex-ovb, eval = FALSE}
set.seed(20)
n <- 200

results <- replicate(2000, {
  x1 <- rnorm(n)
  x2 <- 0.5 * x1 + rnorm(n)              # delta_1 = 0.5 by construction
  y  <- 1 + 2 * x1 + 3 * x2 + rnorm(n, sd = 2)
  c(short = coef(lm(y ~ x1))[2], long = coef(lm(y ~ x1 + x2))[2])
})

c(mean_short = mean(results["short.x1", ]),
  predicted  = 2 + 3 * 0.5,
  mean_long  = mean(results["long.x1", ]))
```

The short regression averages about 3.5, matching
$\beta_1 + \beta_2\delta_1 = 2 + 3(0.5)$. The long regression averages about 2,
the true value.

The bias does not shrink with sample size. Increasing $n$ to 20,000 leaves it
where it is — bias is not sampling error, and no amount of data removes it.
:::

**8.12** Investigate the cost of a redundant control. Regress `sem_gpa` on
`attendance` alone, then add `cum_gpa`, then add a variable constructed to be
almost identical to `attendance`. Compare the standard errors and explain.

::: {.answer}
```{r mlr-ex-collin, eval = FALSE}
students <- read.csv("data/attendance-grades.csv")
set.seed(5)
students$attendance_copy <- students$attendance + rnorm(nrow(students), sd = 0.4)

se <- function(m) round(coef(summary(m))["attendance", 2], 4)

c(alone      = se(lm(sem_gpa ~ attendance, data = students)),
  plus_cum   = se(lm(sem_gpa ~ attendance + cum_gpa, data = students)),
  plus_copy  = se(lm(sem_gpa ~ attendance + attendance_copy, data = students)))
```

Adding `cum_gpa` barely moves the standard error and removes a real source of
bias. Adding the near-copy inflates it enormously while contributing no
information at all: $R_j^2$ for attendance is now close to one, so $(1-R_j^2)$ is
close to zero.

This is the practical content of Section \@ref(mlr-assumptions). Controls are not
free, and a control that duplicates the variable of interest is all cost.
:::

## Question bank {#mlr-question-bank}

### Multiple choice {-}

**QB 8.1** In the model $y = \beta_0 + \beta_1x_1 + \beta_2x_2 + u$, the
coefficient $\beta_1$ measures

a. the total change in $y$ when $x_1$ rises by one unit
b. the change in average $y$ when $x_1$ rises by one unit, holding $x_2$ fixed
c. the correlation between $x_1$ and $y$
d. the share of variation in $y$ explained by $x_1$

**QB 8.2** The degrees of freedom for a $t$-test on a single coefficient in a
regression with $n = 150$ and four explanatory variables is

a. 146 &nbsp;&nbsp; b. 145 &nbsp;&nbsp; c. 148 &nbsp;&nbsp; d. 149

**QB 8.3** Perfect collinearity between two explanatory variables means

a. the OLS estimates are biased
b. the OLS estimates have large standard errors
c. the OLS estimates cannot be computed at all
d. the $R^2$ will be exactly one

**QB 8.4** An omitted variable causes no bias if

a. it is uncorrelated with the dependent variable
b. it is uncorrelated with the included explanatory variables
c. its coefficient is large
d. the sample is large

**QB 8.5** The $F$ test for exclusion restrictions is always

a. two-tailed
b. left-tailed
c. right-tailed
d. tailed according to the alternative hypothesis

### Short reasoning {-}

**QB 8.6** Explain, without using the phrase "holding constant", what the
Frisch–Waugh–Lovell result says a multiple regression coefficient is computed
from.

**QB 8.7** A colleague argues that since adding controls always reduces
precision, one should include as few as possible. Assess this.

**QB 8.8** Why does the numerator of the $F$ statistic divide by $q$ and the
denominator by $n-k-1$? What would go wrong if neither division were performed?

**QB 8.9** In a study of the effect of an irrigation scheme on farm incomes, the
researcher controls for landholding, soil quality and district. Name two
variables likely still omitted, and use the bias formula to argue in which
direction each would push the estimate.

**QB 8.10** A regression of infant mortality on the number of doctors per
thousand people across Indian districts returns a *positive* coefficient. Explain
using the omitted variable bias formula, and suggest a control.

**QB 8.11** The overall $F$ statistic in a regression is significant but no
individual coefficient is. What does this suggest about the explanatory
variables, and what would you do next?

**QB 8.12** Explain why an omitted variable that strongly affects $y$ but is
unrelated to the explanatory variable of interest costs precision but causes no
bias.

### Numerical {-}

**QB 8.13** A researcher estimates $\widehat{\text{wage}} = 4200 + 380\,educ$ and
then $\widehat{\text{wage}} = 3100 + 240\,educ + 95\,exper$. Compute the implied
regression slope of experience on education, and comment on its sign.

**QB 8.14** An unrestricted model with $n = 120$, $k = 6$ has $R^2_{ur} = 0.54$.
Imposing four exclusion restrictions gives $R^2_r = 0.41$. Compute the $F$
statistic and state its degrees of freedom.

**QB 8.15** A regression with $n = 500$ and $k = 3$ reports
$\sum \hat{u}_i^2 = 892$. Compute $\hat{\sigma}^2$ and $\hat{\sigma}$, and state
why the divisor is not $n$.

**QB 8.16** For a regression of yield on fertiliser, irrigation and labour, the
$R^2$ from regressing fertiliser on the other two is 0.76. By what percentage is
the standard error on fertiliser inflated relative to a hypothetical case in
which fertiliser were uncorrelated with the other regressors?

## Looking ahead {#mlr-looking-ahead}

This unit took the assumptions and the variables as given. Both deserve
questioning.

The model has assumed throughout that a one-unit rise in $x$ has the same
consequence wherever it occurs — that the tenth year of schooling is worth what
the second was, and that the effect of experience never flattens. Economics
rarely believes this. The next unit shows how logarithms, squares and interaction
terms let a linear model describe relationships that are not linear at all, at no
cost to any of the machinery developed here.

The unit also treated the choice of controls as settled. It is not. Some
variables belong in a regression, some are harmless, and some make the estimate
worse than leaving them out — a possibility that surprises most readers, since
the instinct is that controlling for more is always safer. Distinguishing good
controls from bad ones is the second theme of the next unit, and it depends on
exactly the bias formula derived here.

## Mathematical note: multiple regression in matrix form {#mlr-matrix-note}

::: {.deeper}
Nothing in this unit depends on what follows. It is here for readers who would
like to see the general case written compactly, and it can be skipped without
consequence.

**The estimator.** With $k$ explanatory variables, writing out $k+1$ normal
equations and solving them by hand is unpleasant. Matrix notation does the same
work in three lines.

Collect the observations into an $n \times 1$ vector $Y$ and an
$n \times (k+1)$ matrix $X$ whose first column is all ones and whose remaining
columns are the explanatory variables. The model is

$$Y = X\beta + u$$

The sum of squared residuals is $(Y - X\beta)'(Y - X\beta)$. Differentiating
with respect to $\beta$ and setting the result to zero,

$$-2X'Y + 2X'X\beta = 0
\qquad\Longrightarrow\qquad
X'X\hat{\beta} = X'Y$$

which is the whole system of normal equations in one line. Provided $X'X$ can be
inverted,

$$\hat{\beta} = (X'X)^{-1}X'Y$$

**The variance.** Substituting $Y = X\beta + u$ gives
$\hat{\beta} = \beta + (X'X)^{-1}X'u$. The second term has expectation zero
under CLRM4, which is unbiasedness. Taking the variance and using
$\mathrm{Var}(Au) = A\,\mathrm{Var}(u)\,A'$, then imposing homoskedasticity,

$$\mathrm{Var}(\hat{\beta}) = \sigma^2 (X'X)^{-1}$$

This is a $(k+1) \times (k+1)$ matrix. Its diagonal holds the variances of the
individual coefficients — the square roots of which are the standard errors R
prints — and its off-diagonal entries hold the covariances between them, which
is what the $F$ test of Section \@ref(mlr-f) needs in order to test several
coefficients jointly.

**Checking it.** The formula is short enough to compute directly.

```{r mlr-matrix}
X <- cbind(1, students$attendance, students$cum_gpa, students$admission_score)
y <- students$sem_gpa

b_hat <- solve(t(X) %*% X) %*% t(X) %*% y
round(drop(b_hat), 5)
```

The same four numbers `lm()` reported in Section \@ref(mlr-model).

**Where collinearity appears.** The requirement in Section
\@ref(mlr-assumptions) that no explanatory variable be an exact linear
combination of the others is exactly the requirement that $X'X$ be invertible.
If one column of $X$ can be written in terms of the others, $X$ is not of full
rank, $X'X$ is singular, and there is no $\hat{\beta}$ to compute — which is
what R is reporting when it returns `NA`.
:::
