# Which Model Should We Fit? {#specification}

```{r setup-09, include = FALSE}
source("_common.R")
library(ggplot2)
```

Every regression so far arrived with its model already written. Semester GPA on
attendance, prior GPA and admission score; earnings on education, age, sex and
residence. We estimated what was put in front of us.

In practice nobody hands you the equation. Two decisions have to be made before
anything can be estimated, and neither is settled by the data alone.

The first is the **form** of the relationship. We have assumed throughout that a
one-unit rise in $x$ produces the same change in $y$ wherever it occurs — that
the tenth year of schooling is worth what the second was. Economics rarely
believes this.

The second is **which variables belong**. Unit \@ref(multiple-regression) showed
that omitting a relevant variable biases the estimate. It is tempting to
conclude that more controls are always safer.

They are not. Some variables belong in a regression, some are merely wasteful,
and some make the estimate worse than leaving them out.

This unit takes the two decisions in turn.

## Is a straight line the right model? {#ff-motivation}

::: {.question}
A slope of $-0.006$ km per litre per kilogram. Is that a lot?
:::

Section \@ref(ols-practice) fitted fuel efficiency on weight for 74 cars and
returned a slope of $-0.00601$.

Nothing about that number tells you whether the effect is large.

```{r ff-units}
cars <- read.csv("data/car-mileage.csv")

c(per_kg     = coef(lm(mileage_kmpl ~ weight_kg, data = cars))[2],
  per_tonne  = coef(lm(mileage_kmpl ~ I(weight_kg / 1000), data = cars))[2])
```

The same relationship, the same data, the same fit — and two numbers differing
by a factor of a thousand. Measure efficiency in litres per hundred kilometres
instead and the sign flips.

Two problems follow, and one transformation addresses both.

**Units are arbitrary.** A coefficient in kilograms and one in tonnes describe
the same relationship, but neither is easier to judge than the other. We would
prefer a measure that does not change when someone changes their mind about
units.

**Straight lines are not always the right shape.** Section \@ref(sr-prf) showed
that the conditional mean of GPA against attendance was not exactly linear, and
took a line as an approximation. Sometimes a better approximation is available
at no cost, by transforming a variable before fitting.

Taking logarithms addresses both.

## The four functional forms {#ff-four}

::: {.question}
Logarithms of what, and what happens to the interpretation?
:::

Either variable may be entered in levels or in logarithms, giving four
combinations. Everything else about the regression is unchanged: it is still
linear in the parameters, still fitted by least squares, still tested with the
same $t$ and $F$ statistics.

What changes is the meaning of $\beta_1$.

Table: (\#tab:ff-forms) The four functional forms and what the slope means in each.

| Model | Equation | Interpretation of $\beta_1$ |
|:--|:--|:--|
| Level–level | $y = \beta_0 + \beta_1 x + u$ | A one-unit rise in $x$ changes $y$ by $\beta_1$ **units** |
| Log–level | $\ln y = \beta_0 + \beta_1 x + u$ | A one-unit rise in $x$ changes $y$ by $100\beta_1$ **per cent** |
| Level–log | $y = \beta_0 + \beta_1 \ln x + u$ | A 1% rise in $x$ changes $y$ by $\beta_1/100$ **units** |
| Log–log | $\ln y = \beta_0 + \beta_1 \ln x + u$ | A 1% rise in $x$ changes $y$ by $\beta_1$ **per cent** |

The last is an **elasticity**, and it is unit-free: it does not matter whether
weight is in kilograms or tonnes, or price in rupees or lakhs.

::: {.theory}
All four interpretations follow from one fact of calculus. For small changes,

$$\Delta \ln y \;\approx\; \frac{\Delta y}{y}$$

that is, a change in the logarithm is approximately a *proportional* change in
the variable. Multiplying by 100 turns it into a percentage.

**Log–level.** Differentiating $\ln y = \beta_0 + \beta_1 x$ with respect to $x$,

$$\frac{1}{y}\frac{dy}{dx} = \beta_1
\qquad\Longrightarrow\qquad
\frac{dy/y}{dx} = \beta_1$$

so a one-unit rise in $x$ raises $y$ by a proportion $\beta_1$, which is
$100\beta_1$ per cent.

**Level–log.** From $y = \beta_0 + \beta_1 \ln x$,

$$dy = \beta_1 \, d(\ln x) = \beta_1 \frac{dx}{x}$$

A 1% rise in $x$ means $dx/x = 0.01$, so $dy = 0.01\beta_1 = \beta_1/100$ units.

**Log–log.** From $\ln y = \beta_0 + \beta_1 \ln x$,

$$\frac{dy/y}{dx/x} = \beta_1$$

which is the definition of an elasticity: the ratio of the proportional change
in $y$ to the proportional change in $x$.
:::

::: {.watchout}
The percentage reading is an approximation, and it degrades as the coefficient
grows.

For $\beta_1 = 0.0942$ the approximation says 9.42% and the exact figure is
$100(e^{0.0942} - 1) = `r round(100*(exp(0.0942)-1), 2)`$%. Close enough.

For $\beta_1 = 0.7$ the approximation says 70% and the truth is
$`r round(100*(exp(0.7)-1), 0)`$%. Above about 0.2, use the exact expression.
:::

## Elasticities: weight and fuel efficiency {#ff-elasticity}

::: {.question}
What does the log–log form buy us on the cars?
:::

```{r ff-cars}
level_level <- lm(mileage_kmpl ~ weight_kg, data = cars)
log_log     <- lm(log(mileage_kmpl) ~ log(weight_kg), data = cars)

round(rbind(level_level = c(slope     = coef(level_level)[2],
                            r_squared = summary(level_level)$r.squared),
            log_log     = c(slope     = coef(log_log)[2],
                            r_squared = summary(log_log)$r.squared)), 4)
```

The log–log slope is $-0.825$: **a car that is 1% heavier is about 0.83% less
fuel-efficient.**

That is a sentence anyone can evaluate, and — unlike $-0.00601$ — it would be
unchanged if weight were recorded in tonnes.

```{r ff-cars-plot, echo = FALSE, fig.height = 3.4, cache = TRUE, fig.cap = "The same 74 cars in levels and in logarithms. Both look broadly linear; the logarithmic version is a little tighter around the line, which is the improvement in R-squared from 0.65 to 0.71. The gain here is real but modest -- the case for the log form rests mainly on interpretation."}
both <- rbind(
  data.frame(x = cars$weight_kg, y = cars$mileage_kmpl, panel = "Levels"),
  data.frame(x = log(cars$weight_kg), y = log(cars$mileage_kmpl),
             panel = "Logarithms"))

ggplot(both, aes(x, y)) +
  geom_point(alpha = 0.55, colour = book_palette$ink, size = 1.3) +
  geom_smooth(method = "lm", se = FALSE, colour = book_palette$accent,
              linewidth = 0.9) +
  facet_wrap(~ panel, scales = "free") +
  labs(x = NULL, y = NULL) +
  theme_book()
```

The fit also improves, from $R^2 = 0.65$ to $0.71$. That is a modest gain rather
than a dramatic one, and the case for the log form here rests mainly on
interpretation. Still, the direction is not an accident: a constant
*proportional* penalty for weight describes cars better than the constant
absolute penalty the level–level model imposed.

::: {.watchout}
A great deal of weak empirical work consists of fitting a straight line to a
relationship that is not straight, and reporting the slope as though it
described the whole range.

The residual plot of Section \@ref(ols-hetero) catches this: a *curved* band of
residuals means the functional form is wrong, not that the variance is
non-constant.

Fitting a line is a choice. Check it.
:::

## Percentages: the return to schooling {#ff-schooling}

::: {.question}
Why is almost every wage regression in economics run in logs?
:::

Because wages are discussed in percentages. Nobody asks how many rupees a year
of schooling is worth without immediately wanting to know rupees compared to
what.

```{r ff-wages}
wages <- read.csv("data/wages-india-synthetic.csv")

wages$male  <- as.integer(wages$sex == "Male")
wages$urban <- as.integer(wages$residence == "Urban")

log_simple <- lm(log(annual_earnings) ~ education, data = wages)
log_full   <- lm(log(annual_earnings) ~ education + age + male + urban,
                 data = wages)

round(coef(summary(log_full))[, 1:2], 5)

round(c(education_alone         = coef(log_simple)["education"],
        education_with_controls = coef(log_full)["education"]), 5)
```

The simple regression gives 0.0942: **one more year of schooling is associated
with about 9.4% higher annual earnings.** Multiply by 100 and read as a
percentage; that is the whole rule.

With the controls of Unit \@ref(multiple-regression) in place it falls to
0.0762, or about 7.6%. The same lesson as Section \@ref(mlr-motivation), now in
percentage terms: part of the apparent return to education was the fact that the
educated in this population are disproportionately urban and male.

::: {.intuition}
Notice what the log form did to the other coefficients as well.

The coefficient on `urban` is 0.80, so urban workers earn roughly 80% more than
otherwise comparable rural workers — or, exactly,
$`r round(100*(exp(0.8045)-1), 0)`$% more, since 0.80 is well above the range
where the approximation is safe.

In levels that same gap was ₹51,287, a number that means nothing until you also
know what rural workers earn.
:::

## When logs remove what looked like heteroskedasticity {#ff-hetero}

::: {.question}
Section \@ref(ols-hetero) met a fan-shaped residual plot and reached for robust
standard errors. Was that the only option?
:::

Robust standard errors accept the heteroskedasticity and correct for it. The
other response is to ask why it was there.

Earnings are a highly skewed variable, and skewed variables often produce
fan-shaped residuals simply because their upper tail is long. Taking logarithms
compresses that tail.

```{r ff-hetero-compare}
spread_by_quartile <- function(m) {
  q <- cut(fitted(m), quantile(fitted(m), 0:4/4), include.lowest = TRUE)
  s <- tapply(residuals(m), q, sd)
  c(round(max(s) / min(s), 3))
}

c(levels = spread_by_quartile(lm(annual_earnings ~ education, data = wages)),
  logs   = spread_by_quartile(log_simple))
```

In levels, the residuals in the top quarter of fitted values are 3.6 times as
spread out as those in the bottom quarter. In logs the ratio is 1.01 — the
spread is essentially identical across quartiles.

::: {.keyidea}
The heteroskedasticity was not a deep feature of the world. It was a consequence
of measuring earnings in rupees rather than in proportional terms.

This is the second thing a fan-shaped residual plot can be telling you. Section
\@ref(ols-hetero) treated it as a fact about the variance, to be accepted and
corrected. Sometimes it is instead a fact about the **scale** of the variable,
and changing the scale removes it entirely.
:::

Robust standard errors would have widened the intervals around the problem.
Taking logs asked why it was there.

## Choosing a form {#ff-choosing}

::: {.question}
Four options and no test that settles it. How is the choice made?
:::

::: {.keyidea}
Three considerations, in order of importance.

**What is the natural way to talk about the variable?** Wages, prices and GDP
are discussed in percentages, so logs. Test scores, class attendance and family
size are discussed in units, so levels.

**Does the relationship look multiplicative?** If a proportional change in $x$
plausibly produces a proportional change in $y$ — as with weight and fuel use —
logs will fit better and mean more.

**Does the scatter suggest it?** Skewed variables often straighten out in logs,
and heteroskedasticity often shrinks at the same time.
:::

What should *not* drive the choice is which version gives the larger $R^2$ or
the smaller $p$-value. Trying transformations until something is significant
produces findings that will not replicate — and Section \@ref(ff-r2-limits)
explains why $R^2$ is particularly unsuited to settling the question.

::: {.watchout}
**Logarithms are undefined at zero and for negative numbers.**

Earnings data routinely contain zeros — people who did not work. Taking logs
silently deletes them, and since those are rarely a random subset of the sample,
the remaining regression answers a different question from the one asked.

```{r ff-zero}
sum(wages$annual_earnings <= 0)
```

There are none here, which is a property of this simulated dataset rather than
of earnings data in general. With real data, check before transforming and say
what you did.
:::

## Does adding a variable ever cost anything? {#ff-irrelevant}

::: {.question}
If omitting a relevant variable causes bias, why not include everything
available?
:::

That is the second decision, and the instinct behind it is understandable.
Section \@ref(mlr-causal) showed that leaving out a relevant variable biases the
estimate. Including it fixes the problem. Why not include everything?

Start with the easy case: a variable that does not belong in the model at all.
Its true coefficient is zero.

::: {.keyidea}
Including an **irrelevant variable** does not bias anything.

The unbiasedness proof of Section \@ref(ols-unbiased) does not require that
every included variable matter. If $\beta_3 = 0$, then OLS estimates $\beta_3$
as zero on average and leaves the other coefficients centred on the truth.
:::

So the cost is not bias. It is precision, and how much precision depends
entirely on one thing.

```{r ff-irrelevant-sim, cache = TRUE}
set.seed(7)

out <- replicate(3000, {
  x1  <- rnorm(100)
  irr_alone <- rnorm(100)                                   # unrelated to x1
  irr_close <- 0.9 * x1 + sqrt(1 - 0.81) * rnorm(100)       # correlated 0.9 with x1
  y <- 1 + 2 * x1 + rnorm(100, sd = 2)                      # neither belongs in the model

  c(b_none  = coef(lm(y ~ x1))[2],
    b_alone = coef(lm(y ~ x1 + irr_alone))[2],
    b_close = coef(lm(y ~ x1 + irr_close))[2],
    se_none  = coef(summary(lm(y ~ x1)))[2, 2],
    se_alone = coef(summary(lm(y ~ x1 + irr_alone)))[2, 2],
    se_close = coef(summary(lm(y ~ x1 + irr_close)))[2, 2])
})

round(rowMeans(out), 4)
```

The true slope is 2. All three estimates average 2, exactly as promised — adding
a variable with no effect costs nothing in bias whether or not it is related to
$x_1$.

The standard errors tell a different story. Adding the unrelated variable moves
the standard error from 0.202 to 0.203, which is nothing. Adding the correlated
one moves it to 0.468, more than doubling it.

::: {.intuition}
This is the $(1-R_j^2)$ factor of Section \@ref(mlr-assumptions), and it is
worth seeing that it does not care whether the added variable matters.

A control that duplicates the variable of interest consumes its variation
whether or not it has any effect on the outcome. The regression is left
estimating a coefficient from whatever variation survives, and pays for the
privilege in precision.
:::

## R-squared always rises {#ff-adjr2}

::: {.question}
Can we use $R^2$ to decide whether a variable belongs?
:::

The suggestion is natural. $R^2$ measures how much of the variation the model
accounts for, so a variable that improves the model should raise it.

The difficulty is that *every* variable raises it.

```{r ff-r2-noise, cache = TRUE}
students <- read.csv("data/attendance-grades.csv")

set.seed(2)
noise <- as.data.frame(matrix(rnorm(nrow(students) * 10), ncol = 10))
augmented <- cbind(students, noise)

base <- lm(sem_gpa ~ attendance + cum_gpa + admission_score, data = students)
junk <- lm(reformulate(c("attendance", "cum_gpa", "admission_score",
                         names(noise)), "sem_gpa"), data = augmented)

round(c(r2_base        = summary(base)$r.squared,
        r2_plus_junk   = summary(junk)$r.squared,
        adj_base       = summary(base)$adj.r.squared,
        adj_plus_junk  = summary(junk)$adj.r.squared), 4)
```

The ten added variables are pure random noise, generated without reference to
the data. They cannot possibly explain semester GPA.

$R^2$ rises anyway, from 0.5421 to 0.5464.

::: {.keyidea}
$R^2$ can never fall when a variable is added.

Least squares could always set the new coefficient to zero and reproduce the
old fit exactly. Since it minimises the sum of squared residuals, it will only
depart from zero if doing so fits *this sample* better — which pure noise
occasionally does, by chance.

A measure that never falls cannot be used to decide whether something belongs.
:::

### Adjusted R-squared {-}

The repair is to charge for each parameter estimated.

$$\bar{R}^2 = 1 - \frac{\text{SSR}/(n-k-1)}{\text{SST}/(n-1)}
\qquad\text{equivalently}\qquad
\bar{R}^2 = 1 - \frac{(1-R^2)(n-1)}{n-k-1}$$

The ordinary $R^2$ compares two sums of squares directly. The adjusted version
divides each by its degrees of freedom first, so adding a variable raises $k$,
shrinks $n-k-1$, and inflates the numerator unless the variable earns its place.

In the output above, adjusted $R^2$ **falls** from 0.5401 to 0.5375. It has
detected what $R^2$ could not.

::: {.watchout}
Adjusted $R^2$ can be negative, and it is not a proportion of anything. It is a
model comparison device, not a measure of fit.

It is also a weak one. It penalises parameters, not irrelevance, and a variable
that is genuinely useless but happens to correlate with the outcome in this
sample will still raise it.
:::

## What R-squared cannot tell you {#ff-r2-limits}

::: {.question}
The model explains 54% of the variation. Is it a good model?
:::

There is no answer to that question, because $R^2$ was never a measure of
whether a model is correct.

::: {.keyidea}
$R^2$ answers exactly one question: **how much of the variation in $y$ does the
fitted model account for in this sample?**

It says nothing about whether the coefficients are unbiased, whether the
standard errors are right, whether the functional form is sensible, or whether
anything can be interpreted causally.
:::

Four specific limitations are worth stating plainly.

**A high $R^2$ does not mean the estimate is unbiased.** Omitted variable bias
depends on what is *missing* from the model. A regression can account for 95% of
the variation and still have every coefficient wrong, because the variable that
would have corrected them was never included. Nothing in $R^2$ can detect a
variable that is not there.

**A low $R^2$ does not mean the model is useless.** Individual behaviour is
mostly idiosyncratic. Cross-sectional regressions on household data routinely
report $R^2$ of 0.1 or 0.2, and the coefficients may still be precisely
estimated and economically important. Section \@ref(ff-schooling) explains 32% of
the variation in log earnings; the return to schooling is not thereby 32% true.

**$R^2$ cannot be compared across different dependent variables.** The earnings
regression in levels and the same regression in logs have different $R^2$
because they are explaining the variation in different quantities. Choosing
between functional forms on $R^2$ compares two things that are not comparable.

**$R^2$ rewards prediction, not explanation.** A variable measured *after* the
outcome, or one that is a consequence of it, will improve the fit
spectacularly and destroy the interpretation. That is the subject of the next
section.

::: {.intuition}
The two purposes of regression are worth separating here.

If the goal is **prediction** — forecasting next year's output, flagging which
loans will default — then fit is close to the whole point, and $R^2$ is a
reasonable thing to look at.

If the goal is **explanation** — estimating what a year of schooling is worth —
then fit is nearly beside the point. What matters is whether the comparison
being made is the right one, and $R^2$ is silent about that.

Most of economics is the second kind of question.
:::

## Good controls and bad controls {#ff-controls}

::: {.question}
Adding variables costs precision but not bias. So is a control ever actively
harmful?
:::

It is, and this is the most important idea in the unit.

Section \@ref(ff-irrelevant) established that a variable with no effect on $y$
is harmless. That result is narrower than it looks, because it assumed the
variable's true coefficient was zero. The dangerous cases are variables that do
belong to the story — just not in the regression.

Three variables, three simulations, identical structure. In each the true effect
of $x$ on $y$ is 1, and in each we compare the estimate with and without the
control.

### A variable that causes both {-}

This is the case Unit \@ref(multiple-regression) was built around. Ability
raises schooling and independently raises earnings.

```{r ff-confounder, cache = TRUE}
set.seed(19)
n <- 2000

confounder <- replicate(1500, {
  ability <- rnorm(n)
  x <- 0.7 * ability + rnorm(n)              # ability raises x
  y <- 1 * x + 1.5 * ability + rnorm(n)      # and raises y directly
  c(without_control = coef(lm(y ~ x))[2],
    with_control    = coef(lm(y ~ x + ability))[2])
})

round(rowMeans(confounder), 4)
```

Without the control the estimate is 1.71 against a truth of 1. With it, 1.00.

The control repaired the estimate, exactly as advertised. A variable that causes
both $x$ and $y$ is a **confounder**, and controlling for it is what multiple
regression is for.

### A variable that x causes {-}

Now change one thing. The added variable is not a cause of $x$; it is a
consequence of it.

Education raises occupational rank, and occupational rank raises earnings. Part
of how education pays is by getting people into better jobs.

```{r ff-mediator, cache = TRUE}
set.seed(19)

mediator <- replicate(1500, {
  educ <- rnorm(n)
  occupation <- 0.6 * educ + rnorm(n)                  # education raises rank
  earnings   <- 0.4 * educ + 1.0 * occupation + rnorm(n)
  c(without_control = coef(lm(earnings ~ educ))[2],
    with_control    = coef(lm(earnings ~ educ + occupation))[2])
})

round(rowMeans(mediator), 4)
```

The total effect of education is 1.0: a direct effect of 0.4, plus 0.6 through
occupation.

Without the control we recover 1.00 — correct. **With** the control we get 0.40.

::: {.watchout}
Controlling for occupation did not remove bias. It removed part of the effect.

Holding occupation fixed means comparing people who ended up in the same job.
Among such people, education can no longer operate through the route by which
most of it operates. What survives is only the direct effect.

A variable that lies on the causal path from $x$ to $y$ is a **mediator**, and
controlling for it answers a narrower question than the one asked — often
without the researcher noticing that the question changed.
:::

### A variable that both x and the error cause {-}

The third case is the least intuitive and the most damaging.

Suppose $x$ has a genuine effect on $y$ and there is no confounding whatever, so
a simple regression would have been correct. Now control for a variable that is
caused by both $x$ and the unobserved part of $y$.

```{r ff-collider, cache = TRUE}
set.seed(19)

collider <- replicate(1500, {
  x <- rnorm(n)
  u <- rnorm(n)                                  # unobserved, affects y
  y <- 1 * x + u                                 # no confounding: x and u independent
  selected <- 0.8 * x + 0.8 * u + rnorm(n)       # caused by both
  c(without_control = coef(lm(y ~ x))[2],
    with_control    = coef(lm(y ~ x + selected))[2])
})

round(rowMeans(collider), 4)
```

Without the control the estimate is 1.00, and correctly so — there was nothing
to correct.

Adding the control breaks it, giving 0.61.

::: {.keyidea}
Here the control **created** bias where none existed.

A variable caused by both $x$ and the error term is a **collider**. Conditioning
on it induces a relationship between $x$ and $u$ that was not there before,
which is precisely the failure of CLRM4 that Unit \@ref(multiple-regression)
warned about — manufactured by the researcher.
:::

::: {.intuition}
The mechanism is easier to see in a concrete case.

Suppose a firm hires workers who are either well-qualified or visibly talented,
and that qualifications and talent are unrelated in the population at large.
Among the people it hires, the two will be negatively related: someone hired
despite weak qualifications must have been unusually talented, and the reverse.

Restricting attention to those hired — or controlling for being hired —
manufactures a correlation between two things that were independent.

This is why controlling for variables measured *after* the treatment, or for
anything determining who appears in the sample, is dangerous in a way that has
nothing to do with sample size.
:::

### What this means in practice {-}

The three simulations are identical in form. Each adds one control to a
regression, and the results are opposite.

Table: (\#tab:ff-controls) The same operation, three times, with three different consequences. The true effect is 1 in every case.

| The control is | Without it | With it | Verdict |
|:--|:--|:--|:--|
| A cause of $x$ and $y$ (**confounder**) | 1.71 | 1.00 | Include it |
| Caused by $x$ (**mediator**) | 1.00 | 0.40 | Changes the question |
| Caused by $x$ and by $u$ (**collider**) | 1.00 | 0.61 | Leave it out |

::: {.keyidea}
No property of the data distinguishes these three cases.

All three controls are correlated with $x$ and with $y$. All three change the
coefficient. All three are statistically significant. Every diagnostic in this
book passes in all three regressions.

What separates them is the direction of causation between the variables — which
is knowledge about the world, not a fact the sample contains.
:::

This is the honest answer to "which variables should I control for?" The
question cannot be settled by looking at output. It requires an argument about
what causes what, made before the regression is run and defended in the text.

::: {.watchout}
The common practice of adding every available variable and reporting whichever
specification looks best is not caution. It is an invitation to all three
problems at once, and it is undetectable in the results.
:::

## Summary {#ff-summary}

::: {.keyidea}
- Two decisions precede any estimate: the **form** of the relationship and
  **which variables** enter. Neither is settled by the data.
- A level–level slope depends on the units of both variables. Logarithms remove
  that dependence and often improve the approximation as well.
- The four forms are level–level, log–level, level–log and log–log, with slopes
  read as units, per cent, units and per cent respectively. All follow from
  $\Delta \ln y \approx \Delta y / y$.
- A **log–log slope is an elasticity** and is unit-free. Cars: a 1% heavier car
  is 0.83% less fuel-efficient.
- The **log–level form** is standard for wages. A year of schooling is worth
  about 9.4% before controls and 7.6% after them.
- The percentage reading is an approximation. Above about 0.2, use
  $100(e^{\beta} - 1)$.
- A fan-shaped residual plot has two possible meanings: non-constant variance,
  or a badly scaled variable. In the earnings data, logs cut the ratio of
  residual spreads from 3.6 to 1.01, so it was the second.
- Logs are undefined at zero, and dropping the zeros silently changes the
  question being answered.
- An **irrelevant variable** causes no bias. Its cost is precision, and the cost
  depends on how strongly it is related to the variable of interest, not on
  whether it affects the outcome.
- **$R^2$ can never fall when a variable is added**, so it cannot decide whether
  one belongs. Ten pure noise variables raised it from 0.5421 to 0.5464.
- **Adjusted $R^2$** charges for each parameter and fell from 0.5401 to 0.5375 on
  the same regression. It is a comparison device, not a measure of fit, and a
  weak one.
- $R^2$ says nothing about bias, is not comparable across different dependent
  variables, and is low in most cross-sectional work for reasons that have
  nothing to do with model quality.
- Controls are not all alike. A **confounder** should be included; a
  **mediator** changes the question; a **collider** creates bias where none
  existed.
- Nothing in the data distinguishes the three. The decision rests on an argument
  about what causes what, made before the regression is run.
:::

## Practice problems {#ff-practice}

::: {.yourturn}
Work each one before opening the solution. The
[question bank](#ff-question-bank) that follows has no answers.
:::

### Numerical problems {-}

**9.1** A regression of log monthly household expenditure on log monthly income
across 900 Indian households gives a slope of 0.72 with a standard error of
0.04.

i. Interpret the slope.
ii. Test whether the elasticity equals one at the 5% level.
iii. What would an elasticity of one imply about how expenditure shares change
     with income?

::: {.answer}
**(i)** A 1% rise in household income is associated with a 0.72% rise in
expenditure. It is an elasticity, and unchanged if both were measured in
thousands of rupees.

**(ii)** $H_0: \beta_1 = 1$. Note this is *not* the test R prints, which uses
zero.

$$t = \frac{0.72 - 1}{0.04} = -7$$

With 898 degrees of freedom the critical value is about 1.96, so we reject
decisively. The elasticity is below one.

**(iii)** An elasticity of one means expenditure rises in exact proportion to
income, so the share of income spent is constant. At 0.72 the share *falls* as
income rises — richer households save a larger fraction, which is a very old
empirical regularity.
:::

**9.2** A wage regression in log–level form returns a coefficient of 0.31 on a
training programme indicator.

i. What does the approximation say the programme is worth?
ii. Compute the exact percentage.
iii. Why do they differ by so much here when they agreed in Section
     \@ref(ff-schooling)?

::: {.answer}
**(i)** $100 \times 0.31 = 31\%$.

**(ii)** $100(e^{0.31} - 1) = `r round(100*(exp(0.31)-1), 1)`\%$.

**(iii)** The approximation $\Delta \ln y \approx \Delta y / y$ holds for small
changes and degrades as the change grows. The schooling coefficient was 0.094,
well inside the safe range; 0.31 is not. The error is always in the same
direction — the approximation understates the true percentage.
:::

**9.3** A model with $n = 200$ and $k = 5$ has $R^2 = 0.42$.

i. Compute adjusted $R^2$.
ii. A sixth variable is added and $R^2$ rises to 0.425. Compute the new adjusted
    $R^2$ and say whether the variable earned its place.

::: {.answer}
**(i)** With $k = 5$, $n - k - 1 = 194$:

$$\bar{R}^2 = 1 - \frac{(1-0.42)(199)}{194} = `r round(1-(1-0.42)*199/194, 4)`$$

**(ii)** With $k = 6$, $n - k - 1 = 193$:

$$\bar{R}^2 = 1 - \frac{(1-0.425)(199)}{193} = `r round(1-(1-0.425)*199/193, 4)`$$

Adjusted $R^2$ rose, from 0.4051 to 0.4071, so on this criterion the variable
paid for itself.

Note how little it rose. A gain of 0.002 is well within the range that a
variable of pure noise can produce by chance, as Exercise 9.10 shows. Adjusted
$R^2$ is a weak criterion, and a near-tie like this one should not decide a
specification on its own.
:::

**9.3 (continued)** Why does a rise in adjusted $R^2$ not establish that the
sixth variable belongs in the model?

::: {.answer}
Because adjusted $R^2$ charges a fixed price per parameter and asks only whether
the improvement in fit exceeds it. It knows nothing about where the variable
sits in the causal structure.

A mediator or a collider will often raise adjusted $R^2$ — both are strongly
related to the outcome — while making the coefficient of interest worse. Section
\@ref(ff-controls) shows all three cases raising the fit and only one of them
belonging.
:::

**9.4** Using `india-production.csv`, estimate a Cobb–Douglas production
function $\ln Y = \beta_0 + \beta_1 \ln L + \beta_2 \ln K + u$.

i. Interpret both slopes.
ii. What do they sum to, and what does that mean?
iii. Test whether returns to scale are constant.

::: {.answer}
```{r ff-ex-cobb}
prod <- read.csv("data/india-production.csv")
cobb <- lm(logy ~ logl + logk, data = prod)
round(coef(summary(cobb)), 4)
```

**(i)** A 1% rise in labour is associated with a 0.71% rise in output, holding
capital fixed; a 1% rise in capital with a 0.79% rise, holding labour fixed.
Both are elasticities.

**(ii)** They sum to 1.50. Doubling both inputs would raise output by about 150%
rather than 100% — **increasing returns to scale**.

**(iii)** $H_0: \beta_1 + \beta_2 = 1$. This is a restriction on a *combination*
of coefficients, so neither printed $t$ statistic tests it.

Impose the restriction by substituting $\beta_2 = 1 - \beta_1$ into the model
and rearranging:

$$\ln Y - \ln K = \beta_0 + \beta_1(\ln L - \ln K) + u$$

This is the restricted model. Its residuals are directly comparable with the
unrestricted ones, because subtracting $\ln K$ from both sides moves a known
quantity and leaves the error untouched — so the $F$ statistic of Section
\@ref(mlr-f) can be computed from the two sums of squared residuals.

```{r ff-ex-cobb-f}
restricted <- lm(I(logy - logk) ~ I(logl - logk), data = prod)

ssr_r  <- sum(residuals(restricted)^2)
ssr_ur <- sum(residuals(cobb)^2)
n <- nrow(prod); k <- 2; q <- 1

f_stat <- ((ssr_r - ssr_ur) / q) / (ssr_ur / (n - k - 1))

round(c(ssr_restricted = ssr_r, ssr_unrestricted = ssr_ur,
        F = f_stat, critical_5pct = qf(0.95, q, n - k - 1),
        p_value = 1 - pf(f_stat, q, n - k - 1)), 5)
```

$F = 59.2$ against a critical value of 4.11, so constant returns to scale is
rejected decisively.

::: {.watchout}
`anova(restricted, cobb)` will **not** do this comparison, and will not warn
you. The two models have different dependent variables — $\ln Y - \ln K$ against
$\ln Y$ — so R has no basis for treating them as nested and silently returns
something else. Computing the statistic from the two sums of squared residuals
is safer, and shows what the test is doing.
:::

With 40 annual observations and inputs that trend together, this result is
fragile. Labour and capital are highly collinear over this period, which is
exactly the situation of Section \@ref(mlr-assumptions).
:::

### Concept checks {-}

**9.5** A researcher reports that adding six control variables raised $R^2$ from
0.31 to 0.44 and concludes the new model is better. Assess.

::: {.answer}
The rise in $R^2$ is not evidence of anything: $R^2$ cannot fall when variables
are added, so it would have risen even had all six been random noise.

Adjusted $R^2$ would at least charge for the six parameters, but it too is a
weak criterion.

More importantly, "better" is undefined without a purpose. If the goal is
prediction, some measure of fit is relevant, though one computed on data not
used for fitting. If the goal is estimating an effect, fit is nearly irrelevant
and the question is whether the six controls are confounders, mediators or
colliders — which $R^2$ cannot address.
:::

**9.6** Explain why controlling for occupation in a regression of earnings on
education does not produce "the effect of education net of confounding".

::: {.answer}
Occupation is largely a *consequence* of education, not a common cause of both.
It is a mediator.

Controlling for it compares people who ended up in the same occupation, which
removes the route through which much of education's effect operates. The
resulting coefficient is the direct effect — education's value conditional on
job — not the total effect, and not a confounding-corrected version of it.

The estimate is not biased for what it now measures. It measures something
narrower than what was asked.
:::

**9.7** Two researchers analyse the same data. One reports the level–level model
because its $R^2$ is 0.48; the other reports log–log because its $R^2$ is 0.51.
What is wrong with both arguments?

::: {.answer}
The two $R^2$ values are not comparable. One describes the share of variation
explained in $y$, the other in $\ln y$. These are different quantities, so the
comparison is meaningless regardless of which is larger.

Beyond that, neither researcher has given a reason of the kind that should
decide the question: how the variable is naturally discussed, whether the
relationship is plausibly multiplicative, and what the scatter looks like.
:::

**9.8** Why does an irrelevant variable that is highly correlated with $x_1$
inflate $\mathrm{se}(\hat{\beta}_1)$ even though its own coefficient is zero in
the population?

::: {.answer}
The standard error depends on $(1-R_j^2)$, where $R_j^2$ comes from regressing
$x_1$ on the other regressors. That quantity is about the relationship *among
the explanatory variables* and has nothing to do with the outcome.

A variable strongly related to $x_1$ raises $R_j^2$ whether or not it affects
$y$. By the argument of Section \@ref(mlr-fwl), the coefficient is estimated
from the part of $x_1$ the other regressors cannot explain, and a strongly
correlated variable leaves little of it.
:::

### R exercises {-}

**9.9** Using `wages-india-synthetic.csv`, fit the four functional forms
relating annual earnings to education. Report each slope with its
interpretation in a sentence, and explain why the four $R^2$ values should not
be compared.

::: {.answer}
```{r ff-ex-four, eval = FALSE}
wages <- read.csv("data/wages-india-synthetic.csv")
w <- subset(wages, education > 0)          # log(0) is undefined

forms <- list(
  level_level = lm(annual_earnings ~ education, data = w),
  log_level   = lm(log(annual_earnings) ~ education, data = w),
  level_log   = lm(annual_earnings ~ log(education), data = w),
  log_log     = lm(log(annual_earnings) ~ log(education), data = w))

t(sapply(forms, function(m) c(slope = coef(m)[2],
                              r2    = summary(m)$r.squared)))
```

Level–level: an extra year is worth about ₹5,400. Log–level: about 9.4% more.
Level–log: a 1% rise in years of schooling is worth about slope/100 rupees.
Log–log: a 1% rise in years of schooling raises earnings by the slope, in per
cent.

The four $R^2$ values are not comparable because the first and third explain
variation in earnings while the second and fourth explain variation in log
earnings.

Note also that dropping the zero-education observations changes the sample. The
log forms answer a question about a different group of people from the level
forms — which is the caution in Section \@ref(ff-choosing) arriving in practice.
:::

**9.10** Demonstrate that $R^2$ cannot fall. Take any regression, add a variable
of pure noise, and confirm. Then repeat 500 times and report how often adjusted
$R^2$ rises.

::: {.answer}
```{r ff-ex-r2, eval = FALSE}
students <- read.csv("data/attendance-grades.csv")
set.seed(31)

base <- lm(sem_gpa ~ attendance + cum_gpa, data = students)

results <- replicate(500, {
  students$junk <- rnorm(nrow(students))
  m <- lm(sem_gpa ~ attendance + cum_gpa + junk, data = students)
  c(r2_rose  = summary(m)$r.squared    >= summary(base)$r.squared,
    adj_rose = summary(m)$adj.r.squared > summary(base)$adj.r.squared)
})

rowMeans(results)
```

$R^2$ rises in 100% of trials, as the algebra guarantees.

Adjusted $R^2$ rises in about 31% of them. That is not a flaw: a noise
variable will occasionally correlate with the outcome by chance, and adjusted
$R^2$ only charges a fixed price per parameter. It filters noise on average
without doing so reliably in any single comparison.
:::

**9.11** Reproduce the collider simulation of Section \@ref(ff-controls), then
vary the strength of the collider's dependence on $x$ and on $u$. Which matters
more for the size of the bias?

::: {.answer}
```{r ff-ex-collider, eval = FALSE}
bias_at <- function(a, b, reps = 400, n = 2000) {
  est <- replicate(reps, {
    x <- rnorm(n); u <- rnorm(n)
    y <- 1 * x + u
    s <- a * x + b * u + rnorm(n)
    coef(lm(y ~ x + s))[2]
  })
  round(mean(est) - 1, 4)
}

set.seed(5)
c(weak_both   = bias_at(0.3, 0.3),
  strong_on_x = bias_at(1.2, 0.3),
  strong_on_u = bias_at(0.3, 1.2),
  strong_both = bias_at(1.2, 1.2))
```

The bias requires *both* dependences. If the collider does not depend on $x$,
conditioning on it induces no relationship between $x$ and $u$; if it does not
depend on $u$, there is no $u$ to induce a relationship with.

This mirrors the omitted variable bias formula of Section \@ref(mlr-causal),
which was also a product of two terms and vanished if either was zero.
:::

## Question bank {#ff-question-bank}

### Multiple choice {-}

**QB 9.1** In the model $\ln y = \beta_0 + \beta_1 x + u$, a coefficient of
0.045 means a one-unit rise in $x$ is associated with

a. a 0.045 unit rise in $y$
b. a 4.5% rise in $y$
c. a 0.045% rise in $y$
d. a 4.5 unit rise in $y$

**QB 9.2** A log–log coefficient is

a. always between 0 and 1
b. an elasticity, and unit-free
c. equal to the correlation between the variables
d. interpretable only if both variables are positive integers

**QB 9.3** Adding an explanatory variable to a regression

a. can raise or lower $R^2$
b. can never lower $R^2$
c. can never lower adjusted $R^2$
d. always raises both

**QB 9.4** Controlling for a variable that lies on the causal path between $x$
and $y$

a. removes omitted variable bias
b. estimates the direct rather than the total effect
c. has no effect on the coefficient
d. always increases $R^2$ without changing the estimate

**QB 9.5** Including an irrelevant variable uncorrelated with the other
regressors

a. biases the coefficients downward
b. leaves the coefficients unbiased and barely changes their standard errors
c. leaves the coefficients unbiased but sharply raises their standard errors
d. makes the model unidentifiable

### Short reasoning {-}

**QB 9.6** A study of fertiliser and yield reports a level–level slope of 0.83.
A colleague asks whether that is a large effect. Explain what further
information is needed, and what functional form would have avoided the question.

**QB 9.7** Explain why $R^2$ is a reasonable thing to examine when forecasting
next year's electricity demand, but a poor guide when estimating the effect of a
school meals programme on attendance.

**QB 9.8** A researcher studying the effect of a job training programme on
earnings controls for whether the participant is currently employed. Which of
the three cases in Section \@ref(ff-controls) is this, and what will it do to
the estimate?

**QB 9.9** Why can adjusted $R^2$ be negative? What does a negative value tell
you about the model?

**QB 9.10** In a regression of infant mortality on household income across
districts, a researcher adds "number of hospitals per capita" as a control and
the income coefficient shrinks by half. Give two different causal stories
consistent with that result, one in which the control belongs and one in which
it does not.

**QB 9.11** A dataset records earnings for people who are employed. Explain why
regressing earnings on education using only these observations is a version of
the collider problem, and what it implies for interpreting the coefficient.

**QB 9.12** Explain why the choice between level–log and log–log cannot be made
by comparing $R^2$, but the choice between two different sets of controls in a
log–log model could in principle be informed by it.

### Numerical {-}

**QB 9.13** A log–level regression of earnings on years of experience gives a
coefficient of 0.026. Give both the approximate and exact percentage returns to
one more year, and to ten more years.

**QB 9.14** A model with $n = 150$, $k = 4$ has $\text{SSR} = 240$ and
$\text{SST} = 600$. Compute $R^2$ and adjusted $R^2$.

**QB 9.15** Two nested models on the same 300 observations have $R^2$ of 0.38
with 3 regressors and 0.41 with 7. Compute both adjusted $R^2$ values, and the
$F$ statistic for the four added variables. Do the two criteria agree?

**QB 9.16** A demand equation estimated in log–log form gives a price elasticity
of $-1.4$ with a standard error of 0.15. Test whether demand is unit-elastic,
and state what the sign and magnitude imply about total revenue when price
rises.

## Looking ahead {#ff-looking-ahead}

Every explanatory variable in this book has been a number — years, rupees,
classes attended. Two of them were not, and we handled them by writing 0 and 1
without ever asking what that meant.

The next unit takes indicator variables seriously. It shows how a single
regression can fit different intercepts for different groups, how interaction
terms let it fit different *slopes* as well, and how the $F$ test of Section
\@ref(mlr-f) becomes a formal test of whether two groups obey the same
relationship at all.

That last question — whether one model describes everybody — is where the
machinery of this unit and the previous one comes together.
