# Dummy Variables: Does One Line Fit Everyone? {#dummy-variables}

```{r setup-10, include = FALSE}
source("_common.R")
library(ggplot2)
```

Every explanatory variable so far has been a number. Years of schooling, classes
attended, capital per worker — quantities that can be larger or smaller, and for
which "a one-unit increase" means something.

Much of what economists want to compare is not like that. Rural and urban.
Treated and untreated. Before and after the reform. These are categories, not
quantities, and there is no sense in which one is one unit more than another.

Twice already we have handled such a variable by writing 0 and 1 and moving on.
It is time to ask what that actually did — and then to ask a question the
technique makes possible.

Every regression in this book has fitted **one line to everybody**. One return
to schooling for all workers, one effect of attendance for all students. That is
an assumption, and a strong one. It may be that the relationship differs across
groups, and that fitting a single line describes nobody in particular.

This unit shows how to put groups into a regression, how to let a relationship
differ between them, and how to test whether it does.

## Variables that are not numbers {#dv-motivation}

::: {.question}
How do you put "urban" into an equation?
:::

The wage data record where each worker lives.

```{r dv-look}
wages <- read.csv("data/wages-india-synthetic.csv")
wages$lw <- log(wages$annual_earnings)

table(wages$residence)
```

There is no arithmetic to do here. Urban minus rural is not a number, and no
amount of care about units will make it one.

What we can do is record whether a condition holds.

::: {.keyidea}
A **dummy variable** — also called an indicator, or a binary variable — takes
the value 1 when a condition is true and 0 when it is false.

$$\text{urban}_i =
\begin{cases}
1 & \text{if worker } i \text{ lives in an urban area} \\
0 & \text{otherwise}
\end{cases}$$

It is a number, so a regression can use it. What makes it different from other
numbers is that it takes only two values, so "a one-unit increase" means
switching from one group to the other.
:::

```{r dv-make}
wages$urban <- as.integer(wages$residence == "Urban")
wages$male  <- as.integer(wages$sex == "Male")

head(data.frame(residence = wages$residence, urban = wages$urban))
```

The choice of which category gets the 1 is arbitrary. Coding rural as 1 instead
would flip the sign of the coefficient and change nothing else.

## One dummy: a shift in the intercept {#dv-intercept}

::: {.question}
What does the coefficient on a 0/1 variable mean?
:::

Start with the simplest possible regression: log earnings on the urban dummy and
nothing else.

$$\ln(\text{earnings}) = \beta_0 + \beta_1 \,\text{urban} + u$$

Work out what the model predicts for each group by substituting the only two
values the variable can take.

::: {.theory}
For a rural worker, $\text{urban} = 0$:

$$E[\ln(\text{earnings}) \mid \text{rural}] = \beta_0$$

For an urban worker, $\text{urban} = 1$:

$$E[\ln(\text{earnings}) \mid \text{urban}] = \beta_0 + \beta_1$$

Subtracting the first from the second:

$$\beta_1 = E[\ln(\text{earnings}) \mid \text{urban}]
- E[\ln(\text{earnings}) \mid \text{rural}]$$
:::

::: {.keyidea}
The intercept is the mean of the **base group** — the one coded 0.

The coefficient is the **difference in means** between the two groups.

Nothing about a slope has changed conceptually. It is still the change in $y$
for a one-unit change in $x$; it is only that the single available unit is the
whole distance from one group to the other.
:::

```{r dv-one}
one_dummy <- lm(lw ~ urban, data = wages)
round(coef(summary(one_dummy)), 5)

round(tapply(wages$lw, wages$residence, mean), 5)
```

The intercept, 9.818, is exactly the rural mean. The coefficient, 1.076, is
exactly the gap between the two means.

Because the outcome is in logarithms, Section \@ref(ff-four) tells us how to
read it: urban workers earn about
$`r round(100*(exp(1.075997)-1), 0)`$% more than rural workers. The coefficient
is far above 0.2, so the exact expression is required and the approximation of
108% would be badly wrong.

### Something familiar {-}

Look again at what this regression is doing. It is comparing the mean of one
group with the mean of another and reporting a standard error for the
difference.

That is the two-sample $t$-test of Unit \@ref(two-sample).

```{r dv-ttest}
c(regression_t = coef(summary(one_dummy))[2, 3],
  t_test       = abs(t.test(lw ~ residence, data = wages,
                            var.equal = TRUE)$statistic))
```

Not merely similar. The same number, to every digit R prints.

::: {.keyidea}
**A regression on a single dummy variable is a two-sample $t$-test.**

The pooled two-sample test of Section \@ref(ts-unknown) assumed a common
variance in the two groups; the regression assumes homoskedasticity, which here
says the same thing. Given the same assumption, they are the same procedure
written two ways.
:::

::: {.intuition}
This is worth pausing on, because it reframes everything that came before.

Unit \@ref(two-sample) looked like a self-contained topic with its own formula.
It was a special case of regression all along — the case with one explanatory
variable that happens to be binary.

And regression can do what the $t$-test cannot. Add controls and the comparison
becomes a difference between groups *among people alike in other respects*,
which is precisely what Unit \@ref(multiple-regression) was about.
:::

```{r dv-controlled}
controlled <- lm(lw ~ urban + education + age + male, data = wages)
round(coef(summary(controlled))[, 1:2], 5)
```

The urban premium falls from 1.076 to 0.805 once education, age and sex are held
fixed. Part of the raw gap was the fact that urban workers are better educated —
not a return to living in a city.

## More than two categories {#dv-categories}

::: {.question}
Residence has two categories. What about a variable with seven?
:::

The data also record religion, which takes seven values.

```{r dv-religion-table}
table(wages$religion)
```

The temptation is to code these 1 through 7 and enter that as a variable. It
would run, and it would be meaningless: it would impose that the step from
category 1 to category 2 has the same effect as the step from 6 to 7, and that
category 4 sits midway between 3 and 5. The numbers are labels, not quantities.

The correct approach is one dummy per category — but not quite all of them.

### The dummy variable trap {-}

Suppose we build a dummy for urban and another for rural, and include both.

```{r dv-trap}
wages$rural <- 1 - wages$urban

coef(lm(lw ~ urban + rural, data = wages))
```

`NA` again, and for the reason established in Section \@ref(mlr-assumptions):
$\text{urban} + \text{rural} = 1$ for every worker, which is exactly the
intercept's column. One column is an exact linear combination of the others, so
CLRM7 fails.

::: {.keyidea}
**The dummy variable trap.** With $g$ categories, include $g-1$ dummies, not
$g$.

The omitted category is the **base group**. Every coefficient is a comparison
with it, and the intercept describes it.

This is not a statistical subtlety but the same arithmetic as Section
\@ref(mlr-assumptions): a full set of category dummies always sums to one, and
the regression already has a column of ones.
:::

An alternative is to drop the intercept instead and keep all $g$ dummies, in
which case each coefficient is a group mean rather than a difference. Both are
defensible; what is not defensible is keeping both the intercept and all $g$
dummies.

### Letting R do it {-}

R handles this automatically if the variable is a factor.

```{r dv-religion}
religion_fit <- lm(lw ~ factor(religion), data = wages)
round(coef(summary(religion_fit))[, 1:2], 4)
```

Six coefficients for seven categories. The missing one is Buddhist, which R
chose because it comes first alphabetically — that is the only reason.

::: {.watchout}
The base group is arbitrary, and the default is rarely the best choice.

Here it is a category with 174 observations out of 20,000. Every coefficient in
the table is a comparison with a small group, and every standard error is
inflated by that group's imprecision. Nothing is *wrong* — the fitted values,
$R^2$ and $F$ statistic are identical whatever the base — but the table is
harder to read than it needs to be.

A large or otherwise natural category makes a better reference.
:::

```{r dv-relevel}
wages$religion <- relevel(factor(wages$religion), ref = "Hindu")
round(coef(summary(lm(lw ~ religion, data = wages)))[, 1:2], 4)
```

With the largest group as the base, each coefficient is a comparison with it and
the standard errors are smaller.

::: {.intuition}
Notice that changing the base group changed every coefficient without changing
the model at all. The fit is identical; only the comparison being reported has
moved.

A coefficient on a dummy is never meaningful on its own. It is meaningful
relative to a base group, and a reader who does not know which group that is
cannot interpret the table.
:::

## Interactions: letting the slope differ {#dv-interactions}

::: {.question}
A dummy shifts the line up or down. What if the groups differ in more than that?
:::

Everything so far has moved the intercept. The model

$$\ln(\text{earnings}) = \beta_0 + \beta_1\,\text{educ} + \beta_2\,\text{urban} + u$$

fits two lines, one for each group, but forces them to be **parallel**. Urban
workers earn more at every level of education, by the same amount.

That is an assumption, and there is no reason to believe it. The return to a
year of schooling might itself differ between rural and urban labour markets.

To allow that, add the product of the two variables.

$$\ln(\text{earnings}) = \beta_0 + \beta_1\,\text{educ} + \beta_2\,\text{urban}
+ \beta_3\,(\text{educ} \times \text{urban}) + u$$

The product is called an **interaction term**.

::: {.theory}
Substitute the two possible values of the dummy, as before.

For rural workers, $\text{urban} = 0$, so both terms containing it vanish:

$$E[\ln(\text{earnings})] = \beta_0 + \beta_1\,\text{educ}$$

For urban workers, $\text{urban} = 1$:

$$
\begin{aligned}
E[\ln(\text{earnings})]
&= \beta_0 + \beta_1\,\text{educ} + \beta_2 + \beta_3\,\text{educ} \\[4pt]
&= (\beta_0 + \beta_2) + (\beta_1 + \beta_3)\,\text{educ}
\end{aligned}
$$

So the model fits two completely separate lines:

- $\beta_2$ is the difference in **intercepts**;
- $\beta_3$ is the difference in **slopes**.
:::

```{r dv-interaction}
interacted <- lm(lw ~ education * urban, data = wages)
round(coef(summary(interacted)), 5)
```

R expands `education * urban` into `education + urban + education:urban`.

Read the four numbers as two lines:

```{r dv-two-lines}
b <- coef(interacted)

round(rbind(rural = c(intercept = b[1],           slope = b[2]),
            urban = c(intercept = b[1] + b[3],    slope = b[2] + b[4])), 5)
```

For rural workers a year of schooling is worth about 8.0%; for urban workers,
about 6.9%. The interaction coefficient is $-0.0109$ with a $t$ statistic of
$-3.08$, so the difference is more than sampling variation would produce.

::: {.watchout}
**With an interaction in the model, $\beta_1$ is not "the effect of education".**

It is the effect of education *in the base group only* — here, rural workers.
The urban return requires adding the interaction coefficient.

This is the same warning as the squared term of Section \@ref(ff-choosing), and
for the same reason: once a variable appears twice, no single coefficient
answers the question on its own.
:::

```{r dv-lines-plot, echo = FALSE, fig.height = 3.8, cache = TRUE, fig.cap = "Log earnings against education for rural and urban workers, with a 2,000-worker sample drawn for legibility. The solid lines are fitted separately to each group; the dashed line is the single pooled regression that ignores residence. The two group lines differ in level and, less obviously, in steepness -- the urban line is slightly flatter. The pooled line describes neither group."}
set.seed(4)
shown <- wages[sample(nrow(wages), 2000), ]

pooled_fit <- lm(lw ~ education, data = wages)
grid <- expand.grid(education = 0:15, urban = 0:1)
grid$lw <- predict(interacted, grid)
grid$Residence <- ifelse(grid$urban == 1, "Urban", "Rural")

pool <- data.frame(education = 0:15)
pool$lw <- predict(pooled_fit, pool)

ggplot(shown, aes(education, lw)) +
  geom_jitter(aes(colour = residence), width = 0.25, alpha = 0.16, size = 0.7) +
  geom_line(data = pool, aes(education, lw), colour = "grey25",
            linetype = "dashed", linewidth = 0.8) +
  geom_line(data = grid, aes(education, lw, colour = Residence),
            linewidth = 1.1) +
  scale_colour_manual(values = c(Rural = book_palette$reject,
                                 Urban = book_palette$accent)) +
  labs(x = "Years of education", y = "Log annual earnings", colour = NULL) +
  theme_book() +
  theme(legend.position = "top")
```

::: {.intuition}
An interaction is how a regression says "it depends".

Without one, the model insists that education is worth the same everywhere and
that living in a city is worth the same at every level of schooling. With one,
each answer is allowed to depend on the other.

Note that the dependence runs both ways, and reading it either way is
legitimate. The return to education differs by residence; equivalently, the
urban premium differs by education. The same coefficient says both, because
$\text{educ} \times \text{urban}$ does not know which variable we regard as
primary.
:::

## Do two groups share the same line? {#dv-chow}

::: {.question}
The intercepts differ and the slopes differ. Is one regression adequate for
both groups, or not?
:::

We now have two things that might differ between groups, and a question that
asks about both at once:

$$H_0 : \beta_2 = 0 \;\text{ and }\; \beta_3 = 0$$

Under this null, urban and rural workers share the same intercept *and* the same
slope — a single line describes everyone.

Section \@ref(mlr-f) established that a claim about two coefficients at once
needs an $F$ test, not two $t$ tests. Applied to a group split, this test has a
name of its own: the **Chow test**.

### Two ways to see it {-}

The first way is the one the test is usually taught with. Fit the regression
three times — once on the pooled sample, once on each group separately — and ask
whether splitting the sample improved the fit by more than chance.

```{r dv-chow-hand}
pooled <- lm(lw ~ education, data = wages)
rural  <- lm(lw ~ education, data = subset(wages, urban == 0))
urban  <- lm(lw ~ education, data = subset(wages, urban == 1))

ssr <- function(m) sum(residuals(m)^2)

round(c(pooled = ssr(pooled), rural = ssr(rural), urban = ssr(urban),
        separate_total = ssr(rural) + ssr(urban)), 2)
```

Separate regressions must fit at least as well as a pooled one, for the reason
given in Section \@ref(mlr-f): the pooled model is the separate model with the
differences restricted to zero. The question is whether the improvement is
larger than restricting nothing would produce by chance.

::: {.theory}
With $k$ parameters in each group's regression and $n_1$, $n_2$ observations,

$$F = \frac{\big[\text{SSR}_P - (\text{SSR}_1 + \text{SSR}_2)\big] \big/ k}
{(\text{SSR}_1 + \text{SSR}_2) \big/ (n_1 + n_2 - 2k)}$$

The numerator is the improvement per restriction; the denominator estimates
$\sigma^2$ from the unrestricted model. The degrees of freedom are $k$ and
$n_1 + n_2 - 2k$.
:::

```{r dv-chow-f}
n <- nrow(wages); k <- 2

f_chow <- ((ssr(pooled) - (ssr(rural) + ssr(urban))) / k) /
          ((ssr(rural) + ssr(urban)) / (n - 2 * k))

round(c(F = f_chow, critical_5pct = qf(0.95, k, n - 2 * k),
        p_value = 1 - pf(f_chow, k, n - 2 * k)), 4)
```

$F = 1198.5$ against a critical value of 3.00. The rural and urban wage
equations are not the same.

### The second way is the interaction model {-}

The other route is to fit the interacted model of Section
\@ref(dv-interactions) and test its two extra coefficients jointly.

```{r dv-chow-anova}
anova(pooled, interacted)
```

The same $F$ statistic and the same degrees of freedom. R has rounded it to
1199 in the table; the two are the same number.

::: {.keyidea}
**Running separate regressions and fitting one interacted regression are the
same thing.**

```{r dv-chow-identity}
round(c(separate_ssr  = ssr(rural) + ssr(urban),
        interacted_ssr = ssr(interacted)), 4)
```

The sums of squared residuals are identical, because the two models fit
identical lines to identical data. Section \@ref(dv-interactions) already showed
the coefficients matching: the interacted model's base coefficients are the
rural regression, and adding the interaction terms gives the urban one.
:::

::: {.intuition}
Given that they are the same test, the interaction approach is usually the more
useful one.

Splitting the sample gives a yes-or-no answer about whether the groups differ.
The interacted regression gives that *and* tells you which coefficient differs,
by how much, and with what standard error — here, that the intercepts differ
enormously and the slopes differ by $-0.011$.

The split version also cannot easily hold other variables fixed across the
comparison, and it estimates a separate error variance for each group where the
interacted model pools them.
:::

### What rejecting does and does not mean {-}

::: {.watchout}
The Chow test says the groups differ. It does not say why.

Urban and rural workers differ in education, occupation, industry, and much
else. A rejection is consistent with a genuinely different labour market and
equally consistent with the two groups differing in variables the model omits —
the omitted variable problem of Section \@ref(mlr-causal), which does not go
away because a group indicator was added.
:::

There is also a failure mode in the other direction. With 20,000 observations,
this test detects a slope difference of 0.011 — a difference that may be too
small to matter for any decision anyone would make.

::: {.keyidea}
A significant Chow test means the groups are *detectably* different, not
*importantly* different.

That distinction is the one from Section \@ref(os-errors): with enough data, any
difference that is not exactly zero will eventually be rejected. Whether it
matters is a question about economics, and the test cannot answer it.
:::

## Summary {#dv-summary}

::: {.keyidea}
- A **dummy variable** takes the value 1 when a condition holds and 0 otherwise.
  It converts a category into something a regression can use, and "a one-unit
  increase" becomes "switching groups".
- With one dummy and nothing else, the intercept is the **base group mean** and
  the coefficient is the **difference in means**. On the wage data these are
  9.818 and 1.076, matching the two group means exactly.
- **A regression on a single dummy is the two-sample $t$-test** of Unit
  \@ref(two-sample). The $t$ statistics agree to every digit. Unit
  \@ref(two-sample) was a special case of regression all along.
- Adding controls turns that raw comparison into a comparison among people alike
  in other respects. The urban premium falls from 1.076 to 0.805.
- With $g$ categories, include $g-1$ dummies. Including all $g$ alongside an
  intercept is the **dummy variable trap** — a violation of CLRM7, since the
  dummies sum to the intercept's column of ones.
- The omitted category is the **base group**. Changing it changes every
  coefficient and no part of the fit. A dummy coefficient means nothing until
  the reader knows the base.
- Never code an unordered category as 1, 2, 3. That imposes equal spacing and an
  ordering the variable does not have.
- An **interaction term** $x \times D$ lets the *slope* differ across groups.
  Without one the two group lines are forced parallel.
- In the interacted model, $\beta_2$ is the difference in intercepts and
  $\beta_3$ the difference in slopes, so the two groups get entirely separate
  lines: 8.0% per year of schooling in rural areas against 6.9% in urban.
- With an interaction present, the coefficient on $x$ alone is the effect **in
  the base group**, not the effect of $x$.
- The **Chow test** asks whether one line fits both groups:
  $H_0: \beta_2 = \beta_3 = 0$. It is the $F$ test of Section \@ref(mlr-f)
  applied to a group split.
- Separate regressions and one interacted regression are the same model. Their
  sums of squared residuals are identical, and both give $F = 1198.5$ here. The
  interacted version is more informative and easier to control.
- A rejection means the groups are detectably different, not that the difference
  matters, and not that the difference is causal.
:::

## Practice problems {#dv-practice}

::: {.yourturn}
Work each one before opening the solution. The
[question bank](#dv-question-bank) that follows has no answers.
:::

### Numerical problems {-}

**10.1** A regression of monthly household expenditure (₹) on a dummy for urban
residence gives

$$\widehat{\text{expenditure}} = 8{,}400 + 3{,}150\,\text{urban}$$

i. What is average expenditure among rural households?
ii. Among urban households?
iii. If the dummy had been coded 1 for rural instead, what would the two
     coefficients be?

::: {.answer}
**(i)** ₹8,400. With `urban = 0` only the intercept remains.

**(ii)** $8{,}400 + 3{,}150 = ₹11{,}550$.

**(iii)** The intercept becomes the urban mean, ₹11,550, and the coefficient
becomes $8{,}400 - 11{,}550 = -₹3{,}150$. The fitted values, residuals and $R^2$
are all unchanged — only the reported comparison has reversed.
:::

**10.2** A wage regression uses four occupational categories: agriculture,
manufacturing, services and government.

i. How many dummies should be included alongside an intercept?
ii. With agriculture as the base group, interpret a coefficient of 0.32 on
    services in a log-wage regression.
iii. What happens to that coefficient if government becomes the base group?

::: {.answer}
**(i)** Three. Including four alongside an intercept is the dummy variable trap.

**(ii)** Service workers earn about
$100(e^{0.32}-1) = `r round(100*(exp(0.32)-1), 1)`\%$ more than agricultural
workers. As the coefficient exceeds 0.2, the exact expression is needed.

**(iii)** It changes, because it now measures services against government rather
than against agriculture. The difference between services and agriculture is
still recoverable from the new table — it is the difference between their two
coefficients.
:::

**10.3** From a log-earnings regression with an interaction:

$$\widehat{\ln w} = 9.2 + 0.081\,\text{educ} + 0.74\,\text{urban}
- 0.015\,(\text{educ} \times \text{urban})$$

i. Write the fitted line for rural workers and for urban workers.
ii. What is the return to a year of schooling in each group?
iii. At what level of education do the two groups have equal predicted earnings?

::: {.answer}
**(i)** Rural ($\text{urban}=0$): $\widehat{\ln w} = 9.2 + 0.081\,\text{educ}$.

Urban ($\text{urban}=1$):
$\widehat{\ln w} = (9.2 + 0.74) + (0.081 - 0.015)\,\text{educ}
= 9.94 + 0.066\,\text{educ}$.

**(ii)** About 8.1% per year in rural areas, 6.6% in urban areas.

**(iii)** Set the two equal:

$$9.2 + 0.081e = 9.94 + 0.066e
\quad\Longrightarrow\quad
0.015e = 0.74
\quad\Longrightarrow\quad
e = `r round(0.74/0.015, 1)`$$

At about 49 years of education — far beyond anything observed. Within any real
range of schooling the urban line lies above the rural one, and the negative
interaction narrows the gap without closing it. This is the same lesson as the
turning point in Section \@ref(ff-choosing): a crossing point outside the data
is an artefact of extrapolation, not a finding.
:::

**10.4** A Chow test on rural and urban income–consumption regressions gives
$\text{SSR}_P = 38{,}710.67$, $\text{SSR}_R = 8{,}661.60$ and
$\text{SSR}_U = 10{,}107.18$, with $n_R = n_U = 150$.

i. Compute the $F$ statistic.
ii. State the degrees of freedom and conclude at the 5% level.

::: {.answer}
**(i)** $k = 2$ (intercept and slope).

$$F = \frac{[38{,}710.67 - (8{,}661.60 + 10{,}107.18)]/2}
{(8{,}661.60 + 10{,}107.18)/(300 - 4)}
= \frac{19{,}941.89/2}{18{,}768.78/296}
= \frac{9{,}970.95}{63.41} = `r round((38710.67-18768.78)/2/(18768.78/296), 1)`$$

**(ii)** Degrees of freedom 2 and 296; the 5% critical value is about
`r round(qf(0.95, 2, 296), 2)`. Since $157.25$ vastly exceeds it, reject: rural
and urban households do not share the same income–consumption relationship.
:::

### Concept checks {-}

**10.5** A student includes dummies for all five states in a sample, plus an
intercept, and reports that R returned `NA` for the last one. Explain what
happened and give two ways to fix it.

::: {.answer}
The five dummies sum to 1 for every observation, which is exactly the
intercept's column. One column is an exact linear combination of the others, so
CLRM7 fails and there is no unique solution — the situation of Section
\@ref(mlr-assumptions).

Two fixes: drop one dummy and treat that state as the base group, or drop the
intercept and keep all five, in which case each coefficient is a state mean
rather than a difference.
:::

**10.6** Explain why entering a five-category variable as the numbers 1 to 5 is
usually wrong, and describe the one circumstance in which it might be defensible.

::: {.answer}
It imposes that the categories are equally spaced and ordered — that moving from
1 to 2 has the same effect as moving from 4 to 5, and that 3 lies between 2 and
4. For unordered categories such as religion or state, none of that is
meaningful.

It might be defensible if the categories are genuinely ordered *and* roughly
equally spaced on the scale that matters — for example, five income bands of
equal width. Even then, entering dummies and inspecting whether the coefficients
rise evenly is the safer choice, because it tests the assumption instead of
imposing it.
:::

**10.7** In a regression of log wages on experience, a female dummy, and their
interaction, the coefficient on experience is 0.043 and the interaction
coefficient is $-0.011$. A colleague reports "the return to experience is 4.3%".
What is wrong, and what are the two correct figures?

::: {.answer}
0.043 is the return to experience **in the base group** — men, since the dummy
marks women. It is not the return for the sample as a whole.

The two correct figures are about 4.3% per year for men and
$0.043 - 0.011 = 0.032$, about 3.2% per year, for women.
:::

**10.8** A Chow test on 40,000 observations rejects at $p < 0.001$, but the two
estimated slopes are 0.0512 and 0.0524. What should be reported?

::: {.answer}
That the groups are statistically distinguishable and economically
indistinguishable.

The slopes differ by roughly one part in forty. With 40,000 observations the
test has enough power to detect differences far too small to affect any
decision, exactly as Section \@ref(os-errors) described. Reporting only the
$p$-value would give a seriously misleading impression; reporting both numbers
alongside it does not.
:::

### R exercises {-}

**10.9** Using `wages-india-synthetic.csv`, confirm that a regression on a
single dummy reproduces the two-sample $t$-test for the male–female earnings
gap. Report the coefficient, the two group means and both $t$ statistics.

::: {.answer}
```{r dv-ex-ttest, eval = FALSE}
wages <- read.csv("data/wages-india-synthetic.csv")
wages$lw   <- log(wages$annual_earnings)
wages$male <- as.integer(wages$sex == "Male")

fit <- lm(lw ~ male, data = wages)

c(coefficient = coef(fit)[2],
  female_mean = mean(wages$lw[wages$male == 0]),
  male_mean   = mean(wages$lw[wages$male == 1]),
  reg_t       = coef(summary(fit))[2, 3],
  ttest_t     = abs(t.test(lw ~ sex, data = wages, var.equal = TRUE)$statistic))
```

The intercept is the female mean, the coefficient is the gap, and the two $t$
statistics agree exactly.

Note what this does *not* establish. The gap is a difference in average log
earnings between two groups; it is not a measure of discrimination, since men
and women in these data differ in education, age and residence. Adding those as
controls changes the question to a narrower one, and Section \@ref(ff-controls)
is the reason even that would need an argument.
:::

**10.10** Using `student-survey-sleep.csv`, test whether the relationship
between social media hours and sleep differs between urban and rural students.
Fit the interacted model, interpret each coefficient, and carry out the Chow
test.

::: {.answer}
```{r dv-ex-sleep, eval = FALSE}
sleep <- read.csv("data/student-survey-sleep.csv")

pooled     <- lm(sleep ~ socialmediahours, data = sleep)
interacted <- lm(sleep ~ socialmediahours * residence, data = sleep)

round(coef(summary(interacted)), 4)
anova(pooled, interacted)
```

Nothing is significant. The interaction coefficient is 0.068 with a standard
error of 0.237, and the joint $F$ test does not come close to rejecting.

The honest conclusion is that these 106 students provide no evidence that the
relationship differs by residence — not that it is the same. With 32 rural
students the test has very little power, so failing to reject was close to
guaranteed whatever the truth. This is Section \@ref(os-errors) in practice, and
it is the opposite failure mode from the wage data, where 20,000 observations
made a trivial difference significant.
:::

**10.11** Show that the base group affects no part of the fit. Regress log
earnings on religion using two different base categories and compare the fitted
values, $R^2$ and $F$ statistic.

::: {.answer}
```{r dv-ex-base, eval = FALSE}
wages$rel <- factor(wages$religion)

fit_a <- lm(lw ~ rel, data = wages)
fit_b <- lm(lw ~ relevel(rel, ref = "Hindu"), data = wages)

c(max_fitted_difference = max(abs(fitted(fit_a) - fitted(fit_b))),
  r2_a = summary(fit_a)$r.squared, r2_b = summary(fit_b)$r.squared,
  f_a  = summary(fit_a)$fstatistic[1],
  f_b  = summary(fit_b)$fstatistic[1])
```

The fitted values agree to machine precision and both summary statistics are
identical. Only the coefficient table differs, because only the comparison being
reported has changed.
:::

**10.12** Fit a model allowing the return to education to differ by sex *and* by
residence simultaneously. Report the four implied slopes and comment on whether
the three-way structure is worth the complexity.

::: {.answer}
```{r dv-ex-two-way, eval = FALSE}
m <- lm(lw ~ education * male + education * urban, data = wages)
round(coef(summary(m)), 5)

b <- coef(m)
outer_slopes <- c(
  rural_female = b["education"],
  rural_male   = b["education"] + b["education:male"],
  urban_female = b["education"] + b["education:urban"],
  urban_male   = b["education"] + b["education:male"] + b["education:urban"])
round(outer_slopes, 5)
```

This specification allows each dummy to shift the education slope but assumes
those shifts add — the urban adjustment is the same for men and women. A full
three-way interaction would relax that too, at the cost of more parameters and a
table few readers will follow.

The judgement is the one from Section \@ref(ff-irrelevant): every additional
interaction consumes variation and widens standard errors. Interactions should
be included because a question requires them, not because they are available.
:::

## Question bank {#dv-question-bank}

### Multiple choice {-}

**QB 10.1** In the regression $y = \beta_0 + \beta_1 D + u$ where $D$ is a dummy,
$\beta_0$ is

a. the mean of $y$ in the whole sample
b. the mean of $y$ in the group with $D = 0$
c. the mean of $y$ in the group with $D = 1$
d. the difference between the two group means

**QB 10.2** A categorical variable with six categories requires

a. six dummies and an intercept
b. five dummies and an intercept
c. six dummies and no intercept, or five and an intercept
d. one variable coded 1 to 6

**QB 10.3** In $y = \beta_0 + \beta_1 x + \beta_2 D + \beta_3(xD) + u$, the slope
for the group with $D = 1$ is

a. $\beta_1$ &nbsp;&nbsp; b. $\beta_3$ &nbsp;&nbsp; c. $\beta_1 + \beta_3$
&nbsp;&nbsp; d. $\beta_1 + \beta_2 + \beta_3$

**QB 10.4** A regression of $y$ on a single dummy variable is equivalent to

a. a one-sample $t$-test
b. a paired $t$-test
c. a two-sample $t$-test assuming equal variances
d. a chi-squared test

**QB 10.5** The Chow test compares

a. two coefficients within one regression
b. a pooled regression against separate regressions for each group
c. two different dependent variables
d. $R^2$ across two models with different sample sizes

### Short reasoning {-}

**QB 10.6** Explain why including a dummy for every category alongside an
intercept is an arithmetic problem rather than a statistical one.

**QB 10.7** A researcher reports that switching the base group "changed the
results". What has actually changed, and what has not?

**QB 10.8** Why does the interacted regression give the same $F$ statistic as
fitting two separate regressions? State what the interacted version tells you
that the split does not.

**QB 10.9** A study interacts a treatment dummy with baseline income and finds a
significant interaction. Explain what that means in words, and give the two
equivalent readings of the coefficient.

**QB 10.10** Explain why a large sample makes a Chow test more likely to reject
even when the two groups are economically almost identical.

**QB 10.11** In a regression of crop yield on rainfall with an irrigation dummy
and its interaction, the interaction coefficient is negative. Interpret this,
and say why it is economically plausible.

**QB 10.12** A colleague argues that since a dummy regression is just a
$t$-test, nothing is gained by using regression. Give two things regression can
do that the $t$-test cannot.

### Numerical {-}

**QB 10.13** A log-wage regression on a public-sector dummy gives a coefficient
of 0.28. Compute the exact percentage premium and explain why $28\%$ would be
the wrong figure to report.

**QB 10.14** For $\widehat{y} = 40 + 6x + 15D - 2(xD)$, compute the predicted
value for both groups at $x = 5$ and $x = 10$, and find the value of $x$ at which
the groups are predicted to be equal.

**QB 10.15** A Chow test on two groups of 80 and 120 observations, with a
regression containing an intercept and three slopes, gives
$\text{SSR}_P = 940$, $\text{SSR}_1 = 310$ and $\text{SSR}_2 = 505$. Compute $F$
and state its degrees of freedom.

**QB 10.16** Three regional dummies are entered with the North as base, giving
coefficients of $-0.14$ (South), $0.09$ (East) and $-0.22$ (West) on log output.
What is the predicted difference in log output between the South and the West,
and which region has the highest predicted output?

## Looking ahead {#dv-looking-ahead}

The relationships in this unit were allowed to differ across groups, but every
observation was still a separate individual, observed once.

A great deal of economic data is not like that. The same states are observed
year after year; the same households are revisited; the same firms report
annually. Such data carry information that a single cross-section cannot,
because a unit observed twice can be compared **with itself**.

That comparison is the closest this book will come to the thought experiment of
Section \@ref(sr-motivation) — the same student, observed twice, differing only
in attendance. It cannot be arranged in a cross-section at all. It can sometimes
be arranged when the same unit is observed more than once, and doing so removes
every difference between units that stays fixed over time, whether or not it was
ever measured.

That is the subject of the next unit, and it is the first genuinely new answer
to the causal problem since Unit \@ref(multiple-regression).
