# Functional Forms {#functional-forms}

<!-- PARKED. Written for Unit 7 and lifted out on 2026-08-05: functional form
     belongs later in the book, once multiple regression is in place and
     interpretation has more to do. Not in _bookdown.yml.

     Depends on: cars <- read.csv("data/car-mileage.csv") and
     wages <- read.csv("data/wages-india-synthetic.csv"), plus the standard
     _common.R setup. The log-transformation material at the end arrived here
     from the heteroskedasticity section and is framed as the alternative
     response to a fan-shaped residual plot.

     QB 7.15 in Unit 7 asked for a log-log regression on india-production.csv
     and was removed at the same time; it should return alongside this. -->

```{r setup-ff, include = FALSE}
source("_common.R")
library(ggplot2)
cars  <- read.csv("data/car-mileage.csv")
wages <- read.csv("data/wages-india-synthetic.csv")
```

## Functional forms {#ols-forms}

::: {.question}
A slope of $-0.006$ km per litre per kilogram. Is that a lot?
:::

The regression of fuel efficiency on weight in Section \@ref(ols-practice)
returned a slope of $-0.00601$. Nothing about that number tells you whether the
effect is large, because it depends entirely on the units chosen. Measure weight
in tonnes and the slope becomes $-6.01$. Measure efficiency in litres per
hundred kilometres and the sign flips.

Two problems follow, and one transformation solves both.

**Units are arbitrary.** A coefficient in kilograms and one in tonnes describe
the same relationship, but only one of them is easy to talk about. We would
prefer a measure that does not change when someone changes their mind about units.

**Straight lines are not always the right shape.** Section \@ref(sr-prf) showed
that the conditional mean of GPA against attendance was not exactly linear, and
took a line as an approximation. Sometimes a better approximation is available at
no cost, by transforming a variable before fitting.

Taking logarithms addresses both.

### The four forms {-}

Either variable may be entered in levels or in logarithms, giving four
combinations. The interpretation of $\beta_1$ changes in each.

Table: (\#tab:ols-forms) The four functional forms and what the slope means in each.

| Model | Equation | Interpretation of $\beta_1$ |
|:--|:--|:--|
| Level–level | $y = \beta_0 + \beta_1 x + u$ | A one-unit rise in $x$ changes $y$ by $\beta_1$ **units** |
| Log–level | $\ln y = \beta_0 + \beta_1 x + u$ | A one-unit rise in $x$ changes $y$ by $100\beta_1$ **per cent** |
| Level–log | $y = \beta_0 + \beta_1 \ln x + u$ | A 1% rise in $x$ changes $y$ by $\beta_1/100$ **units** |
| Log–log | $\ln y = \beta_0 + \beta_1 \ln x + u$ | A 1% rise in $x$ changes $y$ by $\beta_1$ **per cent** |

The last is an **elasticity**, and it is unit-free: it does not matter whether
weight is in kilograms or tonnes, or price in rupees or lakhs.

::: {.theory}
All four interpretations follow from one fact of calculus. For small changes,

$$\Delta \ln y \;\approx\; \frac{\Delta y}{y}$$

that is, a change in the logarithm is approximately a *proportional* change in
the variable. Multiplying by 100 turns it into a percentage.

**Log–level.** Differentiating $\ln y = \beta_0 + \beta_1 x$ with respect to $x$,

$$\frac{1}{y}\frac{dy}{dx} = \beta_1
\qquad\Longrightarrow\qquad
\frac{dy/y}{dx} = \beta_1$$

so a one-unit rise in $x$ raises $y$ by a proportion $\beta_1$, which is
$100\beta_1$ per cent.

**Level–log.** From $y = \beta_0 + \beta_1 \ln x$,

$$dy = \beta_1 \, d(\ln x) = \beta_1 \frac{dx}{x}$$

A 1% rise in $x$ means $dx/x = 0.01$, so $dy = 0.01\beta_1 = \beta_1/100$ units.

**Log–log.** From $\ln y = \beta_0 + \beta_1 \ln x$,

$$\frac{dy/y}{dx/x} = \beta_1$$

which is the definition of an elasticity: the ratio of the proportional change in
$y$ to the proportional change in $x$.
:::

### Fuel efficiency, in elasticity form {-}

```{r ols-forms-cars}
cars <- read.csv("data/car-mileage.csv")

level_level <- lm(mileage_kmpl ~ weight_kg, data = cars)
log_log     <- lm(log(mileage_kmpl) ~ log(weight_kg), data = cars)

round(rbind(level_level = c(slope = coef(level_level)[2],
                            r_squared = summary(level_level)$r.squared),
            log_log     = c(slope = coef(log_log)[2],
                            r_squared = summary(log_log)$r.squared)), 4)
```

The log–log slope is $-0.825$: **a car that is 1% heavier is about 0.83% less
fuel-efficient.** That is a sentence anyone can evaluate, and it would be
unchanged if weight were recorded in tonnes.

The fit also improves, from $R^2 = 0.65$ to $0.71$. That is a modest gain rather
than a dramatic one, and it is worth being clear that the case for the log form
here rests mainly on interpretation rather than on fit. Still, the direction is
not an accident: a constant *proportional* penalty for weight describes cars
better than the constant absolute penalty the level–level model imposed.

```{r ols-forms-plot, echo = FALSE, fig.height = 3.4, fig.cap = "The same 74 cars in levels and in logarithms. Both look broadly linear; the logarithmic version is a little tighter around the line, which is the improvement in R-squared from 0.65 to 0.71. The gain here is real but modest -- the case for the log form rests mainly on interpretation."}
d1 <- data.frame(x = cars$weight_kg, y = cars$mileage_kmpl, panel = "Levels")
d2 <- data.frame(x = log(cars$weight_kg), y = log(cars$mileage_kmpl),
                 panel = "Logarithms")
both <- rbind(d1, d2)

ggplot(both, aes(x, y)) +
  geom_point(alpha = 0.55, colour = book_palette$ink, size = 1.3) +
  geom_smooth(method = "lm", se = FALSE, colour = book_palette$accent,
              linewidth = 0.9) +
  facet_wrap(~ panel, scales = "free") +
  labs(x = NULL, y = NULL) +
  theme_book()
```

::: {.watchout}
This is where the warning about linearity belongs.

A great deal of weak empirical work consists of fitting a straight line to a
relationship that is not straight, and reporting the slope as though it described
the whole range. The residual plot of Section \@ref(ols-hetero) catches this too:
a curved band of residuals means the functional form is wrong, not that the
variance is non-constant.

Fitting a line is a choice. Check it.
:::

### The return to schooling {-}

The log–level form is the workhorse of labour economics, because wages are
naturally discussed in percentages.

```{r ols-forms-wages}
wages <- read.csv("data/wages-india-synthetic.csv")

fit_w <- lm(log(annual_earnings) ~ education, data = wages)
round(coef(summary(fit_w)), 5)
```

The slope of 0.0942 says that **one more year of schooling is associated with
about 9.4% higher annual earnings.** Multiply by 100 and read as a percentage;
that is the whole rule.

::: {.watchout}
Two cautions on that number.

It is a simple regression, so it carries every confounder discussed in Section
\@ref(ols-causal). People with more schooling differ in family background and
much else, and the coefficient absorbs all of it. Adding controls in the next
unit lowers it appreciably.

And the percentage reading is an approximation that degrades for large changes.
For a coefficient of 0.0942 the exact figure is
$100(e^{0.0942} - 1) = `r round(100*(exp(0.0942)-1), 2)`$%, which is close
enough. For a coefficient of 0.7 the approximation says 70% and the truth is
$`r round(100*(exp(0.7)-1), 0)`$%. Above about 0.2, use the exact expression.
:::

### When the variable is the problem, not the variance {-}

Robust standard errors accept the heteroskedasticity and correct for it. The
other response is to ask why it is there.

Earnings are a highly skewed variable, and skewed variables often produce
fan-shaped residuals simply because their upper tail is long. Taking logarithms
compresses that tail.

```{r ols-het-logs}
fit_log <- lm(log(annual_earnings) ~ education, data = wages)

q_log <- cut(fitted(fit_log), quantile(fitted(fit_log), 0:4/4),
             include.lowest = TRUE)
round(tapply(residuals(fit_log), q_log, sd), 4)
```

The spread is now essentially identical across quartiles — a ratio of 1.01
against 3.61 in levels. The heteroskedasticity was not a deep feature of the
world; it was a consequence of measuring earnings in rupees rather than in
proportional terms.

This is the second thing a fan-shaped residual plot can be telling you. Section
\@ref(ols-hetero) treated it as a fact about the variance to be corrected with
robust standard errors. Sometimes it is instead a fact about the *scale* of the
variable, and changing the scale removes it entirely.

Robust standard errors would have accepted the heteroskedasticity and widened the
intervals around it. Taking logs asks why it was there.


### Choosing a form {-}

::: {.keyidea}
Three considerations, in order of importance.

**What is the natural way to talk about the variable?** Wages, prices and GDP are
discussed in percentages, so logs. Test scores, class attendance and family size
are discussed in units, so levels.

**Does the relationship look multiplicative?** If a proportional change in $x$
plausibly produces a proportional change in $y$ — as with weight and fuel use —
logs will fit better and mean more.

**Does the scatter suggest it?** Skewed variables often straighten out in logs,
and heteroskedasticity often shrinks at the same time, since logs compress the
upper tail.

What should *not* drive the choice is which version gives the larger $R^2$ or the
smaller $p$-value. Trying transformations until something is significant produces
findings that will not replicate.
:::

::: {.watchout}
**Logarithms are undefined at zero and for negative numbers.**

Earnings data routinely contain zeros — people who did not work. Taking logs
silently deletes them, and since those are rarely a random subset of the sample,
the remaining regression answers a different question from the one asked.

```{r ols-forms-zero}
sum(wages$annual_earnings <= 0)
```

There are none here, which is a property of this simulated dataset rather than of
earnings data in general. With real data, check before transforming and say what
you did.
:::

