<!--
  PARKED — not in _bookdown.yml, not built.

  "The average is not the distribution" was written for Unit 1 and removed
  deliberately. It is not an introduction to inference; it is the opening
  argument of distributional economics. It belongs wherever inequality,
  poverty, quantile methods or heterogeneous treatment effects are first
  discussed, where students will have something to attach it to.

  It expects `earn` and `workers` (data/wages-india-synthetic.csv) in scope.
-->

## The average is not the distribution {#u1-average}

::: {.question}
And what does an average hide?
:::

Before leaving this unit, one qualification deserves attention.

Almost all of inferential statistics concerns averages. Most of the methods in
this book estimate a mean, compare two means, or fit a line through conditional
means. That emphasis is not arbitrary: means are well behaved, their sampling
distributions are tractable, and both of the results we have just studied are
statements about them.

The mean, however, is only one feature of a distribution, and frequently not
the feature that a policy question concerns.

```{r u1-skew}
c(mean   = mean(earn),
  median = median(earn))

mean(earn) / median(earn)
```

The mean is more than **twice** the median. When those two diverge that
sharply, the average has stopped describing a typical worker.

```{r u1-skew-plot, echo = FALSE, fig.height = 4, fig.cap = "Annual earnings of 20,000 Indian workers. The mean sits far to the right of most people, dragged there by a thin tail of high earners."}
ggplot(workers, aes(annual_earnings / 1000)) +
  geom_histogram(bins = 60, fill = book_palette$fill, colour = "white",
                 linewidth = 0.15) +
  geom_vline(xintercept = median(earn) / 1000,
             colour = book_palette$ink, linetype = "dashed", linewidth = 0.7) +
  geom_vline(xintercept = mean(earn) / 1000,
             colour = book_palette$reject, linewidth = 0.9) +
  annotate("label", x = mean(earn) / 1000, y = Inf, vjust = 1.4,
           label = "mean", colour = book_palette$reject, fill = "white",
           size = 3.5, label.size = 0.3) +
  annotate("label", x = median(earn) / 1000, y = Inf, vjust = 3.2,
           label = "median", colour = book_palette$ink, fill = "white",
           size = 3.5, label.size = 0.3) +
  coord_cartesian(xlim = c(0, 400)) +
  labs(x = "annual earnings (₹ thousand)", y = "number of workers") +
  theme_book()
```

How concentrated is it?

```{r u1-concentration}
top10 <- earn[earn >= quantile(earn, 0.90)]

round(100 * sum(top10) / sum(earn))              # % of all earnings, top tenth
round(quantile(earn, 0.90) / median(earn), 1)    # top-decile threshold vs median
```

The best-paid tenth of workers take a large share of all earnings, and the
threshold for entering that tenth is several times the median wage.

::: {.intuition}
Consider what "average earnings in India" is able to answer.

For total earnings it is exact: the mean multiplied by the number of workers is
the total, by construction.

For the earnings of a typical worker it is misleading. Very few people earn
anything close to the mean, and most earn considerably less.

For whether a policy has helped it can be actively deceptive. If the gains
accrue at the top, the mean may rise while the majority are worse off. **An
average can move in the opposite direction to almost everyone within it.**
:::

::: {.keyidea}
Ask two questions of every average you meet, including your own:

1. **What is the spread?** A mean with no measure of dispersion is half a
   sentence.
2. **Who is where?** Means are the wrong tool for questions about the bottom of
   a distribution — poverty, vulnerability, who is hurt by a shock. Those are
   questions about *tails*, and the mean is precisely the statistic that
   averages tails away.
:::

This is not a philosophical aside; it changes what you estimate. Whole
literatures in development economics exist because the mean was the wrong
target. Poverty measures weight the poor rather than averaging over everyone.
Vulnerability measures ask who is at risk of falling below a line.
Distributional analysis asks whether a policy that raised average income raised
the income of anyone who was badly off to begin with.

We will return to this repeatedly. When we fit a regression we will be
modelling a conditional *mean*, and it will be worth asking each time what that
choice conceals. When we reach binary explanatory variables and estimate an
average treatment effect, the word "average" in that phrase will deserve every
bit of suspicion you can bring to it.

::: {.watchout}
None of this makes averages bad. They are the right tool for a great many
questions, and this book spends most of its length on them for good reason.

It means an average is an **answer to a specific question**, not a neutral
summary of the data. Decide which question you are asking before deciding the
mean answers it.
:::

