---
title: "From Chance to Decisions"
subtitle: "Statistics and Econometrics with R"
author: "Kedar Kulkarni"
date: "`r format(Sys.Date(), '%B %Y')`"
site: bookdown::bookdown_site
documentclass: book
description: |
  A first course in statistics and econometrics for Indian students, built on
  the progression from question to intuition to rigour to computation to
  interpretation.
url: https://kedarkulkarni1991.github.io/stats-econometrics-r/
github-repo: kedarkulkarni1991/stats-econometrics-r
---

# Welcome {-}

```{r setup-index, include = FALSE}
source("_common.R")
```

This book is the result of several years of teaching introductory statistics
and econometrics at Azim Premji University. It was shaped not only by lectures
and classroom discussions, but also by countless questions from students who
wanted to understand *why* statistical methods work, rather than simply *how*
to apply them.

Those questions made me rethink the way statistics is usually introduced.

Many introductory courses place considerable emphasis on mathematical
derivations, definitions and problem-solving techniques. These are important,
and this book does not seek to diminish their value. Yet, in my experience,
many students complete such courses having learned the mechanics of statistical
procedures without fully appreciating the logic that makes them work. They know
the steps involved in carrying out a hypothesis test or estimating a regression
model, but the underlying ideas — the beauty of reasoning from a sample to an
entire population — often remain hidden beneath the mathematics.

This book begins from a different premise. Before introducing notation,
formulas or proofs, it asks a simple question: **what problem are we trying to
solve?** Once that question is clear, intuition follows naturally. The
mathematics is then introduced not as something to be memorised, but as a
precise language for expressing ideas that the reader has already encountered
through examples, simulation and reasoning.

The philosophy of the book can be summarised in five words:

> **Question → Intuition → Rigour → Computation → Interpretation**

Every chapter follows this progression. We begin with a real question, usually
drawn from economics or public policy. We build intuition through examples and
simulation. We then develop the mathematical foundations carefully and without
unnecessary shortcuts. Finally, every idea is implemented in R using real data
so that the transition from theory to empirical analysis becomes seamless.

This integration of intuition, mathematical reasoning and programming reflects
the way statistics is practised today. There is no separate chapter devoted to
software. Programming is not an optional add-on; it is part of the process of
learning statistical thinking. Every table, figure and numerical result in this
book is generated directly from R, allowing readers to reproduce every analysis
for themselves.

The examples throughout the book are grounded, wherever possible, in the Indian
economy. Rather than relying on unfamiliar datasets and examples, we study
questions involving wages, education, poverty, agriculture, inflation,
employment, health and public policy. My hope is that students will not only
learn statistical methods but also develop a deeper understanding of the
economic and social issues that shape contemporary India.

This emphasis is particularly important because empirical economics has changed
dramatically over the past two decades. Modern applied research increasingly
seeks to answer causal questions rather than merely document statistical
relationships. Although this book develops the foundations of frequentist
statistics, estimation and regression analysis, it does so with this broader
objective in mind. The methods introduced here are intended not simply as
computational tools, but as the foundations upon which modern empirical
economics is built.

This book also draws extensively on the outstanding textbooks that have shaped
the teaching of statistics and econometrics over many decades. I have learned
enormously from those works, and this book should be viewed as a companion to
that tradition rather than a replacement for it. Its contribution lies not in
presenting new statistical methods, but in presenting familiar ideas through a
pedagogical approach that places intuition, conceptual understanding and
empirical application at its centre.

The book is written primarily for undergraduate students in India who are
studying quantitative social science — economics, but equally public policy,
sociology, political science, development studies and related disciplines.

I want to be explicit about one group of readers in particular. Many students
arrive in these programmes without a mathematics training in school, and find
statistics and econometrics genuinely difficult as a result. In my experience
the difficulty is almost never a lack of ability. It is that most textbooks are
written for readers who already have the background, and a student without it
is left trying to learn the ideas and the notation simultaneously, from a book
that explains only the second.

This book assumes no such background. Every idea is introduced before the
symbols for it, mathematical steps are shown rather than left to the reader, and
nothing is described as obvious. Motivated readers should be able to work
through it independently.

My hope is that readers finish the book not only knowing how to perform
statistical analyses in R, but also understanding the logic of frequentist
inference and developing the habits of mind needed to think critically about
empirical evidence.

I owe a considerable debt to my colleagues in economics at Azim Premji
University. Good pedagogy is treated here as a serious professional
undertaking rather than an afterthought, and the care that colleagues take —
day after day, and largely unremarked — to become better teachers and better
practitioners has been invaluable. That environment has done a great deal to
bring this project to light.

Finally, this book would never have been written without my students. Every
question asked after class, every thoughtful challenge to an explanation, every
mistake that forced me to rethink the way I taught, and every conversation
about data and economics has left its mark on these pages. I owe a sincere debt
of gratitude to all the students who have taken my courses in introductory
statistics and econometrics over the years. Your curiosity, patience and
willingness to ask difficult questions have continually pushed me to become a
better teacher. This book is the result of that shared journey, and I hope it
helps future students discover the same excitement for statistics that you
brought into my classroom.

## How to use this book {-}

Statistics is not a subject that rewards passive reading. Understanding comes
from asking questions, working through examples, making mistakes and revisiting
ideas. This book is written with that philosophy in mind, and you will get the
most out of it if you engage with it actively.

### Read the book in order {-}

The first few units are designed as a sequence rather than a collection of
independent topics. Ideas introduced early — such as variation, repeated
sampling and statistical inference — form the foundation for everything that
follows. While later chapters can be used as references, I strongly recommend
reading the book from beginning to end the first time.

### Work through the mathematics {-}

Do not skip the derivations because they look unfamiliar. Every mathematical
result is developed gradually, with intermediate steps shown. Try to work
through each derivation yourself before reading ahead. The objective is not to
memorise proofs, but to understand where formulas come from and why they work.

### Think in words before thinking in mathematics {-}

One piece of advice has stayed with me throughout my academic career. During my
PhD, my advisor once told me:

> If you can explain the problem clearly in simple English, you have already
> solved half of it. The mathematics is just a matter of practice.

Over the years I have found this to be remarkably true, and nowhere more so
than in statistics and econometrics.

Students often assume that success in these subjects depends primarily on being
good at mathematics. Mathematical reasoning is undoubtedly important, but most
difficulties arise much earlier than the algebra. They arise because we lose
sight of the question we are trying to answer. Before writing down a formula,
ask yourself: what am I trying to learn? What information do I have? Why should
this method work at all? If you can answer those questions clearly in words,
the mathematics usually becomes far easier to follow.

Throughout this book I encourage you to explain ideas to yourself in plain
English before expressing them in symbols. Statistical notation is a powerful
language, but it is still a language. Its purpose is to communicate ideas
precisely, not to replace logical reasoning.

Whenever you find yourself stuck in a derivation or confused by a formula,
pause and return to the underlying question. More often than not, the
mathematics becomes much less intimidating once the logic is clear.

### Write the code yourself {-}

Every analysis in this book is accompanied by R code. Resist the temptation to
copy and paste it. Typing the code yourself forces you to pay attention to
every line and every argument. Once the code works, experiment with it. Change
a parameter, modify the data, or deliberately introduce an error. Some of the
most valuable lessons come from understanding why something fails rather than
why it succeeds.

### Think before calculating {-}

Whenever a new method is introduced, pause before looking at the computation.
Ask yourself what you expect the answer to be, and why. Developing statistical
intuition is just as important as obtaining the correct numerical result.

### Attempt the exercises first {-}

The exercises are an essential part of the book, not an optional supplement.
Attempt every question before consulting the solutions, even if you are unsure
how to proceed. Struggling with a problem is often where the deepest learning
takes place. The solutions are intended to explain your reasoning, not replace
it.

### Use the book as a reference {-}

Once you have worked through the material, the book can also serve as a
reference. Use the search function at the top of the page to revisit concepts,
definitions, examples or R commands. You are not expected to remember every
formula or every line of code. What matters is understanding the underlying
ideas and knowing where to find them when needed.

### Learn by connecting ideas {-}

Throughout the book, try to ask not only how a method works, but why it was
introduced in the first place. Every chapter begins with a question and
develops the statistical tools needed to answer it. If you understand the
question, the mathematics that follows will feel much more natural.

Above all, remember that statistics is a way of thinking rather than a
collection of techniques. The goal of this book is not simply to help you
perform statistical analyses in R, but to help you reason carefully about
evidence, uncertainty and data. If you finish the book asking better questions
than when you began, then it will have achieved its purpose.

## A note on the data {-}

Every dataset used in this book is either built into R, supplied with the book,
or generated directly by code that you can read and run yourself. This means
that every table, figure and numerical result can be reproduced by running the
accompanying script from beginning to end.

Reproducibility is a fundamental principle of statistical work. If an analysis
cannot be recreated from the original data and the accompanying code, its
results cannot be independently verified. Good statistical practice therefore
means leaving behind a complete record of every step that produced the final
result.

::: {.keyidea}
A result that cannot be reproduced by running a script from top to bottom is
not yet a result.
:::

One dataset deserves a special mention. The class survey used throughout the
book was collected from students enrolled in this course. Before it was
included here, all identifying information was removed and each student was
assigned an anonymous identification number. Protecting the privacy of
participants is one of the first responsibilities of anyone working with data
about people.

Later in the book we return to questions of sampling, data collection and
research ethics. For now it is enough to remember that every dataset represents
real individuals, and good statistical practice begins with treating their
information responsibly.

## What you need {-}

This book requires two free programs, which should be installed in the
following order.

1. **R** — the statistical programming language, available from
   [cran.r-project.org](https://cran.r-project.org).
2. **RStudio Desktop** — an integrated development environment that provides a
   more convenient interface for working with R, available from
   [posit.co/download/rstudio-desktop](https://posit.co/download/rstudio-desktop).

Several chapters use additional R packages. These need to be installed only
once, by running the following command in the RStudio console:

```{r install-pkgs, eval = FALSE}
install.packages(c("tidyverse", "BSDA", "wooldridge", "AER",
                   "stargazer", "sandwich", "lmtest"))
```

No other software is required, and everything used in this book is freely
available.

## Mathematical background {-}

This book assumes a basic acquaintance with introductory probability and
descriptive statistics. Specifically, it will help to have met:

- basic probability rules and notation;
- random variables and their probability distributions;
- probability mass functions, probability density functions and cumulative
  distribution functions.

Everything else the book relies on is built up within it. Measures of central
tendency and dispersion are reviewed in Unit \@ref(towards-inference) before they are
used, and random variables, expected values and variances are developed there
from first principles rather than assumed. If those terms are unfamiliar, that
is where they are introduced.

The focus throughout is statistical inference and data analysis rather than
probability theory itself. Where an idea from probability is needed, it is
recalled briefly at the point of use.

A word to readers who are uncertain whether they have enough mathematics for
this. Almost everything in the list above is notation and vocabulary rather
than technique, and the difficulty of statistics lies elsewhere — in
understanding what a question is asking and what the data can answer. If you
can follow an argument carefully and are willing to work through the
derivations rather than skip them, you have what this book requires. Begin at
Unit \@ref(towards-inference) and consult a probability text only if something there
genuinely does not yield.

## Licence and citation {-}

This book is free, in both senses. The text, figures, exercises and datasets
are released under
[Creative Commons Attribution 4.0](https://creativecommons.org/licenses/by/4.0/),
and the code under the [MIT licence](https://opensource.org/licenses/MIT).

You may copy it, print it, translate it, remix it, teach from it, and build on
it — for any purpose, including commercially. The only condition is that you
credit the source.

```{r fix-errors, echo = FALSE, results = "asis"}
if (knitr::is_html_output()) cat(
"If you spot an error, there is a **pencil icon at the top of every page**.",
"It opens that chapter's source on GitHub, where you can fix the mistake and",
"send it back without installing anything.\n")
if (!knitr::is_html_output()) cat(
"If you spot an error, the source of every chapter is on GitHub and",
"corrections are welcome as pull requests.\n")
```

Corrections from students are genuinely welcome; several of the sharper points
in this book exist because someone asked why.

To cite:

> Kulkarni, K. (2026). *From Chance to Decisions: Statistics and
> Econometrics with R.*
> Azim Premji University.
> <https://kedarkulkarni1991.github.io/stats-econometrics-r/>
> Licensed under CC BY 4.0.
