---
title: "Statistics and Econometrics with R"
author: "Kedar Kulkarni"
date: "`r format(Sys.Date(), '%B %Y')`"
site: bookdown::bookdown_site
documentclass: book
description: |
  A first course in statistics and econometrics, where every idea is developed
  on paper and then immediately carried out in R.
url: https://kedarkulkarni1991.github.io/stats-econometrics-r/
github-repo: kedarkulkarni1991/stats-econometrics-r
---

# Welcome {-}

```{r setup-index, include = FALSE}
source("_common.R")
```

This book grew out of two courses I teach at Azim Premji University —
*Introduction to Statistics and Programming* and *Econometric Methods*.

In both courses I noticed the same thing. Students would learn a formula in one
session and learn the R command for it in another, and the two never quite met.
They could compute a $t$-statistic by hand and they could type `t.test()`, but
the connection between the two stayed fuzzy.

This book is my attempt to fix that. **There is no separate "R chapter."** Every
idea appears three times, in the same breath:

::: {.intuition}
First, in plain words — what are we actually trying to find out, and why would
anyone care?
:::

::: {.theory}
Then, on paper — the algebra, worked slowly, with every step shown. You should
be able to reproduce it in an exam with nothing but a pen.
:::

::: {.inr}
Then, in R — the same example, the same numbers, run as real code. Every result
you see printed in this book was computed when the page was built. If the code
were wrong, the number would be wrong, and I would notice.
:::

## How to use this book {-}

**Read it in order, at least the first time.** Chapters build on each other;
Part II assumes Part I.

**Type the code out yourself.** Copy-paste teaches you very little. Every code
block has a copy button in the top-right corner, but resist it on the first
pass. Change a number and see what breaks — that is where the learning is.

**Do the exercises before opening the solutions.** Solutions are hidden behind a
click, like this:

<details class="solution">
<summary>Show the solution</summary>

Try it first. Genuinely. The click will still be here in ten minutes.

</details>

**Use the search box.** The magnifying glass at the top-left searches the whole
book. If you have forgotten what a $p$-value is, that is faster than scrolling.

## What you need {-}

You need two free programs, installed in this order:

1. **R** — the language itself, from [cran.r-project.org](https://cran.r-project.org)
2. **RStudio Desktop** — a much friendlier window onto R, from
   [posit.co/download/rstudio-desktop](https://posit.co/download/rstudio-desktop)

A handful of chapters use extra packages. Install them once, by running this in
the RStudio console:

```{r install-pkgs, eval = FALSE}
install.packages(c("tidyverse", "BSDA", "wooldridge", "AER", "stargazer"))
```

You will not need anything else, and you will not need to pay for anything.

## Notation used throughout {-}

| Symbol | Means |
|:--|:--|
| $\mu,\ \sigma$ | population mean, population standard deviation |
| $\bar{x},\ s$ | sample mean, sample standard deviation |
| $n$ | sample size |
| $\hat{\beta}$ | an *estimate* of the parameter $\beta$ |
| $\alpha$ | significance level (the risk we accept of being wrong) |
| $H_0,\ H_A$ | null hypothesis, alternative hypothesis |

Greek letters are things about the **population** that we never observe. Roman
letters with hats or bars are things we **compute from a sample** in the hope
that they are close to the Greek ones. Almost all of statistics is the gap
between those two columns.

## A note on the data {-}

Every dataset used in this book is either built into R, shipped alongside the
book, or generated in front of you by code you can read. You will never see
`read.csv(file.choose())` in these pages — that command needs a human to point
at a file, which means nobody can reproduce your result, including you, six
months from now.

::: {.keyidea}
If a result cannot be reproduced by running a script from top to bottom, it is
not yet a result. It is a rumour.
:::
