---
title: "Statistics and Econometrics with R"
author: "Kedar Kulkarni"
date: "`r format(Sys.Date(), '%B %Y')`"
site: bookdown::bookdown_site
documentclass: book
description: |
  A first course in statistics and econometrics for Indian students, built on
  the progression from question to intuition to rigour to computation to
  interpretation.
url: https://kedarkulkarni1991.github.io/stats-econometrics-r/
github-repo: kedarkulkarni1991/stats-econometrics-r
---

# Welcome {-}

```{r setup-index, include = FALSE}
source("_common.R")
```

This book is the result of several years of teaching introductory statistics
and econometrics at Azim Premji University. It was shaped not only by lectures
and classroom discussions, but also by countless questions from students who
wanted to understand *why* statistical methods work, rather than simply *how*
to apply them.

Those questions made me rethink the way statistics is usually introduced.

Many introductory courses place considerable emphasis on mathematical
derivations, definitions and problem-solving techniques. These are important,
and this book does not seek to diminish their value. Yet, in my experience,
many students complete such courses having learned the mechanics of statistical
procedures without fully appreciating the logic that makes them work. They know
the steps involved in carrying out a hypothesis test or estimating a regression
model, but the underlying ideas — the beauty of reasoning from a sample to an
entire population — often remain hidden beneath the mathematics.

This book begins from a different premise. Before introducing notation,
formulas or proofs, it asks a simple question: **what problem are we trying to
solve?** Once that question is clear, intuition follows naturally. The
mathematics is then introduced not as something to be memorised, but as a
precise language for expressing ideas that the reader has already encountered
through examples, simulation and reasoning.

The philosophy of the book can be summarised in five words:

> **Question → Intuition → Rigour → Computation → Interpretation**

Every chapter follows this progression. We begin with a real question, usually
drawn from economics or public policy. We build intuition through examples and
simulation. We then develop the mathematical foundations carefully and without
unnecessary shortcuts. Finally, every idea is implemented in R using real data
so that the transition from theory to empirical analysis becomes seamless.

This integration of intuition, mathematical reasoning and programming reflects
the way statistics is practised today. There is no separate chapter devoted to
software. Programming is not an optional add-on; it is part of the process of
learning statistical thinking. Every table, figure and numerical result in this
book is generated directly from R, allowing readers to reproduce every analysis
for themselves.

The examples throughout the book are grounded, wherever possible, in the Indian
economy. Rather than relying on unfamiliar datasets and examples, we study
questions involving wages, education, poverty, agriculture, inflation,
employment, health and public policy. My hope is that students will not only
learn statistical methods but also develop a deeper understanding of the
economic and social issues that shape contemporary India.

This emphasis is particularly important because empirical economics has changed
dramatically over the past two decades. Modern applied research increasingly
seeks to answer causal questions rather than merely document statistical
relationships. Although this book develops the foundations of frequentist
statistics, estimation and regression analysis, it does so with this broader
objective in mind. The methods introduced here are intended not simply as
computational tools, but as the foundations upon which modern empirical
economics is built.

This book also draws extensively on the outstanding textbooks that have shaped
the teaching of statistics and econometrics over many decades. I have learned
enormously from those works, and this book should be viewed as a companion to
that tradition rather than a replacement for it. Its contribution lies not in
presenting new statistical methods, but in presenting familiar ideas through a
pedagogical approach that places intuition, conceptual understanding and
empirical application at its centre.

The book is intended primarily for undergraduate and postgraduate students of
economics, but it will also be useful to students of public policy, sociology,
political science, development studies and related disciplines. It assumes only
a modest mathematical background and is written so that motivated readers can
work through it independently. My hope is that readers finish the book not only
knowing how to perform statistical analyses in R, but also understanding the
logic of frequentist inference and developing the habits of mind needed to
think critically about empirical evidence.

Finally, this book would never have been written without my students. Every
question asked after class, every thoughtful challenge to an explanation, every
mistake that forced me to rethink the way I taught, and every conversation
about data and economics has left its mark on these pages. I owe a sincere debt
of gratitude to all the students who took my courses in introductory statistics
and econometrics over the years. Your curiosity, patience and enthusiasm
continually reminded me that good teaching begins with good questions. This
book is, in many ways, a product of those conversations, and I hope it serves
future students as well as you have served as my teachers.

## How to use this book {-}

**Read it in order, at least the first time.** Each unit builds on the one
before it, and later units assume the ideas developed earlier.

**Type the code out yourself.** Copy-paste teaches you very little. Change a
number and see what breaks — that is where the learning is.

**Do the exercises before reading the solutions.** They are worth nothing
unless you have already tried.

```{r how-to-web, echo = FALSE, results = "asis"}
if (knitr::is_html_output()) cat(
"**Use the search box.** The magnifying glass at the top-left searches the",
"whole book. If you have forgotten what a $p$-value is, that is faster than",
"scrolling.\n")
```

## What you need {-}

You need two free programs, installed in this order:

1. **R** — the language itself, from [cran.r-project.org](https://cran.r-project.org)
2. **RStudio Desktop** — a much friendlier window onto R, from
   [posit.co/download/rstudio-desktop](https://posit.co/download/rstudio-desktop)

A handful of chapters use extra packages. Install them once, by running this in
the RStudio console:

```{r install-pkgs, eval = FALSE}
install.packages(c("tidyverse", "BSDA", "wooldridge", "AER", "stargazer"))
```

You will not need anything else, and you will not need to pay for anything.

## Notation used throughout {-}

| Symbol | Means |
|:--|:--|
| $\mu,\ \sigma$ | population mean, population standard deviation |
| $\bar{x},\ s$ | sample mean, sample standard deviation |
| $n$ | sample size |
| $E[X]$ | the long-run average of a random quantity |
| $\alpha$ | significance level (the risk we accept of being wrong) |
| $H_0,\ H_A$ | null hypothesis, alternative hypothesis |

Greek letters are things about the **population** that we never observe. Roman
letters with hats or bars are things we **compute from a sample** in the hope
that they are close to the Greek ones. Almost all of statistics is the gap
between those two columns.

## A note on the data {-}

Every dataset used in this book is either built into R, shipped alongside the
book, or generated in front of you by code you can read. You will never see
`read.csv(file.choose())` in these pages — that command needs a human to point
at a file, which means nobody can reproduce your result, including you, six
months from now.

::: {.keyidea}
If a result cannot be reproduced by running a script from top to bottom, it is
not yet a result. It is a rumour.
:::

One dataset deserves a word. The class survey was collected from students in
this course, and originally carried their names. Those were replaced with
anonymous ids before this book was ever published — which is itself the first
lesson of working with data about people. You will meet the same idea again
when we discuss what a sample is, and what you owe the people in it.

## Licence and citation {-}

This book is free, in both senses. The text, figures, exercises and datasets
are released under
[Creative Commons Attribution 4.0](https://creativecommons.org/licenses/by/4.0/),
and the code under the [MIT licence](https://opensource.org/licenses/MIT).

You may copy it, print it, translate it, remix it, teach from it, and build on
it — for any purpose, including commercially. The only condition is that you
credit the source.

```{r fix-errors, echo = FALSE, results = "asis"}
if (knitr::is_html_output()) cat(
"If you spot an error, there is a **pencil icon at the top of every page**.",
"It opens that chapter's source on GitHub, where you can fix the mistake and",
"send it back without installing anything.\n")
if (!knitr::is_html_output()) cat(
"If you spot an error, the source of every chapter is on GitHub and",
"corrections are welcome as pull requests.\n")
```

Corrections from students are genuinely welcome; several of the sharper points
in this book exist because someone asked why.

To cite:

> Kulkarni, K. (2026). *Statistics and Econometrics with R.*
> Azim Premji University.
> <https://kedarkulkarni1991.github.io/stats-econometrics-r/>
> Licensed under CC BY 4.0.
